### 2020-01-16 [2017-12-15]_HPC_(variance_models)/Presentation/presentation.tex
Created paper repository with all of my work.

% Template for overheads, landscape, fancy VT footlines.
%
\magnification=1800 \rightskip=0pt plus 5em
\hsize=9.0truein \vsize=6.5truein
\parindent=0pt \parskip=0pt plus 4truept
\baselineskip=21truept plus 3truept minus 2truept \lineskiplimit=3truept
\lineskip=3truept plus 3truept \tolerance=1000

%fonts
\font\helvr=phvr at 18truept
\font\helvb=phvb at 18truept
\font\timesi=ptmri at 18truept
\font\helvbXX=phvb at 20truept
\font\helvbXXIV=phvb at 24truept
\font\newfont=pbkli

%% Math blackboard fonts {\bb}
\font\tenamsb=msbm10 \font\sevenamsb=msbm7 \font\fiveamsb=msbm5
\newfam\bbfam
\textfont\bbfam=\tenamsb
\scriptfont\bbfam=\sevenamsb
\scriptscriptfont\bbfam=\fiveamsb

\def\bb{\fam\bbfam}

\def\rm{\fam=0\helvr} \let\bf=\helvb \let\tt=\helvb \let\it=\timesi
\let\sl=\timesi

%definitions
\def\cl#1{\centerline{#1}}
\def\head#1{\centerline{\helvbXXIV#1}}
{\obeyspaces\global\let =\ }
\def\verbatim{\parindent=0pt\tt\obeylines\obeyspaces}
\def\l{\hfill\&}        
\def\shift{\hsize=8.5truein\parindent=25truept\indent}
\def\bull{\par\hangindent=20pt\hangafter=1
\indent\hbox to 20pt{\hfil$\bullet$ }\ignorespaces}

% for spacing in tables
\newdimen\digitwidth \setbox0=\hbox{\rm0} \digitwidth=\wd0

\input epsf % For Postscript figures:{\epsfxsize=  \epsffile{}}
\helvr

%**********************************  Page 1 ******************************
\cl{\helvbXXIV PREDICTIVE MODELING OF I/O CHARACTERISTICS}\smallskip
\cl{\helvbXXIV IN HIGH PERFORMANCE COMPUTING SYSTEMS}
\bigskip
\bigskip
\cl{\bf Thomas Lux$^1$, Layne Watson$^{123}$, Tyler Chang$^1$,}
\cl{\bf Jon Bernard$^1$, Bo Li$^1$, Li Xu$^4$, Godmar Back$^1$,}
\cl{\bf Ali Butt$^1$, Kirk Cameron$^1$, and Yili Hong$^4$}
\bigskip
\bigskip
\cl{Departments of Computer Science$^1$, Mathematics$^2$,}
\cl{Aerospace \& Ocean Engineering$^3$, and Statistics$^4$}
\bigskip
\cl{Virginia Polytechnic Institute and State University}
\cl{Blacksburg, VA 24061-0106  USA} 
\vfil
\vtop to 0pt{\vskip -30truept
\cl{\epsfxsize=1.3in \epsffile{VT_Logo.eps}}\vss}
\footline={\hfil}
\eject

%% Redefine footline for the rest of the presentation

\footline={\fiverm\folio\quad\leaders\hrule
height3truept\hfil\quad \epsfxsize=.8in \epsffile{VT_Logo_CompSci.eps}}

%**********************************  Page 2 ******************************
\head{Problem Setting}
\bigskip \bigskip
When the same computer is used to execute the same program repeatedly, most measurable performance characteristics will vary. This paper will consider the execution of IOzone, a file I/O benchmark for HPC systems.
\cl{\epsfxsize=5in \epsffile{Raw_Throughput.eps}}
%% \cl{\hbox{\epsfxsize=30truepc\epsffile{triangleplane.eps}}}
This variance in performance can make it difficult to identify an optimal configuration that meets minimum requirements.
\vfil
\eject

%**********************************  Page 3 ******************************
\head{Proposed Approach}
\bigskip \bigskip
Use multivariate interpolation and regression to model the performance of a computer with respect to the system and application configuration.
%% \bull Bulletted text..
\bigskip\bigskip
Multivariate {\it interpolation} and {\it regression} are defined when there exists $f:{\bb R}^d \rightarrow {\bb R}$ and a set $X$ of $n$ points in ${\bb R}^d$ along with response values $f(x)$ for all $x \in X$.
\bigskip\bigskip
{\bf Interpolation} constructs an approximation $\hat f: {\bb R}^d \rightarrow {\bb R}$ such that $\hat f(x) = f(x)$ for all $x \in X$. The form of the true underlying function $f$ is often unknown, however it is still desirable to construct an approximation $\hat f$ with minimum approximation error at $y \notin X$.
\bigskip\bigskip
{\bf Regression} relaxes the conditions of interpolation by minimizing the error in $\hat f$ at $x \in X$ while maintaining some parametric form with parameters $P$. This can be written as $\min_{P} ||\hat f(X) - f(X)||$, where $f(X)$ is a vector of $f(x)$ for all $x \in X$ and $||\cdot||$ is an appropriate measure.
\vfil
\eject

%**********************************  Page 4 ******************************
\head{Regression Algorithms}
\bigskip
{\bf Multivariate Adaptive Regression Splines} (MARS)
is an iterative approximation of the form
$$ B_{2s-1}(x) = B_l(x) [c(x_i-v)]_+ ,$$
$$ B_{2s}(x) = B_k(x) [c(x_i-v)]_- ,$$
where $s$ is the iteration number, $B_l(x)$ and $B_k(x)$ are basis
functions from the previous iteration, $c, v \in {\bb R}$, 
$$w_+ = \cases{ w, & $w \geq 0$ \cr 0, & $w < 0$},$$
and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit.
\medskip
In effect, MARS creates a locally component-wise tensor
product approximation of the data. The computational
complexity of MARS is ${\cal O}(n d m^3)$ where $m$ is the
maximum number of underlying basis functions.
\vfil\eject

%**********************************  Page 4 ******************************
\head{Regression Algorithms}
\bigskip
{\bf Multilayer Perceptron Regressor} (MLP Regressor)
$$ l(u) = \big( u^t W_l \big)_+ ,$$
where $W_l$ is the $i$ by $j$ weight matrix for layer $l$.
\medskip
The multilayer perceptron (MLP) produces a piecewise linear model of the input data. The computational complexity is ${\cal O}(n d m)$, where $m$ is
determined by the sizes of the layers of the network and the stopping
criterion of the error minimizer.
\bigskip
{\bf Support Vector Regressor} (SVR)
$$ p(x)  = \sum_{i=1}^{n}a_i K(x,x^{(i)}) + b ,$$
where $K$ is the selected kernel function, $a \in {\bb R}^n$, $b \in {\bb R}$ are coefficients to be solved for simultaneously. The computational complexity of the SVR is ${\cal O}(n^2dm)$, with $m$ being determined by the minimization convergence criterion.
\vfil\eject

%**********************************  Page 5 ******************************
\head{Interpolation Algorithms}
\bigskip
{\bf Delaunay} constructs a simplicial mesh such that for a $d$-simplex S with vertices $v^{(0)}$, $v^{(1)}$, $\ldots$, $v^{(d)}$, $x \in S$, and data values $f(v^{(i)})$, $i=0$, $\ldots$, $d$, $x$ is a unique convex combination of the vertices, and the interpolant is given by
$$ p(x) = \sum_{i=0}^{d} w_i f(v^{(i)}), $$
where $w_i$ are the convex weights. The computational complexity of the Delaunay triangulation (for the implementation used here) is ${\cal O}(n^{\lceil d/2 \rceil})$.
\bigskip
{\bf Linear Shepard} (LSHEP) blends local linear interpolants and has the form
$$ p(x) = {\sum_{k=1}^{n}W_k(x)P_k(x) \over \sum_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is a locally supported weighting function and $P_k(x)$ is a local linear approximation to the data satisfying $P_k\big(x^{(k)}\big) = f\big(x^{(x)}\big)$. The computational complexity of LSHEP is ${\cal O}(n^2d^3)$.
\vfil \eject

%**********************************  Page 6 ******************************
\head{IOzone Data}
\bigskip \bigskip
This experiment attempts to model {\it throughput variance} as
a function of application and system {\it parameters}.
\medskip
To do so, the {\it IOZone} benchmark is used to read files of varying sizes
on a homogeneous system. 
The variance in the throughput for each read is modelled as a function of
several parameters.
\medskip
The parameters chosen and the values used for them are in the table below:
\smallskip
\offinterlineskip \tabskip=0pt
\def\filler{height2pt&\omit&&\omit&\cr}
\def\tablerule{\noalign{\hrule}}
\centerline{\vbox{
\halign{&\vrule#&\strut\hskip 4pt\hfil#\hskip 4pt\cr
\tablerule\filler
& Parameters \hfill&& Values \hfill&\cr
\filler\tablerule\filler
&\omit\hskip 4pt\hbox{file size being read (in KB)}
&&64, 256, 1024\hfill&\cr
&\omit\hskip 4pt\hbox{record size of file system (in KB)}
&&32, 128, 512\hfill&\cr
&\omit\hskip 4pt\hbox{number of threads for IOZone reader}
&&1, 2, 4, 8, 16, 32, 64, 128, 256\hfill&\cr
&\omit\hskip 4pt\hbox{CPU frequency (in GHz)}
&&1.2, 1.4, 1.5, 1.6, 1.8, 1.9, 2.0, 2.1, 2.3, &\cr
&\hfill
&& 2.4, 2.5, 2.7, 2.8, 2.9, 3.0, 3.001 \hfil&\cr
\filler\tablerule}}}
\medskip
For each combination of the above parameters, 40 runs of IOzone were done and 
the observed throughput variance was computed using:
$$ \sigma^2 = \bigg(\sum_{i=1}^{40}(t_i - \mu)^2 \bigg) / 39 $$
where $t_i$ is the $i$th observed throughput and $\mu$ is the observed
mean over all 40 runs.
\vfil\eject

%**********************************  Page 7 ******************************
\head{Analyzing Predictability: Multi Dimensional Analysis}
\bigskip\bigskip

\item{1.} For all $k = 1$, $\ldots$, $d$ and for all nonempty subsets $F \subset \{1, 2, \ldots, d\}$, reduce the input data to points $(z, f_F(z))$ with $z \in {\bb R}^k$ and $f_F(z) = E\bigl[ \bigl\{ f\bigl(x^{(i)}\bigr) \bigm| \bigl(x^{(i)}_F = z\bigr) \bigr\} \bigr]$, where $E[\cdot]$ denotes the mean and $x^{(i)}_F$ is the subvector of $x^{(i)}$ indexed by $F$.
\bigskip
\item{2.} For all $r$ in $\{5, 10, \ldots, 95\}$, generate $N$ random splits $(train, test)$ of the reduced data with $r$ percentage for training and $100 - r$ percentage for testing.
\bigskip
\item{3.} When generating each of $N$ random $(train, test)$ splits, ensure that all points from $test$ are in the convex hull of points in $train$ (to prevent extrapolation); also ensure that the points in $train$ are well spaced.
\bigskip\bigskip

In order to ensure that training points are well spaced, a statistical method for picking points is used
\bigskip
\itemitem{1.} Generate a sequence of all pairs of points $\bigl(z^{(i_1)},z^{(j_1)}\bigr), \bigl(z^{(i_2)},z^{(j_2)}\bigr), \ldots$ sorted by ascending pairwise Euclidean distance between points, so that $\bigl|\bigl|z^{(i_k)}-z^{(j_k)}\bigr|\bigr|_2 \leq \bigl|\bigl|z^{(i_{k+1})}-z^{(j_{k+1})}\bigr|\bigr|_2$.
\bigskip
\itemitem{2.} Sequentially remove points from candidacy until only $|train|$ remain by randomly selecting one point from the pair $\bigl(z^{(i_m)}, z^{(j_m)}\bigr)$ for $m = 1,\ldots$ if both $z^{(i_m)}$ and $z^{(j_m)}$ are still candidates for removal.

\vfil\eject

%**********************************  Page 8 ******************************
\head{Increasing Samples: Results for Predicting Throughput Mean}
\bigskip\bigskip\bigskip
\cl{\epsfxsize=5in \epsffile{Mean_TT_Ratio.eps}}
\vfil\eject

\head{Increasing Samples: Results for Predicting Throughput Variance}
\bigskip\bigskip\bigskip
\cl{\epsfxsize=5in \epsffile{Var_TT_Ratio.eps}}
\vfil\eject

%**********************************  Page 10 ******************************
\head{Increasing Dimension: Results for Predicting Throughput Mean}
\bigskip\bigskip\bigskip
\cl{\epsfxsize=5in \epsffile{Mean_Dim.eps}}
\vfil\eject

\head{Increasing Dimension: Results for Predicting Throughput Variance}
\bigskip\bigskip\bigskip
\cl{\epsfxsize=5in \epsffile{Var_Dim.eps}}
\vfil\eject

%**********************************  Page 11 ******************************
\head{Conclusion and Future Work}
\bigskip\bigskip\bigskip
Multivariate models of HPC system performance can effectively predict
I/O throughput mean and variance. These multivariate techniques
significantly expand the scope and portability of statistical models
for predicting computer system performance over previous work.
\bigskip\bigskip\bigskip
Delaunay model can make predictions for 821 system configurations with less than 5\% error when trained on only 43 configurations.
\bigskip\bigskip\bigskip
Multivariate methods should be applied to predict distributions instead of summary statistics. This would allow for much deeper insight.
\vfil\eject

%**********************************  Page 12 ******************************
\head{Acknowledgements}
\bigskip \bigskip
The data shown here was collected for the VarSys project at Virginia Tech.
\vfil\eject

\bye

### 2020-01-16 [2017-12-15]_HPC_(variance_models)/paper.tex
Created paper repository with all of my work.

%**************************************************************************
%* SpringSim 2017 Author Kit
%*
%* Word Processing System: TeXnicCenter and MiKTeX
%*
%**************************************************************************

\documentclass{scspaperproc}

\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{mathptmx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{amsthm}
%% \usepackage{color}   % Needed to color text for drafts

%% \usepackage{algorithm}     % needed for the algorithm block
%% \usepackage{algpseudocode} % algorithmicx psuedo-code styling
%\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
%% anchorcolor=black,linkcolor=black]{hyperref}

% custom hyphenation rules
\usepackage{hyphenat}
\hyphenation{op-tical net-works semi-conduc-tor}

% theorem style
\newtheoremstyle{scsthe}% hnamei
{8pt}% hSpace abovei
{8pt}% hSpace belowi
{\it}% hBody fonti
{}% hIndent amounti1
{\bf}% hTheorem head fontbf
{.}% hPunctuation after theorem headi
{.5em}% hSpace after theorem headi2
{}% hTheorem head spec (can be left empty, meaning `normal')i
\theoremstyle{scsthe}
\newtheorem{theorem}{Theorem}
\renewcommand{\thetheorem}{\arabic{theorem}}
\newtheorem{corollary}[theorem]{Corollary}
\renewcommand{\thecorollary}{\arabic{corollary}}
\newtheorem{definition}{Definition}
\renewcommand{\thedefinition}{\arabic{definition}}
\newcommand{\rpm}{\raisebox{.2ex}{$\scriptstyle\pm$}}

% avoid overrunning the right margin
\sloppy

%% ***** NOTE *****

%% The use of the long citation format (e.g. "Brown and Edwards
%% (1993)" rather than "[5]") and at the same time using the hyperref
%% package can lead to hard to trace bugs in case the citation is
%% broken accross the line (usually this will mark the entire
%% paragraph as a hyperlink (clickable) which is easily noticeable and
%% fixed if using colorlinks, but not if the color is black -- as it
%% is now). Worse yet, if a citation spans page boundary, LaTeX
%% compilation can fail, with an obscure error message. Since this
%% depends a lot on the flow of the text and wording, these bugs come
%% and go and can be extremely hard for a beginner to trace. The error
%% message can look like this:
%%
%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting
%%    level than \pdfstartlink.  \AtBegShi@Output ...ipout \box
%%    \AtBeginShipoutBox \fi \fi
%%    l.174 
%%    ! ==> Fatal error occurred, no output PDF file produced!
%%
%% and can be universally fixed by putting an \mbox{} around the
%% citation in question (in this case, at line 174) and maybe adapting
%% the wording a little bit to improve the paragraph typesetting,
%% which is perhaps not immediately obvious.
%****************************************************************************

% begin document
\begin{document}

% Page header (author list)
\SCSpagesetup{Lux, Watson, Chang, Bernard, Li, Xu, Back, Butt, Cameron, and Hong}

% Conference info
\def\SCSconferenceacro{SpringSim}
\def\SCSpublicationyear{2018}
\def\SCSconferencedates{April 15-18}
\def\SCSconferencevenue{Baltimore, MD, USA}
\def\SCSsymposiumacro{HPC} % High Performance Computing Symposium

% title
\title{Predictive Modeling of I/O Characteristics \\ in High
  Performance Computing Systems}

% AUTHOR LIST
% *** NOTE: May need to adjust titlevboxsize in the preamble
\author{Thomas C. H. Lux \\ [8pt]
Dept. of Computer Science \\
Virginia Polytechnic Institute\\
\& State University (VPI\&SU) \\
Blacksburg, VA 24061 \\
tchlux@vt.edu \\
\and
Layne T. Watson \\[8pt]
Dept. of Computer Science\\
Dept. of Mathematics\\
Dept. of Aerospace \& Ocean Eng.\\ 
VPI \& SU \\
\and
\vspace{-30pt}\\
Tyler H. Chang\\ [8pt]
Dept. of Computer Science \\
VPI \& SU \\
}

\maketitle

\vspace{-20pt}

Additional authors are listed in the Additional Authors section.

\section*{Abstract}

Each of high performance computing, cloud computing, and computer
security have their own interests in modeling and predicting the
performance of computers with respect to how they are configured. An
effective model might infer internal mechanics, minimize power
consumption, or maximize computational throughput of a given
system. This paper analyzes a four-dimensional dataset measuring the
input/output (I/O) characteristics of a cluster of identical computers
using the benchmark IOzone. The I/O performance characteristics are
modeled with respect to system configuration using multivariate
interpolation and approximation techniques. The analysis reveals that
accurate models of I/O characteristics for a computer system may be
created from a small fraction of possible configurations, and that
some modeling techniques will continue to perform well as the number
of system parameters being modeled increases. These results have
strong implications for future predictive analyses based on more
comprehensive sets of system parameters.

\textbf{Keywords:} Regression, approximation, interpolation,
performance modeling

%     Introduction     
%======================
\section{Introduction and related work}
\label{sec:introduction}

Performance tuning is often an experimentally complex and time
intensive chore necessary for configuring HPC systems. The procedures
for this tuning vary largely from system to system and are often
subjectively guided by the system engineer(s). Once a desired level of
performance is achieved, an HPC system may only be incrementally
reconfigured as required by updates or specific jobs. In the case that
a system has changing workloads or nonstationary performance
objectives that range from maximizing computational throughput to
minimizing power consumption and system variability, it becomes clear
that a more effective and automated tool is needed for configuring
systems. This scenario presents a challenging and important
application of multivariate approximation and interpolation
techniques.

Predicting the performance of an HPC system is a challenging problem
that is primarily attempted in one of two ways: (1) build a
statistical model of the performance by running experiments on the
system at select settings, or (2) run artificial experiments using a
simplified simulation of the target system to estimate architecture
and application bottlenecks. In this paper the proposed multivariate
modeling techniques rest in the first category, and they represent a
notable increase in the ability to model complex interactions between
system parameters.

Many previous works attempting to model system performance have used
simulated environments to estimate the performance of a system
\shortcite{grobelny2007fase,wang2009simulation,wang2013towards}. Some
of these works refer to statistical models as being oversimplified and
not capable of capturing the true complexity of the underlying
system. This claim is partially correct, noting that a large portion
of predictive statistical models rely on simplifying the model to one
or two parameters
\shortcite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
These limited statistical models have provided satisfactory
performance in very narrow application settings. Many of the
aforementioned statistical modeling techniques claim to generalize,
while simultaneously requiring additional code annotations, hardware
abstractions, or additional application level understandings in order
to generate models. The approach presented here requires no
modifications of the application, no architectural abstractions, nor
any structural descriptions of the input data being modeled. The
techniques used are purely mathematical and only need performance data
as input.

Among the statistical models presented in prior works,
\shortciteN{bailey2005performance} specifically mention that it is
difficult for the simplified models to capture variability introduced
by I/O. System variability in general has become a problem of
increasing interest to the HPC and systems communities, however most
of the work has focused on operating system (OS) induced variability
\shortcite{beckman2008benchmarking,de2007identifying}. The work that
has focused on managing I/O variability does not use any sophisticated
modeling techniques \shortcite{lofstead2010managing}. Hence, this
paper presents a case study applying advanced mathematical and
statistical modeling techniques to the domain of HPC I/O
characteristics. The models are used to predict the mean throughput of
a system and the variance in throughput of a system. The discussion
section outlines how the techniques presented can be applied to any
performance metric and any system.

%% Multivariate models for an HPC system would be a function of the
%% tunable parameters built to accurately model some desired performance
%% metric.

%% \begin{enumerate}
%% \item The value of multivariate Modeling
%% \item The data context
%% \item The proposed method for using multivariate models
%% \item The impact of effective models
%% \end{enumerate}

In general, this paper compares five multivariate approximation
techniques that operate on inputs in $\mathbb{R}^d$ ($d$-tuples of
real numbers) and produce predicted responses in $\mathbb{R}$. In
order to provide coverage of the varied mathematical strategies that
can be employed to solve the continuous modeling problem, three of the
techniques are regression based and the remaining two are
interpolants. The sections below outline the mathematical formulation
of each technique and provide computational complexity bounds with
respect to the size (number of points and dimension) of input
data. Throughout, $d$ will refer to the dimension of the input data,
$n$ is the number of points in the input data, $x^{(i)} \in
\mathbb{R}^d$ is the $i$-th input data point, $x^{(i)}_j$ is the
$j$-th component of $x^{(i)}$, and $f(x^{(i)})$ is the response value
of the $i$-th input data point.

The remainder of the paper is broken up into five major parts. Section
\ref{sec:multivariate} provides an overview of the multivariate
modeling techniques, Section \ref{sec:methodology} outlines the
methodology for comparing and evaluating the performance of the
models, Section \ref{sec:results} presents the IOzone predictions,
Section \ref{sec:discussion} discusses the obvious and subtle
implications of the models' performance, and Section
\ref{sec:conclusion} concludes and offers directions for future work.

\section{Multivariate Models}
\label{sec:multivariate}

\subsection{Regression}
\vspace{-10pt}
Multivariate approximations are capable of accurately modeling a
complex dependence of a response (in $\mathbb{R}$) on multiple
variables (represented as a points in $\mathbb{R}^{d}$). The
approximations to some (unknown) underlying function $f: \mathbb{R}^d
\rightarrow \mathbb{R}$ are chosen to minimize some error measure
related to data samples $f(x^{(i)})$. For example, least squares
regression uses the sum of squared differences between modeled
response values and true response values as an error measure.

\vspace{-10pt}
\subsubsection{Multivariate Adaptive Regression Splines}
\vspace{-10pt}
This approximation was introduced in
\shortciteN{friedman1991multivariate} and subsequently improved to its
current version in \shortciteN{stanford1993fast}, called fast
multivariate adaptive regression splines (Fast MARS). In Fast MARS, a
least squares fit model is iteratively built by beginning with a
single constant valued function and adding two new basis functions at
each iteration of the form

$$ B_{2s-1}(x) = B_l(x) [c(x_i-v)]_+ ,$$
$$ B_{2s}(x) = B_k(x) [c(x_i-v)]_- ,$$

where $s$ is the iteration number, $B_l(x)$ and $B_k(x)$ are basis
functions from the previous iteration, $c, v \in \mathbb{R}$,

$$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$

and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit. In effect, MARS creates a locally component-wise tensor
product approximation of the data. The overall computational
complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
maximum number of underlying basis functions. This paper uses an
implementation of Fast MARS \shortcite{rudy2017pyearth} with $m =
200$.

\vspace{-10pt}
\subsubsection{Multilayer Perceptron Regressor}
\vspace{-10pt}
The neural network is a well studied and widely used method for both
regression and classification tasks
\shortcite{hornik1989multilayer}. When using the rectified linear unit
(ReLU) activation function \shortcite{dahl2013improving} and training
with the BFGS minimization technique \shortcite{moller1993scaled}, the
model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
\rightarrow \mathbb{R}^{j}$ defined by

$$ l(u) = \big( u^t W_l \big)_+ ,$$

where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this
form, the multilayer perceptron (MLP) produces a piecewise linear
model of the input data. The computational complexity of training a
multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is
determined by the sizes of the layers of the network and the stopping
criterion of the BFGS minimization used for finding weights. This
paper uses the scikit-learn MLP regressor \shortcite{scikit-learn}, a
single hidden layer with 100 nodes, ReLU activation, and BFGS for
training.

\subsubsection{Support Vector Regressor}
\vspace{-10pt}
Support vector machines are a common method used in machine learning
classification tasks that can be adapted for the purpose of regression
\shortcite{basak2007support}. How to build a support vector regressor
(SVR) is beyond the scope of this summary, but the resulting
functional fit $p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form

$$ p(x)  = \sum_{i=1}^{n}a_i K(x,x^{(i)}) + b ,$$

%% $$ \text{Minimize } \bigg\{ \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}
%% a_i a_j K(x_i, x_j) + \epsilon \sum_{i=1}^{n} a_i - \sum_{i=1}^{n} y_i
%% a_i \bigg\} $$

%% $$ \text{Subject to } \sum_{i=1}^{n}a_i = 0 \text{ } \text{ and }
%% \text{ } a_i \in [0,C] $$

where $K$ is the selected kernel function, $a \in \mathbb{R}^n$, $b
\in \mathbb{R}$ are coefficients to be solved for simultaneously. The
computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$
being determined by the minimization convergence criterion. This paper
uses the scikit-learn SVR \shortcite{scikit-learn} with a polynomial
kernel function.

\vspace{-10pt}
\subsection{Interpolation}
\vspace{-10pt}
In some cases it is desirable to have a model that can recreate the
input data exactly. This is especially the case when the confidence in
the response values for known data is high. Both interpolation models
analyzed in this paper are based on linear functions.

\vspace{-10pt}
\subsubsection{Delaunay}
\vspace{-10pt}
The Delaunay method of interpolation is a well studied geometric
technique for producing an interpolant \shortcite{lee1980two}. The
Delaunay triangulation of a set of data points into simplices is such
that the sphere defined by the vertices of each simplex contains no
data points in the sphere's interior. For a $d$-simplex S with
vertices $v^{(0)}$, $v^{(1)}$, $\ldots$, $v^{(d)}$, $x \in S$, and
data values $f(v^{(i)})$, $i=0$, $\ldots$, $d$, $x$ is a unique convex
combination of the vertices,

$$ x = \sum_{i=0}^{d} w_i v^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad
w_i \geq 0, \quad i=0,\ldots,d ,$$

and the Delaunay interpolant to $f$ at $x$ is given by

$$ p(x) = \sum_{i=0}^{d} w_i f(v^{(i)}). $$

The computational complexity of the Delaunay triangulation (for the
implementation used here) is $\mathcal{O}(n^{\lceil d/2 \rceil})$,
which is not scalable to $d > 10$ \shortcite{sartipizadeh2016computing}.
The scipy interface \shortcite{scipy} to the QuickHull implementation
\shortcite{barber1996qhull} of the Delaunay triangulation is used here.

\subsubsection{Linear Shepard}
\vspace{-10pt}
The linear Shepard method (LSHEP) is a blending function using local
linear interpolants, a special case of the general Shepard algorithm
\shortcite{thacker2010algorithm}. The interpolant has the form

$$ p(x) = \frac{\sum_{k=1}^{n}W_k(x)P_k(x)}{\sum_{k=1}^{n}W_k(x)} ,$$

where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
is a local linear approximation to the data satisfying
$P_k\big(x^{(k)}\big) = f\big(x^{(x)}\big)$. The computational
complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
Fortran\#95 implementation of LSHEP in SHEPPACK
\shortcite{thacker2010algorithm}.

%% %     Related Work     
%% %======================
%% \section{Related Work}
%% \begin{enumerate}
%% \item Not sure how much to include here? Shooting for thoroughness or
%%   simply necessary coverage? How much background should I expect the
%%   readers of this paper to have in the ``multivariate modeling of
%%   systems'' area?
%% \end{enumerate}

%     Methodology     
%=====================
\section{Methodology}
\label{sec:methodology}
\subsection{Data}
\vspace{-10pt}

In order to evaluate the viability of multivariate models for
predicting system performance, this paper presents a case study of a
four-dimensional dataset produced by executing the IOzone benchmark
from \shortciteN{iozone} on a homogeneous cluster of computers. All
experiments were performed on parallel shared-memory nodes common to
HPC systems. Each system had a lone guest Linux operating system
(Ubuntu 14.04 LTS//XEN 4.0) on a dedicated 2TB HDD on a 2 socket, 4
core (2 hyperthreads per core) Intel Xeon E5-2623 v3 (Haswell)
platform with 32 GB DDR4. The system performance data was collected by
executing IOzone 40 times for each of a select set of system
configurations. A single IOzone execution reports the max I/O
throughput seen for the selected test in kilobytes per second. The 40
executions for each system configuration are converted into the mean
and variance, both values in $\mathbb{R}$ capable of being modeled
individually by the multivariate approximation techniques presented in
Section \ref{sec:multivariate}. The summary of data used in the
experiments for this paper can be seen in Table \ref{tab:data_type}.
Distributions of raw throughput values being modeled can be seen in
Figure \ref{fig:raw_throughput}.

\begin{table}
  \centering
  \begin{tabular}{c|c}
    \hline
    \textbf{System Parameter} & \textbf{Values}\\
    \hline
    File Size & 64, 256, 1024\\
    Record Size & 32, 128, 512\\
    Thread Count & 1, 2, 4, 8, 16, 32, 64, 128, 256\\
    Frequency & \{12, 14, 15, 16, 18, 19, 20, 21, 23, 24, 25, 27, 28, 29, 30, 30.01\} $\times 10^5$\\
    \hline
    \textbf{Response Values} & \\
    \hline
    Throughput Mean & [$2.6 \times 10^5$, $5.9 \times 10^8$]\\
    Throughput Variance & [$5.9\times 10^{10} $, $4.7 \times 10^{16}$]\\
    \hline
  \end{tabular}
  \caption{A description of the system parameters being considered in
    the IOzone tests. Record size must not be greater than file size
    and hence there are only six valid combinations of the two. In
    total there are $6 \times 9 \times 16 = 864$ unique system
    configurations.}
  \label{tab:data_type}
\end{table}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth,trim={0 .5in 0 .4in}]{Raw_Throughput.pdf}
  \caption{Histograms of 100-bin reductions of the PMF of I/O
    throughput mean (top) and I/O throughput variance (bottom). In the
    mean plot, the first 1\% bin (truncated in plot) has a probability
    mass of .45. In the variance plot, the second 1\% bin has a
    probability mass of .58. It can be seen that the distributions of
    throughputs are primarily of lower magnitude with occasional
    extreme outliers.}
  \label{fig:raw_throughput}
\end{figure}

\vspace{-10pt}
\subsection{Dimensional Analysis}
\vspace{-10pt}
This work utilizes an extension to standard $k$-fold cross validation
that allows for a more thorough investigation of the expected model
performance in a variety of real-world situations. Alongside
randomized splits, two extra components are considered: the amount of
training data provided, and the dimension of the input data. It is
important to consider that algorithms that perform well with less
training input also require less experimentation. Although, the amount
of training data required may change as a function of the dimension of
the input and this needs to be studied as well. The framework used
here will be referred to as a multidimensional analysis (MDA) of the
IOzone data.

%% \color{red}

\subsubsection{Multidimensional Analysis}
\vspace{-10pt}
This procedure combines random selection of training and testing
splits with changes in the input dimension and the ratio of training
size to testing size. Given an input data matrix with $n$ rows
(points) and $d$ columns (components), MDA proceeds as follows:
\begin{enumerate}
\item For all $k = 1$, $\ldots$, $d$ and for all nonempty subsets $F
  \subset \{1, 2, \ldots, d\}$, reduce the input data to points $(z,
  f_F(z))$ with $z \in \mathbb{R}^k$ and $f_F(z) = E\bigl[ \bigl\{
    f\bigl(x^{(i)}\bigr) \bigm| \bigl(x^{(i)}_F = z\bigr) \bigr\}
    \bigr]$, where $E[\cdot]$ denotes the mean and $x^{(i)}_F$ is the
  subvector of $x^{(i)}$ indexed by $F$.
\item For all $r$ in $\{5, 10, \ldots, 95\}$, generate $N$ random
  splits $(train, test)$ of the reduced data with $r$ percentage for
  training and $100 - r$ percentage for testing.
\item When generating each of $N$ random $(train, test)$ splits,
  ensure that all points from $test$ are in the convex hull of points
  in $train$ (to prevent extrapolation); also ensure that the points
  in $train$ are well spaced.
\end{enumerate}

In order to ensure that the testing points are in the convex hull of
the training points, the convex hull vertices of each set of (reduced
dimension) points are forcibly placed in the training set. In order to
ensure that training points are well spaced, a statistical method for
picking points from \shortciteN{amos2014algorithm} is used:
\begin{enumerate}
\item Generate a sequence of all pairs of points
  $\bigl(z^{(i_1)},z^{(j_1)}\bigr), \bigl(z^{(i_2)},z^{(j_2)}\bigr),
  \ldots$ sorted by ascending pairwise Euclidean distance between
  points, so that $\bigl|\bigl|z^{(i_k)}-z^{(j_k)}\bigr|\bigr|_2 \leq
  \bigl|\bigl|z^{(i_{k+1})}-z^{(j_{k+1})}\bigr|\bigr|_2$.
\item Sequentially remove points from candidacy until only $|train|$
  remain by randomly selecting one point from the pair
  $\bigl(z^{(i_m)}, z^{(j_m)}\bigr)$ for $m = 1,\ldots$ if both
  $z^{(i_m)}$ and $z^{(j_m)}$ are still candidates for removal.
\end{enumerate}

Given the large number of constraints, level of reduction, and use of
randomness in the MDA procedure, occasionally $N$ unique
training/testing splits may not be created or may not exist. In these
cases, if there are fewer than $N$ possible splits, then
deterministically generated splits are used. Otherwise after $3N$
attempts, only the unique splits are kept for analysis. The MDA
procedure has been implemented in Python\#3 while most regression and
interpolation methods are Fortran wrapped with Python. All randomness
has been seeded for repeatability.

For any index subset $F$ (of size $k$) and selected value of $r$, MDA
will generate up to $N$ multivariate models $f_F(z)$ and predictions
$\hat{f}_F\big(z^{(i)}\big)$ for a point $z^{(i)} \in \mathbb{R}^k$.
There may be fewer than $N$ predictions made for any given
point. Extreme points of the convex hull for the selected index subset
will always be used for training, never for testing. Points that do
not have any close neighbors will often be used for training in order
to ensure well-spacing. Finally, as mentioned before, some index
subsets do not readily generate $N$ unique training and testing
splits. The summary results presented in this work use the median of
the ($N$ or fewer) values $\hat{f}_F(z)$ at each point as the model
estimate for error analysis.

%% is open source, and is freely available at the url
%% \textit{https://github.com/tchlux/VarSys/HPC\_Paper/Code/multi\_dim\_analysis.py}

%% \begin{enumerate}
%% \item Cycling the categorical settings
%% \item Selecting subsets of 1,2,3 up to 4 dimensions
%% \item Cycling different training : testing ratios (5:95 $\rightarrow$ 95:5)
%% \item Generating 200 random training : testing splits
%% \item Ensuring the testing points are on/inside the convex hull of the training.
%% \item Ensuring the training points are well-spaced.
%% \end{enumerate}

%% \subsection{Prediction}
%% \begin{enumerate}
%% \item For each file generated from the dimensional analysis, train on
%%   the training data, evaluate at the testing data points
%% \end{enumerate}

%     Results     
%=================
\section{Results}
\label{sec:results}

A na\"{\i}ve multivariate prediction technique such as nearest
neighbor could experience relative errors in the range $\displaystyle
[0, \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x) ]$ when modeling
a nonnegative function $f(x)$ from data. The IOzone mean data response
values span three orders of magnitude (as can be seen in Table
\ref{tab:data_type}) while variance data response values span six
orders of magnitude. It is expected therefore, that all studied
multivariate models perform better than a na\"{\i}ve approach,
achieving relative errors strictly less than $10^3$ for throughput
mean and $10^6$ for throughput variance. Ideally, models will yield
relative errors significantly smaller than 1. The time required to
compute thousands of models involved in processing the IOzone data
through MDA was approximately five hours on a CentOS workstation with
an Intel i7-3770 CPU at 3.40GHz. In four dimensions for example, each
of the models could be constructed and evaluated over hundreds of
points in less than a few seconds.

\vspace{-10pt}
\subsection{I/O Throughput Mean}
\vspace{-10pt}
Almost all multivariate models analyzed make predictions with a
relative error less than 1 for most system configurations when
predicting the mean I/O throughput of a system given varying amounts
of training data. The overall best of the multivariate models,
Delaunay, consistently makes predictions with relative error less than
$.05$ (5\% error). In Figure \ref{fig:mean_tt_ratio} it can also be
seen that the Delaunay model consistently makes good predictions even
with as low as 5\% training data (43 of the 864 system configurations)
regardless of the dimension of the data.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Mean_Dim.pdf}
  \caption{These box plots show the prediction error of mean with
    increasing dimension. The top box whisker for SVR is 40, 80, 90
    for dimensions 2, 3, and 4, respectively. Notice that each model
    consistently experiences greater magnitude error with increasing
    dimension. Results for all training percentages are aggregated.}
  \label{fig:mean_dim}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Mean_TT_Ratio.pdf}
  \caption{These box plots show the prediction error of mean with
    increasing amounts of training data provided to the models. Notice
    that MARS is the only model whose primary spread of performance
    increases with more training data. Recall that the response values
    being predicted span three orders of magnitude and hence relative
    errors should certainly remain within that range. For SVR the top
    box whisker goes from around 100 to 50 from left to right and is
    truncated in order to maintain focus on better models. Results for
    all dimensions are aggregated. Max training percentage is 96\% due
    to rounding training set size.}
  \label{fig:mean_tt_ratio}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Var_TT_Ratio.pdf}
  \caption{These box plots show the prediction error of variance with
    increasing amounts of training data provided to the models. The
    response values being predicted span six orders of magnitude. For
    SVR the top box whisker goes from around 6000 to 400 (decreasing
    by factors of 2) from left to right and is truncated in order to
    maintain focus on better models. Results for all dimensions are
    aggregated. Max training percentage is 96\% due to rounding
    training set size.}
  \label{fig:var_tt_ratio}
\end{figure}

\vspace{-10pt}
\subsection{I/O Throughput Variance}
\vspace{-10pt} The prediction results for variance resemble those for
predicting mean. Delaunay remains the best overall predictor
(aggregated across training percentages and dimensions) with median
relative error of .47 and LSHEP closely competes with Delaunay having
a median signed relative error of -.92. Outliers in prediction error
are much larger for all models. Delaunay produces relative errors as
large as 78 and other models achieve relative errors around
$10^3$. The relative errors for many models scaled proportional to the
increased orders of magnitude spanned by the variance response
compared with mean response. As can be seen in Figure
\ref{fig:var_tt_ratio}, all models are more sensitive to the amount of
training data provided than their counterparts for predicting mean.

\vspace{-10pt}
\subsection{Increasing Dimension and Decreasing Training Data}
\vspace{-10pt}
As can be seen in Figure \ref{fig:mean_dim}, all of the models suffer
increasing error rates in higher dimension. This is expected, because
the number of possible interactions to model grows
exponentially. However, LSHEP and Delaunay maintain the slowest
increase in relative error. The increase in error seen for Delaunay
suggests that it is capable of making predictions with a range of
relative errors that grows approximately linearly with increasing
dimension input. This trend suggests that Delaunay would remain a
viable technique for accurately modeling systems with 10's of
parameters given only small amounts of training data. All models, with
the exception of MARS, produce smaller errors given more training
data. Increasing the amount of training data most notably reduces the
number of prediction error outliers.

%     Discussion     
%====================
\section{Discussion}
\label{sec:discussion}

The present results demonstrate that a straightforward application of
multivariate modeling techniques can be used to effectively predict
HPC system performance. Some modeling effort on the part of a systems
engineer combined with a significant amount of experimentation (days
of CPU time for the IOzone data used here) can yield a model capable
of accurately tuning an HPC system to the desired performance
specification, although qualitatively correct predictions can be
achieved with much less (10\%, say) effort.

\vspace{-10pt}
\subsection{Modeling the System}
\vspace{-10pt}
The modeling techniques generated estimates of drastically different
quality when predicting I/O throughput mean and variance. A few
observations: SVR has the largest range of errors for all selections
of dimension and amounts of training data; MARS and LSHEP produce
similar magnitude errors while the former consistently underestimates
and the latter consistently overestimates; Delaunay has considerably
fewer outliers than all other methods. SVR likely produces the poorest
quality predictions because the underlying parametric representation
is global and oversimplified (a single polynomial), making it unable
to capture the complex local behaviors of system I/O. It is still
unclear, however, what causes the behaviors of LSHEP, MARS, and
Delaunay. An exploration of this topic is left to future work.

While the Delaunay method appears to be the best predictor in the
present IOzone case study, it is important to note that the Delaunay
computational complexity increases with the dimension of the input
more rapidly than other techniques. The implementation of Delaunay
(QuickHull) used would experience unacceptably large training times
beyond ten-dimensional input. This leaves much room for other
techniques to perform best in higher dimension unless a more efficient
implementation of Delaunay can be used.

Finally, the ability of the models to predict variance was
significantly worse than for the I/O mean. The larger scale in
variance responses alone do not account for the increase in relative
errors witnessed. This suggests that system variability has a greater
underlying functional complexity than the system mean and that latent
factors are reducing prediction performance.

\vspace{-10pt}
\subsection{Extending the Analysis}
\vspace{-10pt}
System I/O throughput mean and variance are simple and useful system
characteristics to model. The process presented in this work is
equally applicable to predicting other useful performance
characteristics of HPC systems such as: computational throughput,
power consumption, processor idle time, context switches, RAM usage,
or any other ordinal performance metric. For each of these there is
the potential to model system variability as well. This work has
chosen variance as a measure of variability, but the techniques used
in this paper could be applied to more precise measures of variability
such as the percentiles of the performance distribution or the entire
distribution itself. A thorough exploration of HPC systems
applications of multivariate modeling constitutes future work.

%% \begin{enumerate}
%% \item Measuring different system characteristics other than I/O
%% \item Recording system performance constantly to maintain model accuracy
%% \end{enumerate}

\section{Conclusion}
\label{sec:conclusion}
Multivariate models of HPC system performance can effectively predict
I/O throughput mean and variance. These multivariate techniques
significantly expand the scope and portability of statistical models
for predicting computer system performance over previous work. In the
IOzone case study presented, the Delaunay method produces the best
overall results making predictions for 821 system configurations with
less than 5\% error when trained on only 43 configurations. Analysis
also suggests that the error in the Delaunay method will remain
acceptable as the number of system parameters being modeled
increases. These multivariate techniques can and should be applied to
HPC systems with more than four tunable parameters in order to
identify optimal system configurations that may not be discoverable
via previous methods nor by manual performance tuning.

%     Future Work     
%=====================
\vspace{-10pt}
\subsection{Future Work}
\vspace{-10pt}
The most severe limitation to the present work is the restriction to
modeling strictly ordinal (not categorical) system parameters.
Existing statistical approaches for including categorical variables
are inadequate for nonlinear interactions in high dimensions. Future
work could attempt to identify the viability of different techniques
for making predictions including categorical system parameters.

There remain many other multivariate modeling techniques not included
in this work that should be evaluated and applied to HPC performance
prediction. For I/O alone, there are far more than the four tunable
parameters studied in this work. Alongside experimentation with more
models, there is room for a theoretical characterization of the
combined model and data properties that allow for the greatest
predictive power.

%% \color{black}

%% \newpage
%% \ \\

%% \vspace{3in}

%% \texttt{\textbf{L}oc\textbf{A}ll\textbf{Y} li\textbf{NE}ar
%%   \textbf{W}eighted \textbf{A}pproxima\textbf{T}ion\textbf{S}
%%   \textbf{O}f fu\textbf{N}tions}\\
%% %% \texttt{L\ \ A\ \ Y \ \ NE\ \ \  W\ \ \ \ \ \ \ \ A\ \ \ \ \ \ \ \ T\ \ \ S O\ \ \ \ N\ \ \ \ \ \ }
%% \newpage

\bibliographystyle{scsproc}
\bibliography{paper}

\section*{Additional Authors}
Jon Bernard\\
Dept. of Computer Science, VPI \& SU, Blacksburg, VA 24061 \\ [8pt]
Bo Li \\
Dept. of Computer Science, VPI \& SU, Blacksburg, VA 24061\\ [8pt]
Li Xu \\
Dept. of Statistics, VPI \& SU, Blacksburg, VA 24061\\ [8pt]
Godmar Back\\
Dept. of Computer Science, VPI \& SU, Blacksburg, VA 24061\\ [8pt]
Ali R. Butt\\
Dept. of Computer Science, VPI \& SU, Blacksburg, VA 24061\\ [8pt]
Kirk W. Cameron\\
Dept. of Computer Science, VPI \& SU, Blacksburg, VA 24061\\ [8pt]
Yili Hong\\
Dept. of Statistics, VPI \& SU, Blacksburg, VA 24061\\

%     Author Biography Block     
%================================
\section*{Author Biographies}

\textbf{\uppercase{Thomas C. H. Lux}} is a Ph.D. student at Virginia Tech
studying computer science under Dr. Layne Watson.

\textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has interests
in numerical analysis, mathematical programming, bioinformatics, and data
science.
He has been involved with the organization of HPCS since 2000.

\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. student at Virginia Tech
studying computer science under Dr. Layne Watson.

\textbf{\uppercase{Jon Bernard}} is a Ph.D. student at Virginia Tech
studying computer science under Dr. Kirk Cameron.

\textbf{\uppercase{Bo Li}} is a senior Ph.D. student at Virginia Tech
studying computer science under Dr. Kirk Cameron.

\textbf{\uppercase{Li Xu}} is a Ph.D. student at Virginia Tech studying
statistics under Dr. Yili Hong.

\textbf{\uppercase{Godmar Back}} (Ph.D., University of Utah, 2002) has
broad interests in computer systems, with a focus on performance and
reliability aspects of operating systems and virtual machines.

\textbf{\uppercase{Ali R. Butt}} (Ph.D., Purdue, 2006) has interests in cloud computing, distributed computing, and operating system induced variability.

\textbf{\uppercase{Kirk W. Cameron}} (Ph.D., Louisiana State, 2000) has
interests in computer systems design, performance analysis, and power and
energy efficiency.

\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) has interests in
engineering statistics, statistical modeling, and data analysis.

\end{document}

### 2020-01-16 [2018-01-09]_ACMSE_(box_meshes)/paper.tex
Created paper repository with all of my work.

\input{paper_header}

\def\mn#1{\left\Vert #1\right\Vert_\infty} 
\section{Introduction}
\label{sec_introduction}

Regression and interpolation are problems of considerable importance that find applications across many fields of science. Pollution and air quality analysis \cite{de2008field}, energy consumption management \cite{lazos2014optimisation}, and student performance prediction \cite{cortez2008using} are a few examples of interdisciplinary applications of multivariate regression for predictive analysis. As discussed later, these techniques can also be applied to prediction problems related to high performance computing (HPC) file input/output (I/O), Parkinson's patient clinical evaluations \cite{tsanas2010accurate}, and forest fire risk assessment \cite{cortez2007data}.

Multivariate interpolation is formally defined when there exists some function $f:\mathbb{R}^d \rightarrow \mathbb{R}$ and a set $X$ of $n$ points in $\mathbb{R}^d$ along with associated response values $f(x)$ for all $x \in X$. The problem is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow \mathbb{R}$ such that $\hat f(x) = f(x)$ for all $x \in X$. It is often the case that the form of the true underlying function $f$ is unknown, however it is still desirable to construct an approximation $\hat f$ with minimum approximation error at $y \notin X$.

Multivariate regression is often used when the underlying function is presumed to be stochastic, or stochastic error is introduced in the evaluation of $f$. Hence, multivariate regression relaxes the conditions of interpolation by minimizing the error in $\hat f$ at $x \in X$ while maintaining some parametric form with parameters $P$. This can be written as $\min_{P} ||\hat f(X) - f(X)||$, where $f(X)$ is a vector of $f(x)$ for all $x \in X$ and $||\cdot||$ is an appropriate measure. The difficult question in the case of regression is often what parametric form to adopt for any given application. This paper proposes basis functions with overlapping regions of support as the general parametric form.

As the dimension of data increases, the number of possible interactions between dimensions grows exponentially. Quantifying all possible interactions becomes intractable and hence beyond three-dimensional data, mostly linear models are used. That is not to say nonlinear models are absent, but nonlinearities are often either preconceived or model pairwise interactions between dimensions at most. Similarly, higher order models of single variables can be generated from the techniques proposed in this paper, however only first-order interactions between variables are considered.

Regression and interpolation have a considerable theoretical base in one dimension \cite{cheney2009course}. Splines in particular are well understood as an interpolation technique in one dimension \cite{de1978practical}, particularly B-splines. Fewer techniques have been discussed and evaluated for two dimensional problems. Though in two and three dimensions tensor product splines remain computable \cite{unther1996interpolating}, tensor products have an unfortunate exponential scaling in parameterization with increasing dimension. Exponential scaling prohibits tensor products from being reasonably applied beyond three-dimensional data. In order to address this dimensional scaling challenge, C. de Boor and others have recently proposed box splines \cite{de2013box}. Two of the three meshes introduced in this work involve box-shaped regions of support and use box splines as the underlying basis functions.

% Art is a subjective form of personal expression and computers are just robots that are going to take over the world. \textit{*deep breath*} In order to communicate my frustration with the technological and information revolution, I'm going to write the rest of this paper in poetry.

% Star light, star bright \ldots

\subsection{Box Splines}
\label{sec_box_splines}

\begin{figure}
  \includegraphics[width=0.23\textwidth,height=2.5cm]{1D-linear.pdf}
  \includegraphics[width=0.23\textwidth,height=2.5cm]{1D-quadratic.pdf}
  \caption{1D linear (order 2) and quadratic (order 3) box splines with direction vector sets $\bigl( 1 \ 1 \bigr)$ and $\bigl( 1 \ 1 \ 1 \bigr)$ respectively. Notice that these direction vector sets form the B-Spline analogues, order 2 composed of two linear components and order 3 composed of 3 quadratic components (colored and styled in plot).}
  \label{fig_1D_boxes}
\end{figure}

\begin{figure}
  \vspace{-.5cm}
  \includegraphics[width=0.23\textwidth,height=3cm]{2D-linear.pdf}
  \includegraphics[width=0.23\textwidth,height=3cm]{2D-quadratic.pdf}
  \caption{2D linear (order 2) and quadratic (order 3) box splines with direction vector sets $\bigl( I \ I \bigr)$ and $\bigl( I \ I \ I \bigr)$ respectively, where $I$ is the identity matrix in two dimensions. Notice that these direction vector sets also produce boxes with $\text{order}^2$ subregions (colored in plot).
  \vspace{-0.1cm}}
  \label{fig_2D_boxes}
\end{figure}

A box spline in $\mathbb{R}^d$ is defined by its \textit{direction vector set} $A$, composed of $s$ $d$-vectors where $s \geq d$. Further, $A$ will be written as a $d \times s$ matrix. The first $m$ column vectors of $A$ are denoted by $A_m$, $m \leq s$. $A_d$ is required to be nonsingular. Consider the unit cube in $s$ dimensions $Q_s = [0,1)^s$. $A_s \bigl( Q_s \bigr)$ is now the image (in $d$ dimensions) of $Q_s$ under the linear map $A$. This image is the region of support for the box spline defined by $A_s$ in $d$ dimensions. The box spline function in $d$ dimensions for $A_d$ is defined as
\begin{equation}
B(x \mid A_d) = \begin{cases} 
(\det(A_d))^{-1}, & x \in A_d(Q_d), \\
0,                & \text{otherwise.}
\end{cases}
\label{eq_box_base}
\end{equation}
For $A_s$ when $s > d$ the box spline is computed as
\begin{equation}
B(x \mid A_s) = \int_0^1 B(x - t v_s \mid A_{s-1}) dt,
\label{eq_box_recursive}
\end{equation}
where $v_s$ is the $s$th direction vector of $A$.

The application of box splines presented in this paper always utilizes the $d$-dimensional identity matrix as $A_d$. This simplifies the computation in Equation \ref{eq_box_base} to be the characteristic function for the unit cube. Composing $A$ strictly out of $k$ repetitions of the identity matrix forms the $k$th order B-spline with knots located at $0$, $1$, $\ldots$, $k-1$, $k$ along each axis (see Figure \ref{fig_1D_boxes}). Furthermore, while the number of subregions for the $k$th order $d$-dimensional box spline grows as $k^d$ (see Figure \ref{fig_2D_boxes}), the symmetry provided by direction vector sets composed of repeated identity matrices allows the computation of box splines to be simplified. The value of a box spline at any location is then the product of all axis-aligned 1-dimensional $k$th order box splines.

The box splines as presented are viable basis functions. Each box spline can be shifted and scaled without modifying the underlying computation (similar to wavelets), yet the underlying computation is simple and scales linearly with dimension. For a more thorough introduction and exploration of box splines in their more general form, readers are referred to \cite{de2013box}.

\section{Interpolation and Regression}
\label{sec_mesh_construction}

\begin{figure}
  \centering
  \begin{tikzpicture}[scale=1]
    \draw[thick] (3,1.1) circle (0.7mm);

    %% Draw arrows for each width
    \draw[thick,->] (3,1.2) -- (3,2.4);
    \draw[thick,->] (3,1.0) -- (3,0.1);
    \draw[thick,->] (3.1,1.1) -- (6.9,1.1);
    \draw[thick,->] (2.9,1.1) -- (0.1,1.1);

    %% Draw boundary of box
    \draw[thick,-] (0,0) -- (7,0);
    \draw[thick,-] (7,0) -- (7,2.5);
    \draw[thick,-] (7,2.5) -- (0,2.5);
    \draw[thick,-] (0,2.5) -- (0,0);

    %% Add text to the picture
    \node at (3.25,1.35) {$c$};
    \node at (1.3,1.35) {$l^c_1$};
    \node at (5.1,1.35) {$u^c_1$};
    \node at (2.75,0.55) {$l^c_2$};
    \node at (2.75,1.8) {$u^c_2$};
  \end{tikzpicture}
  \caption{An example box in two dimensions with anchor $c$, upper widths $u^c_1$, $u^c_2$, and lower widths $l^c_1$, $l^c_2$. Notice that $c$ is not required to be equidistant from opposing sides of the box, that is $u^c_i \not= l^c_i$ is allowed.}
  \label{fig_example_box}
\end{figure}

Throughout this section, the notation will be reused from Section \ref{sec_introduction}. $X \subset \mathbb{R}^d$ is a finite set of points with known response values $f(x)$ for all $x \in X$. Also let $L,U \in \mathbb{R}^d$ define a bounding box for $X$ such that $L < x < U$ for all $x \in X$.

Define a box $b^c = (c,l^c,u^c)$ in $d$ dimensions with anchor $c \in \mathbb{R}^d$, lower width vector $l^c \in \mathbb{R}^d_+$, and upper width vector $u^c \in \mathbb{R}^d_+$ (where $u^c_i$ refers to the $i$th component of $u^c$). A visual example of a box in two dimensions can be seen in Figure \ref{fig_example_box}. Now, define a componentwise rescaling function $g: \mathbb{R}^d \rightarrow \mathbb{R}^d$ at point $x \in \mathbb{R}^d$ to be
\begin{equation}
  \bigl(g^c(x)\bigr)_r = {k \over 2} \left( 1 - {(x_r - c_r)_- \over l^c_r} + {(x_r - c_r)_+ \over u^c_r} \right),
\end{equation}
where $y_+=\max\{y,0\}$, $y_-=(-y)_+$, $k$ is the order of the box spline as described in Section \ref{sec_box_splines}. Finally, each box spline in a box mesh can be evaluated as $B^c(x) = B(g^c(x) \mid A)$ presuming the order of approximation implies $A$. Both box meshes described in the following subsections use box spline basis functions of this form.

A notable property of boxes defined with the linear rescaling function $g^c$, is that $C^0$ and $C^1$ continuity of the underlying box spline are maintained. $C^0$ continuity is maintained through scaling. $C^1$ continuity is maintained for all box splines with $C^1$ continuity (order $\geq 3$) because the scaling discontinuity is located at $c$, where all box splines (of the presented form) have first derivative zero. All continuity beyond the first derivative is lost through the rescaling function $g^c$.

\subsection{Max Box Mesh}

The first of the three meshes produces a set of boxes around chosen control points and each box has maximal distance between the control point and the nearest side of the box. This centrality property is one mechanism for creating the largest reasonable regions of support for the underlying basis functions. The individual boxes are constructed via the following procedure given a set of control points $C \subseteq X$, $c^{(i)} \in C$,
\begin{enumerate}
  \item Initialize a box $b^{c^{(1)}} = \bigl(c^{(1)}, (c^{(1)} - L), (U - c^{(1)})\bigr)$. \label{step_init}
  \item Identify $c^{(i)}$ over $\bigl\{ j \bigm| j \ne 1$, $B^{c^{(1)}}\bigl( c^{(j)} \bigr) \ne 0 \bigr\}$ that minimizes $\bigl\|c^{(j)}-c^{(1)}\bigr\|_\infty$.  \label{step_closest}
  \item Change the the box $b^{c^{(1)}}$ along the first dimension $r$ such that $\bigl|\bigl|c^{(1)} - c^{(i)}\bigr|\bigr|_\infty = \bigl\vert c^{(1)} - c^{(i)}\bigr\vert_r $, to exclude $c^{(i)}$ from the support of $B^{c^{(1)}}$. \label{step_shrink}
  \item Repeat steps \ref{step_closest} and \ref{step_shrink} until no point in $C$ is in the support of $B^{c^{(1)}}$ (at most $2d$ times, once for each boundary of a box).
\end{enumerate}

The same process is used to construct boxes around all control points in $C$. In order to improve the generality of the approximation, a set of control points is initially chosen to be well-spaced using a statistical method from \cite{amos2014algorithm}:

\begin{enumerate}
\item Generate a sequence of all pairs of points sorted by ascending pairwise Euclidean distance between points $\bigl(x^{(i_1)},x^{(j_1)}\bigr)$, $\bigl(x^{(i_2)},x^{(j_2)}\bigr)$, $\ldots$ , so that $\bigl\|x^{(i_k)}-x^{(j_k)}\bigr\|_2 \leq \bigl\|x^{(i_{k+1})}-x^{(j_{k+1})}\bigr\|_2$.
\item Sequentially remove points from candidacy until only $|C|$ remain by randomly selecting a single point from each pair $\bigl(x^{(i_m)}, x^{(j_m)}\bigr)$ for $m = 1,\ldots$ if both $x^{(i_m)}$ and $x^{(j_m)}$ are still candidates for removal.
\end{enumerate}

Once the boxes for a max box mesh have been constructed, the parameters can be identified via a least squares fit. The max box mesh (denoted $MBM$) is used to generate a $|X| \times |C|$ matrix $M$ of box spline basis function evaluations at all points in $X$. The solution to the least squares problem $\min_P \bigl\| M \ P - f(X) \bigr\|_2$ is the parameterization of $MBM$. When $C = X$, $M$ is the $|X| \times |X|$ identity matrix, making the max box mesh approximation $\hat f$ an interpolant.

While setting the number of boxes equal to the number of points causes the max box mesh to be an interpolant, the generality of the max box mesh approximation can often be improved by bootstrapping the selection of control points. Given a user-selected batch size $s \leq |X|$, start with $s$ well-spaced control points. Next, measure the approximation error at $x \notin C$ and if the error is too large (determined by user), pick $s$ points at which the magnitude of approximation error is largest, add those points to $C$, and recompute the max box mesh. The user is left to decide the batch size $s$ and the error tolerance based on validation performance and computability. This work uses a batch size of one.

\begin{definition}
The hyperplane $x_r = c_r + u^c_r$ is the upper boundary of box $b^c$ along dimension $r$, and similarly $x_r = c_r - l^c_r$ is the lower boundary of box $b^c$. When the anchor point $y$, for some box $b^y$, lies in the hyperplane (and facet of box $b^c$) defining either boundary along dimension $r$ of $b^c$ it is said that $b^y$ \textit{bounds} $b^c$ in dimension $r$ and is denoted $\bigl( b^c \bigm\vert_r b^y \bigr)$.
\end{definition}

Throughout all experiments and all repeated trials conducted for this study, all tested interpolation points were covered by at least one box in the $MBM$. However, it is possible for the $MBM$ to not form a covering of $[L,U]$ when there are cyclic boundaries. Consider the following example in three dimensions:
\begin{align*}
  C         &= \bigl\{(0,0,0), (1,0,2/3), (1,1,4/3)\bigr\}, \\
  b^{c^{(1)}} &= \bigl( (0,0,0),  \ (*,*,*),  \ (1,*,4/3) \bigr), \\
  b^{c^{(2)}} &= \bigl( (1,0,2/3),\ (1,*,*),  \ (*,1,*)   \bigr), \\
  b^{c^{(3)}} &= \bigl( (1,1,4/3),\ (*,1,4/3),\ (*,*,*)   \bigr).
\end{align*}
Asterisks are used to represent boxes that are not bounded by other boxes along some dimensions. The point $(2,2,-3)$ is not in any of the max boxes defined above. In this case, there is a cycle in box boundaries that looks like $\bigl( b^{c^{(1)}} \bigm\vert_1 b^{c^{(2)}} \bigm\vert_2 b^{c^{(3)}} \bigm\vert_3 b^{c^{(1)}} \bigr)$. This example demonstrates that it is geometrically possible for the max box mesh to fail to cover a space, however experiments demonstrate that it is empirically unlikely.

The max box mesh remains a viable strategy for computed approximations. Given a maximum of $c$ control points in $d$ dimensions with $n$ points, the computational complexities are: $\mathcal{O}(c^2 d)$ for computing boxes, $\mathcal{O}(c d^2 + d^3)$ for a least squares fit, and $\mathcal{O}(n / s)$ for bootstrapping (which is multiplicative over the fitting complexities). Evaluating the max box mesh requires $\mathcal{O}(c d)$ computations.

\subsection{Iterative Box Mesh}

The iterative box mesh ($IBM$) comprises box-shaped regions that each contain exactly one control point in their interior just as in the $MBM$. However, the mesh is a covering for $[L,U]$ by construction and places boxes in a way that reduces apparent error. The boxes are constructed via the following procedure given a finite set of points $X \subset \mathbb{R}^d$, where $C \subseteq X$ is the (initially empty) set of control points.
\begin{enumerate}
\item Add the box that covers $[L,U]$ anchored at the most central point $x^{(k)} \in X$, add $x^{(k)}$ to $C$, and least squares fit the $IBM$ model to all $x \in X$.
\item Add a new box $[L,U]$ anchored at $x^{(i)} \notin C$ such that $\bigl| IBM(x^{(i)}) - f(x^{(i)}) \bigr| = \max_{x \in X \backslash C} \bigl| IBM(x) - f(x) \bigr|$, reshaping all boxes $b^{x^{(j)}}$ that contain $x^{(i)}$ by bounding the first dimension $r$ such that $\bigl| x^{(j)}_r - x^{(i)}_r \bigr| = \bigl\| x^{(j)} - x^{(i)} \bigr\|_\infty$ (also reshaping the box $b^{x^{(i)}}$ symmetrically), add $x^{(i)}$ to $C$, and then least squares fit the $IBM$ model to all $x \in X$. \label{step_add_box}
\item Repeat Step \ref{step_add_box} until model approximation error is below tolerance $t$.
\end{enumerate}

Just as for the $MBM$, the parameters can be identified via a least squares fit. The iterative box mesh is used to generate a $|X| \times |C|$ matrix $M$ of box spline function evaluations at all points in $X$. Now the box spline coefficients are the solution to the least squares problem $\min_P \bigl|\bigl| M\ P - f(X) \bigr|\bigr|_2$. Also as for the $MBM$, $C = X$ causes $M$ to equal the $|X| \times |X|$ identity, making the iterative box mesh approximation $\hat f$ an interpolant.

As opposed to the max box mesh, the bootstrapping procedure is built into the iterative box mesh. The user is left to decide the most appropriate error tolerance, however a decision mechanism and analysis is presented in Section \ref{sec_performance_analysis}. As mentioned earlier, the iterative box mesh is a covering for $[L,U]$ by construction and this can be proved by an inductive argument.

An $IBM$ least squares fit $\hat f(z) = \sum_{j}P_j B^{c^{(i_j)}}(z)$ can generate approximations at new points $z \in Z \subset \mathbb{R}^d$ by evaluating $\hat f(z)$. The computational complexity for generating the mesh is $\mathcal{O}(c^2 n d)$ where $c$ is the number of control points determined by the minimum error threshold and $n = |X|$. The computational complexity of evaluating the mesh at a single point is $\mathcal{O}(c d)$.

\subsection{Voronoi Mesh}

The final of the three meshes utilizes 2-norm distances to define boundaries rather than max norm distances. A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor inherently utilizes the convex region $v^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than any other point $x^{(j)}$. The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $V: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by

$$ V^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2 \ d(y \mid x^{(i)})} \right)_+, $$
where $x^{(i)}$ is the center of the Voronoi cell, $y \in \mathbb{R}^d$ is an interpolation point, and $d(y \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $v^{x^{(i)}}$ in the direction $y - x^{(i)}$. $V^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $V^{x^{(i)}}$ has local support. While $V^{x^{(i)}}(x^{(i)}) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell. Note that this basis function is $C^0$ because the boundaries of the Voronoi cell are $C^0$. In the case that there is no boundary along the vector $w$, the basis function value is always $1$.

While the cost of computing the exact Voronoi cells for any given set of points grows exponentially \cite{dutour2009complexity}, the calculation of $d$ is linear with respect to the number of control points and dimensions. Given any center $x^{(i)} \in \mathbb{R}^d$, set of control points $C \subseteq X$, and interpolation point $y \in \mathbb{R}^d$, $d\bigl(y \mid x^{(i)}\bigr)$ is the solution to
\begin{equation}
  \max_{c \in C\backslash\{x^{(i)}\}} \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2} \ \frac{y \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}{c \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}.
\end{equation}

The parameters of the $VM$ can now be computed exactly as for the $MBM$ and $IBM$. The Voronoi mesh is used to generate a $|X| \times |C|$ matrix $M$ of basis function evaluations at all points in $X$. Now the $VM$ coefficients are the solution to the least squares problem $\min_P \bigl\| M \ P - f(X) \bigr\|_2$. When $X = C$, $M$ is the identity making the mesh an interpolant. Bootstrapping can be performed with an identical procedure to that for the $IBM$.
\begin{enumerate}
\item Pick the most central point $x^{(k)} \in X$ to be the first control point in $C$ and fit the $VM$ model to all $x \in X$.
\item Identify a control point $x^{(i)} \notin C$ such that $\bigl| VM(x^{(i)}) - f(x^{(i)}) \bigr| = \max_{x \in X \backslash C} \bigl| VM(x) - f(x) \bigr|$, add $x^{(i)}$ to $C$, and then fit the $VM$ model to all $x \in X$. \label{step_add_control}
\item Repeat Step \ref{step_add_control} until approximation error is below tolerance $t$.
\end{enumerate}

Any $VM$ is na\"{\i}vely a covering for $[L,U]$, since any possible interpolation point will have a nearest neighbor control point. The computational complexity of evaluating a parameterized Voronoi mesh with $c$ control points is $\mathcal{O}(c^2 d)$. Bootstrapping the generation of a Voronoi mesh requires $\mathcal{O}(c^2 n d)$ computations for a maximum number of basis functions $c$ determined by the error threshold.

% \subsection{Boxed Natural Neighbor}
% \begin{itemize}
% \item How the mesh is generated
% \item How interpolation is done
% \end{itemize}

\section{Data and Analysis}
In order to evaluate the proposed interpolation and approximation techniques, this paper utilizes three data sets of varying dimension and application. In the following subsections the sources and targets of each data set are described, as well as challenges and limitations related to interpolating and approximating these data sets. The distributions of response values being modeled can be seen in Figure \ref{fig_response_hists}. The preprocessing and approximation processes are described in Section \ref{sec_performance_analysis}.

\begin{figure}
  \includegraphics[width=.5\textwidth,height=2cm]{p-hist.pdf}
  \includegraphics[width=.5\textwidth,height=2cm]{f-hist.pdf}
  \includegraphics[width=.5\textwidth,height=2cm]{h-hist.pdf}
  \caption{Histograms of Parkinsons (total UPDRS), forest fire (area), and HPC I/O (mean throughput) response values respectively. Notice that both the forest fire and HPC I/O data sets are heavily skewed.
  \vspace{-.5cm}}
  \label{fig_response_hists}
\end{figure}

\subsection{Data Summary}

\subsubsection{High Performance Computing I/O ($n = 532, d = 4$)}
The first of three data sets is a four-dimensional data set produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers. The system performance data was collected by executing IOzone 40 times for each of a select set of system configurations. A single IOzone execution reports the max I/O file-read throughput seen. The 40 executions for each system configuration are converted to their mean, which is capable of being modeled by each of the multivariate approximation techniques presented in Section \ref{sec_mesh_construction}. The four dimensions being modeled to predict throughput mean are file size, record size, thread count, and CPU frequency.

\begin{figure*}[htb]
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.95\textwidth,height=4cm]{eval-times.pdf}};
    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Fit Time};
  \end{tikzpicture}
  \caption{Time required to generate model fits for each technique with varying relative error tolerance during bootstrapping.
    \vspace{-.3cm}}
  \label{fig_eval_times}
\end{figure*}

\subsubsection{Forest Fire ($n = 517, d = 12$)}
The forest fire data set \cite{cortez2007data} describes the area of Montesinho park burned on specific days and months of the year in terms of the environmental conditions. The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month and day of year, the FFMC, DMC, DC, and ISI indices (see source for details), the temperature in Celsius, relative humidity, wind speed, and outdoor rain. The original analysis of this data set demonstrated it to be difficult to model, likely due to the skew in response values.

\subsubsection{Parkinson's Telemonitoring ($n = 468, d = 16$)}
The final data set for evaluation \cite{tsanas2010accurate} is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in Parkinson's patients. The function to be predicted is a time-consuming clinical evaluation measure referred to as the UPDRS score. The total UPDRS score given by a clinical evaluation is estimated through 16 real numbers generated from biomedical voice measures of in-home sound recordings.

\subsection{Performance Analysis}
\label{sec_performance_analysis}

The performance of the approximation techniques varies considerably across the three evaluation data sets. Relative errors for the most na\"{\i}ve approximators such as nearest neighbor can range from zero to $\displaystyle \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x)$ when modeling a positive function $f(x)$ from data. Each of the approximation techniques presented remain within these bounds and all errors are presented in signed relative form $(\hat f(x) - f(x)) / f(x)$. Before the models are constructed all data values (components $x^{(i)}_r$ of $x^{(i)} \in X$) are shifted and scaled to be in the unit cube $[0,1]^d$, while the response values are taken in their original form. All models are evaluated with $10$ random $80/20$ splits of the data.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    \textbf{Data Set} & \textbf{Technique} & \textbf{Tolerance} & \textbf{Average Error}\\
    \hline
    HPC I/O & MBM & 1.2 & 0.597\\
    Forest Fire & MBM & 1.8 & 3.517\\
    Parkinson's & MBM & 0.6 & 0.114\\
    \hline
    HPC I/O & IBM & 0.4 & 0.419\\
    Forest Fire & IBM & 1.8 & 3.615\\
    Parkinson's & IBM & 1.8 & 0.121\\
    \hline
    HPC I/O & VM & 0.2 & 0.382\\
    Forest Fire & VM & 1.0 & 4.783\\
    Parkinson's & VM & 2.0 & 1.824\\
    \hline
  \end{tabular}
  \caption{The optimal error tolerance bootstrapping parameters for each technique and each data set as well as the average absolute relative errors achieved by that tolerance. Notice that large relative error tolerances occasionally yield even lower evaluation errors, demonstrating the benefits of approximation over interpolation for noisy data sets.
  \vspace{-.5cm}}
  \label{tab_optimal_tolerance}
\end{table}

Each of the approximation techniques presented incorporates bootstrapping based on an allowable error tolerance $t$. An analysis of the effects of bootstrapping error tolerances on validation accuracy can be seen in Figure \ref{fig_all_performance}. The approximation meshes perform best on the forest fire and Parkinson's data sets when the error tolerance used for fitting is large (smoothing rather than interpolating), while near-interpolation generally produces the most accurate models for HPC I/O. Another performance result of note is that the $MBM$ and $IBM$ have very similar basis functions with largely different outputs.

\begin{figure*}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{all-performance.pdf}};
    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Signed Relative Error};
  \end{tikzpicture}
  \caption{The performance of all three techniques with varied relative error tolerance for the bootstrapping parameter. The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively. The rows are for HPC I/O, Forest Fire, and Parkinson's respectively. Notice the techniques' behavior on the Parkinson's and Forest Fire data sets, performance increases with larger error tolerance.}
  \label{fig_all_performance}
\end{figure*}

\begin{figure*}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{perf-sample.pdf}};
    \node[below=of img, node distance=0cm, yshift=1cm] {Signed Relative Error in Prediction};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{A sample of relative errors for all three techniques with optimal selections of error tolerance. The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively. The rows are for HPC I/O, Forest Fire, and Parkinson's respectively.
    \vspace{-.3cm}}
  \label{fig_perf_sample}
\end{figure*}

The selection of bootstrapping error tolerance also effects the computation time required to fit each of the models to data. Figure \ref{fig_eval_times} presents the time required to construct approximations for each model and each data set with varying $t$. The rapid reduction in computation time required for the forest fire and HPC I/O data sets suggests that large reductions in error can be achieved with relatively few basis functions. The Parkinson's data set however presents a more noisy response, with increasing number of basis functions reducing error much less quickly.

The distributions of errors experienced by each approximation technique when the optimal bootstrapping relative error tolerance is selected can be seen in Figure \ref{fig_perf_sample}. HPC I/O exhibits the most normal approximation errors, which suggests that the models are converging on the random noise of the response for the data set. The worst relative approximation errors are produced by the Voronoi mesh on the forest fire data set. The small magnitude true response values contribute to the larger relative errors. Regardless, the $VM$ errors are unacceptably large.

\section{Discussion}

The bootstrapping procedure presented for each approximation technique still has much room for improvement. Initial analysis suggests that the appropriate relative error tolerance needs to be discovered empirically for each application of a modeling technique. Further analytic studies could arrive at methods for determining optimal error tolerances at runtime, however increases in runtime complexity may not be afforded in many applications.

The box-shaped basis functions and the construction algorithms used for the $MBM$ and $IBM$ could become a source of error when $d$ (the dimension of the data $X$) is comparable to $n$ (the number of known points). The blending regions in which multiple basis functions overlap are always axis aligned and in applications such as image analysis, any single dimension may be unsuitable for approximating the true underlying function. The Voronoi mesh attempts to address this problem by utilizing boundaries between points in multiple dimensions simultaneously. However, it is empirically unclear whether the true benefits of the $VM$ are seen in applications where $d \ll n$.

Each of the case studies presented have fewer than $1000$ points. The complexities of the presented approximation techniques are suitable for large dimension, but the increased complexity associated with brute-force bootstrapping currently prohibits their use on larger data sets. The Voronoi mesh in particular has a large complexity with respect to $n$ which could be significantly improved via more greedy bootstrapping. While each technique requires less than ten seconds on average to produce a fit in the presented case studies, the fit time required quickly grows into minutes around $1000$ points. While these initial results appear somewhat limiting, they demonstrate the viability of each mesh and leave room for further theoretical exploration of techniques to reduce the runtime complexity while maintaining the approximation power and flexibility.

\section{Conclusion}

The Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh each provide novel strategies for effectively approximating multivariate phenomonon. The underlying constructions are theoretically straightforward, yet powerful and flexible. The computational complexities of each make them particularly suitable for applications in many dimensions, while the bootstrapping error tolerance parameter allows a balance between smoothing and interpolation to be explored empirically with each application.

\subsection{Future Work}

A thorough comparison with constituent multivariate approximation techniques including but not limited to, linear Shepard interpolation, multivariate adaptive regression splines, multilayer perceptron regression, and Delaunay triangulation constitutes future work. A more detailed study of alternative bootstrapping techniques may also provide valuable insight.

\bibliographystyle{ACM-Reference-Format}
\bibliography{paper}

\end{document}

### 2020-01-16 [2018-01-09]_ACMSE_(box_meshes)/paper_header.tex
Created paper repository with all of my work.

%%%% Proceedings format for most of ACM conferences (with the exceptions listed below) and all ICPS volumes.
\documentclass[sigconf]{acmart}
%%%% As of March 2017, [siggraph] is no longer used. Please use sigconf (above) for SIGGRAPH conferences.

\usepackage{booktabs} % For formal tables
\usepackage{tikz}     % For drawing example boxes
\usepackage{graphicx}
\usetikzlibrary{positioning}

\begin{document}

%%%% Proceedings format for SIGPLAN conferences 
% \documentclass[sigplan, anonymous, review]{acmart}
%%%% Proceedings format for SIGCHI conferences
% \documentclass[sigchi, review]{acmart}
%%%% To use the SIGCHI extended abstract template, please visit
% https://www.overleaf.com/read/zzzfqvkmrfzn
% Copyright
%% \setcopyright{none}
% \setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}
% DOI
% \acmDOI{}
% ISBN
% \acmISBN{}
%Conference
% \acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El Paso, Texas USA} % \acmConference[ACMSE 2018]{ACM Southeastern conference}{March 2018}{Richmond, Kentucky USA} 
% \acmYear{2018}
% \copyrightyear{2018}
% \acmArticle{4}
% \acmPrice{0.00}
% These commands are optional
%\acmBooktitle{Transactions of the ACM Woodstock conference}
%\editor{Jennifer B. Sartor}

\copyrightyear{2018} 
\acmYear{2018} 
\setcopyright{acmlicensed}
\acmConference[ACM SE '18]{ACM SE '18: Southeast Conference}{March 29--31, 2018}{Richmond, KY, USA}
\acmBooktitle{ACM SE '18: ACM SE '18: Southeast Conference, March 29--31, 2018, Richmond, KY, USA}
\acmPrice{15.00}
\acmDOI{10.1145/3190645.3190687}
\acmISBN{978-1-4503-5696-1/18/03}

\title{Novel Meshes for Multivariate Interpolation and Approximation}
%\titlenote{Produces the permission block, and copyright information}
%\subtitle{Extended Abstract}
%\subtitlenote{The full version of the author's guide is available as \texttt{acmart.pdf} document}

\author{Thomas C. H. Lux}
%\authornote{Dr.~Trovato insisted his name be first.}
%\orcid{1234-5678-9012}
\affiliation{%
  \department{Dept. of Computer Science}
  \institution{Virginia Polytechnic Institute and State University}
  \city{Blacksburg} 
  \state{Virginia} 
  \postcode{24060}
}
\email{tchlux@vt.edu}

\author{Layne T. Watson}
%\authornote{Dr.~Trovato insisted his name be first.}
%\orcid{1234-5678-9012}
\affiliation{%
  \department{Depts. of Computer Science, Mathematics, and Aerospace \& Ocean Engineering}
  \institution{Virginia Polytechnic Institute and State University}
}

\author{Tyler H. Chang}
\author{Jon Bernard}
\author{Bo Li}
\author{Xiaodong Yu}
\affiliation{%
  \department{Dept. of Computer Science}
  \institution{Virginia Polytechnic Institute and State University}
}

\author{Li Xu}
\affiliation{%
  \department{Dept. of Statistics}
  \institution{Virginia Polytechnic Institute and State University}
}

\author{Godmar Back}
\author{Ali R. Butt}
\author{Kirk W. Cameron}
\author{Danfeng Yao}
\affiliation{%
  \department{Dept. of Computer Science}
  \institution{Virginia Polytechnic Institute and State University}
}

\author{Yili Hong}
\affiliation{%
  \department{Dept. of Statistics}
  \institution{Virginia Polytechnic Institute and State University}
}

% The default list of authors is too long for headers.
\renewcommand{\shortauthors}{T. Lux et al.}

\begin{abstract}

A rapid increase in the quantity of data available is allowing all fields of science to generate more accurate models of multivariate phenomena. Regression and interpolation become challenging when the dimension of data is large, especially while maintaining tractable computational complexity. This paper proposes three novel techniques for multivariate interpolation and regression that each have polynomial complexity with respect to number of instances (points) and number of attributes (dimension). Initial results suggest that these techniques are capable of effectively modeling multivariate phenomena while maintaining flexibility in different application domains.

\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>10010520.10010553.10010562</concept_id>
%   <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010575.10010755</concept_id>
%   <concept_desc>Computer systems organization~Redundancy</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010553.10010554</concept_id>
%   <concept_desc>Computer systems organization~Robotics</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10003033.10003083.10003095</concept_id>
%   <concept_desc>Networks~Network reliability</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>  
% \end{CCSXML}

% \ccsdesc[500]{Computer systems organization~Embedded systems}
% \ccsdesc[300]{Computer systems organization~Redundancy}
% \ccsdesc{Computer systems organization~Robotics}
% \ccsdesc[100]{Networks~Network reliability}

\keywords{Interpolation, Approximation, Splines, Multivariate, Regression}

\maketitle

### 2020-01-16 [2018-02-16]_IEEESE_(distribution_models)/paper.tex
Created paper repository with all of my work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8

\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
                                                          % if you need a4paper
%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
\IEEEoverridecommandlockouts                              % This command is only needed 
%% % copyright info
%% \IEEEpubid{\makebox[\columnwidth]{978-1-5386-6133-8/18/\$31.00
%%         ~\copyright~2018 IEEE \hfill}\hspace{\columnsep}
%%         \makebox[\columnwidth]{ }}

% if you want to use the \thanks command
\overrideIEEEmargins
% See the \addtolength command later in the file to balance the column lengths on the last page of the document

% The following packages can be found on http:\\www.ctan.org
\usepackage{tikz}     % For adding axes to plots
\usepackage{graphicx} % For including PDFs
\usetikzlibrary{positioning}

%% \usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\title{\LARGE \bf Nonparametric Distribution Models for Predicting and
  \\ Managing Computational Performance Variability* }

\author{Thomas C. H. Lux$^{1}$, Layne T. Watson$^{2}$, Tyler H. Chang$^{1}$, Jon Bernard$^{1}$, Bo Li$^{1}$, Xiaodong Yu$^{1}$,\\ Li Xu$^{3}$, Godmar Back$^{2}$, Ali R. Butt$^{2}$, Kirk W. Cameron$^{2}$, Yili Hong$^{4}$, Danfeng Yao$^{2}$% <-this % stops a space
\thanks{*This work was supported by the National Science Foundation Grant CNS-1565314.}%
\thanks{$^{1}$Doctoral student, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
\thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{3}$Doctoral student, Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{4}$Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
}

\begin{document}

\maketitle
\thispagestyle{empty}
\pagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Performance variability can have a significant impact on many applications of computing. Cloud computing, high performance computing, and computer security communities each exert considerable effort managing and analyzing variability throughout the system stack. This work presents and evaluates a methodology for predicting precise characteristics of the computational performance variability of an input/output (I/O) application over varying system configurations. Results demonstrate that the presented methodology is capable of precisely modeling performance variability, which could allow applications that tighten service level agreements, maximize computational throughput, and obfuscate system configurations against malicious users.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{INTRODUCTION}
\label{sec:introduction}

Computational variability presents itself in a variety of forms. Processes that are apparently identical in a cloud computing or high performance computing (HPC) environment may take different amounts of time to complete the same job. This variability can cause unintentional violations of service level agreements in cloud computing applications or indicate suboptimal performance in HPC applications. The sources of variability, however, are distributed throughout the system stack and often difficult to identify. The methodology presented in this work is applicable to modeling the expected variability of useful computer system performance measures without any prior knowledge of system architecture. Some examples of interesting performance measures that could be modeled with the techniques in this work include computational throughput, power consumption, processor idle time, number of context switches, and RAM usage, as well as any other ordinal measure of performance.

Predicting performance variability in a computer system is a challenging problem that has primarily been attempted in one of two ways: (1) build a statistical model of the performance data collected by running experiments on the system at select settings, or (2) run artificial experiments using a simplified simulation of the target system to estimate architecture and application bottlenecks. In this paper, the proposed modeling techniques rest in the first category and represent a notable increase in the ability to model precise characteristics of variability.

Many previous works attempting to model system performance have used simulated environments \cite{grobelny2007fase,wang2009simulation,wang2013towards}. Grobelny et al. refer to statistical models as being oversimplified and not capable of capturing the true complexity of the underlying system. Historically, statistical models have been limited to modelling at most one or two system parameters and have therefore not been capable of modeling the complexity of the underlying system. \cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}. These limited statistical models have provided satisfactory performance in narrow application settings. However these techniques require additional code annotations, hardware abstractions, or additional assumptions about the behavior of the application in order to generate models. In contrast, the approach presented here requires no modifications of the application, no architectural abstractions, nor any structural descriptions of the input data being modeled. The techniques used in this work only require performance data as input.

Most existing work on performance variability has focused on operating system (OS) induced variability \cite{beckman2008benchmarking,de2007identifying}. Yet, system I/O variability has been particularly difficult for statistical models to capture \cite{bailey2005performance}. The prior work attempting to model I/O variability has been similarly limited to one or two system parameters \cite{lofstead2010managing}.

This paper presents and evaluates an application of multivariate interpolation to the domain of computer system I/O throughput. The interpolatory models are used to predict the cumulative distribution functions of the expected I/O throughput for a system at previously unseen configurations. Beyond this I/O case study, the techniques in this paper can tractably model tens of interacting system parameters with tens of thousands of known configurations. The primary contribution of this work is a modeling framework using multivariate interpolation to capture precise characteristics (via cumulative distribution functions) of an arbitrary performance measure on any type of computer system.

Section \ref{sec:variability_modeling} details the mathematical techniques used to approximate variability, the chosen measurement of error, and an optimization strategy for improving model performance. Section \ref{sec:data} summarizes the data collection process as well as provides summary statistics of the data used for the case study. Section \ref{sec:results} presents the results of the I/O case study including analyses of approximation accuracy, model convergence, and system parameter importance. Section \ref{sec:discussion} discusses possible explanations for witnessed results, potential applications of the predictive methodology, and future improvements of the models.

%% ===================================================================
\section{VARIABILITY MODELING}
\label{sec:variability_modeling}

Performance and its variability can be summarized by a variety of statistics. Mean, range, standard deviation, variance, and interquartile range are a few summary statistics that describe performance and variability. However, the most precise characterization of any performance measure is the cumulative distribution function (CDF), or its derivative the probability density function (PDF). Previous techniques for predicting system performance have strictly modeled real-valued summary statistics because there exists a large base of mathematical techniques capable of approximating functions of the form $f: \mathbb{R}^d \rightarrow \mathbb{R}$. However, there is little systems work approximating functions $f: \mathbb{R}^d \rightarrow \{g \mid g: \mathbb{R} \rightarrow \mathbb{R}\}$.

A major hurdle when modeling functions such as the CDF or PDF is that certain properties must be maintained. It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is right continuous and nondecreasing. For the purpose of function approximation, the properties of the CDF are more straightforward to maintain. This work utilizes the fact that a convex combination of CDFs results in a valid CDF. Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF. A demonstration of how this is applied can be seen in Figure \ref{fig:prediction_example}. Next, the three techniques used to model CDFs are summarized, a mechanism for identifying important features is proposed, and the error metric used throughout this work is presented. For each of the following, $n \in \mathcal{N}$ refers to the number of training samples (known system configurations), $d \in \mathcal{N}$ refers to the dimension of the data (ordinal system parameters), $x^{(i)} \in \mathbb{R}^d$ is a known system configuration with known performance CDF $F_{x^{(i)}}$, and $y \in \mathbb{R}^d$ refers to a previously unseen system configuration with performance CDF $F_y$.

%% ===================================================================
\subsection{Delaunay}
\label{sec:delaunay}

The Delaunay triangulation is a well-studied geometric technique for producing an interpolant \cite{lee1980two}. The Delaunay triangulation of a set of data points into simplices is such that there are no data points inside the sphere defined by the vertices of each simplex. For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and functions $F_{x^{(i)}}$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique convex combination of the vertices,

$$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$

and the Delaunay interpolant $F_y$ at $y$ is given by

$$ F_y = \sum_{i=0}^{d} w_i F_{x^{(i)}}. $$

\begin{figure}[htb]
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this example, the general methodology for predicting a CDF and evaluating error can be seen. The Delaunay method chose three source system configurations (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom). The weighted sum of the three known CDFs produces the predicted CDF (dashed line). The KS Statistic (red arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example. The KS test null hypothesis is rejected by $p$-value 0.01, however it is not rejected by $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction_example}
\end{figure}

The computational complexity of Delaunay interpolation (for the implementation used) is $\mathcal{O}(n d^4 \log d)$, which is capable of scaling reasonably to $d \leq 50$. In the present application, a Delaunay simplex $S$ containing $y$ is found, then the $d+1$ vertices (system configurations) of $S$ are used to assign weights to each vertex and produce the predicted CDF for system configuration $y$.

%% ===================================================================
\subsection{Max Box Mesh}
\label{sec:max_box_mesh}

The Max Box mesh is an interpolation technique that utilizes overlapping box splines \cite{de2013box} as basis functions that are shifted and scaled to have support over box shaped regions. The boxes are constructed such that each box has exactly one point in its interior while maintaining the largest minimum distance between the interior point and the closest face of the box (a measure of centrality). Given a set of box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the max box properties anchored at interior points $\bigl\{x^{(i)}\bigr\}$,

$$ F_y = \frac{\sum_i b^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy $b^{x^{(i)}}(y) \geq 0$. The computational complexity of interpolation via the Max Box mesh is $\mathcal{O}(n^2 d)$.

%% ===================================================================
\subsection{Voronoi Mesh}
\label{sec:voronoi_mesh}

A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than any other point $x^{(j)}$. The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by

$$ v^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ d(y - x^{(i)} \mid x^{(i)})} \right)_+, $$
where $d(w \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = 0$ and $v^{x^{(i)}}$ has local support, giving the interpolated value

$$ F_y = \frac{\sum_i v^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i v^{x^{(i)}}(y)}, $$
where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of interpolation via the Voronoi mesh is $\mathcal{O}(n^2 d)$. All of the approximations are an interpolant involving a convex combination of known functions $F_{x^{(i)}}$.
%% ===================================================================
\subsection{Feature Weighting}
\label{sec:feature_weighting}

It is well-known that an important procedure in any application of predictive methodologies is identifying those features of the data that are most relevant to making accurate predictions \cite{guyon2003introduction}. Selection strategies such as the floating searches studied in \cite{pudil1994floating} or others compared in \cite{ferri1994comparative} can be too expensive for large approximation problems. Rather, this work poses feature selection as a continuous optimization problem. Let $X$ be an $n \times d$ matrix of $n$ known system configurations with $d$ parameters each normalized to be in $[0,1]$. Define an error function that computes the error of a predictive model trained on $X\hbox{ diag }w$, $w \in \mathbb{R}^d$,  by performing ten random splits with 80\% of the rows of $X\hbox{ diag }w$ for training and 20\% for testing. A minimum of this error function could be considered an optimal weighting of the features of $X$. Minimization is performed using a zero order method in the absence of a readily computable gradient.

%% ===================================================================
\subsection{Measuring Error}

The performance of approximation techniques that predict functions can be analyzed through a variety of summary statistics. Mean error, mean absolute error, mean squared error, and the max absolute error are all popular measures. This work uses the max absolute error, also known as the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov} for its compatibility with the KS test.

The two-sample KS test is a useful nonparametric test for comparing two CDFs while only assuming stationarity, finite mean, and finite variance. The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when

$$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test presented in this work $n_1 = n_2 = 150$. An example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction_example}.

%% The majority of results presented in this work are absent any asymetric preprocessing operations on the training data, some findings related to automatic identification of important system parameters are presented in Table \ref{tab:optimized_p_value_failure_rate} and analyzed in \ref{sec:results}.

%% ===================================================================
\section{DATA}
\label{sec:data}

\begin{table}
  \centering
  \begin{tabular}{c|c}
    \hline
    \textbf{System Parameter} & \textbf{Values}\\
    \hline
    File Size (KB) & 4, 16, 64, 256, 1024, 4096, 8192, 16384\\
    \hline
    Record Size (KB) & \multilinecell{4, 8, 16, 32, 64, 128, 256, 512,\\ 1024, 2048, 4096, 8192, 16384}\\
    \hline
    Thread Count & 1, 8, 16, 24, 32, 40, 48, 56, 64\\
    \hline
    Frequency (GHz) & 1.2, 1.6, 2, 2.3, 2.8, 3.2, 3.5\\
    \hline
    Test Type & \multilinecell{Readers, Rereaders, Random Readers, \\ Initial Writers, Rewriters, Random Writers}\\
    \hline
  \end{tabular}
  \caption{A description of system parameters considered for IOzone. Record size must be $\leq$ file size during execution.
    \vspace{-.5cm}}
  \label{tab:data_description}
\end{table}

\begin{figure}
  \includegraphics[width=.5\textwidth]{plot-histogram-throughput.pdf}
  \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations. The distribution is skewed right, with few tests having significantly higher throughput than most others.
  \vspace{-.5cm}}
  \label{fig:throughput_histogram}
\end{figure}

This paper presents a variability modeling case study with a five-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers. Each node contains two Intel Xeon E5-2637 CPUs offering a total of 16 CPU cores with 16GB of DRAM. While the CPU frequency varies depending on the test configuration, the I/O from IOzone is performed by an ext4 filesystem sitting above an Intel SSDSC2BA20 SSD drive. At the time of data collection, Linux kernel Version 4.13.0 was used. The system performance data was collected over two weeks by executing IOzone 150 times for each of a select set of approximately 18K system configurations, for a total of approximately 2.7M executions of IOzone. A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test type. The summary of the data components in $x^{(i)}$ for the experiments for this paper can be seen in Table \ref{tab:data_description}. Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.

Some mild preprocessing was necessary to prepare the data for modeling and analysis. All features were shifted by their minimum value and scaled by their range, mapping each feature independently into $[0,1]$. This normalization ensures each feature is treated equally by the interpolation techniques and should be performed on all data before building models and making predictions regardless of application. All 150 repeated trials for a system configuration were grouped with that configuration. The only nonordinal feature in this data is the test type. All test types were treated as different applications and were separated for modeling and analysis, i.e., predictions for the ``readers'' test type were made using only known configurations for the ``readers'' test type.

%% ===================================================================

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{plot-histogram-80_20-KS.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of Predictions with Given KS Statistic};
  \end{tikzpicture}
  \caption{Histograms of the prediction error for each modeling algorithm from ten random splits when trained with 80\% of the data aggregated over all different test types. The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution. The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.
  \vspace{-.1cm}}
  \label{fig:ks_histogram_80_20}
\end{figure}

\section{RESULTS}
\label{sec:results}

All three interpolation techniques are used to predict the distribution of I/O throughput values at previously unseen system configurations. In order to improve robustness of the error analysis, ten random selections of 80\% of the IOzone data are used to train each model and the remaining 20\% provide approximation error for each model. The recorded errors are grouped by unique system configuration and then averaged within each group. The samples are identical for each interpolation technique, ensuring consistency in the training and testing sets.

The aggregation of errors across all IOzone tests given 80\% of the data as training can be seen in Figure \ref{fig:ks_histogram_80_20}. Agglomerate errors for each technique resemble a Gamma distribution. The percentages of significant prediction errors with varying $p$-values are on display in Table \ref{tab:p_value_failure_rate}. The primary $p$-value used for analyses in this work is 0.001, chosen because close to 2K predictions are made for each test type. Also, applications executed in cloud and HPC systems that could benefit from statistical modeling will be executed at least thousands of times. In line with this knowledge, it is important to ensure that only a small fraction of interpretable results could occur solely under the influence of random chance. When considering the $p=0.001$ results for each technique, a little under half of the predicted CDFs are significantly different from the measured (and presumed) correct CDFs. A rejection rate of 45\% would seem a poor result, however in this situation the complexity of the problem warrants a slightly different interpretation. These predictions are a very \textit{precise} characterization of performance variability, in fact the cumulative distribution function of a random variable is the strongest possible characterization of variability that can be predicted. Globally, only a little under half of the predictions fail to capture \textit{all} of the characteristics of performance variability at new system configurations. It is also demonstrated later in this Section that this result can likely be improved.

\begin{table}
  \centering
  \begin{tabular}{c|c|c}
    \hline
    \textbf{Algorithm} & \textbf{$P$-Value} & \textbf{\% N.H. Rejections} \\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{58.4\\69.3\\61.9}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{51.1\\58.4\\53.4}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{44.1\\46.9\\45.1}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{31.4\\26.6\\28.7}\\
    \hline
  \end{tabular}
  \caption{Percent of null hypothesis rejections rate by the KS-test when provided different selections of $p$-values. These accompany the percent of null hypothesis rejection results from Figure \ref{fig:ks_histogram_80_20}. 
    \vspace{-.5cm}}
  \label{tab:p_value_failure_rate}
\end{table}

While interpreting null hypothesis rejection rates for these interpolation techniques, it is important to consider how the rejection rate reduces with increasing amounts of training data. Figure \ref{fig:ks_failure_by_training} displays the change in $p=0.001$ null hypothesis rejection rate with increasing density of training data up to the maximum density allowed by this set. Delaunay interpolation provides the best results with the least training data by about 5\%, but these low density rejection rates are unacceptably high (90\%). Figure \ref{fig:ks_failure_by_training} clearly shows that this data set and/or the system variables used in the models of performance variability is inadequate to capture the full variability map from system parameters to performance CDF. Which or both obtains is not clear. A few well chosen data points can significantly improve the interpolants, and thus a careful study of the rejection instances is warranted, besides enlarging the set of system variables being modeled.

It may be misleading to consider the global performance of each prediction technique across all test types, as some test types are more difficult than others to predict and have more apparent latent variables. In Figure \ref{fig:ks_failure_by_training_and_test}, the relative difficulty of each IOzone test type can be compared. The I/O test types analyzing reads are typically approximated with lower error than those test types analyzing writes. Regardless of test type, in the aggregate results the KS statistics hover consistently around 0.15, demonstrating an impressively low KS statistic for predictions. In order to address the opacity of aggregate analysis, another case study and an application of the methodology from Section \ref{sec:feature_weighting} is presented in Table \ref{tab:optimized_p_value_failure_rate}.

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-Failure-by-Training.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
  \end{tikzpicture}
  \caption{The performance of each algorithm on the KS test ($p=0.001$) with increasing amounts of training data averaged over all IOzone test types and ten random splits of the data. The training percentages range from 5\% to 95\% in increments of 5\%. Delaunay is the best performer until 95\% of data is used for training, at which Max Box mesh becomes the best performer by a fraction of a percent.
  \vspace{-.1cm}}
  \label{fig:ks_failure_by_training}
\end{figure}

The results presented in Table \ref{tab:optimized_p_value_failure_rate} are achieved by permitting each approximation technique 300 iterations of simulated annealing. In each iteration, the impact of potential weights on the average KS statistic were considered. All weights were kept in the range [0,2], and were applied to the normalized features for frequency, file size, record size, and number of threads. All three approximation techniques had similar optimal weights achieved by simulated annealing of approximately $(.001, 2, 1.7, 1.5)$ for frequency, file size, record size, and number of threads, respectively. Recall that each interpolation technique uses small distances to denote large influences on predictions, meaning that frequency was the most important feature when predicting variability for the ``readers'' test type, followed not-so-closely by number of threads, then record size.

The ``readers'' test type results demonstrate that the underlying prediction techniques work and are capable of seeing rejection rates below 5\% when tuned for a given application. It is important to emphasize that the roughly 95\% of predictions for which the null hypothesis was not rejected are predicting the \textit{precise} distribution of I/O throughput that will be witnessed at a previously unseen system configuration. To the authors' knowledge, there is no existing methodology that is generically applicable to any system performance measure, agnostic of system architecture, and capable of making such powerful predictions.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    \textbf{Algorithm} & \textbf{$P$-Value} & \multilinecell{\textbf{Unweighted}\\\textbf{\% N.H. Rejection}} & \multilinecell{\textbf{Weighted}\\\textbf{\% N.H. Rejection}}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{24.9\\21.3\\18.7} & \multilinecell{30.2\\21.2\\11.3}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{21.6\\16.4\\14.9} & \multilinecell{27.4\\16.4\\7.0}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{19.7\\13.1\\12.3} & \multilinecell{25.4\\13.1\\4.6}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{17.9\\11.3\\8.5} & \multilinecell{23.4\\11.3\\2.3}\\
    \hline
  \end{tabular}
  \caption{The null hypothesis rejection rates for various $p$-values with the KS-test. These results are strictly for the ``readers'' IOzone test type and show unweighted results as well as the results with weights tuned for minimum error (KS statistic) by 300 iterations of simulated annealing. Notice that the weights identified for the Delaunay model cause data dependent tuning, reducing performance. MaxBoxMesh performance is improved by a negligible amount. VoronoiMesh performance is notably improved.
    \vspace{-.5cm}}
  \label{tab:optimized_p_value_failure_rate}
\end{table}

%% \begin{itemize}
%% \item Case studies of good and bad predictions of distributions made, present 3 (best, median, worst)
%% \item Table / Figure showing error for each of the tests (hardest test to predict)
%% \item Presentation of the top most difficult configurations to predict for each test (and the bad performance at those points)
%% %% \item Comparison to predictions made using the "test" numerical column
%% \end{itemize}

\section{DISCUSSION}
\label{sec:discussion}

The results of the IOzone case study indicate that predicting the CDF of I/O throughput at previously unseen system configurations is a challenging problem. The KS statistic captures the worst part of any prediction and hence provides a conservatively large estimate of approximation error. The average absolute errors in the predicted CDFs are always lower than the KS statistics. However, the KS statistic was chosen because of the important statistical theory surrounding it as an error measure. Considering this circumstance, a nonnegligible volume of predictions provide impressively low levels of error. Powerful predictive tools such as those presented in this work allow for more in-depth analysis of system performance variability. For example, system configurations that are most difficult to predict in these tests are likely ``outlier'' configurations that do not resemble those configurations that share many similar parameters. Analysis of these configurations may provide valuable insight into effective application specific operation of computer systems.

As mentioned in Section \ref{sec:introduction}, no prior work has attempted to model an arbitrary performance measure for a system to such a high degree of precision. All previous statistical modeling attempts capture a few ($<3$) ordinal performance measures. Generating models that have such high degrees of accuracy allows system engineers to identify previously unused configurations that present desired characteristics. Service level agreements (SLAs) in cloud computing environments are cause for capital competition that is affected heavily by system performance \cite{patel2009service}. Users prefer SLAs that allow the most computing power per monetary unit, incentivizing service providers to guarantee the greatest possible performance. Overscheduling and irregular usage patterns force cloud service providers to occasionally overload machines, in which case precise models of system performance can be used to statistically minimize the probability of SLA violation. Similar targeted performance tuning techniques can be applied to HPC system configuration to maximize application throughput or minimize system power consumption.

A final application domain affected by this methodology is computer security. Collocated users on cloud systems have received attention recently \cite{ali2015security}. If a malicious collocated user is capable of achieving specific insight into the configuration of the system, or the activity of other collocated users by executing performance evaluation programs (i.e., IOzone), a new attack vector may present itself. Malicious users could be capable of identifying common performance distributions of vulnerable system configurations and vulnerable active user jobs. This knowledge may allow targeted exploits to be executed. Light inspection of raw IOzone I/O throughputs provides substantial evidence that distinct performance distributions coincide closely with specific system configuration parameters. Conversely, a service provider may defend against such attacks by deliberately obfuscating the performance of the machine. Models such as those presented in this paper could identify optimal staggering and time-delay whose introduction into the system would prevent malicious users from identifying system configurations and active jobs.

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-failure-by-training-and-test.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
  \end{tikzpicture}
  \caption{The percentage of null hypothesis rejections for predictions made by each algorithm on the KS test ($p=0.001$) over different IOzone test types with increasing amounts of training data. Each percentage of null hypothesis rejections is an average over ten random splits of the data. The training percentages range from 5\% to 95\% in increments of 5\%. The read test types tend to allow lower rejection rates than the write test types.
  \vspace{-.1cm}}
  \label{fig:ks_failure_by_training_and_test}
\end{figure}

Results presented in Table \ref{tab:optimized_p_value_failure_rate} are particularly interesting, demonstrating that Delaunay appears most vulnerable to data dependent tuning, Max Box mesh is largely insensitive to such tuning, and Voronoi mesh benefits (for this data set) from the tuning.

There are many avenues for extending this modeling methodology. One extension is to add categorical variables to the models. Presently the rejection rate of distribution predictions can only be reduced with large volumes of performance data, however the judicious choice (via experimental design, e.g.) of new data points may be able to effectively reduce the amount of training data required. Finally, more case studies need to be done to test the robustness of the present modeling techniques to changes in domain and performance measure.

\section{CONCLUSION}
\label{sec:conclusion}

The methodology presented is capable of providing new insights, extending existing analyses, and improving the management of computational performance variability. Delaunay, Max Box mesh, and Voronoi mesh interpolation are viable techniques for constructing approximations of performance cumulative distribution functions. A case study on I/O throughput demonstrated that the models are capable of effectively predicting CDFs for most unseen system configurations for any of the available I/O test types. The present methodology represents a notable increase in the ability to statistically model arbitrary system performance measures involving the interaction of many ordinal system parameters.
%% \begin{itemize}
%% \item Modeling the distributions of system performance is valuable.
%% \item This work presents a steep increase in the power of predictions made by statistical modeling techniques.
%% \end{itemize}

\bibliographystyle{IEEEtran}
\bibliography{paper}

\end{document}

### 2020-01-16 [2018-02-16]_IEEESE_(distribution_models)/Results/check.py

@@ -1,0 +1,55 @@
+import os, pickle
+import numpy as np
+from util.data import read_struct
+
+# Frequency,File Size,Record Size,Num Threads,Test,Algorithm,Training
+# Percentage,Train Size,Test Size,Fit Time,Eval Time,KS
+# Statistic,Average Absolute Error,Median Error,Average Error 
+
+data = read_struct("IEEE_Main_Results.csv", verbose=True)
+# for h in list(data.dtype.names)[]
+# print(low_throughput.shape)
+
+print("Result lines:        ", len(data))
+subdata = data[data["Test"] == "readers"]
+print("'readers' lines:     ", len(subdata))
+subdata = subdata[subdata["Algorithm"] == "Delaunay"]
+print("'Delaunay' lines:    ", len(subdata))
+indices = sorted((i for i in range(len(subdata))), key=lambda i:
+                 data[i]["KS Statistic"])
+median_error = subdata[indices[len(indices)//2]]
+print("Median KS Stat point:", median_error)
+exit()
+
+# subdata = data[data["Training Percentage"] == 95]
+# subdata = data[data["Test"] == ]
+ks_values = [.1568, .1879, .2251, .3110]
+
+for alg in sorted(set(subdata["Algorithm"])):
+    print(alg)
+    perf_data = subdata[subdata["Algorithm"] == alg]["KS Statistic"]
+    for ks in ks_values:
+        print("","%.4f"%(ks),"%.4f"%(sum(np.where(perf_data > ks,1,0)) / len(perf_data)))
+
+
+# names = list(data.dtype.names)
+# print(names)
+# for n in names[11:]:
+#     col = data[n]
+#     print()
+#     print(n)
+#     print("Min: ",min(col))
+#     print("Max: ",max(col))
+#     print("Bad: ", sum(np.where(abs(col) > 1, 1, 0)))
+#     indices = np.arange(len(col))
+#     print("Algs:",sorted(set(data["Algorithm"][indices[abs(col) > 1]])))
+
+# from util.plotly import Plot
+# p = Plot()
+# for alg in sorted(set(data["Algorithm"])):
+#     print("Finding '%s'"%alg)
+#     subdata = data[data["Algorithm"] == alg]["KS Statistic"]
+#     print("Adding...")
+#     p.add_histogram(alg, subdata)
+# p.plot()
+

### 2020-01-16 [2018-02-16]_IEEESE_(distribution_models)/Results/code.py

@@ -1,0 +1,417 @@
+import os, pickle, time
+import numpy as np
+from util.data import read_struct
+from util.stats import cdf_fit_func, ks_diff, ks_p_value
+from util.algorithms import VoronoiMesh, MaxBoxMesh, Delaunay, \
+    CLASS_NAME, SMALL_NUMBER
+
+CLEAN_DATA_CSV = "clean_data.csv"
+CLEAN_DATA_CSV_PKL = "clean_data.csv.pkl"
+DATA_OUTPUT_FILE = "results.csv"
+EXAMPLE_PREDICTION_FILE = "Delaunay_Example_Prediction.csv"
+
+ALGORITHMS = Delaunay, MaxBoxMesh, VoronoiMesh
+TRAINING_PERCS = np.linspace(95,5,19)
+NUM_FOLDS = 10
+RANDOM_SEED = 1
+DEFAULT_TRAINING_PERC = 80
+MAX_COMPUTE_TIME = 60*60*3
+P_VALUES = [0.05, 0.01, 0.001, 1e-6]
+KS_VALUES = [0.1568, 0.1879, 0.2251, 0.3110]
+
+RUN_PERFORMANCE_ANALYSIS = False
+OPTIMIZE_FEATURE_WEIGHTS = False
+FIND_NEW_OPTIMAL_SOLUTIONS = False
+DELAUNAY_EXAMPLE_PREDICTION = True
+
+if not os.path.exists(CLEAN_DATA_CSV_PKL):
+    data = read_struct(CLEAN_DATA_CSV)
+    print("Saving '%s'..."%(CLEAN_DATA_CSV_PKL))
+    with open(CLEAN_DATA_CSV_PKL, "wb") as f:
+        pickle.dump(data,f)
+else:
+    print("Loading '%s'..."%(CLEAN_DATA_CSV_PKL))
+    with open(CLEAN_DATA_CSV_PKL, "rb") as f: 
+        data = pickle.load(f)
+
+header = list(data.dtype.names)
+h_config = header[:4]
+h_test = header[4]
+h_values = header[5:]
+tests = sorted(np.unique(data[h_test]))
+test_len = max(map(len, tests))
+num_configs = len(h_config)
+
+print("Tests:", tests)
+print("Number of configuration columns:", num_configs)
+print()
+
+# print("Converting '%s' to numbers via the following map:"%h_test)
+# test_avg_throughput = {}
+# for t in tests:
+#     subdata = data[data[h_test] == t]
[…]

### 2020-01-16 [2018-02-16]_IEEESE_(distribution_models)/Results/initial_processing.py

@@ -1,0 +1,276 @@
+import os, pickle
+import numpy as np
+from util.data import read_struct
+from util.plotly import Plot, multiplot
+
+DATA_FILE_PKL = "py_data.pkl"
+COUNTS_FILE_PKL = "config_counts.pkl"
+CLEAN_DATA_PKL = "clean_data.pkl"
+DIST_PREDICTION_FILE = "dist_prediction.html"
+TEST_HIST_FILE = "tests_by_machine_(aggregated).html"
+CLEAN_DATA_CSV = "clean_data.csv"
+
+PATH_TO_CLEAN_DATA = os.path.join("/Users","thomaslux","Git","LargeFiles")
+INITIAL_DATA_READ = not os.path.exists(DATA_FILE_PKL)
+PLOT_AGGREGATE_HISTOGRAMS = False
+VALUES_SUMMARY = True
+SHOW_COUNT_SUMMARY = True
+SAMPLE_FUNCTION_PREDICTION = False
+SAVE_CLEAN_DATA_CSV = False
+
+# ====================================================================
+# ====================================================================
+#      Read in new data     
+if INITIAL_DATA_READ:
+    print("Reading data 1...")
+    data1 = read_struct(os.path.join(PATH_TO_CLEAN_DATA,"2018-1.csv"))
+    print("Reading data 2...")
+    data2 = read_struct(os.path.join(PATH_TO_CLEAN_DATA,"2018-2.csv"))
+    print("Reading data 3...")
+    data3 = read_struct(os.path.join(PATH_TO_CLEAN_DATA,"2018-3.csv"))
+    print("Saving all 3 to file...")    
+    with open(DATA_FILE_PKL, "wb") as f:
+        pickle.dump( (data1, data2, data3), f )
+else:
+    print("Loading '%s'..."%(DATA_FILE_PKL))
+    with open(DATA_FILE_PKL, "rb") as f:
+        data = np.concatenate(pickle.load(f))
+header = list(data.dtype.names)
+print("Data has %i rows. (exclusing 18 erroneous rows)"%(len(data)))
+h_width = max(len(h) for h in header)
+config_vars = header[2:6] + [header[7]]
+print("Variables:",config_vars)
+
+# ====================================================================
+# Generate the counts dictionary (pkl contains unique configs and data)
+if not os.path.exists(COUNTS_FILE_PKL):
+    counts = {}
+    size = len(data)
+    for i,row in enumerate(data):
[…]

### 2020-01-16 [2019-02]_NUMA_(interpolant_error_bound)/Data/NA_Tooling_Test/tests.f90
Created paper repository with all of my work.

@@ -1,0 +1,5 @@
+
+SUBROUTINE SQUARE(A)
+  REAL, INTENT(INOUT) :: A
+  A = A**2
+END SUBROUTINE SQUARE

### 2020-01-16 [2019-02]_NUMA_(interpolant_error_bound)/Data/run_predictions.py

@@ -1,0 +1,208 @@
+import os
+import numpy as np
+
+from util.stats import cdf_fit
+from util.data import Data
+from util.system import Timer
+from util.math import pairwise_distance
+from util.approximate import Delaunay, Voronoi, BoxMesh, ShepMod, \
+    LSHEP, NeuralNetwork, MARS, SVR_RBF, class_name
+
+# Make sure all of the prepared and pre-processed data files exist.
+from preprocess_data import cwd, raw_dir, data_dir
+
+# Overwrite the print function so that it forces prints to flush.
+def print(*args, og_print=print, **kwargs):
+    kwargs["flush"] = True
+    return og_print(*args, **kwargs)
+
+
+seed = 0
+folds = 10
+
+data_name = lambda p: os.path.basename(p).replace(".dill","").replace(".csv","")
+
+models = [ShepMod, BoxMesh, Voronoi, Delaunay, 
+          NeuralNetwork, LSHEP, SVR_RBF, MARS,]
+
+# models = [BoxMesh, SVR_RBF]
+
+# Recording data for each set and for each algorithm that looks like:
+# | Data name | Dimension | Fold Size | Fold Number | Model Name | Model fit time |
+# | x | ... | y | ... | Model y | ... | Model nearest | Model max edge | Model furthest | Model num contributors | Model prediction time |
+
+
+intermediate_fit_results = os.path.join(".", "Predictions", "fit-intermediate.dill")
+final_fit_results = os.path.join(".", "Predictions", "fit-final.dill")
+if os.path.exists(intermediate_fit_results):
+    fit_data = Data.load(intermediate_fit_results)
+else:
+    fit_data = Data(names=["Data name", "Model name", "Train Shape", "Test Shape", "Fold number", "Fit time"],
+                    types=[ str,         str,          tuple,         tuple,        int,           float,   ])
+
+# from util.approximate.delaunay import DelaunayP1, DelaunayP2, DelaunayP3
+# models = [Delaunay, DelaunayP1, DelaunayP2, DelaunayP3]
+
+for raw_file in sorted(os.listdir(raw_dir)):
+    raw_file = raw_file.replace(".gz","")
+    raw_path = os.path.join(raw_dir, raw_file)
+    data_path = os.path.join(data_dir, raw_file + '.dill')
+    intermediate_results_file = os.path.join(".", "Predictions",
+                    f"intermediate_results_{data_name(raw_file)}.dill")
+    final_results_file = os.path.join(".", "Predictions",
[…]

### 2020-01-16 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/make_analytic_visuals.py
Created paper repository with all of my work.

@@ -1,0 +1,112 @@
+from run_analytic_test import output_data_name
+from util.data import Data
+from util.plot import Plot, multiplot
+from util.stats import cdf_fit
+from math import log
+
+# algorithms = {}
+algorithms = {"DelaunayP1", "NeuralNetwork"}
+
+BOX_PLOTS = False
+MAX_ERROR_LINE = True
+
+if MAX_ERROR_LINE:
+    test_func_name = "oscillatory"
+    test_size = 1000
+    d = None
+    for dimension in [2,20]:
+        file_name = f"final-{dimension}D-{test_size}-{test_func_name}-analytic.pkl"
+        data = Data.load(file_name)
+        data.add_column([dimension]*len(data), "Dimension", index=data.names.index("Train"))
+        # Generate interesting extra columns.
+        data["Abs Errors"] = ([float(abs(v)) for v in l] for l in data["Errors"])
+        data["Mean Abs Error"] = (sum(l) / len(l) for l in data["Abs Errors"])
+        data["Min Abs Error"] = (min(l) for l in data["Abs Errors"])
+        data["Max Abs Error"] = (max(l) for l in data["Abs Errors"])
+        if (d is None): d = data
+        else:           d += data
+    if (len(algorithms) > 0): d = d[d["Algorithm"] == algorithms]
+    print(d)
+    d.save("all-data.pkl")
+    print()
+    config_cols = ["Function", "Dimension"]
+    configs = d[:,config_cols].unique()
+    configs.sort()
+    configs.max_display=float('inf')
+    print(configs)
+    for conf in configs:
+        plots = []
+        for S in sorted(set(d["SNR"])):
+            d_conf = d[d[:,config_cols+["SNR"]] == (list(conf) + [S])]
+            F, D = conf
+            p = Plot(f"{D}D '{F}' function", f"SNR = {S}", "|<i>error</i>|",
+                     font_family="times") #b"log(|<i>error</i>|)")
+            seen = {}
+            for algorithm in sorted(set(d["Algorithm"])):
+                d_alg = d_conf[d_conf["Algorithm"] == algorithm]
+                print(d_alg)
+                values = []
+                locations = []
+                # Sort the algorithm data by the 
+                d_alg.sort(key=lambda i: i["Train"])
+                for row in d_alg:
+                    if (row["Train"] < row["Dimension"]): continue
+                    locations.append( "N = "+str(row["Train"]) )
+                    values.append( [v for v in row["Abs Errors"]] )
[…]

### 2020-01-16 [2019-02]_NUMA_(interpolant_error_bound)/Data/NA_Tooling_Test/test.py

@@ -1,0 +1,53 @@
+from util.system import Timer
+from util.approximate import Voronoi, Delaunay
+from util.data import Data
+
+t = Timer()
+print("Loading data..")
+d = Data.load("parkinsons_updrs.csv.dill")
+print("Converting to matrix..")
+num = d.to_matrix()
+print("Rescaling data..")
+num.data -= num.shift
+num.data /= num.scale
+print("Extracting matrix..")
+x = num.data[:,:-1]
+y = num.data[:,-1]
+
+print(x.shape)
+print(y.shape)
+
+fit_times = []
+approx_times = []
+median = 0
+
+for k,(train, test) in enumerate(d.k_fold(only_indices=True)):
+    print(f"  {k+1}:10", end="\r", flush=True)
+    train_x = x[train]
+    train_y = y[train]
+    # Time Delaunay "fit" operations
+    m = Delaunay()
+    m.fit(x,y)
+    t.start()
+    m(x[0])
+    t.stop()
+    print("Delaunay 'fit':", t.total)
+    # Time the fit.
+    model = Voronoi()
+    t.start()
+    model.fit(train_x, train_y)
+    t.stop()
+    fit_times.append(t.total)
+    # Cycle test and make predictions.
+    for j,i in enumerate(test):
+        print(f"  {k+1}:10  {j+1}:{len(test)} {median}", end="\r", flush=True)
+        t.start()
+        model(x[i])
+        t.stop()
+        approx_times.append(t.total)
+        median = sorted(approx_times)[len(approx_times)//2]
+    print(" "*70, end="\r")
+    print("Fit:    ", sorted(fit_times)[len(fit_times)//2])
+    print("Approx: ", sorted(approx_times)[len(approx_times)//2])
+    # print("Skipped:", sorted(model.skipped)[len(model.skipped)//2])
+    print()

### 2020-01-16 [2019-02]_NUMA_(interpolant_error_bound)/Data/NA_Tooling_Test/time_delaunay.py

@@ -1,0 +1,87 @@
+import os
+import numpy as np
+
+from util.data import Data
+from util.system import Timer
+from util.approximate import Delaunay, class_name
+
+# Make sure all of the prepared and pre-processed data files exist.
+from preprocess_data import cwd, raw_dir, data_dir
+
+to_test = 30
+
+
+# Overwrite the print function so that it forces prints to flush.
+def print(*args, og_print=print, **kwargs):
+    kwargs["flush"] = True
+    return og_print(*args, **kwargs)
+
+
+seed = 0
+folds = 10
+
+data_name = lambda p: os.path.basename(p).replace(".dill","").replace(".csv","")
+
+# Recording data for each set and for each algorithm that looks like:
+# | Data name | Dimension | Fold Size | Fold Number | Model Name | Model fit time |
+# | x | ... | y | ... | Model y | ... | Model nearest | Model max edge | Model furthest | Model num contributors | Model prediction time |
+
+for raw_file in sorted(os.listdir(raw_dir)):
+    raw_file = raw_file.replace(".gz","")
+    raw_path = os.path.join(raw_dir, raw_file)
+    data_path = os.path.join(data_dir, raw_file + '.dill')
+    intermediate_results_file = os.path.join(".", "Predictions",
+                    f"intermediate_results_{data_name(raw_file)}.dill")
+    final_results_file = os.path.join(".", "Predictions",
+                    f"results_{data_name(raw_file)}.dill")
+    print('-'*70)
+    print(data_name(raw_path))
+    print(raw_path)
+    print(final_results_file)
+    d = Data.load(data_path)
+    target = d.names[-1]
+    # Declare the indices..
+    d["Indices"] = range(len(d))
+    # Compute the number of training columns.
+    num_train_columns = d.names.index(target)
+    print(f"Data shape: {d.shape} predicting '{target}'")
+    # Start the folds.
+    for i,(train, test) in enumerate(d.k_fold(k=folds, seed=seed)):
+        # Get the numeric representation of this data based on training,
+        # ignore extra added columns and the "target" column.
+        num = train[:,:num_train_columns].to_matrix()
+        train_x = num.data
+        train_y = list(train[target])
+        # Generate the numeric testing set basesd on training.
+        test_x = np.array([num.to_real(row[:num_train_columns]) for row in test])
+        test_y = list(test[target])
+        # Normalize the training and testing data to the unit hypercube.
+        train_x = (train_x - num.shift) / num.scale
[…]

### 2020-01-16 [2019-10]_ISTAS_(ethical_prediction)/paper/paper.tex
Created paper repository with all of my work.

\documentclass[conference]{IEEEtran}

%% \documentclass[sigconf]{acmart}

% The following packages can be found on http:\\www.ctan.org
\usepackage{tikz}     % For adding axes to plots
\usepackage{graphicx} % For including PDFs
\usetikzlibrary{positioning}
\usepackage{hyperref}

\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{textcomp} % For the text tilde low
\usepackage{tabularx}
    \newcolumntype{L}{>{\raggedright\arraybackslash}X}

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

%% Command for a text tilde
\newcommand{\thomastilde}{\raisebox{0.1ex}{\texttildelow}}

%% Make citations group together.
\usepackage[noadjust]{cite}

\begin{document}

\title{A Case Study on a Sustainable Framework for Ethically Aware Predictive Modeling}

\author{\IEEEauthorblockN{Anonymous Author(s)}}

%% \author{%
%%   \IEEEauthorblockN{Thomas Lux}%
%%   \IEEEauthorblockA{Department of Computer Science\\%
%%     Virginia Polytechnic Institute and State University\\%
%%     Blacksburg, Virginia 24060\\%
%%     Email: tchlux@vt.edu}%
%%   \and%
%%   \IEEEauthorblockN{Stefan Nagy}%
%%   \IEEEauthorblockA{Department of Computer Science\\%
%%     Virginia Polytechnic Institute and State University\\%
%%     Blacksburg, Virginia 24060}%
%%   \and%
%%   \IEEEauthorblockN{Mohammed Almanaa}%
%%   \IEEEauthorblockA{Department of Civil Engineering\\%
%%     Virginia Polytechnic Institute and State University\\%
%%     Blacksburg, Virginia 24060}%
%%   \and%
%%   \IEEEauthorblockN{Sirui Yao}%
%%   \IEEEauthorblockA{Department of Computer Science\\%
%%     Virginia Polytechnic Institute and State University\\%
%%     Blacksburg, Virginia 24060}%
%%   \and%
%%   \IEEEauthorblockN{Reid Bixler}%
%%   \IEEEauthorblockA{Department of Computer Science\\%
%%     Virginia Polytechnic Institute and State University\\%
%%     Blacksburg, Virginia 24060}%
%% }

% \thanks{$^{1}$ Student in Department of Civil Engineering}%
% \thanks{$^{2}$ Student in Department of Computer Science}%

% \thispagestyle{empty}
% \pagestyle{empty}

\maketitle

%% \IEEEpeerreviewmaketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Large volumes of data allow for modern application of statistical and mathematical models to practical social issues. Many applications of predictive models like criminal activity heatmapping, recidivism estimation, and child safety scoring rely on data that may be incomplete, incorrect, or biased. Many sensitive social and historical issues can unintentionally be incorporated into predictions causing ethical mistreatment. This work proposes a mechanism for continuously mitigating model bias by using algorithms that produce predictions from reasonably small sets of data, allowing a human-in-the-loop approach to model application. The benefits offered by this framework are twofold: (1) bias can be identified either statistically or by human users on a per-prediction basis; (2) data can be cleaned for bias on a per-prediction basis. A modeling and data management methodology similar to that presented here could strengthen the ethical application of data science and make the process of cleaning and validating data manageable in the long term.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% \IEEEkeywords{Ethics, Predictive Models, Multilayer Perceptron, Decision Tree, Local}

\section{INTRODUCTION}
\label{sec:introduction}
Data collection efforts are often aimed at providing or improving services for our communities and society as a whole. Large scale predictive models have coevolved with the explosion of available data since the start of the digital age and these models are practically necessary for success in domains as diverse as healthcare, employment, finance, and government. Creators of predictive models were motivated by a critical need to anticipate the future and to take action in advance. When the models succeed they can create life-saving value \cite{tomavsev2019clinically} or change economies \cite{siegel2016predictive}, but when they fail they can reinforce historical social injustice \cite{angwinmachine} and leave at-risk children without needed help \cite{eubanks2018child}. It is not surprising then that interpretability is an important research area for commonly used predictive models \cite{vellido2012making,doshi2018considerations,du2018techniques}.

A model is often considered unethical when it violates the expected practices and procedures of a formal agency like a government or business. For example, a model built from anonymous historical data and designed to assist a business in effective employee hiring may recover protected demographic information on the race, sex, or religion of the applicant. However, Title VII of United States federal code prohibits the use of this information in employment decisions. The predictive model would be acting unethically and unlawfully, but the user (the business) may be entirely unaware.

Many researchers have unveiled clear signs of unequal and racial treatment of individuals by predictive models \cite{squires2003racial,stoll2004black,angwinmachine}.  For example, Angwin et al. analyzed a well-known criminal risk assessment: COMPAS and concluded the outcomes of the model are racially biased \cite{angwinmachine}. They showed that ``Black defendants were twice as likely as white defendants to be misclassified as a higher risk of violent recidivism, and white recidivists were misclassified as low risk 63.2 percent more often than black defendants.'' The U.S. government has acknowledged the enormous potential for negative impact in biased models and the White House published several reports highlighting the potential bias that could adversely affect individuals or groups \cite{smith2016big}. Research is constantly underway that proposes ethical frameworks and potential solutions to protect principles of ethics and privacy \cite{dedeo2014wrong,vayena2016elements,kass2001ethics}.

In an effort to prevent biased and discriminatory decisions, policy makers and practitioners have classified unwanted discriminatory attributes as protected information that cannot be used as predictors in a model. However, a model made from anonymized unprotected data can become unethical even when it is not provided protected information. The model needs only to infer protected characteristics from the provided data in the prediction process \cite{dedeo2014wrong,pedreshi2008discrimination}. The possibility of protected inference makes developing unbiased models challenging, particularly when large amounts of data are involved. A burden is placed on users, who must identify ways to ensure predictions remain ethically sound. Generally there are only two methods to ensure a model is not violating ethical expectations: either carefully monitor and analyze all predictions for ethical mistreatment, or guarantee that the data utilized by the model (and how that data is used to predict) is ethically sound. This work makes an argument for the latter, as a common class of models can make good predictions from human-digestible amounts of data.

This paper makes an argument for per-prediction ethical validation of data and models via local approximation methods that produce interpretable results. The improvements offered by this methodology are twofold: (1) bias can be identified either statistically or by human users on a per-prediction basis; (2) data can be cleaned for bias on a per-prediction (regular) basis. Modeling techniques similar to those presented in this work could not only strengthen the ethical application of data science, but also make the process of cleaning and validating data manageable in the long term.

\section{RELATED WORK}
\label{sec:related}
Algorithmic bias has long been a subject of research on machine learning \cite{kirkpatrick2016battling, hajian2016algorithmic, baeza2016data}. 
Mitchell \cite{mitchell1980need} initially defined machine learning bias as ``any basis for choosing one generalization over another, other than strict consistency with the instances''.
Mooney \cite{mooney1996comparative} expanded this definition with the following assertions: (1) that every model bears some inherent bias, and (2) that detecting a model's bias requires comparison against others.
In recent years, the broad adoption of machine learning has made algorithmic bias a factor in real-world data discrimination \cite{squires2003racial, stoll2004black, angwinmachine}.

Several prior works have explored the problem of measuring algorithmic bias \cite{romei2014multidisciplinary, kamiran2013quantifying, zliobaite2015survey}.
Calders \& Verwer \cite{Calders2010} formalized discrimination as, given an input characteristic, the unequal distribution of outputs for different groups. 
Calders-Verwer (CV) scoring is frequently \cite{kamishima2011fairness, kamishima2012fairness, galhotra2017fairness, kamiran2012decision, vzliobaite2011handling} used to measure \emph{group discrimination} \cite{zafar2015learning, zemel2013learning}. For example, if a loan classifier produces dissimilar outcomes for both sexes (e.g. $P(Y=loan|S=male) > P(Y=loan|S=female)$), then its CV score is measured as the difference in outcomes between those groups (e.g. $P(Y=loan|S=male) - P(Y=loan|S=female)$.
Many works have since expanded on CV scoring for preventing discrimination \cite{zafar2017fairness, kamiran2013techniques, galhotra2017fairness}. 
Four general approaches exist: (1) \emph{Suppression} -- removing attributes most correlating with discrimination-sensitive attributes; (2) \emph{Dataset ``massaging''} -- altering labels of some objects to mitigate unwanted classifier outcomes; (3) \emph{Reweighting} -- assigning data carefully chosen weights to lessen the degree of discrimination; (4) \emph{Sampling} -- under- and over-sampling certain groups to compensate.

A recent focus of machine learning and data science research has been on improving model interpretability \cite{lipton2016mythos, juang2013data}. More specific goals include transparency for assessing model ethicality \cite{goodman2017european, herman2017data}, augmenting informativeness \cite{kim2015interactive, yang2018visually, wu2017interpretable}, and inferring causality \cite{syed2009enzyme}. An obstacle to model interpretability is the lack of transparency of black-box classifiers such as deep neural networks \cite{casillas2013interpretability, cortez2013using, krause2016interacting}.  Post-hoc interpretability \cite{lipton2016mythos} (e.g. visualizations or explanations) represents a promising alternative, however, formalizing  ``model-agnostic'' methodologies remains an ongoing research problem \cite{ribeiro2016should}.

\section{Model Descriptions}

In order to construct a predictive model from data, it is usually assumed that the phenomenon being predicted has some underlying \textit{function} that can be approximated. Many algorithms exist for constructing approximations of unknown functions given data. The following subsections roughly outline the mathematical formulation of both the (arguably more popular) \textit{global} models and the (less popular) \textit{local} models analyzed in the present case study.

\subsection{Global Predictive Models}
Classic machine learning and data science techniques applied today often rely on solving a very specific problem. They create a global predictive model with the aim of capturing trends that exist across (hundreds of) thousands of examples. In general, these global models are constructed given real valued data matrix $X \in \mathbb{R}^{n\ \times\ d}$, a truth function $f: \mathbb{R}^d \rightarrow \mathbb{R}$, and \textit{labels} $f(x^{(i)})$ for row vectors $x^{(i)} \in X$, $1 \leq i \leq n$. These models find the solution to

$$ \min_P \bigl\|\ \hat f_P(X) - f(X)\ \bigr\|, $$

\noindent where $\hat f_P: \mathbb{R}^d \rightarrow \mathbb{R}$ is the parametric approximation, $f(X)$ is used to denote the vector with components $f(X)_i = f(x^{(i)})$, and $\| \cdot \|$ is an appropriate measure. The labels may be real numbers, like probability of recidivism estimates, or categories such as ``safe'' or ``not safe'' for an at-risk child.

The difficulty with these models is that the minimization search which identifies the parameters for the model is performed over \textit{all} data. Whenever it is time to explain a prediction, the answer is often ``all data was used to capture this trend''. The models of this form that will be applied are a multilayer perceptron (MLP) and a decision tree (DT).

\subsubsection{Multilayer Perceptron}

The neural network is a well studied and widely used method for both regression and classification tasks \cite{hornik1989multilayer}. When using the rectified linear unit (ReLU) activation function \cite{dahl2013improving} and training with the BFGS minimization technique \cite{moller1993scaled}, the model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i} \rightarrow \mathbb{R}^{j}$ defined by

$$ l(u) = \big( u^t W_l \big)_+ ,$$

\noindent where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this form, the multilayer perceptron produces a piecewise linear model of the input data. The computational complexity of training a multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is determined by the sizes of the layers of the network and the stopping criterion of the BFGS minimization used for finding weights. This paper uses the scikit-learn MLP regressor \cite{scikit-learn}, a single hidden layer with 100 nodes, ReLU activation, and BFGS for training.

\subsubsection{Decision Tree}

The decision tree is used because of the relatively straightforward interpretation of the prediction process. Model construction is out of the scope of this description, but is a well-studied process \cite{quinlan1986induction}. A prediction at a point $z \in \mathbb{R}^d$ for a decision tree constructed over a real vector space is made by traversing nested axis-aligned conditionals of the form
$$\hat f(z) = \hat f(z | z_{k^{(1)}} \geq v^{(1)}, \ldots).$$

This paper uses the scikit-learn Decision Tree regressor \cite{scikit-learn}, no maximum depth or number of nodes, and the Gini impurity measure of information gain.

\subsection{Local Predictive Models}

The construction of approximation functions $\hat f_P: \mathbb{R}^d \rightarrow \mathbb{R}$ as described for global models can instead be approached on a per-prediction basis. A model is henceforth referred to as \textit{local} when any prediction made at a point $z \in \mathbb{R}^d$ is only a function of a set of points $L \subset X$, where membership in $L$ is determined by a distance metric. The advantage of using a local model is a more compact description of \textit{how} a prediction is made that is derived from a manageable subset of all known data. \textit{Local} models become particularly useful when predictions regard ethically sensitive issues and need to be rigorously evaluated for bias. The source data for any prediction can be checked on the spot for fairness of representation against any number of protected attributes.

The following sections describe the two local approximation techniques that will be used to predict recidivism likelihood in this work.

\subsubsection{Nearest Neighbor}

A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}. This algorithm (using the 2-norm to measure distance) will represent a baseline for comparison because it is the most mathematically simple \textit{local} model in this study. A prediction is made for Nearest Neighbor at point $z \in \mathbb{R}^d$ by
$$ \hat f(z) = f \big ( \text{\textit{argmin}}_{x^{(i)} \in X} \| z - x^{(i)} \|_2 \big). $$
This approximation technique can be applied in a wide range of applications, however it must be noted that the approximation surface it produces is not $C^0$ (continuous in value). An extension of this model that is not applied in this work, but could yield useful results is the $k$ nearest neighbor algorithm. This is further mentioned in Section \ref{sec:discussion}.

\subsubsection{Voronoi Mesh}

While nearest neighbor inherently utilizes the convex region $v^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than any other point $x^{(j)}$. The Voronoi mesh \cite{lux2018novel} smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $V: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by

$$ V^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2 \ d(y \mid x^{(i)})} \right)_+, $$
where $x^{(i)}$ is the center of the Voronoi cell, $y \in \mathbb{R}^d$ is an interpolation point, and $d(y \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $v^{x^{(i)}}$ in the direction $y - x^{(i)}$. $V^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $V^{x^{(i)}}$ has local support. While $V^{x^{(i)}}(x^{(i)}) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell. Note that this basis function is $C^0$ because the boundaries of the Voronoi cell are $C^0$. In the case that there is no boundary along the vector $w$, the basis function value is always $1$.

While the cost of computing the exact Voronoi cells for any given set of points grows exponentially \cite{dutour2009complexity}, the calculation of $d$ is linear with respect to the number of control points and dimensions. Given any center $x^{(i)} \in \mathbb{R}^d$, set of control points $C \subseteq X$, and interpolation point $y \in \mathbb{R}^d$, $d\bigl(y \mid x^{(i)}\bigr)$ is the solution to
$$
  \max_{c \in C\backslash\{x^{(i)}\}} \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2} \ \frac{y \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}{c \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}.
$$

The resulting algorithm is capable of producing predictions in $\mathcal{O}(n^2 d)$ computation time, which is relatively fast for all but tens of millions of points. Most importantly there is no \textit{training} for this algorithm, so as data is updated the fundamental model itself is changed accordingly.

%% ===================================================================

\begin{table*}
\begin{center}
\begin{tabular}{ c | c }
 \hline
 \textbf{FEATURE NAME} & \textbf{VALUES} \\
 \hline
 Sex & F, M \\
  \hline
 Age At Release & Under 25, 25-34, 35-44, 45-54, 55 and Older \\  
  \hline
 Convicting Offense Classification & Aggravated Misdemeanor, Serious Misdemeanor, Sexual Predator Community Supervision, etc \\
  \hline
Convicting Offense Subtype & Murder, Alcohol, Weapons, Drug Possession, Assault, Traffic, Burglary, Forgery/Fraud, Animals, Theft, etc\\  
 \hline
Release Type & Special Sentence, Parole Granted, Discharged - Expiration of Sentence, Released to Special Sentence, etc \\  
 \hline
Main Supervising District & 1JD, 2JD, 3JD, 4JD, 5JD, etc\\  
 \hline
Race -- Ethnicity & White (non-Hispanic), Black or African American, White (Hispanic), American Indian or Alaskan\\
 \hline
\end{tabular}
\end{center}
\caption{This table provides a sample of the values for features relevant to the recidivism prediction task.}
\label{table:1}
\end{table*}

\section{DATA AND RESULTS}
\label{sec:data}

A case study is presented to demonstrate the comparative performance of \textit{local} models versus \textit{global} models on an ethically sensitive prediction task. The comparison rests on a 3-Year Recidivism for Offenders Released from Prison in Iowa dataset, publicly available on the U.S. Government open data (\href{https://www.data.gov}{data.gov}) website. This dataset reports whether or not an offender is convicted of another crime within three years of being released from prison. This dataset naively provides a classification task (predict whether or not a released offender would recidivate) or, as will be considered here, it also provides a regression task (predict the probability of recidivism). By collapsing all entries with identical predictor values, this work converts each offender description into a collection of matching individuals with a single real-valued estimate for the probability of recidivism.

This dataset provides $21,646$ instances, among which $14,619$ instances do not recidivate and $7,027$ instances do recidivate. The distribution of ethnicities in this data are distinctly different from the ethnic distribution for the actual population of Iowa in 2016 (See Figure \ref{fig:mesh1}). There exists obvious gaps especially for White non-Hispanic ($67.4\%$ in data \textit{decreased} from $88.7\%$ of actual population) and Black or African American ($23.6\%$ in data \textit{increased} from  $2.9\%$ of actual population).

\begin{figure}
  \includegraphics[width=.48\textwidth]{regression_values_histogram.pdf}
  \caption{A 20-bin histogram of the values for recidivism likelihood in the regression task composed of 6,632 samples.}
  \label{fig:histogram_of_recidivism}
\end{figure}

\begin{figure}
    \centering
\includegraphics[scale=0.2]{figure_2.png}
\includegraphics[scale=0.2]{figure_1.png}
    \caption{The distribution across ethnicity groups of the dataset (left) and the actual population in Iowa (right)}
    \label{fig:mesh1}
\end{figure}

The data provides 17 unique pieces of information related to each individual, including the crime committed in the case of recidivism (blank for non-recidivating individuals). The relevant features to this prediction task can be observed in Table \ref{table:1}. The omitted features are either not relevant to the prediction task, or are uniquely determined by one of the chosen features. For this work the ``Race -- Ethnicity'' is reserved as protected attribute (not available to models) and is later used to evaluate bias in predictions. Any categories with fewer than 100 samples are discarded in order to reduce the resulting dimension and prevent predictions from being made with statistically insignificant amounts of supporting data. Categorical features are encoded with $c$ unique categories into $\mathbb{R}^{c-1}$ by mapping each category onto one of the vertices of a regular simplex centered at 0 where all vertices $v$ satisfy $\|v\|_2 = 1$. The feature named ``Age At Release'' was tested as both the mean age of the range and as a mapped categorical. Results demonstrated insignificant differences in outcome, so the numerical mean age is used for experiments.

In all experiments, predictions are evaluated against the labeled recidivism probability while the race \& ethnicity information is used to evaluate model bias. After preprocessing, the original $21,018$ instances with $17$ features are reduced to $6,632$ in a $50$ dimensional real vector space. The distribution of recidivism probabilities can be seen in Figure \ref{fig:histogram_of_recidivism}.

In order to estimate the performance of the different algorithms, \textit{k-fold} cross validation as described in \cite{kohavi1995study} with $k=10$ is used. All algorithms are given the same ten folds of randomized training and testing data in order to maintain comparative fairness. Note that in this scheme there will be exactly one prediction made for each data point, meaning all analysis of results is done with the same sized data as described in Section \ref{sec:data}.

Figure \ref{fig:regression_performance} displays the evaluation of all algorithms. The Voronoi Mesh and Neural Network produce the best results. The prediction outcomes are promising, demonstrating that 50\% of recidivism likelihood predictions have less than a $16\%$ absolute error without any attempt at problem-specific tuning.

%% ===================================================================

\section{DISCUSSION}
\label{sec:discussion}

As can be observed in Figure \ref{fig:regression_performance}, the Voronoi Mesh (VM) algorithm can make a prediction based on roughly $2d$ (\thomastilde{}100) points from data and compete with the global fitting MLP that uses all data (\thomastilde{}7K) points. This demonstrates that local predictive models are capable of producing equally accurate predictions when compared with global predictive models with far more compact support from data.

The benefit of using the VM (or any local model) to make predictions in ethically sensitive applications is that \textit{every} prediction has a manageable set of source data that describes how a prediction is produced. Addressing the two points in Section \ref{sec:introduction}, statistical tests can be run on these source data points to reduce prediction bias in desirable ways (e.g. data could be filtered until the predictive population matches the demographics of the state). Human-readable source for a prediction also opens the door to regular validation of source data, an important aspect of model maintenance.

In order to maintain ethical conscientiousness in predictive modeling, it is vital that predictions can be audited for fairness and adherence to standards. Local predictive models provide a meaningful avenue for pursing the legal right to a representative sample, and the legal right to a fair prediction.

This case study provides only a glimpse at the prospective application of local models. There are far more advanced (and potentially more accurate models) with very similar properties: Delaunay triangulations (simplicial meshes) \cite{chang2018polynomial}, $k$ nearest neighbor, Linear Shepard \cite{thacker2010algorithm} (and other Shepard methods), and Box Spline Meshes \cite{lux2018novel}, name only a few.

% Almost all machine learning algorithms can be treated local models in which we use solely a fixed-length moving time window for training. This step would make the computational time smaller, giving us a great chance to apply it on online complex problems that cannot handle any delay in the prediction. Yet, this approach would not be ideal if there is a core knowledge that cannot be discovered without running the entire dataset.

% One must also note that the Asian or Pacific Islander category was also the smallest population size by far, meaning that having the highest percentage of bias from false positives is expected for all 4 models.

\begin{figure*}
\centering
\begin{tabular}{l r}

  % NearestNeighbor
  \begin{tabular}{c c}
    \vspace{3mm} \\
    \multicolumn{2}{c}{\textit{NearestNeighbor}} \\
    \multicolumn{2}{c}{ \includegraphics[width=2in,height=2in]{NearestNeighbor_regression.pdf}} \vspace{3mm} \\
    \begin{tabular}{c|c}
        Min & 0.000 \\
        $25^{th}$ & 0.033 \\
        $50^{th}$ & 0.167 \\
        $75^{th}$ & 0.400 \\
        Max & 1.000 \\
    \end{tabular} &
    \begin{tabular}{|c l|}
      \hline
      0.226 & White - NH \\
      0.250 & Black - NH \\
      0.200 & White - H \\
      0.167 & AI or NA \\
      0.306 & Asian or PI \\
      \hline
    \end{tabular} \vspace{3mm} \\
  \end{tabular}

  &

  % VoronoiMesh
  \begin{tabular}{c c}
    \vspace{3mm} \\
    \multicolumn{2}{c}{\textit{VoronoiMesh}} \\
    \multicolumn{2}{c}{ \includegraphics[width=2in,height=2in]{VoronoiMesh_regression.pdf}} \vspace{3mm} \\
    \begin{tabular}{c|c}
        Min & 0.000 \\
        $25^{th}$ & 0.072 \\
        $50^{th}$ & 0.156 \\
        $75^{th}$ & 0.291 \\
        Max & 0.974 \\
    \end{tabular} &
    \begin{tabular}{|c l|}
      \hline
      0.148 & White - NH \\
      0.161 & Black - NH \\
      0.125 & White - H \\
      0.129 & AI or NA \\
      0.139 & Asian or PI \\
      \hline
    \end{tabular} \vspace{3mm} \\
  \end{tabular}

  \\

  % DecisionTreeRegressor
  \begin{tabular}{c c}
    \vspace{3mm} \\
    \multicolumn{2}{c}{\textit{DecisionTreeRegressor}} \\
    \multicolumn{2}{c}{ \includegraphics[width=2in,height=2in]{DecisionTreeRegressor_regression.pdf}} \vspace{3mm} \\
    \begin{tabular}{c|c}
        Min & 0.000 \\
        $25^{th}$ & 0.040 \\
        $50^{th}$ & 0.187 \\
        $75^{th}$ & 0.433 \\
        Max & 1.000 \\
    \end{tabular} &
    \begin{tabular}{|c l|}
      \hline
      0.250 & White - NH \\
      0.286 & Black - NH \\
      0.246 & White - H \\
      0.264 & AI or NA \\
      0.257 & Asian or PI \\
      \hline
    \end{tabular} \vspace{3mm} \\
  \end{tabular}

  &

  % MLPRegressor
  \begin{tabular}{c c}
    \vspace{3mm} \\
    \multicolumn{2}{c}{\textit{MLPRegressor}} \\
    \multicolumn{2}{c}{ \includegraphics[width=2in,height=2in]{MLPRegressor_regression.pdf}} \vspace{3mm} \\
    \begin{tabular}{c|c}
        Min & 0.000 \\
        $25^{th}$ & 0.068 \\
        $50^{th}$ & 0.145 \\
        $75^{th}$ & 0.279 \\
        Max & 1.273 \\
    \end{tabular} &
    \begin{tabular}{|c l|}
      \hline
      0.147 & White - NH \\
      0.153 & Black - NH \\
      0.134 & White - H \\
      0.113 & AI or NA \\
      0.114 & Asian or PI \\
      \hline
    \end{tabular} \vspace{3mm} \\
  \end{tabular}

\end{tabular}

  \caption{These four plots show the true recidivism probability versus guessed recidivism probability for each of the regression techniques. The top two regression algorithms make predictions based only on local data while the bottom two algorithms are global fitting techniques. The left vertical table beneath each figure displays the percentiles of absolute errors when predicting the probability of recidivism with that algorithm. The right table beneath each figure shows the median error in recidivism likelihood for those predictions which were over-estimated (false positives) broken down by race. Notice that without additional constraints, the neural network produces predictions outside of the range [0,1].}
  \label{fig:regression_performance}
\end{figure*}

%% ===================================================================

\section{CONCLUSION}
\label{sec:conclusion}

This paper presents an argument for the application of local models to ethically sensitive prediction tasks. Sample results demonstrate that algorithms which rely only on local support are capable of producing predictions of comparable accuracy to popular global techniques while maintaining an enhanced level of predictive transparency. The potential for operating under more concise legal definitions and meaningful statistical analyses further supports the implementation of local prediction methodologies. A recidivism case study demonstrates that the use of more human-interpretable models for prediction could not only strengthen the ethical application of data science, but also make the process of cleaning and validating data manageable in the long term.

%% ===================================================================

\bibliographystyle{IEEEtran}
\bibliography{paper}

\end{document}

### 2020-01-16 [2019-10]_ISTAS_(ethical_prediction)/[2]-organize_results.py

@@ -1,0 +1,77 @@
+from setup import *
+
+for dt in data_type:
+    if "Classification" in dt:
+        algorithms = classification_algorithms
+    else:
+        algorithms = regression_algorithms
+
+    for at in age_type:
+        print(dt, at)
+        #      Get the predictor and target columns     
+        # ==============================================
+        if "Classification" in dt:
+            if "Cat" in at:
+                predictors = class_cat_preds
+            else:
+                predictors = class_num_preds
+        else:
+            if "Cat" in at:
+                predictors = regrs_cat_preds
+            else:
+                predictors = regrs_num_preds
+
+        #      Cycle through predicting for each algorithm     
+        # =====================================================
+        folder = os.path.join(main_folder, dt, at)
+        
+        raw_data = Struct().load(os.path.join(folder, "raw_data.csv"), sep=",")
+        nums, info = raw_data.to_numpy_real()
+        # Remove the "numeric" columns that are counts of people from normalization
+        to_normalize = info.nums[:]
+        while any(n >= 45 for n in to_normalize):
+            to_normalize.remove(max(to_normalize))
+        # Get the normalization numbers used
+        shift = np.min(nums[:,to_normalize], axis=0)
+        nums[:,to_normalize] -= shift
+        scale = np.max(nums[:,to_normalize], axis=0)
+        nums[:,to_normalize] /= scale
+        # Remove this useless numeric data
+        del nums
+
+        # Initialize a holder for all testing data
+        all_tests = Struct()
+
+        # Cycle through all folds
+        for fold in range(NUM_FOLDS):
+            fold += 1
+            test_file = os.path.join(folder, f"{fold:02d}-10_test.csv")
+            # Load the test data file
+            test_data = Struct().load(test_file)
+            # Get the names of the predictors and the target
+            predictor_names = [test_data.names[i] for i in predictors]
+            # Iterate over the algorithms
+            for alg in algorithms:
+                # Generate an output file and save the data to it
+                output_file = os.path.join(folder,"Predictions",f"{alg.__name__}_{fold:02d}-10.csv")
+                # Load the prediction output
+                prediction_data = Struct().load(output_file, sep=",")
+                # Add the prediction output for this algorithm to the test data
+                test_data.add_column(prediction_data[:,0], prediction_data.names[0])
+            # Concatenate newest test data
+            all_tests += test_data
+
+        # Convert the test results back into their categorical form.
[…]

### 2020-01-16 [2019-10]_ISTAS_(ethical_prediction)/[3]-analyze_results.py
Created paper repository with all of my work.

@@ -1,0 +1,191 @@
+from setup import *
+
+for dt in data_type:
+    if "Classification" in dt:
+        algorithms = classification_algorithms
+    else:
+        # Delaunay
+        algorithms = regression_algorithms
+
+    for at in age_type:
+        print(dt, at)
+        #      Get the predictor and target columns     
+        # ==============================================
+        if "Classification" in dt:
+            if "Cat" in at:
+                predictors = class_cat_preds
+            else:
+                predictors = class_num_preds
+        else:
+            if "Cat" in at:
+                predictors = regrs_cat_preds
+            else:
+                predictors = regrs_num_preds
+
+        #      Cycle through predicting for each algorithm     
+        # =====================================================
+        folder = os.path.join(main_folder, dt, at)
+        
+        # Load the raw data and get the transformation function
+        raw_data = Struct().load(os.path.join(folder, "raw_data.csv"), sep=",")
+        nums, info = raw_data.to_numpy_real()
+
+        # Load the output of all predictions
+        results = Struct().load(os.path.join(folder,"all_predictions.csv"), sep=",")
+        target = [n for n in results.names if "Recidivism" in n][0]
+        predictions = [n for n in results.names if "Prediction" in n]
+
+        if "Classification" in dt:
+            _, info = Struct(data=[["Yes"],["No"]]).to_numpy_real()
+            truth = np.array([info.to_real([v]) for v in results[target]])
+            print()
+            print(f"  Yes -> {info.to_real(['Yes'])}")
+            print(f"  No --> {info.to_real(['No'])}")
+            print()
+        else:
+            truth = np.array(list(results[target]))
+            p = Plot("","Probability of Recidivism","Count",font_family="times")
+            p.add_histogram("", truth, show_in_legend=False, num_bins=20)
+            p.plot(width=400, height=200, file_name="regression_values_histogram.html")
+
+        guess, _ = results[predictions].to_numpy_real()
+
+        if "Classification" in dt:
+            # Get the name of the protected column
+            protected = 0
+            tgt_idx = results.names.index(target)
+            races = sorted(set(v for v in results[results.names[protected]]))
+            print()
+            for i in range(len(algorithms)):
+                print(f"  {algorithms[i]}")
+                cm = metric(truth, guess[:,i])
[…]

### 2020-01-16 [2019-10]_ISTAS_(ethical_prediction)/example.py

@@ -1,0 +1,16 @@
+
+# data_file = "3-Year_Recidivism_for_Offenders_Released_from_Prison.csv"
+
+# from util.data import Struct
+
+# data = Struct()
+# data.load(data_file)
+# data.summarize()
+
+# real_data = data.to_numpy_real()
+
+# from util.algorithms import Delaunay
+
+# model = Delaunay()
+# model.fit(real_data[:.5*len(real_data)])
+# model(real_data

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/paper/paper.tex
Created paper repository with all of my work.

%**************************************************************************
%* SpringSim 2020 Author Kit
%*
%* Word Processing System: TeXnicCenter and MiKTeX
%*
%**************************************************************************

\documentclass{scspaperproc}

\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{mathptmx}

%
%****************************************************************************
% AUTHOR: You may want to use some of these packages. (Optional)
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{amsthm}
%****************************************************************************

%
%****************************************************************************
% AUTHOR: If you do not wish to use hyperlinks, then just comment
% out the hyperref usepackage commands below.

%% This version of the command is used if you use pdflatex. In this case you
%% cannot use ps or eps files for graphics, but pdf, jpeg, png etc are fine.

\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black,bookmarks=false]{hyperref}

%% The next versions of the hyperref command are used if you adopt the
%% outdated latex-dvips-ps2pdf route in generating your pdf file. In
%% this case you can use ps or eps files for graphics, but not pdf, jpeg, png etc.
%% However, the final pdf file should embed all fonts required which means that you have to use file
%% formats which can embed fonts. Please note that the final PDF file will not be generated on your computer!
%% If you are using WinEdt or PCTeX, then use the following. If you are using
%% Y&Y TeX then replace "dvips" with "dvipsone"

%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
%% anchorcolor=black,linkcolor=black]{hyperref}

%% The use of the long citation format (e.g. "Brown and Edwards (1993)" rather than "[5]") and at the same
%% time using the hyperref package can lead to hard to trace bugs in case the citation is broken accross the
%% line (usually this will mark the entire paragraph as a hyperlink (clickable) which is easily noticeable and fixed
%% if using colorlinks, but not if the color is black -- as it is now). Worse yet, if a citation spans page boundary,
%% LaTeX compilation can fail, with an obscure error message. Since this depends a lot on the flow of the text
%% and wording, these bugs come and go and can be extremely hard for a beginner to trace. The error
%% message can look like this:
%%
%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink.
%%    \AtBegShi@Output ...ipout \box \AtBeginShipoutBox 
%%    \fi \fi 
%%    l.174 
%%    ! ==> Fatal error occurred, no output PDF file produced!
%%
%% and can be universally fixed by putting an \mbox{} around the citation in question (in this case, at line 174)
%% and maybe adapting the wording a little bit to improve the paragraph typesetting, which is perhaps not
%% immediately obvious.
%****************************************************************************

%****************************************************************************
%*
%* AUTHOR: YOUR CALL!  Document-specific macros can come here.
%*
%****************************************************************************

%% Use a package to allow for multi-row cells in a table.
\usepackage{multirow}

%% Use a custom `algorithm` command for this paper.
\input{lux-algorithm}

% add custom hyphenation rules here
\usepackage{hyphenat}
\hyphenation{op-tical net-works semi-conduc-tor}

% If you use theorems
\newtheoremstyle{scsthe}% hnamei
{8pt}% hSpace abovei
{8pt}% hSpace belowi
{\it}% hBody fonti
{}% hIndent amounti1
{\bf}% hTheorem head fontbf
{.}% hPunctuation after theorem headi
{.5em}% hSpace after theorem headi2
{}% hTheorem head spec (can be left empty, meaning `normal')i

\theoremstyle{scsthe}
\newtheorem{theorem}{Theorem}
\renewcommand{\thetheorem}{\arabic{theorem}}
\newtheorem{corollary}[theorem]{Corollary}
\renewcommand{\thecorollary}{\arabic{corollary}}
\newtheorem{definition}{Definition}
\renewcommand{\thedefinition}{\arabic{definition}}

% avoid overrunning the right margin; you are welcome to remove this, provided that you take care not to overrun the right margin anywhere in your paper
\sloppy

%#########################################################
%*
%*  The Document.
%*
\begin{document}

%***************************************************************************
% AUTHOR: Thomas C.H. Lux et al.
% FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
%
%		You need to change the author listing below!
%               Please list ALL authors using last name only, separate by a comma except
%               for the last author, separate with "and"

\SCSpagesetup{Lux, Watson, Chang, Xu, Wang, and Hong}

% AUTHOR: Uncomment ONE of these correct conference names.
\def\SCSconferenceacro{SpringSim'20}
%\def\SCSconferenceacro{SummerSim}
%\def\SCSconferenceacro{AutumnSim}
%\def\SCSconferenceacro{PowerPlantSim}

% AUTHOR: Set the correct year of the conference.
\def\SCSpublicationyear{2020}

% AUTHOR: Set the correct month and dates; the dates are separated by a single minus sign
% with no spaces and no leading zeros, the month is a full name (e.g. April) with the first letter
% capitalized. For example, "April 8-13".
\def\SCSconferencedates{May 19-May 21}

% AUTHOR: Set the correct venue in the form "City, State, Country", for example "Los Angeles, CA, USA".
\def\SCSconferencevenue{Fairfax, VA, USA}

% AUTHOR: Enter the title, all letters in upper case
\title{An Algorithm for Constructing Monotone \\ Quintic Interpolating Splines}

% AUTHOR: Enter the authors of the article, see end of the example document for further examples
\author{
Thomas C. H. Lux\\
Layne T. Watson\\
Tyler H. Chang\\ [12pt]
Departments of Computer Science, Mathematics,\\
and Aerospace and Ocean Engineering\\
Virginia Polytechnic Institute \& State University\\
2000 Torgersen Hall, Blacksburg, VA, USA\\
\{tchlux,ltw,thchang\}@vt.edu\\
\and
Li Xu\\%
Yueyao Wang\\%
Yili Hong\\ [12pt]
Department of Statistics\\
Virginia Polytechnic Institute \& State University\\
213 Hutcheson Hall, Blacksburg, VA, USA\\
\{lix1992,yueyao94,yilihong\}@vt.edu\\
}

\maketitle

%% Examples and notes.
%% 
%% \begin{table}[htb]
%%   \centering
%%   \caption{Caption above the table!}\label{tab:example}
%%   \begin{tabular}{r|l}
%%     First & row \\ \hline
%%     second & row \\
%%     third & row \\
%%   \end{tabular}
%% \end{table}
%% 
%% \begin{figure}[htb]
%%   \centering
%%   \includegraphics[width=0.50\textwidth]{path_without_extension}
%%   \caption{Caption below the figure!}\label{fig:example}
%% \end{figure}
%% 
%% \cite{ref}  - Make a citation that is in parenthesis
%% \citeN{ref} - Make a citation out of parenthesis (for noun usage)
%% \shortcite{ref}  - Make a citation for references with >3 authors.
%% \shortciteN{ref} - Make a noun-form citation for references with >3 authors.
%% 
%% 
%% 5 - 12 page paper mandated
%% 150 word abstract
%% 5 keywords maximum

\section*{Abstract}

An algorithm for computing monotone order six piecewise polynomial
interpolants is proposed. Algebraic constraints for enforcing
monotonicity are provided that align with quintic monotonicity
theory. The algorithm is implemented, tested, and applied to several
sample problems to demonstrate the improved accuracy of monotone
quintic spline interpolants compared to that for existing monotone
cubic spline interpolants.

\textbf{Keywords:} monotone, quintic spline, Hermite interpolation, sixth order polynomial.

\section{Introduction and Motivation}
\label{sec:introduction}

%% \begin{figure}[htb]
%%   \centering
%%   \includegraphics[width=0.50\textwidth]{demo_projection}
%%   \caption{These are the projection mechanisms that make a cubic polynomial piece monotone.}\label{fig:projection}
%% \end{figure}

%% The ability to interpolate or regress a function with continuous value and derivatives has proven fundamentally useful. In many applications, there are known shapes or characteristics that are expected of approximations. For example, a cumulative distribution function in statistics cannot decrease, the entropy of a closed thermodynamic system can never increase, an electric motor should not suddenly change direction when slowing down or speeding up. In each of these cases it is important that any constructed spline approximation is monotone.

Many domains of science rely on smooth approximations to real-valued functions over a closed interval. Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate nonparametric approximations in statistics \cite{knott2012interpolating}. While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.

It is often the case that the true underlying function or phenomenon being modeled has known properties e.g., convexity, positivity, various levels of continuity, or monotonicity. Given a reasonable amount of data, it quickly becomes difficult to achieve desirable properties in a single polynomial function. In general, the maintenance of function properties through interpolation/regression is referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}. The specific shapes this work will achieve in approximations are monotonicity and $C^2$ continuity. These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.

In statistics especially, the construction of a monotone interpolating spline that is $C^2$ continuous is meaningfully useful \cite{ramsay1988monotone}. A function with these properties could approximate a cumulative distribution function to a high level of accuracy with relatively few intervals. A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a property many standard parametric distributions maintain.

The currently available software for monotone piecewise polynomial interpolation includes quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases. In addition, a statistical method for bootstrapping the construction of an arbitrarily smooth monotone fit exists \cite{leitenstorfer2006generalized}, but the method does not take advantage of known analytic properties related to quintic polynomials. Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, however this theory has not yet been used to construct a monotone quintic spline interpolation routine. Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \shortcite{xie2018semiparametric}.

\begin{figure}
  \centering
  \includegraphics[width=.9\textwidth]{spline_demonstration}
  \caption{Example polynomials that interpolate function values at the ends of the interval $[0,1]$. The first only interpolates the function values $g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial $g_2(x) = 1 - x$. For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x + 1/2$, which is order four and interpolates the values $g_4(0) = 1/2$, $g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$. Finally the third plot shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 + 50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$, $g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$, $g_6''(1) = -100$. Notice that interpolating the same fixed number of function values at each endpoint will always result in an even order interpolating polynomial.
  }\label{fig:spline_demonstration}
\end{figure}

The importance of piecewise quintic interpolation over lower order approximations can be simply demonstrated. In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values. As demonstrated in Figure \ref{fig:spline_demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two. The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on quintic splines.

%% Software exists that does monotone piecewise quintic regression \cite{}, but the available routines do not computationally take advantage of the quintic monotonicity theory. ???

The major contribution of this work is an algorithm for constructing monotone quintic interpolating splines that utilizes existing quintic monotonicity theory. The remainder of this paper is structured as follows: Section \ref{sec:monotone-cubic} summarizes the existing monotone cubic spline interpolation methodology, Section \ref{sec:monotone-quintic} presents an algorithm for constructing monotone quintic spline interpolants, Section \ref{sec:results} offers experiments and results with cubic and quintic methods, Sections \ref{sec:discussion} and \ref{sec:conclusion} discuss results and conclude.

\subsection{Computing a Monotone Cubic Interpolant}
\label{sec:monotone-cubic}

The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by

\begin{align*}
h_i &= k_{i+1} - k_{i}, \quad
u(t) = 3t^2 - 2t^3, \quad
p(t) = t^3 - t^2,\\
\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
p\big((x-k_i)/h_i\big).
\end{align*}

Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$. However, such a spline has too many \textit{wiggles} for most applications. \citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. Consider the terms $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$. If $f(k_i) = f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$. Monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{cubic_projection_demonstration}
  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone. The regions themselves are numbered 1--4 corresponding to their original description in Fritsch and Carlson (1980). One point is projected onto each region as a demonstration.}\label{fig:projection}
\end{figure}

The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions. The user must decide which region is most appropriate for the projections based on the application, Fritsch and Carlson recommend using Region 2.

While the cubic monotonicity case affords such a concise solution, the region of monotonicity is not so simple in the quintic case. In the next section, an algorithm for performing a projection similar to those for cubic polynomials is proposed.

\section{Monotone Quintic Interpolation}
\label{sec:monotone-quintic}

The following section is composed of three algorithms that together are used to construct a monotone quintic spline interpolant. Without loss of generality, the algorithms will only consider the monotone increasing (nondecreasing) case. The monotone decreasing (nonincreasing) case is handled similarly. Algorithm \ref{alg:check-monotone} checks monotonicity,
Algorithm \ref{alg:make-monotone} enforces monotonicity on an order six
polynomial piece of the quintic spline, and Algorithm
\ref{alg:monotone-spline} uses the previous two algorithms to enforce
monotonicity for the entire quintic spline.

In pseudocode, a function \texttt{binary\_search($g,$ $a,$ $b$)} is used, where $a,b\in S\subset \mathbb{R}^p$ for convex $S$, $g: S \rightarrow \{$\texttt{FALSE}, \texttt{TRUE}$\}$ is a right continuous Boolean function, $g(b) = \hbox{\texttt{TRUE}}$, and for $\mu \in [0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{\texttt{TRUE}} \Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{\texttt{TRUE}}$ for $\mu \le \nu\le1$. The search returns $\big((1-c)a + c b\big)$ for the smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) = \hbox{\texttt{TRUE}}$.

These algorithms make use of first and second derivatives of the approximated function. In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values. The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.

\algorithm
    {is\_monotone$\big(x_0, x_1, f \big)$}
    {where $x_0$, $x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
  $Df(x_1)$, $D^2f(x_1)$. Returns \texttt{TRUE} if $f$ is monotone increasing on $[x_0,x_1]$.}
    {\label{alg:check-monotone}
  \item if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
    \indented{return FALSE}
    end if
  \item if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$
    \indented{return FALSE}
    end if
    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$. The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$. Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \citeN{schmidt1988positivity}. If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
  \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
  \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
  \item \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE}
    \label{alg1-alpha}
  \item \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \label{alg1-beta}
  \item \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE}
    \label{alg1-gamma}
  \item \indented{return TRUE} \label{alg1-simplified-true}
    end if
    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
  \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \citeN{ulrich1994positivity}, first defined after Equation (18).}
  \item $\displaystyle \gamma_0 := 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
    \\ $\displaystyle \gamma_1 := \frac{x_1 - x_0}{Df(x_1)} (B/A)^{3/4}$
    \\ $\alpha_0 := 4 (B/A)^{1/4}$
    \\ $\displaystyle \alpha_1 := -\frac{x_1 - x_0}{Df(x_1)} (B/A)^{1/4}$
    \\ $\displaystyle \beta_0 := 30 - \frac{12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
    \\ $\displaystyle \beta_1 := \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \label{alg1-linearized-abg}
    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \citeN{ulrich1994positivity}. These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.}
  \item $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$
    \\ $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$
    \\ $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$
    \label{alg1-abg}
  \item if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$
    \\ else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$
    \\ end if \label{alg1-last}
}
%% ----------------------------------------------------------------------

The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later. The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval. Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.

%% ----------------------------------------------------------------------
\algorithm
    {make\_monotone$\big (x_0$,$x_1$,$f \big)$}
    {where $x_0,\ x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1)$. Returns $f$ monotone on $[x_0,x_1]$.}
    {\label{alg:make-monotone}
    \item if $\big(f(x_1) = f(x_0)\big)$
      \indented{$Df(x_0) := Df(x_1) := D^2f(x_0) := D^2f(x_1) := 0$}
      \indented{return}
      end if
    \item $Df(x_0) :=\ $median$\displaystyle \bigg(0,\ Df(x_0),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \bigg)$
      \\ $Df(x_1) :=\ $median$\displaystyle \bigg(0,\ Df(x_1),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\bigg)$ \label{alg2-init-df}
      \comment{The selection of values in Step \ref{alg2-init-df} for $Df(x_0)$ and $Df(x_1)$ is suggested by \citeN{ulrich1994positivity} and also by \citeN{huynh1993accurate}. These rules quickly enforce upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
      \comment{The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$. Steps \ref{alg2-dx-conditions} and \ref{alg2-rescale-dx} ensure that the following bounds on $D^2f(x_0)$ are compatible while Step \ref{alg2-ddx0-upper} ensures that the conditions on $D^2f(x_1)$ are compatible. To conclude the simplified case, Step \ref{alg2-gamma} first enforces $\gamma \geq \delta$, then Steps \ref{alg2-alpha} and \ref{alg2-beta} enforce $\alpha \geq 0$ and $\beta \geq \alpha$ respectively. Combined, these three conditions derived from Equation (2.18) in \citeN{schmidt1988positivity} guarantee monotonicity of this polynomial piece.}
    \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
    \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
    \item \indented{if $\big(5 Df(x_0) + 4Df(x_1) > 20v / w)$} \label{alg2-dx-conditions}
    \item \indented{\indented{$Df(x_0) := Df(x_0) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}}
      \indented{\indented{$Df(x_1) := Df(x_1) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}} \label{alg2-rescale-dx}
      \indented{end if}
    \item \indented{$D^2f(x_0) :=\ $min$\displaystyle \bigg(D^2f(x_0), \frac{4 (2 Df(x_0)+Df(x_1))+20 v/w}{w} \bigg)$} \label{alg2-ddx0-upper}
    \item  \indented{$D^2f(x_0) :=\ $max$\big(D^2f(x_0), 3 Df(x_0) / w \big)$} \label{alg2-gamma}
    \item \indented{$D^2f(x_1) :=\ $min$\big(D^2f(x_1), {-4}Df(x_1) / w \big)$} \label{alg2-alpha}
    \item  \indented{$D^2f(x_1) :=\ $max$\displaystyle \bigg(D^2f(x_1), \frac{3w D^2f(x_0) - 24 Df(x_0) - 32 Df(x_1) + 60v/w}{5w} \bigg)$} \label{alg2-beta}
      \indented{return}
      end if
      \comment{The following code considers the case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
    \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
      \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \item if $\big($max$(A,B) > 6\big)$
      \indented{$Df(x_0) := 6 Df(x_0) \big/ $max$(A,B)$}
      \indented{$Df(x_1) := 6 Df(x_1) \big/ $max$(A,B)$}
      end if
      \comment{This ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \citeN{ulrich1994positivity}).}
    \item $\eta := \big(D^2f(x_0), D^2f(x_1)\big)$
      \\ $\displaystyle \eta_0 := \left({-\frac{\sqrt{A}}{4}}\big(7\sqrt{A} + 3\sqrt{B}\big),\ \frac{\sqrt{B}}{4}\big(3\sqrt{A} + 7\sqrt{B}\big) \right)$
    \item $\bigl(D^2f(x_0)$, $D^2f(x_1)\bigr) :=$ binary\_search$\bigl(g,
       \eta,\eta_0\bigr)$
      \comment{Here $g(z) =$\texttt{ is\_monotone}$(x_0,x_1,f)$ where $x_0$, $x_1$, $f(x_0)$, $f(x_1)$, $Df(x_0)$, $Df(x_1)$ are fixed, and the variable $z=\bigl(D^2f(x_0)$,$D^2f(x_1)\bigr)$. This binary search identifies feasible values of $D^2f(x_0)$ and $D^2f(x_1)$ that are nearest to the current values $\eta$. This search is valid because $\eta_0$ is guaranteed to produce a monotone function, which can be seen in Equation (23) of \citeN{ulrich1994positivity}.}
    return
}
%% ----------------------------------------------------------------------

Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have $\mathcal{O}(1)$ runtime  assuming a fixed level of relative precision. A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}). The search requires a fixed number of steps to achieve any predetermined relative precision. Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search. Next, an algorithm for constructing a monotone quintic spline interpolant is presented.

%% ----------------------------------------------------------------------

\algorithm
    {monotone\_spline$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$}
    {where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real numbers, $f$ is an order six piecewise polynomial with breakpoints $k_1$, $\ldots$, $k_n$ defined by the data $\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$, $\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink factor.}
    {\label{alg:monotone-spline}
    \item create queue
      \\ create changed$(1$, $\ldots$, $n) := $ FALSE
      \\ create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$ \label{step:create}
      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity. The \texttt{queue} is a standard first in first out (FIFO) queue with \texttt{enqueue} and \texttt{dequeue} operations. The array \texttt{changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity. The \texttt{step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
    \item for $i := 1$, $\ldots$, $n-1$
      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
    \item while $($size$($queue$) > 0)$
    \item \indented{$(x_j$, $x_{j+1}) := $ dequeue}
    \item \indented{if $\big($not is\_monotone$(x_j$, $x_{j+1}$, $f)\big)$}
    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) -{}$step\_size$(j)$}}
      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) -{}$step\_size$(j+1)$}}
      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another. In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
    \item \indented{\indented{make\_monotone$(x_j,\ x_{j+1},\ f)$}}
    \item \indented{\indented{changed$(j) :=$ TRUE}}
      \indented{\indented{changed$(j+1) :=$ TRUE}} \label{step:changed}
    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not\in$ queue$\big)$ enqueue $\bigl((x_{j-1}, x_j)\bigr)$}}
      \indented{\indented{enqueue $\bigl((x_j, x_{j+1})\bigr)$}}
      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((x_{j+1}, x_{j+2})\bigr)$}} \label{step:requeue}
      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to \texttt{queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}      
      \indented{end if}
      end while
}
%% ----------------------------------------------------------------------

It is mentioned in \citeN{ulrich1994positivity} that for sufficiently small $Df(k_i)$ and $Df(k_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large. It can also be seen that decreasing the assigned derivative values towards zero will eventually cause Steps \ref{alg1-alpha} through \ref{alg1-gamma} of Algorithm \ref{alg:check-monotone} to all fail, resulting in Step \ref{alg1-simplified-true} returning \texttt{TRUE}.

Given the potential for successive monotonicity fixes across all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ breakpoints and shrink factor $s$. In practice this worst case has been observed to be unlikely.

\section{Experimental Results}
\label{sec:results}

Given the increased order of approximation, it is naturally expected that some accuracy benefit could be gained by using a quintic spline over a cubic spline. Experiments 1 and 2 below test some accuracy differences between monotone cubic and quintic splines. Experiment 3 analyzes the number of operations performed by Algorithm \ref{alg:monotone-spline} with increasingly large samples of monotone data.

\subsection{Approximating a Trigonometric Function}

For this experiment, the function $\sin(x) + x$ is considered over the interval $[0,(5/2)\pi]$ as seen in Figure \ref{fig:cubic_quintic_sin}. Given an increasing number of points, the maximum error of cubic and quintic approximations is shown in Figure \ref{fig:experiment_1}. The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points. There is an increase in the error gap between the two approximations as more points are added. The quintic spline requires only $700$ points to approximate the function to the same accuracy achieved by the cubic with $1000$ points.

\begin{figure}
  \centering
  \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
  \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.  Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points. The quintic interpolant captures both first and second derivative information at the points.}
  \label{fig:cubic_quintic_sin}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.7\textwidth]{experiment_1_errors}
  \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points. Both axes are log scaled, where $\mu$ means $\times 10^{-6}$ and n means $\times 10^{-9}$ on the vertical axis. The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points. In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at max error $10^{-8}$.}
  \label{fig:experiment_1}
\end{figure}

\subsection{Random Number Generation}

Consider the cumulative distribution function (CDF) defined by a mixture of three Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}. The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$. $200$ random samples are generated from this distribution with both cubic and quintic approximations from four points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.

While an increased number of points could be used to decrease the error in either approximation, the purpose of this experiment is to demonstrate the effect approximation error has on random number generation. The errors in the CDFs are magnified by the variance involved in randomly sampling from a distribution. As a result, while the cubic and quintic spline approximations have similar maximum error, the cubic distribution is significantly distorted around the first and second modes.

\begin{figure}
  \centering
  \includegraphics[width=.8\textwidth]{experiment_2_distribution}
  \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in Experiment 2. The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$. Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at random numbers uniformly distributed in the range $[0,1]$.}
  \label{fig:experiment_2_distribution}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.4\textwidth]{experiment_2_cubic_dist}
  \includegraphics[width=.4\textwidth]{experiment_2_quintic_dist}
  \caption{Both cubic and quintic splines are used to approximate the CDF of a Gaussian mixture distribution with three components. Four CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials. The resulting cloud of empirical distribution function (EDF) values is seen above. The maximum approximation error in the CDFs are similar, $.08$ for the cubic and $.05$ for the quintic. However, the maximum EDF differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
  \label{fig:experiment_2_results}
\end{figure}

%% - generate large sets of random monotone data, show Table of number
%%   of metrics growing with increasing number of points

%% - show visual of the \textit{error}

\subsection{Random Monotone Data}

This experiment studies the number of times the algorithms \texttt{is\_monotone} and \texttt{make\_monotone} are executed for increasingly large sequences of random monotone data. The minimum, median, and maximum number of times that Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} are executed is recorded, as well as the number of steps taken in all calls to \texttt{binary\_search}. The number of calls and steps grows linearly with $n$ as expected, requiring roughly one call to \texttt{make\_monotone} and $20$ binary search steps to create a monotone piece. Some (less common) problems require an average of three calls to \texttt{make\_monotone} per interval.

\begin{table}
  \centering
  \begin{tabular}{c|c c c|c c c|c c c}
    \hline
    \multirow{2}{*}{$n$}
      & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
      & min & median & max & min & median & max & min & median & max \\
    \hline
    5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
    25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
    50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
    75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
    100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
    \hline
  \end{tabular}
  \caption{Here we see a table of the minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data. Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
  \label{table:e3_results}
\end{table}

\section{Discussion}
\label{sec:discussion}

The monotone quintic spline interpolant provides a distinct increase in accuracy over the monotone cubic variant. Although the complexity of the algorithm is increased, the number of points required to achieve the same level of accuracy has decreased. In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.

Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more complex than the cubic case, gains in accuracy or a decreased necessary number of points are often worth the computational effort.

\section{Conclusion and Future Work}
\label{sec:conclusion}

This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants. Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory.

There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations. If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem. Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.

%% \section*{Acknowledgments}
%% The NSF is acknowledged for its support of this project.

% Please don't change the bibliographystyle style
\bibliographystyle{scsproc}
% AUTHOR: Include your bib file here
\bibliography{paper}

\section*{Author Biographies}

\textbf{\uppercase{THOMAS C. H. LUX}} is a Ph.D. candidate in computer
science at Virginia Polytechnic Institute and State University
(VPI\&SU) studying under Dr. Layne Watson. His research interests
include computational science, approximation theory, optimization, and
artificial intelligence. His email address is \email{tchlux@vt.edu}.

\textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has
interests in numerical analysis, mathematical programming,
bioinformatics, and data science.  He has been involved with the
organization of HPCS since 2000.

\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. candidate at VPI\&SU
studying computer science under Dr. Layne Watson.

\textbf{\uppercase{Li Xu}} is a Ph.D. candidate at VPI\&SU studying
statistics under Dr. Yili Hong.

\textbf{\uppercase{Yueyao Wang}} is a Ph.D. candidate at VPI\&SU
studying statistics under Dr. Yili Hong.

\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) has interests in
engineering statistics, statistical modeling, and data analysis.

%% \newpage

%% \appendix

%% \section{Author Checklist}
%% We strive for a consistent appearance in all papers published in the proceedings. If you used the template and styles within this author’s kit, then almost all of the requirements in this checklist will be automatically satisfied, and there is very little to check.

%% Please \textbf{print a hardcopy of your paper}, and go over your printed paper to make sure it adheres to the following requirements. \textit{Thank you!}

%% \begin{enumerate}
%% 	\item Abstract
%%   \begin{enumerate}
%% 	  \item 150 or fewer words.
%% 	  \item Provide 3-5 keywords (mandatory). This set of keywords will identify your paper in indices and databases.
%% 	\end{enumerate}
%% 	\item Paper Length
%%   \begin{enumerate}
%% 	  \item At least 5, but no more than 12 pages.
%% 	  \item Page size is letter size (8.5’’ x 11’’, or 216 mm x 279 mm).
%% 	\end{enumerate}
%% 	\item All text is in 11-Point Times New Roman except title, header and footer (default in this template).
%% 	\item Paper title is in 12-Point Times New Roman \textbf{BOLDFACE ALL CAPS} (default in this template).
%% 	\item The paper has been spellchecked using U.S. English. 
%% 	\item Spacing and Margins
%%   \begin{enumerate}
%% 	  \item Single spaced.
%% 	  \item Left and right margins are each 1 inch (default in this template).
%% 	  \item Top and bottom margins are according to the template.
%% 	  \item Title starts 1.25 inches from the top of the page (default in this template).
%% 	\end{enumerate}
%% 	\item Section Headings
%%   \begin{enumerate}
%% 	  \item Left justified and set in \textbf{BOLDFACE ALL CAPS} (default in this template).
%% 	  \item Numbered, except for the abstract, acknowledgments, references and author biographies.
%% 	  \item Subsection headings are set in \textbf{Boldface Headline Style} (default in this template).
%% 	\end{enumerate}
%% 	\item No footnotes or page numbers.
%% 	\item The copyright notice on the first page contains the name of the symposium or a track where you want to submit your paper, and the running head on subsequent pages is the surnames of all authors.
%% 	\item Multiple authors are formatted correctly.
%% 	\item Equations are centered and any equation numbers are in parentheses and right-justified (default).
%% 	\item Figures and Tables
%%   \begin{enumerate}
%% 	  \item All text in figures and tables is readable.
%% 	  \item Table captions appear above the table.
%% 	  \item Figure captions appear below the figure.
%% 	\end{enumerate}
%% 	\item Citations and References
%%   \begin{enumerate}
%% 	  \item Citations are by author and year, and are enclosed in parentheses, not brackets (default \BibTeX). 
%% 	  \item References are in the hangref style, and are listed alphabetically by the last names(s) of the author(s) (also default \BibTeX\ setting in this template). 
%% 	\end{enumerate}
%% 	\item Author biographies are one paragraph per author.
%% 	\item Hyperlinks
%%   \begin{enumerate}
%% 	  \item Be sure that hyperlinks will probably work in the future as well.
%% 	  \item Live hyperlinks are blue. Nonlive hyperlinks are black (default in this template).
%% 	\end{enumerate}
%% 	\item All fonts must be embedded in the resulting \texttt{.pdf}. If you use \texttt{pdflatex}, this should already be true--unless you included figures in the \texttt{.eps} or \texttt{.pdf} format which may introduce additional font dependencies. You can use the \texttt{pdffonts} command to see if all the fonts are embedded (the column ``emb'' should say yes for all rows). You can use GhostScript to force embed the fonts, like so: {\texttt{gs -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -dEmbedAllFonts=true -sOutputFile=out.pdf -f in.pdf}} (in Windows, use \texttt{gswin32} or \texttt{gswin32c} to invoke GhostScript).
%% \end{enumerate}

%% After verifying that your paper meets these requirements, please go to the final submission page at conference website and submit your paper. Be sure to complete the transfer of copyright form and upload the \texttt{.pdf} receipt. \textit{Thank you for contributing to the SCS conferences!}

%% \section{Most Observed Mistakes}

%% The following list comprises \textbf{the most common sources of error} that had to be corrected by previous editors. Please make sure to go through the following list and check that your paper is formatted correctly:
%% \begin{enumerate}
%% 	\item	The paper is less than 5 or more than 12 pages long.
%% 	\item	Paper title and section titles are in \textbf{BOLD ALL CAPS}, subsections are \textbf{Bold and Capitalize the First Letter of Important Words}. Please use the templates.
%% 	\item	Paper is in A4 format, not letter format. Please use the required margins.
%% 	\item	The copyright notice is incorrect.
%% 	\item	The running heads are incorrect. Do not forget that the LastNameLastAuthor is preceded by ``, and ''.
%% 	\item	The citation format is incorrect. Use \BibTeX\ for citations and do not change the format used by the template.
%% 	\item	The biographies are missing. Do not forget the ``author biographies'' section.
%% 	\item	Figures or tables are not referenced in the text or have the incorrect caption format.
%% 	\item	The author section after the title is not formatted correctly, the number of organizations does not define the number of blocks, or the number of blocks does not define the layout.
%% 	\item In the heading on the title page, country names are in all capitals.
%% 	\item	Paragraphs are not indented.
%% 	\item Some fonts in the resulting \texttt{.pdf} are not embedded.
%% \end{enumerate}

\end{document}

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/experiments/code/__init__.py

@@ -1,0 +1,1 @@
+'''Testing'''

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/experiments/code/monotone.py

@@ -1,0 +1,388 @@
+# This file provides the function `monotone_quintic_spline` which
+# constructs spline interpolant over provided data. It starts by using
+# divided differences to construct approximations for the first and
+# second derivative at each of the data points. Then it iteratively
+# adjusts the derivatives and second derivatives until all quintic
+# polynomial pieces are monotone. A `Spline` object is returned.
+
+from polynomial import fit
+IS_MONOTONE_CALLS = 0
+
+# Given x and y values, construct a monotonic quintic spline fit.
+#  shrink_rate = .9    -> 120 steps max
+#              = .99   -> 386 steps max
+#              = .999  -> 1221 steps max
+#              = .9999 -> 3861 steps max
+def monotone_quintic_spline(x, y=None, values=None, ends=2, mids=2, exact=True, 
+                            max_steps=100, verbose=False, **kwargs):
+    if verbose: print()
+    from polynomial import Spline
+    # Convert 'x' to exact values.
+    if exact:
+        from fraction import Fraction
+        x = list(map(Fraction, x))
+    # Build a set of 'values' if they were not provided.
+    if (y is not None):
+        if exact: y = list(map(Fraction, y))
+        # Check for which type of monotonicity this function maintains.
+        if   all(y[i+1] >= y[i] for i in range(len(y)-1)): kwargs.update(dict(min_d1=0))
+        elif all(y[i] >= y[i+1] for i in range(len(y)-1)): kwargs.update(dict(max_d1=0))
+        else: raise(Exception("Provided 'y' data is not monotone."))
+        # Pass in the method of fitting end and midpoints as keyword arguments.
+        kwargs.update(dict(mids=mids, ends=ends))
+        # Construct an initial fit that is twice continuous.
+        f = fit(x, y, continuity=2, **kwargs)
+        values = f.values
+    elif (values is not None):
+        if exact: values = [list(map(Fraction, row)) for row in values]
+        # Verify monotonicity.
+        if   all(values[i+1][0] >= values[i][0] for i in range(len(values)-1)): pass
+        elif all(values[i][0] >= values[i+1][0] for i in range(len(values)-1)): pass
+        else: raise(Exception("Provided 'y' data is not monotone."))
+        # Make a spline over the points and derivative values.
[…]

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/experiments/code/plot.py

@@ -1,0 +1,1961 @@
+# This module serves to provide a simplified interface to *offline*
+# python plotly plotting. The user can produce plots without ever
+# interacting directly with the dictionary objects that plotly
+# expects. This module currently supports 2D and 3D scatter plots with
+# numerical axes, histograms, subplots (with varying numbers of plots
+# in each row), animations, box-plots, and plot annotations.
+# 
+# Required packages:
+#   random, numbers, os, webbrowser, sys, re, tempfile
+#   numpy
+#   scipy
+# 
+# Imports nested in appropriate functions:
+#   import plotly
+#   from scipy.spatial import ConvexHull
+#   from scipy.spatial import Delaunay
+# 
+# INSTALLATION:
+# 
+#  Installation requires a SPECIFIC VERSION OF PLOTLY. Here is the
+#  standard set of packages required for usage:
+# 
+#    pip install scipy
+#    pip install numpy
+#    pip install plotly==2.0.15
+# 
+#  This package will not work with newer versions of plotly because
+#  they changed the underlying storage data types for figures.
+#  Any plotly update for this body of code is unlikely.
+# 
+# 
+# USAGE:
+# 
+#  The available (user accessible) functions are:
+# 
+#    plot.Plot        -- The primary class for holding / creating plots.
+#    plot.multiplot   -- A mechanism for plotting multiple Plot
+#                        objects in the same window.
+#    plot.create_html -- A function for generating a local HTML file
+#                        from a figure object (in Plotly terms).
+#    plot.iplot       -- A convenience wrapper for generating
+#                        interactive plots in a Jupyter notebook with
+#                        multiplot functionality as well.
+# 
+
+# --------------------------------------------------------------------
+#                       DEVELOPER COMMENTS
+# 
+# TODO: Plot.add_zero_line(func)
+# TODO: Plot.add_frame(..., persist=True)
+# TODO: Adding multiple frames where the first has no edges and the
+#       rest have edges causes all frames to look like first.
+# 
+# --------------------------------------------------------------------
+
+import random, numbers, os, webbrowser, sys, re, tempfile
+import numpy as np
+
+NOTEBOOK_MODE = False    # Jupyter notebook mode
+PLOT_MARGIN = 50         # In pixels
+PLOT_POINTS = 1000       # Number of samples
+BRIGHTNESS_RANGE = 0.6   # For default shading of points
+RANDOM_SEED = 0          # Seed used for new color generation
+MIN_PALATTE_COLORS = 40  # Number of palatte entries to create
[…]

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/experiments/example_1.py

@@ -1,0 +1,53 @@
+# Include the "code" directory into the path and import code custom to this project.
+import sys, os
+sys.path += [os.path.abspath("code")]
+from fraction import Fraction as F
+from polynomial import polynomial_piece, Polynomial, NewtonPolynomial
+from plot import Plot, multiplot
+sys.path.pop(-1)
+
+
+x = [F(0),F(1)]
+# A linear function.
+y1 = [
+    [F(1)],
+    [F(0)]
+]
+# A cubic function.
+y2 = [
+    [F(1,2),F(4)],
+    [F(1,2),F(4)]
+]
+# A quintic function.
+y3 = [
+    [F(1), F(-7), F( 100,1)],
+    [F(0), F(-7), F(-100,1)]
+]
+
+f1 = Polynomial(polynomial_piece( *y1, x ))
+f2 = Polynomial(polynomial_piece( *y2, x ))
+f3 = Polynomial(polynomial_piece( *y3, x ))
+print("f1: ",f1)
+print("f2: ",f2)
+print("f3: ",f3)
+
+p1 = Plot("","","", font_family="times", font_size=20)
+p1.add_func("f1", f1, x, color=p1.color(1))
+p1.add_annotation("(1)&nbsp;&nbsp;to&nbsp;&nbsp;(0)", .5, .5, ax=.75,
+                  ay=.8, font_family="times", font_size=16)
+
+p2 = Plot("","","", font_family="times", font_size=20)
+p2.add_func("f2", f2, x, color=p1.color(2))
+p2.add_annotation("(½, 4)&nbsp;&nbsp;to&nbsp;&nbsp;(½, 4)", .5, .5, ax=40,
+                  ay=.8, font_family="times", font_size=16)
+
+
+p3 = Plot("","","", font_family="times", font_size=18)
+p3.add_func("f3", f3, x, color=p1.color(3))
+p3.add_annotation("(1,-7,100)&nbsp;&nbsp;to&nbsp;&nbsp;(0,-7,-100)",
+                  .5, .5, ax=15, ay=.8, font_family="times",
+                  font_size=16)
+
+
+multiplot([p1,p2,p3], gap=.06, shared_y=True, show_legend=False,
+          width=680, height=100, file_name="spline_demonstration.html")

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/experiments/experiment_3.py

@@ -1,0 +1,134 @@
+import numpy as np
+
+# Include the "code" directory into the path and import code custom to this project.
+import sys, os
+sys.path += [os.path.abspath("code")]
+from monotone import monotone_quintic_spline, monotone_cubic_spline
+import monotone
+from fraction import Fraction
+from data import Data
+from plot import Plot
+sys.path.pop(-1)
+
+
+SEED = 0
+POINTS = 10
+
+results_file = "e3_results.pkl"
+results = Data(names=["num points", "seed", "total checks", "total fixes", "search steps", "checks dict", "fixes dict"],
+               types=[int,          int,    int,            int,           int,            dict,          dict])
+
+# Try to load existing results if they exist.
+try: results.load(results_file)
+except: pass
+
+# # Run the experiment and collect the data.
+# for points in range(2, 101):
+#     print("points: ",points)
+#     for seed in range(100):
+#         print("  seed: ",seed, end="\r")
+#         # Skip already checked values.
+#         if ((points,seed) in results["num points","seed"]): continue
+#         # Generate random data to test the monotonic fit function.
+#         np.random.seed(seed)
+#         x = np.linspace(0, 1, points)
+#         y = sorted(np.random.normal(size=(points,)))
+#         x -= min(x); x /= max(x)
+#         y -= min(y); y /= max(y); y -= max(y)
+#         # Convert these arrays to exact arithmetic.
+#         x = list(map(Fraction, x))
+#         y = list(map(Fraction, y))
+#         interval = [float(min(x)), float(max(x))]
+#         # Compute the monotone quintic spline.
+#         stdout = sys.stdout
+#         # Block all print statements from the function.
+#         with open(os.devnull, "w") as devnull:
+#             sys.stdout = devnull
+#             monotone.IS_MONOTONE_CALLS = 0
+#             f, checks, changes = monotone_quintic_spline(x,y=y, verbose=True)
+#         sys.stdout = stdout
+#         # Store the results.
+#         results.append(
+#             [points, seed, sum(checks.values()), sum(changes.values()),
+#              monotone.IS_MONOTONE_CALLS, checks, changes]
+#         )
+#     # Save the intermediate results.
+#     results.sort()
+#     results.save(results_file)
+
+print(" "*70)
+
+print(results[results["num points"] > 10])
+
[…]

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/paper/template/scs20paper.tex
Created paper repository with all of my work.

%**************************************************************************
%* SpringSim 2020 Author Kit
%*
%* Word Processing System: TeXnicCenter and MiKTeX
%*
%**************************************************************************

\documentclass{scspaperproc}

\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{mathptmx}

%
%****************************************************************************
% AUTHOR: You may want to use some of these packages. (Optional)
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{amsthm}
%****************************************************************************

%
%****************************************************************************
% AUTHOR: If you do not wish to use hyperlinks, then just comment
% out the hyperref usepackage commands below.

%% This version of the command is used if you use pdflatex. In this case you
%% cannot use ps or eps files for graphics, but pdf, jpeg, png etc are fine.

\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black,bookmarks=false]{hyperref}

%% The next versions of the hyperref command are used if you adopt the
%% outdated latex-dvips-ps2pdf route in generating your pdf file. In
%% this case you can use ps or eps files for graphics, but not pdf, jpeg, png etc.
%% However, the final pdf file should embed all fonts required which means that you have to use file
%% formats which can embed fonts. Please note that the final PDF file will not be generated on your computer!
%% If you are using WinEdt or PCTeX, then use the following. If you are using
%% Y&Y TeX then replace "dvips" with "dvipsone"

%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
%% anchorcolor=black,linkcolor=black]{hyperref}

%% The use of the long citation format (e.g. "Brown and Edwards (1993)" rather than "[5]") and at the same
%% time using the hyperref package can lead to hard to trace bugs in case the citation is broken accross the
%% line (usually this will mark the entire paragraph as a hyperlink (clickable) which is easily noticeable and fixed
%% if using colorlinks, but not if the color is black -- as it is now). Worse yet, if a citation spans page boundary,
%% LaTeX compilation can fail, with an obscure error message. Since this depends a lot on the flow of the text
%% and wording, these bugs come and go and can be extremely hard for a beginner to trace. The error
%% message can look like this:
%%
%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink.
%%    \AtBegShi@Output ...ipout \box \AtBeginShipoutBox 
%%    \fi \fi 
%%    l.174 
%%    ! ==> Fatal error occurred, no output PDF file produced!
%%
%% and can be universally fixed by putting an \mbox{} around the citation in question (in this case, at line 174)
%% and maybe adapting the wording a little bit to improve the paragraph typesetting, which is perhaps not
%% immediately obvious.
%****************************************************************************

%
%****************************************************************************
%*
%* AUTHOR: YOUR CALL!  Document-specific macros can come here.
%*
%****************************************************************************

% add custom hyphenation rules here
\usepackage{hyphenat}
\hyphenation{op-tical net-works semi-conduc-tor}

% If you use theorems
\newtheoremstyle{scsthe}% hnamei
{8pt}% hSpace abovei
{8pt}% hSpace belowi
{\it}% hBody fonti
{}% hIndent amounti1
{\bf}% hTheorem head fontbf
{.}% hPunctuation after theorem headi
{.5em}% hSpace after theorem headi2
{}% hTheorem head spec (can be left empty, meaning `normal')i

\theoremstyle{scsthe}
\newtheorem{theorem}{Theorem}
\renewcommand{\thetheorem}{\arabic{theorem}}
\newtheorem{corollary}[theorem]{Corollary}
\renewcommand{\thecorollary}{\arabic{corollary}}
\newtheorem{definition}{Definition}
\renewcommand{\thedefinition}{\arabic{definition}}

% avoid overrunning the right margin; you are welcome to remove this, provided that you take care not to overrun the right margin anywhere in your paper
\sloppy

%#########################################################
%*
%*  The Document.
%*
\begin{document}

%***************************************************************************
% AUTHOR: AUTHOR NAMES GO HERE
% FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
%
%		You need to change the author listing below!
%               Please list ALL authors using last name only, separate by a comma except
%               for the last author, separate with "and"
%

\SCSpagesetup{LastName1, LastName2, LastName3 ... and LastNameLastAuthor}

% AUTHOR: Uncomment ONE of these correct conference names.
\def\SCSconferenceacro{SpringSim'20}
%\def\SCSconferenceacro{SummerSim}
%\def\SCSconferenceacro{AutumnSim}
%\def\SCSconferenceacro{PowerPlantSim}

% AUTHOR: Set the correct year of the conference.
\def\SCSpublicationyear{2020}

% AUTHOR: Set the correct month and dates; the dates are separated by a single minus sign
% with no spaces and no leading zeros, the month is a full name (e.g. April) with the first letter
% capitalized. For example, "April 8-13".
\def\SCSconferencedates{May 19-May 21}

% AUTHOR: Set the correct venue in the form "City, State, Country", for example "Los Angeles, CA, USA".
\def\SCSconferencevenue{Fairfax, VA, USA}

% AUTHOR: Enter the title, all letters in upper case
\title{Instructions for Authors of Papers Using \LaTeX}

% AUTHOR: Enter the authors of the article, see end of the example document for further examples
\author{
\\%To level with the author block on the right.
Theresa M. K. Roeder \\ [12pt]
Department of Decision Sciences \\
San Francisco State University \\
1600 Holloway Avenue \\
San Francisco, CA, USA \\
scs19roeder@gmail.com\\
% Multiple authors are entered as follows.
% If they share the same affiliation, they must be allocated within the same block
% Every author name must be separated by \\ but the last one within the block, that will be separated by \\ [12pt]
% You may also need to adjust the titlevbox size in the preamble - search for titlevboxsize
\and
Peter I. Frazier \\
Henry Wotton \\ [12pt]
School of Operations Research \\ and Information Engineering \\
Cornell University \\
206 Rhodes Hall, Ithaca, NY, USA \\
\{scs19frazier,scs19wotton\}@gmail.com\\
\and
Roberto Szechtman \\ 
Jonathan Harker \\ [12pt]
Department of Operations Research \\
Naval Postgraduate School \\
1 University Circle \\
Monterey, CA, USA \\
\{scs19szechtman,scs19harker\}@gmail.com\\
\and
\\%To level with the author block on the right.
Enlu Zhou \\ [12pt]
H. Milton School of Industrial \& Systems Engineering \\
Georgia Institute of Technology \\
755 Ferst Drive NW \\
Atlanta, GA, USA \\
scs19zhou@gmail.com
}

\maketitle

\section*{Abstract}

This set of instructions for producing a proceedings paper for The Society for Modeling \& Simulation International (SCS) with \LaTeX\ also serves as a sample file that you can edit to produce your submission, and a checklist to ensure that your submission meets the SCS requirements. Please follow the guidelines herein when preparing your paper. Failure to do so may result in a paper being rejected, returned for appropriate revision, or edited without your knowledge.

\textbf{Keywords:} instructions, author’s kit, conference publication. (3-5 keywords separated by a comma.)
%% AUTHOR:
% This is a list of no more than five keywords that will identify your paper in indices and databases (required).
% Do not use the words “computer”, “simulation”, “model”, or “modeling”, since these are all assumed.

\section{Introduction}
\label{sec:intro}

This paper provides instructions for the preparation of papers for the Society for Modeling \& Simulation International (SCS) using \LaTeX. There is a companion paper that provides instructions for the preparation of papers with Microsoft Word. The easiest way to write a paper using \LaTeX\ that complies with the requirements is to edit the source file, \texttt{scs19paper.tex} for this document. Please do not use an older version, as some specification may have changed. An author kit is available via conference website. The author kit includes this \LaTeX\ document and its Microsoft Word companion. This document was typeset using \texttt{pdflatex}.

\section{General Guidelines}

\subsection{Language}

The paper should be prepared using U.S. English in the interest of consistency across the proceedings. Please carefully check the spelling of words before you submit your paper. There are spell checkers for \LaTeX\ as well.
Some examples of software which supports spell checking are TexnicCenter, TexMaker, and TexClipse.

\subsection{Objectivity}
The content of the paper should be objective and without any appearance of commercialism.  In general, comparisons of commercial software should be avoided unless they are central to the topic.  If a comparison of commercial software is included, it should be based on objective analysis that includes criteria, description of ranking methodology on each criteria, and the rankings themselves to arrive at the conclusion.
If an approach other than a detailed objective analysis is used to select the simulation software used for the study being reported, such as, availability of the software, or the familiarity of the analyst with the software, it should be clearly identified. To ensure suitability for an international audience, please pay attention to the following:

\begin{itemize}
\item{Write in a straightforward style.}
\item{Try to avoid long or complex sentence structures.}
\item{Briefly define or explain all technical terms that may be unfamiliar to readers.}
\item{Explain all acronyms the first time they are used in your text – e.g., “Digital Signal Processing (DSP)”.}
\item{Explain local references (e.g., not everyone knows all city names in a particular country).}
\item{Explain “insider” comments. Ensure that your whole audience understands any reference whose meaning you do not describe (e.g., do not assume that everyone has used a Macintosh or a particular application).}
\item{Explain colloquial language and puns. Understanding phrases like “red herring” may require a local knowledge of English. Humor and irony are difficult to translate.}
\item{Use unambiguous forms for culturally localized concepts, such as times, dates, currencies and numbers (e.g., “1-5- 97” or “5/1/97” may mean 5 January or 1 May and “12/10/11” can be even more confusing, and “seven o’clock” may mean 7:00 am or 7:00 pm).  For currencies, indicate English equivalences – e.g., “Participants were paid 10,000 lire, or roughly \$5.”}
\item{Be careful with the use of gender-specific pronouns (he, she) and other gendered words (chairman, manpower, man-months). Use inclusive language that is gender-neutral (e.g., she or he, they, s/he, chair, staff, staff-hours, person-years).}
\item{If possible, use the full (extended) alphabetic character set for names of persons, institutions, and places (e.g., {Gr\o nb\ae k}, Lafreni\'ere, S\'anchez, Universit\"at, {Wei\ss enbach}, Z\"ullighoven, {\AA rhus}, etc.).}
\end{itemize}

\subsection{Paper Submission}

Please, refer to the author's guidelines document in the author's kit.

\subsection{Length Constraints}

\subsubsection{The Abstract and Keywords}
The abstract should be at most 150 words. Since abstracts of all papers accepted for publication in the proceedings will also appear in the final program, the length limit of 150 words will be strictly enforced for each abstract. The abstract should consist of a single paragraph, and it should not contain references or mathematical symbols.

The list of keywords should have no more than five keywords that will identify your paper in indices and databases (required). Do not use the words “computer”, “simulation”, “model”, or “modeling”, since these are all assumed.

\subsubsection{Length of the Paper}
The page size in the proceedings must be 8.5 inches by 11 inches (21.6 cm by 27.9 cm). The overall length of the paper should be at least 5 proceedings pages. \textbf{Papers should be at most 12 pages}.

\subsubsection{Font Specification and Spacing}
The paper should be set in the Times New Roman font using a 11-point font size and it should be single spaced. Do not use other fonts; the use of other fonts means the proceedings editors will need to send the paper back to you to change the font.

These settings are automatically applied by the class file used for this document, thus you are asked not to change these settings in your paper. If you want to use bold Greek symbols you should use the \texttt{bm} package.

\subsubsection{Margins}
\label{sec:margins}

The width of the text area is 6.5 inches (16.0 cm). The left and right margins should be 1 inch (2.54 cm) on each page. The top margin shall be 0.89 inches (2,24 cm) including the header, the bottom of the pages shall be 0.67 inches (1.7 cm). These settings are automatically applied by the class file used for this document, thus you are asked not to change these settings in your paper.

\subsubsection{Justification}
Headings of sections, subsections, and subsubsections should be left-justified. One-line captions for figures or tables should be centered. A multiline caption for a figure or table should be fully justified. All other text should be fully justified across the page (that is, the text should line up on the right-hand and left-hand sides of the page). These settings are automatically applied by the class file used for this document.

\subsection{Headings of Sections, Subsections, and Subsubsections}
Section, subsection, and subsubsection headings should appear flush left, set in the bold font style, and numbered as shown in this document. The headings for the Abstract, Acknowledgments, References and Author Biographies sections are not numbered. These settings are automatically applied by the class file used for this document.

To suppress the section numbers, use the \verb+\section*{}+ command.

Section headings should be set in \textbf{\uppercase{full capitals like this phrase}}, while subsection and subsubsection headings should be \textbf{Capitalized
in Headline Style like This Phrase}. These settings are automatically applied by the class file used for this document.

\subsubsection{Paragraphs}
The paragraphs should not be indented. Normal style will add a space between paragraphs, do not insert additional space between paragraphs.

\subsubsection{Footnotes}
\textbf{Do not use footnotes}; instead incorporate such material into the text directly or parenthetically.

\subsubsection{Page Numbers}
\textbf{Do not include page numbers.} Page numbers are generated by the proceedings editors once all accepted papers are ordered for the final proceedings.

\section{Formatting the First Page}

\subsection{Running Heads}
The running head (provided in the template) in the lower left-hand corner of the first page (\textit{which should read the conference name, location and the name of the Society}) is left-justified and set in the 9-point italic font style. %You do not have to change the content of the first page copyright notice; the first page header and footer are automatically provided by the class file.
You will need to \textbf{uncomment the appropriate version of the track or symposium name} at the beginning of \texttt{scs19paper.tex}, depending on where you are going to submit your paper.

Running heads on the second and subsequent pages should contain the last names of the authors, centered and set in the 10-point italic font style. For example, running heads for papers with varying numbers of authors would appear like \emph{Yilmaz} (single author), or \emph{Yilmaz and Chan} (two authors), or \emph{Yilmaz, Chan, and Moon} (three authors), or \emph{Yilmaz, Chan, Moon, and Roeder} (four authors). These are created by using the following macro:

\noindent \begin{verbatim}
\SCSpagesetup{LastName1, LastName2, and LastNameLastAuthor}
\end{verbatim}

located just after the \verb+\begin{document}+ of this \texttt{.tex} file.

Please use this macro to set up the running heads, as it sets further parameters important for the correct layout of the document. List all authors; do not use \textit{et al.} The author names are listed in the same order as they appear on the title page. This will be the same order when providing the author biographies at the end of the paper.

\subsection{Title and Authors}

Center the title of the paper on the page and set it in bold 12pt \textbf{\uppercase{full capitals}}. The top edge of the title should begin at least 1.25 inches from the top of the page. The correct placement is automatically done by the class file as well. Just use the \verb+\title+ and \verb+\maketitle+ commands as it is done in the source of this document. Multiline titles should have about the same amount of text on each line. There should be 1 blank line between the title and the authors’ names (will be inserted by the class file if the \verb+\author+, \verb+\title+, and \verb+\maketitle+ commands are used).

Each author’s name should be centered on a new line, with the author’s first name first and no job title or honorific. Insert 1 blank line between the author’s name and address. The organization or institution that the author is affiliated to should be typed first. Next type the complete street address, without abbreviations, followed by the city, standard two-letter state or province abbreviation, and country. The address should be centered. For papers with multiple authors, the authors should be listed in order of decreasing contribution, with authors from the same institution listed separately if possible. Different formats for multiple authors are shown as examples in Figures~\ref{fig2same} through \ref{fig2equals1different} at the end of this document. There should be 1 blank line between the author names and the text of the paper. You should use the \verb+\author+ command to enter author names, separated using the command \verb+\and+. See the \texttt{.tex} source file for this document. Authors with multiple affiliations shall pick the main affiliation for the title page and mention other affiliations in their biography.

\section{Formatting Subsequent Pages}
Please refer to \autoref{sec:margins} for the correct margins.

\subsection{Mathematical Expressions in Text and in Displays}
Display only the most important equations, and number only the displayed equations that are explicitly referenced in the text. To conserve space, simple mathematical expressions such as \mbox{$\bar Y = n^{-1} \sum_{i=1}^n Y_i$} may be incorporated into the text. Mathematical expressions that are more complicated or that must be referenced later should be displayed, as in

\begin{equation}
s^2 = \frac 1 {n-1} \sum_{i=1}^n (Y_i - \bar Y)^2\;. \nonumber
\end{equation}

If a display is referenced in the text, then enclose the equation number in parentheses and place it flush with the right-hand margin of the column. For example, the quadratic equation has the general form

\begin{equation} \label{eq:quadratic}
ax^2 + bx + c = 0, \mbox{ where } a \ne 0\;.
\end{equation}

In the text, each reference to an equation number should also be enclosed in parentheses. For example, the solution to \eqref{eq:quadratic} is given in \eqref{eq:quadratic sol} in \autoref{app:quadratic}.

If the equation is at the end of a sentence, then you should end the equation with a period. If the sentence in question continues beyond the equation, then you should end the equation with the appropriate punctuation~-~that is, a comma, semicolon, or no punctuation mark.

\subsection{Displayed Lists}
A displayed list is a list that is set off from the text, as opposed to a run-in list that is incorporated into the text. The bulleted list given below provides more information about the format of a displayed list. 
\begin{itemize}
	\item Use standard bullets instead of checks, arrows, etc.\ for bulleted lists.
	\item For numbered lists, the labels should not be Arabic numerals enclosed in parentheses because such labels cannot be distinguished from equation numbers.
\end{itemize}

The paragraph after the list is not indented.
\begin{enumerate}
	\item Use standard numbering instead of special characters for numbered lists.
	\item For numbered lists, the labels should not be Arabic numbers enclosed in parentheses because such labels cannot be distinguished from equation numbers.
	\item You may need to restart the numbering on numbered lists.
\end{enumerate}

\subsection{Definitions and Theorems}
Definitions, theorems, propositions, etc. should be formatted like a normal paragraph with a boldface heading as shown in the examples below. Number these items separately and sequentially. You may choose to separately number theorems, propositions, corollaries, etc., as opposed to the example below where corollaries and theorems are numbered together. Search the source of this document to see how these environments were defined. The key command is \verb+\newtheorem+. Use a period after the definition, theorem, corollary or proposition number.

\begin{definition}
In colloquial New Zealand English, the term \emph{dopey mongrel} is used to refer to someone who has exhibited less than stellar intelligence.
\end{definition}

\begin{theorem}
If a proceedings editor from New Zealand accidentally deletes his draft of the author kit shortly after completing it, he would be considered to be a dopey mongrel.
\end{theorem}

\begin{corollary}
One of the proceedings editors is a dopey mongrel.
\end{corollary}

The paragraph after the list is not indented.

\subsection{Figures and Tables}
\label{sec:graphics}
Figures and tables should be centered within the text and should not extend beyond the right and left margins of the paper. Figures and tables can make use of color since the SCS produces electronic proceedings. However, try to select colors that can be differentiated when printing in black and white in consideration of vast majority of people using such printers. Figures and tables are numbered sequentially, but separately, using Arabic numerals. All tables and figures should be explicitly referenced in the text and they should not be placed before they are referenced.

Each table should appear in the document after the paragraph in which the table is first referenced. However, if the table is getting split across pages, it is okay to include it after a few paragraphs from its first reference. One-line table captions are centered, while multiline captions are left justified. The table captions appear \emph{above} the table.

Captions can be written using normal sentences with full punctuation. All captions should end with a period. It is fine to have multiple-sentence captions that help to explain the table. See Tables~\ref{tab:first} and \ref{tab:second} for examples, note that the captions are above the table.

\begin{table}[htb]
\caption{Table captions appear above the table, and if they are longer than one line they are left justified. Captions are written using normal sentences with full punctuation. It is fine to have multiple-sentence captions that help to explain the table.}\label{tab:first}
\centering
\begin{tabular}{rll}
\hline
Creature & IQ & Diet\\ \hline
dog & 70 & anything\\
cat & 75 & almost nothing\\
human & 60 & ice cream \\
dolphin & 120 & fish fillet\\
\hline
\end{tabular}
\end{table}

\begin{table}[htb]
\centering
\caption{Counting in Maori.}\label{tab:second}
\begin{tabular}{r|l}
English & Maori \\ \hline
one & tahi \\
two & rua \\
three & toru \\
four & wha \\
\end{tabular}
\end{table}

Each figure should appear in the document after the paragraph in which the figure is first referenced. Figure captions appear below the figure. Single-line captions are centered, while multiline captions are left justified. Captions end with a period. See Figures~\ref{fig:tahi} and \ref{fig:rua} for examples.

\begin{figure}[htb]
{
\centering
%\includegraphics[width=0.9\columnwidth]{MathExpandExpression.jpg}
\includegraphics[width=0.50\textwidth]{MathExpandExpression}
\caption{An unusual answer to a question.}\label{fig:tahi}
}
\end{figure}

\begin{figure}[htb]
{
\centering
%\includegraphics[width=0.9\columnwidth]{puzzle.png}
\includegraphics[width=0.50\textwidth]{puzzle}
\caption{The area of the square is 64 squares, while that of the rectangle is 65 squares, yet they are made of the same pieces! How
is this possible?}\label{fig:rua}
}
\end{figure}

References to tables and figures are given as \autoref{fig:tahi} or \autoref{tab:first}. For example, ``We see in \autoref{tab:second} that \ldots'' and ``We see in \autoref{fig:rua} \ldots'' are both correct. Be sure to use the \verb+\label+ command within the figure or table environment and refer to the associated figure or table using \verb+\autoref{your label here}+.

Please do not use hard coded figure/table numbers. This is error prone, and the references will not be hyperlinks.

Please ensure that your graphics files use standard fonts (Times New Roman, Symbol, etc.) or that those are embedded in the final figure files.
If they are not embedded, and if the font is not available on the editor's computer, then the font will not be included in the final PDF.
This may lead to a problem with displaying the final PDF file on computers without an appropriate font.
At best you select a format which allows to embed the fonts in all non bitmap figure files.

Including graphics files in your document can be complicated. In general you have 2 options. Either (a) you use \texttt{.jpg}, \texttt{.png} or \texttt{.pdf} files (\texttt{.eps} files can be also used, provided that the \texttt{epstopdf} package is added in the preamble) or (b) all of your files must be either \texttt{.ps} or \texttt{.eps}.
Note that there are tools to convert these formats into one another. The main difference between the formats is how they store the images and how well suited they are for specific graphics. You can choose between bitmap and vector graphics.

Bitmap graphics are well suited for photographs (\texttt{.jpg} is very common here) or for screenshots (\texttt{.png} is a lossless encoding in contrast to \texttt{.jpg}, and is thus better suited for all those cases where you have sharp edges in your graphics).

Vector graphics are the encoding to be chosen for all kinds of drawings (diagrams, charts, \ldots). In contrast to bitmap formats, they can be scaled to any size without any loss of sharpness. This makes it possible to read such graphics even if two pages are printed on one sheet of paper, or if the documents are read electronically.

So what to choose for your \LaTeX\ document? As a rule of thumb you should always prefer \texttt{.pdf} or \texttt{.eps}.
In general these two formats can contain both, bitmap and vector graphics. But there is no need (and no use) to convert your bitmaps to either of these.

If you follow Option (a), then you must use the \texttt{PDFLaTeX} command to generate your PDF file, as was done with this file. PDFLaTeX is nowadays the standard - so option (a) should be the natural choice. The final file format is PDF.

If you follow Option (b), then you must use the outdated \texttt{latex - dvips - ps2pdf} route for generating a PDF file. You may run into a problem if using both the \texttt{hyperref} package and the \texttt{graphicx} package; there seems to be a clash there.
In that case, you might either not use the \texttt{hyperref} package and continue to use \texttt{graphicx}, or continue to use the \texttt{hyperref} package and use the \texttt{epsfig} package in place of the \texttt{graphicx} package.

If you persevere with \texttt{hyperref} then be sure to use the appropriate version of the \verb+\usepackage+ command; see the preamble in the source of this file for details. See also \autoref{sec:hyper} below. It is important to note that if you remove the \texttt{hyperref} package, then you have to deal with the correct formatting of hyperlinks on your own.

Whatever option you choose: if you include figures via \texttt{includegraphics}, then please do so without the file ending (e.g., skip \texttt{.pdf}, \texttt{.eps}, \ldots).

\subsection{Hyperlinks}
\label{sec:hyper}

A \emph{hyperlink} specifies a Web address (URL) or an e-mail address. The use of hyperlinks allows authors to give readers access to external electronic information, such as a dynamic simulation or animation. 
But please note: hyperlinks (to web pages) might not work forever (web pages might be removed), and thus using hyperlinks intensively may make a paper (or parts thereof) less useful in future. If enough information is provided in the main body of the paper to enable searching for the cited content in any case and if the inclusion of the web address does not hurt the appearance of the paper, then the web address can be included in the main body of the paper itself.

\textbf{While the use of hyperlinked text is encouraged in the main body of the paper, it is recommended that corresponding web addresses and other identifying information should be provided in list of references.} For example, instead of spelling out the web address of the conference website, one would refer to conference website and the corresponding entry in the reference section will spell out the associated web address and other relevant information such as author(s) and/or organization that published the content. This would allow readers to search for the content using the author(s), organization, etc. in case the actual web-address is changed. This also allows for a cleaner appearance of the main body of the paper.

Each hyperlink should be set in the same font as the text. Hyperlinks are \emph{not} underlined. A live hyperlink (or hot link) - that is, a hyperlink that will activate your Web browser and take it to an external Web site or that will activate your e-mail software for sending a message to a specific e-mail address - should be colored blue. You can see examples of such hyperlinks in this paper. The use of live hyperlinks is at the discretion of the author(s).

Non-live hyperlinks that is, the hyperlinks that are included for the reader’s information but do not actually invoke the reader’s Web browser or e-mail software should be colored black (use package \texttt{url} and \verb+\url{http://...}+). To use live hyperlinks in a proceedings paper, use the \texttt{hyperref} package. If you are using \texttt{PDFLaTeX} to generate your PDF file then, as was done for this file, you should use the following as the last \verb+\usepackage+ command that is already in the preamble.

\begin{verbatim}
\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,
anchorcolor=black,linkcolor=black]{hyperref}
\end{verbatim}

On the other hand, if you are using the traditional \texttt{latex - dvips - ps2pdf} route, then users of MiKTeX or PCTeX for Windows should add the command

\begin{verbatim}
\usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,
anchorcolor=black,linkcolor=black]{hyperref}
\end{verbatim}

as the last \verb+\usepackage+ command in the preamble, while users of Y\&Y TeX should add the command

\begin{verbatim}
\usepackage[dvipsone,colorlinks=true,urlcolor=blue,citecolor=black,
anchorcolor=black,linkcolor=black]{hyperref}
\end{verbatim}

as the last \verb+\usepackage+ command in the preamble.

In general the \verb+\usepackage+ command above that works for MiKTeX running on a Windows system should also work for most implementations of \LaTeX\ running on a Unix or Apple system.
Thus the hypertext link \href{http://www.scs.org}{conference website}~\cite{SCS} to the conference website can be established by the command

\begin{verbatim}
\href{http://www.scs.org}{conference website}~\cite{SCS}
\end{verbatim}

It is recommended to add all hypertext references to the \texttt{.bib} file and to refer to them from the text as it is done in the example above.

If the authors use hyperlinked text in the main body of the paper, they must ensure that each hyperlink includes a citation following the hyperlinked text, a corresponding entry is provided in the list of references, and the associated web address displayed for the hyperlink is complete and correct so that a reader who has only a hard copy of the paper can still access the cited material by typing the relevant part of the displayed text of the hyperlink into the address bar of a Web browser. If the authors opt for including the web address in the main body of the text itself, they must ensure that the hyperlink is complete and correct for the same reason. Again, it is recommended that corresponding web addresses and email addresses should be provided in the references.

If you use the package \texttt{hyperref} as suggested here, and if you use citation commands to handle references, then your citations will
become click-able hyperlinks (as in this document). The same happens to all the references to equations, figures a tables. This can aid the reader in navigating the document.

%\subsection{Citing a Reference}
%To cite a reference in the text, use the author-date method. Thus,~\citeN{chi89} could also be cited parenthetically~\cite{chi89}. For a work by four or more authors, use an abbreviated form. For example, a work by Banks, Carson, Nelson~and Nicol would be cited in one of the following ways:~\shortciteN{bcnn:simulation} or~\shortcite{bcnn:simulation}.
%
%Parenthetical citations are enclosed in parentheses $(~)$, not square brackets $[~]$. The items in a series of such citations are usually separated by commas. If an item in the series of parenthetical citations contains punctuation because (for example) it refers to a work with three or more coauthors, then all items must be separated by semicolons.
%
%The following is a list of \textbf{correct} forms of citations:
%\begin{itemize}
%\item Brown and Edwards (1993),
%\item (Brown and Edwards 1993),
%\item (Smith 1987, Brown and Edwards 1993), and
%\item (Smith 1987; Arnold, Brown, and Edwards 1992; Brown et al. 1997).
%\end{itemize}
%
%The following is a list of \textbf{incorrect} forms of citations:
%\begin{itemize}
%\item Brown and Edwards [1993],
%\item (Brown and Edwards, 1993),
%\item (Smith 1987; Brown and Edwards, 1993), and
%\item (Smith 1987, Arnold Brown and Edwards 1992, Brown et al. 1995)
%\end{itemize}
%
%For further details, please refer to \emph{The Chicago Manual of Style}~\cite{chicago03}. In \autoref{sec:bibtex} you can see how correct citations can easily be achieved by using \BibTeX.
%
%\subsection{List of References}
%Place the list of references after the appendices. The section heading is \textbf{REFERENCES}, and is not numbered. List only references that are cited in the text. Arrange the references in alphabetical order (chronologically for a particular author or group of authors); do not number the references.
%Give complete references without abbreviations. To identify multiple references by the same authors and year, append a lower case letter to the year of publication, for example, 1984a and 1984b. Use hanging indentation to distinguish individual entries. Do not insert additional space between references.
%
%You can enter the references using (a) \emph{\BibTeX\ as discussed in \autoref{sec:bibtex}}, (b) using the environment \texttt{thebibliography} via the \verb+\bibitem+ and \verb+\cite+ commands, or (c) the \texttt{hangref} environment as shown below.
%Please note that neither (b) nor (c) are recommended. These alternatives may mean extra work for you and the editor during the editing process. Option (c) means in addition that the references will not be hyperlinks - as the proceedings are electronic proceedings this is not recommended at all.
%
%\emph{Whatever you do: the list of entries/items need to be included in your submission! And it needs to be included in such a way that the document can be compiled by the editor.}
%
%The bibliographic style for a journal article is: \\
%\textless{}Surname of first author\textgreater{}, \textless{}First author initials\textgreater{},
%\textless{}Initials and surnames of other authors\textgreater{}. \textless{}year\textgreater{}.
%\textless{}Capitalized article title in quotes\textgreater{}. \textless{}\emph{Journal Name in
%Headline Italics}\textgreater\ \textless{}Volume number\textgreater{}, \textless{}page numbers\textgreater{}.
%
%The format for other types of reference can be inferred from the examples in the references, which include:
%\begin{itemize}
%\item a technical report~\cite{chi89},
%\item a proceedings article~\cite{cheng:input94},
%\item a journal article~\cite{gupta:mnormal},
%\item a book by 2 authors~\cite{hammersley:montecarlo},
%\item a chapter in a book~\cite{sch79},
%\item an unpublished thesis or dissertation~\cite{ste99},
%\item a book with no identified authors~\cite{chicago03}, and
%\item a document available on the web~\cite{SCS}.
%\end{itemize}
%
%Be sure that references include all of the necessary information such as full name of the publication, year, name of the proceedings, journal or book title, page numbers, etc.
%
%Again, for further details and examples, please refer to \emph{The Chicago Manual of Style}~\cite{chicago03}. Please note that the examples given in the reference section of this document are based on the 16th edition of the Chicago Manual of Style. Clarity and consistency should be your primary concern.

\section{Using \BibTeX}
\label{sec:bibtex}
To cite a reference in the text, the author-date method is used. Thus,~\citeN{chi89} could also be cited parenthetically~\cite{chi89}. For a work by four or more authors, use an abbreviated form. For example, a work by Banks, Carson, Nelson~and Nicol would be cited in one of the following ways:~\shortciteN{bcnn:simulation} or~\shortcite{bcnn:simulation}.

Using \BibTeX\ for referencing is the recommended way. Please, do \textbf{not} typeset references and citations manually. Indeed, the references in this document were generated using \BibTeX, so the source for this document serves as an example of how to use \BibTeX\ to meet the formatting requirements.

One benefit of using \BibTeX\ is that bibliography formatting and referencing can be greatly simplified: the correct citation and reference list style is automatically created. We assume that you already know how to use \BibTeX.

Software to manage \BibTeX\ files, for example JabRef (Java based), can support you on managing and creating valid \texttt{bib} files.
\emph{Please open your bib file with a software like JabRef BEFORE you submit your final version. Experience shows that almost all manually edited bib files contain duplicated bib keys (which means a random selection of references), broken entries which usually lead to missing bibliographic information, invalid keys, and last but not least invalid tokens in bib files. Bib files DO NOT support comments. \BibTeX\ should not report any error for your final submitted document.}

The \BibTeX\ input file \texttt{scsproc.bst} and the \LaTeX\ macros found in \texttt{scsprocbib.tex} are provided, so no other files (apart from your bibliography) are required. The macros in these files have been tested with \LaTeX. The simplest way to write a paper that uses \BibTeX\ is to take the source file for this document, and modify it to generate your article.

\subsection{Set Up the \BibTeX\ Input Files}

\BibTeX\ requires a bibliography style file (extension \texttt{.bst}) and a bibliography database file (extension \texttt{.bib}).  This is achieved using

\begin{verbatim}
\bibliographystyle{scsproc}
\bibliography{demobib}
\end{verbatim}

just before the AUTHOR BIOGRAPHY section.  The file \texttt{demobib}\ in the \verb+\bibliography+ command should be replaced with the base names of your \BibTeX\ \texttt{.bib} files that you use for your bibliography.  \BibTeX\ is then run as usual to create a bibliography file (\texttt{.bbl}).

\subsection{Use the Citation Macros}
There are a number of macros available to cite references in the \LaTeX\ source document.  The \verb+\cite+ macro can be used to give a list of references in parentheses.  For example,

\begin{verbatim}
\cite{law:simulationc,cheng:queuehetero}
\end{verbatim} % no space after the comma!

results in the citation~\cite{law:simulationc,cheng:queuehetero}. A reference that functions as a noun is created with the \verb+\citeN+ macro. For example,

\begin{verbatim}
\citeN{law:simulationc} say \ldots
\end{verbatim}

results in:~\citeN{law:simulationc} say \ldots\,.

Citations within parentheses do not need the extra parentheses provided by the above citation commands. To suppress the inclusion of extra parentheses, use the \verb+\citeNP+ macro. To obtain (\citeNP{cheng:queuehetero}, \citeNP{law:simulationc}), for example, use:

\begin{verbatim}
(\citeNP{cheng:queuehetero}, \citeNP{law:simulationc}).
\end{verbatim}

When there are four or more authors, the name of the first author should be given along with the text ``et al.''  This can be achieved with the \verb+\shortcite+ macro. To obtain~\shortcite{bcnn:simulation}, for example, use:

\begin{verbatim}
\shortcite{bcnn:simulation}
\end{verbatim}

The macros \verb+\shortciteN+ and \verb+\shortciteNP+ are also available to obtain `et al.' when a citation with many authors is used as a noun.

When citing a reference inside a figure or table caption, please use \verb+\protect\cite+ sequence to avoid build errors.

\subsection{Generate the Bibliography File}

Run \texttt{PDFLaTeX} (or \LaTeX), then \BibTeX, and then \texttt{PDFLaTeX} two more times. Running \texttt{PDFLaTeX} the first time creates the \texttt{.aux} file. Running \BibTeX\ creates the \texttt{.bbl} file.  Running \texttt{PDFLaTeX} again (twice) fixes the bibliography and citation references.

\subsection{Include the Bibliography File in Your Submission}
\label{sec:submitbib}

Be sure to include your \texttt{.bib} file(s) or your \texttt{.bbl} file as part of your submission. If you only include the \texttt{.bbl} file, then please verify that you include the most up-to-date version reflecting changes during the editing process by rerunning \BibTeX\ one last time before submission.

Please be aware that submitting the \texttt{.bbl} file instead of the \texttt{.bib} file means extra work for the editing team and for you, as any changes to the reference list need to be done by you in this case.
Please open and save the file before every submission with a software like JabRef to see whether the file is correct and to check for duplicated entries and/or bib keys in the file.

\section{Author Checklist}
We strive for a consistent appearance in all papers published in the proceedings. If you used the template and styles within this author’s kit, then almost all of the requirements in this checklist will be automatically satisfied, and there is very little to check.

Please \textbf{print a hardcopy of your paper}, and go over your printed paper to make sure it adheres to the following requirements. \textit{Thank you!}

\begin{enumerate}
	\item Abstract
  \begin{enumerate}
	  \item 150 or fewer words.
	  \item Provide 3-5 keywords (mandatory). This set of keywords will identify your paper in indices and databases.
	\end{enumerate}
	\item Paper Length
  \begin{enumerate}
	  \item At least 5, but no more than 12 pages.
	  \item Page size is letter size (8.5’’ x 11’’, or 216 mm x 279 mm).
	\end{enumerate}
	\item All text is in 11-Point Times New Roman except title, header and footer (default in this template).
	\item Paper title is in 12-Point Times New Roman \textbf{BOLDFACE ALL CAPS} (default in this template).
	\item The paper has been spellchecked using U.S. English. 
	\item Spacing and Margins
  \begin{enumerate}
	  \item Single spaced.
	  \item Left and right margins are each 1 inch (default in this template).
	  \item Top and bottom margins are according to the template.
	  \item Title starts 1.25 inches from the top of the page (default in this template).
	\end{enumerate}
	\item Section Headings
  \begin{enumerate}
	  \item Left justified and set in \textbf{BOLDFACE ALL CAPS} (default in this template).
	  \item Numbered, except for the abstract, acknowledgments, references and author biographies.
	  \item Subsection headings are set in \textbf{Boldface Headline Style} (default in this template).
	\end{enumerate}
	\item No footnotes or page numbers.
	\item The copyright notice on the first page contains the name of the symposium or a track where you want to submit your paper, and the running head on subsequent pages is the surnames of all authors.
	\item Multiple authors are formatted correctly.
	\item Equations are centered and any equation numbers are in parentheses and right-justified (default).
	\item Figures and Tables
  \begin{enumerate}
	  \item All text in figures and tables is readable.
	  \item Table captions appear above the table.
	  \item Figure captions appear below the figure.
	\end{enumerate}
	\item Citations and References
  \begin{enumerate}
	  \item Citations are by author and year, and are enclosed in parentheses, not brackets (default \BibTeX). 
	  \item References are in the hangref style, and are listed alphabetically by the last names(s) of the author(s) (also default \BibTeX\ setting in this template). 
	\end{enumerate}
	\item Author biographies are one paragraph per author.
	\item Hyperlinks
  \begin{enumerate}
	  \item Be sure that hyperlinks will probably work in the future as well.
	  \item Live hyperlinks are blue. Nonlive hyperlinks are black (default in this template).
	\end{enumerate}
	\item All fonts must be embedded in the resulting \texttt{.pdf}. If you use \texttt{pdflatex}, this should already be true--unless you included figures in the \texttt{.eps} or \texttt{.pdf} format which may introduce additional font dependencies. You can use the \texttt{pdffonts} command to see if all the fonts are embedded (the column ``emb'' should say yes for all rows). You can use GhostScript to force embed the fonts, like so: {\texttt{gs -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -dEmbedAllFonts=true -sOutputFile=out.pdf -f in.pdf}} (in Windows, use \texttt{gswin32} or \texttt{gswin32c} to invoke GhostScript).
\end{enumerate}

After verifying that your paper meets these requirements, please go to the final submission page at conference website and submit your paper. Be sure to complete the transfer of copyright form and upload the \texttt{.pdf} receipt. \textit{Thank you for contributing to the SCS conferences!}

\section*{Acknowledgments}
Place the acknowledgments section, if needed, after the main text, but before any appendices and the references. The section heading is not numbered. These instructions are adapted from WSC instructions with permission from WSC BoD~\cite{WSC} that have been iteratively updated and improved by proceedings editors and several other individuals, who are too numerous to name separately since the first set of instructions were written by Barry Nelson for the 1991 WSC.

\appendix

\section{Appendices} \label{app:quadratic}
Place any appendices after the acknowledgments. Note that the default style labels them A, B, C, and so forth (please, do not change this). 
The solution to \eqref{eq:quadratic} has the form:

\begin{equation} \label{eq:quadratic sol}
x = \frac{-b \pm \sqrt{b^2-4ac}}{2a} \mbox{ if } a \ne 0\;.
\end{equation}

\section{Getting Help}
If you need help in preparing your paper, contact the proceedings editors. You can reach the entire team by writing to our unified point of contact at \href{mailto://scs@scs.org}{scs@scs.org}.

\section{Most Observed Mistakes}

The following list comprises \textbf{the most common sources of error} that had to be corrected by previous editors. Please make sure to go through the following list and check that your paper is formatted correctly:
\begin{enumerate}
	\item	The paper is less than 5 or more than 12 pages long.
	\item	Paper title and section titles are in \textbf{BOLD ALL CAPS}, subsections are \textbf{Bold and Capitalize the First Letter of Important Words}. Please use the templates.
	\item	Paper is in A4 format, not letter format. Please use the required margins.
	\item	The copyright notice is incorrect.
	\item	The running heads are incorrect. Do not forget that the LastNameLastAuthor is preceded by ``, and ''.
	\item	The citation format is incorrect. Use \BibTeX\ for citations and do not change the format used by the template.
	\item	The biographies are missing. Do not forget the ``author biographies'' section.
	\item	Figures or tables are not referenced in the text or have the incorrect caption format.
	\item	The author section after the title is not formatted correctly, the number of organizations does not define the number of blocks, or the number of blocks does not define the layout.
	\item In the heading on the title page, country names are in all capitals.
	\item	Paragraphs are not indented.
	\item Some fonts in the resulting \texttt{.pdf} are not embedded.
\end{enumerate}

% Please don't change the bibliographystyle style
\bibliographystyle{scsproc}
% AUTHOR: Include your bib file here
\bibliography{demobib}

\section*{Author Biographies}

\textbf{\uppercase{THERESA M. K. ROEDER}} is an Associate Professor of Decision Sciences at San Francisco State University. She holds a PhD in Industrial Engineering and Operations Research from UC Berkeley. Her research interests lie in O.R. education and simulation modeling, especially in healthcare and higher education. Her email address is \email{scs19roeder@gmail.com}.

\textbf{\uppercase{PETER I. FRAZIER}} is an Associate Professor in the School of Operations Research and Information Engineering at Cornell University. He holds a Ph.D. in Operations Research and Financial Engineering from Princeton University. His research interests include optimal learning, e-commerce, and the physical sciences.  His email address is \email{scs19frazier@gmail.com}.

\textbf{\uppercase{HENRY WOTTON}} is a Full Professor in the School of Operations Research and Information Engineering at Cornell University. He holds a Ph.D. in Industrial Engineering from UC Berkeley, His research interests include modeling and simulation. His email address is \email{scs19henrywotton@gmail.com}.

\textbf{\uppercase{ROBERTO SZECHTMAN}} received his Ph.D. from Stanford University, and currently is Associate Professor in the Operations Research Department at the Naval Postgraduate School. His research interests include applied probability and military operations research. His email address is \email{scs19szechtman@gmail.com}.

\textbf{\uppercase{JONATHAN HARKER}} received his Ph.D. from UCLA, and currently is Assistant Professor in the Operations Research Department at the Naval Postgraduate School. His research interests include distributed simulations and military research. His email address is \email{scs19jonathanharker@gmail.com}.

\textbf{\uppercase{ENLU ZHOU}} is an Assistant Professor in the H. Milton Stewart School of Industrial \& Systems Engineering at Georgia Institute of Technology. She holds a Ph.D. in Electrical Engineering from the University of Maryland, College Park. Her research interests include stochastic control and simulation optimization, with applications in financial engineering and revenue management. Her email address is \email{scs19zhou@gmail.com}.

\newpage

\begin{figure*}[htb]
{
\centering
First Name Last Name 1 \\
First Name Last Name 2 \\
\vspace{12pt}
Affiliation \\
Address, City, Country \\
E-mail address
\caption{Example title page heading with 2 authors from the same institution.\label{fig2same}}
}
\end{figure*}

\begin{figure*}[htb]
{
\centering
\begin{tabular}{ccc}
\phantom{Entries to adjust spacing - ignore} & \phantom{intermediate space} & \phantom{Entries to adjust spacing - ignore} \\
First Name Last Name 1 & & First Name Last Name 2 \\
\\
Affiliation 1 & & Affiliation 2 \\
Address, City, Country 1 & & Address, City, Country 2 \\
E-mail address 2 & & E-mail address 2 
\end{tabular}
\caption{Example title page heading with 2 authors from different institutions.\label{fig2different}}
}
\end{figure*}

\begin{figure*}[htb]
{
\centering
\begin{tabular}{ccc}
\phantom{This adjusts spacing - ignore} & \phantom{This adjusts spacing - ignore} & \phantom{This adjusts spacing - ignore} \\
First Name Last Name 1 & & First Name Last Name 2 \\
\\
Affiliation 1 & & Affiliation 2 \\
Address, City, Country 1 & & Address, City, Country 2 \\
E-mail address 1 & & E-mail address 2 \\
\\ \\
& First Name Last Name 3 \\
\\
& Affiliation 3\\
& Address, City, Country 3\\
& E-mail address 3 
\end{tabular}
\caption{Alternate example title page heading with 3 authors from different institutions. \label{fig3different}}
}
\end{figure*}

\begin{figure*}[htb]
{
\centering
\begin{tabular}{ccc}
\phantom{Adjust spacing using these entries} & \phantom{intermediate space} & \phantom{Adjust spacing using these entries} \\
First Name Last Name 1 & & First Name Last Name 2 \\
\\
Affiliation 1 & & Affiliation 2 \\
Address, City, Country 1 & & Address, City, Country 2 \\
E-mail address 1 & & E-mail address 2 \\
\\ \\
First Name Last Name 3 & & First Name Last Name 4 \\
\\
Affiliation 3 & & Affiliation 4 \\
Address, City, Country 3 & & Address, City, Country 4 \\
E-mail address 3 & & E-mail address 4 
\end{tabular}
\caption{Example title page heading with 4 authors from different institutions.\label{fig4different}}
}
\end{figure*}

\begin{figure*}[htb]
{
\centering
\begin{tabular}{ccc}
\phantom{Entries to adjust spacing - ignore} & \phantom{intermediate space} & \phantom{Entries to adjust spacing - ignore} \\
First Name Last Name 1 & & First Name Last Name 3 \\
First Name Last Name 2 & & \\
\\
Affiliation & & Affiliation 3 \\
Address, City, Country & & Address, City, Country 3 \\
E-mail address & & E-mail address 3 
\end{tabular}
\caption{Example title page heading with 3 authors from 2 different institutions. \label{fig2equals1different}}
}
\end{figure*}

\end{document}

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/experiments/code/system.py

@@ -1,0 +1,13 @@
+# Save "data" in a file titled "file_name" using pickle.
+def save(data, file_name="_save.pkl"):
+    import dill as pickle
+    with open(file_name, "wb") as f:
+        pickle.dump(data, f)
+
+
+# Load data from a pickle file titled "file_name".
+def load(file_name="_save.pkl"):
+    import dill as pickle
+    with open(file_name, "rb") as f:
+        data = pickle.load(f)
+    return data

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/experiments/example_2.py

@@ -1,0 +1,69 @@
+import numpy as np
+# Include the "code" directory into the path and import code custom to this project.
+import sys, os
+sys.path += [os.path.abspath("code")]
+from plot import Plot
+from polynomial import polynomial_piece
+sys.path.pop(-1)
+
+c = "rgba(50,50,50,.5)"
+a = .8
+p = Plot("","α","β", font_family="times", font_size=18)
+p.add("Region 1", [0, 3, 3], [3, 3, 0], mode="lines", fill="tonext",
+      color=p.color(0), group=0)
+p.add_func("Region 2", lambda x: (9 - x**2)**(1/2), [0,3], group=1,
+           color=p.color(2), fill="tonext")
+p.add("Region 3", [0, 3], [3, 0], mode="lines", fill="tonext", 
+      color=p.color(3), group=2)
+p.add("Region 4", [0, 1, 3], [3, 1, 0], mode="lines", fill="tozeroy", 
+      color=p.color(1), group=3)
+
+p.add("Point 1 to Region 1", [3,2], [4.5,3], color=c, group=0,
+      line_color=p.color(0,alpha=a), mode="markers+lines", dash=None)
+p.add("Point 2 to Region 2", [4,3*4/(4**2+3**2)**(1/2)], [3,3*3/(4**2+3**2)**(1/2)], 
+      group=1, color=c, line_color=p.color(2, alpha=a),
+      mode="markers+lines", dash="dot")
+p.add("Point 3 to Region 3", [4,12/5], [1,3/5], color=c, group=2,
+      line_color=p.color(3,alpha=a), mode="markers+lines", dash="dash")
+p.add("Point 4 to Region 4", [1.5,9/13], [3.5,21/13], color=c, group=3,
+      line_color=p.color(1,alpha=a), mode="markers+lines", dash="dashdot")
+
+p.add_annotation(" 4 ", .3, .3, font_family="times", font_size=17,
+                 show_arrow=False, border_width=1, x_anchor="left")
+p.add_annotation(" 3 ", 1.02, 1.02, font_family="times", font_size=17,
+                 show_arrow=False, border_width=1, x_anchor="left")
+p.add_annotation(" 2 ", 1.55, 1.55, font_family="times", font_size=17,
+                 show_arrow=False, border_width=1, x_anchor="left")
+p.add_annotation(" 1 ", 2.3, 2.3, font_family="times", font_size=17,
[…]

### 2020-01-16 [2019-11]_HPC_(quintic_spline)/experiments/test.py

@@ -1,0 +1,22 @@
+import numpy as np
+from util.system import Timer
+from parallel import map, builtin_map
+
+def floated(v):
+    try: return [float(i) for i in v]
+    except: return float(v)
+
+t = Timer()
+
+t.start()
+out = builtin_map(floated, list(range(100000000)))
+t.stop()
+print(np.array(out))
+print("Serial time:", t())
+
+t.start()
+out = map(floated, list(range(10000)))
+t.stop()
+print(np.array(out))
+print("Parallel time:", t())
+

### 2020-01-16 [2019-12-16]_HPC_(mpca)/code/util/data.py
Created paper repository with all of my work.

@@ -1,0 +1,2006 @@
+#      data.py     
+# =================
+COMMON_SEPERATORS = [",", " ", "	", ";"]
+
+UPDATE_FREQUENCY = .5  # How much time (in seconds) between updates when reading data.
+MAX_ERROR_PRINTOUT = 10 # Only print out this many errors when processing data
+
+import numpy as np
+NP_TYPES = {str:(np.str_, 16),    # Which numpy types correspond to 
+            int:(np.int64,),      # the python types of variables
+            float:(np.float64,)}
+PY_TYPES = {value[0]:key for (key,value) in NP_TYPES.items()}
+GENERATOR_TYPE = type(_ for _ in ())
+
+# coding: future_fstrings
+# from future import print_function
+
+# TODO:  Add tests for empty data object.
+# TODO:  Add "Descriptor" class for Data.names and Data.types so that
+#        they can be directly modified by users safely (with checks).
+# TODO:  Add "Row" class that protects item assignment types over rows
+#        and allows for item-wise equality operator.
+# TODO:  Non-singleton comparison operator for Column Column comparison.
+# TODO:  Add addition, subtraction, multiplication, and division
+#        operators to column objects.
+# TODO:  Make "predict" not require that target columns be convertible
+#        to numeric form, only that they operate under a weighted sum.
+# TODO:  Make printout shorter for data with lots of columns.
+# TODO:  Check the __iadd__ for data object columns, breaking "missing".
+# TODO:  Update __iadd__ to behave differently when adding data:
+#        when adding data with subset of known columns, add rows and
+#        columns (for unknown) and None for missing.
+# TODO:  Make "to_matrix" automatically flatten elements of columns
+#        that are typed as numeric numpy arrays.
+
+QUOTES = {'"'}
+DEFAULT_DISPLAY_WAIT = 3.
+
+# Define a class for reading a video file without actually storing it in memory.
+class IndexedVideo:
+    import cv2
+    import numpy as np
+    class FailedSet(Exception): pass
+    class FailedRead(Exception): pass
+    class BadIndex(Exception): pass
+    class IndexTooLarge(Exception): pass
+    # Initialize by getting the information about this video from file.
+    def __init__(self, video_file_name, flatten=False, use_float=False):
+        # Store the 'flatten' attirbute for flattening images.
+        self.flatten = flatten
+        self.float = use_float
[…]

### 2020-01-16 [2019-12-16]_HPC_(mpca)/code/util/math/pairs.py

@@ -1,0 +1,134 @@
+import numpy as np
+
+# This function maps an index in the range [0, count*(count - 1) // 2] 
+# to a tuple of integers in the range [0,count). The mapping will cover
+# all pairs if you use all indices between [0, count*(count - 1) // 2].
+def pair_to_index(p1, p2):
+    if (p1 < p2): p1, p2 = p2, p1
+    return (p1 * (p1 - 1) // 2) + p2
+
+# This function maps an index in the range [0, count*(count - 1) // 2] 
+# to a tuple of integers in the range [0,count). The mapping will cover
+# all pairs if you use all indices between [0, count*(count - 1) // 2].
+def index_to_pair(index):
+    val = int(((1/4 + 2*index)**(1/2) + 1/2))
+    remainder = index - val*(val - 1)//2
+    return (val, remainder)
+
+# Compute the number of original elements when given the number of pairs.
+# This is done by realizing the following:
+# 
+#      n(n-1) / 2 = f(n)
+#   (n^2 - n) / 2 = f(n)
+#         n^2 - n = 2 f(n)
+#   n^2 - n + 1/4 = 2 f(n) + 1/4
+#     (n - 1/2)^2 = 2 f(n) + 1/4
+#         n - 1/2 = (2 f(n) + 1/4)^{1/2}
+#               n = (2 f(n) + 1/4)^{1/2} + 1/2
+#              2n = (8 f(n) + 1)^{1/2} + 1
+#               n = ((8 f(n) + 1)^{1/2} + 1) / 2
+# 
+def num_from_pairs(num_pairs):
+    return int(round( ((8*num_pairs + 1)**(1/2) + 1) / 2 ))
+
+# Define a custom error for passing arrays of the wrong shape to pairwise.
+class ShapeError(Exception): pass
+
+# Compute the distance between all pairs of elements in a single list,
+# or the distance of all combined pairs between two lists.
+def pairwise_distance(x1, x2=None):
+    # Define an error for having the wrong array shape.
+    if (len(x1.shape) != 2): raise(ShapeError("Only 2D NumPy arrays are allowed."))
+    # Determine use case.
+    if (type(x2) == type(None)):
+        # Compute the pairwise distances.
+        x1_sq = np.sum(x1**2, axis=1, keepdims=True)
+        d = (x1_sq + x1_sq[:,0]) - 2 * np.matmul(x1, x1.T)
[…]

### 2020-01-16 [2019-12-16]_HPC_(mpca)/code/util/stats/distributions.py

@@ -1,0 +1,312 @@
+from util.stats import *
+
+# ===================================
+#      CDF and PDF Fit Functions     
+# ===================================
+
+# A class for handling standard operators over distributions. Supports:
+#    addition with another distribution
+#    subtraction from another distribution (KS-statistic)
+#    multiplication by a constant in [0,1]
+class Distribution():
+    def __init__(self, func):
+        self._function = func
+        standard_attributes = ["min", "max", "derivative", 
+                               "integral", "inverse", "nodes"]
+        # Copy over the attributes of the distribution function.
+        for attr in standard_attributes:
+            if hasattr(func, attr): setattr(self, attr, getattr(func, attr))
+
+    # Define addition with another distribution.
+    def __radd__(self, func):
+        # Special case, when distributions are used in a "sum" then an
+        # integer will be added as the "start" of the sum. Ignore this.
+        if (type(func) == int): return self
+        # Get the new minimum and maximum of this distribution and "func".
+        new_min = min(self.min, func.min)
+        new_max = max(self.max, func.max)
+        def new_func(x=None, func=func):
+            if (type(x) == type(None)): return (new_min, new_max)
+            else:                       return self._function(x) + func(x)
+        new_func.min = new_min
+        new_func.max = new_max
+        # Copy over the attributes of the distribution function.
+        standard_attributes = ["derivative", "integral"]
+        for attr in standard_attributes:
+            if (hasattr(self, attr) and hasattr(func, attr)):
+                mine = getattr(self, attr)
+                other = getattr(func, attr)
+                setattr(new_func, attr, lambda x: mine(x) + other(x))
+        # Return a distribution object.
+        return Distribution(new_func)
+    __add__ = __radd__
+
+    # Define multiplication by a number (usually a float).
+    def __rmul__(self, number):
+        def new_func(x=None, m=number, f=self._function):
+            if (type(x) == type(None)): return f(None)
+            else:                       return m * f(x)
+        new_dist = Distribution(self)
+        new_dist._function = new_func
+        # Copy over the attributes of the distribution function.
+        standard_attributes = ["derivative", "integral"]
+        for attr in standard_attributes:
+            if hasattr(self, attr):
+                f = getattr(self, attr)
+                setattr(new_dist, attr, lambda x: number * f(x))
+        return new_dist
[…]

### 2020-01-16 [2019-12-16]_HPC_(mpca)/paper.tex
Created paper repository with all of my work.

%**************************************************************************
%* SpringSim 2020 Author Kit
%*
%* Word Processing System: TeXnicCenter and MiKTeX
%*
%**************************************************************************

\documentclass{scspaperproc}

\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{mathptmx}

%
%****************************************************************************
% AUTHOR: You may want to use some of these packages. (Optional)
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{amsthm}
%****************************************************************************

%
%****************************************************************************
% AUTHOR: If you do not wish to use hyperlinks, then just comment
% out the hyperref usepackage commands below.

%% This version of the command is used if you use pdflatex. In this case you
%% cannot use ps or eps files for graphics, but pdf, jpeg, png etc are fine.

\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black,bookmarks=false]{hyperref}

%% The next versions of the hyperref command are used if you adopt the
%% outdated latex-dvips-ps2pdf route in generating your pdf file. In
%% this case you can use ps or eps files for graphics, but not pdf, jpeg, png etc.
%% However, the final pdf file should embed all fonts required which means that you have to use file
%% formats which can embed fonts. Please note that the final PDF file will not be generated on your computer!
%% If you are using WinEdt or PCTeX, then use the following. If you are using
%% Y&Y TeX then replace "dvips" with "dvipsone"

%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
%% anchorcolor=black,linkcolor=black]{hyperref}

%% The use of the long citation format (e.g. "Brown and Edwards (1993)" rather than "[5]") and at the same
%% time using the hyperref package can lead to hard to trace bugs in case the citation is broken accross the
%% line (usually this will mark the entire paragraph as a hyperlink (clickable) which is easily noticeable and fixed
%% if using colorlinks, but not if the color is black -- as it is now). Worse yet, if a citation spans page boundary,
%% LaTeX compilation can fail, with an obscure error message. Since this depends a lot on the flow of the text
%% and wording, these bugs come and go and can be extremely hard for a beginner to trace. The error
%% message can look like this:
%%
%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink.
%%    \AtBegShi@Output ...ipout \box \AtBeginShipoutBox 
%%    \fi \fi 
%%    l.174 
%%    ! ==> Fatal error occurred, no output PDF file produced!
%%
%% and can be universally fixed by putting an \mbox{} around the citation in question (in this case, at line 174)
%% and maybe adapting the wording a little bit to improve the paragraph typesetting, which is perhaps not
%% immediately obvious.
%****************************************************************************

%****************************************************************************
%*
%* AUTHOR: YOUR CALL!  Document-specific macros can come here.
%*
%****************************************************************************

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multiline}[2][c]{{%
  \begin{tabular}%
    [#1]{@{}c@{}}#2%
  \end{tabular}%
}}

%% Use a package to allow for multi-row cells in a table.
\usepackage{multirow}

% add custom hyphenation rules here
\usepackage{hyphenat}
\hyphenation{op-tical net-works semi-conduc-tor}

% If you use theorems
\newtheoremstyle{scsthe}% hnamei
{8pt}% hSpace abovei
{8pt}% hSpace belowi
{\it}% hBody fonti
{}% hIndent amounti1
{\bf}% hTheorem head fontbf
{.}% hPunctuation after theorem headi
{.5em}% hSpace after theorem headi2
{}% hTheorem head spec (can be left empty, meaning `normal')i

\theoremstyle{scsthe}
\newtheorem{theorem}{Theorem}
\renewcommand{\thetheorem}{\arabic{theorem}}
\newtheorem{corollary}[theorem]{Corollary}
\renewcommand{\thecorollary}{\arabic{corollary}}
\newtheorem{definition}{Definition}
\renewcommand{\thedefinition}{\arabic{definition}}

% avoid overrunning the right margin; you are welcome to remove this, provided that you take care not to overrun the right margin anywhere in your paper
\sloppy

%#########################################################
%*
%*  The Document.
%*
\begin{document}

%***************************************************************************
% AUTHOR: Thomas C.H. Lux and Tyler H. Chang
% FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
%
%		You need to change the author listing below!
%               Please list ALL authors using last name only, separate by a comma except
%               for the last author, separate with "and"

\SCSpagesetup{Lux and Chang}

% AUTHOR: Uncomment ONE of these correct conference names.
\def\SCSconferenceacro{SpringSim'20}
%\def\SCSconferenceacro{SummerSim}
%\def\SCSconferenceacro{AutumnSim}
%\def\SCSconferenceacro{PowerPlantSim}

% AUTHOR: Set the correct year of the conference.
\def\SCSpublicationyear{2020}

% AUTHOR: Set the correct month and dates; the dates are separated by a single minus sign
% with no spaces and no leading zeros, the month is a full name (e.g. April) with the first letter
% capitalized. For example, "April 8-13".
\def\SCSconferencedates{May 19-May 21}

% AUTHOR: Set the correct venue in the form "City, State, Country", for example "Los Angeles, CA, USA".
\def\SCSconferencevenue{Fairfax, VA, USA}

% AUTHOR: Enter the title, all letters in upper case
\title{Metric Principle Component Analysis: \\ On Identifying Important Subspaces for Approximation}

% AUTHOR: Enter the authors of the article, see end of the example document for further examples
\author{
Thomas C. H. Lux\\
Tyler H. Chang\\ [12pt]
Department of Computer Science\\
Virginia Polytechnic Institute \& State University\\
2000 Torgersen Hall, Blacksburg, VA, USA\\
\{tchlux,thchang\}@vt.edu\\
}

\maketitle

%% Examples and notes.
%% 
%% \begin{table}[htb]
%%   \centering
%%   \caption{Caption above the table!}\label{tab:example}
%%   \begin{tabular}{r|l}
%%     First & row \\ \hline
%%     second & row \\
%%     third & row \\
%%   \end{tabular}
%% \end{table}
%% 
%% \begin{figure}[htb]
%%   \centering
%%   \includegraphics[width=0.50\textwidth]{path_without_extension}
%%   \caption{Caption below the figure!}\label{fig:example}
%% \end{figure}
%% 
%% \cite{ref}  - Make a citation that is in parenthesis
%% \citeN{ref} - Make a citation out of parenthesis (for noun usage)
%% \shortcite{ref}  - Make a citation for references with >3 authors.
%% \shortciteN{ref} - Make a noun-form citation for references with >3 authors.
%% 
%% 
%% 5 - 12 page paper mandated
%% 150 word abstract
%% 5 keywords maximum

\section*{Abstract}

Many modern data science problems for approximation have high
dimension while the true underlying phenomenon can be accurately
represented from a fraction of the provided information. A variety of
dimension reduction techniques exist, but most target unsupervised
applications rather than approximation problems or only accommodate a
small set of approximation problems. This paper presents metric
principle component analysis (MPCA), an extension of standard
principle component analysis (PCA) that accounts for metric variation
in applied approximation problems like classification and
regression. MPCA is theoretically motivated, shown to be a superset of
classical PCA, and its usefulness as a dimension reduction technique
for approximation problems is demonstrated. Initial results reveal
both that MPCA can be significantly better in some cases, and that in
other applications it performs similarly to standard PCA.

\textbf{Keywords:} principle component analysis, classification,
regression, approximation, metric learning

\section{Introduction}
\label{sec:introduction}

Dimension reduction is an important problem in data science and, more
generally, function approximation. Consider a multivariable function
$f : \mathbb{R}^d \rightarrow (S, m)$ where $(S,m)$ denotes a metric
space with metric $m$. In the context of data science, $(S,m)$ could
represent a discrete classification space, a continuous real-valued
space, or could be descriptive of some graph structure. In this paper,
the supervised learning problem is considered, where a finite set $X
\subset \mathbb{R}^d$ of data points are given along with labeled
function values $Y = \{y_i = f(x_i) | x_i \in X\}$.  There are many
algorithms in the field of data science, mathematics, and machine
learning for predicting general functions such as multivariate spline
approximations, neural networks, clustering algorithms, support vector
machines/regressors, and more. One particularly common yet effective
class of solutions to these problems are distance based approximations
such as k-nearest neighbor (KNN).  Though effective for many classes
of problems, these methods depend on the premise that distance in the
input space can be associated with change in the output class or
value. While this assumption holds for most of the common "big data"
problems, it is often the case that some number of directions in the
input space have no bearing on the output class/value. Without
applying dimension reduction, these directions will increase the cost
of any distance based approximation techniques, in some cases making
the prediction techniques computationally intractable.  Furthermore,
such "meaningless directions" cannot improve prediction accuracy, and
often decrease accuracy by disassociating distance from change in
class/value.

In this paper, a novel technique for performing dimension reduction is
proposed that considers how the output $f(x)$ changes given the data
$X$ and function value/labels $Y$. Using this information, an
orthonormal basis is constructed for a new subspace of $\mathbb{R}^d$
where every direction is associated with some change in $f$, thereby
increasing the accuracy of distance based methods while often
significantly reducing the dimension. The technique is model agnostic,
meaning that it can be applied as a "pre-conditioning" step before any
distance based approximation. The effectiveness of the technique is
demonstrated on one theoretical and three real-world problems, using
four different distance based approximation algorithms as well as a
multi-layer perceptron (MLP).

The remainder of this paper is structured as follows: Section
\ref{sec:background} considers existing dimension reduction techniques
for approximation problems, Section \ref{sec:methodology} explains the
methodology of computing MPCA and provides a theoretical example,
Section \ref{sec:results} shows results on real-world problems,
Section \ref{sec:discussion} discusses the implications of results,
and finally Section \ref{sec:conclusion} concludes.

\section{Background}
\label{sec:background}

\subsection{Related Work}
\label{sec:related}

One of the most standard methods for performing dimension reduction is
correlation analysis, whereby a general model\footnote{The model may be a
multivariable linear or quadratic model, possibly with interaction
terms.} is fit\footnote{Most commonly fits are performed via least-squares
approximation.} to the data and the coefficients of each term are
analyzed to determine whether that term has any effect on the
output. After dropping all terms that have no statistically
significant effect on the function value, it can be inferred that only
the directions associated with the remaining terms affect the output
class or function value. These techniques are highly effective when an
appropriate model is chosen, but are highly model dependent and
generally fail for a black-box output function $f$. For example, this
technique is most commonly applied by fitting a linear model to
labeled data. Then, if the function $f$ is not itself approximately
linear, the above technique can fail to produce any meaningful
results.

Principle component analysis (PCA) is another commonly used dimension
reduction technique. An explanation of Principle Component Analysis
and its applications can be found in
\citeN{PCAtutorial,powell_lehe_2014}. Given a finite set of unlabeled
points $X \subset \mathbb{R}^d$, PCA identifies an ordered set of
orthonormal basis vectors $\{v_1, \ldots, v_d\},$ such that the
projection of $X$ onto any subspace defined by the span of $\{v_1,
\ldots, v_k\},$ $k \leq d$ has maximal variance. Equivalently, PCA is
the dimension reduction technique that minimizes reconstruction error
for $X$ with respect to mean squared error (MSE). In its standard
application, PCA is an unsupervised technique, and does not consider
the effect of each dimension on output class or function values.

Recently a modified variant of PCA was proposed which incorporates
function values into the subspace basis and the technique is referred
to as \textit{supervised PCA}
\shortcite{barshan2011supervised}. However this technique relies on
the coefficients of a linear fit and faces the same nonlinear
approximation weaknesses (as will be demonstrated in a theoretical
example).

Other notable techniques include linear discriminant analysis (LDA)
and neural network autoencoders. LDA depends heavily on linear
separability of the output classes, and therefore can fail for a
general function $f$. In their most naive implementation, autoencoders
are equivalent to standard PCA. However, they can be modified to also
consider output labels, though no convergence guarantees can be made
and the results are purely heuristic. Other feature weighting
techniques have been studied for clustering and generic learning
algorithms
\shortcite{FeatureWeightingForClustering,FeatureWeightingReview}, and
in general a lot of work has been done on feature selection given a
model \shortcite{FeatureSurvey,tenenbaum2000global}. Those survey
papers extensively cover existing techniques and approaches for
weighting (continuous valued) and selecting (binary operation)
features of data for making accurate approximations.

\subsection{Approximation Techniques}
\label{sec:approximation}

Four distance based approximation techniques are used to validate the
proposed dimension reduction technique and additionally a classic
neural network is tested. The first approximation algorithm is
k-nearest neighbor using the 2-norm distance with $k=1,$ and the
second is the same method with $k=10.$ By sorting points in a tree,
k-nearest neighbors is able to scale linearly with the input dimension
and near logarithmically with respect to number of input points. The
second technique is the modified linear Shepard's method called LSHEP
\shortcite{thacker2010algorithm}. LSHEP incorporates a local linear
fit into the the original Shepard's method \shortcite{shepard1968two},
an inverse distance weighting technique that scales linearly with
dimension and linearly with number of points. The final distance based
technique uses the Delaunay triangulation, a piecewise linear
interpolant based on a simplicial mesh of the same name
\shortcite{chang2018polynomial}. Of the above techniques, Delaunay is
the most computationally expensive, scaling approximately linearly
with number of points, but growing prohibitively expensive in
dimension greater than 50. The last technique for general comparison
is a classic neural network called a Multi-Layer Perceptron
(MLP). This work uses the implementation available at
\shortcite{scikit-learn} while choosing the Rectified Linear Unit
(ReLU) activation function and Stochastic Gradient Descent (SGD) error
minimizer. 1000 gradient steps are allowed to be taken and the default
number of hidden nodes (100) is used with one hidden layer.

\section{Methodology}
\label{sec:methodology}

In this section, a novel application of Principle Component Analysis
(PCA) is proposed in order to perform dimension reduction in a
supervised learning (approximation) context. The method will be
referred to as Metric PCA (MPCA).

Let $f : \mathbb{R}^d \rightarrow (S,m)$ be an arbitrary function of
interest, where $(S,m)$ is a metric space and $d$ is a positive
integer.  Let $X \subset \mathbb{R}^d$ be a finite set of points with
known function values (labels) $Y = \{y_i = f(x_i) | x_i \in X\}$.
Consider the set $Z$ of vectors

\begin{equation*}
  Z = \bigg\{ \frac{(x_i - x_j)\  m(y_i, y_j)}{\|(x_i - x_j)\|_2^2} : \quad x_i, x_j \in X, \quad y_i, y_j \in Y \bigg\},
\end{equation*}

\noindent with covariance matrix $Cov(Z).$ Note that since $Cov(Z)$ is
symmetric, its Eigenvectors are orthogonal and form a basis for
$\mathbb{R}^d$. Note that if the original set $X$ contains
$n$ points, then the new set $Z$ must contain $\mathcal{O}(n^2)$
points, since it consists of all vectors \textit{between} points in $X$,
rescaled by change in their corresponding function values (in $Y$).

An Eigenvector decomposition is performed on $Cov(Z),$ then the points
are reduced into the lower-dimensional space spanned by the first $k$
Eigenvectors ranked by the magnitude of their corresponding
Eigenvalues (i.e., PCA on the new set $Z$). A \textit{magnitude} is
assigned to each vector by computing the \textit{total variation} of
the function along each Eigenvector and normalizing the sum of
magnitudes to be one. In practice, instead of performing an
Eigenvector decomposition on $Cov(Z),$ a singular value decomposition
is performed on the matrix $\displaystyle Z^\star$, whose rows are
transposed vectors in $Z$. Then, $Z^\star = U \Sigma V^T$, and the
columns of $V$ are exactly the Eigenvectors of $Cov(Z).$

\subsection{Equivalence of PCA and MPCA Under Euclidean Distance Metric}

In this section it is shown that in the special case where $m(y_i,y_j)
= \|x_i - x_j\|_2$, MPCA is identical to PCA.

Given $X$ as stated above, if the set of all pairwise difference
vectors $Z$ is considered, then $PCA(Z) = PCA(X).$ Denote $Z_j$ to be
the subset of $Z$ about vertex $x_j \in X,$ $Z_j = \{x_j - x_i : x_i
\in X\}.$ Consider the first principle component of $X,$ $v_1 \in
\mathbb{R}^d.$ It is noted that the first principle component of $Z$
is the same, and will follow that all remaining components are
identical.

\begin{align*}
  \max_{\|v\|_2 = 1} \|Zv\|_2^2
  =& \max_{\|v\|_2 = 1} \sum_{z \in Z} \langle z, v \rangle^2
  = \max_{\|v\|_2 = 1} \sum_{Z_j \subset Z} \sum_{z \in Z_j} \langle z, v \rangle^2 \\
  \leq&\ \sum_{Z_j \subset Z} \max_{\|v\|_2 = 1} \sum_{z \in Z_j} \langle z, v \rangle^2 
  = \ \sum_{Z_j \subset Z} \sum_{z \in Z_j} \langle z, v_1 \rangle^2 \\
  =& \sum_{z \in Z} \langle z, v_1 \rangle^2 
  = \ \|Zv_1\|_2^2 \\
  \implies&\max_{\|v\|_2 = 1} \|Zv\|_2 = \|Zv_1\|_2.
\end{align*}

After removing the first component from each vector in $Z$, the same
technique can be reapplied to find the second component. This
methdology can be applied to the remaining components in an inductive
fashion, to show total equivalence of the principle components for $X$
and $Z.$

\begin{figure}
  \centering
  \includegraphics[width=0.30\textwidth]{1-mpca-pca.png}
  \hspace{5mm}
  \includegraphics[width=0.53\textwidth]{1-mpca-pca.pdf}
  \caption{Depicted above from left to right are source data for an
    approximation problem (red dots, on left) and a true underlying
    function (blue mesh, on left), the constructed point set $Z$ with
    metric principle components (middle), and the source point set $X$
    with standard principle components (right). Notice that the MPCA
    components are ordered and weighted in proportion to their
    relevance to the true function, while the PCA components are
    unrelated to the underlying function.}
  \label{fig:analytic}
  %% {{https://people.cs.vt.edu/tchlux/files/[2018-11-28]_mpca_demo_2.html}}
\end{figure}

\subsection{Analytic Demonstration}

Consider the following example. In this, $X^{(50\ \times\ 2)}$ is a
random sample of points drawn from the unit cube and the values come
from a function $f: \mathbb{R}^2 \rightarrow \mathbb{R},$ defined as
$f(x,y) = x^2$ and note that this only depends on the $x$
direction. The results of applying PCA and MPCA to this problem can be
seen in Figure \ref{fig:analytic}. It is clear that the example set of
points are not distributed according the direction of greatest change
in the underlying function. This causes PCA to produce distinctly
different vectors from MPCA, noting that the standard principle
components are not only irrelevant to approximation, they are ordered
incorrectly in terms of importance. It should be noted that this
symmetric underlying function would cause \textit{supervised PCA} to
fail, since no meaningful linear fit can be produced.

\begin{table}
  \centering
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{l l}
    \hline
    \textit{Methods} & PCA, MPCA \\
    \hline
    \textit{Approximators} & KNN (1), KNN (10), LSHEP, Delaunay, MLP \\
    \hline
    \textit{Number of samples (for MPCA)} & n, 10n \\
    \hline
    \textit{Fraction of original dimension (for reduction)} & $1/3$, $1/10$, $1/100$ \\
    \hline
  \end{tabular}
  \vspace{5mm}
  \caption{All possible combinations of the above settings are
    considered in experiments. Four distance based techniques are
    tested as well as a classical neural network. Since MPCA requires
    the construction of a pairwise difference set of points, random
    samples of size $n$ and $10n$ are drawn to prevent drastic
    increase in the amount of data.}
  \label{tab:evaluation}
\end{table}

\subsection{Evaluation}

Given theoretical interest in MPCA, it is important to evaluate the
performance of MPCA by applying it to real approximation problems. In
this section, MPCA is applied to three such real problems, two image
classification problems (MNIST, CIFAR10) and one regression problem
(Yelp rating prediction). First, the performance of four common
approximation algorithms applied to the raw training data is evaluated
using four-fold cross validation. Then, the dimension is reduced to
various percentages of the total dimension using both MPCA and PCA and
the performance of each algorithm is reevaluated. Recall that when $X$
contains $n$ data points, the transformed set $Z$ contains
$\mathcal{O}(n^2)$ points. Therefore, MPCA requires the singular value
decomposition of a matrix $\displaystyle Z^\star$ with $d$ columns and
$\mathcal{O}(n^2)$ rows.  Since this is not computationally feasible
when $n$ is large, a random sample of vectors in $Z$ can be
taken. Throughout Section \ref{sec:results}, all combinations of the
values listed in Table \ref{tab:evaluation} are tested.

\section{Results}
\label{sec:results}

\textbf{\textit{Yelp:}} \shortciteN{yelp2018vegas} is a collection of 479 American-style
restaurant ratings from Las Vegas. Most of the data is composed of
categorical descriptors, there are 63 features. This is a regression
problem, where the algorithm predicts the star rating of a restaurant
on 0-5 scale with .5 intervals.

\textbf{\textit{MNIST:}} \shortciteN{lecun_cortes_burges_2008} provide a collection 60,000
images that are randomly reduced to 10,000 black and white images,
each with shape [28 x 28] having 784 channels. The data poses a
classification problem with 10 unique classes, which are 10 digits
handwritten.

\textbf{\textit{CIFAR10:}} \shortciteN{krizhevsky2009learning} share a
collection of 50,000 images that are randomly reduced to 10,000 color
images, each with shape [32 x 32 x 3] having 3072 channels in
total. The data poses a classification problem with 10 unique classes
that are common objects seen in each image.

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{2-mpca-MNIST-components.png}
  \caption{Above are the first 12 components produced by MPCA over the
    MNIST hand-written digit classification data set. The original
    MNIST data is in strictly black and white, these images are
    normalized to unit 2-norm for visualization purposes. The red
    coloring indicates large magnitude negative values, black is
    neutral, and white is large positive values. Notice the
    resemblance of these components to natural digits.}
  \label{fig:mnist}
  %% {{https://people.cs.vt.edu/tchlux/files/[2018-11-28]_MNIST_12_MPCA_Components.png}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{2-mpca-CIFAR10-components.png}
  \caption{Above are the first 12 components produced by MPCA over the
    CIFAR10 general image classification data set. These images are
    normalized to unit 2-norm before visualization. Notice that the
    components strongly resemble the early terms in a Fourier
    decomposition, suggesting that low frequency spatial oscillations
    contain the most information in images, followed by more varied
    spatial patterns.}
  \label{fig:cifar10}
  %% {{https://people.cs.vt.edu/tchlux/files/[2018-11-28]_CIFAR10_12_MPCA_Components.png}}
\end{figure}

The first 12 components produced by MPCA for the image classification
problems can be seen in Figures \ref{fig:mnist} and
\ref{fig:cifar10}. Notice how the first components are drastically
different even though both are computer vision classification tasks.
This is expected, because recognizing the differences between
handwritten digits and generic objects (with 3-dimensional
orientations) are drastically different problems.

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{3-prediction-errors.pdf}
  \caption{The prediction errors of KNN are computed via 4-fold cross
    validation. From left to right, the three bars within each group
    show the performance of KNN with $\{1/100$, $1/10$, $1/3\}$ of all
    features respectively. In general, MPCA and PCA perform similarly
    for all three prediction tasks.}
  \label{fig:errors}
  %% {{https://people.cs.vt.edu/tchlux/files/[2018-11-28]_tgds_errors_hist.html}}
\end{figure}

The prediction errors with varying amounts of dimension reduction by
MPCA and PCA are presented in Figure \ref{fig:errors}. The results
show that common real approximation problems seem to have data
variance that aligns well with the underlying phenomenon being
modeled. This result is somewhat expected, as data sets are often
collected in a way that assumes each piece of information is
specifically relevant to the underlying phenomenon being approximated.

MPCA is used with varying levels of dimension reduction to improve KNN
predictions in Table \ref{tab:mpca-results}. For each approximation
problem, a different amount of reduction produces the best
approximation accuracy. The Yelp data requires the smallest subspace,
MNIST is near the middle, and the CIFAR10 data requires the largest
subspace that is not the raw images.

Finally, results combining all algorithms, data sets, and both PCA and
MPCA are presented in Table \ref{tab:all-results}. Notice that the
empirical results demonstrate that MPCA is not guaranteed to be the
best reduction technique even for strictly distance based prediction
techniques. On the \textit{Yelp} data, PCA still produces better
results for 10-nearest neighbors. It is suspected that the worse
performance of MPCA is due to the reduction on $Z$ and the increased
variability introduced by the random sampling and metric scaling.

These results should not be considered exhaustive, nor should these
results devalue MPCA accordingly. In almost all cases the differences
between the best PCA and MPCA predictors are less than $1\%.$ Aside
from performing similarly, MPCA has the added benefit of accommodating
less careful data collection procedures.

\begin{table}
  \centering
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{c c c c c}
    \hline
    \textbf{Data Name} & \textbf{Raw data}
      & \multiline{\textbf{1/3 MPCA}\\[-3mm]\textbf{components}}
      & \multiline{\textbf{1/10 MPCA}\\[-3mm] \textbf{components}}
      & \multiline{\textbf{1/100 MPCA}\\[-3mm] \textbf{components}} \\
    \hline \hline
    \multiline{\textit{Yelp}\\[-3mm]64 dim}
      & 0.493 (stars) & \underline{0.493 (stars)} & 0.496 (stars) & 0.530 (stars) \\
    \multiline{\textit{MNIST}\\[-3mm]784 dim}
      & 5.29\% & 5.26\% & \underline{4.63\%} & 13.43\% \\
    \multiline{\textit{CIFAR10}\\[-3mm]3072 dim}
      & 70.99\% & 70.23\% & 69.46\% & \underline{62.46\%} \\
    \hline
  \end{tabular}
  \vspace{5mm}
  \caption{This table shows the prediction accuracy of a KNN model
    when provided the raw data, and varying size subspaces from MPCA
    for the three real approximation problems. In each problem, the
    optimal size subspace varies. For Yelp, the smallest size subspace
    is best, while for CIFAR10, the largest subspace is best.}
  \label{tab:mpca-results}
\end{table}

\section{Discussion}
\label{sec:discussion}

The error results for reduced-dimension problems using PCA and Metric
PCA (MPCA) are remarkably similar. In almost all cases the MPCA
reduction causes an improvement in prediction accuracy for nearest
neighbor predictions, though the improvement is small. Although an
analytic example demonstrates the potentially large difference between
MPCA and PCA, in practice the two appear to often be nearly
equivalent. One could speculate that this is a contrived phenomenon,
the only data that is kept in curated sets is data that represents a
phenomenon of interest. MPCA is constructed to disregard data that is
not useful, but for the data sets chosen in this work the underlying
data is already distributed according to the functions being modeled.

\begin{table}
  \centering
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{c c c c c}
    \hline
    \textbf{Data Set} & \textbf{Approximator}
      & \multiline{\textbf{Reduction}\\[-1mm]\textbf{Technique}}
      & \multiline{\textbf{Reduced}\\[-1mm]\textbf{Dimension}}
      & \multiline{\textbf{Lowest}\\[-1mm]\textbf{Mean}\\[-1mm]\textbf{Error}} \\
    \hline \hline
    \multirow{5}{*}{\textit{Yelp}}
    & Delaunay & PCA & 21 & 0.512 \\
    & KNN (1) & MPCA & 6 & 0.610 \\
    & KNN (10) & PCA & 6 & \underline{0.493} \\
    & LSHEP & raw & 63 & 0.678 \\
    & MLPRegressor & PCA & 1 & 0.515 \\
    \hline
    \multirow{5}{*}{\textit{MNIST}}
    & Delaunay & PCA & 8 & 0.132 \\
    & KNN (1) & MPCA & 78 & \underline{0.046} \\
    & KNN (10) & MPCA & 78 & 0.051 \\
    & LSHEP & PCA & 8 & 0.139 \\
    & MLPRegressor & PCA & 261 & 0.133 \\
    \hline
    \multirow{5}{*}{\textit{CIFAR10}}
    & Delaunay & PCA & 31 & \underline{0.631} \\
    & KNN (1) & MPCA & 31 & 0.681 \\
    & KNN (10) & MPCA & 31 & 0.643 \\
    & LSHEP & PCA & 31 & 0.650 \\
    & MLPRegressor & raw & 3072 & 0.897 \\
    \hline
  \end{tabular}
  \vspace{5mm}
  \caption{In this expanded set of results, the optimal reduction
    technique and subspace size is presented for each algorithm and
    for each real approximation problem. Notice that MPCA is not the
    most common \textit{best} reduction technique, which is
    counterintuitive. This suggests that the data provided for each
    approximation problem is in fact representative of the underlying
    phenomenon being modeled. It also suggests that the noise
    introduced in randomly sampling pairwise vectors for MPCA may
    inhibit its performance on approximation problems that are
    well-represented by available data.}
  \label{tab:all-results}
\end{table}

\section{Conclusion}
\label{sec:conclusion}

The proposed technique, Metric PCA, demonstrates strong potential as a
dimension reduction strategy for approximation problems. Analytically,
Metric PCA is not susceptible to adverse data conditions that may
cause PCA to disregard important dimensions for
approximation. Empirically, the results obtained by MPCA and PCA are
quite similar, suggesting that many curated data sets have the
property that data variance corresponds closely to variance
in the underlying function. Analytic and empirical results combined
demonstrate the effectiveness of MPCA as a robust dimension reduction
strategy for approximation.

%% \begin{figure}[htb]
%%   \centering
%%   \includegraphics[width=0.50\textwidth]{demo_projection}
%%   \caption{These are the projection mechanisms that make a cubic polynomial piece monotone.}\label{fig:projection}
%% \end{figure}

% Please don't change the bibliographystyle style
\bibliographystyle{scsproc}
% AUTHOR: Include your bib file here
\bibliography{paper}

\section*{Author Biographies}

\textbf{\uppercase{THOMAS C. H. LUX}} is a Ph.D. candidate in computer
science at Virginia Polytechnic Institute and State University
(VPI\&SU) studying under Dr. Layne Watson. His research interests
include computational science, approximation theory, optimization, and
artificial intelligence. His email address is \email{tchlux@vt.edu}.

\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. candidate at VPI\&SU
studying computer science under Dr. Layne Watson.

\end{document}

### 2020-01-16 [2019-12-16]_HPC_(mpca)/code/util/math/weakly_admissible_meshes.py

@@ -1,0 +1,31 @@
+import numpy as np
+
+# Given "n" points and "d" dimensions, generate a weakly admissible
+# mesh. WARNING: Forced to dimension = 2.
+def polar_wam(n, dimension=2):
+    # Initialize points only containing the center.
+    points = [[0,0]]
+    for length in range(1,n):
+        l = 1/2 * (1 + np.cos((length * np.pi) / n))
+        for angle in range(0,2*n+1):
+            a = (2 * np.pi * angle) / (2 * n + 1)
+            # Convert the length and angle into a 2D point.
+            x = l * np.cos(a)
+            y = l * np.sin(a)
+            points.append( [x,y] )
+    # Convet the points into a row-vector matrix and shift into the unit box.
+    return (np.array(points) + 1) / 2
+
+# Given "n" points and "d" dimensions, generate a weakly admissible
+# mesh. WARNING: Forced to dimension = 2.
+def box_wam(n, dimension=2):
+    # Initialize points only containing the center.
+    points = []
+    for length_1 in range(0,n+1):
+        l1 = 1/2 * (1 + np.cos((length_1 * np.pi) / n))
+        for length_2 in range(0,n+1):
+            l2 = 1/2 * (1 + np.cos((length_2 * np.pi) / n))
+            # Convert the length and angle into a 2D point.
+            points.append( [l1,l2] )
+    # Convet the points into a row-vector matrix.
+    return np.array(points)

### 2020-01-16 [2019-12-16]_HPC_(mpca)/code/util/stats/__init__.py

@@ -1,0 +1,35 @@
+# Import all of the submodules so that users can access all from "util.stats".
+from util.stats.rank          import *
+from util.stats.ks            import *
+from util.stats.difference    import *
+from util.stats.distributions import *
+from util.stats.samples       import *
+from util.stats.modes         import *
+from util.stats.metric_pca    import *
+from util.stats.plotting      import *
+
+# Backwards compatibility with warning for deprecation.
+
+def cdf_fit_func(*args, **kwargs):
+    print("\nWARNING: 'cdf_fit_func' is a deprecated function. Use 'cdf_fit' instead.\n")
+    return cdf_fit(*args, **kwargs)
+
+def pdf_fit_func(*args, **kwargs):
+    print("\nWARNING: 'pdf_fit_func' is a deprecated function. Use 'pdf_fit' instead.\n")
+    return pdf_fit(*args, **kwargs)
+
+
+# ../../development/testing/test_stats.py 
+if __name__ == "__main__":
+    import numpy as np
+    from util.random import cdf
+    np.random.seed(1)
+    data = cdf()
+    
+    from util.random import cdf
+    np.random.seed(1)
+    data = cdf(nodes=3).inverse(np.random.random(100))
+    # modes(data)
+
+    # Look for "largest" regions of separation 
+

### 2020-01-16 [2019-12-16]_HPC_(mpca)/code/util/stats/metric_pca.py

@@ -1,0 +1,95 @@
+import numpy as np
+from util.math import abs_diff
+
+# =============================================
+#      Metric Principle Component Analysis     
+# =============================================
+
+# Generate vector between scaled by metric difference. Give the metric
+# the indices of "vectors" in the provided matrix.
+def gen_random_metric_diff(matrix, index_metric, power=2, count=None):
+    from util.random import pairs
+    # Iterate over random pairs, skip those with no difference.
+    for (p1, p2) in pairs(len(matrix), count):
+        metric_diff = index_metric(p1, p2)
+        if (metric_diff <= 0): continue
+        vec = matrix[p1] - matrix[p2]
+        length = np.linalg.norm(vec)**power
+        if length > 0: yield metric_diff * vec / length
+
+# Given a set of row-vectors, compute a convex weighting that is
+# proportional to the inverse total variation of metric distance
+# between adjacent points along each vector (component).
+def normalize_error(points, values, metric, display=True):
+    # Compute the magnitudes using total variation.
+    if display: print(" estimating error slope per axis.. ", end="\r", flush=True)
+    avg_slope = np.zeros(points.shape[1])
+    update = end = ""
+    for axis in range(points.shape[1]):
+        update = "\b"*len(end)
+        end = f"{axis+1} of {points.shape[1]}"
+        if display: print(update, end=end, flush=True)
+        # Sort points according to this axis.
+        ordering = np.argsort(points[:,axis])
+        for i in range(len(ordering)-1):
+            p1, p2 = ordering[i], ordering[i+1]
+            diff = (points[p2,axis] - points[p1,axis])
+            if (diff > 0): avg_slope[axis] += metric(values[p1], values[p2]) / (diff * points.shape[1])
+    if display: print("\r                                   ", end="\r", flush=True)
+    # If there are dimensions with no slope, then they are the only ones needed.
+    if (min(avg_slope) <= 0.0): avg_slope = np.where(avg_slope == 0, 1., float('inf'))
+    # We want to minimize the expected error associated with the 2-norm.
+    # 
+    # E[f(x) - f(y)] = sum( E[f(x)1 - f(y)1]^2 + ... + E[f(x)d - f(y)d]^2 )^(1/2)
+    #               -> { E[f(x)1-f(y)1]^(-2), ..., E[f(x)d-f(y)d]^(-2) }
+    # 
+    # This is the normalizing vector we want, scaled to unit determinant.
[…]

### 2020-01-16 [2019-12]_IEEESE_(quintic_applied)/code/monotone.py
Created paper repository with all of my work.

@@ -1,0 +1,394 @@
+# This file provides the function `monotone_quintic_spline` which
+# constructs spline interpolant over provided data. It starts by using
+# divided differences to construct approximations for the first and
+# second derivative at each of the data points. Then it iteratively
+# adjusts the derivatives and second derivatives until all quintic
+# polynomial pieces are monotone. A `Spline` object is returned.
+
+from polynomial import fit
+IS_MONOTONE_CALLS = 0
+
+# Given x and y values, construct a monotonic quintic spline fit.
+#  shrink_rate = .9    -> 120 steps max
+#              = .99   -> 386 steps max
+#              = .999  -> 1221 steps max
+#              = .9999 -> 3861 steps max
+def monotone_quintic_spline(x, y=None, values=None, ends=2, mids=2, exact=True, 
+                            max_steps=100, verbose=False, **kwargs):
+    if verbose: print()
+    from polynomial import Spline
+    # Convert 'x' to exact values.
+    if exact:
+        from fraction import Fraction
+        x = list(map(Fraction, x))
+    # Build a set of 'values' if they were not provided.
+    if (y is not None):
+        if exact: y = list(map(Fraction, y))
+        # Check for which type of monotonicity this function maintains.
+        if   all(y[i+1] >= y[i] for i in range(len(y)-1)): kwargs.update(dict(min_d1=0))
+        elif all(y[i] >= y[i+1] for i in range(len(y)-1)): kwargs.update(dict(max_d1=0))
+        else: raise(Exception("Provided 'y' data is not monotone."))
+        # Pass in the method of fitting end and midpoints as keyword arguments.
+        kwargs.update(dict(mids=mids, ends=ends))
+        # Construct an initial fit that is twice continuous.
+        f = fit(x, y, continuity=2, **kwargs)
+        values = f.values
+        # Make the first derivative 0 at the right end.
+        values[-1][1] = 0
+        # Make the second derivatives 0 at the ends.
+        values[0][2] = values[-1][2] = 0
+    elif (values is not None):
+        if exact: values = [list(map(Fraction, row)) for row in values]
+        # Verify monotonicity.
+        if   all(values[i+1][0] >= values[i][0] for i in range(len(values)-1)): pass
[…]

### 2020-01-16 [2019-12]_IEEESE_(quintic_applied)/code/polynomial.py

@@ -1,0 +1,717 @@
+# This file provides utilities for constructing polynomial interpolants.
+# 
+# The following objects are provided:
+# 
+#   Spline           -- A piecewise polynomial interpolant with
+#                       evaluation, derivative, and a string method.
+#   Polynomial       -- A monomial with stored coefficients,
+#                       evaluation, derivative, and a string method.
+#   NewtonPolynomial -- A Newton polynomial with stored coefficients,
+#                       points, evaluation, derivative, and a string method.
+# 
+# The following functions are provided:
+# 
+#   polynomial       -- Given x and y values, this produces a minimum
+#                       degree Newton polynomial that interpolates the
+#                       provided points (this is a *single* polynomial).
+#   polynomial_piece -- Given a function value and derivatives,
+#                       produce a polynomial that interpolates these
+#                       values at the endpoints of an interval.
+#   fit              -- Use 'fill_derivative' to construct an interpolating
+#                       Spline with a specified level of continuity.
+#   fill_derivative  -- Compute all derivatives at points to be
+#                       reasonable values using either a linear or a
+#                       quadratic fit over neighboring points. 
+#   solve_quadratic  -- Given three points, this will solve the equation
+#                       for the quadratic function which interpolates
+#                       all 3 values. Returns coefficient 3-tuple. 
+# 
+
+from fraction import Fraction
+
+# This class-method wrapper function ensures that the class method
+# recieves a fraction and returns a "float" if a fraction was not
+# provided as input (ensuring internal methods only recieve fractions).
+def float_fallback(class_method):
+    def wrapped_method(obj, x):
+        # Check for vector usage.
+        try: return [wrapped_method(obj, v) for v in x]
+        except:
+            # Return perfect precision if that was provided.
+            if (type(x) == Fraction): return class_method(obj, x)
+            # Otherwise, use exact arithmetic internally and return as float.
+            else: return float(class_method(obj, Fraction(x)))
+    return wrapped_method
+
+# A piecewise polynomial function that supports evaluation,
+# differentiation, and stitches together an arbitrary sequence of
+# function values (and any number of derivatives) at data points
+# (knots). Provides evaluation, derivation, and string methods.
[…]

### 2020-01-16 [2019-12]_IEEESE_(quintic_applied)/paper.tex
Created paper repository with all of my work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8

\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
                                                          % if you need a4paper
%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
\IEEEoverridecommandlockouts                              % This command is only needed 
%% % copyright info
%% \IEEEpubid{\makebox[\columnwidth]{978-1-5386-6133-8/18/\$31.00
%%         ~\copyright~2018 IEEE \hfill}\hspace{\columnsep}
%%         \makebox[\columnwidth]{ }}

% if you want to use the \thanks command
\overrideIEEEmargins
% See the \addtolength command later in the file to balance the column lengths on the last page of the document

% The following packages can be found on http:\\www.ctan.org
\usepackage{tikz}     % For adding axes to plots
\usepackage{graphicx} % For including PDFs
\usetikzlibrary{positioning}
\usepackage{balance} % For balancing the last column.

%% \usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\title{\LARGE \bf Effective Nonparametric Distribution Modeling \\ for Distribution Approximation Applications }

%% \author{Thomas C. H. Lux$^{1}$, Layne T. Watson$^{2,5}$, Tyler H. Chang$^{1}$, Li Xu$^{3}$, Yueyao Wang$^{3}$,\\ Jon Bernard$^{1}$, Yili Hong$^{4}$, Kirk W. Cameron$^{2}$% <-this % stops a space
%% \thanks{*This work was supported by the National Science Foundation Grant CNS-1565314.}%
%% \thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
%% \thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
%% \thanks{$^{3}$Doctoral candidate, Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
%% \thanks{$^{4}$Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
%% \thanks{$^{5}$Departments of Mathematics and Aerospace Engineering, Virginia Polytechnic Institute and State University, Virginia 24060, USA}%
%% }

\begin{document}

\maketitle
\thispagestyle{empty}
\pagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}

Many fields of science rely on the collection of samples and estimation of true population distributions from those samples. There are several effective nonparametric methods for approximating a true distribution from empirical data, however it is unclear which methods produce the best approximations in practice. This work presents a case study on the effectiveness of various distribution approximations. Results show that piecewise linear approximations produce the smallest maximum absolute error, while the classic empirical distribution function (EDF) produces the smallest median absolute error as well as the smallest absolute error at lower percentiles when approximating a distribution from a sample. When building distribution prediction models, the piecewise quintic and cubic approximations produce the lowest absolute error at most error percentiles. These initial results encourage more research on the best methods of fitting empirical data with smooth functions to generate accurate distribution approximations.

\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{INTRODUCTION}
\label{sec:introduction}

Empirical samples play a pivotal role in science. Experiments are run, data is recorded, and that data is used to draw conclusions about the \textit{truth}. When an experiment is run many times with varying outcomes, it is common to describe the truth as a random variable. In this work continuous (numeric) outcome random variables are considered. In this context, a random variable $X$ is precisely defined by its absolutely continuous \textit{cumulative distribution function} (CDF) $F_X$ and the derivative of the CDF, the \textit{probability density function} (PDF) $f_X$. As samples are drawn from $X$, the value of the CDF can be estimated at the sample points by measuring the probability at which other samples are less than or equal to that value. In this sense empirical data defines an empirical distribution function (EDF). The points that define the EDF are used to construct approximations of the true underlying CDF.

An approximation of a CDF can have varying levels of smoothness and, for empirical purposes the approximations are constructed over a closed interval. Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, and most importantly to this work they can provide accurate nonparametric approximations in statistics \cite{knott2012interpolating}. While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.

\begin{figure}
  \vspace{-.3cm}
  \includegraphics[width=.5\textwidth]{fl-sample-prediction.pdf}
  \caption{A demonstration of the linear fit and the classic EDF over a sample of 10 points. This data is normalized to be in the range $[0,1]$ and the right half of the domain is cropped out of this plot. Notice that the EDF points do not necessarily (and often do not) exactly equal the true CDF at any given position. The sampling inherently introduces variance into measurements of the CDF. This makes producing an accurate approximation difficult, as will be seen.
  \vspace{-.5cm}}
  \label{fig:sample-prediction}
\end{figure}

In this statistical work, the construction of a monotone interpolating spline that is continuous in its derivatives could be meaningfully useful \cite{ramsay1988monotone}. A function with $C^1$, and especially $C^2$ continuity could approximate a cumulative distribution function to a high level of accuracy with relatively few intervals. A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a desirable property many standard parametric distributions maintain.

There is publicly available software for monotone piecewise polynomial interpolation, including quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases. Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}. Recently an algorithm for the construction of monotone quintic splines has been produced as well (citation pending).%% \cite{lux2020algorithm}
This work considers the EDF, a linear interpolant, a monotone cubic spline interpolant, and a monotone quintic spline interpolant as candidate approximations.

In the next section, the methodology of this work is outlined and experimental setup is detailed. In Section \ref{sec:results}, three experiments related to the approximation of cumulative distribution functions are presented and analyzed. Finally Section \ref{sec:conclusion} concludes.

\begin{figure}
  \vspace{-.3cm}
  \includegraphics[width=.5\textwidth]{cq-sample-prediction.pdf}
  \caption{A demonstration of the monotone cubic and quintic spline interpolants over a sample of 10 points. This data is normalized to be in the range $[0,1]$ and the right half of the domain is cropped out of this plot. Notice that the difference between EDF points and the true CDF reduce the accuracy of first derivative (for both) and second derivative (for quintic) estimates in the interpolants, magnifying the error in the empirical estimations of CDF values.
  %% \vspace{-.5cm}
  }
  \label{fig:sample-prediction}
\end{figure}

%% ===================================================================
\section{METHODOLOGY AND DATA}
\label{sec:methodology}

In order to identify the best performing approximation techniques for distribution estimation, a case study on real-world data is presented and three experiments are run on that data to test various fits. First the accuracy of distribution approximations with varying sample sizes is studied, then the likelihood of any technique having the smallest maximum error is estimated, and finally the approximations are used to make predictions in a distribution prediction application similar to \cite{lux2018nonparametric}. Each experiment presents a unique, albeit with limited data, perspective on the quality of approximation provided by the four distribution approximation techniques.

The CDF modeling case study is constructed from a four-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers. Each node contains two Intel Xeon E5-2637 CPUs offering a total of 16 CPU cores with 16GB of DRAM. While the CPU frequency varies depending on the test configuration, the I/O from IOzone is performed by an ext4 filesystem sitting above an Intel SSDSC2BA20 SSD drive. At the time of data collection, Linux kernel Version 4.13.0 was used. The system performance data was collected over two months by executing IOzone  $10000$ times for each of a select set of $17$ system configurations, for a total of $170$ thousand executions of IOzone. A single IOzone execution reports the max I/O throughput in bytes per second seen for the selected test type. For this case study, only the results of a ``readers'' test are considered, where bytes are sequentially read from the SSD. The summary of the data for the experiments for this paper can be seen in Table \ref{tab:data_description}. Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.

The performance of approximation techniques that predict probability functions can be analyzed through a variety of summary statistics. The first two experiments study the distribution of absolute differences between approximated CDFs and the true CDFs. This distribution over many trials gives an idea of the \textit{expected} error. The last experiment in this work analyzes the max absolute difference between approximated and true CDFs, also known as the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov} for its compatibility with the KS test.

The two-sample KS test is a useful nonparametric test for comparing two CDFs while only assuming stationarity, finite mean, and finite variance. The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test presented in this work $n_1 = n_2$.

Finally an example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction-example}. The Delaunay triangulation is a well-studied geometric technique for producing an interpolant \cite{lee1980two}. The Delaunay triangulation of a set of data points into simplices is such that there are no data points inside the sphere defined by the vertices of each simplex. For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and functions $F_{x^{(i)}}$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique convex combination of the vertices,
$$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
and the Delaunay interpolant $F_y$ at $y$ is given by
$$ F_y = \sum_{i=0}^{d} w_i F_{x^{(i)}}. $$
In the case of these experiments, the file size, record size, thread count, and CPU frequency are first normalized to the unit hypercube. After normalization, Delaunay predictions are made via a leave-one-out method resulting in throughput data from five known configurations being used to predict one unknown throughput distribution. The Delaunay method is used to select the five source distributions as well as provide the convex weights associated with each of the source distributions that will most likely produce the predicted distribution.

\begin{figure}
  \vspace{-.3cm}
  \includegraphics[width=.5\textwidth]{throughput-histogram.pdf}
  \caption{A histogram of all throughput values across the 17 unique system configurations and 170 thousand executions of a ``readers'' test in IOzone.}
  \label{fig:throughput_histogram}
\end{figure}

\begin{table}
  \centering
  \begin{tabular}{c|c}
    \hline
    \textbf{System Parameter} & \textbf{Values}\\
    \hline
    File Size (KB) & 4, 64, 256, 1024, 8192, 16384, 32768, 65536\\
    \hline
    Record Size (KB) & 4, 8, 16, 32, 64, 128, 4096, 8192, 16384\\
    \hline
    Thread Count & 8, 16, 24, 32, 40, 48, 56, 64\\
    \hline
    Frequency (GHz) & 1.2, 1.6, 2, 2.3, 2.8, 3.2, 3.5\\
    \hline
  \end{tabular}
  \caption{A description of system parameters considered for IOzone. Record size must be less than or equal to file size during execution. In all, $10$ thousand repeated trials are run at $17$ unique system configurations.
    \vspace{-.5cm}}
  \label{tab:data_description}
\end{table}

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this example, the general methodology for predicting a CDF and evaluating error can be seen. The Delaunay method chose three source distributions (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom at arrow). The weighted sum of the three known CDFs produces the predicted CDF (dashed line). The KS Statistic (arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example. The KS test null hypothesis is rejected by $p$-value 0.01, however it is not rejected by $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

\section{RESULTS}
\label{sec:results}

All experiments that follow will consider the differences between CDF approximations given by the EDF, a linear interpolant, a monotone cubic interpolant, and a monotone quintic interpolant. In the first experiment, all four methods are used to approximate the ``true'' distributions of the ten thousand throughput values at each system configuration. One hundred random collections of $k$ samples are drawn from each true distribution for $k = 10$, $50$, $200$ and all methods are used to approximate the true distribution from each sample. In Figure \ref{fig:abs-error-dist}, the distributions of absolute difference between the ``true'' CDF and each approximated CDF at $1000$ equally spaced percentiles is shown. The sample of $50$ produces almost exactly half the errors observed with $10$ samples, and the error is roughly halved again when increasing to $200$ samples. Aside from the reduction in accuracy caused by sample size, most approximations appear to be nearly identical. When only $10$ samples are observed, the EDF has slightly less absolute error than other techniques between the median and third quartile. This difference is very minor, and difficult to observe.

\begin{figure}
  \vspace{-.3cm}
  \includegraphics[width=.45\textwidth]{abs-errors-10-samples.pdf}
  \includegraphics[width=.45\textwidth]{abs-errors-50-samples.pdf}
  \includegraphics[width=.45\textwidth]{abs-errors-200-samples.pdf}
  \caption{The distribution of absolute errors with $10$ (top), $50$ (middle), and $200$ (bottom) samples for each approximation method. Notice that the EDF performs slightly better than other methods (more small absolute errors) when there are only $10$ samples. Given $50$ or more samples, all the approximations produced nearly identical errors.
  \vspace{-.5cm}}
  \label{fig:abs-error-dist}
\end{figure}

The second experiment considers the KS statistic (maximum absolute error) rather than the aggregate of errors. An increased collection of sample sizes is considered, with $k =$ 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 150, 200, 250, 300, 400, 500. For all $k$, 100 random samples of $k$ throughput values are drawn from the true distributions and approximate CDFs are constructed via each of the four methods. The error of each approximate CDF is measured at 1000 equally spaced percentiles and the maximum absolute error is recorded. At each sample size, Figure \ref{fig:ks-rank-0} depicts the probability that any given approximation method produces a distribution estimate with the lowest KS statistic. In this measure of error, the piecewise quintic and cubic approximations to the CDF provide the most accurate estimate of the true CDF. This result reveals that the chosen measure of error is important in determining which CDF approximation method is most suited to an applications. The EDF provides the lowest expected absolute error, however, the more smooth approximations appear to provide lower maximum absolute error.

\begin{figure*}
  \vspace{-.3cm}
  \includegraphics[width=1\textwidth]{KS-rank-0-probability.pdf}
  \caption{The probability that any one of the distribution approximation techniques has the lowest KS statistic among all the techniques when given a varying number of samples. Notice that the linear approximation remains the most likely to have the smallest KS statistic for all sample sizes, closely followed by the cubic.
  \vspace{-.5cm}}
  \label{fig:ks-rank-0}
\end{figure*}

The third experiment considers a distribution prediction application. In this case, the Delaunay method was applied to choose five source distributions and convex weights that will predict the throughput distribution at the system configuration with a 2.8GHz CPU frequency, 16 megabyte (MB) file size, 16MB record size, and 64 threads. This configuration is predicted because it is the only interpolation point (configuration inside or on the convex hull of other configurations) among the 17 available system configurations. An example of the prediction methodology can be seen in Figure \ref{fig:prediction-example}. The distribution of absolute errors measured at 1000 equally spaced percentiles of the true CDF for all four approximation methods can be seen in Figure \ref{fig:delaunay-errors}. In this case the piecewise quintic and cubic approximations provide the best predictions overall. The quintic and cubic methods produce smaller errors than the piecewise linear and EDF approximations at more than 90\% of measurements.

\begin{figure}
  \vspace{-.3cm}
  \includegraphics[width=.5\textwidth]{delaunay-predictions.pdf}
  %% \includegraphics[width=.5\textwidth]{abs-errors-100-samples.pdf}
  \caption{The result of using Delaunay to predict distributions. Notice that the best performing, lowest error (to the left), approximation is the EDF. Interestingly, the piecewise linear fit does not produce results as accurate as the EDF. This likely has to do with the inherent smoothing that happens when multiple different EDFs are combined to make one prediction.
  \vspace{-.5cm}}
  \label{fig:delaunay-errors}
\end{figure}

The three experiments presented in this section have each tested a unique facet of distribution approximation. The first experiment analyzes the approximation of a distribution from a sample when measuring the aggregate absolute CDF approximation error; in this case the EDF largely produces the best approximations. The second experiment analyzes the maximum absolute error when approximating a CDF from a sample; in this case the monotone quintic and cubic spline interpolants are most likely to provide the best approximation. The third and final experiment analyzes the aggregate absolute error when predicting unobserved distributions, for which the monotone quintic and cubic splines produced the best overall CDF approximations.

%% \section{DISCUSSION}
%% \label{sec:discussion}

%% Discussion text..

\section{CONCLUSION}
\label{sec:conclusion}

This preliminary case study on the best functions for approximating cumulative distribution functions from empirical data has given some actionable insights. Specifically, the choice of error measure is important in deciding which distribution approximation method to apply to a sample. The standard empirical distribution function has the lowest expected absolute error when approximating a CDF given a sample of data. However, more smooth estimates like monotone quintic splines produce lower expected KS statistics (maximum absolute error), and are expected to have lower error in distribution prediction applications.

Future research may extend this case study to other data sets and include a more exhaustive distribution prediction test suite (experiment three). Tangentially, future work may target improved estimates of the first and second derivatives of the true CDF given a sample of data in order to reduce the approximation error of the monotone quintic and cubic spline methods.

\balance

\bibliographystyle{IEEEtran}
\bibliography{paper}

%% \newpage

%% \begin{figure}
%%   \includegraphics[width=.5\textwidth]{plot-histogram-throughput.pdf}
%%   \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations. The distribution is skewed right, with few tests having significantly higher throughput than most others.
%%   \vspace{-.5cm}}
%%   \label{fig:throughput_histogram}
%% \end{figure}

%% This paper presents a variability modeling case study with a five-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers. Each node contains two Intel Xeon E5-2637 CPUs offering a total of 16 CPU cores with 16GB of DRAM. While the CPU frequency varies depending on the test configuration, the I/O from IOzone is performed by an ext4 filesystem sitting above an Intel SSDSC2BA20 SSD drive. At the time of data collection, Linux kernel Version 4.13.0 was used. The system performance data was collected over two weeks by executing IOzone 150 times for each of a select set of approximately 18K system configurations, for a total of approximately 2.7M executions of IOzone. A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test type. The summary of the data components in $x^{(i)}$ for the experiments for this paper can be seen in Table \ref{tab:data_description}. Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.

%% Some mild preprocessing was necessary to prepare the data for modeling and analysis. All features were shifted by their minimum value and scaled by their range, mapping each feature independently into $[0,1]$. This normalization ensures each feature is treated equally by the interpolation techniques and should be performed on all data before building models and making predictions regardless of application. All 150 repeated trials for a system configuration were grouped with that configuration. The only nonordinal feature in this data is the test type. All test types were treated as different applications and were separated for modeling and analysis, i.e., predictions for the ``readers'' test type were made using only known configurations for the ``readers'' test type.

%% %% ===================================================================

%% \begin{figure}
%%   \begin{tikzpicture}
%%     \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{plot-histogram-80_20-KS.pdf}};
%%     \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
%%     \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of Predictions with Given KS Statistic};
%%   \end{tikzpicture}
%%   \caption{Histograms of the prediction error for each modeling algorithm from ten random splits when trained with 80\% of the data aggregated over all different test types. The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution. The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.
%%   \vspace{-.1cm}}
%%   \label{fig:ks_histogram_80_20}
%% \end{figure}

%% \begin{table}
%%   \centering
%%   \begin{tabular}{c|c|c}
%%     \hline
%%     \textbf{Algorithm} & \textbf{$P$-Value} & \textbf{\% N.H. Rejections} \\
%%     \hline
%%     \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{58.4\\69.3\\61.9}\\
%%     \hline
%%     \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{51.1\\58.4\\53.4}\\
%%     \hline
%%     \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{44.1\\46.9\\45.1}\\
%%     \hline
%%     \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{31.4\\26.6\\28.7}\\
%%     \hline
%%   \end{tabular}
%%   \caption{Percent of null hypothesis rejections rate by the KS-test when provided different selections of $p$-values. These accompany the percent of null hypothesis rejection results from Figure \ref{fig:ks_histogram_80_20}. 
%%     \vspace{-.5cm}}
%%   \label{tab:p_value_failure_rate}
%% \end{table}

%% \begin{figure}
%%   \begin{tikzpicture}
%%     \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-Failure-by-Training.pdf}};
%%     \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
%%     \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
%%   \end{tikzpicture}
%%   \caption{The performance of each algorithm on the KS test ($p=0.001$) with increasing amounts of training data averaged over all IOzone test types and ten random splits of the data. The training percentages range from 5\% to 95\% in increments of 5\%. Delaunay is the best performer until 95\% of data is used for training, at which Max Box mesh becomes the best performer by a fraction of a percent.
%%   \vspace{-.1cm}}
%%   \label{fig:ks_failure_by_training}
%% \end{figure}

%% \begin{figure}
%%   \begin{tikzpicture}
%%     \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-failure-by-training-and-test.pdf}};
%%     \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
%%     \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
%%   \end{tikzpicture}
%%   \caption{The percentage of null hypothesis rejections for predictions made by each algorithm on the KS test ($p=0.001$) over different IOzone test types with increasing amounts of training data. Each percentage of null hypothesis rejections is an average over ten random splits of the data. The training percentages range from 5\% to 95\% in increments of 5\%. The read test types tend to allow lower rejection rates than the write test types.
%%   \vspace{-.1cm}}
%%   \label{fig:ks_failure_by_training_and_test}
%% \end{figure}

\end{document}

### 2020-01-16 [2019-12]_IEEESE_(quintic_applied)/code/0-clean-data.py

@@ -1,0 +1,59 @@
+import os
+from util.data import Data
+
+# --------------------------------------------------------------------
+
+# STEP 1
+sample = None
+prefix = '[0]-' if sample is None else '[0]-sample-'
+test = "readers" # "initial-writers" # "readers"
+
+# The conditions on taking each of the steps.
+step_1 = not os.path.exists(prefix+"mu-reduced.csv.gz")
+step_2 = not os.path.exists(prefix+test+".csv.gz")
+step_3 = not os.path.exists(prefix+test+".csv.pkl")
+
+print()
+print("Step 1..")
+if step_1:
+    print(" loading 'mu.csv.gz' and reducing..")
+    data = Data.load("mu.csv.gz", sample=sample)
+    to_keep = ['', 'host', 'media', 'freq', 'fsize', 'rsize', 'threads', 'iter', 'subiter', 'test', 'throughput']
+    data = data[to_keep]
+    data.save(prefix+"mu-reduced.csv.gz")
+elif (step_2 or step_3):
+    print(" loading 'mu-reduced.csv.gz'..")
+    data = Data.load(prefix+"mu-reduced.csv.gz", sample=sample)
+print(" done.")
+
+# STEP 2
+print()
+print("Step 2..")
+if step_2:
+    print(f" reducing to '{test}' test..")
+    full_data = data
+    # Pick which columns to remove.
+    to_remove = ['', 'host', 'media',  'iter', 'subiter', 'test']
+    to_keep = [n for n in data.names if n not in to_remove]
+    # reduce to the specific test, and throw out the columns I don't want
+    data = data[data["test"] == test][to_keep].copy()
+    del(full_data)
+    print(data)
+    # Stack up the throughput values.
+    data.stack('throughput')
+    data.inflate('throughput')
+    print(data)
+    data.save(prefix+test+".csv.gz")
+elif step_3:
+    data = Data.load(prefix+test+".csv.gz", verbose=False, sample=sample)
+print(" done.")
+
+# STEP 3
+print()
+print("Step 3..")
+if step_3:
+    print(f" saving a '.pkl' version for fast loading..")
+    data.save(prefix+test+".pkl")
+print(" done.")
+
+# --------------------------------------------------------------------

### 2020-01-16 [2019-12]_IEEESE_(quintic_applied)/code/1-run-experiments.py

@@ -1,0 +1,174 @@
+MAKE_HISTOGRAM_PLOT = False
+MAKE_SAMPLE_PREDICTION_FIGURE = True
+COLLECT_ALL_DATA = not (MAKE_HISTOGRAM_PLOT or MAKE_SAMPLE_PREDICTION_FIGURE)
+
+previous_step = __import__("0-clean-data")
+test_name = previous_step.test
+data_file_path = previous_step.prefix + test_name + ".pkl"
+
+from util.data import Data
+from util.system import AtomicOpen
+from random import seed, sample
+
+
+data = Data.load(data_file_path)
+data.max_display = len(data)
+print(data)
+print()
+print("Configuration values:")
+print("  freq:   ", sorted(set(data["freq"])))
+print("  fsize:  ", sorted(set(data["fsize"])))
+print("  rsize:  ", sorted(set(data["rsize"])))
+print("  threads:", sorted(set(data["threads"])))
+
+interesting_config = [2800000, 16384, 16384, 64] # <- the
+# configuration to use for distribution model testing.
+
+config_cols = len(interesting_config)
+interest_row = data[:,:config_cols].index( interesting_config )
+
+print()
+print("interest_row: ",interest_row)
+
+# Get the configuration data and values for the example row.
+config = data[interest_row, :config_cols]
+values = sorted(data[interest_row, config_cols:])
+print()
+print("        ", data.names[:config_cols])
+print("config: ",config)
+print("values: ",len(values))
+print()
+
+# Normalize the range of the values.
+min_value = values[0]
+max_value = values[-1]
+values = [(v - min_value) / (max_value - min_value) for v in values]
+
+# Sample out the values (only keep 1000) of them.
+from numpy import linspace
+keep_count = 1000
+values = [values[int(round(i))] for i in linspace(0,len(values)-1,keep_count)]
+
+# Setup the experiment parameters.
+trials = 100
+sample_sizes = [10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 
+                150, 200, 250, 300, 400, 500]
+
+# Get all of the different fits into one list.
+from fits import cdf_points, flat_fit, linear_fit, cubic_fit, quintic_fit
+functions = [flat_fit, linear_fit, cubic_fit, quintic_fit]
+
+# Fit the "truth" function.
+truth = linear_fit(values)
+true_min_max = [0, 1]
+true_x, true_y = cdf_points(values)
+
+if MAKE_HISTOGRAM_PLOT:
[…]

### 2020-01-16 [2019-12]_IEEESE_(quintic_applied)/code/fits.py

@@ -1,0 +1,74 @@
+# Given a list of numbers, generate two lists, one of the CDF x points
+# and one of the CDF y points (for fitting).
+def cdf_points(data):
+    import numpy as np
+    SMALL = 2**(-26)
+    # Sort the data (without changing it) and get the min and max
+    data = np.array(sorted(data))
+    min_pt = data[0]
+    max_pt = data[-1]
+    # Initialize a holder for all CDF points
+    data_vals = []
+    # Add all of the CDF points for the data set
+    for i,val in enumerate(data):
+        if ((i > 0) and (val == data[i-1])): data_vals[-1] = (i+1)/len(data)
+        else:                                data_vals.append( (i+1)/len(data) )
+    # Add the 100 percentile point if it was not added already
+    if (data_vals[-1] != 1.0): data_vals[-1] = 1.0
+    # Convert data into its unique values.
+    data = np.array(sorted(set(data)))
+    # Convert it all to numpy format for ease of processing
+    data_vals = np.array(data_vals)
+    # Return the two lists that define the CDF points.
+    from fraction import Fraction
+    data = list(map(Fraction, data))
+    data_vals = list(map(Fraction, data_vals))
+    return data, data_vals
+
+
+# Return a discontinuous fit.
+def flat_fit(data):
+    from numpy import searchsorted
+    x, y = cdf_points(data)
+    def edf(z):
+        try:    return [value(_) for _ in z]
+        except: return y[min(len(y)-1, max(0,searchsorted(x, z, side="left")-1)
+                             + (1 if z == max(x) else 0))]
+    return edf
+
+# Return a linear fit function.
+def linear_fit(data):
+    from polynomial import fit
+    x, y = cdf_points(data)    
+    cdf = fit(x, y, continuity=0)
+    def fit(z): return max(0,min(1,cdf(z)))
+    fit.derivative = cdf.derivative
+    return fit
+
+# Return a cubic fit function.
+def cubic_fit(data):
+    from monotone import monotone_cubic_spline
+    x, y = cdf_points(data)
+    cdf = monotone_cubic_spline(x, y)
+    def fit(z):
+        val = max(0,min(1,cdf(z)))
+        if (z < x[0]): val = min(val, y[0])
+        if (z > x[-1]): val = max(val, y[-1])
+        return val
+    fit.derivative = cdf.derivative
+    return fit
+
+# Return a quintic fit function.
[…]

### 2020-01-17 [2020-01]_IEEESE_(convex_optimization)/paper.tex
Created paper repository with all of my work. Convex optimization files.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8

\documentclass[conference]{ieeeconf}  % Comment this line out
                                                          % if you need a4paper
%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
\IEEEoverridecommandlockouts                              % This command is only needed 
%% % copyright info
%% \IEEEpubid{\makebox[\columnwidth]{978-1-5386-6133-8/18/\$31.00
%%         ~\copyright~2018 IEEE \hfill}\hspace{\columnsep}
%%         \makebox[\columnwidth]{ }}

% if you want to use the \thanks command
\overrideIEEEmargins
% See the \addtolength command later in the file to balance the column lengths on the last page of the document

% The following packages can be found on http:\\www.ctan.org
\usepackage{tikz}     % For adding axes to plots
\usepackage{graphicx} % For including PDFs
\usetikzlibrary{positioning}
\usepackage{balance} % For balancing the last column.

%% \usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\title{\LARGE \bf Analytic Test Functions for Generalizable
  \\ Evaluation of Convex Optimization Techniques}

%% \author{Thomas C. H. Lux$^{1}$, Tyler H. Chang$^{1}$% <-this % stops a space
%% \thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu, thchang at vt.edu)}}%
%% }
\author{Anonymous Authors}

\begin{document}

\maketitle
\thispagestyle{empty}
\pagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}

Convex optimization algorithms such as gradient descent, quasi-Newton
methods, and their variants are designed to find the global minimum of
strongly convex functions. When these algorithms are applied to the
minimization of non-convex functions they offer no robust theoretical
guarantees. Despite the lack of guarantees, many methods still find
good solutions in practice and are widely used in academia and
industry to solve non-convex problems. In this paper, a set of
analytic test functions and transformations are presented that can be
used to quantify the expected performance of optimization algorithms
on difficult (non-convex) optimization problems. The test functions
and transformations in this set are used to compare and evaluate the
convergence rates of stochastic gradient descent, L-BFGS, AdaGrad, and
Adam.

\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{sec:introduction}

Convex optimization techniques such as stochastic gradient descent
(SGD), Newton's method, and their variants are widely used in machine
learning applications.  Perhaps most notable is the usage of convex
optimization techniques for minimizing neural network loss functions.
Convex optimization is a well studied field, with many theoretical and
practical guarantees on algorithm convergence when applied to convex
objective functions.  Unfortunately, given the non-convex nature of
most loss landscapes, none of the standard theoretical analyses apply
in the context of neural network training.

Therefore in the context of neural network training, the efficiency of
optimization algorithms is measured empirically.  The common
experiment-based analysis poses an issue for designing effective
optimization algorithms, since often the only way to measure
performance is by running the algorithms on complex real world
problems. Even then, an algorithm that performs well on a handful of
problems is not guaranteed to perform well on other problems.  While
the authors of an optimization algorithm may argue that their
algorithm should heuristically perform better given some class of
neural network loss functions, it is almost impossible to make any
theoretical guarantees.

To address this issue, this work presents a framework for empirical
analysis based on the optimization of four analytic objective
functions. This framework is applied to test four well-known
optimization algorithms.  Each of the four objective functions in this
paper has been carefully designed to exhibit some property that
contrasts with those of a \textit{nice} convex function.  Based on
literature and industry usage, the four optimization algorithms
considered here are SGD \cite{nemirovski2009robust}, L-BFGS
\cite{nocedal1980updating,liu1989limited}, AdaGrad
\cite{duchi2013proximal}, and Adam \cite{kingma2014adam}.  It is our
hope that by empirically analyzing the convergence of optimization
algorithms on hand-crafted analytic objective functions, the effects
of each design decision on convergence can be observed and quantified.
This work offers deeper theoretical insight into how the choice of
optimization algorithm can be effected by the expected properties of
the objective function landscape.

In the following paper, first the four optimization algorithms of
interest are introduced along with a summary of their convergence
properties on convex functions.  Next, four analytic objective
functions are introduced that each have a specific rational and
purpose for testing the expected convergence rate of an optimization
algorithm. Following the objective functions, three meaningful
transformations that can be used to tune the difficulty of an
optimization problem are presented. Next, an experimental methodology
is outlined and applied to the chosen optimization algorithms.
Finally, some experimental results are visualized and interpreted.

\section{Algorithms}
\label{sec:algorithms}

For this analysis, four convex optimization algorithms commonly used
in machine learning are considered, specifically these algorithms are
applied to training neural networks. The four algorithms are SGD,
L-BFGS, AdaGrad, and Adam. A summary of each algorithm along with
theoretical convergence guarantees on convex objective functions
follows.

\subsection{Stochastic Gradient Descent}

SGD \cite{nemirovski2009robust} is a slight modification to the classic
gradient descent algorithm:
$$ x^{(k+1)} = x^{(k)} - \alpha^{(k)} \nabla f\left(x^{(k)}\right) $$
where $x^{(k)}$ denotes the $k$th iterate, $\alpha^{(k)}$ denotes the
$k$th step size, and $\nabla f(x)$ denotes the gradient of $f$ at the
point $x$.  Gradient descent can be thought of as an iterative
minimization of the original function $f$ based on its first-order
Taylor expansion:
$$ f(x) \approx f(x^{(k)}) + \nabla f(x^{(k)})^T x $$
subject to the constraint that $\|x^{(k+1)} - x^{(k)}\| \leq
\alpha^{(k)}/\| f(x^{(k)})\|$.

The difference between SGD and the standard gradient descent
algorithm, is that SGD only assumes access to an approximation $g
\approx \nabla f$.  In theory, the condition on the approximation $g$
is that for all $x$,
$$ \mathbb{E}[g(x)] = \nabla f(x). $$
Because $g$ is an approximation to $\nabla f$, it is possible that
each $g(x^{(k)})$ could actually be an ascent direction, making the
convergence of SGD non-monotone, even for strongly convex functions.
Because of these \textit{bad directions}, for a fixed step size
$\alpha^{(k)} = \alpha$, SGD only converges to within some problem
dependent radius of the true optimum $x^\star$, at which point the
convergence stalls.  To achieve further convergence, the step size
must be decayed.  In practice, $\alpha^{(k)}$ is often held constant
for many iterations, then decayed by some factor $\tau \in (0,1)$.

For a strongly convex function and a deterministic gradient, SGD
reduces to standard gradient descent and its convergence is linear.
I.e., given $t$ iterations,
$$ |f(x^{(k)}) - f(x^\star)| \approx \mathcal{O}\left( c^t ) \right) $$
for some constant $0 < c < 1$.  If $g$ is indeed a stochastic estimate
to $\nabla f$, the convergence rate is reduced to
$\mathcal{O}\left(\frac{1}{t}\right)$.  If furthermore the objective
function is only convex (as opposed to strongly convex), this rate is
further reduced to $\mathcal{O}\left(\frac{1}{\sqrt{t}}\right)$.

\subsection{L-BFGS}

In general, quasi-Newton methods use an approximation to the Hessian
$\nabla^2 f$ to allow for bigger step sizes in directions of low
variance.  For a perfect quadratic, this allows Newton methods to
\textit{jump} straight to the minima; for strongly convex functions,
this accommodates poorly conditioned objective functions by normalizing
the sub-level sets of $f$.  The classic Newton update can be derived
from a second-order Taylor approximation to $f$, and is given by:
$$ x^{(k+1)} = x^{(k)} - \left(\nabla^2 f(x^{(k)})\right)^{-1}\nabla f(x^{(k)}). $$

The original Broyden-Fletcher-Goldfarb-Shanno algorithm (BFGS)
algorithm iteratively refines an approximation to the Hessian matrix
$H^{(k)}$ by applying Rank-1 matrix updates to its current
approximation, each of which satisfies the secant condition:
$$ H^{(k)}\left(x^{(k)}-x^{(k-1)}\right) = \nabla f(x^{(k)}) - \nabla f(x^{(k-1)}).$$
Intuitively, this can be thought of as iteratively refining the
Hessian based on a planar fit to each observed gradient.  The Newton
update is defined in terms of the inverse Hessian, but BFGS avoids
performing a matrix inversion by leveraging the Rank-1
Sherman-Morrison-Woodbury matrix identity:
$$(H + xy^T)^{-1} = H^{-1} - H^{-1}x(I + v^T H^{-1}x)v^T H^{-1}$$
where $xy^T$ is a Rank-1 matrix, and $I$ is the identity.  By
leveraging this formula, BFGS is able to keep the iteration cost
computationally cheap, since the cost of the Rank-1 update is
significantly cheaper than the cost of matrix inversion.

L-BFGS \cite{nocedal1980updating,liu1989limited} is a slight
modification to BFGS, which further reduces iteration and storage
costs for high-dimensional problems.  Instead of keeping track of the
entire Hessian matrix $H$, L-BFGS stores only the previous $m$ update
vectors ($x$ and $y$ in the Woodbury matrix formula), then
reconstructs each $H^{(k)}$ in each iteration.  If $m=1$, then L-BFGS
is reduced to the secant method.  If $m=k_{max}$ (the max-iteration
cost) then L-BFGS is equivalent to BFGS, though the storage and
computational cost may be greater or lesser depending on whether
$k_{max}$ is greater than or less than the dimension.  For a typical
application, $m$ is strictly less than the dimension, making this a
computationally efficient algorithm.  As a useful consequence, the
memory limit ensures that L-BFGS can accomodate non-constant Hessians.

For a strongly convex objective function, L-BFGS converges
superlinearly to the optimum $x^\star$.  That is, L-BFGS is faster
than linear but slower than the quadratic convergence rate
$\mathcal{O}\left( c^{b^t} \right)$ (where both $c$ and $b$ are
positive numbers less than one).  When convexity assumptions are
dropped, L-BFGS has no convergence guarantees.  In fact, in the
presence of local maxima and saddle-points, most quasi-Newton methods
will converge to both \cite{dauphin2014identifying}.

\begin{figure*}
  \centering
  \includegraphics[width=0.45\textwidth]{Figures/final-subquadratic}
  \hspace{5mm}
  \includegraphics[width=0.45\textwidth]{Figures/final-saddle}
  \\\vspace{5mm}
  \includegraphics[width=0.45\textwidth]{Figures/final-superquadratic}
  \hspace{5mm}
  \includegraphics[width=0.45\textwidth]{Figures/final-multimin}
  \caption{One dimensional visualizations of each of the four analytic
    objective functions with derivative adjacent to each
    function. These functions are designed to be extended into many
    dimensions by repeating the same 1D function along each
    component.}
  \label{fig:objectives}
\end{figure*}

\subsection{AdaGrad}

AdaGrad \cite{duchi2013proximal} attempts to recreate the benefits of
Newton's method without explicitly approximating the Hessian.  To
achieve this, AdaGrad directly measures the \textit{variance} in
function value with respect to each basis direction.  Specifically,
AdaGrad derives a variance matrix $G$ that captures the same
approximate information as L-BFGS, but with much lower cost since $G$
is always diagonal given an orthonormal basis.  It should be noted
that for a strongly convex function, the Hessian $\nabla^2 f$ is
always symmetric positive definite (SPD), which immediately implies
that it is columnwise diagonally dominant.  Therefore, for strongly
convex functions, a diagonal matrix $G$ derived from variance
information generally makes a reasonable approximation to the true
Hessian $\nabla^2 f$.

The $k$th variance estimate for each dimension of $G$ is given by
$$ G^{(k)} = diag\left(\sqrt{\sum_{n=1}^k (g^{(n)})^2} + \varepsilon\right) $$

where each $g^{(n)}$ is a previous gradient (estimate) and
$\varepsilon$ is an error-correction factor, introduced to prevent $G$
from becoming singular in degenerate cases.  Note that since $G^{(k)}$
is diagonal, it can be readily inverted.  Also, by storing
$\left(G^{(k)}\right)^2 - \varepsilon I$ and performing the square
root and error-accommodating operations on demand, $G$ can be
iteratively refined without tracking previous gradients.  To normalize
the sublevel sets and improve the conditioning of $f$, a scaled
version of $G$ can be directly plugged in for $H$ in the quasi-Newton
update:
$$ x^{(k+1)} = x^{(k)} - \alpha^{(k)} (G^{(k)})^{-1} g(x^{(k)}) $$
where $\alpha^{(k)}$ is a step size, and $g$ is an approximation to
$\nabla f$ in the stochastic case and $g = \nabla f$ in the
deterministic case.  More intuitively, AdaGrad can also be thought of
as a trust region method, where the variance estimate $G$ allows for
larger steps in directions of low variance.

AdaGrad is guaranteed the same convergence as SGD, but the constant
terms that are ignored by the Big-O notation are siginificantly better
for AdaGrad when the problem is poorly conditioned.

\subsection{Adam}

Adam \cite{kingma2014adam} combines the idea of variance estimation
from AdaGrad, with the idea of momentum.  Intuitively, momentum places
some weight on previous iterates by replacing the current gradient
estimate $g$ with a weighted average of $g$ and the previously seen
gradients:
$$ \hat{g}^{(k+1)}(x^{(k)}) = \beta g(x^{(k)}) + (1-\beta)\hat{g}^{(k)}. $$
When $g$ is a stochastic estimate, this has the effect of smoothing
over noise and avoiding wild oscillations. For poorly conditioned
problems, this prevents the iterates $x^{(k)}$ from wildly oscillating
about the optimum descent direction; for non-convex functions, this
can allow Adam to step through sharp minima, which often correspond to
poor generalization error.

Leveraging the variance estimate $G$ used by AdaGrad, Adam applies
momentum not only to the gradient (first moment) estimate, but also
applies momentum to the variance matrix $G$ (second moment).
Therefore, the Adam update can be summarized by:
\begin{align*}
  x^{(k+1)} &= x^{(k)} - \alpha \left(\sqrt{\beta_2 g^2(x^{(k)}) + (1-\beta_2)(G^{(k)})^2}\right)^{-1}\\
  &\quad\cdot\left(\beta_1 g(x^{(k)}) + (1-\beta_1)\hat{g}\right)
\end{align*}
where $\beta_1$ and $\beta_2$ are the first and second moment
coefficients respectively, and $G^{(k)}$ is the $k$th non-corrected
variance estimate from AdaGrad.  Large momentum coefficients are most
helpful for noisy and poorly conditioned problems.  However, if the
momentum coefficient is too large with respect to the step size
$\alpha$, Adam can fail to converge.  Most interesting problems are
noisy and poorly conditioned, and most algorithms tend to converge
well for any well-conditioned problem.  So, it is common practice to
set $\beta_1 \approx 1$ and $\beta_2 \approx 1$, then choose the
largest convergent step size $\alpha$.

Similarly to AdaGrad, Adam converges at the same rate as SGD but with
more favorable hidden constants when the problem is poorly
conditioned.

\begin{figure*}
  \centering
  \includegraphics[width=0.45\textwidth]{Figures/final-algorithm-500}
  \hspace{5mm}
  \includegraphics[width=0.45\textwidth]{Figures/final-algorithm-50K}
  \caption{Convergence results averaged over all objective functions,
    dimensions, amounts of noise, rotation, and skew. The thickest
    line in each series represents the median, while the thin lines of
    the same style (and color) represent the 10th and 90th
    percentiles. The left figure depicts only the first 500 steps
    while the right figure depicts all 50 thousand steps. Adam and SGD
    are the best performers beyond two thousand steps. L-BFGS
    performs best in the first 50 steps, then remains second to Adam
    until SGD overtakes it at roughly two thousand steps.}
  \label{fig:results-all}
\end{figure*}

\section{Analytic Objective Functions}

In order to empirically evaluate each optimization technique in a
generalizable way, four analytic functions for minimization are
presented.  Each of these test functions has a single global minimum
and specially designed challenges for typical convex optimization
techniques.

\subsection{Sub-quadratic}

The sub-quadratic is a convex function that appears to come to a
sudden point at the global minimum.  Many common optimization
algorithms will tend to overshoot this minimum due to over stepping.
The function is defined as follows
$$ \sum_{i=1}^{d} |x_i|^{\frac{2k}{2k-1}} $$
where $d$ is the dimension of the problem and $k > 1$.  A
one-dimensional plot of the function and its derivative is shown in
Figure \ref{fig:objectives}.

\subsection{Super-quadratic}

The super-quadratic function is a convex function used to mimic a
phenomenon observed in practice where the region surrounding an
optimal point has a gradient whose magnitude goes to zero at a rapidly
decreasing rate.  Visually, this manifests as a \textit{flattening}
surrounding the global minimum.  The function is defined as follows
$$ \sum_{i=1}^{d} x_i^{2k} $$
where $d$ is the dimension of the problem and $k > 1$.  A
one-dimensional plot of the function and its derivative is shown in
Figure \ref{fig:objectives}.

\subsection{Saddle}

In problems with tens or more dimensions, the likelihood of
non-uniform curvature between dimensions becomes increasingly likely.
When dimensions have opposing curvature, {\it saddle points} are
created.  Recent work \cite{dauphin2014identifying} has shown that
saddle points are a very common occurrence when training neural
networks.  Analytically we define the following function that has
exponentially more saddle points with growing dimension.
$$ \sum_{i=1}^{d} \frac{s^4 x^2}{2} - \frac{s^2 x^4}{2} + \frac{x^6}{6}, $$
where $d$ is the dimension of the problem and $s$ is the constant
defining the absolute value of the location of saddle points
per-dimension.  A one-dimensional plot of the function and its
derivative is shown in Figure \ref{fig:objectives}.

\subsection{Multimin}

Many problems that require optimization have local minima. Using
Chebyshev polynomials, a function is constructed that has a prescribed
number of local minima whose occurrence grows exponentially with
increasing dimension.
$$ \sum_{i=1}^{d} \big [  1 + a x_i^2 + f_{2m + 1}(x_i) \big ], $$
\begin{align*}
  f_0(x_i) &= 1, \\
  f_1(x_i) &= x_i, \\
  f_{n+1}(x_i) &= 2 x_i f_{n}(x_i) - f_{n-1}(x_i), \\
\end{align*}
Here, $d$ is the dimension of the data, $a$ is a multiplier for
determining the relative effect size of the quadratic term, and $m$ is
the number of local minima per dimension. The number of local minima
in the space will grow as $m^d$. When $m$ is odd there will be one
global minimum; this is recommended.  When $m$ is even there will be
$2^d$ global minimizers. A one-dimensional plot of the function and
its derivative is shown in Figure \ref{fig:objectives}.

\begin{figure*}
  \centering
  \includegraphics[width=0.45\textwidth]{Figures/final-function-1}
  \includegraphics[width=0.45\textwidth]{Figures/final-function-2}
  \caption{Convergence results broken up by function and averaged over
    all dimensions, amounts of noise, rotation, and skew. The thickest
    line in each series represents the median, while the thin lines of
    the same style (and color) represent the 10th and 90th
    percentiles. Adam converges more quickly, but is overtaken by SGD
    after tens of thousands of steps for the sub- and super-quadratic
    functions. Adam and L-BFGS perform best on the saddle objective,
    while SGD and Adam perform best for the multi-minimum objective.}
  \label{fig:results-function}
\end{figure*}

\section{Function Transformations}

Along with the four different objective functions, three analytic
transformations are presented that allow for careful tuning of both
the type and the difficulty of challenges presented to convex
optimization techniques. These three transformations are chosen to
mimic common problems faced in real-world applications.

\subsection{Noise}

The analytic functions that have been presented thus far all have
well-defined, with deterministic gradients almost everywhere.  In most
neural network training applications, a subset of the total training
data volume is used in each gradient evaluation, resulting in a
stochastic estimate of the true gradient.  To simulate this reality,
various amounts of uniform random noise are added to each gradient
evaluation.  SGD and other first order methods (such as Adam and
AdaGrad) are expected to still converge under these conditions
\cite{nemirovski2009robust}.  Though the analysis mentions no
constraint on the variance of the noise, the maximum magnitude of the
uniform noise has been selectively capped at 25\% of the maximum
magnitude of the gradient for our evaluations.  Let $\|g\|_{L^\infty}$
denote the maximum magnitude of the gradient.  For each objective
function, optimizations are attempted with no noise, uniform noise
with 12.5\% of $\|g\|_{L^\infty}$, and uniform noise with 25\% of
$\|g\|_{L^\infty}$. These three amounts of noise are further referred
to as $0$ noise, $.5$ (12.5\%) noise, and $1$ (25\%) noise.

\subsection{Skew}

The condition number of the sub-level set $C$ of a function $f$ is
defined as the ratio between the maximum diameter $W_{max}$ and the
minimum diameter $W_{min}$ across $C$:
$$ \kappa(C) = \frac{W_{max}}{W_{min}}.$$

For $f$ convex, this is proportional to the conditioning of $f$ as an
operator.  Notice that the sublevel sets for all the presented
objective functions are approximately square.  This means that without
modification, the problems are all well-conditioned, i.e., $\kappa(f)
\approx 1$.  To simulate poor problem conditioning, which is common in
machine learning applications, various amounts of skew are introduced
on $f$.  For each objective function, optimizations are attempted with
no skew, an inverse conditioning ratio of $\frac{W_{min}}{W_{max}} =
0.5,$ and an inverse conditioning of $\frac{W_{min}}{W_{max}} = 0.01.$

\subsection{Rotation}

Finally note that each of the presented objective functions is
completely separable, in that it can be decomposed into the sum of its
components in each dimension, which can be optimized separately.  For
Adam and AdaGrad, which use diagonal matrices to capture variance in
each basis dimension, this means that all the necessarry information
can be captured without need for off-diagonal elements.  However, as
the functions are rotated to a maximum angle of $\pi / 8$ radians, the
objective functions become non-separable, making Adam and AdaGrad's
variance approximations poor proxies for the true Hessian.  To
simulate non-separability, optimization algorithms are applied to each
objective function with no rotation, $\pi / 16$ radian rotation ($.5$
rotation), and full $\pi / 8$ degree rotation ($1$ rotation).

\begin{figure}
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    & $m$ & $k$ & $a$ & $s$ \\
    \hline
    Sub-quadratic & N/A & 2 & N/A & N/A \\
    Super-quadratic & N/A & 2 & N/A & N/A \\
    Saddle Point & N/A & N/A & N/A & 0.75 \\
    Mult-Min & 3 & N/A & 2 & N/a \\
    \hline
  \end{tabular}
  \caption{The constants that are used within each objective function. $k$ is the power multiplier used for the sub- and super-quadratic functions. $m$ is the number of minima per dimension in the multi-min objective. And $s$ determines the (positive and negative) locations at which the saddle function will have a directional derivative of zero along each dimension.}
  \label{tab:const}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figures/final-noise}
  \caption{Convergence results broken up by noise and averaged over
    all functions, dimensions, amounts of rotation, and skew. The
    thickest line in each series represents the median, while the thin
    lines of the same style (and color) represent the 10th and 90th
    percentiles. The introduction of any noise slows the convergence
    of all algorithms while SGD becomes the best. Adam is similar, but
    lacks the later convergence achieved by SGD.}
  \label{fig:results-noise}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figures/final-skew}
  \caption{Convergence results broken up by skew and averaged over all
    functions, dimensions, amounts of noise, and rotation. The
    thickest line in each series represents the median, while the thin
    lines of the same style (and color) represent the 10th and 90th
    percentiles. The existence of mild skew causes L-BFGS to become
    the best technique, while large amounts of skew and no skew both
    lend to Adam remaining the best.}
  \label{fig:results-skew}
  \vspace{1.2cm}
\end{figure}

\begin{figure}
  \centering
  \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    & $\alpha$ & $\beta_1$ & $\beta_2$ & $\tau$ & $\varepsilon$ \\
    \hline
    SGD     & $0.1$  & -- & -- & $0.5$ & -- \\
    L-BFGS & $0.99$ & -- & -- & -- & -- \\
    AdaGrad & $0.01$ & -- & -- & -- & $10^{-6}$ \\
    Adam    & $0.01$ & $0.9$ & $0.99$ & -- & $10^{-8}$ \\
    \hline
  \end{tabular}
  \caption{The hyperparameter settings used for each optimization
    algorithm. The values chosen are those either recommended in the
    source paper, or tuned lightly for this test set (in the case of
    $\alpha$ for SGD). The value of $m$ used for L-BFGS was chosen as
    the maximum of ten and the square root of the dimension.}
  \label{tab:hypeparam}
  \vspace{0.4cm}
\end{figure}

\section{Implementation and Data Collection}

All of the objective functions were implemented in Python, and their
gradients were generated using the Python automatic differentiation
tool {\tt autograd}.  The constants used for the objective functions
are presented in Figure \ref{tab:const}.

The four optimization algorithms discussed have been coded in Python.
For all algorithms the hyperparameter settings were based upon
recommended settings in source papers, specifically for $\beta_1$,
$\beta_2$, $\tau$, and $\varepsilon$, while the step size $\alpha$ was
tuned for reasonable performance. The value of $m$ used for L-BFGS
was chosen as the maximum of ten and the square root of the
dimension. Figure \ref{tab:hypeparam} shows the selected hyperparameter
values for each algorithm.  For SGD, the decay factor $\tau$ was
applied after every five thousand iterations.

Each algorithm was run on each noise level, skew, and rotation
individually for all combinations of all objective functions in
dimensions 10, 100, and 1000. For each objective function and each
level of noise, skew, and rotation, 100 independent trials were run
from (common) random starting points uniformly distributed in the unit
hypercube. The optimization algorithms were permitted 50 thousand
objective function evaluations. Note that each of the four objective
functions has a single global minimum where $f(x) = 0$, and is upper
bounded by $f(x) = 1$.

\section{Results}

The two overall best performers for the analytic objective functions
given varying noise, skew, and rotation were Adam and SGD.  Figure
\ref{fig:results-all} shows the median objective value obtained versus
number of steps for each optimization algorithm over all test
functions and values for noise, skew, and rotation.

The results are broken up by function in Figure
\ref{fig:results-function}, where multiple (expected) behaviors of
each optimization algorithm can be observed. For the sub- and
super-quadratic functions, the decreasing step size of SGD allows
better tail convergence than any other technique. Adam performs best
on the saddle function because its momentum and strictly positive
second-order estimate of objective function curvature allow it to
continue walking closer to the minimum without getting stuck at a
local plateau in the gradient. None of the optimization algorithms are
able to successfully minimize the saddle and multi-min problems, as
these are incredibly difficult in high dimension.

Some unexpected and difficult-to-explain behaviors also occur.  It is
unclear why SGD obtains better tail performance than Adam on the
multi-min problem. Perhaps the step size becomes just the right
size to step out of the local minima.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figures/final-rotation}
  \caption{Convergence results broken up by rotation and averaged over
    all functions, dimensions, amounts of noise, and skew. The
    thickest line in each series represents the median, while the thin
    lines of the same style (and color) represent the 10th and 90th
    percentiles. The incorporation of rotation has little to no effect
    on the convergence for the evaluated optimization algorithms.}
  \label{fig:results-rotation}
\end{figure}

\subsection{Convergence by Noise}

In Figure \ref{fig:results-noise}, the effect of increased noise in
the gradient of the objective function is studied.

As expected, the addition of noise significantly slows the convergence
of all the algorithms for all the functions.  However, the addition of
noise seems to allow SGD better performance than Adam on average,
especially for amounts of noise that match the hyperparameters of SGD
well (as appears to be the case for a noise of 0.5).

\subsection{Convergence by Skew}

In Figure \ref{fig:results-skew}, the effect of increased skew (i.e.,
deteriorating the conditioning) of the objective functions is studied.

Interestingly, SGD seems to be unaffected by skew. All of Adam,
AdaGrad, and L-BFGS are capable of compensating for skew, so it is
expected that their performance would not be impacted by changing
skew.

\subsection{Convergence by Rotation}

In Figure \ref{fig:results-rotation}, the effect of increased rotation
of the objective functions (which corresponds to non-separability) is
studied.  It was assumed that this could negatively impact Adam and
AdaGrad, but none of the algorithms are significantly affected.

\section{Conclusion}

In this paper, a test set of analytic objective functions and
transformations were presented and used to analyze the convergence of
four common convex optimization algorithms. The specific challenges
posed by the analytic objective functions and the associated
transformations were constructed to match expected behaviors of
real-world problems.  Empirical results confirm common observations
with regards to ADAM and SGD being good choices for non-convex
optimization. However, empirical results also suggest that an initial
burst of 10 to 100 steps of L-BFGS may improve the early convergence
of optimization algorithms in practice. Further theoretical insight
and improved generalizability of results may be allowed by continued
usage of this analytic objective function test set or ones like it.

\balance

\bibliographystyle{IEEEtran}
\bibliography{paper}

\end{document}

### 2020-01-17 [2020-01]_IEEESE_(convex_optimization)/Old/lbfgs_fortran/main.f90

@@ -1,0 +1,30 @@
+PROGRAM TEST
+USE LBFGS_MOD
+IMPLICIT NONE
+INTEGER, PARAMETER :: D = 10
+REAL(KIND=R8) :: X(D)
+REAL(KIND=R8), ALLOCATABLE :: HIST(:,:)
+INTEGER :: IERR, I
+X(:) = 1.0_R8
+
+CALL LBFGS(D, QUAD_GRAD, X, IERR, HISTORY=HIST, EPS=EPSILON(0.0_R8)**2.0_R8)
+IF (IERR .NE. 0) PRINT *, "ERROR : ", IERR
+
+PRINT *, "X = ", X(:)
+
+CONTAINS
+
+SUBROUTINE QUAD_GRAD(X, G)
+  REAL(KIND=R8), INTENT(IN) :: X(:)
+  REAL(KIND=R8), INTENT(OUT) :: G(SIZE(X))
+  REAL(KIND=R8) :: NOISE(SIZE(X))
+  G(:) = 2.0_R8 * X(:)
+  ! G(:) = 4.0_R8 * (X(:) ** 3.0_R8)
+  ! G(:) = 6.0_R8 * (X(:) ** 5.0_R8)
+  CALL RANDOM_NUMBER(NOISE(:))
+  NOISE(:) = (NOISE(:) - .5_R8) * 2_R8
+  G(:) = G(:) + .01_R8 * NOISE(:)
+  RETURN
+END SUBROUTINE QUAD_GRAD
+
+END PROGRAM TEST

### 2020-01-17 [2020-01]_IEEESE_(convex_optimization)/Old/code/objective.py
Created paper repository with all of my work. Convex optimization files.

@@ -1,0 +1,261 @@
+# Get numpy data structures (in differentiable form)
+from autograd.numpy import zeros, ones, arange, linspace, \
+    identity, roll, radians, random, linalg
+# Get numpy mathematical expressions (in differentiable form)
+from autograd.numpy import pi, sum, exp, sin, cos, log, outer, dot, matmul
+# Use an automatic elementwise-gradient function
+# from autograd import grad
+from autograd import elementwise_grad as grad
+
+# Set up a repeatable random number generator
+SEED = 0
+RANDOM = random.RandomState(SEED)
+# Store all the objective functions in a global variable
+FUNCTIONS = []
+
+
+# Decorator for converting a function into an objective function
+def objectify(lower=None, upper=None, max_val=None, solution=None, gradient=None):
+    # Define the "normalize" decorator that handles given arguments
+    def decorator(func):
+        def obj_func(x, *args, **kwargs):
+            # Rescale the x out of [-1,1] and into the domain of the
+            # function if a normalization was necessary
+            if (lower is not None) and (upper is not None):
+                x = ((x + 1) * (upper - lower) / 2) + lower
+            # Return the objective value
+            value = func(x, *args, **kwargs)
+            # Rescale the objective value if necessary
+            if hasattr(x, "__len__"):
+                if (max_val is not None):
+                    value /= max_val*len(x)
+            else: value /= max_val
+            # Return the value.
+            return value
+        # Set a random point generator (in the -1,1 range)
+        obj_func.rand = lambda d, gen=RANDOM: gen.rand(d) * 2 - 1
+        # Set a solution
+        if solution: obj_func.sol  = solution
+        else:        obj_func.sol  = lambda d: zeros(d)
+        # Set a gradient
+        if gradient: obj_func.grad = gradient
+        else:        obj_func.grad = grad(obj_func)
+        obj_func.lower, obj_func.upper = -1., 1.
+        # Store the maximum gradient magnitude
+        obj_func.max_grad = max(abs(obj_func.grad(ones(1)*obj_func.lower)),
+                                abs(obj_func.grad(ones(1)*obj_func.upper)))
+        # Copy the name over to the new function
+        obj_func.__name__ = func.__name__ + "_(objectified)"
+        obj_func.func = func
+        # Add the objective function to the list of functions
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/test_one.py
Updated IEEESE paper with revisions suggested by Dr Hong. Added backwards compatible 'util' subdirectory to the NUMA paper to begin working on revisions.

@@ -1,0 +1,90 @@
+import os, time, random
+import numpy as np
+from util.system import load, save, Timer
+from util.math import fekete_points
+from util.approximate import DelaunayP1, NeuralNetwork
+from util.plot import Plot
+
+seed = 0
+dimension = 2
+test_size = 1000
+train_sizes = [2 * 4**(i) for i in range(1,6+1)]
+train = train_sizes[-1]
+snr = 0.0
+# The models to test.
+models = [NeuralNetwork(epochs=epochs) for epochs in (100, 1000, 5000, 10000)]
+names = [f"{epochs} epochs" for epochs in (100, 1000, 5000, 10000)]
+
+# A modified version of the oscillatory test problem from:
+#    'High Dimensional Polynomial Interpolation' by Barthelman et al. (2000)
+# It is a cosine function of the norm of the input vectors. This
+# function has no componentwise structure to take advantage of.
+def oscillatory(x, c=1):
+    if len(x.shape) <= 1:
+        return np.cos(c*np.pi + np.linalg.norm(x))
+    elif len(x.shape) == 2:
+        return np.cos(c*np.pi + np.linalg.norm(x, axis=1))
+    else:
+        class UsageError(Exception): pass
+        raise(UsageError(f"This objective function only supports arrays and matrices, received shape {x.shape}."))
+
+test_function = oscillatory
+# Get the Fekete points.
+fekete_file = os.path.join(
+    "fekete_points", f"fekete-{train}-{dimension}.pkl")
+print(f"Searching for '{fekete_file}'..")
+try:
+    train_points = load(fekete_file)
+except:
+    train_points = fekete_points(train, dimension)
+    save(train_points, fekete_file)
+
+print("generating test poitns..")
+test_points = np.random.random(size=(test_size, dimension))
+test_values = test_function(test_points)
+print("  using train points:", train_points.shape)
+print("  using test points: ", test_points.shape)
+# Cycle through different "signal-to-noise ratios"
+print()
+print("-"*35)
+# Re-seed the random number generators for each test.
+random.seed(seed)
+np.random.seed(seed)
+snr_string = f"{snr:.2f}"
+print("SNR", snr_string)
+# Evaluate the test function at those points.
+train_values = test_function(train_points)
+# Generate noise (relative to values) and add it on to the values.
+noise_ratio = snr * (np.random.random(size=train_values.shape)*2 - 1)
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/approximate/modified_shepard/__init__.py

@@ -1,0 +1,100 @@
+# Makes available all custom algorithms that I have worked with
+import os
+import numpy as np
+import fmodpy
+from util.approximate import WeightedApproximator
+
+# This directory
+CWD = os.path.dirname(os.path.abspath(__file__))
+# Import the fortran source code.
+modified_shepard = fmodpy.fimport(
+    os.path.join(CWD,"modified_shepard.f95"),
+    output_directory=CWD)
+shepmod = modified_shepard.shepmod
+shepmodval = modified_shepard.shepmodval
+
+
+class TooFewDataPoints(Exception):
+    def __init__(self, d, n):
+        super().__init__(f"Only {n} points were provided but {d+2} are needed to define a modified shepard interpolant.")
+
+# Wrapper class for using the modified shepard fortran code
+class ShepMod(WeightedApproximator):
+    def __init__(self):
+        self.ierrors = {}
+        self.x = self.rw = None
+
+    # Use fortran code to compute the boxes for the given data
+    def _fit(self, control_points, **kwargs):
+        # Local operations
+        self.x = np.asfortranarray(control_points.T)
+        self.rw = np.ones(shape=(self.x.shape[1],), order="F")
+        # In-place update of self.rw (radius of influence).
+        _, ier = shepmod(self.x.shape[0], self.x.shape[1],
+                         self.x, self.rw, **kwargs)
+        if (ier != 0):
+            raise(TooFewDataPoints(*self.x.shape))
+
+    # Use fortran code to evaluate the computed boxes
+    def _predict(self, x):
+        # Calculate all of the response values
+        indices = []
+        weights = []
+        base = np.arange(self.x.shape[1])
+        for x_pt in x:
+            x_pt = np.array(np.reshape(x_pt,(self.x.shape[0],)), order="F")
+            ierr = 0
+            wts = np.zeros(self.x.shape[1], order="F")
+            _, ierr = shepmodval(x_pt, *self.x.shape, self.x, self.rw, wts)
+            # Store the error flag (for processing later).
+            self.ierrors[ierr] = self.ierrors.get(ierr, 0) + 1
+            # Get those indices where there is a nonzero weight.
+            ids = base[wts > 0]
+            # Append the impactful indices and weights.
+            indices.append( ids )
+            weights.append( wts[ids] )
+        # Return the source indices and weights associated with all
+        # interpolation points.
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/approximate/testing.py

@@ -1,0 +1,88 @@
+import numpy as np
+
+# Given a model, test the time it takes to make a prediction
+def test_time(model, ns=range(10,10011,1000), ds=range(2,11,2)):
+    import time
+    print("Starting")
+    for d in ds:
+        print(f"D: {d}")
+        print()
+        for n in ns:
+            print(f" N: {n}")
+            x = np.random.random((n,d))
+            y = np.random.random((n,))
+            start = time.time()
+            model.fit(x, y)
+            print(f"  fit:     {time.time() - start:5.2f}")
+            start = time.time()
+            model(x[0])
+            print(f"  predict: {time.time() - start:5.2f}")
+            print()
+        print()
+
+# Given a model, generate a test plot demonstrating the surface it procudes
+def test_plot(model, low=0, upp=1, plot_points=3000, p=None,
+              fun=lambda x: 3*x[0]+.5*np.cos(8*x[0])+np.sin(5*x[-1]),
+              N=20, D=2, noise=0., random=True, seed=0, x=None, 
+              y=None, classifier=False):
+    np.random.seed(seed)
+    provided_points = (type(x) != type(None)) and (type(y) != type(None))
+    if (type(x) == type(None)):
+        # Generate x points
+        if random:
+            x = np.random.random(size=(N,D))
+        else:
+            N = int(round(N ** (1/D)))
+            x = np.array([r.flatten() for r in np.meshgrid(*[np.linspace(0,1,N)]*D)]).T
+    if (type(y) == type(None)):
+        # Calculate response values
+        y = np.array([round(fun(v) + np.random.random()*noise)
+                      if classifier else
+                      fun(v)+np.random.random()*noise for v in x])
+    # Fit the model to the points
+    model.fit(x,y, classifier=classifier)
+    # Generate the plot
+    from util.plot import Plot
+    if type(p) == type(None): p = Plot()
+    if not provided_points: p.add("Training Points", *x.T, y)
+    p.add_func(str(model), model, *([(low-.1,upp+.1)]*D),
+               plot_points=plot_points, vectorized=True)
+    # p.add_func("truth", fun, *([(low-.1,upp+.1)]*D),
+    #            plot_points=plot_points)
+    return p, x, y
+
+# Given a model, use the method "points_and_weights" to identify the
+# regions of support for each of the interpolation points.
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/approximate/wrappers.py

@@ -1,0 +1,133 @@
+import numpy as np
+from util.math import abs_diff
+
+MAX_DIM = None
+MAX_SAMPLES = None
+DEFAULT_COND_SCALE = True
+DEFAULT_CONDITIONER = "PCA" # MPCA
+
+# A wrapper for an approximator that must fit unique points. Use this
+# wrapper on an un-initialized "weighted approximator" to inheret all
+# methods while modifying the "fit" and "predict" methods.
+def unique(weighted_approximator):
+    # ----------------------------------------------------------------
+    # A class for wrapping an approximator.
+    class UniquePointsApproximator(weighted_approximator):
+        original_points = None
+        unique_points = None
+        unique_indices = None
+        original_values = None
+
+        # Wrap the fit method to capture only unique points.
+        def fit(self, points, values=None, *args, **kwargs):
+            if ((not issubclass(type(points), np.ndarray)) or (len(points.shape) != 2)):
+                from util.approximate.base import UnexpectedType
+                raise(UnexpectedType("Expected 2D numpy array as first argument."))
+            self.original_points = points
+            self.unique_points = {}
+            for i,pt in enumerate(points):
+                pt = tuple(pt)
+                self.unique_points[pt] = self.unique_points.get(pt, []) + [i]
+            # Store the indices of the first occurrence of each unique point.
+            self.unique_indices = np.array(sorted(
+                self.unique_points[pt][0] for pt in self.unique_points))
+            # Average the response value for the points that are identical.
+            if (values is not None):
+                self.original_values = values
+                to_add = set(self.unique_points)
+                avg_values = []
+                for pt in self.original_points:
+                    pt = tuple(pt)
+                    if pt in to_add:
+                        indices = self.unique_points[pt]
+                        wt = 1. / len(indices)
+                        avg_values.append( sum(values[i]*wt for i in indices) )
+                        to_add.remove(pt)
+                args = args + (avg_values,)
+            # Call the fit method on parent with unique points only.
+            return super().fit(self.original_points[self.unique_indices,:], 
+                               *args, **kwargs)
+            
+        # Wrap the predict method to return original points.
+        def predict(self, points, *args, **kwargs):
+            if ((not issubclass(type(points), np.ndarray)) or (0 >= len(points.shape) < 2)):
+                from util.approximate.base import UnexpectedType
+                raise(UnexpectedType("Provided 'points' should be a 1D or 2D numpy array."))
+            # If values were provided, return usual prediction.
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/development/testing/test_random.py

@@ -1,0 +1,109 @@
+from util.random import *
+
+# ====================================================================
+#                           util.random     
+# ====================================================================
+
+def test_latin(display=False):
+    from util.random import latin
+    print("Testing latin..", end=" ")
+    if display: print()
+
+    if display:
+        from util.plot import Plot
+        D = 2
+        N = 400
+
+        p = Plot("Latin hyper cube design")
+        # Add grid lines for the cube.
+        for i in range(N+1):
+            p.add(f"gly {i}", [0,1], [i / N]*2, mode="lines", group="grid",
+                  color="rgba(.6,.1,.1,.2)", show_in_legend=(i==0))
+            p.add(f"glx {i}", [i / N]*2, [0,1], mode="lines", group="grid",
+                  color="rgba(.6,.1,.1,.2)", show_in_legend=False)
+        # Add the random points.
+        p.add("Points", *(latin(N,D).T))
+        p.show(file_name="/Users/thomaslux/Desktop/lhc.html")
+
+    from numpy import sort
+    N = 10000
+    D = 1000
+    pts = latin(N, D)
+    for i in range(D):
+        values = sort(pts[:,i])
+        max_adj_diff = max(abs(values[:-1] - values[1:]))
+        if max_adj_diff >= (2/N):
+            print("BAD!", "   ", i, max_adj_diff, 1/N, 2/N)
+        assert(max_adj_diff < (2/N))
+
+    print("passed.")
+
+
+def test_random_range(display=False):
+    # print("Testing random_range..", end=" ")
+    # if display: print()
+    from util.random import random_range
+
+    if display:
+        print()
+        print("-"*70)
+        print("Begin tests for 'random_range'")
+
+    import random
+    # Check unique values witnessed with sliced range.
+    seen = {}
+    for i in range(1000):
+        random.seed(i)
+        v = tuple(random_range(4))
+        seen[v] = seen.get(v,0) + 1
+    if display:
+        for row in sorted(seen):
+            print("%5d"%seen[row], row)
+    # vals = {}
+    # mults = {}
+    # for i in range(10000):
+    #     results = tuple(random_range(20))
+    #     key, (val, mult) = results[:-1], results[-1]
+    #     diff = set(key[i-1] - key[i] for i in range(len(key)))
+    #     if len(diff) <= 2:
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/development/testing/test_stats.py

@@ -1,0 +1,530 @@
+import numpy as np
+from util.stats import *
+
+# ====================================================================
+#                           util.stats     
+# ====================================================================
+
+def _test_mpca(display=False):
+    if display:
+        print()
+        print("-"*70)
+        print("Begin tests for 'mpca'")
+
+    GENERATE_APPROXIMATIONS = True
+    BIG_PLOTS = False
+    SHOW_2D_POINTS = False
+
+    import random
+    # Generate some points for testing.
+    np.random.seed(4) # 4
+    random.seed(0) # 0
+    rgen = np.random.RandomState(1) # 10
+    n = 100
+    points = (rgen.rand(n,2) - .5) * 2
+    # points *= np.array([.5, 1.])
+
+    # Create some testing functions (for learning different behaviors)
+    funcs = [
+        lambda x: x[1]               , # Linear on y
+        lambda x: abs(x[0] + x[1])   , # "V" function on 1:1 diagonal
+        lambda x: abs(2*x[0] + x[1]) , # "V" function on 2:1 diagonal
+        lambda x: x[0]**2            , # Quadratic on x
+        lambda x: (x[0] + x[1])**2   , # Quadratic on 1:1 diagonal
+        lambda x: (2*x[0] + x[1])**3 , # Cubic on 2:1 diagonal
+        lambda x: (x[0]**3)          , # Cubic on x
+        lambda x: rgen.rand()        , # Random function
+    ]
+    # Calculate the response values associated with each function.
+    responses = np.vstack(tuple(tuple(map(f, points)) for f in funcs)).T
+
+    # Reduce to just the first function
+    choice = 3
+    func = funcs[choice]
+    response = responses[:,choice]
+
+    # Run the princinple response analysis function.
+    components, values = mpca(points, response)
+    values /= np.sum(values)
+    conditioner = np.matmul(components, np.diag(values))
+
+    if display:
+        print()
+        print("Components")
+        print(components)
+        print()
+        print("Values")
+        print(values)
+        print()
+        print("Conditioner")
+        print(conditioner)
+        print()
+
+    components = np.array([[1.0, 0.], [0., 1.]])
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/misc/image.py

@@ -1,0 +1,176 @@
+# Import python default packages
+import os, sys, time
+import numpy as np
+from PIL import Image
+
+# Define globals
+PERSPECTIVE_RANGE = [-.2, .2] # Range of real-world replicable changes
+ROTATION_RANGE = [-20, 20] # Steepest roads in America
+SCALE_RANGE = [-3,5] # Some common and reasonable zoom ranges
+
+# Define some errors that might be raised in this code
+class CommandError(Exception): pass
+class MissingPackage(Exception): pass
+
+# Check the python version
+if (sys.version_info[0] < 3):
+    print("WARNING: This code was written and tested for Python3 only.")
+
+# Find the coefficients of the trasnformation matrix necessary to
+# perform an affine trasnformation that maps the points in
+# "input_points" to the points in "output_points"
+def find_coeffs(input_points, output_points):
+    matrix = []
+    for p1, p2 in zip(input_points, output_points):
+        matrix.append([p1[0], p1[1], 1, 0, 0, 0, -p2[0]*p1[0], -p2[0]*p1[1]])
+        matrix.append([0, 0, 0, p1[0], p1[1], 1, -p2[1]*p1[0], -p2[1]*p1[1]])
+    A = np.matrix(matrix, dtype=np.float)
+    B = np.array(output_points).reshape(8)
+    res = np.dot(np.linalg.inv(A.T * A) * A.T, B)
+    return np.array(res).reshape(8)
+
+# Perspective change matrix generator
+def perspective(img, magnitude):
+    left_change = right_change = 0
+    if magnitude < 0:
+        left_change = abs(img.size[1] * magnitude)
+    elif magnitude > 0:
+        right_change = abs(img.size[1] * magnitude)    
+    input_points = [(0,0+left_change),
+                    (img.size[0],0+right_change),
+                    (img.size[0],img.size[1]-right_change),
+                    (0,img.size[0]-left_change),]
+    output_points = [(0,0),
+                     (img.size[0],0),
+                     (img.size[0], img.size[1]),
+                     (0,img.size[0]),]
+    return find_coeffs(input_points, output_points)
+
+def random_transformation(img, p_range=PERSPECTIVE_RANGE, 
+                          r_range=ROTATION_RANGE, s_range=SCALE_RANGE):
+    p_change = np.random.random()*(np.diff(p_range)[0])+p_range[0]
+    r_change = np.random.random()*(np.diff(r_range)[0])+r_range[0]
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/misc/paper.py

@@ -1,0 +1,206 @@
+from util.math import transpose
+from util.decorators import type_check
+
+TABLE_SEPERATOR = "\\\\\n    "
+# TABLE_TEXT%(<column_layout>, <num_columns>, <contents>)
+TABLE_TEXT = """\\begin{table}
+  \centering
+  \\begin{tabular}{%s}
+    \hline
+    %s
+    \hline
+  \end{tabular}
+  \caption{}
+  \label{table:}
+\end{table}"""
+TABLE_MULTI_COL = "\multicolumn{%i}{%s}{%s}"
+N_COLS = lambda n, side: side+"|".join(["c"]*n)+side
+
+
+# Given a float 'value' and a format 'fmt' (with 'e' in it), generate
+# a scientific notation string, but simplify numbers with exponents
+# whose absolute value is less than or equal to 'min_exp'.
+def simplify_scientific(value, fmt='.2e', min_exp=2):
+    v, e = f"{value:{fmt}}".split("e")
+    if "." in v: digits = len(v) - v.index(".") - 1
+    else:        digits = v
+    e = int(e)
+    # If we are simplifying some sig figs, fix them.
+    if (abs(e) <= min_exp):
+        v = list(v)
+        # (numerically stable) expansion of exponential form
+        if   (e < 0): v = ['0',v.pop(1)] + ['0']*(-e-1) + v
+        elif (e > 0): v = [v[0]]+v[2:2+e] + ['0']*(e-digits) + ['.'] + v[2+e:]
+        v, e = ''.join(v + (['0'] if v[-1] == '.' else [])), 0
+    string = f"{v} \\times 10^{{{e}}}"
+    if (min_exp > 0): string = string.replace(" \\times 10^{0}", "")\
+                                     .replace("10^{1}","10")
+    return string
+
+# Generate the text for a latex table block from a two dimensional
+# python list. Emphasize numeric column values based on their sorted
+# index and provided "wrap_" keyword arguments.
+# 
+#   "py_list"        -> A list of lists containing table rows.
+# 
+#   "header"         -> A row that will be placed above the table with
+#                       an additional \hline separating it from contents.
+# 
+#   "fmt"            -> A format string for the numeric values.
+# 
+#   "wrap_nums"      -> A tuple of (before,after) strings to wrap
+#                       around numbers, defaults to math mode dollar.
+# 
+#   "side_bars"      -> True if table should have outer-edge side bars.
+# 
+#   "simplify_sig"   -> The maximum absolute value exponent that will
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/optimize/line.py

@@ -1,0 +1,206 @@
+
+# Perform a binary search on a line over a boolean objective function.
+# Find the point at which the the function switches outputs.
+# 
+# INPUTS:
+#   f -- A boolean function of the form f(x) = (x <= z) for (l <= z < u).
+#   l -- The lower bound for the range to search, f(l) = True.
+#   u -- The upper bound for the range to search, f(u) = False.
+# 
+# OPTIONAL INPUTS:
+#   accuracy  -- The maximum allowable distance from the correct solution.
+#   round     -- A rounding function to set the values of l and u.
+#   max_steps -- The maximum number of steps allowed in the search.
+# 
+# RETURNS:
+#   x -- A point within `accuracy` of the position where f transitions
+#        from True to False.
+#    
+def binary(f, l, u, accuracy=0, round=lambda x: x, max_steps=1000):
+    # Verify the inputs provided are valid.
+    assert(f(l) == True)
+    assert(f(u) == False)
+    assert(l <= u)
+    assert(accuracy > 0)
+    # Perform the binary search until we reach desired accuracy.
+    step = 1
+    while (abs(u - l) > accuracy) and (step <= max_steps):
+        step += 1
+        mid = round(l/2 + u/2)
+        # If we've reached the limit of achievable accuracy, return.
+        if (mid == l) or (mid == u): return mid
+        # Transition the search window.
+        if f(mid): l = mid
+        else:      u = mid
+    # Return the midpoint of the upper and lower bounds.
+    return round(l/2 + u/2)
+
+
+
+# Perform a Golden section search on a line over a convex 1D
+# objective function. Find the minimum point of the function.
+# 
+# INPUTS:
+#   f -- A real valued function with a single minimum on (l,u).
+#   l -- The lower bound for the range to search.
+#   u -- The upper bound for the range to search.
+# 
+# OPTIONAL INPUTS:
+#   accuracy  -- The maximum allowable distance from the correct solution.
+#   round     -- A rounding function to set the values of l and u.
+#   max_steps -- The maximum number of steps allowed for the algorithm.
+# 
+# RETURNS:
+#   x -- A point within `accuracy` of the position that minimizes f.
+#    
+def golden_section(f, l, u, accuracy=0, round=lambda x: x, max_steps=1000):
+    # Verify the inputs provided are valid.
+    assert(l <= u)
+    assert(accuracy >= 0)
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/stats/rank.py

@@ -1,0 +1,165 @@
+from util.stats import *
+
+import os, pickle, time
+from itertools import combinations
+
+# Pre:  "indexable" is a list type, "value" is an indexable whose
+#       elements have comparison operators, "key" is a function to use
+#       for sorting
+# Post: A binary search over the indexable is performed and "value" is
+#       inserted in a sorted order left of anything with equal key,
+#       "key" defaults to sorting by direct element value comparison.
+def insert_sorted(indexable, value, key=lambda i:i):
+    low = 0
+    high = len(indexable)
+    index = (low + high) // 2
+    while high != low:
+        if key(indexable[index]) >= key(value):   high = index
+        elif key(indexable[index]) < key(value):  low = index + 1
+        index = (low + high) // 2
+    indexable.insert(index, value)
+
+# Pre:  "numbers" is a python list of numbers
+# Post: All of the values in "numbers" multiplied together
+def product(numbers):
+    value = 1
+    for n in numbers:
+        value *= n
+    return value
+
+# Pre:  "indexable" is an object that can be index accessed
+#       "indexable" must be in sorted order
+#       "value" is a value that can be compared with values in "indexable"
+# Post: The number of elements in "indexable" that are greater than "value"
+def count_greater(indexable, value):
+    low = 0
+    high = len(indexable)
+    index = (low + high) // 2    
+    while high != low:
+        if   indexable[index] >  value:
+            high = index
+        elif indexable[index] <= value:  
+            low  = index + 1
+        else:
+            string = "High: %i\tLow: %i\tIndex: %i\n"%(high,low,index)
+            string += "Indexable:\n%s"%str(indexable)
+            raise(Exception("Encountered invalid number in "+
+                            "comparison.\n\n"+string))
+        index = (low + high) // 2
+    return len(indexable) - index
+
+# Pre:  "interest" is an iterable of length > 0
+#       "other_data" is a list of iterables each of length > 0
+#       "rank" is a 0-indexed position in an ordering, 0'th being best
+#       "order" is 'True' if order matters for data set values.
+# Post: The probability that if random single elements are selected
+#       from each of the sets in "other_data" and "interest", that the
+#       value selected from "interest" will be in position "rank" from
+#       the smallest value
+def rank_probability(interest, other_data, rank=0, order=False):
+    # Check basic assumptions that should hold at the start of this function.
+    assert(len(interest) > 0)
+    assert(len(other_data) > 0)
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/stats/samples.py

@@ -1,0 +1,152 @@
+from util.stats import *
+
+# =====================================================
+#      Approximating the desired number of samples     
+# =====================================================
+
+
+# Compute the confidence of "num_samples" samples having less than or
+# equal to "max_error" at exactly 1/2. This is achieved by factoring
+# the expression (sum_{i=j}^k choose(n,i) / 2^n) to reduce operations.
+def _half_confidence(num_samples, max_error):
+    from util.math import Fraction, choose
+    # Calculate the maximum number of steps away from the mid-point
+    # that we can take *before* violating the error condition.
+    odd = num_samples % 2
+    start = (num_samples+1) // 2
+    #         initial_error + steps * step_error <= max_error
+    #   (max_error - initial_error) / step_error >= steps
+    #  (max_error - initial_error) * num_samples >= steps
+    # knowing that                    step_error  = 1 / num_samples
+    min_error = odd * Fraction(1, (2*num_samples))    
+    steps = int((max_error - min_error) * num_samples)
+    # Put steps into allowable bounds.
+    steps = max(0, min(num_samples - start, steps))
+    # Handle two cases where there is no functioning confidence bound.
+    if (odd) and (max_error < min_error): return Fraction()
+    # Compute the fraction.
+    numerator   = 1
+    denominator = 2**(num_samples)
+    # First compute the constant multiple outside of the sum.
+    for i in range(start+steps+1, num_samples+1): numerator   *= i
+    for i in range(2,             start+1):       denominator *= i
+    # Compute the sum of the inner parts of the distribution.
+    total = 0
+    for i in range(start, start+steps+1):
+        v = 1
+        for j in range(i+1,             start+steps+1): v *= j
+        for j in range(1+num_samples-i, start+1):       v *= j
+        if (v > 1): total += v * (2 if ((i != start) or odd) else 1)
+    # Perform the final numerator update.
+    if (total > 0): numerator *= total
+    # Return the (reduced) fraction.
+    return Fraction(numerator, denominator)
+
+
+# Given a list of numbers, return True if the given values provide an
+# estimate to the underlying distribution with confidence bounded error.
+def samples(size=None, error=None, confidence=None, at=None):
+    # Determine what to calculate based on what was provided.
+    from util.math import choose, Fraction
+    if   (size is None):       to_calculate = "samples"
+    elif (error is None):      to_calculate = "error"
+    elif (confidence is None): to_calculate = "confidence"
+    else:                      to_calculate = "verify"
+    # Default evaluation point is at (1/2), where the error is greatest.
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/response_1.txt
Updated IEEESE paper with revisions suggested by Dr Hong. Added backwards compatible 'util' subdirectory to the NUMA paper to begin working on revisions.

Response to Reviewers

Thank you all for the thoughtful reviews and comments on the paper. We
have revised the paper and believe the revisions address all concerns.
The major endowed improvements include:

  (1) a new proof for lemma 3 that results in a sharp bound on the
      error of the gradient approximation (Section 5, page 10),

  (2) a new example that demonstrates the sharpness of the theoretical
      error bound (Section 5, page 11),

  (3) a new subsection illustrating the modeling accuracy of a
      regression technique and an interpolation technique on a known
      analytic test function when given an increasing number of
      well spaced approximation points (Section 5.1, pages 12 and 13),

  (4) a radial basis function kernel support vector regressor (SVR)
      was substituted in place of the polynomial kernel SVR,

  (5) a new minor Section in the appendix defines relevant statistical
      terminology used throughout the paper (Appendix, page 25).

Responses to individual requests / comments by reviewers follows.

  * missing regression based on radial basis functions

Addressed by (4), updated description in last paragraph of Section 2.3
(page 4) to coincide with (4).

  * missing regression / interpolation based on sparse grids

The box spline method described in Section 3.4 uses a sparse grid to
construct its approximation (described in detail in reference [33]).
The phrase "sparse grid" was added to the first sentence of 3.4 to
make this more clear.

  * missing interpolation over [approximate Fekete points]

Addressed by (3), the "well spaced" points used for model construction
are approximate Fekete points. See Section 5.1 paragraph 3, page 12.

  * missing numerical evidence about the error bound from the theorem
    in Section 5, is it sharp?

Addressed by (2), Section 5 page 11 paragraph 2. An example shows that
(with modified lemma 3, per (1)), the error bound is sharp.

  * would expect some convergence plot showing error decreasing with
    more points

Included in (3), Section 5.1 Figures 2 and 3, pages 12 and 13.

  * missing a comparison on simulated data
  * recommend tests approximating functions with analytic expression
    (refer to Barthelmann et al. 2000)

Included in (3), Section 5.1 pages 12 through 14. Used a modified
version of the "Oscillatory" test function from provided reference.
The original dot product structure of the test function strongly
resembles the internal structure of the multilayer perceptron. The
slight modification was made to reduce bias in the comparison.

  * should include comparison on simulated data both with and without
    randomness incorporated into the data

Addressed by (3), where the left and right halves of Figures 2 and 3
in Section 5.1 (pages 12 and 13) now depict the effect randomness has
on approximation error.

  * in case the models are not approximating CDFs, alternatives to the
    KS statistic should be discussed

A sentence was added to the last paragraph of Section 7 that mentions
the quantification of error with alternative measures.

  * right now the box-plots visualize the KS statistic, not the KS
    measure

The KS statistic is the max norm difference (the KS measure). This was
clarified in the caption of Figure 13 (page 19) and the second
paragraph of Section 6.5 (page 19). We refer to the visual in Figure 1
for further clarification, which illustrates that the KS statistic is
the max norm difference between two cumulative distribution functions.

  * readers may have a weaker background in statistics, terms like
    null hypothesis, p-value, and box-plots cannot be given for
    granted

Addressed by (5), which is also now referred to in the last sentence
of Section 4 (page 8).

  * statistical discussion in the Figures should be moved to main text

This suggestion is appreciated. Revision (5) was preferred for fear
that removing important definitions from Figure text may reduce
comprehension for less thorough readers of the paper.

  * state clearer in the introduction which are the main objectives
    and contributions of this work

The second to last paragraph of Section 1 (page 3) has been modified
to clearly state both the objectives and contributions of this work.

### 2020-02-13 [2019-12]_IEEESE_(quintic_applied)/paper.tex
Updated IEEESE paper with revisions suggested by Dr Hong. Added backwards compatible 'util' subdirectory to the NUMA paper to begin working on revisions.

@@
 
-%% \author{Thomas C. H. Lux$^{1}$, Layne T. Watson$^{2,5}$, Tyler H. Chang$^{1}$, Li Xu$^{3}$, Yueyao Wang$^{3}$,\\ Jon Bernard$^{1}$, Yili Hong$^{4}$, Kirk W. Cameron$^{2}$% <-this % stops a space
-%% \thanks{*This work was supported by the National Science Foundation Grant CNS-1565314.}%
-%% \thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
-%% \thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
-%% \thanks{$^{3}$Doctoral candidate, Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
-%% \thanks{$^{4}$Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
-%% \thanks{$^{5}$Departments of Mathematics and Aerospace Engineering, Virginia Polytechnic Institute and State University, Virginia 24060, USA}%
-%% }
+\author{Thomas C. H. Lux$^{1}$, Layne T. Watson$^{2,5}$, Tyler H. Chang$^{1}$, Li Xu$^{3}$, Yueyao Wang$^{3}$,\\ Jon Bernard$^{1}$, Yili Hong$^{4}$, Kirk W. Cameron$^{2}$% <-this % stops a space
+\thanks{*This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.}%
+\thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
+\thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
+\thanks{$^{3}$Doctoral candidate, Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
+\thanks{$^{4}$Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
+\thanks{$^{5}$Departments of Mathematics and Aerospace Engineering, Virginia Polytechnic Institute and State University, Virginia 24060, USA}%
+}
 
@@
 This work presents a case study on the effectiveness of various distribution approximations.
-Results show that piecewise linear approximations produce the smallest maximum absolute error, while the classic empirical distribution function (EDF) produces the smallest median absolute error as well as the smallest absolute error at lower percentiles when approximating a distribution from a sample.
+Results show that piecewise linear approximations produce the smallest maximum absolute error, while the classic empirical distribution function (EDF) produces the smallest median absolute error as well as the smallest first quartile and minimum absolute error when approximating a distribution from a sample.
 When building distribution prediction models, the piecewise quintic and cubic approximations produce the lowest absolute error at most error percentiles.
-These initial results encourage more research on the best methods of fitting empirical data with smooth functions to generate accurate distribution approximations.
+This case study encourages more research on the best methods of fitting empirical data with smooth functions to generate accurate distribution approximations.
 
@@
 A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a desirable property many standard parametric distributions maintain.
+In effect, an improved distribution estimate can increase the accuracy of distribution predicting models \cite{xu2020modeling}.
 
@@
 Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}.
-Recently an algorithm for the construction of monotone quintic splines has been produced as well (citation pending).%% \cite{lux2020algorithm}
+Recently an algorithm for the construction of monotone quintic splines has been produced as well \cite{lux2020algorithm}.
 This work considers the EDF, a linear interpolant, a monotone cubic spline interpolant, and a monotone quintic spline interpolant as candidate approximations.
@@
 
+\begin{figure}
+  \vspace{-.3cm}
+  \includegraphics[width=.5\textwidth]{throughput-histogram.pdf}
+  \caption{A histogram of all throughput values across the 17 unique system configurations and 170 thousand executions of a ``readers'' test in IOzone.}
+  \label{fig:throughput_histogram}
+\end{figure}
+
 The performance of approximation techniques that predict probability functions can be analyzed through a variety of summary statistics.
@@
 The Delaunay method is used to select the five source distributions as well as provide the convex weights associated with each of the source distributions that will most likely produce the predicted distribution.
-
-\begin{figure}
-  \vspace{-.3cm}
-  \includegraphics[width=.5\textwidth]{throughput-histogram.pdf}
-  \caption{A histogram of all throughput values across the 17 unique system configurations and 170 thousand executions of a ``readers'' test in IOzone.}
-  \label{fig:throughput_histogram}
-\end{figure}
 
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
+    \node (img)  {\includegraphics[width=0.4\textwidth,]{example_distribution_prediction.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
@@
   \caption{In this example, the general methodology for predicting a CDF and evaluating error can be seen.
-The Delaunay method chose three source distributions (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom at arrow).
+The Delaunay method chose three source distributions (dotted lines) and assigned weights \{.1, .3, .6\} (top to bottom at middle).
 The weighted sum of the three known CDFs produces the predicted CDF (dashed line).
-The KS Statistic (arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example.
+The KS Statistic (vertical line) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example.
 The KS test null hypothesis is rejected by $p$-value 0.01, however it is not rejected by $p$-value 0.001.
@@
   \caption{The result of using Delaunay to predict distributions.
-Notice that the best performing, lowest error (to the left), approximation is the EDF.
-Interestingly, the piecewise linear fit does not produce results as accurate as the EDF.
-This likely has to do with the inherent smoothing that happens when multiple different EDFs are combined to make one prediction.
+Notice that the best performing, lowest error (to the left), approximations at almost all absolute error percentiles are the cubic and quintic interpolants, closely followed by the linear interpolant.
+Interestingly, these results are contrary to pure distribution approximations from a sample.
+This likely has to do with the inherent smoothing that happens when multiple different approximations are combined to make one prediction.
   \vspace{-.5cm}}

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/approximate/delaunay/test.py

@@ -1,0 +1,71 @@
+if __name__ == "__main__":
+    import numpy as np
+    from util.plot import Plot
+    from util.approximate import Delaunay
+
+    # Settings for creating the example.
+    N           = 30
+    dim         = 2
+    random      = True
+    plot_points = 1000
+
+    # Function for testing.
+    mult = 5
+    low  = 0
+    upp  = 1
+    fun = lambda x: np.cos(x[0]*mult) + np.sum(np.sin(x[1:]*mult))
+    np.random.seed(0)
+    # Generate the X and Y for the points.
+    if random:
+        x = np.random.random(size=(N,dim))
+    else:
+        width = int(round(N ** (1/dim)))
+        x = np.array([r.flatten() for r in np.meshgrid(
+            * ((np.linspace(low,upp,width),)*dim) )]).T
+        x = x[:N]
+
+    y = np.array([fun(v) for v in x])
+
+    # Shift to add padding
+    padding = .4
+    low -= (upp - low)*padding
+    upp += (upp - low)*padding
+
+
+    # PMODE=1, once with PMODE=2, once with PMODE=3, and once with PMODE unspecified and CHUNKSIZE=10.
+    # Fit the Delaunay model to the points
+    surf = Delaunay(pmode=None)# , chunksize=100
+    surf.fit(x,y)
+
+
+    # Manual test at a specific point.
+    # test = np.array([[.3587, .422]])
+    test = np.array([[.25, .25]])
+
+    # # Parallel surface (wrong)
+    # guess_surf = Delaunay(parallel=True)
+    # guess_surf.fit(x,y)
+    # guess_pts, guess_wts = guess_surf._predict(test.copy())
+    # # Serial surface (right)
+    # true_surf = Delaunay(parallel=False)
+    # true_surf.fit(x,y)
+    # true_pts, true_wts = true_surf._predict(test.copy())
+    # # Display results
+    # print()
+    # print("True points: ", true_pts)
+    # print("    weights: ", true_wts)
+    # print("      value: ", np.sum(y[true_pts] * true_wts))
+    # print()
+    # print("Guess points:", guess_pts)
+    # print("     weights:", guess_wts)
+    # print("      value: ", np.sum(y[guess_pts] * guess_wts))
+    # print()
+    # exit()
+
+    # Create the surface in the plot (calling Delaunay).
+    p = Plot()
+    p.add("Training Points", *x.T, y)
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/data/auxiliary.py

@@ -1,0 +1,1 @@
+internals.py

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/data/data.py

@@ -1,0 +1,2109 @@
+import os, sys
+from util.data.read import read_data
+from util.data.categories import regular_simplex
+from util.data import DEFAULT_DISPLAY_WAIT, GENERATOR_TYPE, NP_TYPES, \
+    SEPARATORS, FILE_SAMPLE_SIZE, MISSING_SAMPLE_SIZE
+
+# TODO:  Instead of just checking file size, check number of columns too,
+#        use a sample by default if the #columns is large (speed!).
+# TODO:  Warnings or versioning for when the source data is modified
+#        in a way that invalidates a view. Need to track unpropagated
+#        changes somehow. Or make permanent changes disallowed? Or copy?
+#        E.g. view is made, columns are rearranged
+# TODO:  Make "to_matrix" automatically consider 'None' a unique value
+#        for categorical columns and add a dimension 'is present' otherwise.
+# TODO:  Make "to_matrix" automatically flatten elements of columns
+#        that contain iterables into their base types. If the types
+#        are numeric, use float, otherwise use string. Use "is numeric".
+# TODO:  Check for duplicate column names, handle duplicate column
+#        names? Do names *have* to be Python strings?
+# TODO:  Support in-place addition with *some* duplicate solumns and
+#        the same number of rows? That or block duplicate col names.
+# 
+# 
+# TODO:  Remove builtin prediction code for anything other than
+#        nearest neighbor prediction. Maybe remove prediction in
+#        general, since it isn't something that can be picked beforehand.
+# TODO:  Running 'predict' with nearest neighbor seg-faults after 4th
+#        round, might involve conditioning, probably involves ball tree.
+# TODO:  'predict' does not work with a view, picks wrong column, it
+#        also appears to pick the wrong column even when not a view.
+# TODO:  Rewrite "predict" to work given a data object. This will
+#        break up the data into collections of unique missing values.
+#        Then for each collection, build a model over the appropriate
+#        matrix-format data and predict that collection. This will
+#        work more nicely with the k-fold method.
+# TODO:  Make "predict" not require that target columns be convertible
+#        to numeric form, only that they operate under a weighted sum.
+# TODO:  Remove the "fill" method in favor of "predict".
+# 
+# 
+# TODO:  Implement "parallel read" that create subprocesses that read
+#        chunks of the file, communicate progress, communicate type
+#        digressions, saves chunks in .pkl, then organizes them all
+#        into the expected ordered data format.
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/data/exceptions.py

@@ -1,0 +1,17 @@
+
+# Local descriptive exceptions
+class NameTypeMismatch(Exception): pass
+class BadSpecifiedType(Exception): pass
+class BadSpecifiedName(Exception): pass
+class NoNamesSpecified(Exception): pass
+class ImproperUsage(Exception): pass
+class MissingModule(Exception): pass
+class UnknownName(Exception): pass
+class Unsupported(Exception): pass
+class BadAssignment(Exception): pass
+class BadElement(Exception): pass
+class BadTarget(Exception): pass
+class BadValue(Exception): pass
+class BadIndex(Exception): pass
+class BadData(Exception): pass
+class Empty(Exception): pass

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/development/bootstrapped_box_splines/bbs_test.py

@@ -1,0 +1,132 @@
+# See why and where the prediction error is so large
+#  - check 4D data first, it's bad in some spots
+#  - verify the convex hull?
+#  - look at plots of th 1D predictions, sanity check
+
+import numpy as np
+import time
+from util.plotly import *
+from util.approximate import BBS, FitBoxMesh, MaxBoxMesh
+
+FUN = lambda x: np.cos(x[0]) * (np.sin(x[1]) if len(x) > 1 else 1.0) + 1
+FBM_NUM_BOXES = float("inf")
+BBS_NUM_BOXES = float("inf")
+BBS_BATCH_SIZE = 4
+
+EXTRA_ARGS = {
+    "FitBoxMesh": [FBM_NUM_BOXES],
+    "BBS": [BBS_NUM_BOXES, BBS_BATCH_SIZE]
+}
+PLOT_POINTS = 1000
+
+# Preparing the data
+normalize_points = False
+random_points = False
+random_perturbation = False
+tight_plot = False
+plot_results = True
+
+perturbation_multiplier = 1.0
+num_points = 2
+dim = 2
+multiplier = 10
+
+algorithms = [BBS()]
+# algorithms = [FitBoxMesh()]
+# algorithms = [MaxBoxMesh()]
+
+# BOXES:   [[ 0.   0. ]
+#           [ 0.   0.1]
+#           [ 0.1  0. ]
+#           [ 0.1  0.1]]
+# WIDTHS:  [ 0.1  0.1  0.1  0.1]
+# WEIGHTS: [ 1.          1.09983342  1.          1.09933467]
+
+np.random.seed(0)
+
+# Generate testing points in a grid pattern
+plot_range = [[-0.2,1.2]]*dim # [[0,1]]*dim # 
+if random_points:
+    # Generate testing points randomly spaced
+    points = np.random.random(size=(num_points**dim,dim)) * (num_points-1)*num_points
+else:
+    points = np.meshgrid(*[np.linspace(0,num_points-1,num_points) for i in range(dim)])
+    points = np.array([p.flatten() for p in points]).T * num_points
+
+if random_perturbation:
+    points += np.random.random(size=points.shape)*perturbation_multiplier
+
+points *= 0.05
+# Calculate the associated response values
+values = np.array([[FUN(pt) for pt in points]]).T
+points = np.concatenate((points, values), axis=1)
+# Sort the points so they resemble grid order
+points = points[points[:,1].argsort()]
+points = points[points[:,0].argsort()]
+
+print()
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/development/ml.py

@@ -1,0 +1,30 @@
+from util.approximate import VoronoiMesh
+
+# Create an instance of an algorithm that can be ued to classify
+# data. The "fit" method causes the classifier to construct the
+# necessary regression models to predict each class.
+class Classifier:
+    def __init__(self, alg=VoronoiMesh):
+        self.alg = alg
+        self.models = []
+
+    # Given x (matrix with instance rows and feature columns) and y
+    # (class labels), optionally a stratified subsample size (the
+    # number of random subsets to model for each class), and
+    # optionally use feature weighting for fitting, generate a model
+    # that can use the internal algorithm to make predictions at
+    # previously unseen x points.
+    def fit(self, x, y, stratification=None, feature_weighting=False):
+        self.classes = sorted(set(y))
+        # Build a model for each class
+        for value in self.classes:
+            response = np.where(y == value, 1, 0)
+            self.models.append( self.alg() )
+            self.models[-1].fit(x, response)
+
+    # Given a new x (or set of x), calculate the value of each class
+    # model for that x and return the (set of) predicted classes.
+    def predict(self, x):
+        classes = [self.classes[idx] for idx in
+                   np.argmax([m(x) for m in self.models], axis=0)]
+        return np.array(classes)

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/development/testing/test_system.py

@@ -1,0 +1,79 @@
+from util.system import *
+
+def test_Timer():
+    print("Testing Timer..", end=" ")
+    import time
+    from util.system import Timer
+    a = Timer()
+    first = a.start
+    assert(a.total() < .01)
+    time.sleep(1)
+    assert(first - a.start == 0)
+    assert(abs(a.total() - 1) < .01)
+    a.start()
+    assert(abs(first - a.start) > 1)
+    assert(a.total() < .001)    
+    a.stop()
+    print("passed.")
+
+
+def test_AtomicOpen(display=False):
+    print("Testing AtomicOpen..", end=" ")
+    import time, os
+    from util.parallel import map
+    from util.system import AtomicOpen, Timer
+
+    # Testing without atomic writes.
+    def write_test(string="testing..", atomic=False):
+        open_file = AtomicOpen if atomic else open
+        with open_file(test_file, "r") as f:
+            if display: print(string, "file opened")
+            if display: print(string, f.read().strip())
+            # print(string, file=f)
+            time.sleep(2)
+        if display: print(string, "file closed")
+    # Testing for atomic writes.
+    def atomic_write_test(string): return write_test(string, atomic=True)
+
+    p = 10
+    import util.parallel
+    util.parallel.MAX_PROCS = 50
+
+    t = Timer()
+    if display: print("Testing parallel regular write..")
+
+    if display: print()
+    test_file = "_test_.txt"
+    with open("_test_.txt", "w") as f:
+        print("'conents'", file=f)
+
+    list(map(write_test, map(str,range(p)), redirect=False))
+    if display:
+        with open(test_file) as f: print(f'{"-"*70}\n{f.read()}{"-"*70}')
+        print(t())
+    # assert( t() < 4 )
+
+    t.start()
+    if display: print("Testing parallel atomic write..")
+
+    if display: print()
+    test_file = "_test_.txt"
+    with open("_test_.txt", "w") as f:
+        print("'conents'", file=f)
+
+    list(map(atomic_write_test, map(str,range(p)), redirect=False))
+    if display:
+        with open(test_file) as f: print(f'{"-"*70}\n{f.read()}{"-"*70}')
+        print(t())
+    # assert( t() > 6 )
+
+    os.remove(test_file)
+    print("passed.")
+
+
+
+def test():
+    print("Testing 'util.system'..", end=" ")
+    test_Timer()
+    test_AtomicOpen(True)
+    print("passed.")

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/math/__init__.py

@@ -1,0 +1,126 @@
+# Make the "Fraction" class available here.
+from util.math.fraction import Fraction
+from util.math.points import mesh, chebyshev, polynomials, \
+    polynomial_indices, fekete_indices, fekete_points
+from util.math.pairs import pair_to_index, index_to_pair, \
+    num_from_pairs, pairwise_distance
+
+# Useful for checking near-equal cases when roundoff error is involved.
+SMALL = 2**(-26)
+#       ^^ SQRT(EPSILON( 0.0_REAL64 ))
+
+# Given a list of lists, flatten it to one dimension.
+def flatten(l): return [v for row in l for v in row]
+
+# Given a list of lists, transpose it and return.
+def transpose(l): return [list(row) for row in zip(*l)]
+
+# Return a boolean "is_numeric"
+def is_numeric(obj):
+    try:
+        abs((.3*obj + 1*obj) - .3*obj)
+        return True
+    except: return False
+
+# Function for performing absolute difference between numbers (or
+# vectors). Falls back to equality for things that don't support
+# difference, absolute value, or sums.
+def abs_diff(v1, v2):
+    if hasattr(v1, "__iter__"):
+        try: return sum(abs(v1 - v2))
+        except: return int(v1 == v2)
+    try:    return abs(v1 - v2)
+    except: return int(v1 == v2)
+
+# Function for computing the product of a list of numbers.
+def product(sequence):
+    v = 1
+    for number in sequence: v *= number
+    return v
+
+# Compute the probability that 1 specific value is picked from a
+# population of size `N` given a set number of `samples`.
+def pick(samples, N):
+    return 1 - product( ((N-i-1)/(N-i) for i in range(samples)) )
+
+# Return True if a number is prime, False otherwise.
+def is_prime(n):
+    for i in range(2,int(n**(1/2))+1):
+        if (not (n%i)): return False
+    return True
+
+# Return all unique prime factors of a number in sorted order.
+def primes(n):
+    factors = {}
+    candidate = 2
+    while candidate**2 <= n:
+        while (n % candidate) == 0:
+            factors[candidate] = factors.get(candidate,0) + 1
+            n //= candidate
+        candidate += 1
+    # Store the last remainder if no more primes factors were found.
+    if (n > 1): factors[n] = factors.get(n,0) + 1
+    # Sort all prime factors by their index.
+    return sorted((k,factors[k]) for k in factors)
+
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/misc/apriori.py

@@ -1,0 +1,198 @@
+import os, time, pickle  # System libraries
+
+# Item tree building constants
+MAX_ITEMS_PER_LAYER = 100    # Max number of items kept per layer
+MIN_SUPPORT = 0.0
+DEFAULT_ORDER = False
+
+def DEFAULT_OUT(*args, **kwargs):
+    kwargs.update({"flush":True}) # Default output forces a "flush"
+    print(*args, **kwargs)
+ 
+def _check_consistency(data):
+    if (type(data) != list or type(data[0]) != list):
+        raise(TypeError("Data given must be a 2D python list."))
+
+
+# This class will generate an associative item list, given minimum
+# support and confidence in a item. Currently only implements support
+class AprioriTree:
+    # "output" should be a function that takes a string as an argument
+    # and has an argument named "end" that declares the line-end behavior
+    def __init__(self, support=MIN_SUPPORT, output=DEFAULT_OUT,
+                 ordered=DEFAULT_ORDER, max_items=MAX_ITEMS_PER_LAYER):
+        self.data = []
+        self.support = support
+        self.ordered = ordered
+        self.max_items = max_items
+        if ordered: 
+            raise(Exception("No ordered implementation available yet. Consider making all values tuples (<column index>, value)."))
+        self.current_items = {}
+        self.base_items = {}
+        self.base_file_name = "ap_tree_items[%i].pkl"
+        self.file_number = 0
+        self.print = output # Function used for output
+        self.item_sets = [] # Where the final results of mining are stored
+
+
+    ######## BEGIN ITEM MINING CODE #########
+
+    # Pre:  "self.data" has been set and checked for consistency
+    # Post: self.item_sets is established with length 1 items
+    def _initialize_items(self):
+        self.base_items = {}
+        for row in self.data:
+            for val in row:
+                if (val,) not in self.base_items: 
+                    self.base_items[(val,)] = 0
+        self.current_items = self.base_items
+        
+
+    # Pre:  "new_items" is a dictionary of items
+    # Post: "self.current_items" are written to file and the new items
+    #       are transitioned into their place
+    def _transition_items(self, new_items):
+        self.file_number += 1
+        with open(self.base_file_name%(self.file_number), "wb") as f:
+            pickle.dump(self.current_items, f)
+        self.current_items = new_items
+
+
+    # Pre:  "self.current_items" and "self.base_items" are populated
+    #       dictionaries 
+    # Post: All the combinatorial possible combinations of the most
[…]

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/util/parallel/parallel.py

@@ -1,0 +1,231 @@
+# TODO: When a function called by a worker uses "exit()", 'None' is
+#       printed and the process remains, sitting there. If all
+#       processes exit, then the master should exit too. A message
+#       should be printed saying that a subprocess exited, though.
+
+# Get functions required for establishing multiprocessing
+from multiprocessing import cpu_count, get_context, current_process
+# Imports that are used by the processes.
+from queue import Empty
+import os
+# Try to import "dill" module for better serialization, otherwise use pickle.
+try:                        from dill   import dumps, loads
+except ModuleNotFoundError: from pickle import dumps, loads
+
+# Make sure new processes started are minimal (not complete copies)
+MAX_PROCS = cpu_count()
+LOG_DIRECTORY = os.path.abspath(os.path.curdir)
+JOB_GET_TIMEOUT = 5
+# Global storage for processes
+MAP_PROCESSES = []
+# Get a reference to the builtin "map" function, so it is not lost
+# when it is overwrittend by "def map" later in this file.
+builtin_map = map
+
+# Iterator that pulls jobs from a queue and stops iterating once the
+# queue is empty (and remains empty for "JOB_GET_TIMEOUT" seconds). On
+# construction, requires a "job_queue" that jobs will come from.
+class JobIterator:
+    def __init__(self, job_queue):
+        self.job_queue = job_queue
+    def __iter__(self): return self
+    def __next__(self):
+        try:          value = self.job_queue.get(timeout=JOB_GET_TIMEOUT)
+        except Empty: raise(StopIteration)
+        return value
+
+# Given an iterable object and a queue, place jobs into the queue.
+def producer(jobs_iter, job_queue):
+    for job in enumerate(jobs_iter):
+        job_queue.put(job)
+
+# Given an iterable object, a queue to place return values, and a
+# function, apply "func" to elements of "iterable" and put return
+# values into "return_queue".
+# 
+# INPUTS:
+#   func         -- A function that takes elements of "iterable" as
+#                   input. It is expected that the function has been
+#                   packaged with with "dill.dumps" to handle lambda's.
+#   iterable     -- An iterable object (presumably a JobIterator).
+#   return_queue -- A Queue object with a 'put' method.
+#   redirect     -- Boolean, True if log files should be created per
+#                   process, otherwise all print goes to standard out.
+def consumer(func, iterable, return_queue, redirect):
+    import sys
[…]

### 2020-02-13 [2019-12]_IEEESE_(quintic_applied)/code/5-delaunay-predict.py

@@ -143,6 +143,10 @@ legend = dict(
     borderwidth=.5
 )
 
+# p.show(layout=dict(margin=dict(l=40, b=40)),
+#        width=700, height=500, legend=legend)
+
+
 p.show(file_name="delaunay-predictions.html",
        layout=dict(margin=dict(l=40, b=40)),
        width=400, height=290, legend=legend)

### 2020-02-13 [2019-12]_IEEESE_(quintic_applied)/code/fits.py

@@ -46,9 +46,9 @@ def linear_fit(data):
     return fit
 
 # Return a cubic fit function.
-def cubic_fit(data):
+def cubic_fit(data=None, x=None, y=None):
     from monotone import monotone_cubic_spline
-    x, y = cdf_points(data)
+    if (x is None) or (y is None): x, y = cdf_points(data)
     cdf = monotone_cubic_spline(x, y)
     def fit(z):
         val = max(0,min(1,cdf(z)))
@@ -59,9 +59,9 @@ def cubic_fit(data):
     return fit
 
 # Return a quintic fit function.
-def quintic_fit(data):
+def quintic_fit(data=None, x=None, y=None):
     from monotone import monotone_quintic_spline
-    x, y = cdf_points(data)
+    if (x is None) or (y is None): x, y = cdf_points(data)
     cdf = monotone_quintic_spline(x, y)
     def fit(z):
         val = max(0,min(1,cdf(z)))

### 2020-03-30 [2019-02]_NUMA_(interpolant_error_bound)/paper.tex
updated work

\RequirePackage{fix-cm}
\documentclass[smallextended,final]{svjour3}  % onecolumn (second format)
\smartqed  % flush right qed marks, e.g. at end of proof

% Insert here the call for the packages your document requires
\usepackage{graphicx}
\usepackage{tikz}     % For adding axes to plots
\usetikzlibrary{positioning}
\usepackage{graphicx} % For including PDFs
\usepackage{amsmath}  % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{booktabs} % For formal tables
\usepackage{tabularx} % For tables with nice column widths.
\usepackage[colorlinks,allcolors=blue]{hyperref}
\usepackage{float}    % For controlling the placement of tables.

%% -------------------------------------------------------------------
%% Command for changing the color of text within a block.
\usepackage{xcolor}   % Color package for coloring text.
\usepackage[normalem]{ulem} % For strikethrough text and the \sout{} command.
\usepackage[xcolor,rightbars]{changebar}% Package for adding bars into margin next to comments.
\renewcommand{\changebarsep}{5mm}   % Push the bar farther into the margin.
\renewcommand{\changebarwidth}{1mm} % Increase the width of the bar.
\cbcolor{red}
\newcommand{\changed}[2]{\cbstart{\color{red}#1}{\color{gray}\sout{#2}}\cbend{}}
%% -------------------------------------------------------------------

%
% please place your own definitions here and don't use \def but
% \newcommand{}{}

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\newcolumntype{d}[1]{D{.}{.}{#1}}

%% Create a ``Lemma'' environment that does not have italics font.
\spnewtheorem{plainlemma}{Lemma}{\bf}{}
%% Create a ``Theorem'' environment that is not numbered.
\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
%% Create a ``Proof'' environment that starts with 'Proof.'
\spnewtheorem*{proofdot}{Proof.}{\it}{}

%% \newcommand{\Bigl}{\mathopen\Big}
%% \newcommand{\Bigr}{\mathclose\Big}
%
% Insert the name of "your journal" with
\journalname{Numerical Algorithms}
%
\begin{document}

\title{Interpolation of Sparse High-Dimensional Data\thanks
  {This work was supported by the National Science Foundation Grant CNS-1565314.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}
%\titlerunning{Short form of title}        % if too long for running head

\author{Thomas C. H. Lux   \and
  Layne T. Watson          \and
  Tyler H. Chang           \and
  Yili Hong                \and
  Kirk Cameron
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Thomas Lux \at
  Virginia Polytechnic Institute \& State University (VPI\&SU) \\
  Blacksburg, VA 24061 \\
  \email{tchlux@vt.edu}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Increases in the quantity of available data have allowed all fields of
science to generate more accurate models of multivariate phenomena.
Regression and interpolation become challenging when the dimension of
data is large, especially while maintaining tractable computational
complexity. Regression is a popular approach to solving approximation
problems with high dimension, however there are often advantages to
interpolation. This paper presents a novel and insightful error bound
for (piecewise) linear interpolation in arbitrary dimension and
contrasts the performance of some interpolation techniques with
popular regression techniques. Empirical results demonstrate the
viability of interpolation for moderately high dimensional
approximation problems, and encourage broader application of
interpolants to multivariate approximation in science.

\keywords{approximation \and regression \and interpolation \and high
  dimension \and error bound}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

%     Introduction     
%======================

%% ===================================================================
\section{Introduction}
\label{sec:introduction}

Regression and interpolation are problems of considerable importance
that find applications across many fields of science. Pollution and
air quality analysis \cite{de2008field}, energy consumption management
\cite{lazos2014optimisation}, and student performance prediction
\cite{cortez2008using,lux2016applications} are a few of many
interdisciplinary applications of multivariate regression for
predictive analysis. As discussed later, these techniques can also be
applied to prediction problems related to forest fire risk assessment
\cite{cortez2007data}, Parkinson's patient clinical evaluations
\cite{tsanas2010accurate}, local rainfall and weather
\cite{williams2009rattle}, credit card transactions
\cite{pozzolo2015calibrating}, and high performance computing (HPC)
file input/output (I/O) \cite{lux2018nonparametric}.

Regression and interpolation have a considerable theoretical base in
one dimension \cite{cheney2009course}. \changed{Common techniques
  include Lagrangian interpolation, piecewise linear interpolation,
  and more generally spline interpolation via B-splines
  \cite{de1978practical}.}{Splines in particular are well understood
  as an interpolation technique in one dimension
  \cite{de1978practical}, particularly B-splines}. Tensor products of
B-splines \cite{unther1996interpolating} or other basis functions have
an unfortunate exponential scaling with increasing
dimension. Exponential scaling prohibits tensor products from being
reasonably applied beyond three-dimensional data. In order to address
this dimensional scaling challenge, C. de Boor and others proposed box
splines \cite{de2013box}, of which one of the approximation techniques
in this work is composed \cite{lux2018novel}.

The theoretical foundation of low dimensional interpolation allows the
construction of strong error bounds that are absent from high
dimensional problems. This work extends some known results regarding
the secant method \cite{dennis1996numerical} to construct an
interpolation error bound for problems of arbitrary dimension. These
error bounds are useful, considering the same cannot be said for
regression algorithms in general. The maximum complexity of an
interpolant is bounded by the amount of data available, while the
maximum complexity of a regressor is bounded by both the amount of
data and the chosen parametric form. For this reason, generic uniform
bounds are largely unobtainable for regression techniques on arbitrary
approximation problems, even when the approximation domain is bounded.

Aside from theoretical motivation for the use of interpolants, there
are often computational advantages as well. Interpolants do not have
the need for \textit{fitting} data, or minimizing error with respect
to model parameters. In applications where the amount of data is large
and the relative number of predictions that need to be made for a
given collection of data is small, the direct application of an
interpolant is much less computationally expensive.

In this work, multivariate interpolation is defined given a closed
convex subset $Y$ of a metrizable topological vector space with metric
$s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points
$X = \bigl\{x^{(1)}$, $\ldots$, $x^{(n)}\bigr\} \subset \mathbb{R}^d$,
along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal
is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$
such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all
$i = 1$, $\ldots$, $n$. It is often the case that the form of the true
underlying function $f$ is unknown, however it is still desirable to
construct an approximation $\hat f$ with small approximation error at
$y \notin X$. The two metric spaces that will be discussed in this
work are the real numbers with metric $s(x,y) = |x-y|$, and the set of
cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
(KS) statistic \cite{lilliefors1967kolmogorov} as a metric.

Multivariate regression is often used when the underlying function is
presumed to be stochastic, or stochastic error is introduced in the
evaluation of $f$. Hence, multivariate regression relaxes the
conditions of interpolation by choosing parameters $P$ defining $\hat
f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f
\bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl
| \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$
in some norm. The difficult question in the case of regression is
often what parametric form to adopt for any given application.

The most challenging problem when scaling in dimension is that the
number of possible interactions between dimensions grows
exponentially. Quantifying all possible interactions becomes
intractable, and hence beyond three-dimensional data mostly linear
models are used. That is not to say nonlinear models are absent, but
nonlinearities are often either preconceived or model pairwise
interactions between dimensions at most. Even globally nonlinear
approximations such as neural networks are constructed from
compositions of summed low-interaction functions
\cite{clevert2015fast}.

Provided the theoretical and practical motivations for exploring
interpolants, the main objective of this work is to study the
empirical performance differences in both accuracy and computational
expense between a set of interpolation techniques and a set of common
regression techniques. The main contributions of this work are a sharp
theoretical error bound for (piecewise) linear interpolation in
arbitrary dimension and multiple empirical comparisons that illustrate
potential advantages of interpolation.

The remainder of this paper is organized as follows. Sections
\ref{sec:regression} and \ref{sec:interpolation} present the
multivariate models. Section \ref{sec:error} gives the error measuring
methodology that is used for collecting empirical results. Section
\ref{sec:theory} contains the theoretical error bounds for
interpolation as well as an analytic test for demonstration. Section
\ref{sec:data} presents the data sets for empirical approximation
analysis and the approximation results for all models on these data
sets. Section \ref{sec:discussion} analyzes and discusses the results,
their implications, and their limitations. Finally, Section
\ref{sec:conclusion} concludes.

%% ===================================================================
\section{Multivariate Regression}
\label{sec:regression}
Multivariate regressors are capable of accurately modeling a complex
dependence of a response (in $Y$) on multiple variables (represented
as a points in $\mathbb{R}^{d}$). The approximations to some (unknown)
underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to
minimize some error measure related to data samples
$f\bigl(x^{(i)}\bigr)$. For example, least squares regression uses the
sum of squared differences between modeled function values and true
function values as an error measure. In this section and the next,
some techniques are limited to approximating real valued functions ($Y
\subset \mathbb{R}$). These techniques can be extended to real
vector-valued ranges by repeating the construction for each component
of the vector output. Throughout the following, $x$ denotes a
$d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th
$d$-tuple data point. Different symbols are used to represent the
approximation function $\hat f$.

\subsection{Multivariate Adaptive Regression Splines}
\label{sec:mars}
This approximation was introduced in \cite{friedman1991multivariate}
and subsequently improved to its current version in
\cite{stanford1993fast}, called fast multivariate adaptive regression
splines (Fast MARS). In Fast MARS, a least squares fit model is
iteratively built by beginning with a single constant valued function
and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where \changed{$j \leq m$ is the iteration number, $m$ is the maximum
  number of underlying basis functions}{}, $1 \le p \le n$, and
$B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$ and
$w_- = (-w)_+$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit. In effect, MARS creates a locally component-wise tensor
product approximation of the data. The overall computational
complexity of Fast MARS is $\mathcal{O}(n d m^3)$ \changed{}{where $m$
  is the maximum number of underlying basis functions}. This paper
uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m =
200$ throughout all experiments.

\subsection{Support Vector Regressor}
\label{sec:svr}
Support vector machines are a common method used in machine learning
classification tasks that can be adapted for the purpose of regression
\cite{basak2007support}. How to build a support vector regressor (SVR)
is beyond the scope of this summary, but the resulting functional fit
$p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
 $$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
$b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with
$m$ being determined by the minimization convergence criterion. This
paper uses the scikit-learn SVR \cite{scikit-learn} with a radial
basis function kernel.

\subsection{Multilayer Perceptron Regressor}
\label{sec:mlp}
The neural network is a well studied and widely used method for both
regression and classification tasks
\cite{hornik1989multilayer,rumelhart1988learning}. When using the
rectified linear unit (ReLU) activation function
\cite{dahl2013improving} and fitting the model with stochastic
gradient descent (SGD) or BFGS minimization techniques
\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
\rightarrow \mathbb{R}^{j}$ defined by
 $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
matrix for layer $l$. In this form, the multilayer perceptron (MLP)
produces a piecewise linear model of the input data. The computational
complexity of training a multilayer perceptron is $\mathcal{O}(n d
m)$, where $m$ is determined by the sizes of the layers of the network
and the stopping criterion of the minimization used for finding
weights. This paper uses an MLP built from Keras and Tensorflow to
perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
with ten hidden layers each having thirty two nodes (a total of
approximately ten thousand parameters), ReLU activation, and
\changed{one hundred}{five} thousand \changed{steps}{epochs} of SGD
for training. \changed{It should be noted that the choice of neural
  network architecture impacts approximation performance, but no
  architecture tuning is performed here}{}.

%% ===================================================================
\section{Multivariate Interpolation}
\label{sec:interpolation}
The following interpolation techniques demonstrate a reasonable
variety of approaches to interpolation. All of the presented
interpolants produce approximations that are continuous in value,
which is often a desirable property in applied approximation problems.

\subsection{Delaunay}
\label{sec:delaunay}

The Delaunay triangulation is a well-studied geometric technique for
producing an interpolant \cite{lee1980two}. The Delaunay triangulation
of a set of data points into simplices is such that there are no data
points inside the sphere defined by the vertices of each simplex. For
a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
$\ldots$, $d$, $y \in S$ is a unique convex combination of the
vertices,
 $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
 $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$ The
computational complexity of Delaunay interpolation (for the
implementation used \cite{chang2018polynomial}) is $\mathcal{O}(k n
d^3)$. Given pathological data the entire triangulation could be
computed with $k = n^{\lceil d / 2 \rceil}$, but the average case yields
$k = d \log d$ and is capable of scaling reasonably to $d \leq 50$. In
the present application, a Delaunay simplex $S$ containing $y$ is
found, then the $d+1$ vertices (points in $X$) of $S$ are used to
assign weights to each vertex and produce the predicted function value
for point $y$.

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth,]{example_delaunay.pdf}
  \includegraphics[width=0.4\textwidth,]{example_not_delaunay.pdf}
  \caption{\color{red} On the left above is a depiction of a Delaunay
    triangulation over four points, notice that the circumball (shaded
    circle) for the left simplex does not contain the fourth point.
    On the right above, a non-Delaunay mesh is depicted. Notice that
    the circumball for the top simplex (shaded circle, clipped at
    bottom edge of the visual) contains the fourth point which
    violates the Delaunay condition for a simplex.
  \vspace{-.1cm}}
  \label{fig:delaunay}
\end{figure}
}{}

\subsection{Modified Shepard}
\label{sec:modified-shepard}

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k - \bigl\|x -
x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x - x^{(k)}\bigr\|_2 \bigr)\bigr)^2$}{}
and $r_k \in \mathbb{R}$ is the smallest radius such that at least
$d+1$ other points $x^{(j)}$, $j \not = k$, are inside the closed
Euclidean ball of radius $r_k$ about $x^{(k)}$. The computational
complexity of ShepMod is $\mathcal{O}(n^2d)$. This paper uses a
Fortran 95 implementation of ShepMod derived from SHEPPACK
\cite{thacker2010algorithm}.

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
  \caption{\color{red} On the left above is a depiction of the radius
    of influence for three chosen points of a collection using the
    modified Shepard criteria. On the right is the shape of the weight
    function $W_k$ for all data points, where $W_k(x) = 0$ when
    $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen that in the limit
    as $W_k$ goes to infinity at $x_k$, the relative weight of $x_k$
    goes to one.
  \vspace{-.1cm}}
  \label{fig:shepard}
\end{figure}
}{}

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local
linear interpolants, a special case of the general Shepard algorithm
\cite{thacker2010algorithm}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
 {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is \changed{the same as for the modified Shepard method}{a locally supported weighting function} and $P_k(x)$
is a local linear approximation to the data satisfying
$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
Fortran 95 implementation of LSHEP in SHEPPACK
\cite{thacker2010algorithm}.

\subsection{Box Splines}
\label{sec:box-splines}

The box spline model used here is \changed{an}{sparse grid}
interpolation technique built from overlapping box splines
\cite{de2013box}. The box splines serve as basis functions that are
shifted and scaled to have support over box shaped regions. The boxes
are constructed in a way to guarantee a covering of the domain
\cite{lux2018novel}. Given a set of box splines
$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
in \cite{lux2018novel} and anchored at interior points
$\bigl\{x^{(i)}\bigr\}$,
 $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy
$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
\geq 0$. The computational complexity of interpolation via the box
spline model is $\mathcal{O}(n^2 d)$.

\subsection{Voronoi}
\label{sec:voronoi}

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth,]{VM_Construction.pdf}
  \caption{\color{red} Above is a depiction of the Voronoi cell
    boundaries (dashed lines) about a set of interpolation points
    (dots). In this example, the Voronoi model basis function about
    the center most point has nonzero weight in the shaded region and
    transitions from a value of one at the point to zero at the
    boundary of the twice expanded Voronoi cell (solid line).
  \vspace{-.1cm}}
  \label{fig:voronoi-basis}
\end{figure}
}{}

The nearest neighbor algorithm \cite{cover1967nearest} is a
well-studied technique for classification and approximation. Nearest
neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
cell \cite{dirichlet1850reduction}) consisting of all points closer to
$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
the nearest neighbor approximation by utilizing the Voronoi cells to
define support via a generic basis function $v: \mathbb{R}^d
\rightarrow \mathbb{R}_+$ given by
 $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
 {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
has local support, giving the interpolated value
 $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
the approximations are an interpolant involving a convex combination
of known function values $f\bigl(x^{(i)}\bigr)$.

%% ===================================================================
\section{Measuring Error}
\label{sec:error}

\begin{figure}[htb]
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen. The Delaunay
    method chose three source distributions (dotted lines) and
    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
    weighted sum of the three known CDFs produces the predicted CDF
    (dashed line). The KS Statistic (arrow) computed between the true
    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
    example. For this example the KS test null hypothesis is rejected
    at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

When the range of an approximation is the real numbers, error is
reported with summary statistics including: min absolute error, max
absolute error, and absolute error quartiles. When the range
of an approximation is the space of cumulative distribution functions,
the Kolmogorov-Smirnov statistic (max-norm difference between the
functions) is used.

A hurdle when modeling function-valued outputs such as cumulative
distribution functions (CDFs) or probability density functions (PDFs)
is that certain properties must be maintained. It is necessary that a
PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. \changed{Instead}{However}, for a CDF
$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
nondecreasing. This work utilizes the fact that a convex combination
of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
this is applied can be seen in Figure \ref{fig:prediction-example}. 
\changed{In this example the KS test null hypothesis is rejected at $p$-value
0.01, however it is not rejected at $p$-value 0.001.}{}

The performance of approximation techniques that predict probability
functions can be analyzed through a variety of summary
statistics. This work uses the max absolute difference, also known as
the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
for its compatibility with the KS test.

The two-sample KS test is a useful nonparametric test for comparing
two CDFs while only assuming stationarity, finite mean, and finite
variance. The null hypothesis (that two CDFs come from the same
underlying distribution) is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
applications of the KS test presented in this work $n_1 = n_2$. An
example of the \changed{end-to-end}{round-trip} prediction methodology from known and
predicted distributions to the calculation of error can be seen in
Figure \ref{fig:prediction-example}. A brief listing of relevant
statistical terms used throughout this work is provided in
the Appendix (Section \ref{sec:appendix}).

\section{Theoretical Error Bound}
\label{sec:theory}

This section presents the theoretical results bounding the error of
(piecewise) linear interpolation. The error analysis relies on linear
interpolation for three reasons: (1) second order results can be
obtained utilizing a Lipschitz constant on the gradient of a function,
rather than standard Lipschitz bounds; (2) the results directly apply
to Delaunay interpolation; and (3) multiple other interpolants in this
paper compute predictions as convex combinations of observed function
values, which may allow for straightforward extensions of this error
bound.

\begin{plainlemma}
  \label{lemma:1}
  Let $S \subset \mathbb{R}^d$ be open and convex, $f: S \rightarrow
  \mathbb{R}$, and let $\nabla f(x)$ be $\gamma$-Lipschitz continuous
  in the $2$-norm. Then for all $x,y \in S$
  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
\end{plainlemma}

\begin{proofdot}
  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
  in the direction $(y - x).$
  \begin{align*}
    \big|f(y) - f(x) - \langle &\nabla f(x), y - x \rangle \big|
        = \big|g(1) - g(0) - g'(0) \big| \\
       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt = \frac{\gamma \|y - x\|_2^2}{2}.
  \end{align*}
  \qed
\end{proofdot}

\hfill

\begin{plainlemma}
  \label{lemma:2}
  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
  $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
  where $\sigma_d$ is the smallest singular value of $M.$
\end{plainlemma}

\begin{proofdot}
  Using the facts that $M$ and $M^t$ have the same singular values,
  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
  \begin{align*}
    \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
                  &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
                  &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
  \end{align*}
  \qed
\end{proofdot}

%% \newpage

\begin{plainlemma}
  \label{lemma:3}
  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \frac{\sqrt{d} \, \gamma k^2}{2 \sigma_d}.$$
\end{plainlemma}

\begin{proofdot}
  \def\g{\gamma_g}
  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
  = \gamma \|x_j-x_0\|_2^2$.

  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
  and so $g'(0) - \g z \le g'(z) = 0$. Now
  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
  \begin{align*}
    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
  \end{align*}
  \noindent a contradiction for the value of $g(\tilde t)$. A similar
  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
  \g/2$. In terms of $f$,
  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
  \implies \|\nabla f(x_0)-c\|_2 \le \frac{\sqrt{d} \, \gamma
    k^2}{2\sigma_d}.$$ \qed
\end{proofdot}

\begin{onetheorem}
  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
    \ref{lemma:3}}, for $z \in S$,
  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d} \, \gamma k^2}{2 \sigma_d} \|z - x_0\|_2.$$
\end{onetheorem}

\begin{proofdot}
  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d}\, \gamma k^2 / (2 \sigma_d)$ by {\it Lemma \ref{lemma:3}}. Now
  \begin{align*}
    \big|f(z) - \hat f(z)\big|
       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \big(\sqrt{d}\,\gamma k^2 / (2 \sigma_d)\big) \|z - x_0\|_2 \\
       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d}\,\gamma k^2}{2 \sigma_d} \|z - x_0\|_2,
  \end{align*}
  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
  %
  \qed
\end{proofdot}

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate.  Only linear convergence to the true function can
be achieved in practice, without the incorporation of additional
observations. Notice that the approximation error is largely
determined by the spacing of observed data. Predictions made by
simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher
error.

That the theoretical error bound is sharp can be observed with the
test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d,$ and the
simplex defined by vertices $X = \{0, e_1, \ldots, e_d\},$ where $e_i$
is the $i$-th standard basis vector in $\mathbb{R}^d,$ $e = (1$,
$\ldots$, $1) \in \mathbb{R}^d,$ and $0$ denotes the zero vector in
any dimension.  The constants relevant to the error bound are
$$ \gamma = 2, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 = 0,
\qquad \hat q(x) = \langle e, x - 0\rangle + q(0). $$
\noindent Noting that $q(0) = 0,$ the approximation error at $z =
-(1/2)e$ is
$$ \big|q(z) - \hat q(z)\big| = \big|\|z\|_2^2 - \langle e,z
\rangle\big| = \big| d/4 + d/2 \big| = 3d/4, $$
\noindent while the error bound from the theorem gives

\vspace{-2mm}
\begin{align*}
  \big|q(z) - \hat q(z)\big|
  &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d}\,\gamma k}
      {2 \sigma_d} \|z - x_0\|_2 \\
  &{}=  \|z\|_2^2 + \sqrt{d} \|z\|_2 \\
  &{}=  d/4 + d/2 = 3d/4.
\end{align*}
\vspace{-2mm}

\noindent Acknowledging that the error bound is sharp, it may be of
interest to observe the error of piecewise linear approximation
techniques on an analytic test function.

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.9\textwidth,]{Fig2.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
  \end{tikzpicture}
  \caption{Delaunay and MLP approximations are constructed from Fekete
    points over the unit cube evaluating the test function $f(x) =
    \cos(\|x\|_2)$ for $x \in \mathbb{R}^2$. The figure shows the
    first/third quartiles at the box bottom/top, the second quartile
    (median) at the white bar, median 95\% confidence interval (cones,
    barely visible in figure), and whiskers at 3/2 of the adjacent
    interquartile ranges, for the absolute prediction error for each
    model at $1000$ random evaluation points. The left plot observes a
    perfect interpolation problem with exact evaluations of $f.$ The
    right plot observes a regression problem with uniform random noise
    giving values in $[.9 f(x),\ 1.1f(x)]$ for each $x.$ Both axes are
    log scaled.}
  \label{fig:convergence-2d}
\end{figure}

%% ===================================================================
\vspace{-2mm}
\subsection{Demonstration on an Analytic Test Function}
\label{sec:analytic}

The theoretical results constructed in Section \ref{sec:theory} for
(piecewise) linear interpolation are promising and apply directly to
Delaunay interpolation, however they are difficult to interpret in
context with approximation algorithms that do not have similar known
uniform error bounds. For that reason, an analytic function is used to
measure the error of a piecewise linear interpolant (Delaunay) and a
piecewise linear regressor (the MLP) when provided an increasing
amount of data with varying levels of random noise. \changed{Only
  Delaunay and the MLP are considered in this demonstration because
  all other mentioned techniques are not strictly piecewise linear
  approximations.}{}

The test function chosen here for analysis resembles the
\textit{oscillatory} function used by \cite{barthelmann2000high}.
However, a slight modification is made to remove the simple dot
product structure (which is favorable for the MLP). Let $f(x)
=\cos(\|x\|_2)$ for $x \in \mathbb{R}^d$ where $d$ is either $2$
(Fig. \ref{fig:convergence-2d}) or $20$
(Fig. \ref{fig:convergence-20d}). This function has a bounded change
in gradient $2$-norm and hence meets the necessary Lipschitz condition
for the error bound. Data points and approximation points for this
test will be within the unit cube $[0,1]^d$.

The error of a linear interpolant constructed from well spaced points
is dominated by the distance to the nearest data point. In order to
uniformly decrease the distance to the nearest data point across the
unit cube, exponentially more data points must be available for
approximation. For this experiment $N = 2(4^i),$ $i = 1,$ $\ldots,$
$6,$ chosen points are kept well spaced by computing approximate
Fekete points. Fekete points have a history in potential theory
\cite{kovari_pommerenke_1968} and are most generally defined as those
points that maximize the absolute value of a Vandermonde
determinant. Here, the QR method outlined in \cite{bos2010computing}
is used to identify approximate Fekete points from a Vandermonde
matrix. A multivariate polynomial basis of size $2N$ is constructed
containing all polynomials of degree $\leq n$ for the largest $n$ such
that ${d+n \choose n} \leq 2N$ and $2N - {d+n \choose n}$
arbitrarily selected polynomials of degree $n + 1.$ The Vandermonde
matrix for these $2N$ polynomials is used to select $N$ approximate
Fekete points.

Finally an additional aspect is added to this test problem by
incorporating random uniform noise into the evaluations of $f.$ For
each test two experiments are executed, one with exact function
evaluations (an interpolation problem) and one with a constant
signal-to-noise (SNR) ratio of 10:1 (a regression problem).

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.9\textwidth,]{Fig3.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
  \end{tikzpicture}
  \caption{Delaunay and MLP approximations are constructed from Fekete
    points over the unit cube evaluating the test function $f(x) =
    \cos(\|x\|_2)$ for $x \in \mathbb{R}^{20}$. The details are the
    same as for Fig. \ref{fig:convergence-2d}.}
  \label{fig:convergence-20d}
\end{figure}

The bound from the theorem suggests that exponentially increasing the
number of \changed{well-spaced}{} data points should result in a
decreasing error. The $d = 2$ test seen on the left half of Figure
\ref{fig:convergence-2d} shows a consistent decrease in error for
Delaunay and also shows the eventual accuracy plateau obtained by a
parametric regression form (the MLP, at roughly 500 points). On the
right hand side of Figure \ref{fig:convergence-2d} the random noise
clearly prohibits Delaunay from converging to $f$, while the MLP is
able to improve its approximation with more data points on average.
The convergence result for a very low-dimensional problem like this is
expected.  However, intuition fails for higher dimensional problems.

Figure \ref{fig:convergence-20d} shows the test function with $d = 20$
presents a significantly more challenging approximation problem than
its counterpart in low dimension. The same increase in number of data
points from 32 to 8192 causes no apparent improvement in approximation
for either the noise-free or the noisy problems. Perhaps unexpectedly,
the interpolation technique performs better than the regression
technique on the noisy data (right) in Figure
\ref{fig:convergence-20d}, and worse than the regression technique on
the noise-free data (left). This result emphasizes the relevance of
interpolation for problems in high dimension. It also reveals that the
outcome of MLP regressions can get worse when adding more data points.

\changed{
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.9\textwidth,]{theorem-convergence.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Euclidean Distance};
  \end{tikzpicture}
  \caption{\color{red}The distribution of absolute error, distance to
    the nearest data point, smallest singular value and the longest
    edge of the simplex containing each approximation point in the
    tests from Fig. \ref{fig:convergence-2d} and
    Fig. \ref{fig:convergence-20d} for Delaunay. In two dimensions it
    can be seen that $\|z - x_0\|_2$, $\sigma_d$, and $k$ all shrink
    at the same rate for well-spaced approximation points, resulting
    in a faster rate of decrease for approximation error. Notice that
    in higher dimension the data remains sparse even with thousands of
    data points. This is seen by the relatively small reduction in
    $k$, and this explains the slow reduction in error seen by
    Delaunay in Fig. \ref{fig:convergence-20d}.}
  \label{fig:data-spacing}
\end{figure}
}{}

\changed{The results achieved for both the $d = 2$ and $d = 20$ align
  with the theoretical error bound as can be seen in Figure
  \ref{fig:data-spacing}. In low dimension, thousands of data points
  meaningfully reduce the sparsity of the approximation problem.
  However, in higher dimension it takes many more points to achieve a
  reduction in data sparsity. The same result holds for a Latin
  hypercube well-spaced point set, suggesting that sparsity is an
  inherent problem in high dimension approximation problems. This
  result is not a product of the chosen test function, but rather the
  exponential increase in complexity that accompanies increased
  dimension.}{}

In summary, the regime of interpolation is \changed{}{much} greater
for moderate to high dimensional problems \changed{due to increased
  sparsity}{}, even for functions with random noise incorporated into
their evaluation. Acknowledging the viability of interpolation for
problems of moderate dimension, the next section will consider
real-world problems of similar proportion (thousands of examples in
tens of dimensions).

%% ===================================================================
\vspace{-2mm}
\section{Data and Empirical Analysis}
\label{sec:data}

This section extends the comparison of interpolation and regression
algorithms to a sample of real-world problems. Five different data
sets of varying dimension and application are utilized to construct
approximations and compare the accuracy of different techniques.

In the following five subsections the sources and targets of each test
data set are described, as well as challenges and limitations related
to approximating these data. The distribution of response values being
modeled is presented followed by the distribution of approximation
errors for each algorithm. The plots for all five data sets have the
same format.

All five data sets are rescaled such that the domain of approximation
is the unit hypercube. The range of the first four data sets is the
real numbers, while the range of the fifth data set is the space of
cumulative distribution functions. All approximation techniques are
applied to the first four data sets, while only those interpolants
whose approximations are convex combinations of observed data are
applied to the final data set.

All approximations are constructed using $k$-fold cross validation as
described in \cite{kohavi1995study} with $k=10$. This approach
randomly partitions data into $k$ (nearly) equal sized sets. Each
algorithm is then evaluated by constructing an approximation over each
unique union of $k-1$ elements of the partition, making predictions
for points in the remaining element. As a result, each observed data
point is used in the construction of $k-1$ different approximations
and is approximated exactly once. The $k$-fold cross validation method
is data-efficient and provides an unbiased estimate of the expected
prediction error \cite{kohavi1995study}, however it should be noted
that neither this method nor others can provide a universally unbiased
estimator for the variance of prediction error \cite{bengio2004no}.

In addition to the figures displaying approximation results for each
data set, tables of accompanying numerical results are located in the
Appendix (Section \ref{sec:appendix}). All of the test data sets
capture underlying functions that are almost certainly stochastic. As
described in Section \ref{sec:introduction}, regression techniques
appear most appropriate for these problems. However, typically data
grows exponentially more sparse with increasing dimension. Given that
sparse data regressors tend towards interpolation and, as demonstrated
in Section \ref{sec:analytic}, interpolants produce similar (if not
identical) results, it is presumed that interpolants are equally
viable approximation techniques on these problems.

%% -------------------------------------------------------------------
\subsection{Forest Fire ($n = 504$, $d = 12$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of forest fire area burned under recorded weather
    conditions. The data is presented on a $\ln$ scale because
    most values are small with exponentially fewer fires on record
    that burn large areas.}
  \label{fig:hist-forest-fire}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the amount of area
    that would be burned given environment conditions. $10$-fold cross
    validation as described in the beginning of Section \ref{sec:data}
    is used to evaluated each algorithm. This results in exactly one
    prediction from each algorithm for each data point. These boxes
    depict the median (middle bar), median $95\%$ confidence interval
    (cones), quartiles (box edges), fences at $3/2$ interquartile
    range (whiskers), and outliers (dots) of absolute prediction error
    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
    errors are presented on a $\ln$ scale. The numerical data
    corresponding to this figure is provided in Table
    \ref{table:error-forest-fire} in the Appendix.}
  \label{fig:error-forest-fire}
\end{figure}

The forest fire data set \cite{cortez2007data} describes the area of
Montesinho park burned over months of the year along with
environmental conditions. The twelve dimensions being used to model
burn area are the $x$ and $y$ spatial coordinates of burns in the
park, month of year (mapped to $x$, $y$ coordinates on a unit circle),
the FFMC, DMC, DC, and ISI indices (see source for details), the
temperature, relative humidity, wind speed, and outdoor rain. The
original analysis of this data set demonstrated it to be difficult to
model, likely due to the skew in response values.

As suggested by Figure \ref{fig:error-forest-fire}, the SVR has the
lowest absolute prediction errors for $80\%$ of the data, with MLP and
Delaunay being the nearest overall competitors. The effectiveness of
SVR on this data suggests the underlying function can be defined by
relatively few parameters, as well as the importance of capturing the
low-burn-area data points.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\subsection{Parkinson's Telemonitoring ($n = 5875$, $d = 19$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig6.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of the Parkinson's patient total UPDRS clinical
    scores that will be approximated by each algorithm.}
  \label{fig:hist-parkinsons}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig7.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the total UPDRS score
    given audio features from patients' home life, using $10$-fold
    cross validation. These boxes depict the median (middle bar),
    median $95\%$ confidence interval (cones), quartiles (box edges),
    fences at $3/2$ interquartile range (whiskers), and outliers
    (dots) of absolute prediction error for each model. The numerical
    data corresponding to this figure is provided in Table
    \ref{table:error-parkinsons} in the Appendix.}
  \label{fig:error-parkinsons}
\end{figure}

The second data set for evaluation \cite{tsanas2010accurate} is
derived from a speech monitoring study with the intent to
automatically estimate Parkinson's disease symptom development in
Parkinson's patients. The function to be approximated is a
time-consuming clinical evaluation measure referred to as the UPDRS
score. The total UPDRS score given by a clinical evaluation is
estimated through 19 real numbers generated from biomedical voice
measures of in-home sound recordings.

Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the
lowest minimum, first quartile, and median of absolute errors for this
problem, while providing the best prediction $66\%$ of the time. The
MLP has the lowest third quartile and provides the best prediction for
$32\%$ of approximations. The dominance of ShepMod may be due in part
to regular-interval total UPDRS scores provided by clinicians,
favoring a nearest-neighbor strategy of prediction.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\subsection{Australian Daily Rainfall Volume ($n = 2609$, $d = 23$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig8.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of daily rainfall in Sydney, Australia, presented
    on a $\ln$ scale because the frequency of larger amounts of
    rainfall is significantly less. There is a peak in occurrence of
    the value $0$, which has a notable effect on the resulting model
    performance.}
  \label{fig:hist-weather}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig9.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the amount of
    rainfall expected on the next calendar day given various sources
    of local meteorological data, using $10$-fold cross validation.
    These boxes depict the median (middle bar), median $95\%$
    confidence interval (cones), quartiles (box edges), fences at
    $3/2$ interquartile range (whiskers), and outliers (dots) of
    absolute prediction error for each model. The errors are presented
    on a $\ln$ scale, mimicking the presentation in Figure
    \ref{fig:hist-weather}. The numerical data corresponding to this
    figure is provided in Table \ref{table:error-weather} in the
    Appendix.}
  \label{fig:error-weather}
\end{figure}

The third data set for the total daily rainfall in Sydney, Australia
\cite{williams2009rattle} provides a slightly higher dimensional
challenge for the interpolants and regressors. This data is composed
of many meteorological readings from the region in a day including:
min and max temperatures, sunshine, wind speed directions (converted
to coordinates on a circle), wind speeds, and humidities throughout
the day, day of the year (converted to coordinates on a circle), and
the model must predict the amount of rainfall tomorrow.

While Figure \ref{fig:hist-weather} makes MARS look far better than
other techniques, it only provides the best prediction for $11\%$ of
points. The MLP has the lowest absolute error for $56\%$ of points and
LSHEP is best for $28\%$. MARS likely achieves such a low first
quartile due to the high occurrence of the value zero in the data.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\subsection{Credit Card Transaction Amount ($n = 5562$, $d = 28$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig10.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of credit card transaction amounts, presented on
    a $\ln$ scale. The data contains a notable frequency peak around
    $\$1$ transactions. Fewer large purchases are made, but some large
    purchases are in excess of five orders of magnitude greater than
    the smallest purchases.}
  \label{fig:hist-credit-card}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig11.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the expected
    transaction amount given transformed (and obfuscated) vendor and
    customer-descriptive features, using $10$-fold cross validation.
    These boxes depict the median (middle bar), median $95\%$
    confidence interval (cones), quartiles (box edges), fences at
    $3/2$ interquartile range (whiskers), and outliers (dots) of
    absolute prediction error for each model. The absolute errors in
    transaction amount predictions are presented on a $\ln$ scale,
    just as in Figure \ref{fig:hist-credit-card}. The numerical data
    corresponding to this figure is provided in Table
    \ref{table:error-credit-card} in the Appendix.}
  \label{fig:error-credit-card}
\end{figure}

The fourth test data set, and the final with a real-valued range, is a
collection of credit card transactions
\cite{pozzolo2015calibrating}. The provided data carries no direct
real-world meaning, being the output of a principle component analysis
on the original hidden source data. This obfuscation is done to
protect the information of the credit card users. This data has the
largest dimension of all considered, at 28. A model for this data
predicts the transaction amount given the vector of principle
component information.

As can be seen in Figure \ref{fig:error-credit-card}, the MLP
outperforms all other algorithms at the first, second, third, and
fourth quartiles. The MLP produces the lowest absolute error
prediction for $80\%$ of transactions, Delaunay bests the remaining
$20\%$. It is likely that with less data, Delaunay would be the best
performer.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\subsection{High Performance Computing I/O ($n = 3016$, $d = 4$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig12.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of the raw throughput values recorded during all
    IOzone tests across all system configurations. The distribution is
    skewed right, with few tests having significantly higher
    throughput than most others. The data is presented on a $\ln$
    scale.}
  \label{fig:hist-throughput}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig13.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughput KS Statistic};
  \end{tikzpicture}
  \caption{The models directly capable of predicting distributions are
    applied to predicting the expected CDF of I/O throughput at a
    previously unseen system configuration, using $10$-fold cross
    validation. The KS statistic (max norm) between the observed
    distribution at each system configuration and the predicted
    distribution is recorded and presented above. Note that the above
    figure is \textit{not} log-scaled like Figure
    \ref{fig:hist-throughput}. The numerical data corresponding to
    this figure is provided in Table \ref{table:error-throughput} in
    the Appendix.}
  \label{fig:error-throughput}
\end{figure}

The final of five data sets is derived from \cite{cameron2019moana},
which provides four-dimensional distribution data by executing the
IOzone benchmark \cite{iozone} on a computer system and varying the
system's file size, record size, thread count, and CPU frequency. At
each configuration, IOzone samples the I/O file-read throughput (in
bytes per second) 150 times. Empirical distribution function points
are computed from each set of 150 executions, which are interpolated
with a piecewise cubic Hermite interpolating polynomial
\cite{fritsch1980monotone} to approximate the CDF. All interpolation
algorithms with the exception of LSHEP are used to approximate these
CDFs from system configurations.

Delaunay achieves the lowest KS statistic (max norm difference) for
$62\%$ of approximations, while Voronoi is best for the remaining
$38\%$. Figure \ref{fig:error-throughput} shows that while Delaunay
may have more best predictions, the behavior of Voronoi may be
preferable.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth]{Fig14.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of KS Statistic};
  \end{tikzpicture}
  \caption{Histograms of the prediction error for each interpolant
    that produces predictions as convex combinations of observed data,
    using $10$-fold cross validation. The histograms show the KS
    statistics for the predicted throughput distribution versus the
    actual throughput distribution. The four vertical lines represent
    cutoff KS statistics given $150$ samples for commonly used
    $p$-values 0.05, 0.01, 0.001, $10^{-6}$, respectively. All
    predictions to the right of a vertical line represent CDF
    predictions that are significantly different (by respective
    $p$-value) from the actual distribution according to the KS
    test. The numerical counterpart to this figure is presented in
    Table \ref{table:null-hypothesis-results}.}
  \label{fig:throughput-prediction}
\end{figure}

\begin{table}
  \renewcommand{\arraystretch}{1.3}
  \centering
  \begin{tabular}{c|c|c|c|c}
               & $p = .05$         & $p = .01$         & $p = .001$        & $p = 10^{-6}$\\
    \hline
    Delaunay   & $\mathbf{50.3}\%$ & $\mathit{43.5}\%$ & $\mathit{36.2}\%$ & $\mathit{24.7}\%$\\
    ShepMod    & $\mathit{51.4}\%$ & $44.8\%$          & $38.1\%$          & $27.7\%$\\
    Voronoi    & $52.6\%$          & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
    BoxSplines & $99.4\%$          & $98.5\%$          & $96.6\%$          & $89.3\%$\\
  \end{tabular}
  \caption{Numerical counterpart of the histogram data presented in
    Figure \ref{fig:throughput-prediction}. The columns display the
    percent of null hypothesis rejections by the KS-test when provided
    different selections of $p$-values for each algorithm. The
    algorithm with the lowest rejection rate at each $p$ is boldface,
    while the second lowest is italicized.}
  \label{table:null-hypothesis-results}
\end{table}

%% ===================================================================

Figure \ref{fig:throughput-prediction} expands on the KS statistic
results presented in Figure \ref{fig:error-throughput}. Agglomerate
errors for each algorithm resemble a Gamma distribution. The
percentages of significant prediction errors with varying $p$-values
are on display in Table \ref{table:null-hypothesis-results}. When
considering the $p=0.001$ results for each technique, a little over
one third of the predicted CDFs are significantly different from the
measured (and presumed) correct CDFs. However, it should be noted that
with 150 samples, the error of an empirical distribution function
(EDF) can reasonably be upwards of $.1$, which serves as a rough
estimate for the lower limit of achievable error. Globally, only a
third of Voronoi predictions fail to capture \textit{all} of the
characteristics of the CDFs at new system configurations.

\section{Discussion}
\label{sec:discussion}

\begin{table}
  \centering
  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l}
    Algorithm & \multicolumn{2}{c|}{Avg. \% Best} & \multicolumn{2}{c|}{\multilinecell{Avg. Fit or \\ Prep. Time (s)}} & \multicolumn{2}{c}{Avg. App. Time (s)}\\
    \hline
    MARS        & \quad\quad$4$&$5$          & \quad\quad$20$&$0$s        & \qquad\quad $0$&$001$s\\
    SVR         & $\mathit{19}$&$\mathit{5}$ & $\mathbf{0}$&$\mathbf{5}$s & $\mathbf{0}$&$\mathbf{0001}$s\\
    MLP         & $\mathbf{43}$&$\mathbf{1}$ & $200$&$0$s                 & $0$&$001$s\\
    Delaunay    & $5$&$2$                    & $1$&$0$s                   & $1$&$0$s\\
    ShepMod     & $18$&$0$                   & $\textit{0}$&$\textit{7}$s & $\mathbf{0}$&$\mathbf{0001}$s\\
    LSHEP       & $8$&$4$                    & $2$&$0$s                   & $\mathbf{0}$&$\mathbf{0001}$s\\
    Voronoi     & $0$&$5$                    & $1$&$0$s                   & $0$&$04$s\\
    BoxSplines  & $3$&$5$                    & $0$&$8$s                   & $\mathit{0}$&$\mathit{0005}$s\\
  \end{tabular}
  \caption{This average of Appendix Tables
    \ref{table:best-forest-fire}, \ref{table:best-parkinsons},
    \ref{table:best-weather}, and \ref{table:best-credit-card}
    provides a gross summary of overall results. The columns display
    (weighted equally by data set, \textit{not} points) the average
    frequency with which any algorithm provided the lowest absolute
    error approximation, the average time to fit/prepare, and the
    average time required to approximate one point. The times have
    been rounded to one significant digit, as reasonably large
    fluctuations may be observed due to implementation hardware.
    Interpolants provide the lowest error approximation for nearly one
    third of all data, while regressors occupy the other two
    thirds. This result is obtained without any customized tuning or
    preprocessing to maximize the performance of any given
    algorithm. In practice, tuning and preprocessing may have large
    effects on approximation performance.}
  \label{table:avg-performance}
\end{table}

Table \ref{table:avg-performance} summarizes results across the four
test data sets with real-valued ranges. The interpolants discussed in
this paper produce the \textit{best} approximations roughly one third
of the time, and produce competitive approximations for almost all
data sets. These test problems are almost certainly
\textit{stochastic} in nature, but the high dimension leads to greater
data sparsity and model construction cost, making interpolation more
competitive.

The major advantages to interpolation lie in the near absence of
\textit{fit} time. Delaunay, LSHEP, and ShepMod all require pairwise
distance calculations, for numerical robustness (Delaunay) and to
determine the radii of influence for data points (LSHEP and
ShepMod). At least hundreds, and sometimes hundreds of thousands of
predictions can be made by the interpolants before the most widely
used regressor (MLP) finishes fitting these relatively small data
sets.  However, the computational complexities of all interpolants
presented are greater than linear in either dimension or number of
points, whereas the regressors' nonlinear complexity in dimension
generally comes from the model fitting optimization.

The new theoretical results presented in Section \ref{sec:theory}
directly apply to Delaunay interpolation, however the performance of
Delaunay does not appear significantly better than other algorithms on
these data sets.  This observation may be due to the stochastic nature
of the data, but it also speaks to the power of the approximations
generated by the different interpolation methods.  The strong
performance of other interpolants suggests that theoretical results
similar to those presented in this work can be achieved for the other
interpolants under reasonable assumptions.

Finally, most of the interpolants presented in this work benefit from
the ability to model \textit{any} function over real $d$-tuples with a
range that is closed under convex combinations. In general, error can
be quantified by any measure (particularly of interest may be
$L^2$, $L^\infty$, etc.). The results of the distribution prediction
case study indicate that interpolants can effectively predict
CDFs. The error analysis for that work relies on the KS statistic,
which captures the worst part of any prediction and hence provides a
conservatively large estimate of approximation error. The average
absolute errors in the predicted CDFs are always lower than the KS
statistics. However, the KS statistic was chosen as a metric because
of the important surrounding statistical theory. A nonnegligible
volume of predictions provide impressively low levels of average
absolute error in that final case study.

\section{Conclusion}
\label{sec:conclusion}

The major contributions of this work are: 1) new uniform theoretical
error bounds for piecewise linear interpolation in arbitrary dimension
(Section \ref{sec:theory}); 2) an empirical evaluation across
real-world problems that demonstrates interpolants produce
competitively accurate models of multivariate phenomenon when compared
with common regressors for sparse, moderately high dimensional
problems (Section \ref{sec:data}); and 3) a demonstration that some
interpolants generalize to interpolation in function spaces (Section
\ref{sec:error}), preserving monotonicity (with CDFs, e.g.), neither
of which common regressors can do.

The various interpolants discussed in this paper have been
demonstrated to effectively approximate multivariate phenomena up to
$30$ dimensions. The underlying constructions are theoretically
straightforward, interpretable, and yield reasonably accurate
predictions. Most of the interpolants' computational complexities make
them particularly suitable for applications in even higher
dimension. The major benefits of interpolation are seen when only a
small number of approximations ($\leq 1000$) are made from data and
when there are relatively few data points for the dimension (for
empirical results presented, $\log_d n \leq 5$). These findings
encourage broader application of interpolants to multivariate
approximation in science.

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

%%====================================================================
%%====================================================================

\bibliographystyle{spmpsci}      % mathematics and physical sciences
\bibliography{paper}

\newpage

\begin{appendix}
\section{Appendix}
\label{sec:appendix}

\begin{normalsize}
\textbf{Statistical Terminology.} A random variable $X$ is precisely
defined by its \textit{cumulative distribution function} (CDF) $F_X$
and the derivative of the CDF, the \textit{probability density
  function} (PDF) $f_X.$ For any possible value $x$ of $X$, the
\textit{percentile} of $x$ is $100$ $F_X(x)$, the percentage values
drawn from $X$ that would be less than or equal to $x$ as the number
of samples tends towards infinity. The \textit{quartiles} of $X$ are
its $25$-th, $50$-th (median), and $75$-th percentiles. The absolute
difference between the median and an adjacent quartile is an
\textit{interquartile range}. Given an independent and identically
distributed sample from $X$ and presuming that $X$ has finite mean and
variance, a \textit{confidence interval} can be drawn about any
percentile estimated from the sample. A \textit{confidence interval}
describes the probability that a value lies within an interval. The
\textit{null hypothesis} is a statement (derived from some test
statistic) that the expected value of the observed statistic is equal
to an assumed population statistic. The $p$ value is the probability
of observing a given statistic if the \textit{null hypothesis} is
true. \changed{For a more detailed introduction to statistics and 
related terminology, see the work of Navidi \cite{navidi_2015}.}{}

\vspace{4mm}
\noindent \textbf{Raw Numerical Results.} The tables that follow show the
precise experimental results for all data sets presented in Section
\ref{sec:data}. The tests were all run serially on an otherwise idle
machine with a CentOS 6.10 operating system and an Intel i7-3770 CPU
operating at 3.4 GHz. The detailed performance results in the tables
that follow are very much dependent on the problem and the algorithm
implementation (e.g., some codes are TOMS software, some industry
distributions, and others are from conference paper venues). Different
typeface is used to show best performers, however not much
significance should be attached to ranking algorithms based on small
time (millisecond) differences. The results serve as a demonstration
of conceptual validity.

\end{normalsize}

\newpage

\vspace*{\fill}

\begin{table}[H]
  \centering
  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
    \hline
    Algorithm & \multicolumn{2}{c}{Min}       & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
    \hline
    MARS      & $0$&$00984$                   & $3$&$11$                      & $7$&$01$                      & $\mathit{11}$&$\mathit{7}$    & $1090$&$0$\\
    SVR       & $0$&$0118$                    & $\mathbf{0}$&$\mathbf{615}$   & $\mathbf{0}$&$\mathbf{931}$   & $\mathbf{5}$&$\mathbf{89}$    & $1090$&$0$\\
    MLP       & $0$&$0426$                    & $2$&$63$                      & $5$&$27$                      & $14$&$0$                      & $1090$&$0$\\
    Delaunay  & $\mathbf{0}$&$\mathbf{0}$     & $1$&$98$                      & $5$&$37$                      & $13$&$1$                      & $\mathit{1080}$&$\mathit{0}$\\
    ShepMod   & $\mathbf{0}$&$\mathbf{0}$     & $1$&$93$                      & $6$&$27$                      & $16$&$0$                      & $1090$&$0$\\
    LSHEP     & $0$&$04$                      & $2$&$17$                      & $8$&$87$                      & $19$&$1$                      & $\mathbf{1070}$&$\mathbf{0}$\\
    Voronoi   & $\mathit{0}$&$\mathit{00982}$ & $3$&$65$                      & $7$&$56$                      & $15$&$6$                      & $1090$&$0$\\
    BoxSpline & $\mathbf{0}$&$\mathbf{0}$     & $\mathit{1}$&$\mathit{27}$    & $\mathit{4}$&$\mathit{61}$    & $12$&$6$                      & $1090$&$0$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in
    Figure \ref{fig:error-forest-fire}. The columns of absolute error
    percentiles correspond to the minimum, first quartile, median,
    third quartile, and maximum absolute errors respectively. The
    minimum of each column is boldface, while the second lowest value
    is italicized. All values are rounded to three significant
    digits.}
  \label{table:error-forest-fire}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
    \cline{1-3}\cline{5-10}
    Algorithm & \multicolumn{2}{c|}{\% Best}   &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
    \cline{1-3}\cline{5-10}
    MARS      & \quad$7$&$3$                   &  & \quad\quad$29$&$1$                      & \quad$0$&$00137$                   & \quad\quad\quad$0$&$0686$\\
    SVR       & $\mathbf{78}$&$\mathbf{0}$     &  & $\mathbf{0}$&$\mathbf{00584}$           & $\mathit{0}$&$\mathit{0000620}$    & $\mathit{0}$&$\mathit{00310}$\\
    MLP       & $0$&$0$                        &  & $32$&$8$                                & $0$&$000871$                       & $0$&$0436$\\
    Delaunay  & $0$&$2$                        &  & $0$&$0151$                              & $0$&$0234$                         & $1$&$18$\\
    ShepMod   & $2$&$0$                        &  & $\mathit{0}$&$\mathit{00634}$           & $0$&$0000644$                      & $0$&$00322$\\
    LSHEP     & $5$&$1$                        &  & $0$&$0275$                              & $\mathbf{0}$&$\mathbf{0000463}$    & $\mathbf{0}$&$\mathbf{00231}$\\
    Voronoi   & $0$&$0$                        &  & $0$&$0152$                              & $0$&$000396$                       & $0$&$0198$\\
    BoxSpline & $\mathit{9}$&$\mathit{7}$      &  & $0$&$00724$                             & $0$&$0000978$                      & $0$&$00489$\\
    \cline{1-3}\cline{5-10}
  \end{tabular}
  \caption{The left above shows how often each algorithm had the
    lowest absolute error approximating forest fire data in Table
    \ref{table:error-forest-fire}. On the right columns are median fit
    time of 454 points, median time for one approximation, and median
    time approximating 50 points.}
  \label{table:best-forest-fire}
\end{table}

\vspace*{\fill}
\newpage

\begin{table}
  \centering
  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
    \hline
    Algorithm & \multicolumn{2}{c}{Min}                   & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
    \hline
    MARS      & $0$&$00948$                               & $9$&$98$                      & $20$&$4$                      & $32$&$9$                      & $863$&$0$\\
    SVR       & $0$&$00233$                               & $3$&$15$                      & $7$&$21$                      & $11$&$3$                      & $\mathbf{28}$&$\mathbf{6}$\\
    MLP       & $0$&$0000239$                             & $\mathit{0}$&$\mathit{533}$   & $\mathit{1}$&$\mathit{25}$    & $\mathbf{2}$&$\mathbf{84}$    & $39$&$3$\\
    Delaunay  & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$2$                       & $3$&$5$                       & $7$&$67$                      & $30$&$7$\\
    ShepMod   & $\mathbf{0}$&$\mathbf{0}$                 & $\mathbf{0}$&$\mathbf{255}$   & $\mathbf{0}$&$\mathbf{908}$   & $\mathit{3}$&$\mathit{43}$    & $34$&$5$\\
    LSHEP     & $0$&$00254$                               & $2$&$93$                      & $7$&$16$                      & $13$&$1$                      & $\mathit{29}$&$\mathit{0}$\\
    Voronoi   & $\mathbf{0}$&$\mathbf{0}$                 & $1$&$29$                      & $3$&$52$                      & $6$&$87$                      & $30$&$1$\\
    BoxSpline & $0$&$006$                                 & $4$&$3$                       & $8$&$91$                      & $15$&$1$                      & $45$&$3$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in
    Figure \ref{fig:error-parkinsons}. The columns of absolute error
    percentiles correspond to the minimum, first quartile, median,
    third quartile, and maximum absolute errors respectively. The
    minimum of each column is boldface, while the second lowest value
    is italicized. All values are rounded to three significant
    digits.}
  \label{table:error-parkinsons}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
    \cline{1-3}\cline{5-10}
    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
    \cline{1-3}\cline{5-10}
    MARS      & \quad\,$0$&$0$               &  & \quad\quad\quad$37$&$9$                 & \quad\,\,$0$&$00253$               & \quad\quad\quad\,\,$1$&$48$\\
    SVR       & $0$&$1$                      &  & $\mathbf{0}$&$\mathbf{859}$             & $\mathit{0}$&$\mathit{000181}$     & $\mathit{0}$&$\mathit{106}$\\
    MLP       & $\mathit{32}$&$\mathit{0}$   &  & $348$&$0$                               & $0$&$00111$                        & $0$&$653$\\
    Delaunay  & $0$&$0$                      &  & $2$&$47$                                & $1$&$22$                           & $720$&$0$\\
    ShepMod   & $\mathbf{66}$&$\mathbf{4}$   &  & $\mathit{1}$&$\mathit{13}$              & $0$&$000182$                       & $0$&$107$\\
    LSHEP     & $0$&$0$                      &  & $2$&$39$                                & $\mathbf{0}$&$\mathbf{000144}$     & $\mathbf{0}$&$\mathbf{0845}$\\
    Voronoi   & $1$&$6$                      &  & $2$&$77$                                & $0$&$0274$                         & $16$&$1$\\
    BoxSpline & $0$&$0$                      &  & $1$&$26$                                & $0$&$000643$                       & $0$&$377$\\
    \cline{1-3}\cline{5-10}
  \end{tabular}
  \caption{The left above shows how often each algorithm had the
    lowest absolute error approximating Parkinson's data in Table
    \ref{table:error-parkinsons}. On the right columns are median fit
    time of 5288 points, median time for one approximation, and median
    time approximating 587 points.}
  \label{table:best-parkinsons}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
    \hline
    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$}              & \multicolumn{2}{c}{$50^{th}$}  & \multicolumn{2}{c}{$75^{th}$}  & \multicolumn{2}{c|}{Max}\\
    \hline
    MARS      & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$                       & $1$&$96$                       & $\mathit{53}$&$\mathit{3}$\\
    SVR       & $0$&$0000915$                              & $0$&$0833$                                 & $0$&$263$                      & $\mathit{1}$&$\mathit{13}$     & $109$&$0$\\
    MLP       & $0$&$0000689$                              & $0$&$0337$                                 & $\mathbf{0}$&$\mathbf{0795}$   & $\mathbf{0}$&$\mathbf{264}$    & $\mathbf{5}$&$\mathbf{31}$\\
    Delaunay  & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$187$                                  & $0$&$919$                      & $2$&$72$                       & $56$&$3$\\
    ShepMod   & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$0957$                                 & $0$&$685$                      & $2$&$9$                        & $73$&$2$\\
    LSHEP     & $\mathbf{0}$&$\mathbf{0}$                  & $\mathit{0}$&$\mathit{0153}$               & $\mathit{0}$&$\mathit{106}$    & $1$&$17$                       & $113$&$0$\\
    Voronoi   & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$43$                                   & $1$&$28$                       & $2$&$94$                       & $83$&$8$\\
    BoxSpline & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$342$                                  & $1$&$59$                       & $5$&$53$                       & $119$&$0$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in
    Figure \ref{fig:error-weather}. The columns of absolute error
    percentiles correspond to the minimum, first quartile, median,
    third quartile, and maximum absolute errors respectively. The
    minimum value of each column is boldface, while the second lowest
    is italicized. All values are rounded to three significant
    digits.}
  \label{table:error-weather}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
    \cline{1-3}\cline{5-10}
    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)}    & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
    \cline{1-3}\cline{5-10}
    MARS      & \,\,\,\,$10$&$7$             &  & \quad\quad\quad$\mathit{0}$&$\mathit{151}$ & \quad$0$&$000117$                  & \quad\quad\quad\,\,$0$&$0304$\\
    SVR       & $0$&$0$                      &  & $\mathbf{0}$&$\mathbf{133}$                & $\mathbf{0}$&$\mathbf{0000947}$    & $\mathbf{0}$&$\mathbf{0246}$\\
    MLP       & $\mathbf{60}$&$\mathbf{9}$   &  & $169$&$0$                                  & $0$&$00137$                        & $0$&$356$\\
    Delaunay  & $0$&$1$                      &  & $0$&$664$                                  & $0$&$886$                          & $230$&$0$\\
    ShepMod   & $3$&$5$                      &  & $0$&$265$                                  & $0$&$000128$                       & $0$&$0333$\\
    LSHEP     & $\mathit{28}$&$\mathit{4}$   &  & $0$&$874$                                  & $\mathit{0}$&$\mathit{0000975}$    & $\mathit{0}$&$\mathit{0254}$\\
    Voronoi   & $0$&$2$                      &  & $0$&$675$                                  & $0$&$0270$                         & $7$&$01$\\
    BoxSpline & $4$&$1$                      &  & $0$&$330$                                  & $0$&$000406$                       & $0$&$106$\\
    \cline{1-3}\cline{5-10}
  \end{tabular}
  \caption{Left table shows how often each algorithm had the lowest
    absolute error approximating Sydney rainfall data in Table
    \ref{table:error-weather}. On the right columns are median fit
    time of 2349 points, median time for one approximation, and median
    time approximating 260 points.}
  \label{table:best-weather}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
    \hline
    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
    \hline
    MARS      & $5$&$36$                                   & $3610$&$0$                    & $4580$&$0$                    & $5450$&$0$                    & $13400$&$0$\\
    SVR       & $0$&$00706$                                & $8$&$14$                      & $\mathit{13}$&$\mathit{0}$    & $41$&$6$                      & $7690$&$0$\\
    MLP       & $0$&$00151$                                & $\mathbf{1}$&$\mathbf{6}$     & $\mathbf{3}$&$\mathbf{86}$    & $\mathbf{8}$&$\mathbf{34}$    & $\mathbf{604}$&$\mathbf{0}$\\
    Delaunay  & $\mathbf{0}$&$\mathbf{0}$                  & $\mathit{2}$&$\mathit{69}$    & $13$&$8$                      & $\mathit{35}$&$\mathit{0}$    & $\mathit{4840}$&$\mathit{0}$\\
    ShepMod   & $\mathbf{0}$&$\mathbf{0}$                  & $4$&$21$                      & $17$&$3$                      & $51$&$6$                      & $6510$&$0$\\
    LSHEP     & $1$&$27$                                   & $199$&$0$                     & $260$&$0$                     & $343$&$0$                     & $6530$&$0$\\
    Voronoi   & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$                      & $32$&$8$                      & $52$&$9$                      & $4860$&$0$\\
    BoxSpline & $\mathbf{0}$&$\mathbf{0}$                  & $12$&$4$                      & $35$&$0$                      & $90$&$1$                      & $7690$&$0$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in
    Figure \ref{fig:error-credit-card}. The columns of absolute error
    percentiles correspond to the minimum, first quartile, median,
    third quartile, and maximum absolute errors respectively. The
    minimum value of each column is boldface, while the second lowest
    is italicized. All values are rounded to three significant
    digits.}
  \label{table:error-credit-card}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
    \cline{1-3}\cline{5-10}
    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
    \cline{1-3}\cline{5-10}
    MARS      & \quad\,$0$&$0$               &  & \quad\quad\quad$22$&$0$                 & \quad$0$&$00148$                   & \quad\quad\quad\,\,$0$&$820$\\
    SVR       & $0$&$0$                      &  & $\mathbf{1}$&$\mathbf{01}$              & $\mathit{0}$&$\mathit{000210}$     & $\mathit{0}$&$\mathit{117}$\\
    MLP       & $\mathbf{79}$&$\mathbf{5}$   &  & $290$&$0$                               & $0$&$000714$                       & $0$&$397$\\
    Delaunay  & $\mathit{20}$&$\mathit{5}$   &  & $3$&$12$                                & $3$&$71$                           & $2070$&$0$\\
    ShepMod   & $0$&$1$                      &  & $\mathit{1}$&$\mathit{4}5$              & $0$&$000220$                       & $0$&$122$\\
    LSHEP     & $0$&$0$                      &  & $3$&$47$                                & $\mathbf{0}$&$\mathbf{000176}$     & $\mathbf{0}$&$\mathbf{0981}$\\
    Voronoi   & $0$&$0$                      &  & $3$&$32$                                & $0$&$0950$                         & $52$&$8$\\
    BoxSpline & $0$&$0$                      &  & $1$&$66$                                & $0$&$000956$                       & $0$&$532$\\
    \cline{1-3}\cline{5-10}
  \end{tabular}
  \caption{The left above shows how often each algorithm had the
    lowest absolute error approximating credit card transaction data
    in Table \ref{table:error-credit-card}. On the right columns are
    median fit time of 5006 points, median time for one approximation,
    and median time approximating 556 points.}
  \label{table:best-credit-card}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
    \hline
    Algorithm & \multicolumn{2}{c}{Min}      & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
    \hline
    Delaunay  & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$  & $\mathbf{0}$&$\mathbf{158}$   & $\mathit{0}$&$\mathit{308}$   & $\mathit{0}$&$\mathit{897}$\\
    ShepMod   & $0$&$0307$                   & $\mathit{0}$&$\mathit{0983}$  & $\mathit{0}$&$\mathit{164}$   & $0$&$335$                     & $1$&$0$\\
    Voronoi   & $\mathit{0}$&$\mathit{0303}$ & $0$&$1$                       & $0$&$165$                     & $\mathbf{0}$&$\mathbf{274}$   & $\mathbf{0}$&$\mathbf{762}$\\
    BoxSpline & $0$&$0679$                   & $0$&$429$                     & $0$&$571$                     & $0$&$752$                     & $1$&$0$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in
    Figure \ref{fig:error-throughput}. The columns of absolute error
    percentiles correspond to the minimum, first quartile, median,
    third quartile, and maximum KS statistics respectively between
    truth and guess for models predicting the distribution of I/O
    throughput that will be observed at previously unseen system
    configurations. The minimum value of each column is boldface,
    while the second lowest is italicized. All values are rounded to
    three significant digits.}
  \label{table:error-throughput}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
    \cline{1-3}\cline{5-10}
    Algorithm & \multicolumn{2}{c|}{\% Best}   &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
    \cline{1-3}\cline{5-10}
    Delaunay  & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad\quad$0$&$344$                & \quad$0$&$00984$                   & \quad\quad\,\,$2$&$71$\\
    ShepMod   & $0$&$0$                        &  & $\mathbf{0}$&$\mathbf{0884}$            & $\mathbf{0}$&$\mathbf{000145}$     & $\mathbf{0}$&$\mathbf{0436}$\\
    Voronoi   & $\mathit{38}$&$\mathit{0}$     &  & $0$&$360$                               & $0$&$00253$                        & $0$&$762$\\
    BoxSpline & $0$&$0$                        &  & $\mathit{0}$&$\mathit{0972}$            & $\mathit{0}$&$\mathit{000210}$     & $\mathit{0}$&$\mathit{0633}$\\
    \cline{1-3}\cline{5-10}
  \end{tabular}
  \caption{The left above shows how often each algorithm had the
    lowest KS statistic on the I/O throughput distribution data in
    Table \ref{table:error-throughput}. On the right columns are
    median fit time of 2715 points, median time for one approximation,
    and median time approximating 301 points.}
  \label{table:best-throughput}
\end{table}

\end{appendix}

\end{document}

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/3-Embedding-Polynomials.tex
updated work

\section{Embedding Polynomials}

Let $\mathbf{v}\in\mathbb{R}^d$ and let $p(\mathbf{v})$ be a multivariate polynomial of finite degree.
The goal of this section is to find real-valued least squares solution to equations of the form $p(\mathbf{v}) = 0$.
To do so, define the squared error function
\begin{equation}
E(\mathbf{v}) = \left( p(\mathbf{v}) \right)^2.
\label{eq:single-error}
\end{equation}
If $p(\mathbf{v}) = 0$ has real-valued solutions, then minimizing $E(\mathbf{v})$ will produce an element of the solution set.
If $p(\mathbf{v}) = 0$ is not exactly solvable for $\mathbf{v}\in\mathbb{R}^d$, then minimizing $E(\mathbf{v})$ will produce a solution in the least squares sense.
Similarly, given a system of polynomial equations $p_1(\mathbf{v})=0$, $\ldots$, $p_m(\mathbf{v})=0$, the total SOS error is given by
\begin{equation}
E(\mathbf{v}) = \left(p_1(\mathbf{v})\right)^2 + \ldots + \left(p_m(\mathbf{v})\right)^2.
\label{eq:system-error}  
\end{equation}
This section will show how to encode $\mathcal{H}(\text{\boldmath$\sigma$}) = E(\mathbf{v}) + \tau$, where $\mathcal{H}(\text{\boldmath$\sigma$})$ is an Ising-Hamiltonian of the form (\ref{eq:hamiltonian}), $E(\mathbf{v})$ is of the form (\ref{eq:single-error}), and $\tau$ is a constant energy offset.
SOS errors of the form (\ref{eq:system-error}) can be implemented trivially by summing over the individual Hamiltonians corresponding to each equation in the system.
As will be shown, the following process considers binary number representations of \textit{any} fixed precision, including those with mixed-precision and integer-valued problems.

Let $x_i\in\{0,1\}$ be the bit value of the $i$th bit of $x$, let $e$ denote a fixed exponent, and let $s\in\{-1,1\}$ denote whether $x$ is signed or unsigned ($-1$ for signed variables and $1$ for unsigned variables).
Then the following equation defines an $n$-bit binary encoding of $x$ using two's complement for signed numbers and a fixed-point precision determined by the exponent $e$.
Note that choosing $e=0$ results in an integer-valued problem.
\begin{equation}
x = 2^e \Bigg ( \sum_{i=1}^{n-1} 2^{i-1} x_i + 2^{n-1}x_{n}s \Bigg )
\label{eq:binary-encoding}
\end{equation}
Now, let $\mathbf{v} = [x^{(1)}, \ldots, x^{(d)}]^T$ and combine (\ref{eq:single-error}) and (\ref{eq:binary-encoding}) to express the energy as a multivariate binary polynomial with real-valued coefficients.
To get a QUBO expression of the form (\ref{eq:QUBO}), terms with three or more interacting variables must be quadratized (reduced to two-local form).
Many quadratization methods exist \cite{dattani2019quadratization}, but all known methods that maintain the squared-error energy landscape (i.e. reproduce the ground state and the full spectrum) utilize ancillary qubits.
This work employs reduction by substitution \cite{rosenberg1975reduction} for its equivalence with the logical AND operation.
The accompanying AND QUBO enforces $z_3 = z_1 \wedge z_2$ by assigning weights
\begin{equation}
C_{\wedge}(\mathbf{z}) = 3z_3 + z_1 z_2 - 2 (z_1 z_3 + z_2 z_3).
\label{eq:and}
\end{equation}

For example, consider the binary energy polynomial ${\tilde E}(\mathbf{x}) = x_1x_2x_3x_4$. Then setting $y_1 = x_1 \wedge x_2$ and $y_2 = x_3 \wedge x_4$ using (\ref{eq:and}) and summing over QUBOs gives 
$$
{\tilde C}(\mathbf{x},\mathbf{y}) = y_1y_2 + 3y_1 + 3y_2 + x_1x_2 + x_3x_4 - 2(x_1y_1 + x_2y_1) - 2(x_3y_2 + x_4y_2).
$$

When quadratizing terms as above, it is important to weight each occurrence of $C_\wedge$ by some large constant $\omega$ so that breaking an AND constraints incurs a large penalty.
Otherwise, the annealer could conceivably break an AND constraint to achieve an impossibly low squared error, resulting in a nonsensical solution.
As a rule of thumb, $\omega$ should be significantly greater than the expected squared residual but small enough so that $\Delta'$ is not unduly affected (as detailed in Section 4).
As long as the AND constraints are satisfied, the resulting QUBO will satisfy $C(\mathbf{x}) \cong E(\mathbf{v})$ and the resulting logical Hamiltonian will satisfy $\mathcal{H}(\text{\boldmath$\sigma$}) \cong E(\mathbf{v}) + \tau$, where the congruence is understood through the binary encoding of each $x^{(k)}$.

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/5-methods.tex
updated work

\section{Experimental Methods}

For the results presented in Section 6, logical Hamiltonians were generated via the process described in Section 3 and run on a D-Wave 2000Q machine.
Before performing quadratization or constructing the logical chains, the initial rescale factor $\lambda_I$ necessary to encode the logical Hamiltonian weights into the range allowed by the physical Hamiltonian was computed.
For all demonstrations the AND weights used for quadratization were set to $\lambda_I / 8$, with $1/8$ being the largest power of 2 multiplier that yields correct solutions on all test problems.
After adding the AND gates to the logical Hamiltonian, the final rescale factor $\lambda_F$ necessary to encode the physical Hamiltonian was computed and the chain strengths were set to $\lambda_F$.
Finally, the {\verb minorminer.find_embedding } tool from D-Wave's Ocean toolkit was used to construct the logical chains and perform the necessary rounding, completing the physical embedding \cite{cai2014practical}.
Any chain breaks were resolved by majority vote, and similarly, any violated AND gates were corrected in a post-processing step.
For each run, the default annealing time of 20 microseconds is used, along with D-Wave's default annealing schedule.

\subsection{Code Availability}

Python code with a user-friendly interface has been produced that automates the process for generating QUBOs, handles AND gate and chain weighting, generates physical Hamiltonian embeddings, runs on either simulated or physical hardware, and resolves AND gate and chain breakage.
Each of these steps is carried out exactly as described above.
This code library was used for generating the Hamiltonians in the Results section and is available on GitHub at \url{https://github.com/tchlux/qaml}.
Note, the library uses the SAPI API, and therefore depends on several D-Wave toolkits for finding physical embeddings on the Chimera graph structure and accessing D-Wave's cloud QA solvers.
The exact code used for replicating the experiments and the corresponding data for Tables \ref{tab:biprime}, \ref{tab:poly_ls}, and \ref{tab:linear_ls} are located in the {\verb qaml/experiments } subdirectory.
It is worth noting how concisely the following experiments are implemented in the provided framework (just 5, 7, and 14 lines of code for experiments 6.1, 6.2, and 6.3, respectively), as the high-level abstractions are conducive to programming and solving arbitrary polynomial systems on a quantum annealer with ease.

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/7-conclusion.tex
updated work

\section{Conclusion and Future Work}

In this work, a methodology is proposed for encoding polynomial SOS minimization for both fixed-point decimal and integer valued problems.
To achieve this, the squared error function is mapped to the QUBO model, which is isomorphic to the Ising-Hamiltonian model, using a two's complement encoding and the quadratization scheme of Rosenberg \cite{rosenberg1975reduction}.
The current state-of-the-art D-Wave Chimera architecture is able to solve the proposed embedding for a wide variety of problems, some of which do not admit analytic classical solutions, with the primary limitation being the qubit requirement for large systems.
The fact that the QA ground state is a minimum forward error solution carries tremendous advantages for ill-conditioned problems.
In general, these results accompany and extend the recently published results of Borle et al. \cite{borle2019analyzing} and Chang et al. \cite{chang2019quantum}.

Overall, this is a powerful methodology with tremendous applications in the areas of numerical analysis, machine learning, computer security, and general scientific computing.
As the state-of-the-art in quantum annealing hardware continues to evolve, this methodology should scale to increasingly complex problems.

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/biprime.tex
updated work

\begin{table}
    \centering
    \begin{tabular}{c|c c c|c c|c|c}
         Bits & $x^{(1)}$ & $x^{(2)}$ & $x^{(1)}x^{(2)}$ & Logical & Physical & Sq. Error & Occurrence \\ \hline
         2 & 2 & 3 & 6 & 8 & 16 & 0 & 180 \\
         3 & 5 & 7 & 35 & 15 & 69 & 0 & 311 \\
         4 & 11 & 13 & 143 & 24 & 148 & 0 & 35 \\
         5 & 29 & 31 & 899 & 35 & 349 & 0 & 82 \\
         6 & 59 & 61 & 3599 & 48 & 658 & 0 & 7 \\
         7 & 113 & 127 & 14351 & 63 & 1293 & 0 & 3 \\
         8 & 241 & 251 & 60491 & 80 & -- & -- & -- \\
    \end{tabular}
    \caption{Using the squared-error energy function methodology, biprimes with various bits of precision are factored. The columns depict the number of bits, the numbers multiplied and their product, the number of logical and physical qubits required, the minimum squared error (minimum energy) achieved by the quantum annealer, and the number of samples that produced the optimal solution. The number of samples taken grows as $500n$, where $n$ is the number of bits. Correct solutions are found for all embedded problems, however at 8-bits no physical embedding could be discovered for the logical Hamiltonian on the available the hardware. The multiplication circuit constructed with this methodology has a dense $n^2$ subgraph as well as a dense $n$ subgraph, making the number of physical qubits required to embed the Hamiltonian grow faster than the number of logical qubits on the Chimera graph structure of the available quantum annealing hardware.}
    \label{tab:biprime}
\end{table}

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/fig2.tex
updated work

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{fig2.pdf}
\caption{Embedding a quantum XOR gate: this figure illustrates the process of programming a quantum annealer, by embedding an XOR gate. The goal is to implement the circuit $x_1 \oplus x_2 = x_3$.
The truth table in (a) shows all combinations of $x_1$, $x_2$, and $x_3$, with all 4 valid input/output combinations highlighted.
Using the QUBO model, this generates the algebraic system of inequalities shown on the right hand side, where $g$ denotes the ground energy of the system.
Because the row containing all zeros is a valid solution, $g$ must be $0$.
The system of inequalities for (a) has no solution, so an ancillary bit $x_4$ is introduced in (b).
There are $2^4$ ways to assign the value of $x_4$ for the $4$ valid states from (a), some of which produce solvable systems and some of which do not.
By trial and error, the rows highlighted in (b) produce a solvable system, shown on the right hand side.
Solving this system produces the QUBO shown in (c), and applying the isomorphism results in the Hamiltonian below.
}
\label{fig:xor-ex}
\end{figure}

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/linear_ls.tex
updated work

\begin{table}
    \centering
    \begin{tabular}{c|c c|c c|c c}
         $K$ & Logical & Physical & \multicolumn{2}{c|}{\shortstack{\textit{Unsigned}\\ S.E. $\ \ $ Occ.}} & \multicolumn{2}{c}{\shortstack{\textit{Signed}\\ S.E. $\ \ $ Occ.}} \\ \hline
         2 & 8 & 22 & 0.000 & 24 & 0.000 & 2 \\
         3 & 12 & 38 & 0.000 & 23 & 0.000 & 27 \\
         4 & 16 & 80 & 0.005 & 3 & 0.013 & 1 \\
         5 & 20 & 130 & 0.026 & 1 & 0.048 & 1 \\
         6 & 24 & 202 & 0.037 & 1 & 0.048 & 1 \\
         7 & 28 & 267 & 2.604 & 1 & 1.882 & 1 \\
         8 & 32 & 312 & 0.917 & 1 & 0.767 & 1
    \end{tabular}
    \caption{Using the squared-error energy function methodology, a linear system is solved. The columns depict the number of variables \textit{and} equations in a test ($K \times K$ matrix), the number of logical and physical qubits required, the minimum squared error (minimum energy) achieved by the quantum annealer rounded to 2 decimal digits, and the number of samples that produced the best achieved solution. $500 K$ samples were drawn from the quantum annealer for all tests. In this test, the numbers are represented in $4$ bit fixed point notation with $3$ bits after the decimal. The embeddings used by the signed and unsigned systems are identical, only QUBO coefficients vary. Best achievable solution has $0$ error for all $K$.}
    \label{tab:linear_ls}
\end{table}

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/paper.tex
updated work

\documentclass[smallextended]{svjour3}       % onecolumn (second format)

\usepackage{amsmath,amsfonts}  % AMS standard symbols and fonts
\usepackage{url}       % pretty clickable links
\usepackage{graphicx}  % including pictures
\usepackage{tabularx}  % center-aligned columns in table
\usepackage{mathtools} % "ceil" command
\usepackage{breqn}
\usepackage{color}

%
\def\UrlBreaks{\do\/\do-\do\_}

% Make line breaks occur on hyphens in URLs for BibTex.

% declare a "ceil" command
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

%% Command for creating multiplie-line cells inside a table
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

% Declare title and authors
\title{Least Squares Solutions to Polynomial Systems of Equations with Quantum Annealing
\thanks{All authors contributed equally to this work. The author order was determined alphabetically.}}
\author{Tyler H. Chang \and Thomas C. H. Lux \and Sai Sindhura Tipirneni}

\titlerunning{LS Solution to Polynomial Systems}
\authorrunning{Chang, Lux, and Tipirneni}

\institute{T. H. Chang \at
   Virginia Polytechnic Institute and State University, Dept. of Computer Science, Blacksburg, VA, USA, 24061 \\
   \email{thchang@vt.edu}
   \and
   T. C. H. Lux \at
   Virginia Polytechnic Institute and State University, Dept. of Computer Science, Blacksburg, VA, USA, 24061 \\
   \email{tchlux@vt.edu}
   \and
   S. S. Tipirneni \at
   Virginia Polytechnic Institute and State University, Dept. of Computer Science, Blacksburg, VA, USA, 24061 \\
   \email{tsaisindhura@vt.edu}
}

\date{Received: date / Accepted: date}

\begin{document}

% Title area
\maketitle

% Abstract
\begin{abstract}
This work proposes and analyzes a methodology for finding least squares solutions to systems of polynomial equations.
Systems of polynomial equations are ubiquitous in computational science, with major applications in machine learning and computer security (i.e., model fitting and integer factorization).
The proposed methodology maps the squared error function for a polynomial equation onto the Ising-Hamiltonian model, ensuring that approximate solutions (by least squares) to real-world problems can be computed on a quantum annealer even when exact solutions do not exist. Hamiltonians for integer factorization and polynomial systems of equations are implemented and analyzed for both logical optimality and physical practicality on modern quantum annealing hardware.
\end{abstract}

\keywords{Least squares \and Quantum annealing \and Polynomial systems of equations \and Prime factorization \and Root finding}

% Body
\input{1-intro.tex}
\input{2-background.tex}
\input{3-Embedding-Polynomials.tex}
\input{4-analysis.tex}
\input{5-methods.tex}
\input{6-results.tex}
\input{7-conclusion.tex}

\begin{acknowledgements}
This is a pre-print of the article published in Quantum Information Processing. The final authenticated version is available online at:
\url{https://www.doi.org/10.1007/s11128-019-2489-x}.
%The authors would like to thank Wu-chun Feng and Mohamed W.\ Hassan for their council and feedback.
%Also, the authors would like to acknowledge the anonymous reviewers for their helpful comments, which greatly improved this paper.
\end{acknowledgements}

% references section
\bibliographystyle{spmpsci}
\bibliography{paper}

\end{document}

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/poly_ls.tex
updated work

\begin{table}
    \centering
    \begin{tabular}{c|c c|c c|c|c}
         Bits & $x$ & $y$ & Logical & Physical & Sq. Error & Occurrence \\ \hline
         2 & $3/4$ & $3/4$ & 10 & 26 & 0.2070 & 455 \\
         3 & $3/4$ & $3/4$ & 21 & 104 & 0.2070 & 50 \\
         4 & $3/4$ & $13/16$ & 36 & 286 & 0.2061 & 82 \\
         5 & $25/32$ & $25/32$ & 55 & 596 & 0.2005 & 21 \\
         6 & $25/32$ & $49/64$ & 78 & 1216 & 0.2004 & 2 \\
         7 & $99/128$ & $99/128$ & 105 & -- & -- & -- \\
    \end{tabular}
    \caption{Using the squared-error energy function methodology, a polynomial least squares problem is solved. The columns depict the number of bits of precision in the variables, the best representable solution to the least squares problem, the number of logical and physical qubits required, the minimum squared error (minimum energy) achieved by the quantum annealer rounded to 4 decimal digits, and the number of samples that produced the optimal solution. $500n$ samples were drawn from the quantum annealer for all tests, where $n$ is the number of bits of precision in the solution. In this test, the numbers are represented in fixed point notation where all bits are after the decimal. This means that the binary digits encode negative powers of two. The obtained solutions are the best possible solutions that can be achieved for their respective precision, however at 7-bits no physical embedding could be discovered for the logical Hamiltonian on the available the hardware.}
    \label{tab:poly_ls}
\end{table}

### 2020-03-30 [2019-02]_NUMA_(interpolant_error_bound)/Data/Analytic/run_spacing_test.py

@@ -1,0 +1,192 @@
+import os, time, random
+import numpy as np
+from util.system import load, save, Timer
+from util.math import fekete_points, pairwise_distance
+from util.approximate import DelaunayP1, NeuralNetwork
+from util.random import latin
+from util.plot import Plot
+from util.data import Data
+
+# A modified version of the oscillatory test problem from:
+#    'High Dimensional Polynomial Interpolation' by Barthelman et al. (2000)
+# It is a cosine function of the norm of the input vectors. This
+# function has no componentwise structure to take advantage of.
+def oscillatory(x, c=1):
+    if len(x.shape) <= 1:
+        return np.cos(c*np.pi + np.linalg.norm(x))
+    elif len(x.shape) == 2:
+        return np.cos(c*np.pi + np.linalg.norm(x, axis=1))
+    else:
+        class UsageError(Exception): pass
+        raise(UsageError(f"This objective function only supports arrays and matrices, received shape {x.shape}."))
+
+# `True` if the plots and mesaurements should be of "error"
+# `False` if the plots and measurements should be about simplices.
+use_vals = False
+# `True` if Fekete points should be used.
+# `False` if a random Latin hypercube design should be used instead.
+use_fekete = True
+# Random seed.
+seed = 0
+dimension = 20 # 2
+test_size = 1000
+train_sizes = [2 * 4**(i) for i in range(1,6+1)]
+train = train_sizes[-1]
+snr = 0.0
+# The models to test.
+model = DelaunayP1
+
+# Set the test function and declare some objects.
+test_function = oscillatory
+t = Timer()
+m = DelaunayP1()
+# Plot the distirbution of nearest points with increasing data
+p = Plot(f"Delaunay in {dimension} Dimensions", "2 × 4^i data points", "Distance",
+         font_family="times", font_size=25)
+
+if use_vals:
+    errors = []
+else:
+    errors = []
+    vertex_distances = []
+    edge_lengths = []
+    sing_values = []
+for train in train_sizes:
+    if (train + 1 < dimension):
+        if use_vals:
+            errors.append( [] )
+        else:
+            errors.append( [] )
+            vertex_distances.append( [] )
+            edge_lengths.append( [] )
+            sing_values.append( [] )
+        continue
+
+    # Seed each algorithm the same (in case it involved ranndomness).
+    random.seed(seed)
+    np.random.seed(seed)
+
[…]

### 2020-03-30 [2020-01]_IEEESE_(quintic_applied)/code/5-delaunay-predict.py

@@ -1,0 +1,152 @@
+import random, os
+from numpy import array, linspace, zeros
+
+from util.data import Data
+from util.plot import Plot
+from util.stats import cdf_fit
+
+from fits import cdf_points, flat_fit, linear_fit, cubic_fit, quintic_fit
+
+
+# --------------------------------------------------------------------
+if not os.path.exists("[5]-small-readers.pkl"):
+    data = Data.load("[0]-readers.pkl")
+    data.max_display = 17
+    measured = data.shape[1] - 4
+    # Collapse the "throughput" columns into one list.
+    data.pack("throughput")
+    # Sample out the values (only keep 1000) of them.
+    from numpy import linspace
+    keep_count = 1000
+    to_keep = [int(round(i)) for i in linspace(0,measured-1,keep_count)]
+    for row in data:
+        row["throughput"].sort()
+        row["throughput"] = [row["throughput"][i] for i in to_keep]
+    # Unpack the list-column again before saving.
+    data.unpack("throughput")
+    print(data)
+    data.save("[5]-small-readers.pkl")
+else:
+    data = Data.load("[5]-small-readers.pkl")
+
+
+# --------------------------------------------------------------------
+#  Load the data.
+data.max_display = 17
+data.pack("throughput")
+
+# --------------------------------------------------------------------
+#  Compute the Delaunay weights for each row.
+print("Computing Delaunay weights for each row..")
+data["extrapolated"] = False
+data["indices"] = None
+data["weights"] = None
+x = array( data[:,:4].to_matrix() )
+shift = x.min(axis=0)
+x -= shift
+scale = x.max(axis=0)
+x /= scale
+indices = list(range(len(x)))
+from util.approximate import Delaunay
+to_remove = set()
+for i in range(len(x)):
+    print(i, x[i]*scale + shift)
+    row = x[indices.pop(i)]
+    # Get the indices and weights for this row.
+    m = Delaunay()
+    m.fit(x[indices])
+    try:
+        ids, wts = m(row, allow_extrapolation=False)
+        data[i,"indices"] = [indices[_] for _ in ids]
+        data[i,"weights"] = list(wts)
+    except: data[i, "extrapolated"] = True
+    # Re-insert the popped index.
+    indices.insert(i,i)
+
+print(f"  data shape: {data.shape}")
+
+
+# --------------------------------------------------------------------
+#   Generate samples to use for fits.
+#   Compute the distribution fits.
[…]

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/1-intro.tex
updated work

\section{Introduction}

Quantum annealing (QA) provides a practical framework to realize adiabatic quantum computing (AQC), where AQC is a general purpose quantum computing framework that is known to be equivalent to the quantum gate model \cite{aharonov2008adiabatic}.
In the quantum gate model, quantum state vectors evolve through time via the application of quantum logic gates (unitary operators) \cite{steane1998quantum}.
On the other hand, AQC conceptually produces the solution to a problem through a single ``annealing'' operation.
This is achieved by transitioning from an initial Hamiltonian to an arbitrary Hamiltonian that encodes the problem, using the adiabatic theorem.
The standard implementation of QA only solves a subset of AQC problems because the allowed operators only span a subspace of all unitary operators.
Specifically, QA is limited to annealing over the span of the stoquastic Pauli-Z matrices, whose eigenstates are strictly real valued \cite{kadowaki1998quantum}.

While the general methodology for embedding classical programs in QA is understood, QA systems can still be tedious to program. D-Wave Systems has improved accessibility through its Ocean software suite that contains high-level programming modules for network science and constraint satisfaction problems.
Pakin also provides tools for embedding high-level constraint logic programs, using a standard cell library \cite{pakin2018performing}.
However, these approaches are inefficient for linear algebra and mathematical programming applications, making hand-crafted solutions preferred in many cases.
Linear algebra problems are ubiquitous in computational and data science, with some notable applications being least squares fitting, solving large sparse or dense systems, and even neural network training.
Such applications would certainly garner interest in high-profile fields such as machine learning, where the term ``quantum machine learning'' was coined by \cite{biamonte2017quantum}, and a quantum variational autoencoder is described in \cite{khoshaman2018quantum}.

In this work, a novel methodology is proposed for least-squares minimization of polynomial systems of equations via QA.
This broad class of problems is often referred to as sum of squares (SOS) polynomial minimization, and is a general case of both the least squares problem and the polynomial real-valued root-finding problem.
Solutions are obtained by mapping the squared error function for any multivariate polynomial onto a Hamiltonian embedding.
To make the proposed solution accessible, a high-level programming framework is provided, which automatically handles low-level details such as fixed-point binary encoding, quadratization, and both physical and logical (minor) embedding of the Ising-Hamiltonian model.
This framework is standalone, but interfaces with D-Wave's SAPI API.
One practical benefit of the squared error methodology is that the minimum forward error solution will correspond to the ground state for the Hamiltonian, and lower forward error solutions will have lower energies.
This causes the annealer to tend towards low energy solutions even when it fails to find the true global minimum.
Another benefit of both the methodology and the framework is that the fixed-point precision of each variable can be independently specified, and so this work immediately extends to nonlinear integer programming problems (such as integer factorization) and mixed-precision problems.

The remainder of this paper will proceed as follows.
Section 2 will present relevant background in QA, programming and algorithmic concerns, and related works.
Section 3 will introduce the proposed methodology for mapping multivariate polynomial systems of equations/SOS problems to the Ising-Hamiltonian model.
Section 4 contains a brief analysis of the proposed algorithm in terms of several performance concerns introduced in Section 2.
Section 5 contains technical details of the implementation, with a focus on the practicalities of the D-Wave machine.
Section 6 shows empirical results for $n$-bit integer factorization, least squares minimization of a system of polynomial equations, and solutions to linear systems of equations on the state-of-the-art D-Wave 2000Q quantum annealer.
Finally, Section 7 contains a conclusion and brief discussion.

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/2-background.tex
updated work

\section{Background}
\label{sec:background}

At its foundation, QA utilizes the physical principle of quantum tunneling \cite{ray1989sherrington} to find grounds states for the Ising-Hamiltonian model
\begin{equation}
\mathcal{H}(\text{\boldmath$\sigma$}) = \sum_{i=1}^N h_i\sigma_i + \sum_{i=1}^N \sum_{j=i+1}^N J_{i,j} \sigma_i \sigma_j
\label{eq:hamiltonian}
\end{equation}
where $N$ denotes the number of qubits (quantum bits) involved in the computation (indexed $q_1,\ldots, q_N$), $\sigma_i \in \{-1,1\}$ denotes an eigenstate of $q_i$, $h_i$ denotes the bias of $q_i$, and $J_{i,j}$ denotes the coupler strength between the entangled pair $q_i$ and $q_j$.

Similarly as in AQC, QA achieves the solution to an arbitrary Hamiltonian by beginning from the ground state of an initial Hamiltonian $\mathcal{H}_0$, whose solution is known.
$\mathcal{H}_0$ is then gradually evolved through time into the problem Hamiltonian $\mathcal{H}_1$, via a mathematical homotopy of the form 
\begin{equation}
\mathcal{H}_s(\text{\boldmath$\sigma$}) = (1-s)\mathcal{H}_0(\text{\boldmath$\sigma$}) + s\mathcal{H}_1(\text{\boldmath$\sigma$}).
\label{eq:homotopy}
\end{equation}
Here, the time dependent variable $s$ transitions from $0$ to $1$ during the annealing time \cite{albash2018adiabatic}.

If the transformation (\ref{eq:homotopy}) is adiabatic, meaning that the surrounding temperature is near absolute zero and the rate of change in $\mathcal{H}_s(\text{\boldmath$\sigma$})$ is smaller than the energy gap $\Delta$ between the ground state of $\mathcal{H}_1$ and the next lowest energy eigenstate (i.e., the first excited state), then this process results in a solution to $\mathcal{H}_1$ with probability one.
Consequently, the annealing time $T$ required to find a true ground state with high probability is proportional to $\mathcal{O}(\Delta^{-2}|\log \Delta|^{6\alpha})$ \cite{elgart2012note}, where $\alpha$ is a Gevrey index, which describes the smoothness of the transition.
Rather than increasing the annealing time $T$ (approximately) quadratically as $\Delta$ shrinks, a more practical approach is to fix $T$ and instead increase the number of samples (i.e., runs of QA).
The probability of not achieving the ground state for {\it any} of the runs decreases exponentially with repeated independent samples, while the probability of a single anneal resulting in the ground state decreases quadratically with a decaying $\Delta$ \cite{kadowaki1998quantum}.
As a result, an exponentially decaying $\Delta$ only requires linearly increasing the number of samples to achieve the same likelihood of a ground state solution \cite{jiang2018quantum}.

Minimization of an arbitrary Ising-Hamiltonian is classically NP-complete \cite{barahona1982computational}, and although QA currently remains less efficient than classical heuristic techniques (such as simulated annealing), both theory \cite{santoro2002theory} and empirical research \cite{boixo2014evidence} suggest that QA will eventually overtake classical techniques as the preferred method for minimizing Ising-Hamiltonian equations.
Currently, the state-of-the-art in QA is pioneered by D-Wave systems, which produces both hardware and software for minimizing (\ref{eq:hamiltonian}) via QA.
The current flagship D-Wave machine is capable of performing QA with over 2000 qubits sparsely connected according to the Chimera graph topology, in which each qubit is connected to at most 6 other qubits. 
The next generation of machines will support QA with over 5000 qubits on a new Pegasus connection topology \cite{boothby2018next}, in which each qubit is connected to at most 15 other qubits.

D-Wave's QA hardware does not implement perfect adiabatic evolution and must account for real world practicalities such as outside noise, bias leakage, and short annealing times.
Therefore, it is useful to differentiate between the logical and physical Hamiltonians.
The {\it logical} Hamiltonian (often referred to in literature as the minor embedding) will be a functional of the form (\ref{eq:hamiltonian}), where $h_i$ and $J_{i,j}$ are real numbers.
Note that not all logical Hamiltonians can be embedded on a current D-Wave machine.
D-Wave requires that $h_i \in [-2,2]$, requires that $J_{i,j}\in[-1,1]$, uses a quantization step size of approximately $0.01$, and only allows for nonzero couplings between qubit pairs $q_i$ and $q_j$ that are connected according to some graph topology $G$ \cite{karimi2019practical}.
Therefore, it is important to distinguish the {\it physical} Hamiltonian as shown below, which could be implemented on the current D-Wave hardware.
\begin{equation}
\Tilde{\mathcal{H}}(\text{\boldmath{$\sigma$}}) = \sum_{i'=1}^{N'}\Tilde{h}_{i'}\sigma_{i'} + \sum_{(i',j') \in G} \Tilde{J}_{i',j'} \sigma_{i'} \sigma_{j'}
\label{eq:physical}
\end{equation}
Here, $N' \geq N$ is the number of physical qubits required to synthesize all the required connections through the ``chaining'' process, as shown in Figure \ref{fig:chains}.
After the chaining process is completed, the entire Hamiltonian is rescaled by some factor $\kappa$, which is the largest factor such that all resulting $\Tilde{h}_{i'} \in [-2,2]$ and $\Tilde{J}_{i',j'} \in [-1,1]$.
D-Wave provides embedding tools, which use heuristics to construct the chains and compute the corresponding constant $\kappa$.

Due to physical hardware limitations, such as the quantization step size, manufacturing imperfections, outside noise, and bias leakage, the problem solved by the D-Wave annealer is generally not identical to the physical Hamiltonian in (\ref{eq:physical}).
Rather, the D-Wave solves a perturbed model, as shown below.
\begin{equation}
\hat{\mathcal{H}}(\text{\boldmath$\sigma$}) = \sum_{i'=1}^{N'} (\Tilde{h}_{i'} + \delta_{i'})\sigma_{i'} + \sum_{(i',j')\in G} (\Tilde{J}_{i',j'} + \delta_{i',j'}) \sigma_{i'} \sigma_{j'}
\label{eq:perturbed}
\end{equation}
The problem (\ref{eq:perturbed}) will have the same ground state as (\ref{eq:physical}) if $\|\Tilde{\mathcal{H}} - \hat{\mathcal{H}}\|_\infty \leq \Delta'$, where $\Delta'$ is the size of the energy gap between the ground and first excited state of $\Tilde{\mathcal{H}}$, and $\|\cdot\|_\infty$ denotes the $H_\infty$ operator norm.
Note that the ratio $\|\Tilde{\mathcal{H}} - \hat{\mathcal{H}}\|_\infty/\Delta'$ can be considered analogous to the conditioning of the problem in this context.

\input{fig1.tex}

To measure the optimality of an implementation, it is typical to consider both the number of physical qubits $N'$ required to embed the Hamiltonian and the size of the physical energy gap $\Delta'$.
In general, to implement a classical circuit on a QA machine, the circuit must be transformed into a Hamiltonian whose ground states correspond to valid input/output combinations as illustrated in Figure \ref{fig:xor-ex}.
During this process, it is common to introduce ancillary qubits, which are important terms in the Hamiltonian, but do not contain any useful information on input or output.
Using basic gates generated via the above process, it is possible to implement general classical circuits using the additive property of Hamiltonians \cite{pakin2018performing}.
While this methodology is convenient, it is wasteful in terms of ancillary bit requirements, and therefore a hand-crafted solution is preferable for most problems.

\input{fig2.tex}

Using the transformation $\sigma_i = 2x_i - 1$ and dropping constant terms (which offset the energy landscape, but do not affect the locations of the minima), the Ising model is isomorphic to the quadratic unconstrained binary optimization (QUBO) model
\begin{equation}
C(\mathbf{x}) = \sum_{i=1}^N a_i x_i + \sum_{i=1}^N \sum_{j=i+1}^N b_{i,j} x_i x_j
\label{eq:QUBO}
\end{equation}
where $x_i\in\{0,1\}$ are abstract binary variables, $a_i = 2\left(h_i-\sum_j J_{i,j}\right)$, and $b_{i,j} = 4J_{i,j}$.
For dealing with binary representations of fixed-point numbers, the QUBO model is more convenient and will be favored in Section 3.
However, for performance analyses (Section 4) it is more accurate to consider the Hamiltonian model, which better describes the physical embedding.

The general problem of finding a least squares solution to a system of polynomial equations is equivalent to globally solving a multivariate SOS polynomial minimization problem.
This class of problems should not to be confused with the class of polynomial least squares problems, which refer to the fitting of a polynomial to data by solving a linear least squares problem.
Since SOS problems are known to be NP-hard, classical solutions focus on iterating toward a solution by solving sequences of linear matrix inequality problems \cite{lasserre2001global}.
In the special case where the system of polynomial equations is consistent, the problem can often be efficiently solved using homotopy methods \cite{watson1997algorithm}.
In the special case of linear and least squares systems, these problems can be solved with cubic complexity using standard matrix factorization techniques \cite{golub2012matrix}.

It should be noted that the current generation of D-Wave machines are not capable of solving any meaningfully large SOS problems.
In fact, the elementary operation of multiplication/factorization is considered difficult for these D-Wave machines, due to the practical limitations of D-Wave's technology \cite{andriyash2016boosting}.
Related to some of the applications discussed, a good deal of work has been put into integer factorization, with a focus on security applications.
While Shor's algorithm has become synonymous with quantum factorization, quantum annealing requires a different model \cite{shor1999polynomial}.
Early work in quantum factorization was pioneered by Peng et al., who produced a general model for integer factorization \cite{peng2008quantum}.
Jiang et al.\ expanded on this work in the special case of biprime factorization \cite{jiang2018quantum}.
Dridi et al.\ proposed an alternative approach to biprime factorization using Gr\"{o}bner bases \cite{dridi2017prime}.
In general, to factor an $n$-bit binary integer, these approaches require $\mathcal{O}(n^2)$ qubits and result in an exponentially decaying physical energy gap $\Delta'$.
It is also worth mentioning a hybrid quantum-classical algorithm for nonlinear integer programming, proposed by Alghassi et al.\ \cite{alghassi2019graver}.

For the special case of solving systems of linear equations, quantum algorithms have been proposed for both the gate model \cite{PhysRevLett.103.150502} and AQC model \cite{PhysRevLett.122.060504,PhysRevA.99.012320} that achieve exponential speedup in cases where the associated matrix has a low condition number.
For QA, there are two closely related works that were recently published, both of which achieve a QA solution by embedding the error function in the Ising-Hamiltonian.
Borle et al. \cite{borle2019analyzing} propose this methodology for solving linear systems of equations and least squares problems, but refrain from analyzing higher-order systems.
Chang et al. \cite{chang2019quantum} propose this method for finding least squares solutions to general systems of polynomial equations, but use a slightly different binary encoding technique, where fixed-precision signed arithmetic is achieved via a linear transformation of unsigned binary integers. The present work accompanies \cite{borle2019analyzing,chang2019quantum} by providing both an extension to higher order systems and an easily accessible coding framework for solving arbitrary polynomial SOS problems on a QA system.

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/4-analysis.tex
updated work

\section{Algorithm Analysis}

This section provides an overview and analysis of the important factors to consider when encoding polynomial systems of equations onto a quantum annealing architecture (resembling modern D-Wave hardware).

\subsection{Energy Gap}

The range of energies grows and shrinks exponentially with the bits of precision $n$, yielding a $\mathcal{O}(2^{-n})$ decay in the physical energy gap $\Delta'$ after the necessary rescaling operations.
When mixed-precision variables interact, $\Delta'$ is determined by the effective precision $\hat n$, which is the difference between the largest and smallest nonzero absolute value representable among all the variables in the system.
The number of additions $\alpha$ and number of multiplications $\mu$ have, respectively, linear and exponential effects on the maximum energy of the logical Hamiltonian.
Therefore, $\Delta'$ shrinks inverse linearly with respect to $\alpha$ and exponentially with respect to $\mu$.
Similarly as with addition, the number of equations $m$ in a system decreases $\Delta'$ at most inverse linearly since the equations do not directly interact, but may share some terms in common.
Combined, this yields a rate of decay for the physical energy gap (after rescaling) of
$$\Delta' \approx \mathcal{O}\bigg(\frac{1}{\alpha m 2^{\hat n \mu}}\bigg).$$
In general, it can be seen that $\Delta'$ will be primarily determined by the exponential decay $2^{-{\hat n}\mu}$.

\subsection{Number of Qubits}

Other than the number of qubits required to represent the binary numbers, some ancillary qubits may be introduced into the logical Hamiltonion during both quadratization and physical embedding.
Quadratization is performed analytically by the high-level tool described in Section 5, and so the ancillary qubit requirements due to quadratization can be easily analyzed.
When addition or subtraction is performed, no additional quadratization is required.
However, multiplication between two variables with $n_0$ and $n_1$ bits of precision requires $n_0 \times n_1$ ancillary qubits (one for each quadratization).
Given a maximum of $\mu$ multiplications per individual polynomial (single equation) and $n$ bits of precision, the number of ancillary qubits introduced through quadratization will be $\mathcal{O}(m n^\mu)$.
Notice that increasing the number of equations $m$ can at most linearly increase the number of ancillary quadratization terms required (when no variable products are shared between equations).
Also notice that the number of variables $d$ does not affect the number of ancillary qubits due to quadratization, for a fixed $\mu$.

When embedding the physical Hamiltonian for a polynomial system, two factors will dominate the number of necessary qubits.
First, the annealer must generate a fully connected subgraph of size $n^{(\mu + 1)/2}$.
Second, given some variable $\hat{\mathbf{x}}$ is multiplied by $k$ other distinct variables throughout a system, $\hat {\mathbf{x}}$ will have at least $k$ connections to each of its bits.
The modern Chimera graph structure is not conducive to fully-connected graphs nor small sets of qubits with high graph centrality, hence experimental results show a growth rate greater than $k n^\mu$ for large problems.

\subsection{Sensitivity to Perturbation}

Given the perturbed Hamiltonian model (\ref{eq:perturbed}), it is reasonable
to wonder how the perturbations $\delta_i$ and $\delta_{i,j}$ will affect
the quality of the solution.
In particular, one might wonder whether the effects could be great enough
so that the ground state of the perturbed problem is no longer the ground
state of the original problem.

Consider first an arbitrary physical Ising-Hamiltonian $\Tilde{\mathcal{H}}$ involving $N'$ physical qubits, as defined in (\ref{eq:physical}) and its corresponding perturbed Hamiltonian $\hat{\mathcal{H}}$ as in (\ref{eq:perturbed}).
Suppose that $|\delta_i|, |\delta_{i,j}| < \delta$, where $\delta>0$ is a constant that represents the largest magnitude perturbation possible on the current hardware.
Combining equations (\ref{eq:physical}) and (\ref{eq:perturbed}), applying the upper bounds $\delta$, and combining like terms yields
$$
\|\Tilde{\mathcal{H}} - \hat{\mathcal{H}}\|_\infty
= \sup_{\sigma} |\mathcal{H}(\sigma) - \hat{\mathcal{H}}(\sigma)|
\leq \sup_\sigma \left|\sum_{i'=1}^{N'} \delta \sigma_{i'} +
\sum_{i'=1}^{N'}\sum_{(i',j')\in G}\delta\sigma_{i'}\sigma_{j'} \right|.
$$

In the worst case, assuming nonzero connections between all physical qubits, the right hand side is upper bounded by $(N' + N' L')\delta$, where $L'$ is the number of connections per node in the physical graph $G$.
For the Chimera graph, each node has at most $6$ undirected edges.
So for the D-Wave Chimera architecture, the condition that guarantees a correct ground state for the physical Hamiltonian $\hat{\mathcal{H}}$ is $7N' \delta < 2\Delta'$.

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/6-results.tex
updated work

\section{Results}

In this section, Hamiltonian embeddings for error functions of the form (\ref{eq:single-error}) and (\ref{eq:system-error}) are implemented for solving biprime factorization, polynomial SOS problems, and linear systems of equations.
For the first two problem, the focus is placed on increasing the bit precision $n$, since
it was shown in Section 4 that increasing the relative precision causes the physical energy gap $\Delta'$ to decay exponentially.
For problems with $n$ bits of precision per variable, the number of samples was assigned to $500n$ (a linear increase).
As outlined in Section \ref{sec:background}, this should maintain a (nearly) constant probability of achieving the ground state under ideal conditions.
However, since the problem conditioning is largely determined by $\Delta'$, it is somewhat concerning that the perturbations in (\ref{eq:perturbed}) could have a significant negative impact on the computation for large precision $n$.
In fact, the problem conditioning does seem to cause some significant decay in the probabilities.
However, the presented results achieve the ground state for every problem that could be embedded on the current D-Wave hardware, indicating that the primary limiting factor is the number of physical qubits $N'$ required to embed the physical Hamiltonian.

For the linear system problem, the focus is shifted to how the proposed methodology is affected by increasing the number of equations and variables, with a fixed four-bit precision.
Since there is no multiplication between variables in a linear system, the analysis results of Section 4 suggest that these types of problems should scale better on the D-Wave 2000Q hardware.
However, somewhat surprisingly, the D-Wave is unable to find exact solutions for relatively small linear systems, even when the true solution is exactly representable for the binary encoding.

\subsection{Factorization}
\label{sec:factorization}

To factor a biprime $M = x^{(1)}x^{(2)}$, where $x^{(1)}$ and $x^{(2)}$ are $n$-bit unsigned integers, the corresponding polynomial is 
$$
p\left(x^{(1)},x^{(2)}\right) = x^{(1)}x^{(2)} - M
$$
and the squared error function is given by
\begin{equation}
E\left(x^{(1)},x^{(2)}\right) = \left(x^{(1)}x^{(2)} - M\right)^2.
\label{eq:primefact}
\end{equation}
Creating $n^2$ ancillary qubits $y_{i,j} = x^{(1)}_i \wedge x^{(2)}_j$, this energy function can be embedded into a QUBO as described in Section 3.
Note that the best available techniques also require $\mathcal{O}(n^2)$ ancillary qubits for prime factorization \cite{dridi2017prime,jiang2018quantum,peng2008quantum}. 
The results for embedding this QUBO as a physical Hamiltonian (as described in Section 5) and running on the D-Wave hardware are shown in Table \ref{tab:biprime}.
Similar to the work done by Jiang et al.\ for biprime factoring, this methodology encodes $M$ directly into the Hamiltonian as a constant, rather than wasting extra qubits to store $M$ as a variable.
In fact, Jiang et al.\ used a nearly identical methodology for arbitrary integer factorization, but favored a multiplication table for factoring biprimes.

\input{biprime.tex}

The most recent work in biprime factorization solves a slightly larger problem \cite{dridi2017prime,jiang2018quantum} by assuming properties of the factors.
In contrast, this methodology can factor arbitrary integers (which may not be biprimes) while still achieving the same number of bits of precision as those best known results.
As presented, arbitrary integer factorization takes the same form as biprime factorization, but has more candidate solutions (and an innately higher likelihood of success).
For that reason, only biprimes are considered here.
The correct factorization is obtained up to $M=14351$ (whose prime factors are both $7$-bit unsigned integers) by minimizing (\ref{eq:primefact}), which is also the largest problem that can be successfully embedded on current quantum annealing hardware.
At $n=8$ bits of precision, the Hamiltonian failed to embed onto the D-Wave 2000Q system, due to size limitations.

\subsection{Nonlinear Least Squares}

\input{fig3.tex}

Let $\mathbf{v} = [x,y]^T \in \mathbb{R}^2$, and consider the following system of polynomial equations.
\begin{equation}
\begin{array}{rcl}
    xy &=& 1\\
    x^2+y^2 &=& 1\\
    x - y &=& 0
    \end{array}
    \label{eq:ls-system}
\end{equation}
This results in the squared energy function
$E(\mathbf{v}) = (xy-1)^2 + (x^2+y^2 - 1)^2 + (x-y)^2$.
A graph of the system (\ref{eq:ls-system}) is shown in Figure \ref{fig:ls-graph}.
To find the solution in the first quadrant, $x$ and $y$ are encoded as unsigned $n$-bit numbers each with an exponent of $e = -n$ using the encoding (\ref{eq:binary-encoding}).
The true solution is $x=y\approx 0.77$, with a minimum squared residual of $r^2 = 0.2$.
For $n=2$, the physical embedding is shown in Table \ref{tab:poly_ls_embedding}.
As seen in Table \ref{tab:poly_ls}, for $n>2$, the physical qubit requirement appears to grow super-linearly but sub-exponentially with the logical qubit requirement.

\input{poly_ls_embedding.tex}

The results collected on the D-Wave machine are shown in Table \ref{tab:poly_ls}.
For all tests through up to $6$ bits of precision, the closest representable solution is found.
For $n=7$ bits of precision, again, the Hamiltonian failed to embed due to size constraints.

\input{poly_ls.tex}

\subsection{Linear System of Equations}

Thus far, the presented results have featured an integer-valued and nonlinear problem, and focused on increasing the bit precision $n$.
In general, integer valued and nonlinear problems are interesting because they do not classically admit analytic solutions, as commented on by Chang et al. \cite{chang2019quantum}.
By contrast, as discussed by Borle et al. \cite{borle2019analyzing}, the time complexity involved in embedding linear and least squares systems of equations into a Hamiltonian rivals the time complexity of solving the system using classical techniques.
However, linear and least squares systems of equations remain interesting problems because of their wide usage in data science and applied mathematics.
Furthermore, the proposed method could still be useful in the context of large sparse systems.

In order to study how the proposed method scales with increasing numbers of variables and equations, consider a $K \times K$ linear system of equations for $K = 2,\ldots,8$.
To demonstrate signed linear algebra, one set of experiments is carried out using the two's complement binary encoding in (\ref{eq:binary-encoding}) with no bits of precision before the decimal and three bits of precision after the decimal.
To demonstrate unsigned linear algebra, another set of experiments is carried out using the unsigned binary encoding from (\ref{eq:binary-encoding}) with one bit of precision before the decimal and three bits of precision after the decimal.
The results are shown in Table \ref{tab:linear_ls}.
Note that under the proposed methodology, it is not apparently ``easier'' to solve a trivial system.
Therefore, for convenience of analysis and reproducability, the embedded system in the signed case is of the form $\mathbf{B}_0 \mathbf{v}=\mathbf{0}$, where the coefficients in the matrix $\mathbf{B}_0$ are randomly generated numbers in the range $(-1,1)$ rounded to the nearest multiple of $0.25$.
In the unsigned case, a similar system $\mathbf{B}_1 \mathbf{v} = \mathbf{c}_1$ is constructed, such that the solution is instead $\mathbf{v}=\mathbf{1}$.
All the presented problems use four bit variables, which is standard for the D-Wave 2000Q system \cite{borle2019analyzing}, and are run for $500 K$ samples.

\input{linear_ls.tex}

Note that Borle et al. \cite{borle2019analyzing} were able to solve least squares systems of size $100\times 8$ to reasonable accuracy, by increasing the annealing time from 20 microseconds (the default value) up to 50 microseconds and performing 10,000 samples.
For the experiments shown here, both the signed and unsigned linear systems are solved and an exact solution is obtained up to a system of size $3\times 3$. 
For larger systems, an inexact approximate solution is still obtained.
Although Section 4 indicates that the proposed methodology should scale better with the number of variables $d$ than with the precision $n$, these results suggest the opposite, since the only limiting factor in Sections 6.1 and 6.2 was the size of the D-Wave 2000Q machine.
It is worth noting that the physical qubit requirements in Table 3 do not challenge the size limitations.

The primary objective of these experiments is to demonstrate generality and provide an accessible method of encoding polynomial SOS problems on modern QA systems.
Although it was not demonstrated, it is reasonable to assume that improved results can be obtained by increasing the annealing time and sample count.

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/fig1.tex
updated work

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{fig1a.pdf}$\qquad$
\includegraphics[width=0.35\textwidth]{fig1b.pdf}
\caption{Finding a physical embedding: this figure shows a toy 32 qubit system demonstrating the Chimera graph structure (left) and a synthesized dense network on an individual unit cell (right). The Chimera graph consists of bipartite 8-qubit unit cells as shown, with sparse connections between cells.
To synthesize a dense 4-qubit network on any individual unit cell, negative coupler weights (shown in bold) are added.
I.e., $J_{0,4} = J_{1,5} = J_{2,6} = J_{3,7} < 0$.
}
\label{fig:chains}
\end{figure}

### 2020-03-30 [2019-05]_QIP_(quantum_ls)/poly_ls_embedding.tex
updated work

\begin{table}
    \centering
    \begin{tabular}{c|c c c|c}
      Logical Bit & Embedding &$\quad$& Logical Bit & Embedding \\ \cline{1-2}\cline{4-5}
      1 & 1376, 1383, 1248 && 6  & 1390, 1382, 1384 \\
      2 & 1380, 1379, 1251 && 7  & 1389, 1381        \\
      3 & 1261, 1258, 1253 && 8  & 1391, 1257, 1385 \\
      4 & 1249, 1377       && 9  & 1252, 1250       \\
      5 & 1388, 1386       && 10 & 1255, 1263
    \end{tabular}
    \caption{The exact embedding used in the polynomial least squares problem for $2$ bits of precision is displayed with the logical bit index in the \textit{Logical Bit} column and the physically embedded bit indices on hardware (all of which are chained to be equal) in the \textit{Embedding} column. Notice that the Chimera graph structure requires 25 physical qubits to represent the 10 logical qubits. The physical embedding was heuristically chosen using { \tt minorminer.find_embedding}, which constructed chains of 2 and 3 qubits to match the logical Hamiltonian to the connectivity of the Chimera graph. Notice also that the physical qubit indices are an automatically selected subset of the 2000 qubits available on the D-Wave 2000Q machine. The logical QUBO uses bits 1--4 to store the variables $x$ and $y$ and bits 5--10 are ancillary bits introduced through quadratization.}
    \label{tab:poly_ls_embedding}
\end{table}

%% Using embedding with 25 qubits:
%%  0 [1376, 1383, 1248]
%%  1 [1380, 1379, 1251]
%%  2 [1261, 1258, 1253]
%%  3 [1249, 1377]
%%  4 [1388, 1386]
%%  5 [1390, 1382, 1384]
%%  6 [1389, 1381]
%%  7 [1391, 1257, 1385]
%%  8 [1252, 1250]
%%  9 [1255, 1263]

### 2020-03-30 [2019-11]_HPC_(quintic_spline)/paper/paper.tex
updated work

@@
 %****************************************************************************
+
+%% ------ Package for adding bars into margin next to comments. ------
+%% Command for changing the color of text within a block.
+\usepackage{xcolor}   % Color package for coloring text.
+\usepackage[normalem]{ulem} % For strikethrough text and the \sout{} command.
+\usepackage[xcolor,rightbars]{changebar}
+\renewcommand{\changebarsep}{5mm}   % Push the bar farther into the margin.
+\renewcommand{\changebarwidth}{1mm} % Increase the width of the bar.
+\cbcolor{red} % Set the color of the change bars to red.
+%% \newcommand\changed[2][]{#2} % <- hidden
+\newcommand\changed[2][]{{\cbstart{\color{red}#2}\cbend}} % <- no old text
+% \newcommand\changed[2][]{{\cbstart{\color{red}#2}{\color{gray}\sout{#1}}\cbend}}
+%% -------------------------------------------------------------------
 
@@
 
-An algorithm for computing monotone order six piecewise polynomial
-interpolants is proposed. Algebraic constraints for enforcing
-monotonicity are provided that align with quintic monotonicity
-theory. The algorithm is implemented, tested, and applied to several
-sample problems to demonstrate the improved accuracy of monotone
-quintic spline interpolants compared to that for existing monotone
-cubic spline interpolants.
+A novel algorithm for computing monotone order six piecewise
+polynomial interpolants is proposed. Algebraic constraints for
+enforcing monotonicity are provided that align with quintic
+monotonicity theory. The algorithm is implemented, tested, and applied
+to several sample problems to demonstrate the improved accuracy of
+monotone quintic spline interpolants compared to the previous
+state-of-the-art monotone cubic spline interpolants.
+
+%% An algorithm for computing monotone order six piecewise polynomial
+%% interpolants is proposed. Algebraic constraints for enforcing
+%% monotonicity are provided that align with quintic monotonicity
+%% theory. \changed{The prior state of the art algorithm for constructing
+%%   monotone spline interpolants is piecewise cubic, matching function
+%%   values as well as first derivatives. The algorithm proposed here
+%%   matches function value, first, and second derivatives achieving
+%%   higher accuracy for $C^2$ functions while producing an approximation
+%%   with continuous first derivative.} The \changed{proposed} algorithm
+%% \changed{for constructing a monotone quintic interpolating spline} is
+%% implemented, tested, and applied to several sample problems to
+%% demonstrate the improved accuracy of monotone quintic spline
+%% interpolants compared to that \changed{of} existing monotone cubic
+%% spline interpolants.
 
@@
 Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate nonparametric approximations in statistics \cite{knott2012interpolating}.
-While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.
+While polynomial interpolants and regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.
 
@@
 As demonstrated in Figure \ref{fig:spline_demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two.
+Hence, only even order (odd degree) interpolating splines are broadly applicable.
 The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on quintic splines.
@@
 Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by
-
 \begin{align*}
@@
 \end{align*}
-
 Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$.
@@
 \citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals.
-Consider the terms $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$.
-If $f(k_i) = f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$.
+If $f(k_i) = f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$, otherwise let $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$.
 Monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.
@@
   \centering
-  \includegraphics[width=0.45\textwidth]{cubic_projection_demonstration}
+  \includegraphics[width=0.40\textwidth]{cubic_projection_demonstration}
   \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone.
@@
 Without loss of generality, the algorithms will only consider the monotone increasing (nondecreasing) case.
-The monotone decreasing (nonincreasing) case is handled similarly.
+The monotone decreasing case is handled similarly.
 Algorithm \ref{alg:check-monotone} checks monotonicity,
@@
 
-These algorithms make use of first and second derivatives of the approximated function.
+These algorithms make use of first and second derivatives of the approximated function where $D$ denotes the differentiation operator.
 In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values.
@@
 
-Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have $\mathcal{O}(1)$ runtime  assuming a fixed level of relative precision.
+Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have $\mathcal{O}(1)$ runtime assuming a fixed level of precision is chosen beforehand.
 A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}).
-The search requires a fixed number of steps to achieve any predetermined relative precision.
+The search requires a fixed number of steps to achieve any predetermined relative precision, since its accuracy is predetermined and not a function of the problem at hand.
 Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
@@
       \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
+      end for
       \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
     \item while $($size$($queue$) > 0)$
-    \item \indented{$(x_j$, $x_{j+1}) := $ dequeue}
-    \item \indented{if $\big($not is\_monotone$(x_j$, $x_{j+1}$, $f)\big)$}
-    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) -{}$step\_size$(j)$}}
-      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) -{}$step\_size$(j+1)$}}
+    \item \indented{$(k_j$, $k_{j+1}) := $ dequeue}
+    \item \indented{if $\big($not is\_monotone$(k_j$, $k_{j+1}$, $f)\big)$}
+    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(k_j) := Df(k_j) -{}$step\_size$(j)$}}
+      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(k_{j+1}) := Df(k_{j+1}) -{}$step\_size$(j+1)$}}
       \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
 In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
-    \item \indented{\indented{make\_monotone$(x_j,\ x_{j+1},\ f)$}}
+    \item \indented{\indented{make\_monotone$(k_j,\ k_{j+1},\ f)$}}
     \item \indented{\indented{changed$(j) :=$ TRUE}}
       \indented{\indented{changed$(j+1) :=$ TRUE}} \label{step:changed}
-    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not\in$ queue$\big)$ enqueue $\bigl((x_{j-1}, x_j)\bigr)$}}
-      \indented{\indented{enqueue $\bigl((x_j, x_{j+1})\bigr)$}}
-      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((x_{j+1}, x_{j+2})\bigr)$}} \label{step:requeue}
+    \item \indented{\indented{if $\big(j > 1$ and $(k_{j-1}, k_j) \not\in$ queue$\big)$ enqueue $\bigl((k_{j-1}, k_j)\bigr)$}}
+      \indented{\indented{enqueue $\bigl((k_j, k_{j+1})\bigr)$}}
+      \indented{\indented{if $\big(j+1 < n$ and $(k_{j+1}, k_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((k_{j+1}, k_{j+2})\bigr)$}} \label{step:requeue}
       \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to \texttt{queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}
@@
 
-Given the potential for successive monotonicity fixes across all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ breakpoints and shrink factor $s$.
+If successive monotonicity fixes are required for all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ breakpoints and shrink factor $s$.
 In practice this worst case has been observed to be unlikely.
@@
 
-\begin{figure}
+\begin{figure}[h]
   \centering
@@
   \end{tabular}
-  \caption{Here we see a table of the minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
+  \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
 Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
@@
 The monotone quintic spline interpolant provides a distinct increase in accuracy over the monotone cubic variant.
-Although the complexity of the algorithm is increased, the number of points required to achieve the same level of accuracy has decreased.
+The computational complexity of this algorithm is identical to the existing cubic method and the cost of evaluating the spline after construction is unchanged.
+While the number of computations required per interval has increased, the number of points required to achieve the same level of accuracy has decreased.
 In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.
+The checks for monotonicity across all intervals are independent and the monotonicity adjustments could be computed independently and intelligently merged with minor changes to the serial algorithm.
 
-Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more complex than the cubic case, gains in accuracy or a decreased necessary number of points are often worth the computational effort.
+Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more expensive than the cubic case, gains in accuracy or decreased necessary number of points are often worth the computational effort.
+In practice the costs of spline interpolation are often dominated by evaluation rather than construction, and the increased accuracy is afforded for a negligible increase in computational cost.
 
@@
 Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory.
-
 There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations.
@@
 
-%% \section*{Acknowledgments}
-%% The NSF is acknowledged for its support of this project.
+\section*{Acknowledgments}
+This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.
 
@@
 science at Virginia Polytechnic Institute and State University
-(VPI\&SU) studying under Dr. Layne Watson. His research interests
-include computational science, approximation theory, optimization, and
-artificial intelligence. His email address is \email{tchlux@vt.edu}.
+(VPI\&SU) studying under Dr. Layne Watson. His email address is
+\email{tchlux@vt.edu}.
 
@@
 
-\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) has interests in
-engineering statistics, statistical modeling, and data analysis.
+\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) is an
+associate professor of Statistics at VPI\&SU.
 

### 2020-03-30 [2020-01]_IEEESE_(convex_optimization)/paper.tex
updated work

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%2345678901234567890123456789012345678901234567890123456789012345678901234567890
-%        1         2         3         4         5         6         7         8
-
-\documentclass[conference]{ieeeconf}  % Comment this line out
+\documentclass[letterpaper, 10 pt, conference]{IEEEtran}  % Comment this line out
                                                           % if you need a4paper
-%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
-\IEEEoverridecommandlockouts                              % This command is only needed 
-%% % copyright info
-%% \IEEEpubid{\makebox[\columnwidth]{978-1-5386-6133-8/18/\$31.00
-%%         ~\copyright~2018 IEEE \hfill}\hspace{\columnsep}
-%%         \makebox[\columnwidth]{ }}
-
-% if you want to use the \thanks command
-\overrideIEEEmargins
-% See the \addtolength command later in the file to balance the column lengths on the last page of the document
 
@@
 \usepackage{balance} % For balancing the last column.
-
-%% \usepackage{graphics} % for pdf, bitmapped graphics files
-%\usepackage{epsfig} % for postscript graphics files
-%\usepackage{mathptmx} % assumes new font selection scheme installed
-%\usepackage{times} % assumes new font selection scheme installed
 \usepackage{amsmath} % assumes amsmath package installed
@@
 
+%% ------------------------------------------
+%% Block to make a ``titleheader''
+\makeatletter
+\newcommand*\titleheader[1]{\gdef\@titleheader{#1}}
+\AtBeginDocument{%
+  \let\st@red@title\@title%
+  \def\@title{%
+    \bgroup\normalfont\large\raggedleft\@titleheader\par\egroup
+    \vskip1.5em\st@red@title}
+}
+\makeatother
+%% ------------------------------------------
+
 \title{\LARGE \bf Analytic Test Functions for Generalizable
@@
 
-%% \author{Thomas C. H. Lux$^{1}$, Tyler H. Chang$^{1}$% <-this % stops a space
-%% \thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu, thchang at vt.edu)}}%
-%% }
-\author{Anonymous Authors}
+\titleheader{IEEE SoutheastCon 2020}
+
+\author{Thomas C. H. Lux$^{1}$, Tyler H. Chang$^{1}$% <-this % stops a space
+\thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu, thchang at vt.edu)}}%
+}
 
@@
 
+%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
+\IEEEoverridecommandlockouts                              % This command is only needed 
+%% % copyright info
+\IEEEpubid{\makebox[\columnwidth]{978-1-7281-6861-6/20/\$31.00~\copyright{}2020 IEEE \hfill} \hspace{\columnsep}\makebox[\columnwidth]{ }}
+
 \maketitle
-\thispagestyle{empty}
-\pagestyle{empty}
 

### 2020-03-30 [2019-02]_NUMA_(interpolant_error_bound)/Data/example_delaunay.py

@@ -1,0 +1,35 @@
+from util.plot import Plot, multiplot
+
+p = Plot()
+p.add_node("top center", 2, -.45, size=335,
+           marker_line_width=1,
+           marker_line_color="rgba(0,0,0,.2)",
+           color='rgba(0,0,0,.05)')
+p.add_node("a", 0, 1, size=10, color=p.color(0))
+p.add_node("b", 2, 0, size=10, color=p.color(1))
+p.add_node("c", 2, 2, size=10, color=p.color(2))
+p.add_node("d", 4, 1, size=10, color=p.color(3))
+p.add_node("", 2, 3, display=False)
+p.add_node("", 2, -1, display=False)
+p.add_edge(['a','b','d','a'])
+p.add_edge(['a','c','d','a'])
+p1 = p.graph(width=400, height=400, file_name="example_not_delaunay.html")
+
+
+p = Plot()
+p.add_node("left center", 1.25, 1, size=170,
+           marker_line_width=1,
+           marker_line_color="rgba(0,0,0,.2)",
+           color='rgba(0,0,0,.05)')
+p.add_node("a", 0, 1, size=10, color=p.color(0))
+p.add_node("b", 2, 0, size=10, color=p.color(1))
+p.add_node("c", 2, 2, size=10, color=p.color(2))
+p.add_node("d", 4, 1, size=10, color=p.color(3))
+p.add_node("", 2, 3, display=False)
+p.add_node("", 2, -1, display=False)
+p.add_edge(['a','b','c','a'])
+p.add_edge(['b','c','d','b'])
+p2 = p.graph(width=400, height=400, file_name="example_delaunay.html")
+# p2 = p.graph(html=False)
+
+# multiplot([p1,p2], file_name="example_delaunay.html")

### 2020-03-30 [2019-02]_NUMA_(interpolant_error_bound)/Data/example_shepard.py

@@ -1,0 +1,84 @@
+from util.plot import Plot, multiplot
+
+# Return a linearly separated set of floats between "start" and "end"
+def linspace(start, end, steps):
+    return [start + (end-start)*i/(steps-1) for i in range(steps)]
+
+# Transpose a python list of lists.
+def transpose(l): return [_ for _ in zip(*l)]
+
+# Given a center and a radius, produce a set of points equally spaced
+# along the radius of the circle.
+def circle_points(center, radius, steps=200):
+    from math import sin, cos, pi
+    points = []
+    for i, step in enumerate(linspace(0, 2*pi, steps)):
+        points.append( [center[0] + cos(step)*radius, center[1] + sin(step)*radius] )
+    return points
+
+
+# Generate an interesting random set of points
+points = [
+    [0.48, 0.52],
+    [0.40, 0.24],
+    [0.65, 0.93],
+    [0.16, 0.56],
+    [0.93, 0.68],
+    [0.70, 0.16],
+]
+
+# Get the radius about each point.
+from util.approximate import ShepMod
+from numpy import array
+m = ShepMod()
+m.fit(array(points, dtype=float))
+# Now "m.rw" holds the radius about each point.
+
+
+p = Plot()
+# Add points.
+for i,(pt,r) in enumerate(zip(points, m.rw)):
+    circle_pts = circle_points(pt, r)
+    if i == 0:
+        p.add("circle", *transpose(circle_pts+[pt]),
+              mode="lines", dash=None, color=(0,0,0,1.0))
+    elif i == 1:
+        p.add("circle", *transpose([pt]+circle_pts[125:]+circle_pts[:125]),
+              mode="lines", dash="dot", color=p.color(i,alpha=.3))
+    elif i == 2:
+        p.add("circle", *transpose([pt]+circle_pts[25:]+circle_pts[:25]),
+              mode="lines", dash="dash", color=p.color(i,alpha=.3))
+    # Add the node second so that it is visually on top.
+    p.add_node(f"point {i+1}", pt[0], pt[1], size=10, color=i)
+
+height = 500
+width = 500
+y_bounds = [-.165, 1.55]
+x_center = 0.5
+x_bounds = [x_center - (y_bounds[1]-y_bounds[0]) * (width/height) / 2,
+            x_center + (y_bounds[1]-y_bounds[0]) * (width/height) / 2]
[…]

### 2020-04-01 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-response-2020-04.tex
Finished first draft of response to reviewers for the NA journal paper.

%% ========================================================================================================================================================================================================

\setcounter{page}{0}\thispagestyle{empty}

{\huge \noindent \bf Response to Reviewers}
\vspace{.25cm}

\noindent \today
\vspace{1cm}

\noindent Dear reviewers,
\vspace{.25cm}

\noindent We have carefully considered all of your comments and feedback, and are grateful for your time and effort. The following revisions have been made to account for the feedback given:
\vspace{.25cm}

\begin{enumerate}[leftmargin=1cm]
  \setlength{\itemsep}{.4cm}
\item A description and set of references to sparse grid methodologies has been placed in the introduction. (Page \hyperlink{page.2}{2}) \label{review:item1}
\item The descriptions of approximation algorithms have been modified for greater clarity, and three new visuals depicting relevant Delaunay, Voronoi, and Modified Shepard properties have been added. (Pages \hyperlink{page.4}{4}--\hyperlink{page.8}{8}) \label{review:item2}
\item Slight textual modifications were made to Lemma \ref{lemma:3} for clarity, and a new visual was constructed to illustrate the logic of the accompanying proof. (Pages \hyperlink{page.11}{11}--\hyperlink{page.12}{12}) \label{review:item3}
\item A new visual was added to the analytic test subsection to demonstrate the link between the theory and practice, accompanying text was added as well. (Pages \hyperlink{page.15}{15}--\hyperlink{page.16}{16}) \label{review:item4}
\item Minor rephrasing and additional references were added throughout the text to address specific concerns and/or misunderstandings. (Pages \hyperlink{page.3}{3}, \hyperlink{page.9}{9}--\hyperlink{page.10}{10}, \hyperlink{page.13}{13}--\hyperlink{page.14}{14}, \hyperlink{page.17}{17}, and \hyperlink{page.28}{28}) \label{review:item5}
\end{enumerate}
\vspace{.7cm}

\noindent For the reviewers' convenience, all newly introduced modifications in this revision are marked with \changed{red text}{} and have a red bar in the margin accompanying them. In addition, all colored links in this response point directly to the associated positions in the revised paper. Responses to individual requests / comments follow.
\vspace{.25cm}

\begin{itemize}[leftmargin=.5cm]
  \setlength{\itemsep}{.4cm}

\item[*] {\it It is not clear why only Delaunay and the MLP are compared in the analytic test.} \vspace{.2cm}

The reason for only studying these two methods is twofold. First and foremost, only Delaunay and MLP are strictly piecewise linear approximations, which means the theorem directly applies to these techniques. Secondly, there is a lot of potential for clutter / overabundance of information in the visuals that follow and little for a reader to gain from seeing all techniques compared on this test. The text was modified to clarify this point at the end of the first paragraph of Section \ref{sec:analytic}.

\newpage\setcounter{page}{0}\thispagestyle{empty}
\item[*] {\it Why does the error of Delaunay not decrease with more points in Figure \ref{fig:convergence-20d}?}\vspace{.2cm}

This question has multiple answers, but in short there are not enough data points to see a meaningful decrease in error. It is our hope that item \ref{review:item4} answers this question thoroughly in the paper.

\item[*] {\it The connection between the analytic test and the theorem is unclear. A simulation in which the error is studied with respect to $d(z,x_0)$ for the Delaunay interpolant should be included.}\vspace{.2cm}

Addressed by item \ref{review:item4}.

\item[*] {\it It is not clear why the performed simulation is in Section 5.1, and not in a new section.}\vspace{.2cm}

As more clearly shown by item \ref{review:item4}, the analytic test specifically demonstrates the validity of the theorem. This is why the analytic test is considered part of Section \ref{sec:theory}.

\item[*] {\it Report the mean computational time in the main text and leave only the details at the appendix.}\vspace{.2cm}

The mean computation time is included in the main text in Table \ref{table:avg-performance}.

\item[*] {\it The statement, "In summary, the regime of interpolation is much greater \ldots" is way too ambitious.}\vspace{.2cm}

The phrasing has been reworked (page \hyperlink{page.16}{16}) to emphasize the underlying fact: the transition point at which regression algorithms become better than interpolation algorithms is (at least partially) a function of the signal-to-noise ratio (SNR) and the data sparsity. When data is more sparse, the range of SNRs for which interpolants are useful grows.

\item[*] {\it Add a mention of piecewise linear and Lagrangian interpolation to the list of ``well understood'' interpolation techniques.}\vspace{.2cm}

Done on page \hyperlink{page.2}{2}.

\item[*] {\it A high dimension interpolation survey is not complete without mention of sparse grids.}\vspace{.2cm}

Addressed by item \ref{review:item1}.

\newpage\setcounter{page}{0}\thispagestyle{empty}
\item[*] {\it ``\ldots generic uniform bounds are largely unobtainable for regression techniques on arbitrary approximation problems.'' Not true, bounds for regression techniques can be found using projection onto orthogonal polynomials.}\vspace{.2cm}

We have reworded the statement and referenced the sparse grid error bounds in the same paragraph (page \hyperlink{page.2}{2}).

\item[*] {\it ``\ldots interactions are either preconceived or model pairwise interactions between dimensions at most.'' This statement is not true for sparse grids.}\vspace{.2cm}

The wording has been clarified to say that this statement applies to problems with more than tens of dimensions (page \hyperlink{page.3}{3}).

\item[*] {\it What is the range of $j$ for MARS? Mention that the choice of the MLP network architecture is crucial. Add a picture of a mesh that satisfies the Delaunay criterion and one that doesn’t. Comment on the singularity at $W^{(k)}(x^{(k)})$ for modified Shepard and plot the shape of $W^{(k)}(x)$. Please remove the sparse grid reference from Box Splines. Add a plot showing the Voronoi tessellation and $v^{x^{(i)}}(y)$.}\vspace{.2cm}

All of these have been addressed in each individual subsection and in part by item \ref{review:item2} above. Regarding the shape of $W_k$ in Modified Shepard and LSHEP, the weight equation was rewritten to provoke a more clear understanding in accompaniment with the visual.

\item[*] {\it How would you measure the error if the range of approximation is the space of continuous functions?}\vspace{.2cm}

This is answered in the last paragraph of Section \ref{sec:discussion}, the Discussion.

\item[*] {\it Rewrite as ``Instead, for a CDF \ldots'' on page \hyperlink{page.9}{9}. Move the sentence in Figure \ref{fig:prediction-example} regarding the KS test to the main text. Add ``empirical'' before ``CDF'' where appropriate.}\vspace{.2cm}

All addressed on page \hyperlink{page.9}{9}.

\item[*] {\it I think that the Appendix provided on the meaning of confidence intervals and null hypothesis is too short for someone without a background in statistics.}\vspace{.2cm}

A reference to an introductory statistics textbook has been provided for readers that need a more thorough introduction to statistics (page \hyperlink{page.28}{28}).

\item[*] {\it I don’t understand the expression “round-trip prediction methodology”.}\vspace{.2cm}

The phrasing has been changed (page \hyperlink{page.10}{10}). This is a reference the process of converting source data into predictions and then measuring the error.

\newpage\setcounter{page}{0}\thispagestyle{empty}
\item[*] {\it Add a picture of $g(t)$, $g'(t)$, and the line $w$ to Lemma \ref{lemma:3}. Give details on these equalities}
  $$ \frac{g'(0) \tilde t}{2} = \int_0^{\tilde t} \bigl( g'(0) - \gamma_g t \bigr) dt$$
  $$ \frac{-g'(0)^2}{2 \gamma_g} = g'(0) - \frac{-g'(0)^2}{2 \gamma_g} - \frac{\gamma_g}{2}$$
\vspace{.2cm}

The first equation is the definition of an integral recalling $\tilde t = g'(0) / \gamma_g$. The second equation is misunderstood. In Lemma \ref{lemma:3} it is correctly written as
  $$ \frac{-g'(0)^2}{2 \gamma_g} \color{red}<\color{black} g'(0) - \frac{\color{red}+\color{black} g'(0)^2}{2 \gamma_g} - \frac{\gamma_g}{2},$$ \newline\noindent which follows directly from the statement that $g'(0) > \gamma_g / 2$. Some text has been added to the lemma to attempt to make this more clear. The requested picture has been added as per item \ref{review:item3}.

\item[*] {\it You should also try the analytic test with other space-filling points.}\vspace{.2cm}

The tests were additionally run with a Latin hypercube design, however nearly identical results were achieved (mentioned on page \hyperlink{page.14}{14}). The purpose of this test is not to provide a review of space filling / well-conditioned designs, but rather to pick a reasonable design and observe the connections between the observed error and the theorem. Fekete points were specifically requested as such a reasonable design by reviewers in the first revision of this paper.

\item[*] {\it Add in Figure \ref{fig:convergence-2d} the reference slopes that show linear and quadratic decay of the error with respect to the number of points.}\vspace{.2cm}

Notice that the number of points is increasing in this visual. There is no obvious ``reference'' for linear and quadratic decay here. This is only possible when something approaches 0 in log-log plots. Instead, readers are given the relative decay in longest edge length to compare with decrease in error per item \ref{review:item4}. In that figure, it is acknowledged that the approximation error is decreasing at a faster rate than the longest edge length.

\item[*] {\it Plot the Fekete points used in the case $d = 2$ (e.g. the case with 128 or 512 points).}\vspace{.2cm}

While potentially interesting for those unaware of Fekete points, this visual is easy to find elsewhere and would distract readers from the purpose of this subsection in the paper. The specific sample design methodology is not as important as decreasing the longest edge length of all simplices at roughly the same rate as the decrease in $\sigma_d$ as seen in Figure \ref{fig:data-spacing}.

\newpage\setcounter{page}{0}\thispagestyle{empty}
\item[*] {\it On the test where $d = 20$: could it be that you just need more points before you can see convergence?}\vspace{.2cm}

Yes certainly more points would help, but in order to achieve meaningful density in 20 dimensions at least millions of points would be needed. Consider that it takes roughly a million points to cover all corners of a box in $d = 20$, and that's with no points on the interior! The intent behind this example is to show that sparsity is very common in tens of dimensions and up.

\item[*] {\it Are the Fekete points in $d = 20$ just bad and you should use another design?}\vspace{.2cm}

In general it is very difficult to construct well-spaced point sets in high dimension. This is a known problem. The Fekete points are a reasonable choice and are expected to perform similarly to other well-spaced point sets.

\item[*] {\it On the test with d = 2: could it be that the MLP stagnation is due to not having done enough epochs of training?}\vspace{.2cm}

No. To check for this the amount of training was extended by five times. No meaningful improvement in performance was achieved.

\item[*] {\it What do you need the polynomials of order $n+1$ for (the Fekete points are obtained by the polynomials of order n)?}\vspace{.2cm}

This is a misunderstanding. The paper reads ``\ldots polynomials of \color{red}degree \color{black} $n+1$.'' and the degree of a polynomial is one less than the order. So order $n$ polynomials are used here.

\item[*] {\it Don’t you perform both interpolation and regression on both the data sets with and without noise?}\vspace{.2cm}

Yes, the phrasing in the parenthetical statements was unclear. It is been reworded to emphasize that having no noise poses an interpolation problem and having noise poses a regression problem (bottom of page \hyperlink{page.14}{14} and top of page \hyperlink{page.15}{15}).

\item[*] {\it The statement, ``\ldots exponentially increasing the number of data points should result in a decreasing error.'' is imprecise: increasing even exponentially won’t help if you place the points badly.}\vspace{.2cm}

The original statement was made in the context of using Fekete points. In general yes, the points must be well spaced for convergence to be achieved. The wording has been adjusted to make this more clear (page \hyperlink{page.15}{15}).

\newpage\setcounter{page}{0}\thispagestyle{empty}
\item[*] {\it Test on forest fire: you should mention that timings and lowest absolute prediction error can be checked in the Appendix (same for the other tests).}\vspace{.2cm}

The wording in the paragraph that introduces the empirical tests has been modified to make this more clear (page \hyperlink{page.17}{17}).

\item[*] {\it Add at least for the forest fire test a scatter plot of predicted values versus correct values, at least for the two/three best methods.}\vspace{.2cm}

This suggestion is appreciated. Such specific outcomes are not obviously related to the error bound theorem, which is the central component of this paper. For this reason, we would like to abstain from including information that might distract readers from the main purposes of the text.

%% \newpage\setcounter{page}{0}\thispagestyle{empty}

\end{itemize}

\newpage
%% ========================================================================================================================================================================================================

### 2020-04-01 [2019-02]_NUMA_(interpolant_error_bound)/Data/example_cdf_prediction/fits.py

@@ -1,0 +1,78 @@
+# Given a list of numbers, generate two lists, one of the CDF x points
+# and one of the CDF y points (for fitting).
+def cdf_points(data):
+    import numpy as np
+    SMALL = 2**(-26)
+    # Sort the data (without changing it) and get the min and max
+    data = np.array(sorted(data))
+    min_pt = data[0]
+    max_pt = data[-1]
+    # Initialize a holder for all CDF points
+    data_vals = []
+    # Add all of the CDF points for the data set
+    for i,val in enumerate(data):
+        if ((i > 0) and (val == data[i-1])):
+            data_vals[-1] = (i+1)/len(data)
+        else:                                data_vals.append( (i+1)/len(data) )
+    # Add the 100 percentile point if it was not added already
+    if (data_vals[-1] != 1.0): data_vals[-1] = 1.0
+    # Convert data into its unique values.
+    data = np.array(sorted(set(data)))
+    assert(len(data) == len(data_vals))
+    # Convert it all to numpy format for ease of processing
+    data_vals = np.array(data_vals)
+    # Return the two lists that define the CDF points.
+    from fraction import Fraction
+    data = list(map(Fraction, data))
+    data_vals = list(map(Fraction, data_vals))
+    return data, data_vals
+
+
+# Return a discontinuous fit.
+def flat_fit(data):
+    from numpy import searchsorted
+    x, y = cdf_points(data)
+    def edf(z):
+        try:    return [value(_) for _ in z]
+        except: return y[min(len(y)-1, max(0,searchsorted(x, z, side="left")-1)
+                             + (1 if z == max(x) else 0))]
+    return edf
+
+# Return a linear fit function.
+def linear_fit(data):
+    from polynomial import fit
+    x, y = cdf_points(data)    
+    print("x: ",len(x))
+    print("y: ",len(y))
+    cdf = fit(x, y, continuity=0)
+    def fit(z): return max(0,min(1,cdf(z)))
+    fit.derivative = cdf.derivative
+    return fit
+
+# Return a cubic fit function.
+def cubic_fit(data=None, x=None, y=None):
+    from monotone import monotone_cubic_spline
+    if (x is None) or (y is None): x, y = cdf_points(data)
+    cdf = monotone_cubic_spline(x, y)
+    def fit(z):
+        val = max(0,min(1,cdf(z)))
+        if (z < x[0]): val = min(val, y[0])
[…]

### 2020-04-01 [2019-02]_NUMA_(interpolant_error_bound)/paper.tex
Finished first draft of response to reviewers for the NA journal paper.

@@
 \usepackage{float}    % For controlling the placement of tables.
+\usepackage{enumitem} % For changing item indents in enumerate.
 
@@
 \cbcolor{red}
-\newcommand{\changed}[2]{\cbstart{\color{red}#1}{\color{gray}\sout{#2}}\cbend{}}
+%% \newcommand{\changed}[2]{\cbstart{\color{red}#1}{\color{gray}\sout{#2}}\cbend{}}
+\newcommand{\changed}[2]{\cbstart{\color{red}#1}{}\cbend{}}
 %% -------------------------------------------------------------------
@@
 
+%% Include the response to reviewers at the beginning of the document.
+\include{tchlux-NA-response-2020-04}
+
+%% Set the title.
 \title{Interpolation of Sparse High-Dimensional Data\thanks
@@
 \date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
+%% The correct dates will be entered by the editor
+
+{\let\newpage\relax\maketitle}
+
+%% \maketitle
+
+\setcounter{page}{1}
 
@@
   and more generally spline interpolation via B-splines
-  \cite{de1978practical}.}{Splines in particular are well understood
+  \cite{de1978practical}}{Splines in particular are well understood
   as an interpolation technique in one dimension
@@
 The theoretical foundation of low dimensional interpolation allows the
-construction of strong error bounds that are absent from high
-dimensional problems. This work extends some known results regarding
-the secant method \cite{dennis1996numerical} to construct an
-interpolation error bound for problems of arbitrary dimension. These
-error bounds are useful, considering the same cannot be said for
-regression algorithms in general. The maximum complexity of an
-interpolant is bounded by the amount of data available, while the
-maximum complexity of a regressor is bounded by both the amount of
-data and the chosen parametric form. For this reason, generic uniform
-bounds are largely unobtainable for regression techniques on arbitrary
-approximation problems, even when the approximation domain is bounded.
+construction of strong error bounds that are \changed{lacking
+  in}{absent from} high dimensional problems. \changed{A large body of
+  literature exists studying the construction and application of
+  sparse grids \cite{bungartz2004sparse,chkifa2014high}, which attempt
+  to address the problem of dimensional scaling. The sparse grid point
+  sets combined with specific approximation functions are used to
+  construct total error bounds by considering projections onto
+  orthogonal polynomials
+  \cite{migliorati2013approximation,nobile2016convergence}. However,
+  sparse grids have exponential size in dimension making them
+  intractable for more than tens of dimensions and do not immediately
+  apply to the problem of scattered data interpolation.}{} This work
+extends some known results regarding the secant method
+\cite{dennis1996numerical} to construct \changed{a uniform error bound
+  for interpolants over arbitrary scattered data in any dimension}{an
+  interpolation error bound for problems of arbitrary
+  dimension}. These error bounds are useful, considering
+\changed{similar uniform bounds}{the same cannot} cannot be
+\changed{constructed}{said} for regression algorithms in general. The
+maximum complexity of an interpolant is bounded by the amount of data
+available, while the maximum complexity of a regressor is bounded by
+both the amount of data and the chosen parametric form. For this
+reason, generic uniform bounds are largely unobtainable for regression
+techniques on arbitrary approximation problems, even when the
+approximation domain is bounded.
 
@@
 exponentially. Quantifying all possible interactions becomes
-intractable, and hence beyond three-dimensional data mostly linear
-models are used. That is not to say nonlinear models are absent, but
-nonlinearities are often either preconceived or model pairwise
-interactions between dimensions at most. Even globally nonlinear
+intractable, and hence \changed{problems}{} beyond \changed{ten to
+  twenty dimensions usually rely on linear models}{three-dimensional
+  data mostly linear models are used}. That is not to say nonlinear
+models are absent, but nonlinearities are often either preconceived or
+model \changed{interactions between small subsets of}{pairwise
+  interactions between} dimensions at most. Even globally nonlinear
 approximations such as neural networks are constructed from
@@
      {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k - \bigl\|x -
-x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x - x^{(k)}\bigr\|_2 \bigr)\bigr)^2$}{}
-and $r_k \in \mathbb{R}$ is the smallest radius such that at least
-$d+1$ other points $x^{(j)}$, $j \not = k$, are inside the closed
-Euclidean ball of radius $r_k$ about $x^{(k)}$. The computational
-complexity of ShepMod is $\mathcal{O}(n^2d)$. This paper uses a
-Fortran 95 implementation of ShepMod derived from SHEPPACK
+where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k -
+  \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x -
+  x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x \neq x^{(k)}$,}{} and $r_k
+\in \mathbb{R}$ is the smallest radius such that at least $d+1$ other
+points $x^{(j)}$, $j \not = k$, are inside the closed Euclidean ball
+of radius $r_k$ about $x^{(k)}$. The computational complexity of
+ShepMod is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95
+implementation of ShepMod derived from SHEPPACK
 \cite{thacker2010algorithm}.
@@
   \caption{\color{red} On the left above is a depiction of the radius
-    of influence for three chosen points of a collection using the
-    modified Shepard criteria. On the right is the shape of the weight
-    function $W_k$ for all data points, where $W_k(x) = 0$ when
-    $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen that in the limit
-    as $W_k$ goes to infinity at $x_k$, the relative weight of $x_k$
-    goes to one.
+    of influence for three chosen points of a collection in two
+    dimensions using the modified Shepard criteria. On the right is
+    the shape of the weight function $W_k$ for all data points, where
+    $W_k(x) = 0$ when $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen
+    that in the limit as $W_k$ tends towards infinity at $x_k$, the
+    relative weight of $x_k$ tends to one.
   \vspace{-.1cm}}
@@
  {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is \changed{the same as for the modified Shepard method}{a locally supported weighting function} and $P_k(x)$
-is a local linear approximation to the data satisfying
+where $W_k(x)$ is \changed{the same as for the modified Shepard
+  method}{a locally supported weighting function} and $P_k(x)$ is a
+local linear approximation to the data satisfying
 $P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
@@
     boundaries (dashed lines) about a set of interpolation points
-    (dots). In this example, the Voronoi model basis function about
-    the center most point has nonzero weight in the shaded region and
-    transitions from a value of one at the point to zero at the
-    boundary of the twice expanded Voronoi cell (solid line).
+    (dots) in two dimensions. In this example, the Voronoi model basis
+    function about the center most point has nonzero weight in the
+    shaded region and transitions from a value of one at the point to
+    zero at the boundary of the twice expanded Voronoi cell (solid
+    line).
   \vspace{-.1cm}}
@@
 
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
-  \end{tikzpicture}
-  \vspace{-0.3cm}
+\changed{
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\textwidth,]{example_distribution_prediction.pdf}
   \caption{In this HPC I/O example, the general methodology for
-    predicting a CDF and evaluating error can be seen. The Delaunay
-    method chose three source distributions (dotted lines) and
-    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
-    weighted sum of the three known CDFs produces the predicted CDF
-    (dashed line). The KS Statistic (arrow) computed between the true
-    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
-    example. For this example the KS test null hypothesis is rejected
-    at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
+    predicting a CDF and evaluating error can be seen\color{red},
+    where M means $\times 10^6$\color{black}. The Delaunay method
+    chose three source distributions (dotted lines) and assigned
+    weights \color{red}\{.1, .3, .6\} \color{black} (top to bottom at
+    \color{red}middle\color{black}). The weighted sum of the three
+    known CDFs produces the predicted CDF (dashed line). The KS
+    Statistic (\color{red}vertical line\color{black}) computed between
+    the true CDF (solid line) and predicted CDF (dashed line) is 0.2
+    for this example. For this example the KS test null hypothesis is
+    rejected at $p$-value 0.01, however it is not rejected at
+    $p$-value 0.001.
   \vspace{-.1cm}}
@@
 \end{figure}
+}{}
 
@@
 The two-sample KS test is a useful nonparametric test for comparing
-two CDFs while only assuming stationarity, finite mean, and finite
-variance. The null hypothesis (that two CDFs come from the same
-underlying distribution) is rejected at level $p \in [0,1]$ when
- $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
-applications of the KS test presented in this work $n_1 = n_2$. An
-example of the \changed{end-to-end}{round-trip} prediction methodology from known and
-predicted distributions to the calculation of error can be seen in
+two \changed{empirical}{} CDFs while only assuming stationarity,
+finite mean, and finite variance. The null hypothesis (that two
+\changed{empirical}{} CDFs come from the same underlying distribution)
+is rejected at level $p \in [0,1]$ when
+ $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)}
+\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$ with distribution sample
+sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test
+presented in this work $n_1 = n_2$. An example of \changed{the process
+  of generating predicted distributions from known distributions and
+  the subsequent}{the round-trip prediction methodology from known and
+  predicted distributions to the} calculation of error can be seen in
 Figure \ref{fig:prediction-example}. A brief listing of relevant
-statistical terms used throughout this work is provided in
-the Appendix (Section \ref{sec:appendix}).
+statistical terms used throughout this work is provided in the
+Appendix (Section \ref{sec:appendix}).
 
@@
   \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
-  = \gamma \|x_j-x_0\|_2^2$.
+  = \gamma \|x_j-x_0\|_2^2$. \changed{The following is visualized in 
+  Figure \ref{fig:lemma3}.}{}
 
@@
   since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
-  and so $g'(0) - \g z \le g'(z) = 0$. Now
+  and so $g'(0) - \g z \le g'(z) = 0$. Now \changed{by integrating}{}
   $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
   (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
+  \changed{acknowledging $g'(0) > \g / 2$}{}
   \begin{align*}
@@
 \end{proofdot}
+
+\changed{
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\textwidth,]{example_lemma_3.pdf}
+  \def\g{\gamma_g}
+  \caption{\color{red} Three different scenarios visualizing {\it
+      Lemma \ref{lemma:3}}, where $g(t)$ is the difference between a
+    piecewise linear interpolant and the approximated function along a
+    normalized line segment between interpolation points, $g'(t)$ is
+    $\g$-Lipschitz continuous, and $w$ and $\tilde t$ are defined in
+    the proof. Leftmost is a randomly chosen permissible shape of $g$
+    and $g'$. The middle is the only possible shape of $g$ and $g'$
+    given $g'(0) = \g / 2$, establishing the case of equality in the
+    lemma. Rightmost is the resulting contradiction when $g'(0) > \g /
+    2$, notice it is impossible to ensure $g'(t)$ is $\g$-Lipschitz
+    continuous and satisfy $g(1) = 0$ (highlighted with red circle on
+    the right).
+  \vspace{-.1cm}}
+  \label{fig:lemma3}
+\end{figure}
+}{}
 
@@
 data only when the diameter of the simplex is also reduced at a
-proportional rate.  Only linear convergence to the true function can
-be achieved in practice, without the incorporation of additional
-observations. Notice that the approximation error is largely
-determined by the spacing of observed data. Predictions made by
-simplices whose vertices are not well-spaced (i.e., have large
-diameter, or are nearly contained in a hyperplane) have higher
-error.
+proportional rate. Only linear convergence to the true function
+\changed{is}{can be} achieved in practice\changed{}{,} without the
+incorporation of additional observations\changed{, by moving closer to
+  interpolation points}{}. Notice that the approximation error is
+largely determined by the spacing of observed data. Predictions made
+by simplices whose vertices are not well-spaced (i.e., have large
+diameter, or are nearly contained in a hyperplane) have higher error.
 
@@
 $6,$ chosen points are kept well spaced by computing approximate
-Fekete points. Fekete points have a history in potential theory
-\cite{kovari_pommerenke_1968} and are most generally defined as those
-points that maximize the absolute value of a Vandermonde
-determinant. Here, the QR method outlined in \cite{bos2010computing}
-is used to identify approximate Fekete points from a Vandermonde
-matrix. A multivariate polynomial basis of size $2N$ is constructed
-containing all polynomials of degree $\leq n$ for the largest $n$ such
-that ${d+n \choose n} \leq 2N$ and $2N - {d+n \choose n}$
-arbitrarily selected polynomials of degree $n + 1.$ The Vandermonde
-matrix for these $2N$ polynomials is used to select $N$ approximate
-Fekete points.
+Fekete points. \changed{The following experiment was also run with a
+  Latin hypercube design \cite{park1994optimal}, which produced almost
+  identical results that are not reported here.}{} Fekete points have
+a history in potential theory \cite{kovari_pommerenke_1968} and are
+most generally defined as those points that maximize the absolute
+value of a Vandermonde determinant. Here, the QR method outlined in
+\cite{bos2010computing} is used to identify approximate Fekete points
+from a Vandermonde matrix. A multivariate polynomial basis of size
+$2N$ is constructed containing all polynomials of degree $\leq n$ for
+the largest $n$ such that ${d+n \choose n} \leq 2N$ and $2N - {d+n
+  \choose n}$ arbitrarily selected polynomials of degree $n + 1.$ The
+Vandermonde matrix for these $2N$ polynomials is used to select $N$
+approximate Fekete points.
 
@@
 each test two experiments are executed, one with exact function
-evaluations (an interpolation problem) and one with a constant
-signal-to-noise (SNR) ratio of 10:1 (a regression problem).
+evaluations (\changed{an interpolation problem}{}) and one with a
+constant signal-to-noise ratio (SNR) of 10:1 (\changed{a regression
+  problem}{}).
 
@@
 
-The bound from the theorem suggests that exponentially increasing the
-number of \changed{well-spaced}{} data points should result in a
-decreasing error. The $d = 2$ test seen on the left half of Figure
+The bound from the theorem suggests that \changed{by}{exponentially}
+increasing the number of \changed{well-spaced}{} data points,
+\changed{approximation error can be reliably decreased}{results in a
+  decreasing error}. The $d = 2$ test seen on the left half of Figure
 \ref{fig:convergence-2d} shows a consistent decrease in error for
@@
     \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Euclidean Distance};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Error, Distance, and Smallest SV };
   \end{tikzpicture}
   \caption{\color{red}The distribution of absolute error, distance to
-    the nearest data point, smallest singular value and the longest
-    edge of the simplex containing each approximation point in the
-    tests from Fig. \ref{fig:convergence-2d} and
+    the nearest data point, smallest singular value (SV) and the
+    longest edge of the simplex containing each approximation point in
+    the tests from Fig. \ref{fig:convergence-2d} and
     Fig. \ref{fig:convergence-20d} for Delaunay. In two dimensions it
@@
     in higher dimension the data remains sparse even with thousands of
-    data points. This is seen by the relatively small reduction in
-    $k$, and this explains the slow reduction in error seen by
-    Delaunay in Fig. \ref{fig:convergence-20d}.}
+    data points, and the decay in data spacing is more prominant. The
+    relatively small reduction in $k$ along with the decrease in
+    $\sigma_d$ explain the minimal reduction in error seen by Delaunay
+    in Fig. \ref{fig:convergence-20d}.}
   \label{fig:data-spacing}
@@
 
-\changed{The results achieved for both the $d = 2$ and $d = 20$ align
-  with the theoretical error bound as can be seen in Figure
+\changed{The results achieved for both $d = 2$ and $d = 20$ align with
+  the theoretical error bound as can be seen in Figure
   \ref{fig:data-spacing}. In low dimension, thousands of data points
@@
   However, in higher dimension it takes many more points to achieve a
-  reduction in data sparsity. The same result holds for a Latin
-  hypercube well-spaced point set, suggesting that sparsity is an
-  inherent problem in high dimension approximation problems. This
-  result is not a product of the chosen test function, but rather the
-  exponential increase in complexity that accompanies increased
-  dimension.}{}
-
-In summary, the regime of interpolation is \changed{}{much} greater
-for moderate to high dimensional problems \changed{due to increased
-  sparsity}{}, even for functions with random noise incorporated into
-their evaluation. Acknowledging the viability of interpolation for
-problems of moderate dimension, the next section will consider
-real-world problems of similar proportion (thousands of examples in
-tens of dimensions).
+  reduction in data sparsity while simultaneously being more difficult
+  to produce well-\color{red}conditioned simplices. This evidences the inherent
+  challenge of data sparsity in high dimension approximation
+  problems. The analytic results presented are not caused by the
+  chosen test function, but rather the exponential increase in
+  complexity that accompanies increased dimension.}{}
+
+In summary, the regime of \changed{signal-to-noise ratios that 
+  result in competitively accurate interpolants}{} is \changed{}{much}
+greater for moderate to high dimensional problems \changed{due to
+  increased sparsity}{, even for functions with random noise
+  incorporated into their evaluation}. Acknowledging the viability of
+interpolation for problems of moderate dimension, the next section
+will consider real-world problems of similar proportion (thousands of
+examples in tens of dimensions).
 
@@
 In addition to the figures displaying approximation results for each
-data set, tables of accompanying numerical results are located in the
-Appendix (Section \ref{sec:appendix}). All of the test data sets
-capture underlying functions that are almost certainly stochastic. As
+data set, tables of accompanying numerical results \changed{including
+  timings and exact prediction errors}{} are located in the Appendix
+(Section \ref{sec:appendix}). All of the test data sets capture
+underlying functions that are almost certainly stochastic. As
 described in Section \ref{sec:introduction}, regression techniques
@@
   \label{fig:hist-forest-fire}
-\end{figure}
-
-\begin{figure}
-  \centering
+
+  \vspace{.3cm}
+
   \begin{tikzpicture}
@@
   \label{fig:hist-parkinsons}
-\end{figure}
-
-\begin{figure}
-  \centering
+
+  \vspace{.3cm}
+
   \begin{tikzpicture}
@@
   \label{fig:hist-weather}
-\end{figure}
-
-\begin{figure}
-  \centering
+
+  \vspace{.3cm}
+
   \begin{tikzpicture}
@@
   \label{fig:hist-credit-card}
-\end{figure}
-
-\begin{figure}
-  \centering
+
+  \vspace{.3cm}
+
   \begin{tikzpicture}
@@
   function} (PDF) $f_X.$ For any possible value $x$ of $X$, the
-\textit{percentile} of $x$ is $100$ $F_X(x)$, the percentage values
+\textit{percentile} of $x$ is $100$ $F_X(x)$, the percentage of values
 drawn from $X$ that would be less than or equal to $x$ as the number

### 2020-04-01 [2019-02]_NUMA_(interpolant_error_bound)/Data/example_cdf_prediction/sample_cdf_prediction.py

@@ -1,0 +1,86 @@
+import numpy as np
+from fits import quintic_fit
+from util.plot import Plot
+
+# Distirbution Prediction Application
+p = Plot("", "I/O Throughput", "CDF Value",
+         font_family="times", font_size=18)
+
+xs = [
+    [0.0, 0.18, 0.53, 0.74, 1.0],
+    [0.0, 0.17, 0.52, 0.75, 1.02],
+    [0.0, 0.46, 0.62, 0.89, .98],
+    [0.0, 0.15, .4, .85, 1.],
+]    
+
+# Modify the x values to look like throughput values.
+scale = 4e6
+shift = 1e6
+rescale = lambda z: z*scale + shift
+
+# Rescale the values.
+for x in xs: x[:] = list(map(rescale, x))
+
+# Generate some reasonable y values.
+ys = [
+    [0.0, 0.49, 0.88, 0.89, 1.0],
+    [0.0, 0.33, 0.66, 0.85, 1.0],
+    [0.0, 0.11, 0.25, 0.34, 1.0],
+    [0.0, 0.16, .4, .7, 1. ],
+]
+# Make fits of the data.
+fs = [quintic_fit(x=x, y=y) for (x,y) in zip(xs, ys)]
+# Pick color numbers for the data.
+cs = [0, 3, 1]
+# Generate a guess function as a convex combination of other functions.
+guess = lambda z, fs=fs: sum(f(z) * w for (f,w) in zip(fs, [.1, .3, .6]))
+
+# # Compute the KS statistic difference between the guess and the truth.
+# vs = np.linspace(0, 1, 1000)
+# diffs = [abs(guess(v) - fs[-1](v)) for v in vs]
+# print("np.max(diffs): ",np.max(diffs))
+# print("np.argmax(diffs): ",np.argmax(diffs))
+# print("vs[np.argmax(diffs)]: ",vs[np.argmax(diffs)])
+# x = vs[np.argmax(diffs)]
+# print("guess(x): ",guess(x))
+# print("fs[-1](x): ",fs[-1](x))
+
+for i,(x,y,f,c) in enumerate(zip(xs,ys,fs,cs)):
+    # Skip the "truth" last one.
+    if i >= 3: continue
+    # p.add("Points", x, y, group=i, color=c)
+    p.add_func("f", f, (min(x), max(x)), color=c, dash="dot", opacity=.5)
+
+
+p.add_func("guess", guess, (min(map(min,xs)), max(map(max,xs))), color=6, dash="dash")
[…]

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-interpolation/Versions/0-paper.tex
Finished NUMA paper and submitted. Might need to do one more round with editors.

--- 2020-03-30 [2019-02]_NUMA_(interpolant_error_bound)/paper.tex
@@
 \usepackage{float}    % For controlling the placement of tables.
-
-%% -------------------------------------------------------------------
-%% Command for changing the color of text within a block.
-\usepackage{xcolor}   % Color package for coloring text.
-\usepackage[normalem]{ulem} % For strikethrough text and the \sout{} command.
-\usepackage[xcolor,rightbars]{changebar}% Package for adding bars into margin next to comments.
-\renewcommand{\changebarsep}{5mm}   % Push the bar farther into the margin.
-\renewcommand{\changebarwidth}{1mm} % Increase the width of the bar.
-\cbcolor{red}
-\newcommand{\changed}[2]{\cbstart{\color{red}#1}{\color{gray}\sout{#2}}\cbend{}}
-%% -------------------------------------------------------------------
 
@@
 Regression and interpolation have a considerable theoretical base in
-one dimension \cite{cheney2009course}. \changed{Common techniques
-  include Lagrangian interpolation, piecewise linear interpolation,
-  and more generally spline interpolation via B-splines
-  \cite{de1978practical}.}{Splines in particular are well understood
-  as an interpolation technique in one dimension
-  \cite{de1978practical}, particularly B-splines}. Tensor products of
+one dimension \cite{cheney2009course}. Splines in particular are well
+understood as an interpolation technique in one dimension
+\cite{de1978practical}, particularly B-splines. Tensor products of
 B-splines \cite{unther1996interpolating} or other basis functions have
@@
 cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
-(KS) statistic \cite{lilliefors1967kolmogorov} as a metric.
+(KS) statistic \cite{lilliefors1967kolmogorov} as metric.
 
@@
 Provided the theoretical and practical motivations for exploring
-interpolants, the main objective of this work is to study the
-empirical performance differences in both accuracy and computational
-expense between a set of interpolation techniques and a set of common
-regression techniques. The main contributions of this work are a sharp
-theoretical error bound for (piecewise) linear interpolation in
-arbitrary dimension and multiple empirical comparisons that illustrate
-potential advantages of interpolation.
+interpolants, this work aims to study the empirical performance
+differences between a set of scalable (moderately) interpolation
+techniques and a set of common regression techniques. These techniques
+are applied to a collection of moderately high dimensional problems
+($5 \le d \le 30$) and the empirical results are discussed.
 
@@
 \ref{sec:theory} contains the theoretical error bounds for
-interpolation as well as an analytic test for demonstration. Section
-\ref{sec:data} presents the data sets for empirical approximation
-analysis and the approximation results for all models on these data
-sets. Section \ref{sec:discussion} analyzes and discusses the results,
-their implications, and their limitations. Finally, Section
-\ref{sec:conclusion} concludes.
+interpolation. Section \ref{sec:data} presents the data sets for
+empirical approximation analysis and the approximation results for all
+models on these data sets. Section \ref{sec:discussion} analyzes and
+discusses the results, their implications, and their
+limitations. Finally, Section \ref{sec:conclusion} concludes.
 
@@
 \end{align*}
-where \changed{$j \leq m$ is the iteration number, $m$ is the maximum
-  number of underlying basis functions}{}, $1 \le p \le n$, and
-$B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
- $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$ and
-$w_- = (-w)_+$. After iteratively constructing a model, MARS then
+where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
+$B_k(x)$ are basis functions from the previous iteration,
+ $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
+and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
 iteratively removes basis functions that do not contribute to goodness
@@
 product approximation of the data. The overall computational
-complexity of Fast MARS is $\mathcal{O}(n d m^3)$ \changed{}{where $m$
-  is the maximum number of underlying basis functions}. This paper
-uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m =
-200$ throughout all experiments.
+complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
+maximum number of underlying basis functions. This paper uses an
+implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
+throughout all experiments.
 
@@
 $m$ being determined by the minimization convergence criterion. This
-paper uses the scikit-learn SVR \cite{scikit-learn} with a radial
-basis function kernel.
+paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial
+kernel function.
 
@@
 regression and classification tasks
-\cite{hornik1989multilayer,rumelhart1988learning}. When using the
+\cite{rumelhart1988learning,hornik1989multilayer}. When using the
 rectified linear unit (ReLU) activation function
@@
 with ten hidden layers each having thirty two nodes (a total of
-approximately ten thousand parameters), ReLU activation, and
-\changed{one hundred}{five} thousand \changed{steps}{epochs} of SGD
-for training. \changed{It should be noted that the choice of neural
-  network architecture impacts approximation performance, but no
-  architecture tuning is performed here}{}.
+approximately ten thousand parameters), ReLU activation, and five
+thousand epochs of SGD for training.
 
@@
 
-\changed{
-\begin{figure}
-  \centering
-  \includegraphics[width=0.4\textwidth,]{example_delaunay.pdf}
-  \includegraphics[width=0.4\textwidth,]{example_not_delaunay.pdf}
-  \caption{\color{red} On the left above is a depiction of a Delaunay
-    triangulation over four points, notice that the circumball (shaded
-    circle) for the left simplex does not contain the fourth point.
-    On the right above, a non-Delaunay mesh is depicted. Notice that
-    the circumball for the top simplex (shaded circle, clipped at
-    bottom edge of the visual) contains the fourth point which
-    violates the Delaunay condition for a simplex.
-  \vspace{-.1cm}}
-  \label{fig:delaunay}
-\end{figure}
-}{}
-
 \subsection{Modified Shepard}
@@
 \textit{metric interpolation}, also called a Shepard method
-\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
+\cite{shepard1968two,gordon1978shepard}. The interpolant has the form
  $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
      {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k - \bigl\|x -
-x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x - x^{(k)}\bigr\|_2 \bigr)\bigr)^2$}{}
-and $r_k \in \mathbb{R}$ is the smallest radius such that at least
-$d+1$ other points $x^{(j)}$, $j \not = k$, are inside the closed
-Euclidean ball of radius $r_k$ about $x^{(k)}$. The computational
-complexity of ShepMod is $\mathcal{O}(n^2d)$. This paper uses a
-Fortran 95 implementation of ShepMod derived from SHEPPACK
-\cite{thacker2010algorithm}.
-
-\changed{
-\begin{figure}
-  \centering
-  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
-  \caption{\color{red} On the left above is a depiction of the radius
-    of influence for three chosen points of a collection using the
-    modified Shepard criteria. On the right is the shape of the weight
-    function $W_k$ for all data points, where $W_k(x) = 0$ when
-    $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen that in the limit
-    as $W_k$ goes to infinity at $x_k$, the relative weight of $x_k$
-    goes to one.
-  \vspace{-.1cm}}
-  \label{fig:shepard}
-\end{figure}
-}{}
+where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
+\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
+smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
+\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
+$x^{(k)}$. The computational complexity of ShepMod is
+$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
+ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
 
@@
  {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is \changed{the same as for the modified Shepard method}{a locally supported weighting function} and $P_k(x)$
+where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
 is a local linear approximation to the data satisfying
@@
 
-The box spline model used here is \changed{an}{sparse grid}
-interpolation technique built from overlapping box splines
-\cite{de2013box}. The box splines serve as basis functions that are
-shifted and scaled to have support over box shaped regions. The boxes
-are constructed in a way to guarantee a covering of the domain
-\cite{lux2018novel}. Given a set of box splines
+The box spline model used here is an interpolation technique built
+from overlapping box splines \cite{de2013box}. The box splines serve
+as basis functions that are shifted and scaled to have support over
+box shaped regions. The boxes are constructed in a way to guarantee a
+covering of the domain \cite{lux2018novel}. Given a set of box splines
 $\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
@@
 \label{sec:voronoi}
-
-\changed{
-\begin{figure}
-  \centering
-  \includegraphics[width=0.6\textwidth,]{VM_Construction.pdf}
-  \caption{\color{red} Above is a depiction of the Voronoi cell
-    boundaries (dashed lines) about a set of interpolation points
-    (dots). In this example, the Voronoi model basis function about
-    the center most point has nonzero weight in the shaded region and
-    transitions from a value of one at the point to zero at the
-    boundary of the twice expanded Voronoi cell (solid line).
-  \vspace{-.1cm}}
-  \label{fig:voronoi-basis}
-\end{figure}
-}{}
 
@@
     CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
-    example. For this example the KS test null hypothesis is rejected
-    at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
+    example. The KS test null hypothesis is rejected at $p$-value
+    0.01, however it is not rejected at $p$-value 0.001.
   \vspace{-.1cm}}
@@
 reported with summary statistics including: min absolute error, max
-absolute error, and absolute error quartiles. When the range
-of an approximation is the space of cumulative distribution functions,
-the Kolmogorov-Smirnov statistic (max-norm difference between the
+absolute error, and absolute error quartiles. When the range of an
+approximation is the space of cumulative distribution functions, the
+Kolmogorov-Smirnov statistic (max-norm difference between the
 functions) is used.
@@
 PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
-\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. \changed{Instead}{However}, for a CDF
+\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
 $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
@@
 this is applied can be seen in Figure \ref{fig:prediction-example}.
-\changed{In this example the KS test null hypothesis is rejected at $p$-value
-0.01, however it is not rejected at $p$-value 0.001.}{}
 
@@
 applications of the KS test presented in this work $n_1 = n_2$. An
-example of the \changed{end-to-end}{round-trip} prediction methodology from known and
+example of the round-trip prediction methodology from known and
 predicted distributions to the calculation of error can be seen in
-Figure \ref{fig:prediction-example}. A brief listing of relevant
-statistical terms used throughout this work is provided in
-the Appendix (Section \ref{sec:appendix}).
+Figure \ref{fig:prediction-example}.
 
@@
 paper compute predictions as convex combinations of observed function
-values, which may allow for straightforward extensions of this error
+values, which may allow for straight forward extensions of this error
 bound.
@@
   \label{lemma:1}
-  Let $S \subset \mathbb{R}^d$ be open and convex, $f: S \rightarrow
-  \mathbb{R}$, and let $\nabla f(x)$ be $\gamma$-Lipschitz continuous
-  in the $2$-norm. Then for all $x,y \in S$
+  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
+  \rightarrow \mathbb{R}$, and $\nabla f \in
+  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
+  continuous functions in the $2$-norm. Then for all $x,y \in S$
   $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
@@
   \begin{align*}
-    \big|f(y) - f(x) - \langle &\nabla f(x), y - x \rangle \big|
-        = \big|g(1) - g(0) - g'(0) \big| \\
-       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
+    \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
+       &= \big|g(1) - g(0) - g'(0) \big| & \\
+       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
+       &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
        &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
        &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
-       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt = \frac{\gamma \|y - x\|_2^2}{2}.
+       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
+       &= \frac{\gamma \|y - x\|_2^2}{2}.
   \end{align*}
@@
 \end{proofdot}
-
-\hfill
 
@@
 
-%% \newpage
-
 \begin{plainlemma}
@@
   \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
-  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \frac{\sqrt{d} \, \gamma k^2}{2 \sigma_d}.$$
+  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
 \end{plainlemma}
@@
 \begin{proofdot}
-  \def\g{\gamma_g}
-  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
-  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
-  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
-  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
-  = \gamma \|x_j-x_0\|_2^2$.
-
-  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
-  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
-  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
-  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
-  and so $g'(0) - \g z \le g'(z) = 0$. Now
-  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
-  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
+  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
+  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
+  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
+  \big\rangle = 0.$ Now
   \begin{align*}
-    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
-    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
-    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
+    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
+        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
+	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
+	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
+	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
+	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
   \end{align*}
-  \noindent a contradiction for the value of $g(\tilde t)$. A similar
-  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
-  \g/2$. In terms of $f$,
-  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
-  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
-  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
-  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
-  \implies \|\nabla f(x_0)-c\|_2 \le \frac{\sqrt{d} \, \gamma
-    k^2}{2\sigma_d}.$$ \qed
+  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
+  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
+  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
+  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
+  \qed
 \end{proofdot}
@@
     \ref{lemma:3}}, for $z \in S$,
-  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d} \, \gamma k^2}{2 \sigma_d} \|z - x_0\|_2.$$
+  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
 \end{onetheorem}
@@
 \begin{proofdot}
-  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d}\, \gamma k^2 / (2 \sigma_d)$ by {\it Lemma \ref{lemma:3}}. Now
+  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
   \begin{align*}
@@
        &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \big(\sqrt{d}\,\gamma k^2 / (2 \sigma_d)\big) \|z - x_0\|_2 \\
-       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d}\,\gamma k^2}{2 \sigma_d} \|z - x_0\|_2,
+       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
+       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
   \end{align*}
@@
 diameter, or are nearly contained in a hyperplane) have higher
-error.
-
-That the theoretical error bound is sharp can be observed with the
-test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d,$ and the
-simplex defined by vertices $X = \{0, e_1, \ldots, e_d\},$ where $e_i$
-is the $i$-th standard basis vector in $\mathbb{R}^d,$ $e = (1$,
-$\ldots$, $1) \in \mathbb{R}^d,$ and $0$ denotes the zero vector in
-any dimension.  The constants relevant to the error bound are
-$$ \gamma = 2, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 = 0,
-\qquad \hat q(x) = \langle e, x - 0\rangle + q(0). $$
-\noindent Noting that $q(0) = 0,$ the approximation error at $z =
--(1/2)e$ is
-$$ \big|q(z) - \hat q(z)\big| = \big|\|z\|_2^2 - \langle e,z
-\rangle\big| = \big| d/4 + d/2 \big| = 3d/4, $$
-\noindent while the error bound from the theorem gives
-
-\vspace{-2mm}
-\begin{align*}
-  \big|q(z) - \hat q(z)\big|
-  &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d}\,\gamma k}
-      {2 \sigma_d} \|z - x_0\|_2 \\
-  &{}=  \|z\|_2^2 + \sqrt{d} \|z\|_2 \\
-  &{}=  d/4 + d/2 = 3d/4.
-\end{align*}
-\vspace{-2mm}
-
-\noindent Acknowledging that the error bound is sharp, it may be of
-interest to observe the error of piecewise linear approximation
-techniques on an analytic test function.
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.9\textwidth,]{Fig2.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
-  \end{tikzpicture}
-  \caption{Delaunay and MLP approximations are constructed from Fekete
-    points over the unit cube evaluating the test function $f(x) =
-    \cos(\|x\|_2)$ for $x \in \mathbb{R}^2$. The figure shows the
-    first/third quartiles at the box bottom/top, the second quartile
-    (median) at the white bar, median 95\% confidence interval (cones,
-    barely visible in figure), and whiskers at 3/2 of the adjacent
-    interquartile ranges, for the absolute prediction error for each
-    model at $1000$ random evaluation points. The left plot observes a
-    perfect interpolation problem with exact evaluations of $f.$ The
-    right plot observes a regression problem with uniform random noise
-    giving values in $[.9 f(x),\ 1.1f(x)]$ for each $x.$ Both axes are
-    log scaled.}
-  \label{fig:convergence-2d}
-\end{figure}
+error. In light of this error bound, an empirical evaluation of
+presented algorithms follows.
 
@@
 \vspace{-2mm}
-\subsection{Demonstration on an Analytic Test Function}
-\label{sec:analytic}
+\section{Data and Empirical Analysis}
+\label{sec:data}
 
@@
 context with approximation algorithms that do not have similar known
-uniform error bounds. For that reason, an analytic function is used to
-measure the error of a piecewise linear interpolant (Delaunay) and a
-piecewise linear regressor (the MLP) when provided an increasing
-amount of data with varying levels of random noise. \changed{Only
-  Delaunay and the MLP are considered in this demonstration because
-  all other mentioned techniques are not strictly piecewise linear
-  approximations.}{}
-
-The test function chosen here for analysis resembles the
-\textit{oscillatory} function used by \cite{barthelmann2000high}.
-However, a slight modification is made to remove the simple dot
-product structure (which is favorable for the MLP). Let $f(x)
-=\cos(\|x\|_2)$ for $x \in \mathbb{R}^d$ where $d$ is either $2$
-(Fig. \ref{fig:convergence-2d}) or $20$
-(Fig. \ref{fig:convergence-20d}). This function has a bounded change
-in gradient $2$-norm and hence meets the necessary Lipschitz condition
-for the error bound. Data points and approximation points for this
-test will be within the unit cube $[0,1]^d$.
-
-The error of a linear interpolant constructed from well spaced points
-is dominated by the distance to the nearest data point. In order to
-uniformly decrease the distance to the nearest data point across the
-unit cube, exponentially more data points must be available for
-approximation. For this experiment $N = 2(4^i),$ $i = 1,$ $\ldots,$
-$6,$ chosen points are kept well spaced by computing approximate
-Fekete points. Fekete points have a history in potential theory
-\cite{kovari_pommerenke_1968} and are most generally defined as those
-points that maximize the absolute value of a Vandermonde
-determinant. Here, the QR method outlined in \cite{bos2010computing}
-is used to identify approximate Fekete points from a Vandermonde
-matrix. A multivariate polynomial basis of size $2N$ is constructed
-containing all polynomials of degree $\leq n$ for the largest $n$ such
-that ${d+n \choose n} \leq 2N$ and $2N - {d+n \choose n}$
-arbitrarily selected polynomials of degree $n + 1.$ The Vandermonde
-matrix for these $2N$ polynomials is used to select $N$ approximate
-Fekete points.
-
-Finally an additional aspect is added to this test problem by
-incorporating random uniform noise into the evaluations of $f.$ For
-each test two experiments are executed, one with exact function
-evaluations (an interpolation problem) and one with a constant
-signal-to-noise (SNR) ratio of 10:1 (a regression problem).
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.9\textwidth,]{Fig3.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
-  \end{tikzpicture}
-  \caption{Delaunay and MLP approximations are constructed from Fekete
-    points over the unit cube evaluating the test function $f(x) =
-    \cos(\|x\|_2)$ for $x \in \mathbb{R}^{20}$. The details are the
-    same as for Fig. \ref{fig:convergence-2d}.}
-  \label{fig:convergence-20d}
-\end{figure}
-
-The bound from the theorem suggests that exponentially increasing the
-number of \changed{well-spaced}{} data points should result in a
-decreasing error. The $d = 2$ test seen on the left half of Figure
-\ref{fig:convergence-2d} shows a consistent decrease in error for
-Delaunay and also shows the eventual accuracy plateau obtained by a
-parametric regression form (the MLP, at roughly 500 points). On the
-right hand side of Figure \ref{fig:convergence-2d} the random noise
-clearly prohibits Delaunay from converging to $f$, while the MLP is
-able to improve its approximation with more data points on average.
-The convergence result for a very low-dimensional problem like this is
-expected.  However, intuition fails for higher dimensional problems.
-
-Figure \ref{fig:convergence-20d} shows the test function with $d = 20$
-presents a significantly more challenging approximation problem than
-its counterpart in low dimension. The same increase in number of data
-points from 32 to 8192 causes no apparent improvement in approximation
-for either the noise-free or the noisy problems. Perhaps unexpectedly,
-the interpolation technique performs better than the regression
-technique on the noisy data (right) in Figure
-\ref{fig:convergence-20d}, and worse than the regression technique on
-the noise-free data (left). This result emphasizes the relevance of
-interpolation for problems in high dimension. It also reveals that the
-outcome of MLP regressions can get worse when adding more data points.
-
-\changed{
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.9\textwidth,]{theorem-convergence.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Euclidean Distance};
-  \end{tikzpicture}
-  \caption{\color{red}The distribution of absolute error, distance to
-    the nearest data point, smallest singular value and the longest
-    edge of the simplex containing each approximation point in the
-    tests from Fig. \ref{fig:convergence-2d} and
-    Fig. \ref{fig:convergence-20d} for Delaunay. In two dimensions it
-    can be seen that $\|z - x_0\|_2$, $\sigma_d$, and $k$ all shrink
-    at the same rate for well-spaced approximation points, resulting
-    in a faster rate of decrease for approximation error. Notice that
-    in higher dimension the data remains sparse even with thousands of
-    data points. This is seen by the relatively small reduction in
-    $k$, and this explains the slow reduction in error seen by
-    Delaunay in Fig. \ref{fig:convergence-20d}.}
-  \label{fig:data-spacing}
-\end{figure}
-}{}
-
-\changed{The results achieved for both the $d = 2$ and $d = 20$ align
-  with the theoretical error bound as can be seen in Figure
-  \ref{fig:data-spacing}. In low dimension, thousands of data points
-  meaningfully reduce the sparsity of the approximation problem.
-  However, in higher dimension it takes many more points to achieve a
-  reduction in data sparsity. The same result holds for a Latin
-  hypercube well-spaced point set, suggesting that sparsity is an
-  inherent problem in high dimension approximation problems. This
-  result is not a product of the chosen test function, but rather the
-  exponential increase in complexity that accompanies increased
-  dimension.}{}
-
-In summary, the regime of interpolation is \changed{}{much} greater
-for moderate to high dimensional problems \changed{due to increased
-  sparsity}{}, even for functions with random noise incorporated into
-their evaluation. Acknowledging the viability of interpolation for
-problems of moderate dimension, the next section will consider
-real-world problems of similar proportion (thousands of examples in
-tens of dimensions).
-
-%% ===================================================================
-\vspace{-2mm}
-\section{Data and Empirical Analysis}
-\label{sec:data}
-
-This section extends the comparison of interpolation and regression
-algorithms to a sample of real-world problems. Five different data
-sets of varying dimension and application are utilized to construct
+uniform error bounds. For that reason, this paper utilizes five
+different data sets of varying dimension and application to construct
 approximations and compare the accuracy of different techniques.
@@
 data set, tables of accompanying numerical results are located in the
-Appendix (Section \ref{sec:appendix}). All of the test data sets
-capture underlying functions that are almost certainly stochastic. As
-described in Section \ref{sec:introduction}, regression techniques
-appear most appropriate for these problems. However, typically data
-grows exponentially more sparse with increasing dimension. Given that
-sparse data regressors tend towards interpolation and, as demonstrated
-in Section \ref{sec:analytic}, interpolants produce similar (if not
-identical) results, it is presumed that interpolants are equally
-viable approximation techniques on these problems.
+Appendix. All of the test data sets capture underlying functions that
+are almost certainly stochastic. As described in Section
+\ref{sec:introduction}, regression techniques appear most appropriate
+for these problems. However, typically data grows exponentially more
+sparse with increasing dimension. Given sparse data, regressors tend
+towards interpolation.
 
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig2.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig3.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig6.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig7.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig8.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig6.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig9.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig7.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig10.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig8.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig11.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig9.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig12.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig10.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughput};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig13.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig11.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughput KS Statistic};
@@
     previously unseen system configuration, using $10$-fold cross
-    validation. The KS statistic (max norm) between the observed
-    distribution at each system configuration and the predicted
-    distribution is recorded and presented above. Note that the above
-    figure is \textit{not} log-scaled like Figure
-    \ref{fig:hist-throughput}. The numerical data corresponding to
-    this figure is provided in Table \ref{table:error-throughput} in
-    the Appendix.}
+    validation. The KS statistic between the observed distribution at
+    each system configuration and the predicted distribution is
+    recorded and presented above. Note that the above figure is
+    \textit{not} log-scaled like Figure \ref{fig:hist-throughput}. The
+    numerical data corresponding to this figure is provided in Table
+    \ref{table:error-throughput} in the Appendix.}
   \label{fig:error-throughput}
@@
 
-Delaunay achieves the lowest KS statistic (max norm difference) for
-$62\%$ of approximations, while Voronoi is best for the remaining
-$38\%$. Figure \ref{fig:error-throughput} shows that while Delaunay
-may have more best predictions, the behavior of Voronoi may be
-preferable.
+Delaunay achieves the lowest KS statistic for $62\%$ of
+approximations, while Voronoi is best for the remaining $38\%$. Figure
+\ref{fig:error-throughput} shows that while Delaunay may have more
+best predictions, the behavior of Voronoi may be preferable.
 %% -------------------------------------------------------------------
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth]{Fig14.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth]{Fig12.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
@@
     Voronoi & $52.6\%$ & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
-    BoxSplines & $99.4\%$          & $98.5\%$          & $96.6\%$          & $89.3\%$\\
+    BoxSpline & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
   \end{tabular}
@@
     \hline
-    MARS        & \quad\quad$4$&$5$          & \quad\quad$20$&$0$s        & \qquad\quad $0$&$001$s\\
-    SVR         & $\mathit{19}$&$\mathit{5}$ & $\mathbf{0}$&$\mathbf{5}$s & $\mathbf{0}$&$\mathbf{0001}$s\\
-    MLP         & $\mathbf{43}$&$\mathbf{1}$ & $200$&$0$s                 & $0$&$001$s\\
-    Delaunay    & $5$&$2$                    & $1$&$0$s                   & $1$&$0$s\\
-    ShepMod     & $18$&$0$                   & $\textit{0}$&$\textit{7}$s & $\mathbf{0}$&$\mathbf{0001}$s\\
-    LSHEP       & $8$&$4$                    & $2$&$0$s                   & $\mathbf{0}$&$\mathbf{0001}$s\\
-    Voronoi     & $0$&$5$                    & $1$&$0$s                   & $0$&$04$s\\
-    BoxSplines  & $3$&$5$                    & $0$&$8$s                   & $\mathit{0}$&$\mathit{0005}$s\\
+    MARS & \quad\quad$4$&$6$ & \,\quad$20$&$0$ & \quad\quad$0$&$001$\\
+    SVR & $\mathit{21}$&$\mathit{1}$ & $\mathbf{0}$&$\mathbf{5}$ & $\mathbf{0}$&$\mathbf{0001}$\\
+    MLP & $\mathbf{42}$&$\mathbf{1}$ & $200$&$0$ & $0$&$001$\\
+    Delaunay & $6$&$0$ & $1$&$0$ & $1$&$0$\\
+    ShepMod & $18$&$4$ & $\mathit{0}$&$\mathit{7}$ & $\mathbf{0}$&$\mathbf{0001}$\\
+    LSHEP & $8$&$4$ & $2$&$0$ & $\mathbf{0}$&$\mathbf{0001}$\\
+    Voronoi & $0$&$5$ & $1$&$0$ & $0$&$04$\\
+    BoxSpline & $0$&$2$ & $0$&$8$ & $\mathit{0}$&$\mathit{0005}$\\
   \end{tabular}
@@
 Finally, most of the interpolants presented in this work benefit from
-the ability to model \textit{any} function over real $d$-tuples with a
-range that is closed under convex combinations. In general, error can
-be quantified by any measure (particularly of interest may be
-$L^2$, $L^\infty$, etc.). The results of the distribution prediction
-case study indicate that interpolants can effectively predict
-CDFs. The error analysis for that work relies on the KS statistic,
-which captures the worst part of any prediction and hence provides a
-conservatively large estimate of approximation error. The average
-absolute errors in the predicted CDFs are always lower than the KS
-statistics. However, the KS statistic was chosen as a metric because
-of the important surrounding statistical theory. A nonnegligible
-volume of predictions provide impressively low levels of average
-absolute error in that final case study.
+the ability to model any function over real $d$-tuples with a range
+that is closed under convex combinations. The results of the
+distribution prediction case study indicate that interpolants can
+effectively predict CDFs. The error analysis for that work relies on
+the KS statistic, which captures the worst part of any prediction and
+hence provides a conservatively large estimate of approximation
+error. The average absolute errors in the predicted CDFs are always
+lower than the KS statistics. However, the KS statistic was chosen as
+a metric because of the important surrounding statistical theory. A
+nonnegligible volume of predictions provide impressively low levels of
+average absolute error in that final case study.
 
@@
 encourage broader application of interpolants to multivariate
-approximation in science.
+approximation in science
 
@@
 \begin{appendix}
-\section{Appendix}
+\section*{Appendix}
 \label{sec:appendix}
 
-\begin{normalsize}
-\textbf{Statistical Terminology.} A random variable $X$ is precisely
-defined by its \textit{cumulative distribution function} (CDF) $F_X$
-and the derivative of the CDF, the \textit{probability density
-  function} (PDF) $f_X.$ For any possible value $x$ of $X$, the
-\textit{percentile} of $x$ is $100$ $F_X(x)$, the percentage values
-drawn from $X$ that would be less than or equal to $x$ as the number
-of samples tends towards infinity. The \textit{quartiles} of $X$ are
-its $25$-th, $50$-th (median), and $75$-th percentiles. The absolute
-difference between the median and an adjacent quartile is an
-\textit{interquartile range}. Given an independent and identically
-distributed sample from $X$ and presuming that $X$ has finite mean and
-variance, a \textit{confidence interval} can be drawn about any
-percentile estimated from the sample. A \textit{confidence interval}
-describes the probability that a value lies within an interval. The
-\textit{null hypothesis} is a statement (derived from some test
-statistic) that the expected value of the observed statistic is equal
-to an assumed population statistic. The $p$ value is the probability
-of observing a given statistic if the \textit{null hypothesis} is
-true. \changed{For a more detailed introduction to statistics and 
-related terminology, see the work of Navidi \cite{navidi_2015}.}{}
-
-\vspace{4mm}
-\noindent \textbf{Raw Numerical Results.} The tables that follow show the
-precise experimental results for all data sets presented in Section
-\ref{sec:data}. The tests were all run serially on an otherwise idle
-machine with a CentOS 6.10 operating system and an Intel i7-3770 CPU
-operating at 3.4 GHz. The detailed performance results in the tables
-that follow are very much dependent on the problem and the algorithm
-implementation (e.g., some codes are TOMS software, some industry
-distributions, and others are from conference paper venues). Different
-typeface is used to show best performers, however not much
-significance should be attached to ranking algorithms based on small
-time (millisecond) differences. The results serve as a demonstration
-of conceptual validity.
-
-\end{normalsize}
-
-\newpage
+The tables that follow show the precise experimental results for all
+data sets presented in Section \ref{sec:data}. The tests were all run
+serially on an otherwise idle machine with a CentOS 6.10 operating
+system and an Intel i7-3770 CPU operating at 3.4 GHz. The detailed
+performance results in the tables that follow are very much dependent
+on the problem and the algorithm implementation (e.g., some codes are
+TOMS software, some industry distributions, and others are from
+conference paper venues). Different typeface is used to show best
+performers, however not much significance should be attached to
+ranking algorithms based on small time (millisecond) differences. The
+results serve as a demonstration of conceptual validity.
 
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}       & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
     MARS & $0$&$00984$ & $3$&$11$ & $7$&$01$ & $\mathit{11}$&$\mathit{7}$ & $1090$&$0$\\
-    SVR       & $0$&$0118$                    & $\mathbf{0}$&$\mathbf{615}$   & $\mathbf{0}$&$\mathbf{931}$   & $\mathbf{5}$&$\mathbf{89}$    & $1090$&$0$\\
-    MLP       & $0$&$0426$                    & $2$&$63$                      & $5$&$27$                      & $14$&$0$                      & $1090$&$0$\\
-    Delaunay  & $\mathbf{0}$&$\mathbf{0}$     & $1$&$98$                      & $5$&$37$                      & $13$&$1$                      & $\mathit{1080}$&$\mathit{0}$\\
-    ShepMod   & $\mathbf{0}$&$\mathbf{0}$     & $1$&$93$                      & $6$&$27$                      & $16$&$0$                      & $1090$&$0$\\
-    LSHEP     & $0$&$04$                      & $2$&$17$                      & $8$&$87$                      & $19$&$1$                      & $\mathbf{1070}$&$\mathbf{0}$\\
-    Voronoi   & $\mathit{0}$&$\mathit{00982}$ & $3$&$65$                      & $7$&$56$                      & $15$&$6$                      & $1090$&$0$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{0}$     & $\mathit{1}$&$\mathit{27}$    & $\mathit{4}$&$\mathit{61}$    & $12$&$6$                      & $1090$&$0$\\
+    SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
+    MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
+    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
+    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
+    LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
+    Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
     \hline
@@
     \cline{1-3}\cline{5-10}
-    MARS      & \quad$7$&$3$                   &  & \quad\quad$29$&$1$                      & \quad$0$&$00137$                   & \quad\quad\quad$0$&$0686$\\
-    SVR       & $\mathbf{78}$&$\mathbf{0}$     &  & $\mathbf{0}$&$\mathbf{00584}$           & $\mathit{0}$&$\mathit{0000620}$    & $\mathit{0}$&$\mathit{00310}$\\
+    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
+    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathbf{0}$&$\mathbf{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
     MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
-    Delaunay  & $0$&$2$                        &  & $0$&$0151$                              & $0$&$0234$                         & $1$&$18$\\
-    ShepMod   & $2$&$0$                        &  & $\mathit{0}$&$\mathit{00634}$           & $0$&$0000644$                      & $0$&$00322$\\
+    Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
+    ShepMod & $3$&$7$ &  & $\mathit{0}$&$\mathit{00634}$ & $0$&$0000644$ & $0$&$00322$\\
     LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
     Voronoi & $0$&$0$ &  & $0$&$0152$ & $0$&$000396$ & $0$&$0198$\\
-    BoxSpline & $\mathit{9}$&$\mathit{7}$      &  & $0$&$00724$                             & $0$&$0000978$                      & $0$&$00489$\\
+    BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
     \cline{1-3}\cline{5-10}
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}                   & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
     MARS & $0$&$00948$ & $9$&$98$ & $20$&$4$ & $32$&$9$ & $863$&$0$\\
-    SVR       & $0$&$00233$                               & $3$&$15$                      & $7$&$21$                      & $11$&$3$                      & $\mathbf{28}$&$\mathbf{6}$\\
+    SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
     MLP & $0$&$0000239$ & $\mathit{0}$&$\mathit{533}$ & $\mathit{1}$&$\mathit{25}$ & $\mathbf{2}$&$\mathbf{84}$ & $39$&$3$\\
-    Delaunay  & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$2$                       & $3$&$5$                       & $7$&$67$                      & $30$&$7$\\
+    Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
     ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
@@
     Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
-    BoxSpline & $0$&$006$                                 & $4$&$3$                       & $8$&$91$                      & $15$&$1$                      & $45$&$3$\\
+    BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
     \hline
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$}              & \multicolumn{2}{c}{$50^{th}$}  & \multicolumn{2}{c}{$75^{th}$}  & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
     MARS & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$ & $1$&$96$ & $\mathit{53}$&$\mathit{3}$\\
-    SVR       & $0$&$0000915$                              & $0$&$0833$                                 & $0$&$263$                      & $\mathit{1}$&$\mathit{13}$     & $109$&$0$\\
+    SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
     MLP & $0$&$0000689$ & $0$&$0337$ & $\mathbf{0}$&$\mathbf{0795}$ & $\mathbf{0}$&$\mathbf{264}$ & $\mathbf{5}$&$\mathbf{31}$\\
     Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
-    ShepMod   & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$0957$                                 & $0$&$685$                      & $2$&$9$                        & $73$&$2$\\
+    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
     LSHEP & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
-    Voronoi   & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$43$                                   & $1$&$28$                       & $2$&$94$                       & $83$&$8$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$342$                                  & $1$&$59$                       & $5$&$53$                       & $119$&$0$\\
+    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{0}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
     \hline
@@
     MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\quad$\mathit{0}$&$\mathit{151}$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
-    SVR       & $0$&$0$                      &  & $\mathbf{0}$&$\mathbf{133}$                & $\mathbf{0}$&$\mathbf{0000947}$    & $\mathbf{0}$&$\mathbf{0246}$\\
-    MLP       & $\mathbf{60}$&$\mathbf{9}$   &  & $169$&$0$                                  & $0$&$00137$                        & $0$&$356$\\
+    SVR & $4$&$1$ &  & $\mathbf{0}$&$\mathbf{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
+    MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
     Delaunay & $0$&$1$ &  & $0$&$664$ & $0$&$886$ & $230$&$0$\\
@@
     Voronoi & $0$&$2$ &  & $0$&$675$ & $0$&$0270$ & $7$&$01$\\
-    BoxSpline & $4$&$1$                      &  & $0$&$330$                                  & $0$&$000406$                       & $0$&$106$\\
+    BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
     \cline{1-3}\cline{5-10}
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
     MARS & $5$&$36$ & $3610$&$0$ & $4580$&$0$ & $5450$&$0$ & $13400$&$0$\\
-    SVR       & $0$&$00706$                                & $8$&$14$                      & $\mathit{13}$&$\mathit{0}$    & $41$&$6$                      & $7690$&$0$\\
-    MLP       & $0$&$00151$                                & $\mathbf{1}$&$\mathbf{6}$     & $\mathbf{3}$&$\mathbf{86}$    & $\mathbf{8}$&$\mathbf{34}$    & $\mathbf{604}$&$\mathbf{0}$\\
-    Delaunay  & $\mathbf{0}$&$\mathbf{0}$                  & $\mathit{2}$&$\mathit{69}$    & $13$&$8$                      & $\mathit{35}$&$\mathit{0}$    & $\mathit{4840}$&$\mathit{0}$\\
+    SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
+    MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
+    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
     ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
@@
     Voronoi & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$ & $32$&$8$ & $52$&$9$ & $4860$&$0$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{0}$                  & $12$&$4$                      & $35$&$0$                      & $90$&$1$                      & $7690$&$0$\\
+    BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
     \hline
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}      & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
     Delaunay & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$ & $\mathbf{0}$&$\mathbf{158}$ & $\mathit{0}$&$\mathit{308}$ & $\mathit{0}$&$\mathit{897}$\\
-    ShepMod   & $0$&$0307$                   & $\mathit{0}$&$\mathit{0983}$  & $\mathit{0}$&$\mathit{164}$   & $0$&$335$                     & $1$&$0$\\
-    Voronoi   & $\mathit{0}$&$\mathit{0303}$ & $0$&$1$                       & $0$&$165$                     & $\mathbf{0}$&$\mathbf{274}$   & $\mathbf{0}$&$\mathbf{762}$\\
-    BoxSpline & $0$&$0679$                   & $0$&$429$                     & $0$&$571$                     & $0$&$752$                     & $1$&$0$\\
+    ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
+    Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
+    BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
     \hline

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-interpolation/Versions/2-response.tex
Finished NUMA paper and submitted. Might need to do one more round with editors.

--- 2020-04-01 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-response-2020-04.tex
@@
 Addressed by item \ref{review:item1}.
+However, sparse grids are not included in the comparison because they do not scale beyond tens of dimensions and are primarily used for applications where data points can be chosen (although as a reviewer mentioned, they can be applied to scattered data problems).
 
@@
 All of these have been addressed in each individual subsection and in part by item \ref{review:item2} above.
-Regarding the shape of $W_k$ in Modified Shepard and LSHEP, the weight equation was rewritten to provoke a more clear understanding in accompaniment with the visual.
+Regarding the shape of $W_k$ in Modified Shepard and LSHEP, the weight equation was rewritten to provide a more clear understanding in accompaniment with the visual.
 
@@
 
-The phrasing has been changed (page \hyperlink{page.10}{10}). This is a reference the process of converting source data into predictions and then measuring the error.
+The phrasing has been changed (page \hyperlink{page.10}{10}). This is a reference to the process of converting source data into predictions and then measuring the error.
 
@@
 This is a misunderstanding.
-The paper reads ``\ldots polynomials of \color{red}degree \color{black} $n+1$.'' and the degree of a polynomial is one less than the order.
-So order $n$ polynomials are used here.
+The paper reads ``\ldots polynomials of \color{red}degree \color{black} $n+1$.'' For these Fekete points we chose a polynomial basis of size $2N$ and have fixed $d$, but there is not necessarily a value of $n$ for which ${n + d \choose n} = 2N$, so we must augment the ${n+d \choose n}$ basis polynomials of degree $n$ with a few higher degree ones to get exactly $2N$.
 
@@
 
-This suggestion is appreciated.
-Such specific outcomes are not obviously related to the error bound theorem, which is the central component of this paper.
-For this reason, we would like to abstain from including information that might distract readers from the main purposes of the text.
+Done on page \hyperlink{page.29}{29} and referenced at the beginning of the empirical results section on page \hyperlink{page.17}{17}
 

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-interpolation/Versions/3-paper.tex
Finished NUMA paper and submitted. Might need to do one more round with editors.

--- 2020-04-01 [2019-02]_NUMA_(interpolant_error_bound)/paper.tex
@@
 
-%% Include the response to reviewers at the beginning of the document.
-\include{tchlux-NA-response-2020-04}
+%% %% Include the response to reviewers at the beginning of the document.
+%% \include{tchlux-NA-response-2_2020-04}
 
@@
 Regression and interpolation have a considerable theoretical base in
-one dimension \cite{cheney2009course}. \changed{Common techniques
-  include Lagrangian interpolation, piecewise linear interpolation,
-  and more generally spline interpolation via B-splines
-  \cite{de1978practical}}{Splines in particular are well understood
-  as an interpolation technique in one dimension
-  \cite{de1978practical}, particularly B-splines}. Tensor products of
-B-splines \cite{unther1996interpolating} or other basis functions have
-an unfortunate exponential scaling with increasing
-dimension. Exponential scaling prohibits tensor products from being
-reasonably applied beyond three-dimensional data. In order to address
-this dimensional scaling challenge, C. de Boor and others proposed box
-splines \cite{de2013box}, of which one of the approximation techniques
-in this work is composed \cite{lux2018novel}.
+one dimension \cite{cheney2009course}. Common techniques include
+Lagrangian interpolation, piecewise linear interpolation, and more
+generally spline interpolation via B-splines
+\cite{de1978practical}. Tensor products of B-splines
+\cite{unther1996interpolating} or other basis functions have an
+unfortunate exponential scaling with increasing dimension. Exponential
+scaling prohibits tensor products from being reasonably applied beyond
+three-dimensional data. In order to address this dimensional scaling
+challenge, C. de Boor and others proposed box splines
+\cite{de2013box}, of which one of the approximation techniques in this
+work is composed \cite{lux2018novel}.
 
 The theoretical foundation of low dimensional interpolation allows the
-construction of strong error bounds that are \changed{lacking
-  in}{absent from} high dimensional problems. \changed{A large body of
-  literature exists studying the construction and application of
-  sparse grids \cite{bungartz2004sparse,chkifa2014high}, which attempt
-  to address the problem of dimensional scaling. The sparse grid point
-  sets combined with specific approximation functions are used to
-  construct total error bounds by considering projections onto
-  orthogonal polynomials
-  \cite{migliorati2013approximation,nobile2016convergence}. However,
-  sparse grids have exponential size in dimension making them
-  intractable for more than tens of dimensions and do not immediately
-  apply to the problem of scattered data interpolation.}{} This work
-extends some known results regarding the secant method
-\cite{dennis1996numerical} to construct \changed{a uniform error bound
-  for interpolants over arbitrary scattered data in any dimension}{an
-  interpolation error bound for problems of arbitrary
-  dimension}. These error bounds are useful, considering
-\changed{similar uniform bounds}{the same cannot} cannot be
-\changed{constructed}{said} for regression algorithms in general. The
-maximum complexity of an interpolant is bounded by the amount of data
-available, while the maximum complexity of a regressor is bounded by
-both the amount of data and the chosen parametric form. For this
-reason, generic uniform bounds are largely unobtainable for regression
-techniques on arbitrary approximation problems, even when the
-approximation domain is bounded.
+construction of strong error bounds that are lacking in high
+dimensional problems. A large body of literature exists studying the
+construction and application of sparse grids
+\cite{bungartz2004sparse,chkifa2014high}, which attempt to address the
+problem of dimensional scaling. The sparse grid point sets combined
+with specific approximation functions allow for the construction of
+interpolation error bounds as well as bounds on total error for
+regressors by considering projections onto orthogonal polynomials
+\cite{nobile2016convergence}. However, sparse grids are not studied
+here because their size still grows quite fast with dimension
+(although less than exponentially) making them not readily tractable
+for more than a few tens of dimensions and are primarily used for
+problems that allow data points to be chosen. This work extends some
+known results regarding the secant method \cite{dennis1996numerical}
+to construct a uniform error bound for interpolants over arbitrary
+scattered data in any dimension. These error bounds evidence that
+interpolants are useful, considering similar uniform bounds cannot be
+constructed for regressors in general
+\cite{migliorati2013approximation}. The maximum complexity of an
+interpolant is bounded by the amount of data available, while the
+maximum complexity of a regressor is bounded by both the amount of
+data and the chosen parametric form. For this reason, generic uniform
+bounds are largely unobtainable for regression techniques on arbitrary
+approximation problems, even when the approximation domain is bounded.
 
@@
 exponentially. Quantifying all possible interactions becomes
-intractable, and hence \changed{problems}{} beyond \changed{ten to
-  twenty dimensions usually rely on linear models}{three-dimensional
-  data mostly linear models are used}. That is not to say nonlinear
-models are absent, but nonlinearities are often either preconceived or
-model \changed{interactions between small subsets of}{pairwise
-  interactions between} dimensions at most. Even globally nonlinear
-approximations such as neural networks are constructed from
-compositions of summed low-interaction functions
+intractable, and hence problems beyond ten to twenty dimensions
+usually rely on linear models. That is not to say nonlinear models are
+absent, but nonlinearities are often either preconceived or model
+interactions between small subsets of dimensions at most. Even
+globally nonlinear approximations such as neural networks are
+constructed from compositions of summed low-interaction functions
 \cite{clevert2015fast}.
@@
 \end{align*}
-where \changed{$j \leq m$ is the iteration number, $m$ is the maximum
-  number of underlying basis functions}{}, $1 \le p \le n$, and
+where $j \leq m$ is the iteration number, $m$ is the maximum
+  number of underlying basis functions, $1 \le p \le n$, and
 $B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
- $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$ and
-$w_- = (-w)_+$. After iteratively constructing a model, MARS then
-iteratively removes basis functions that do not contribute to goodness
-of fit. In effect, MARS creates a locally component-wise tensor
-product approximation of the data. The overall computational
-complexity of Fast MARS is $\mathcal{O}(n d m^3)$ \changed{}{where $m$
-  is the maximum number of underlying basis functions}. This paper
-uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m =
-200$ throughout all experiments.
+ $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
+
+$w_- = (-w)_+$, and $i$, $p$, $l$, and $k$ are chosen to produce the
+best fit to the data with the addition of $B_{2j-1}$ and $B_{2j}$ to
+the existing basis, among all possible choices of $i$, $p$, $l$, and
+$k$. After iteratively constructing a model, MARS then iteratively
+removes basis functions that do not contribute to goodness of fit. In
+effect, MARS creates a locally component-wise tensor product
+approximation of the data. The overall computational complexity of
+Fast MARS is $\mathcal{O}(n d m^3)$. See the work of Friedman
+\cite{stanford1993fast} for a more thorough description of this
+approximation and its construction. This paper uses an
+implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
+throughout all experiments.
 
@@
 with ten hidden layers each having thirty two nodes (a total of
-approximately ten thousand parameters), ReLU activation, and
-\changed{one hundred}{five} thousand \changed{steps}{epochs} of SGD
-for training. \changed{It should be noted that the choice of neural
-  network architecture impacts approximation performance, but no
-  architecture tuning is performed here}{}.
+approximately ten thousand parameters), ReLU activation, and one
+hundred thousand steps of SGD for training. It should be noted that
+the choice of neural network architecture affects approximation
+performance, but no architecture tuning is performed here.
 
@@
 producing an interpolant \cite{lee1980two}. The Delaunay triangulation
-of a set of data points into simplices is such that there are no data
-points inside the sphere defined by the vertices of each simplex. For
-a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
-$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
-$\ldots$, $d$, $y \in S$ is a unique convex combination of the
-vertices,
+of a set of points into simplices is such that there
+are no points inside the sphere defined by the
+vertices of each simplex, visualized in Figure \ref{fig:delaunay}.
+For a $d$-simplex S with vertices $x^{(0)}$,
+$x^{(1)}$, $\ldots$, $x^{(d)}$, and function values
+$f\bigl(x^{(i)}\bigr)$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique
+convex combination of the vertices,
  $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
@@
 
-\changed{
 \begin{figure}
@@
   \includegraphics[width=0.4\textwidth,]{example_not_delaunay.pdf}
-  \caption{\color{red} On the left above is a depiction of a Delaunay
+  \caption{ On the left above is a depiction of a Delaunay
     triangulation over four points, notice that the circumball (shaded
@@
 \end{figure}
-}{}
 
@@
 \cite{gordon1978shepard,shepard1968two}. The interpolant has the form
- $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
-     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k -
-  \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x -
-  x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x \neq x^{(k)}$,}{} and $r_k
-\in \mathbb{R}$ is the smallest radius such that at least $d+1$ other
-points $x^{(j)}$, $j \not = k$, are inside the closed Euclidean ball
-of radius $r_k$ about $x^{(k)}$. The computational complexity of
-ShepMod is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95
-implementation of ShepMod derived from SHEPPACK
-\cite{thacker2010algorithm}.
-
-\changed{
+$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)} {\sum\limits_{k=1}^{n}W_k(x)} ,$$ 
+where $p\bigl(x^{(k)}\bigr) = f\bigl(x^{(k)}\bigr)$, $W_k(x) =
+\bigl(\bigl(r_k - \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k
+\bigl\|x - x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x \neq x^{(k)}$, and
+$r_k \in \mathbb{R}$ is the smallest radius such that at least $d+1$
+other points $x^{(j)}$, $j \not = k$, are inside the closed Euclidean
+ball of radius $r_k$ about $x^{(k)}$. The interpolant $p(x)$ is
+continuous because the singularities at the $x^{(k)}$ are
+removable. The weight function is visualized in Figure \ref{fig:shepard}.
+The computational complexity of ShepMod is $\mathcal{O}(n^2d)$. This
+paper uses a Fortran 95 implementation of ShepMod derived from
+SHEPPACK \cite{thacker2010algorithm}.
+
 \begin{figure}
   \centering
-  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
-  \caption{\color{red} On the left above is a depiction of the radius
+  \includegraphics[width=0.4\textwidth,]{example_shepard_radius.pdf}
+  \hspace{.2cm}
+  \includegraphics[width=0.4\textwidth,]{example_shepard_weight.pdf}
+  \caption{On the left above is a depiction of the radius
     of influence for three chosen points of a collection in two
-    dimensions using the modified Shepard criteria. On the right is
-    the shape of the weight function $W_k$ for all data points, where
-    $W_k(x) = 0$ when $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen
-    that in the limit as $W_k$ tends towards infinity at $x_k$, the
-    relative weight of $x_k$ tends to one.
+    dimensions using the modified Shepard criteria. On the right a
+    third axis shows the relative weight for the center most
+    interpolation point $x^{(i)}$ with the solid line representing its
+    radius of influence, where $W_i(x)$ is $0$ for $\|x-x^{(i)}\|_2
+    \geq r_i$ and $W_i(x) / \sum_{k=1}^n W_k(x) \to 1$ as $x \to
+    x^{(i)}$.
   \vspace{-.1cm}}
@@
  {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is \changed{the same as for the modified Shepard
-  method}{a locally supported weighting function} and $P_k(x)$ is a
-local linear approximation to the data satisfying
+where $W_k(x)$ is the same as for the modified Shepard method and
+$P_k(x)$ is a local linear approximation to the data satisfying
 $P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
@@
   \includegraphics[width=0.6\textwidth,]{VM_Construction.pdf}
-  \caption{\color{red} Above is a depiction of the Voronoi cell
+  \caption{Above is a depiction of the Voronoi cell
     boundaries (dashed lines) about a set of interpolation points
@@
 $x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
-the nearest neighbor approximation by utilizing the Voronoi cells to
+the nearest neighbor approximation by utilizing the Voronoi cells
+(visualized in Figure \ref{fig:voronoi-basis}) to
 define support via a generic basis function $v: \mathbb{R}^d
@@
   \caption{In this HPC I/O example, the general methodology for
-    predicting a CDF and evaluating error can be seen\color{red},
-    where M means $\times 10^6$\color{black}. The Delaunay method
-    chose three source distributions (dotted lines) and assigned
-    weights \color{red}\{.1, .3, .6\} \color{black} (top to bottom at
-    \color{red}middle\color{black}). The weighted sum of the three
-    known CDFs produces the predicted CDF (dashed line). The KS
-    Statistic (\color{red}vertical line\color{black}) computed between
-    the true CDF (solid line) and predicted CDF (dashed line) is 0.2
-    for this example. For this example the KS test null hypothesis is
-    rejected at $p$-value 0.01, however it is not rejected at
-    $p$-value 0.001.
+    predicting a CDF and evaluating error can be seen, where M means
+    $\times 10^6$. The Delaunay method chose three source
+    distributions (dotted lines) and assigned weights \{.1, .3, .6\}
+    (top to bottom at middle). The weighted sum of the three known
+    CDFs produces the predicted CDF (dashed line). The KS Statistic
+    (vertical line) computed between the true CDF (solid line) and
+    predicted CDF (dashed line) is 0.2 for this example. For this
+    example the KS test null hypothesis is rejected at $p$-value
+    threshold 0.01, however it is not rejected at $p$-value threshold
+    0.001.
   \vspace{-.1cm}}
@@
 PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
-\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. \changed{Instead}{However}, for a CDF
+\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. Instead, for a CDF
 $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
@@
 The two-sample KS test is a useful nonparametric test for comparing
-two \changed{empirical}{} CDFs while only assuming stationarity,
+two empirical CDFs while only assuming stationarity,
 finite mean, and finite variance. The null hypothesis (that two
-\changed{empirical}{} CDFs come from the same underlying distribution)
+empirical CDFs come from the same underlying distribution)
 is rejected at level $p \in [0,1]$ when
- $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)}
-\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$ with distribution sample
-sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test
-presented in this work $n_1 = n_2$. An example of \changed{the process
-  of generating predicted distributions from known distributions and
-  the subsequent}{the round-trip prediction methodology from known and
-  predicted distributions to the} calculation of error can be seen in
-Figure \ref{fig:prediction-example}. A brief listing of relevant
-statistical terms used throughout this work is provided in the
-Appendix (Section \ref{sec:appendix}).
+$$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
+with distribution sample sizes $n_1,n_2 \in \mathbb{N}$
+(\textbf{not} the number of approximation points). For all
+applications of the KS test presented in this work $n_1 = n_2$. An
+example of the process of generating predicted distributions from
+known distributions and the subsequent calculation of error can be
+seen in Figure \ref{fig:prediction-example}. In this example the KS
+test null hypothesis is rejected at $p$-value threshold 0.01, however it is not
+rejected at $p$-value threshold 0.001. A brief listing of relevant statistical
+terms used throughout this work is provided in the Appendix (Section
+\ref{sec:appendix}).
 
@@
   \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
-  = \gamma \|x_j-x_0\|_2^2$. \changed{The following is visualized in 
-  Figure \ref{fig:lemma3}.}{}
+  = \gamma \|x_j-x_0\|_2^2$. The following is visualized in 
+  Figure \ref{fig:lemma3}.
 
@@
   since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
-  and so $g'(0) - \g z \le g'(z) = 0$. Now \changed{by integrating}{}
+  and so $g'(0) - \g z \le g'(z) = 0$. Now by integrating
   $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
   (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
-  \changed{acknowledging $g'(0) > \g / 2$}{}
+  using $g'(0) > \g / 2$
   \begin{align*}
@@
   \def\g{\gamma_g}
-  \caption{\color{red} Three different scenarios visualizing {\it
-      Lemma \ref{lemma:3}}, where $g(t)$ is the difference between a
+  \caption{ Three different scenarios visualizing {\it Lemma
+      \ref{lemma:3}}, where $g(t)$ is the difference between a
     piecewise linear interpolant and the approximated function along a
@@
 data only when the diameter of the simplex is also reduced at a
-proportional rate. Only linear convergence to the true function
-\changed{is}{can be} achieved in practice\changed{}{,} without the
-incorporation of additional observations\changed{, by moving closer to
-  interpolation points}{}. Notice that the approximation error is
-largely determined by the spacing of observed data. Predictions made
-by simplices whose vertices are not well-spaced (i.e., have large
-diameter, or are nearly contained in a hyperplane) have higher error.
-
-That the theoretical error bound is sharp can be observed with the
+proportional rate. Only linear convergence to the true function is
+achieved in practice without the incorporation of additional
+observations, by moving closer to interpolation points. Notice that
+the approximation error is largely determined by the spacing of
+observed data. Predictions made by simplices whose vertices are not
+well-spaced (i.e., have large diameter, or are nearly contained in a
+hyperplane) have higher error.
+
+The fact that the theoretical error bound is sharp can be observed with the
 test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d,$ and the
@@
 
+%% ===================================================================
+\vspace{-2mm}
+\subsection{Demonstration on an Analytic Test Function}
+\label{sec:analytic}
+
 \begin{figure}
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.9\textwidth,]{Fig2.pdf}};
+    \node (img)  {\includegraphics[width=0.9\textwidth,]{convergence-2d.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
@@
 
-%% ===================================================================
-\vspace{-2mm}
-\subsection{Demonstration on an Analytic Test Function}
-\label{sec:analytic}
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.9\textwidth,]{convergence-20d.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
+  \end{tikzpicture}
+  \caption{Delaunay and MLP approximations are constructed from Fekete
+    points over the unit cube evaluating the test function $f(x) =
+    \cos(\|x\|_2)$ for $x \in \mathbb{R}^{20}$. The details are the
+    same as for Fig. \ref{fig:convergence-2d}.}
+  \label{fig:convergence-20d}
+\end{figure}
 
@@
 piecewise linear regressor (the MLP) when provided an increasing
-amount of data with varying levels of random noise. \changed{Only
-  Delaunay and the MLP are considered in this demonstration because
-  all other mentioned techniques are not strictly piecewise linear
-  approximations.}{}
+amount of data with varying levels of random noise. Only Delaunay and
+the MLP are considered in this demonstration because all other
+mentioned techniques are not strictly piecewise linear approximations.
 
@@
 $6,$ chosen points are kept well spaced by computing approximate
-Fekete points. \changed{The following experiment was also run with a
+Fekete points. The following experiment was also run with a
   Latin hypercube design \cite{park1994optimal}, which produced almost
-  identical results that are not reported here.}{} Fekete points have
+  identical results that are not reported here. Fekete points have
 a history in potential theory \cite{kovari_pommerenke_1968} and are
@@
 each test two experiments are executed, one with exact function
-evaluations (\changed{an interpolation problem}{}) and one with a
-constant signal-to-noise ratio (SNR) of 10:1 (\changed{a regression
-  problem}{}).
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.9\textwidth,]{Fig3.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
-  \end{tikzpicture}
-  \caption{Delaunay and MLP approximations are constructed from Fekete
-    points over the unit cube evaluating the test function $f(x) =
-    \cos(\|x\|_2)$ for $x \in \mathbb{R}^{20}$. The details are the
-    same as for Fig. \ref{fig:convergence-2d}.}
-  \label{fig:convergence-20d}
-\end{figure}
-
-The bound from the theorem suggests that \changed{by}{exponentially}
-increasing the number of \changed{well-spaced}{} data points,
-\changed{approximation error can be reliably decreased}{results in a
-  decreasing error}. The $d = 2$ test seen on the left half of Figure
-\ref{fig:convergence-2d} shows a consistent decrease in error for
-Delaunay and also shows the eventual accuracy plateau obtained by a
-parametric regression form (the MLP, at roughly 500 points). On the
-right hand side of Figure \ref{fig:convergence-2d} the random noise
-clearly prohibits Delaunay from converging to $f$, while the MLP is
-able to improve its approximation with more data points on average.
-The convergence result for a very low-dimensional problem like this is
-expected.  However, intuition fails for higher dimensional problems.
-
-Figure \ref{fig:convergence-20d} shows the test function with $d = 20$
-presents a significantly more challenging approximation problem than
-its counterpart in low dimension. The same increase in number of data
-points from 32 to 8192 causes no apparent improvement in approximation
-for either the noise-free or the noisy problems. Perhaps unexpectedly,
-the interpolation technique performs better than the regression
-technique on the noisy data (right) in Figure
-\ref{fig:convergence-20d}, and worse than the regression technique on
-the noise-free data (left). This result emphasizes the relevance of
-interpolation for problems in high dimension. It also reveals that the
-outcome of MLP regressions can get worse when adding more data points.
-
-\changed{
+evaluations (an interpolation problem) and one with a constant
+signal-to-noise ratio (SNR) of 10:1 (a regression problem).
+
 \begin{figure}
@@
   \end{tikzpicture}
-  \caption{\color{red}The distribution of absolute error, distance to
+  \caption{The distribution of absolute error, distance to
     the nearest data point, smallest singular value (SV) and the
@@
 \end{figure}
-}{}
-
-\changed{The results achieved for both $d = 2$ and $d = 20$ align with
-  the theoretical error bound as can be seen in Figure
+
+The bound from the theorem suggests that by increasing the number of
+well-spaced data points, approximation error can be reliably
+decreased. The $d = 2$ test seen on the left half of Figure
+\ref{fig:convergence-2d} shows a consistent decrease in error for
+Delaunay and also shows the eventual accuracy plateau obtained by a
+parametric regression form (the MLP, at roughly 500 points). On the
+right hand side of Figure \ref{fig:convergence-2d} the random noise
+clearly prohibits Delaunay from converging to $f$, while the MLP is
+able to improve its approximation with more data points on average.
+The convergence result for a very low-dimensional problem like this is
+expected.  However, intuition fails for higher dimensional problems.
+
+Figure \ref{fig:convergence-20d} shows the test function with $d = 20$
+presents a significantly more challenging approximation problem than
+its counterpart in low dimension. The same increase in number of data
+points from 32 to 8192 causes no apparent improvement in approximation
+for either the noise-free or the noisy problems. Perhaps unexpectedly,
+the interpolation technique performs better than the regression
+technique on the noisy data (right) in Figure
+\ref{fig:convergence-20d}, and worse than the regression technique on
+the noise-free data (left). This result emphasizes the relevance of
+interpolation for problems in high dimension. It also reveals that the
+outcome of MLP regressions can get worse when adding more data points.
+
+The results achieved for both $d = 2$ and $d = 20$ align with the
+theoretical error bound as can be seen in Figure
 \ref{fig:data-spacing}. In low dimension, thousands of data points
@@
 reduction in data sparsity while simultaneously being more difficult
-  to produce well-\color{red}conditioned simplices. This evidences the inherent
-  challenge of data sparsity in high dimension approximation
-  problems. The analytic results presented are not caused by the
-  chosen test function, but rather the exponential increase in
-  complexity that accompanies increased dimension.}{}
-
-In summary, the regime of \changed{signal-to-noise ratios that 
-  result in competitively accurate interpolants}{} is \changed{}{much}
-greater for moderate to high dimensional problems \changed{due to
-  increased sparsity}{, even for functions with random noise
-  incorporated into their evaluation}. Acknowledging the viability of
-interpolation for problems of moderate dimension, the next section
-will consider real-world problems of similar proportion (thousands of
-examples in tens of dimensions).
+to produce well-conditioned simplices. This evidences the
+inherent challenge of data sparsity in high dimension approximation
+problems. The analytic results presented are not caused by the chosen
+test function, but rather the exponential increase in complexity that
+accompanies increased dimension.
+
+In summary, the regime of signal-to-noise ratios that result in
+competitively accurate interpolants is greater for moderate to high
+dimensional problems due to increased sparsity. Acknowledging the
+viability of interpolation for problems of moderate dimension, the
+next section will consider real-world problems of similar proportion
+(thousands of examples in tens of dimensions).
 
@@
 approximations and compare the accuracy of different techniques.
+The main goal is to demonstrate the viability of interpolants
+for real-world approximation problems and hence only real-world data
+sets (none that are synthetic) are utilized in this comparison.
 
@@
 In addition to the figures displaying approximation results for each
-data set, tables of accompanying numerical results \changed{including
-  timings and exact prediction errors}{} are located in the Appendix
-(Section \ref{sec:appendix}). All of the test data sets capture
-underlying functions that are almost certainly stochastic. As
-described in Section \ref{sec:introduction}, regression techniques
-appear most appropriate for these problems. However, typically data
-grows exponentially more sparse with increasing dimension. Given that
-sparse data regressors tend towards interpolation and, as demonstrated
-in Section \ref{sec:analytic}, interpolants produce similar (if not
+data set, scatter plots of predicted versus actual values and tables
+of accompanying numerical results including timings and quartiles of
+prediction error are located in the Appendix (Section
+\ref{sec:appendix}). All of the test data sets capture underlying
+functions that are almost certainly stochastic. As described in
+Section \ref{sec:introduction}, regression techniques appear most
+appropriate for these problems. However, typically data grows
+exponentially more sparse with increasing dimension. Given that sparse
+data regressors tend towards interpolation and, as demonstrated in
+Section \ref{sec:analytic}, interpolants produce similar (if not
 identical) results, it is presumed that interpolants are equally
@@
   \label{fig:hist-throughput}
-\end{figure}
-
-\begin{figure}
-  \centering
+
+%% \end{figure}
+%% \begin{figure}
+  %% \centering
+\vspace{4mm}
+
   \begin{tikzpicture}
@@
 
+Table \ref{table:avg-performance} summarizes results across the four
+test data sets with real-valued ranges. The interpolants discussed in
+this paper produce the \textit{best} approximations roughly one third
+of the time, and produce competitive approximations for almost all
+data sets. These test problems are almost certainly
+\textit{stochastic} in nature, but the high dimension leads to greater
+data sparsity and model construction cost, making interpolation more
+competitive.
+
+The major advantages to interpolation lie in the near absence of
+\textit{fit} time. Delaunay, LSHEP, and ShepMod all require pairwise
+distance calculations, for numerical robustness (Delaunay) and to
+determine the radii of influence for data points (LSHEP and
+ShepMod). At least hundreds, and sometimes hundreds of thousands of
+predictions can be made by the interpolants before the most widely
+used regressor (MLP) finishes fitting these relatively small data
+sets.  However, the computational complexities of all interpolants
+presented are greater than linear in either dimension or number of
+points, whereas the regressors' nonlinear complexity in dimension
+generally comes from the model fitting optimization.
+
 \begin{table}
@@
     MLP         & $\mathbf{43}$&$\mathbf{1}$ & $200$&$0$s                 & $0$&$001$s\\
+    \hline
     Delaunay    & $5$&$2$                    & $1$&$0$s                   & $1$&$0$s\\
@@
 
-Table \ref{table:avg-performance} summarizes results across the four
-test data sets with real-valued ranges. The interpolants discussed in
-this paper produce the \textit{best} approximations roughly one third
-of the time, and produce competitive approximations for almost all
-data sets. These test problems are almost certainly
-\textit{stochastic} in nature, but the high dimension leads to greater
-data sparsity and model construction cost, making interpolation more
-competitive.
-
-The major advantages to interpolation lie in the near absence of
-\textit{fit} time. Delaunay, LSHEP, and ShepMod all require pairwise
-distance calculations, for numerical robustness (Delaunay) and to
-determine the radii of influence for data points (LSHEP and
-ShepMod). At least hundreds, and sometimes hundreds of thousands of
-predictions can be made by the interpolants before the most widely
-used regressor (MLP) finishes fitting these relatively small data
-sets.  However, the computational complexities of all interpolants
-presented are greater than linear in either dimension or number of
-points, whereas the regressors' nonlinear complexity in dimension
-generally comes from the model fitting optimization.
-
 The new theoretical results presented in Section \ref{sec:theory}
@@
 \bibliographystyle{spmpsci}      % mathematics and physical sciences
-\bibliography{paper}
+\bibliography{tchlux-NA-interpolation}
 
@@
 statistic) that the expected value of the observed statistic is equal
-to an assumed population statistic. The $p$ value is the probability
-of observing a given statistic if the \textit{null hypothesis} is
-true. \changed{For a more detailed introduction to statistics and 
-related terminology, see the work of Navidi \cite{navidi_2015}.}{}
+to an assumed population statistic. The $p$-value of a given statistic
+value $\rho$ for a given data set (sample from a distribution) is the
+probability of observing a statistic at least as extreme as $\rho$ for
+other data sets (samples from that same distribution), assuming the
+null hypothesis holds. The smaller the $p$-value, the stronger the
+statistical evidence is for rejecting the null hypothesis. For a more
+detailed introduction to statistics and related terminology, see the
+work of Navidi \cite{navidi_2015}.
 
@@
 
+\cbend{}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.45\textwidth,]{scatter_forestfires.pdf}
+  \hspace{.6cm}
+  \includegraphics[width=0.45\textwidth,]{scatter_parkinsons.pdf}
+  \vspace{.6cm}
+
+  \includegraphics[width=0.45\textwidth,]{scatter_weather.pdf}
+  \hspace{.6cm}
+  \includegraphics[width=0.45\textwidth,]{scatter_creditcard.pdf}
+  \vspace{.3cm}
+
+  \caption{Scatter plots for predicted versus actual
+    values for the top three models on each of the four real valued
+    approximation problems. Top left is forest fire data, top right is
+    Parkinson's data, bottom left is rainfall data, and bottom right
+    is credit card transaction data. Each approximation algorithm has 
+    a unique style and the top three algorithms are listed in order of 
+    ranking in the legends. There are a large of number of 0-valued 
+    entries in the forest fire and rainfall data sets that are not
+    included in the visuals making the true ranking of the models
+    appear to disagree with the observed outcomes.}
+  \label{fig:scatter_plots}
+\end{figure}
+\cbend{}
+
+\vspace*{\fill}
+\newpage
+
 \begin{table}[H]
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}       & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}       & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
@@
     MLP       & $0$&$0426$                    & $2$&$63$                      & $5$&$27$                      & $14$&$0$                      & $1090$&$0$\\
+    \hline
     Delaunay  & $\mathbf{0}$&$\mathbf{0}$     & $1$&$98$                      & $5$&$37$                      & $13$&$1$                      & $\mathit{1080}$&$\mathit{0}$\\
@@
     is italicized. All values are rounded to three significant
-    digits.}
+    digits.\vspace{.2cm}}
   \label{table:error-forest-fire}
-\end{table}
-
-\begin{table}[H]
-  \centering
+
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
     MLP       & $0$&$0$                        &  & $32$&$8$                                & $0$&$000871$                       & $0$&$0436$\\
+    \cline{1-3}\cline{5-10}
     Delaunay  & $0$&$2$                        &  & $0$&$0151$                              & $0$&$0234$                         & $1$&$18$\\
@@
     time of 454 points, median time for one approximation, and median
-    time approximating 50 points.}
+    time approximating 50 points.\vspace{-1cm}}
   \label{table:best-forest-fire}
@@
 
-\vspace*{\fill}
-\newpage
-
-\begin{table}
+\begin{table}[H]
   \centering
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}                   & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}                   & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
@@
     MLP       & $0$&$0000239$                             & $\mathit{0}$&$\mathit{533}$   & $\mathit{1}$&$\mathit{25}$    & $\mathbf{2}$&$\mathbf{84}$    & $39$&$3$\\
+    \hline
     Delaunay  & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$2$                       & $3$&$5$                       & $7$&$67$                      & $30$&$7$\\
@@
     is italicized. All values are rounded to three significant
-    digits.}
+    digits.\vspace{.2cm}}
   \label{table:error-parkinsons}
-\end{table}
-
-\begin{table}
-  \centering
+
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
     MLP       & $\mathit{32}$&$\mathit{0}$   &  & $348$&$0$                               & $0$&$00111$                        & $0$&$653$\\
+    \cline{1-3}\cline{5-10}
     Delaunay  & $0$&$0$                      &  & $2$&$47$                                & $1$&$22$                           & $720$&$0$\\
@@
     time of 5288 points, median time for one approximation, and median
-    time approximating 587 points.}
+    time approximating 587 points.\vspace{-1cm}}
   \label{table:best-parkinsons}
@@
 
-\begin{table}
+\begin{table}[H]
   \centering
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$}              & \multicolumn{2}{c}{$50^{th}$}  & \multicolumn{2}{c}{$75^{th}$}  & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$}              & \multicolumn{2}{c}{$50^{th}$}  & \multicolumn{2}{c}{$75^{th}$}  & \multicolumn{2}{c}{Max}\\
     \hline
@@
     MLP       & $0$&$0000689$                              & $0$&$0337$                                 & $\mathbf{0}$&$\mathbf{0795}$   & $\mathbf{0}$&$\mathbf{264}$    & $\mathbf{5}$&$\mathbf{31}$\\
+    \hline
     Delaunay  & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$187$                                  & $0$&$919$                      & $2$&$72$                       & $56$&$3$\\
@@
     is italicized. All values are rounded to three significant
-    digits.}
+    digits.\vspace{.2cm}}
   \label{table:error-weather}
-\end{table}
-
-\begin{table}
-  \centering
+
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
     MLP       & $\mathbf{60}$&$\mathbf{9}$   &  & $169$&$0$                                  & $0$&$00137$                        & $0$&$356$\\
+    \cline{1-3}\cline{5-10}
     Delaunay  & $0$&$1$                      &  & $0$&$664$                                  & $0$&$886$                          & $230$&$0$\\
@@
     \ref{table:error-weather}. On the right columns are median fit
-    time of 2349 points, median time for one approximation, and median
-    time approximating 260 points.}
+p    time of 2349 points, median time for one approximation, and median
+    time approximating 260 points.\vspace{-1cm}}
   \label{table:best-weather}
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
@@
     MLP       & $0$&$00151$                                & $\mathbf{1}$&$\mathbf{6}$     & $\mathbf{3}$&$\mathbf{86}$    & $\mathbf{8}$&$\mathbf{34}$    & $\mathbf{604}$&$\mathbf{0}$\\
+    \hline
     Delaunay  & $\mathbf{0}$&$\mathbf{0}$                  & $\mathit{2}$&$\mathit{69}$    & $13$&$8$                      & $\mathit{35}$&$\mathit{0}$    & $\mathit{4840}$&$\mathit{0}$\\
@@
     is italicized. All values are rounded to three significant
-    digits.}
+    digits.\vspace{.2cm}}
   \label{table:error-credit-card}
-\end{table}
-
-\begin{table}
-  \centering
+
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
     MLP       & $\mathbf{79}$&$\mathbf{5}$   &  & $290$&$0$                               & $0$&$000714$                       & $0$&$397$\\
+    \cline{1-3}\cline{5-10}
     Delaunay  & $\mathit{20}$&$\mathit{5}$   &  & $3$&$12$                                & $3$&$71$                           & $2070$&$0$\\
@@
     median fit time of 5006 points, median time for one approximation,
-    and median time approximating 556 points.}
+    and median time approximating 556 points.\vspace{-1cm}}
   \label{table:best-credit-card}
@@
 
-\begin{table}
+\vspace*{\fill}
+
+\begin{table}[H]
   \centering
@@
     \hline
-    Algorithm & \multicolumn{2}{c}{Min}      & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}      & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
@@
     while the second lowest is italicized. All values are rounded to
-    three significant digits.}
+    three significant digits.\vspace{.2cm}}
   \label{table:error-throughput}
-\end{table}
-
-\begin{table}
-  \centering
+
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
 
+\vspace*{\fill}
+
 \end{appendix}

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Predictions/visual_make_scatter.py

@@ -1,0 +1,119 @@
+
+import os, sys
+
+old_path = sys.path
+sys.path = [os.path.abspath(os.curdir)] + old_path
+from util.data import Data
+from util.plot import Plot
+
+results = ["results_iozone_150.dill",       
+           "results_forestfires.dill",      
+           "results_parkinsons_updrs.dill", 
+           "results_weatherAUS.dill",       
+           "results_creditcard.dill",       
+]
+
+targets = {
+    "forestfires":("area", ["SVR", "BoxMesh", "MARS"], True),
+    "parkinsons":("total_UPDRS", ["ShepMod", "NeuralNetwork", "Voronoi"], False),
+    "weather":("Rainfall Tomorrow", ["NeuralNetwork", "LSHEP", "MARS"], True),
+    "creditcard":("Amount", ["NeuralNetwork", "Delaunay", "ShepMod"], True),
+}
+
+
+legend = dict(
+    xanchor = "center",
+    yanchor = "top",
+    x = .5,
+    y = .95,
+    orientation = "h",
+    bgcolor="white",
+    bordercolor="grey",
+    borderwidth=.5,
+    traceorder='reversed'
+)
+ms = 5 # Marker size
+
+all_symbols = ["circle", "diamond", "cross", "x",
+               "star", "pentagon", "diamond-wide",
+               "star-triangle-down"]
+all_colors = [(126,153,210), (110,200,109), (210,105,109),
+              (200,153,200), (255,180,40), (110,200,200),
+              (150,150,180), (255,180,40)]
+
+all_names = set(sum((v[1] for v in targets.values()),[]))
+all_names = sorted(all_names, key=lambda n: -sum(n in v[1] for v in targets.values()))
+all_names = [n.replace("NeuralNetwork","MLP")\
+              .replace("BoxMesh", "BoxSpline") for n in all_names]
+
+all_symbols = {n:s for (n,s) in zip(all_names,all_symbols)}
+all_colors = {n:c for (n,c) in zip(all_names,all_colors)}
+all_colors["Voronoi"], all_colors["LSHEP"] = all_colors["LSHEP"], all_colors["Voronoi"]
+all_symbols["SVR"], all_symbols["LSHEP"] = all_symbols["LSHEP"], all_symbols["SVR"]
+print("all_names: ",all_names)
+print("all_symbols: ",all_symbols)
+print("all_colors: ",all_colors)
+
+
+for r_path in results:
+    for t in targets:
[…]

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Revision/fix_old_data/fix.py

@@ -1,41 +1,0 @@
-import os, dill
-
-# Update the system path to only point to the local directory (no imports from elsewhere).
-import sys
-to_remove = [p for p in sys.path if "fix_old_data" in p]
-if len(to_remove) > 0:
-    for p in to_remove: sys.path.pop(sys.path.index(p))
-# Import the master util Data object.
-from util.data import Data
-print()
-print(sys.path)
-print(Data)
-# Now delete the already-imported module so it doesn't interfere with ".util"
-_ = sys.modules.pop("util")
-del _
-_ = sys.modules.pop("util.data")
-del _
-# Now reset the path so the local util is used for the import.
-sys.path = to_remove[:1] + sys.path
-
-
-# Update the 'fit-final.dill' file.
-f = "fit-final.dill"
-with open(f, "rb")  as raw: d = dill.load(raw)
-nd = Data(d, names=d.names, types=d.types)
-nd.save("old-" + f)
-
-# Update all results files.
-for f in sorted(os.listdir()):
-    if "results_" != f[:len("results_")]: continue
-    print()
-    print(f)
-    with open(f, "rb") as raw: d = dill.load(raw)
-    print()
-    print(d.names)
-    print(d.types)
-    print()
-    nd = Data(d, names=d.names, types=d.types)
-    nd.save("old-" + f)
-    print("Updated..")
-

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Revision/fix_old_data/new_data/data.py

@@ -1,1856 +1,0 @@
-import os, sys
-from .read import read_data
-from .categories import regular_simplex
-from . import DEFAULT_DISPLAY_WAIT, GENERATOR_TYPE, NP_TYPES, \
-    SEPARATORS, FILE_SAMPLE_SIZE, MISSING_SAMPLE_SIZE
-
-# TODO:  Make access of a single row (in slicing) return a Row object.
-# TODO:  Make "save" and "load" to CSV support having iterables of
-#        readable types. Expand them out to columns with particular
-#        names (for compatibility with other software), read those
-#        particular names back into a stack.
-# TODO:  Make "to_matrix" automatically flatten elements of columns
-#        that contain iterables into their base types. If the types
-#        are numeric, use float, otherwise use string. Use "is numeric".
-# TODO:  Make "to_matrix" automatically consider 'None' a unique value
-#        for categorical columns and add a dimension 'is present' otherwise.
-# TODO:  Wrap the necessary "Numeric" attributes into the matrix,
-#        using a custom ndarray object from numpy.
-# TODO:  Rewrite "predict" to work given a data object. This will
-#        break up the data into collections of unique missing values.
-#        Then for each collection, build a model over the appropriate
-#        matrix-format data and predict that collection. This will
-#        work more nicely with the k-fold method.
-# TODO:  Remove the "fill" method in favor of "predict".
-# TODO:  Make "predict" not require that target columns be convertible
-#        to numeric form, only that they operate under a weighted sum.
-# TODO:  Fix "type digression" message to say specifically what column
-#        changed types and show the value that triggered the change.
-# TODO:  Implement "parallel read" that create subprocesses that read
-#        chunks of the file, communicate progress, communicate type
-#        digressions, saves chunks in .pkl, then organizes them all
-#        into the expected ordered data format.
-# TODO:  Make printout shorter for data with lots of columns.
-# TODO:  Implmenent 'IndexedAudio' reader that mimics the
-#        'IndexedVideo' reader and allows for access to peaks.
-# TODO:  Implemenet 'IndexedText' reader that mimics the
-#        'IndexedVideo' reader and allows for access to characters.
-# TODO:  Implement 'IndexedCSV' reader that reads rows out of a very
-#        large CSV file.
-# TODO:  Implement 'LazyData' whose underlying '.data' object is a
-#        lazy-evaluated 'IndexedCSV' object.
-# TODO:  Implement 'CachedData' whose underlying '.data' object caches
-#        groups of rows to keep the memory footprint low.
-
-
[…]

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Revision/fix_old_data/new_data/test.py

@@ -1,524 +1,0 @@
-from util.data.data import Data
-
-# Some tests for the data
-def test_data():
-    import os
-    import numpy as np
-
-    print("Testing Data...", end=" ")
-
-    # ----------------------------------------------------------------
-    a = Data()
-
-    # Verify append
-    a.append([1,"a"])
-    a.append([2,"b"])
-    a.append([3,"c"])
-
-    # Verify add column (with edge case, none type)
-    a.add_column([None,None,None])
-    assert(a.types[-1] == type(None))
-
-    # Reassign the missing value column to floats, verify type update
-    a[a.names[-1]] = [1.2, 3.0, 2.4]
-    assert(a.types[-1] == float)
-    # WARNING: Need to check doing (<int>, <str>) assignment
-
-    # Verify automatic type generation AND automatic type-casting
-    a.append([1,2,3])
-    assert(tuple(a[-1]) == tuple([1,"2",3.0]))
-    assert(tuple(map(type,a[-1])) == (int, str, float))
-
-    # Add some missing values.
-    a.append([-1,None,None])
-
-    #  Done modifying "a", rest of tests leave it constant.
-    # ----------------------------------------------------------------
-
-    # Verify that slicing works on descirptors.
-    assert(a.names[:-1] == ['0','1'])
-
-    # Verify that slicing works on descriptors of views.
-    b = a[:,1:]
-    assert(b.names[1:] == ['2'])
-
-    # Verify in-place addition of rows (different length data with same column names)
-    b = a[:-2].copy()
-    c = a[1:-2]
-    b += c
-    assert(tuple(b["0"]) == tuple([1,2,3,2,3]))
-
-    # Verify in-place addition of new columns (same length data with new columns)
-    b = a[:,:-1].copy()
-    c = a[:,-1:].copy()
-    c["3"] = range(len(c))
-    b += c
-    assert(tuple(b[-2]) == (1,"2",3.0,3))
-    assert(tuple(b[-1]) == (-1,None,None,4))
-
-    # Verify in-place addition of new columns (same length data with fewer columns)
-    b = a[:,:-1].copy()
-    c = a[:,:].copy()
-    c["3"] = range(len(c))
-    c += b
-    assert(tuple(c[-2]) == (1,"2",None,None))
[…]

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Revision_1/fix_old_data/new_data/auxiliary.py

@@ -1,0 +1,323 @@
+from .exceptions import BadValue, ImproperUsage
+
+# This class is a protected type of list where the contents are
+# supposed to be of a specific type. It enforces that assignment and
+# append operations do not violate the type specifiation of the object.
+class Descriptor:
+    # Initialize this descriptor, make sure all elements of 'values'
+    # have the same type if provided.
+    def __init__(self, data, values=None, types=type(None)):
+        # Store the parent data object of this descriptor.
+        self.data = data
+        # Store the provided data (or make a new set)
+        if (values is None): values = []
+        self.values = values
+        # Initialize the type.
+        self.type = types
+        for value in self.values:
+            if (self.type == type(None)): self.type = type(value)
+            elif (self.type != type(value)):
+                raise(BadValue(f"Only type '{self.type}' objects can be added to this descriptor."))
+
+    # Check for the type of the added value.
+    def append(self, value):
+        if (self.data.view): raise(ImproperUsage("Cannot modify descriptor of view."))
+        if (self.type == type(None)): self.type = type(value)
+        if (self.type != type(value)):
+            raise(BadValue(f"Only type '{self.type}' objects can be added to this descriptor."))
+        return self.values.append(value)
+
+    # Check for type in the "setitem" operation.
+    def __setitem__(self, index, value):
+        if (self.data.view): raise(ImproperUsage("Cannot modify descriptor of view."))
+        if (self.type == type(None)): self.type = type(value)
+        if (self.type != type(value)):
+            raise(BadValue(f"Only type '{self.type}' objects can be added to this descriptor."))
+        self.values[index] = value
+
+    # Return the appropriately mapped values through getitem.
+    def __getitem__(self, index):
+        # If this is a slice, then return a 
+        if (type(index) == slice):
+            return [self.values[self.data.col(i)] for i in range(self.data.shape[1])[index]]
+        # Use the standard getitem operator.
+        return self.values[self.data.col(index)]
+
+    # Define a "len" operator.
+    def __len__(self): return self.data.shape[1]
+
+    # Define a "str" operator.
+    def __str__(self): return str(list(self))
+
+    # Define an "insert" function.
+    def insert(self, idx, value):
+        if (self.data.view): raise(ImproperUsage("Cannot modify descriptor of view."))
+        if (self.type == type(None)): self.type = type(value)
+        if (self.type != type(value)):
+            raise(BadValue(f"Only type '{self.type}' objects can be added to this descriptor."))
[…]

### 2020-08-13 [2020-01]_IEEESE_(convex_optimization)/code/raw/2-summarize_data.py

@@ -1,135 +1,0 @@
-# Make sure the previous step has been executed.
-previous_step = __import__("1-reduce_data")
-data_file = previous_step.output_file
-
-import os
-from util.data import Data
-
-reduced_data = Data.load(data_file)
-reduced_data.sort()
-
-output_file = "2-data-summarized.csv"
-output_folder = "Summarized"
-
-if not os.path.exists(output_file):
-    summary_data = Data(names=["plot", "series", "path"],
-                        types=[str,    str,      str])
-
-    print()
-    print("Reduced data:",reduced_data)
-    print()
-
-    config_cols = ["dimension", "function", "skew", "rotation", "noise"]
-
-    # Considering each algorithm, average over all values by:
-    #   'function'
-    #   'skew'
-    #   'rotation'
-    #   'noise'
-
-    series_by_col = "algorithm"
-    plot_by_cols = ["algorithm", "function", "skew", "rotation", "noise"]
-
-    # Cycle each of the differennt plot columns.
-    for plot_by_col in plot_by_cols:
-        print()
-        print('-' * 70)
-        # Check for the existence of the output file.
-        if (plot_by_col == series_by_col):
-            data_file = os.path.join(output_folder, f"{series_by_col}-mean.csv")
-        else:
-            data_file = os.path.join(output_folder, f"{series_by_col}-mean-by-{plot_by_col}.csv")
-
-        # Store the output path.
-        summary_data.append([plot_by_col, series_by_col, data_file])
-        # Check to see if this output already exists, if so skip.
-        if os.path.exists(data_file):
-            print(f"skipping '{data_file}'..")
-            continue
-
-        # Identify the different plots that will be created.
-        print("plot_by_col: ",plot_by_col)
-        print()
-        plot_by_values = sorted(set(reduced_data[plot_by_col]))
-
-        # Figure out which configurations to keep (need to be same across all subsets).
-        config_cols_to_check = [c for c in config_cols if c not in {plot_by_col, series_by_col}]
-        configs_to_keep = reduced_data[config_cols_to_check].unique()
-        to_keep = set(range(len(configs_to_keep)))
-        for plot_by_value in plot_by_values:
-            # Get the data that would be used for this plot, and the configurations.
-            plot_data = reduced_data[ reduced_data[plot_by_col] == plot_by_value ]
-            plot_configs = plot_data[config_cols_to_check].unique()
-            # Remove all configurations that do not appear in this plot's data.
[…]

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-interpolation/Versions/1-paper.tex
Finished NUMA paper and submitted. Might need to do one more round with editors.

--- 2020-03-30 [2019-02]_NUMA_(interpolant_error_bound)/paper.tex
@@
      {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k - \bigl\|x -
-x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x - x^{(k)}\bigr\|_2 \bigr)\bigr)^2$}{}
-and $r_k \in \mathbb{R}$ is the smallest radius such that at least
-$d+1$ other points $x^{(j)}$, $j \not = k$, are inside the closed
-Euclidean ball of radius $r_k$ about $x^{(k)}$. The computational
-complexity of ShepMod is $\mathcal{O}(n^2d)$. This paper uses a
-Fortran 95 implementation of ShepMod derived from SHEPPACK
-\cite{thacker2010algorithm}.
-
-\changed{
-\begin{figure}
-  \centering
-  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
-  \caption{\color{red} On the left above is a depiction of the radius
-    of influence for three chosen points of a collection using the
-    modified Shepard criteria. On the right is the shape of the weight
-    function $W_k$ for all data points, where $W_k(x) = 0$ when
-    $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen that in the limit
-    as $W_k$ goes to infinity at $x_k$, the relative weight of $x_k$
-    goes to one.
-  \vspace{-.1cm}}
-  \label{fig:shepard}
-\end{figure}
-}{}
+where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
+\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
+smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
+\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
+$x^{(k)}$. The computational complexity of ShepMod is
+$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
+ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
 
@@
 
-The box spline model used here is \changed{an}{sparse grid}
-interpolation technique built from overlapping box splines
-\cite{de2013box}. The box splines serve as basis functions that are
-shifted and scaled to have support over box shaped regions. The boxes
-are constructed in a way to guarantee a covering of the domain
-\cite{lux2018novel}. Given a set of box splines
-$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
-in \cite{lux2018novel} and anchored at interior points
-$\bigl\{x^{(i)}\bigr\}$,
+The box spline model used here is a sparse grid interpolation
+technique built from overlapping box splines \cite{de2013box}. The box
+splines serve as basis functions that are shifted and scaled to have
+support over box shaped regions. The boxes are constructed in a way to
+guarantee a covering of the domain \cite{lux2018novel}. Given a set of
+box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box
+properties outlined in \cite{lux2018novel} and anchored at interior
+points $\bigl\{x^{(i)}\bigr\}$,
  $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
@@
 applications of the KS test presented in this work $n_1 = n_2$. An
-example of the \changed{end-to-end}{round-trip} prediction methodology from known and
+example of the round-trip prediction methodology from known and
 predicted distributions to the calculation of error can be seen in
@@
 
-%% \newpage
+\newpage
 
@@
 piecewise linear regressor (the MLP) when provided an increasing
-amount of data with varying levels of random noise. \changed{Only
-  Delaunay and the MLP are considered in this demonstration because
-  all other mentioned techniques are not strictly piecewise linear
-  approximations.}{}
+amount of data with varying levels of random noise.
 
@@
 each test two experiments are executed, one with exact function
-evaluations (an interpolation problem) and one with a constant
-signal-to-noise (SNR) ratio of 10:1 (a regression problem).
+evaluations (for interpolation) and one with a constant signal-to-noise
+(SNR) ratio of 10:1 (for regression).
 
@@
 The bound from the theorem suggests that exponentially increasing the
-number of \changed{well-spaced}{} data points should result in a
-decreasing error. The $d = 2$ test seen on the left half of Figure
-\ref{fig:convergence-2d} shows a consistent decrease in error for
-Delaunay and also shows the eventual accuracy plateau obtained by a
-parametric regression form (the MLP, at roughly 500 points). On the
-right hand side of Figure \ref{fig:convergence-2d} the random noise
-clearly prohibits Delaunay from converging to $f$, while the MLP is
-able to improve its approximation with more data points on average.
-The convergence result for a very low-dimensional problem like this is
-expected.  However, intuition fails for higher dimensional problems.
+number of data points should result in a decreasing error. The $d = 2$
+test seen on the left half of Figure \ref{fig:convergence-2d} shows a
+consistent decrease in error for Delaunay and also shows the eventual
+accuracy plateau obtained by a parametric regression form (the MLP, at
+roughly 500 points). On the right hand side of Figure
+\ref{fig:convergence-2d} the random noise clearly prohibits Delaunay
+from converging to $f$, while the MLP is able to improve its
+approximation with more data points on average.  The convergence
+result for a very low-dimensional problem like this is expected.
+However, intuition fails for higher dimensional problems.
 
@@
 
-\changed{
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.9\textwidth,]{theorem-convergence.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Euclidean Distance};
-  \end{tikzpicture}
-  \caption{\color{red}The distribution of absolute error, distance to
-    the nearest data point, smallest singular value and the longest
-    edge of the simplex containing each approximation point in the
-    tests from Fig. \ref{fig:convergence-2d} and
-    Fig. \ref{fig:convergence-20d} for Delaunay. In two dimensions it
-    can be seen that $\|z - x_0\|_2$, $\sigma_d$, and $k$ all shrink
-    at the same rate for well-spaced approximation points, resulting
-    in a faster rate of decrease for approximation error. Notice that
-    in higher dimension the data remains sparse even with thousands of
-    data points. This is seen by the relatively small reduction in
-    $k$, and this explains the slow reduction in error seen by
-    Delaunay in Fig. \ref{fig:convergence-20d}.}
-  \label{fig:data-spacing}
-\end{figure}
-}{}
-
-\changed{The results achieved for both the $d = 2$ and $d = 20$ align
-  with the theoretical error bound as can be seen in Figure
-  \ref{fig:data-spacing}. In low dimension, thousands of data points
-  meaningfully reduce the sparsity of the approximation problem.
-  However, in higher dimension it takes many more points to achieve a
-  reduction in data sparsity. The same result holds for a Latin
-  hypercube well-spaced point set, suggesting that sparsity is an
-  inherent problem in high dimension approximation problems. This
-  result is not a product of the chosen test function, but rather the
-  exponential increase in complexity that accompanies increased
-  dimension.}{}
-
-In summary, the regime of interpolation is \changed{}{much} greater
-for moderate to high dimensional problems \changed{due to increased
-  sparsity}{}, even for functions with random noise incorporated into
-their evaluation. Acknowledging the viability of interpolation for
-problems of moderate dimension, the next section will consider
-real-world problems of similar proportion (thousands of examples in
-tens of dimensions).
+In summary, the regime of interpolation is much greater for moderate
+to high dimensional problems, even for functions with random noise
+incorporated into their evaluation. Acknowledging the viability of
+interpolation for problems of moderate dimension, the next section
+will consider real-world problems of similar proportion (thousands of
+examples in tens of dimensions).
 
@@
 of observing a given statistic if the \textit{null hypothesis} is
-true. \changed{For a more detailed introduction to statistics and 
-related terminology, see the work of Navidi \cite{navidi_2015}.}{}
+true.
 

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-interpolation/Versions/2-paper.tex
Finished NUMA paper and submitted. Might need to do one more round with editors.

--- 2020-04-01 [2019-02]_NUMA_(interpolant_error_bound)/paper.tex
@@
 %% Include the response to reviewers at the beginning of the document.
-\include{tchlux-NA-response-2020-04}
+\include{tchlux-NA-response-2_2020-04}
 
@@
   to address the problem of dimensional scaling. The sparse grid point
-  sets combined with specific approximation functions are used to
-  construct total error bounds by considering projections onto
+  sets combined with specific approximation functions allow for the
+  construction of interpolation error bounds as well as bounds on
+  total error for regressors by considering projections onto
   orthogonal polynomials
   \cite{migliorati2013approximation,nobile2016convergence}. However,
-  sparse grids have exponential size in dimension making them
-  intractable for more than tens of dimensions and do not immediately
-  apply to the problem of scattered data interpolation.}{} This work
-extends some known results regarding the secant method
-\cite{dennis1996numerical} to construct \changed{a uniform error bound
-  for interpolants over arbitrary scattered data in any dimension}{an
-  interpolation error bound for problems of arbitrary
-  dimension}. These error bounds are useful, considering
-\changed{similar uniform bounds}{the same cannot} cannot be
-\changed{constructed}{said} for regression algorithms in general. The
-maximum complexity of an interpolant is bounded by the amount of data
-available, while the maximum complexity of a regressor is bounded by
-both the amount of data and the chosen parametric form. For this
-reason, generic uniform bounds are largely unobtainable for regression
-techniques on arbitrary approximation problems, even when the
-approximation domain is bounded.
+  sparse grids are not studied here because they have exponential size
+  in dimension making them intractable for more than tens of
+  dimensions and are primarily used for problems that allow data
+  points to be chosen.}{} This work extends some known results
+regarding the secant method \cite{dennis1996numerical} to construct
+\changed{a uniform error bound for interpolants over arbitrary
+  scattered data in any dimension}{an interpolation error bound for
+  problems of arbitrary dimension}. These error bounds
+\changed{evidence that interpolants}{} are useful, considering
+\changed{similar uniform bounds cannot be constructed}{the same cannot
+  be said} for \changed{regressors}{regression algorithms} in
+general. The maximum complexity of an interpolant is bounded by the
+amount of data available, while the maximum complexity of a regressor
+is bounded by both the amount of data and the chosen parametric
+form. For this reason, generic uniform bounds are largely unobtainable
+for regression techniques on arbitrary approximation problems, even
+when the approximation domain is bounded.
 
@@
 for training. \changed{It should be noted that the choice of neural
-  network architecture impacts approximation performance, but no
+  network architecture affects approximation performance, but no
   architecture tuning is performed here}{}.
@@
 producing an interpolant \cite{lee1980two}. The Delaunay triangulation
-of a set of data points into simplices is such that there are no data
-points inside the sphere defined by the vertices of each simplex. For
-a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
-$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
-$\ldots$, $d$, $y \in S$ is a unique convex combination of the
-vertices,
+of a set of \changed{}{data} points into simplices is such that there
+are no \changed{}{data} points inside the sphere defined by the
+vertices of each simplex. For a $d$-simplex S with vertices $x^{(0)}$,
+$x^{(1)}$, $\ldots$, $x^{(d)}$, and function values
+$f\bigl(x^{(i)}\bigr)$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique
+convex combination of the vertices,
  $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
@@
  $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
-     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k -
-  \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x -
-  x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x \neq x^{(k)}$,}{} and $r_k
-\in \mathbb{R}$ is the smallest radius such that at least $d+1$ other
-points $x^{(j)}$, $j \not = k$, are inside the closed Euclidean ball
-of radius $r_k$ about $x^{(k)}$. The computational complexity of
-ShepMod is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95
-implementation of ShepMod derived from SHEPPACK
-\cite{thacker2010algorithm}.
+     {\sum\limits_{k=1}^{n}W_k(x)} ,$$ where
+     \changed{$p\bigl(x^{(k)}\bigr) = f\bigl(x^{(k)}\bigr)$, $W_k(x) =
+       \bigl(\bigl(r_k - \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/
+       \bigl( r_k \bigl\|x - x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x
+       \neq x^{(k)}$,}{} and $r_k \in \mathbb{R}$ is the smallest
+     radius such that at least $d+1$ other points $x^{(j)}$, $j \not =
+     k$, are inside the closed Euclidean ball of radius $r_k$ about
+     $x^{(k)}$. \changed{The interpolant $p(x)$ is continuous because
+       the singularities at the $x^{(k)}$ are removable.}{} The
+     computational complexity of ShepMod is $\mathcal{O}(n^2d)$. This
+     paper uses a Fortran 95 implementation of ShepMod derived from
+     SHEPPACK \cite{thacker2010algorithm}.
 
@@
   \centering
-  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
+  \includegraphics[width=0.4\textwidth,]{example_shepard_radius.pdf}
+  \hspace{.2cm}
+  \includegraphics[width=0.4\textwidth,]{example_shepard_weight.pdf}
   \caption{\color{red} On the left above is a depiction of the radius
     of influence for three chosen points of a collection in two
-    dimensions using the modified Shepard criteria. On the right is
-    the shape of the weight function $W_k$ for all data points, where
-    $W_k(x) = 0$ when $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen
-    that in the limit as $W_k$ tends towards infinity at $x_k$, the
-    relative weight of $x_k$ tends to one.
+    dimensions using the modified Shepard criteria. On the right a
+    third axis shows the relative weight for the center most
+    interpolation point $x^{(i)}$ with the solid line representing its
+    radius of influence, where $W_i(x)$ is $0$ for $\|x-x^{(i)}\|_2
+    \geq r_i$ and $W_i(x) / \sum_{k=1}^n W_k(x) \to 1$ as $x \to
+    x^{(i)}$.
   \vspace{-.1cm}}
@@
   (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
-  \changed{acknowledging $g'(0) > \g / 2$}{}
+  \changed{using $g'(0) > \g / 2$}{}
   \begin{align*}
@@
 In addition to the figures displaying approximation results for each
-data set, tables of accompanying numerical results \changed{including
-  timings and exact prediction errors}{} are located in the Appendix
-(Section \ref{sec:appendix}). All of the test data sets capture
-underlying functions that are almost certainly stochastic. As
+data set, \changed{scatter plots of predicted versus actual values
+  and}{} tables of accompanying numerical results \changed{including
+  timings and quartiles of prediction error}{} are located in the
+Appendix (Section \ref{sec:appendix}). All of the test data sets
+capture underlying functions that are almost certainly stochastic. As
 described in Section \ref{sec:introduction}, regression techniques
@@
 
+\cbend{}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.45\textwidth,]{scatter_forestfires.pdf}
+  \hspace{.6cm}
+  \includegraphics[width=0.45\textwidth,]{scatter_parkinsons.pdf}
+  \vspace{.6cm}
+
+  \includegraphics[width=0.45\textwidth,]{scatter_weather.pdf}
+  \hspace{.6cm}
+  \includegraphics[width=0.45\textwidth,]{scatter_creditcard.pdf}
+  \vspace{.3cm}
+
+  \caption{\color{red} Scatter plots for predicted versus actual
+    values for the top three models on each of the four real valued
+    approximation problems. Top left is forest fire data, top right is
+    Parkinson's data, bottom left is rainfall data, and bottom right
+    is credit card transaction data. The blue circles correspond to
+    the best model, green diamonds to the second best, and red crosses
+    to the third best model for each data set. There are a large of
+    number of 0-valued entries in the forest fire and rainfall data
+    sets that are not included in the visuals making the true ranking
+    of the models appear to disagree with the observed outcomes.}
+  \label{fig:scatter_plots}
+\end{figure}
+\cbend{}
+
+\vspace*{\fill}
+\newpage
+
 \begin{table}[H]

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Predictions/util/data/data.py

@@ -1,0 +1,2109 @@
+import os, sys
+from util.data.read import read_data
+from util.data.categories import regular_simplex
+from util.data import DEFAULT_DISPLAY_WAIT, GENERATOR_TYPE, NP_TYPES, \
+    SEPARATORS, FILE_SAMPLE_SIZE, MISSING_SAMPLE_SIZE
+
+# TODO:  Instead of just checking file size, check number of columns too,
+#        use a sample by default if the #columns is large (speed!).
+# TODO:  Warnings or versioning for when the source data is modified
+#        in a way that invalidates a view. Need to track unpropagated
+#        changes somehow. Or make permanent changes disallowed? Or copy?
+#        E.g. view is made, columns are rearranged
+# TODO:  Make "to_matrix" automatically consider 'None' a unique value
+#        for categorical columns and add a dimension 'is present' otherwise.
+# TODO:  Make "to_matrix" automatically flatten elements of columns
+#        that contain iterables into their base types. If the types
+#        are numeric, use float, otherwise use string. Use "is numeric".
+# TODO:  Check for duplicate column names, handle duplicate column
+#        names? Do names *have* to be Python strings?
+# TODO:  Support in-place addition with *some* duplicate solumns and
+#        the same number of rows? That or block duplicate col names.
+# 
+# 
+# TODO:  Remove builtin prediction code for anything other than
+#        nearest neighbor prediction. Maybe remove prediction in
+#        general, since it isn't something that can be picked beforehand.
+# TODO:  Running 'predict' with nearest neighbor seg-faults after 4th
+#        round, might involve conditioning, probably involves ball tree.
+# TODO:  'predict' does not work with a view, picks wrong column, it
+#        also appears to pick the wrong column even when not a view.
+# TODO:  Rewrite "predict" to work given a data object. This will
+#        break up the data into collections of unique missing values.
+#        Then for each collection, build a model over the appropriate
+#        matrix-format data and predict that collection. This will
+#        work more nicely with the k-fold method.
+# TODO:  Make "predict" not require that target columns be convertible
+#        to numeric form, only that they operate under a weighted sum.
+# TODO:  Remove the "fill" method in favor of "predict".
+# 
+# 
+# TODO:  Implement "parallel read" that create subprocesses that read
+#        chunks of the file, communicate progress, communicate type
+#        digressions, saves chunks in .pkl, then organizes them all
+#        into the expected ordered data format.
[…]

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Predictions/util/data/test.py

@@ -1,0 +1,695 @@
+
+from util.data.data import Data, flatten
+
+# TODO:  Test for file that has a type digression happen on a line
+#        with empty strings in it. This caused a crash [2019-12-17].
+
+
+# Some tests for the data
+def test_data():
+    import os
+    import numpy as np
+
+    print("Testing Data...", end=" ")
+
+    # ----------------------------------------------------------------
+    a = Data()
+
+    # Verify append
+    a.append([1,"a"])
+    a.append([2,"b"])
+    a.append([3,"c"])
+
+    # Verify add column (with edge case, none type)
+    a.add_column([None,None,None])
+    assert(a.types[-1] == type(None))
+
+    # Reassign the missing value column to floats, verify type update
+    a[a.names[-1]] = [1.2, 3.0, 2.4]
+    assert(a.types[-1] == float)
+    # WARNING: Need to check doing (<int>, <str>) assignment
+
+    # Verify automatic type generation AND automatic type-casting
+    a.append([1,2,3])
+    assert(tuple(a[-1]) == tuple([1,"2",3.0]))
+    assert(tuple(map(type,a[-1])) == (int, str, float))
+
+    # Add some missing values.
+    a.append([-1,None,None])
+
+    #  Done modifying "a", rest of tests leave it constant.
+    # ----------------------------------------------------------------
+
+    # Verify that "index" works correctly.
+    assert( a.index([1,"2",3.0]) == 3 )
+    assert( a.index([-1,None,None]) == 4 )
+
+    # Verify that slicing works on descirptors.
+    assert(a.names[:-1] == ['0','1'])
+
+    # Verify that slicing works on descriptors of views.
+    b = a[:,1:]
+    assert(b.names[1:] == ['2'])
+
+    # Verify in-place addition of rows (different length data with same column names)
+    b = a[:-2].copy()
+    c = a[1:-2]
+    b += c
+    assert(tuple(b["0"]) == tuple([1,2,3,2,3]))
+
+    # Verify in-place addition of new columns (same length data with new columns)
+    b = a[:,:-1].copy()
+    c = a[:,-1:].copy()
+    c["3"] = range(len(c))
+    b += c
+    assert(tuple(b[-2]) == (1,"2",3.0,3))
[…]

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Predictions/util/plot/plot.py

@@ -1,0 +1,1968 @@
+# This module serves to provide a simplified interface to *offline*
+# python plotly plotting. The user can produce plots without ever
+# interacting directly with the dictionary objects that plotly
+# expects. This module currently supports 2D and 3D scatter plots with
+# numerical axes, histograms, subplots (with varying numbers of plots
+# in each row), animations, box-plots, and plot annotations.
+# 
+# Required packages:
+#   random, numbers, os, webbrowser, sys, re, tempfile
+#   numpy
+#   scipy
+# 
+# Imports nested in appropriate functions:
+#   import plotly
+#   from scipy.spatial import ConvexHull
+#   from scipy.spatial import Delaunay
+# 
+# INSTALLATION:
+# 
+#  Installation requires a SPECIFIC VERSION OF PLOTLY. Here is the
+#  standard set of packages required for usage:
+# 
+#    pip install scipy
+#    pip install numpy
+#    pip install plotly==2.0.15
+# 
+#  This package will not work with newer versions of plotly because
+#  they changed the underlying storage data types for figures.
+#  Any plotly update for this body of code is unlikely.
+# 
+# 
+# USAGE:
+# 
+#  The available (user accessible) functions are:
+# 
+#    plot.Plot        -- The primary class for holding / creating plots.
+#    plot.multiplot   -- A mechanism for plotting multiple Plot
+#                        objects in the same window.
+#    plot.create_html -- A function for generating a local HTML file
+#                        from a figure object (in Plotly terms).
+#    plot.iplot       -- A convenience wrapper for generating
+#                        interactive plots in a Jupyter notebook with
+#                        multiplot functionality as well.
+# 
+
+# --------------------------------------------------------------------
+#                       DEVELOPER COMMENTS
+# 
+# TODO: Plot.add_zero_line(func)
+# TODO: Plot.add_frame(..., persist=True)
+# TODO: Adding multiple frames where the first has no edges and the
+#       rest have edges causes all frames to look like first.
+# 
+# --------------------------------------------------------------------
+
+import random, numbers, os, webbrowser, sys, re, tempfile
+import numpy as np
+
+NOTEBOOK_MODE = False    # Jupyter notebook mode
+PLOT_MARGIN = 50         # In pixels
+PLOT_POINTS = 1000       # Number of samples
+BRIGHTNESS_RANGE = 0.6   # For default shading of points
+RANDOM_SEED = 0          # Seed used for new color generation
+MIN_PALATTE_COLORS = 40  # Number of palatte entries to create
[…]

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Revision/fix_old_data/new_data/__init__.py

@@ -1,30 +1,0 @@
-import numpy as np
-
-QUOTES = {'"'}
-DEFAULT_DISPLAY_WAIT = 3.
-COMMON_SEPARATORS = [",", " ", "	", ";"]
-UPDATE_FREQUENCY = .5  # How much time (in seconds) between updates when reading data.
-MAX_ERROR_PRINTOUT = 10 # Only print out this many errors when processing data
-
-# Definition of separators by file extension when saving.
-SEPARATORS = {"csv":",", "tsv":"\t", "ssv":" "}
-
-# Declare some constants.
-NP_TYPES = {str:(np.str_, 16),    # Which numpy types correspond to 
-            int:(np.int64,),      # the python types of variables
-            float:(np.float64,)}
-PY_TYPES = {value[0]:key for (key,value) in NP_TYPES.items()}
-# 
-GENERATOR_TYPE = type(_ for _ in ())
-
-FILE_SAMPLE_SIZE = 2**20
-MISSING_SAMPLE_SIZE = 10000
-
-is_none = lambda v: type(v) == type(None)
-
-# coding: future_fstrings
-# from future import print_function
-
-from .data import Data
-from .categories import regular_simplex, category_ratio
-# from util.data.test import *

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/Data/Revision/test_func.py

@@ -1,49 +1,0 @@
-import numpy as np
-
-from util.plot import Plot
-
-# --------------------------------------------------
-#      Suggested functions (not useful here).
-def oscillatory(x, c=1, w=0):
-    return np.cos(2*np.pi*w + np.sum(c*x))
-def product_peak(x, c=1, w=0):
-    return np.prod((c**(-2) + (x-w)**2)**(-1))
-def corner_peak(x, c=1):
-    try:    d = len(x)
-    except: d = 1
-    return (1 + np.sum(c*x))**(-d-1)
-def gaussian(x, c=1, w=0, t=1):
-    return np.exp( - np.sum(c**2 * t * (x - w)**2) )
-def continuous(x, c=1, w=0):
-    return np.exp( - np.sum(c * np.abs(x - w)) )
-# --------------------------------------------------
-
-# A standard component-wise cosine function.
-def cosine(x):
-    return np.sum( np.cos(x*np.pi) )
-
-# A quadratic 2-norm bowl about the origin.
-def quadratic(x):
-    return np.linalg.norm(x)**2
-
-# Return the length of x.
-def linear(x):
-    return np.sum(x)
-
-dim = 2
-test_funcs = [linear, quadratic]
-
-p = Plot()
-
-name = lambda f: str(f).split()[1]
-
-x_range = [0,1] #[-2,2]
-y_range = None #[-3,3]
-
-for f in test_funcs:
-    p.add_func(name(f), f, *([x_range]*dim))
-    
-p.add_func("Error", lambda x: linear(x)-quadratic(x),
-           *([x_range]*dim), opacity=.5)
-
-p.show(y_range=y_range)

### 2020-08-13 [2020-01]_IEEESE_(convex_optimization)/code/raw/visualize.py

@@ -1,182 +1,0 @@
-# Make sure the previous step has been executed.
-previous_step = __import__("2-summarize_data")
-data_file = previous_step.output_file
-
-import os
-from util.data import Data
-from util.plot import Plot, PLOT_MARGIN
-PLOT_MARGIN = 0
-
-
-summary_data = Data.load(data_file)
-summary_data.sort()
-
-print()
-print("summary_data: ",summary_data)
-
-legend = dict(
-    xanchor = "center",
-    yanchor = "top",
-    x = .525,
-    y = .92,
-    orientation = "h",
-    bgcolor="white",
-    bordercolor="grey",
-    borderwidth=.5
-)
-
-layout = dict(
-    margin = dict(t=10,b=35,l=42,r=20),
-)
-
-no_x_title_layout = dict(
-    margin = dict(t=10,b=20,l=42,r=20),
-)
-
-# Pixel offset of label from highest value.
-label_x_location = 25000
-label_y_offset = -1500
-
-# Create a dictionary that returns a default value for missing keys.
-class DefaultDict(dict):
-    def __init__(self, default=None, *args, **kwargs):
-        self.default_value = default
-        super().__init__(*args, **kwargs)
-    def __getitem__(self, key):
-        if (self.default_value is None): return super().get(key, key)
-        else: return super().get(key, self.default_value)
-
-# Generate a method for sorting the plots and series.
-sort_order = DefaultDict(
-    default = DefaultDict(),
-    function = DefaultDict(
-        default = float('inf'),
-        subquadratic = 0,
-        superquadratic = 1,
-        saddle = 2,
-        multimin = 3
-    ),
-    algorithm = DefaultDict(
-        default = float('inf'),
-        SGD = 0,
-        LBFGS = 1,
-        ADAGRAD = 2,
-        ADAM = 3,
-    )
-)
-col_name = None
-sort_func = lambda key: sort_order[col_name][key]
-# Dash style order.
-dash_styles = [None, "dot", "dash", "dashdot"]
-
-
-# Function for making a set of plots given the column and values from
-# that column to make plots for as well as the column to use for the
-# series in each plot.
-def make_plots(plot_col, series_col, plot_values, data, suffix=""):
-    # Cycle each plot column value, append plots to one file.
-    for plot_col_value in plot_values:
-        is_first = plot_col_value == plot_values[0]
-        is_last = plot_col_value == plot_values[-1]
-        # Get the (sub)set of data for this plot.
[…]

### 2020-08-22 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
MQSI paper, making final changes.

\bigskip\hrule\bigskip\medskip
\heading{1. INTRODUCTION}

Many domains of science rely on smooth approximations to real-valued
functions over a closed interval. Piecewise polynomial functions
(splines) provide the smooth approximations for animation in graphics
[Herman et al.\ 2015; Quint 2003], aesthetic structural support in
architecture [Brennan 2020], efficient aerodynamic surfaces in
automotive and aerospace engineering [Brennan 2020], prolonged
effective operation of electric motors [Berglund et al.\ 2009], and
accurate nonparametric approximations in statistics [Knott
2012]. While polynomial interpolants and regressors apply broadly,
splines are often a good choice because they can approximate globally
complex functions while minimizing the local complexity of an
approximation.

It is often the case that the true underlying function or phenomenon
being modeled has known properties like convexity, positivity,
various levels of continuity, or monotonicity. Given a reasonable
amount of data, it quickly becomes difficult to achieve desirable
properties in a single polynomial function. In general, the
maintenance of function properties through interpolation/regression is
referred to as {\it shape preserving} [Fritsch and Carlson 1980;
Gregory 1985]. The specific properties the present algorithm will
preserve in approximations are monotonicity and $C^2$ continuity. In
addition to previously mentioned applications, these properties are
crucially important in statistics to the approximation of a cumulative
distribution function and subsequently the effective generation of
random numbers from a specified distribution [Ramsay 1988].  A spline
function with these properties could approximate a cumulative
distribution function to a high level of accuracy with relatively few
intervals. A twice continuously differentiable approximation to a
cumulative distribution function (CDF) would produce a corresponding
probability density function (PDF) that is continuously
differentiable, which is desirable.

The currently available software for monotone piecewise polynomial
interpolation includes quadratic [He and Shi 1998], cubic [Fritsch and
Carlson 1980], and (with limited application) quartic [Wang and Tan
2004; Piah and Unsworth 2011; Yao and Nelson 2018] cases. In
addition, a statistical method for bootstrapping the construction of
an arbitrarily smooth monotone fit exists [Leitenstorfer and Tutz
2006], but the method does not take advantage of known analytic
properties of quintic polynomials. The code by Fritsch [1982] for
$C^1$ cubic spline interpolation is the predominantly utilized code
for constructing monotone interpolants at present. Theory has been
provided for the quintic case [Ulrich and Watson 1994; He{\ss} and
Schmidt 1994] and that theory was recently utilized in a proposed
algorithm [Lux 2020] for monotone quintic spline construction, however
no published mathematical software exists.

The importance of piecewise quintic interpolation over lower order
approximations can be simply observed. In general, the order of a
polynomial determines the number of function (and derivative) values
it can interpolate, which in turn determines the growth rate of error
away from interpolated values. $C^2$ quintic (order six) splines
match the function value and two given derivatives at each breakpoint.
This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
necessary and sufficient conditions in Ulrich and Watson [1994] for the
construction of monotone quintic spline interpolants of monotone data.
Precisely, the problem is, given a strictly increasing sequence $X_1<X_2<
\cdots <X_n$ of breakpoints with corresponding monotone increasing
function values $Y_1\le Y_2\le \cdots \le Y_n$, find a $C^2$ monotone
increasing quintic spline $Q(x)$ with the same breakpoints satisfying
$Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI} actually does something slightly
more general, producing $Q(x)$ that is monotone increasing (decreasing)
wherever the data is monotone increasing (decreasing).)

The remainder of this paper is structured as follows: Section 2 provides
the algorithms for constructing a $C^2$ monotone quintic spline interpolant
to monotone data, Section 3 outlines the method of spline representation
($B$-spline basis) and evaluation, Section 4 analyzes the complexity and
sensitivity of the algorithms in {\tt MQSI}, and Section 5 presents 
timing data and some graphs of constructed interpolants.

### 2020-08-22 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
MQSI paper, making final changes.

\heading{2. MONOTONE QUINTIC INTERPOLATION}

In order to construct a monotone quintic interpolating spline, two
primary problems must be solved. First, reasonable derivative values
at data points need to be estimated. Second, the estimated derivative
values need to be modified to enforce monotonicity on all polynomial
pieces.

Fritsch and Carlson [1980] originally proposed the use of central differences
to estimate derivatives, however this often leads to extra and unnecessary
{\it wiggles} in the spline when used to approximate second derivatives.
In an attempt to capture the local shape of the data, this package uses a
facet model from image processing [Haralick and Watson 1981] to estimate
first and second derivatives at breakpoints. Rather than picking a local
linear or quadratic fit with minimal residual, this work uses a quadratic
facet model that selects the local quadratic interpolant with minimum
magnitude curvature.

%% ===================================================================
%%                   Algorithm 1: QUADRATIC_FACET
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 1:} {
\tt QUADRATIC\_FACET}$\bigl(X(1{:}n)$, $Y(1{:}n)$, $i \bigr)$

where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
\le n$, and $n \ge 3$. Returns the slope and curvature at $X_i$ of the
local quadratic interpolant with minimum magnitude curvature.

}
{\parindent=3mm
%% -------------------------------------------------------------------
\item{} {\tt if} $\bigl((i\ne1\wedge Y_i \approx Y_{i-1})$ {\tt or}
$(i\ne n\wedge Y_i \approx Y_{i+1})\bigr)$ {\tt then return $(0,0)$}

\item{} {\tt else if} $i=1$ {\tt then}
\itemitem{} $f_1\:\hbox{interpolant to }(X_1,Y_1)$,
  $(X_2,Y_2)$, and $(X_3,Y_3)$.
\itemitem{} {\tt if} $\bigl(Df_1(X_1)(Y_2-Y_1)<0\bigr)$ {\tt then return}
  $\bigl(0,0\bigr)$
\itemitem{} {\tt else return} $\bigl(Df_1(X_1),D^2f_1\bigr)$
\itemitem{} {\tt endif}

\item{} {\tt else if} $i=n$ {\tt then}
\itemitem{} $f_1\:\hbox{interpolant to }(X_{n-2},Y_{n-2})$,
  $(X_{n-1},Y_{n-1})$, and $(X_n,Y_n)$.
\itemitem{} {\tt if} $\bigl(Df_1(X_n)(Y_n-Y_{n-1})<0\bigr)$ {\tt then return}
  $\bigl(0,0\bigr)$
\itemitem{} {\tt else return} $\bigl(Df_1(X_n),D^2f_1\bigr)$
\itemitem{} {\tt endif}

\item{} {\tt else if} $\bigl(1<i<n\wedge (Y_{i+1} - Y_i)(Y_i - Y_{i-1}) < 0
\bigr)$ {\tt then}
\itemitem{} The point $(X_i$, $Y_i)$ is an extreme point. The
quadratic with minimum magnitude
curvature that has slope zero at $X_i$ will be the facet chosen.
\itemitem{} $f_1\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$, $(X_i,Y_i)$,
  and $Df_1(X_i) = 0$.
\itemitem{} $f_2\:\hbox{interpolant to }(X_i,Y_i)$, $(X_{i+1},Y_{i+1})$,
  and $Df_2(X_i) = 0$.
\itemitem{} {\tt if} $\bigl(|D^2f_1| \leq |D^2f_2|\bigr)$ {\tt then
  return} $\bigl(0$, $D^2f_1\bigr)$
\itemitem{} {\tt else return} $\bigl(0$, $D^2f_2\bigr)$
\itemitem{} {\tt endif}

\item{} {\tt else}
\itemitem{} {The point $(X_i$, $Y_i)$ is in a monotone segment of
data. In the following, it is possible that $f_1$ or $f_3$ do
not exist because $i \in \{2, n-1\}$. In those cases, the minimum
magnitude curvature among existing quadratics is chosen.}
\itemitem{} $f_1\:\hbox{interpolant to }(X_{i-2},Y_{i-2})$,
  $(X_{i-1},Y_{i-1})$, and $(X_i,Y_i)$.
\itemitem{} $f_2\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$,
  $(X_i,Y_i)$, and $(X_{i+1},Y_{i+1})$.
\itemitem{} $f_3\:\hbox{interpolant to }(X_i,Y_i)$,
  $(X_{i+1},Y_{i+1})$, and $(X_{i+2},Y_{i+2})$.

\itemitem{} {\tt if} $\bigl(Df_1(X_i)(Y_i-Y_{i-1}) \ge 0 \wedge
  |D^2f_1| = \min\bigl\{ |D^2f_1|,
  |D^2f_2|, |D^2f_3|\bigr\} \bigr)$ {\tt then}
\itemitem{} \codent {\tt return} $\bigl(Df_1(X_i),D^2f_1\bigr)$

\itemitem{} {\tt else if} $\bigl(Df_2(X_i)(Y_i-Y_{i-1}) \ge 0 \wedge
  |D^2f_2| = \min\bigl\{ |D^2f_1|,
  |D^2f_2|, |D^2f_3|\bigr\} \bigr)$
\itemitem{} \codent {\tt then return} $\bigl(Df_2(X_i),D^2f_2\bigr)$

\itemitem{} {\tt else if} $\bigl(Df_3(X_i)(Y_{i+1}-Y_i) \ge 0\bigr)$ {\tt then}
\itemitem{} \codent {\tt return} $\bigl(Df_3(X_i),D^2f_3\bigr)$
\itemitem{} {\tt else return} $(0, 0)$
\itemitem{} {\tt endif}
\item{} {\tt endif}
}
\vskip 5mm
%% ----------------------------------------------------------------------

The estimated derivative values by the quadratic facet model are not
guaranteed to produce monotone quintic polynomial segments. Ulrich and
Watson [1994] established tight constraints on the monotonicity of a
quintic polynomial piece, while deferring to He{\ss} and Schmidt
[1994] for a relevant simplified case. The following algorithm
implements a sharp check for monotonicity by considering the
nondecreasing case. The nonincreasing case is handled similarly.

%% ===================================================================
%%                    Algorithm 2: IS_MONOTONE
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 2:}

{\tt IS\_MONOTONE}$\bigl(x_0$, $x_1$, $f(x_0)$, $Df(x_0)$,
$D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1) \bigr)$

where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$.

}
{\parindent=6mm
%% -------------------------------------------------------------------
\item{1.} {\tt if} $\bigl(f(x_0) \approx f(x_1)\bigr)$ {\tt then}
\item{2.} \codent {\tt return} $\bigl( 0 = Df(x_0) = Df(x_1)
  = D^2f(x_0) = D^2f(x_1) \bigr)$
\item{3.} {\tt endif}
\item{4.} {\tt if} $\bigl(Df(x_0) < 0$ {\tt or} $Df(x_1) < 0\bigr)$ {\tt
  then return FALSE endif}
\item{5.} $w \: x_1 - x_0$
\item{6.} $z \: f(x_1) - f(x_0)$

\item{} {The necessity of Steps 1--4 follows directly from the
  fact that $f$ is $C^2$. The following Steps 7--13 coincide with a
  simplified condition for quintic monotonicity that reduces to one of
  cubic positivity studied by Schmidt and He{\ss} [1988]. Given
  $\alpha$, $\beta$, $\gamma$, and $\delta$ as defined by Schmidt and
  He{\ss}, monotonicity results when $\alpha \geq 0$, $\delta \geq 0$,
  $\beta \geq \alpha - 2 \sqrt{\alpha \delta}$, and $\gamma \geq
  \delta - 2 \sqrt{\alpha \delta}$.  Step 4 checked for $\delta < 0$,
  Step 8 checks $\alpha < 0$, Step 10 checks $\beta < \alpha - 2
  \sqrt{\alpha \delta}$, and Step 11 checks $\gamma < \delta - 2
  \sqrt{\alpha \delta}$. If none of the monotonicity conditions are
  violated, then the order six piece is monotone and Step 12
  concludes.}

\item{7.} {\tt if} $\bigl(Df(x_0) \approx 0$ {\tt or} $Df(x_1) \approx
0\bigr)$ {\tt then}
\item{8.} \codent {\tt if} $\bigl(D^2f(x_1)w > 4Df(x_1)$ {\tt then
return FALSE endif}
\item{9.} \codent $t \: 2 \sqrt{Df(x_0) (4Df(x_1) - D^2f(x_1) w) }$
\item{10.} \codent {\tt if} $\bigl(t + 3Df(x_0) + D^2f(x_0)w < 0 \bigr)$
  {\tt then return FALSE endif}
\item{11.} \codent {\tt if} $\bigl(60z - w\bigl(24Df(x_0) + 32Df(x_1) - 2t
  + w(3D^2f(x_0) - 5D^2f(x_1))\bigr) < 0\bigr)$
\item{}   \codent \codent {\tt then return FALSE endif}
\item{12.} \codent {\tt return TRUE}
\item{13.} {\tt endif}

\item{} {The following code considers the full quintic monotonicity
case studied by Ulrich and Watson [1994]. Given $\tau_1$, $\alpha$,
$\beta$, and $\gamma$ as defined by Ulrich and Watson, a quintic
piece is proven to be monotone if and only if
$\tau_1 > 0$, and $\alpha, \gamma > -(\beta+2)/2$ when $\beta \leq 6$,
and $\alpha, \gamma > -2 \sqrt{\beta-2}$ when $\beta > 6$.
Step 14 checks $\tau_1 \le 0$, Steps 19 and 20 determine monotonicity based
on $\alpha$, $\beta$, and $\gamma$.}

\item{14.} {\tt if} $\bigl( w\bigl(2\sqrt{Df(x_0)\,Df(x_1)} - 3(Df(x_0) +
  Df(x_1))\bigr) - 24z \leq 0 \bigr)$
\item{} \codent {\tt then return FALSE endif}
\item{15.} $t \: \bigl(Df(x_0)\, Df(x_1)\bigr)^{3/4}$
\item{16.} $\alpha \: (4 Df(x_1) - D^2f(x_1)w) \sqrt{Df(x_0)} / t$
\item{17.} $\gamma \: (4 Df(x_0) - D^2f(x_0)w) \sqrt{Df(x_1)} / t$
\item{18.} $\displaystyle \beta \: {60z/w + 3\bigl(w(D^2f(x_1) -
  D^2f(x_0)) - 8(Df(x_0) + Df(x_1))\bigr) \over 2 \sqrt{Df(x_0)\,Df(x_1)}}$
\item{19.} {\tt if} $(\beta \leq 6)$ {\tt then return}
$\bigl( \min\{\alpha,\gamma\} > - (\beta + 2) / 2 \bigr)$
\item{20.} {\tt else} {\tt return}
$\bigl( \min\{\alpha,\gamma\} > -2 \sqrt{\beta - 2}\,\bigr)$
\item{21.} {\tt endif}
}
\vskip 5mm
%% ----------------------------------------------------------------------

It is shown by Ulrich and Watson [1994] that when $0 = DQ(X_i) =
DQ(X_{i+1}) = D^2Q(X_i) = D^2Q(X_{i+1})$, the quintic polynomial
over $[X_i$, $X_{i+1}]$ is guaranteed to be monotone. Using this fact, the
following algorithm shrinks (in magnitude) initial derivative estimates
until a monotone spline is achieved and outlines the core routine in the
accompanying package.

%% ===================================================================
%%       Algorithm 3: MQSI - Monotone quintic spline interpolant
\vskip 3mm
{\parindent 0mm
{\bf Algorithm 3:} {\tt MQSI}$\bigl(X(1{:}n), Y(1{:}n) \bigr)$

where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are data
points. Returns monotone quintic spline interpolant $Q(x)$ such that
$Q(X_i) = Y_i$ and is monotone increasing (decreasing) on all
intervals that $Y_i$ is monotone increasing (decreasing).

}
{\parindent=6mm
\item{} Approximate first and second derivatives at $X_i$ with {\tt
  QUADRATIC\_FACET}.
\item{} {\tt for} $i\:1$ {\tt step} 1 {\tt until} $n$ {\tt do}
\item{} \codent $(u_i$, $v_i)\:$ {\tt QUADRATIC\_FACET}$(X$, $Y$, $i)$
\item{} {\tt enddo}
\item{} Identify and store all intervals where $Q$ is nonmonotone in a
queue $\cal Q$.
\item{} {\tt for} $i\:1$ {\tt step} 1 {\tt until} $n-1$ {\tt do}
\item{} \codent {\tt if not IS\_MONOTONE}$(X_i$, $X_{i+1}$, $Y_i$, $u_i$,
$v_i$, $Y_{i+1}$, $u_{i+1}$, $v_{i+1})$ {\tt then}
\item{} \codent\codent Add interval $\bigl(X_i$, $X_{i+1}\bigr)$ to
queue $\cal Q$.
\item{} \codent {\tt endif}
\item{} {\tt enddo}
\item{} {\tt do while} $\bigl($ queue $\cal Q$ of intervals is nonempty $\bigr)$
\itemitem{} Shrink (in magnitude) $DQ$ (in $u$) and $D^2Q$ (in $v$)
  that border intervals where $Q$ is nonmonotone.  
\itemitem{} Identify and store remaining intervals where $Q$ is
  nonmonotone in queue $\cal Q$.
\item{} {\tt enddo}
\item{} Construct and return a $B$-spline representation of $Q(x)$.
}
\vskip 3mm
%% ----------------------------------------------------------------------

Since {\tt IS\_MONOTONE} can handle both nondecreasing and
nonincreasing simultaneously by taking into account the sign of $z$,
Algorithm 3 produces $Q(x)$ that is monotone increasing (decreasing)
over exactly the same intervals that the data $(X_i$, $Y_i)$ is
monotone increasing (decreasing).

Given the minimum magnitude curvature nature of the initial derivative
estimates, it is desirable to make the smallest necessary changes to
the initial interpolating spline $Q$ while enforcing monotonicity. In
practice a binary search for the boundary of monotonicity is used in
place of solely shrinking $DQ$ and $D^2Q$ at breakpoints adjoining
{\it active\/} intervals: intervals over which $Q$ is nonmonotone at
least once during the search. The binary search considers a Boolean
function $B_i(s)$, for $0 \le s \le 1$, that is true if the order six
polynomial piece of $Q(x)$ on $[X_i, X_{i+1}]$ matching derivatives
$Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and 
derivatives $Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, 
$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false otherwise. 
The binary search is only applied at those breakpoints adjoining intervals
$[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and hence $B_i(1)$ is false.
It is further assumed that there exists $0 \le s^* \le 1$ such that $B_i(s)$
is true for $0 \le s \le s^*$ and false for some $1 > s > s^*$. Since the
derivative conditions at interior breakpoints are shared by intervals left
and right of the breakpoint, the binary search is performed at all
breakpoints simultaneously.  Specifically, the monotonicity of $Q$ is
checked on all active intervals in each step of the binary search to
determine the next derivative modification at each breakpoint. The goal
of this search is to converge on the boundary of the monotone region in
the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
Watson [1994]) for all intervals. This multiple-interval binary search
allows the value zero to be obtained for all (first and second) derivative
values in a fixed maximum number of computations, hence has no effect on
computational complexity order. This binary search algorithm is outlined
below.

%% ===================================================================
%%              Algorithm 4: Global binary search
\vskip 5mm {\parindent 0mm {\bf Algorithm 4:} {\tt
    BINARY\_SEARCH}$\bigl(X(1{:}n), Y(1{:}n), u(1{:}n), v(1{:}n) \bigr)$

where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are
data points, and $Q(x)$ is a quintic spline interpolant such that
$Q(X_i) = Y_i$, $DQ(X_i) = u_i$, $D^2Q(X_i) = v_i$. Modifies
derivative values ($u$ and $v$) of $Q$ at data points to ensure
{\tt IS\_MONOTONE} is true for all intervals defined by adjacent data
points, given a desired precision $\mu \in {\bbb R}$. proj$\bigl(w$,
int$(a,b)\bigr)$ denotes the projection of $w$ onto the closed
interval with endpoints $a$ and $b$.

}
{\parindent=6mm
\item{} Initialize the step size $s$, make a copy of data defining $Q$,
and construct three queues necessary for the multiple-interval binary search.
\item{} $s \: 1$
\item{} $(\hat u,\hat v) \: (u,v)$
\item{} {\tt searching} $\:$ {\tt TRUE}
\item{} {\tt checking} $\:$ empty queue for holding left indices of {\it intervals}
\item{} {\tt growing} $\:$ empty queue for holding indices of {\it data points}
\item{} {\tt shrinking} $\:$ empty queue for holding indices of {\it data points}
\item{} {\tt for} $i\:1$ {\tt step} 1 {\tt until} $n-1$ {\tt do}
\itemitem{} {\tt if not IS\_MONOTONE}$\bigl( X_i$, $X_{i+1}$, $Y_i$, $u_i$,
$v_i$, $Y_{i+1}$, $u_{i+1}$, $v_{i+1} \bigr)$ {\tt then}
\itemitem{} \codent Add data indices $i$ and $i+1$ to queue {\tt shrinking}.
\itemitem{} {\tt endif}
\item{} {\tt enddo}
\item{} {\tt do while} $\bigl(${\tt searching or} ({\tt shrinking} is
nonempty)$\bigr)$
\itemitem{} Compute the {\it step size} $s$ for this iteration of the search.
\itemitem{} {\tt if  searching  then} $s \: \max\{\mu, s/2\}$
{\tt else} $s \: 3s/2$ {\tt endif}
\itemitem{} {\tt if} $(s=\mu)$ {\tt then searching} $\:$ {\tt FALSE};
  clear queue {\tt growing endif}
\itemitem{} Increase in magnitude $u_i$ and $v_i$ for all data indices
  $i$ in {\tt growing} such that the points $X_i$ are strictly adjoining
  intervals over which $Q$ is monotone.
\itemitem{} {\tt for} ($i \in$ {\tt growing}) {\tt and} ($i \not\in$
{\tt shrinking}) {\tt do}
\itemitem{} \codent $u_i \: \hbox{proj}\bigl(u_i + s\,\hat u_i, \hbox{int}
  (0,\hat u_i)\bigr)$
\itemitem{} \codent $v_i \: \hbox{proj}\bigl(v_i + s\,\hat v_i, \hbox{int}
  (0,\hat v_i)\bigr)$
\itemitem{} \codent Add data indices $i-1$ (if not 0) and $i$ (if not $n$) to queue
  {\tt checking}.
\itemitem{} {\tt enddo}
\itemitem{} Decrease in magnitude $u_i$ and $v_i$ for all data indices
  $i$ in {\tt shrinking} and ensure those data point indices are placed
  into {\tt growing} when {\tt searching}.
\itemitem{} {\tt for} $i \in$ {\tt shrinking do}
\itemitem{} \codent If {\tt searching}, then add index $i$ to queue
  {\tt growing} if not already present.
\itemitem{} \codent $u_i \: \hbox{proj}\bigl(u_i - s\,\hat u_i, \hbox{int}
  (0,\hat u_i)\bigr)$
\itemitem{} \codent $v_i \: \hbox{proj}\bigl(v_i - s\,\hat v_i, \hbox{int}
  (0,\hat v_i)\bigr)$
\itemitem{} \codent Add data indices $i-1$ (if not 0) and $i$ (if not $n$) to queue
  {\tt checking}.
\itemitem{} {\tt enddo}
\itemitem{} Empty queue {\tt shrinking}, then check all intervals left-indexed
  in queue {\tt checking} for monotonicity with {\tt IS\_MONOTONE},
  placing data endpoint indices of intervals over which $Q$ is nonmonotone
  into queue {\tt shrinking}.
\itemitem{} Clear queue {\tt shrinking}.
\itemitem{} {\tt for} $i \in$ {\tt checking do}
\itemitem{} \codent
  {\tt if not IS\_MONOTONE}$\bigl( X_i$, $X_{i+1}$, $Y_i$, $u_i$,
  $v_i$, $Y_{i+1}$, $u_{i+1}$, $v_{i+1} \bigr)$ {\tt then}
\itemitem{} \codent \codent Add data indices $i$ and $i+1$ to {\tt shrinking}.
\itemitem{} \codent {\tt endif}
\itemitem{} {\tt enddo}
\itemitem{} Clear queue {\tt checking}.
\item{} {\tt enddo }
}
\vskip 5mm
%% ----------------------------------------------------------------------

%% \itemitem{} Take search steps either modifying derivative values
%%   towards monotonicity (zero) or towards the original values over all
%%   intervals that are either nonmonotone or previously modified.

In the subroutine {\tt MQSI}, $\mu = 2^{-26}$, which results in 26
guaranteed search steps for all intervals that are initially
nonmonotone. An additional 43 steps could be required to reduce a
derivative magnitude to zero with step size growth rate of $3/2$. This
can only happen when $Q$ becomes nonmonotone on an interval for the
first time while the step size equals $\mu$, but for which the only
viable solution is a derivative value of zero. The maximum number of
steps is due to the fact that $\sum_{i=0}^{42} \mu (3/2)^i > 1$. In
total {\tt BINARY\_SEARCH} search could require 69 steps.

### 2020-08-22 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
MQSI paper, making final changes.

\heading{5. PERFORMANCE AND APPLICATIONS}

This section contains graphs of sample {\tt MQSI} results given various
data configurations. Execution times for Algorithms 1 and 4 are given; the
total execution time of {\tt MQSI} is utterly dominated by the time for a
banded linear system solve computing the $3n$ $B$-spline coefficients,
which is ${\cal O}(n)$.  The files {\tt sample\_main.f90} and {\tt
sample\_main.dat} accompanying the subroutine {\tt MQSI} illustrate Fortran
2003 subroutine usage with optional arguments and data points from a file.
Compilation instructions and the full package contents are specified in
the {\tt README} file.

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/2-piecewise-polynomial.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/3-large-tangent.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/4-signal-decay.eps}}
{\narrower\noindent\rmVIII Fig.\ 2.
{\ttVIII MQSI} results for three of the functions in the included test
suite. The {\itVIII piecewise polynomial} function (top) shows the
interpolant capturing local linear segments, local flats, and
alternating extreme points. The {\itVIII large tangent} (middle)
problem demonstrates outcomes on rapidly changing segments of data.
The {\itVIII signal decay} (bottom) alternates between extreme values
of steadily decreasing magnitude.
\par}
\vskip 6mm
\centerline{\epsfxsize=4truein \epsffile{vis/5-real-data.eps}}
{\narrower\noindent\rmVIII Fig.\ 3.
{\ttVIII MQSI} results when approximating the cumulative distribution
function of system throughput (bytes per second) data for a computer
with a 3.2 GHz CPU performing file read operations from Cameron et al.
[2019]. The empirical distribution of 30 thousand throughput values is
shown in the red dashed line, while the solid line with stylized
markers denotes the approximation made with MQSI given equally spaced
empirical distribution points from a sample of size 100.
\par}
\endinsert

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/6-random-monotone.eps}}
{\narrower\noindent\rmVIII  Fig.\ 4.
The {\itVIII random monotone} test poses a particularly challenging
problem with large variations in slope. Notice that despite drastic
shifts in slope, the resulting monotone quintic spline interpolant
provides smooth and reasonable estimates to function values between data.
\par}
\iffalse
\vskip 15mm
\centerline{\epsfxsize=4truein \epsffile{vis/7-runtimes-10.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/8-runtimes-100.eps}}
{\everymath={\scriptstyle} \narrower\noindent\rmVIII Fig.\ 5.
Median total runtime of the routine {\ttVIII MQSI} for number of data points
$n=10$
to 100 (top) and $n=100$ to 1000 (bottom), generated from 100
repeated trials averaged over $14$ different testing functions. The
timings are colored by major algorithmic component, but the vast
majority of execution time is spent solving the banded linear system
of equations in the {\ttVIII FIT\_SPLINE} routine. The runtimes for
the quadratic facet (Algorithm 1) take roughly one microsecond
($10^{-6}$ seconds) per breakpoint, while the binary search
(Algorithm 4) takes roughly four microseconds per breakpoint.
\par}
\fi
\endinsert

Throughout, all visuals have points that are stylized by local
monotonicity conditions. Blue circles denote extreme points, purple
squares are in {\it flat} regions with no change in function value,
red down triangles are monotone decreasing, and green up triangles are
monotone increasing.

Figure 2 offers examples of the interpolating splines produced by the
routine {\tt MQSI} on various hand-crafted sets of data. These same
data sets are used for testing local installations in the provided
program {\tt test\_all.f90}. Notice that the quadratic facet model
perfectly captures the local linear segments of data in the piecewise
polynomial test for Figure 2. Figure 3 depicts an approximation of a
cumulative distribution function made by {\tt MQSI} on a computer
systems application by Cameron et al. [2019] that studies the
distribution of throughput (in bytes per second) when reading files
from a storage medium. Figure 4 provides a particularly difficult
monotone interpolation challenge using randomly generated monotone
data.

On a computer running MacOS 10.15.5 with a 2 GHz Intel Core i5 CPU, the
quadratic facet (Algorithm 1) takes roughly one microsecond ($10^{-6}$
seconds) per breakpoint, while the binary search (Algorithm 4) takes roughly
four microseconds per breakpoint; these times were generated from 100
repeated trials averaged over 14 different testing functions.  The vast
majority of execution time is spent solving the banded linear system of
equations in the routine {\tt FIT\_SPLINE} for the $B$-spline coefficients.
For large problems ($n > 100$) it would be faster to construct splines
over intervals independently (each interval requiring a $6 \times 6$ linear
system to be solved for a local $B$-spline representation, or the
construction of the Newton form of the interpolating polynomial from the
derivative information at the endpoints), however the single linear system
is chosen here for the decreased redundancy in the spline description. An
optional argument to {\tt MQSI} returns the derivative information at the
breakpoints (interval endpoints) for the Newton form of the interpolating
polynomial (piece of $Q(x)$) over each interval.

### 2020-08-22 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
MQSI paper, making final changes.

% Algorithm XXXX: MQSI---Monotone quintic spline interpolation,
% Thomas C.H. Lux, Layne T. Watson, Tyler H. Chang,
% and William Thacker
% ACM Trans. Math. Software, ?/?/2020.
% 
%
\magnification=1000 \hsize=30pc \vsize=47pc
\parindent=10pt \parskip=0pt plus 3pt
\baselineskip=13pt plus 2pt minus 1pt \lineskiplimit=2pt
\lineskip=2pt plus 2pt %\raggedright
\tolerance=800
%\abovedisplayskip=6pt plus 6pt minus 3pt
%\abovedisplayshortskip=0pt plus 3pt
%\belowdisplayskip=6pt plus 6pt minus 3pt
%\belowdisplayshortskip=6pt plus 2pt minus 3pt

% \cite is a placeholder to use for references while writing.
\def\cite#1{[{\tt REF}]}

% Color definitions.
\input colordvi
\def\BEGINC{\textPeach} \def\ENDC{\textBlack}
\def\red#1{\BEGINC#1\ENDC}
\let\beginred=\BEGINC
\let\endred=\ENDC

% Commands for inserting an 'algorithm' section.
\def\thickrule{\hrule\hrule\hrule}
\def\codent{\hskip 1.0em} % <- 'tab' for indentation.

% fonts
\font\itVIII=cmti8 \font\slVIII=cmsl8 \font\bfVIII=cmbx8
\font\ttVIII=cmtt8 \font\caps=cmcsc10 \font\capsVIII=cmcsc10 at 8pt
\font\ttXVIII=cmtt12 at 18pt \font\eightmi=cmmi8 \font\eightsy=cmsy8
\font\rmeight=cmr8  \def\rmVIII{\rmeight \baselineskip=10pt plus 2pt
minus 1pt \lineskiplimit=2pt \lineskip=2pt plus 1pt
\parskip=0pt \textfont0=\rmeight \textfont1=\eightmi \textfont2=\eightsy}
\font\hlvXVIII=phvr at 18pt \font\hlvVIII=phvr at 8pt \font\hlv=phvr

% math blackboard font
\font\tenamsb=msbm10 \font\sevenamsb=msbm7 \font\fiveamsb=msbm5
\newfam\bbfam
\textfont\bbfam=\tenamsb
\scriptfont\bbfam=\sevenamsb
\scriptscriptfont\bbfam=\fiveamsb
\def\bbb{\fam\bbfam}

%Postscript figures.
\input epsf
\def\lcaption#1#2{{\noindent\rmVIII Fig.~#1. #2\par}}
\def\ltabcaption#1#2{{\noindent\rmVIII Table #1. #2\par}}

% other definitions.
\def\h{\par\hangindent=10pt\hangafter=1\noindent}
\def\heading#1{\par\bigskip\leftline{\hlv#1}\nobreak\smallskip
\noindent\ignorespaces}
\newdimen\headdigit{\setbox11=\hbox{\hlv 8. }\global\headdigit=\wd11}
\def\footnoterule{\bigskip\kern-3pt \hrule \kern 2.6pt \smallskip}
\def\bullitem{\par\vskip 2pt plus 2pt\hangindent=15pt \hangafter=1
\noindent\hbox to 15pt{---\hfil}\ignorespaces}
\def\bulllitem{\par\vskip 2pt plus 2pt\hangindent=30pt
\hangafter=1 \noindent\hskip 15pt\hbox to
15pt{---\hfil}\ignorespaces}
\def\Th#1 #2\par{\par\medskip{\it{\caps#1}. #2}\par\medskip}
\def\btable#1{\topinsert \rmVIII \let\bf=\bfVIII \textfont0=\scriptfont0
\scriptfont0=\scriptscriptfont0 \textfont1=\scriptfont1
\scriptfont1=\scriptscriptfont1 \centerline{#1}}
\let\xpar=\par
\def\ds{\displaystyle}\def\ab{\allowbreak}

% math definitions
\def\tn#1{\left\Vert #1\right\Vert_2}
\def\mn#1{\left\Vert #1\right\Vert_\infty}
\def\lee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord<$}\nointerlineskip
\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
\def\gee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord>$}\nointerlineskip
\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
\def\real{\mathop{\rm I\!R}\nolimits}
\def\R{{\bf R}}
\def\Rd{{\bf R}^d}
\def\Rdd{{\bf R}^{d\times d}}
%\def\bull{\hfill \vrule height6pt width6pt depth0pt}
\def\:{\mathrel{:\mathord=}}

% OUTPUT
\def\rightheadline{\hfill\hlvVIII\rightrh\hskip20pt$\scriptstyle\bullet$\hskip
20pt\folio}
\def\leftheadline{\hlvVIII\folio\hskip20pt$\scriptstyle\bullet$\hskip
20pt\leftrh\hfill}
\footline={\hfill}%no page number on first page
\output{\plainoutput}
\def\plainoutput{\shipout\vbox{\makeheadline\pagebody\makefootline}
  \advancepageno \global\footline={\hfill}
  \ifodd\pageno
        \global\headline={\rightheadline}
  \else
        \global\headline={\leftheadline}
  \fi
  \ifnum\outputpenalty>-20000 \else\dosupereject\fi}
\def\pagebody{\vbox to\vsize{\boxmaxdepth\maxdepth \pagecontents}}
\def\makeheadline{\vbox to 0pt{\vskip-22.5pt
  \line{\vbox to8.5pt{}\the\headline}\vss}\nointerlineskip}
\def\makefootline{\baselineskip24pt\vskip-6pt\line{\the\footline}}
\def\dosupereject{\ifnum\insertpenalties>0pt
  \line{}\kern-\topskip\nobreak\vfill\supereject\fi}
%\def\footnoterule{\vskip10pt\kern-3pt \hrule width 3em \kern 2.6pt}

\def\leftrh{Lux et al.}
\def\rightrh{Algorithm XXXX: MQSI}

{\baselineskip=24pt
\leftline{\hlvXVIII Algorithm XXXX: MQSI---Monotone Quintic}
\leftline{\hlvXVIII Spline Interpolation}}
\bigskip\medskip
\leftline{\hlv THOMAS C. H. LUX and LAYNE T. WATSON}
\smallskip
\leftline{\hlv Virginia Polytechnic Institute and State University}
\smallskip
\leftline{\hlv TYLER H. CHANG}
\smallskip
\leftline{\hlv Argonne National Laboratory}
\smallskip
\leftline{\hlv WILLIAM I. THACKER}
\smallskip
\leftline{\hlv Winthrop University}
\bigskip

\hrule\bigskip\smallskip \footnote{}{\hskip
-\parindent{\parindent=0pt\rmVIII
This work was supported by National Science Foundation Grants CNS-1565314,
CNS-1838271, and DGE-154362.
Authors' addresses:
T. C. H. Lux, L. T. Watson, Departments of Computer Science, Mathematics,
and Aerospace and Ocean Engineering, Virginia Polytechnic Institute \&
State University, Blacksburg, VA 24061; e-mails: {\ttVIII tchlux@vt.edu,
ltwatson@computer.org}; T. H. Chang, Mathematics and Computer Science
Division, Argonne National Laboratory, 9700 South Cass Avenue, Bldg. 240,
Lemont, IL 60439; e-mail: {\ttVIII thchang@vt.edu}; W. I. Thacker, Winthrop
University, Rock Hill, SC 29733; {\ttVIII thackerw@winthrop.edu}.

\xpar
% Permission to make digital/hard copy of
%part or all of this work for personal or classroom use is granted
%without fee provided that the copies are not made or distributed for
%profit or commercial advantage, the copyright notice, the title of
%the publication, and its date appear, and notice is given that
%copying is by permission of the ACM, Inc. To copy otherwise, to
%republish, to post on servers, or to redistribute to lists, requires
%specific permission and/or fee. \xpar\copyright\ 2018 by the
%Association for Computing Machinery, Inc. \xpar 
}}

% Abstract and Introduction
{\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
monotone quintic spline interpolants to monotone data. Using sharp
theoretical monotonicity constraints, first and second derivative estimates
at data provided by a quadratic facet model are refined to produce a
$\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
details, complexity and sensitivity analyses, usage information, and a
brief performance study are included.

\medskip
Categories and Subject Descriptors: G.1.1 [{\bfVIII Numerical Analysis}]:
Interpolation --- Spline and piecewise polynomial interpolation;
J.2 [{\bfVIII Computer Applications}]: Physical Science and Engineering
--- {\itVIII Mathematics};
G.4 [{\bfVIII Mathematics of Computing}]: Mathematical Software

\medskip
General Terms: Algorithms, Monotonicity, Spline

\medskip
Additional Key Words and Phrases: quintic spline, interpolation

}

%% Section 1
\input 1-introduction

%% Section 2
\input 2-monotone-quintic

%% Section 3
\input 3-spline-rep

%% Section 4
\input 4-analysis

%% Section 5
\input 5-applications

%% Acknowledgements
%% \heading{Acknowledgments}
%% This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.

% ====================================================================
%                            Bibliography
% \vfil\goodbreak
\heading{\hlvVIII BIBLIOGRAPHY} {
\baselineskip=2pt%\parskip=100mm
\frenchspacing
% Temporary 'caps', 'sl', and 'rm'.
\let\caps=\capsVIII
\let\sl=\slVIII
\let\rm=\rmeight \rmVIII

% \ys makes a half-em horizontal space.
\def\ys{\hskip 1em minus .5em}

%\refj is for regular journal articles
%  #1 -- Author(s)
%  #2 -- Year
%  #3 -- Title
%  #4 -- Journal
%  #5 -- Volume, number, and pages
\def\refj#1#2#3#4#5{\noindent\hangindent=10pt\hangafter=1
{\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\vskip 6pt minus 3pt}

%\refb is for books
%  #1 -- Author(s)
%  #2 -- Year
%  #3 -- Title
%  #4 -- Publisher, City, State
\def\refb#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
{\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\vskip 6pt minus 3pt}

%\reft is for tech. reports, theses, and conferences
%  #1 -- Author(s)
%  #2 -- Year
%  #3 -- Title
%  #4 -- Technical report ID, Publisher, City, State
\def\reft#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
{\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\vskip 6pt minus 3pt}

%% -------------------------------------------------------------------

% berglund2009planning
\refj{Berglund, T., Brodnik, A., Jonsson, H., Staffanson, M., and
  Soderkvist, I.}{2009}{Planning smooth and obstacle-avoiding B-spline
  paths for autonomous mining vehicles}{IEEE Transactions on
  Automation Science and Engineering}{7(1):167--172}

% brennan2019measure
\refj{Brennan, A.}{2020}{Measure, modulation and metadesign: NC
  fabrication in industrial design and architecture}{Journal of Design
  History}{33(1):66--82}

% deboor1978practical
\refb{de Boor, C.}{1978}{A Practical Guide to Splines}{Springer, Verlag,
New York}

% cameron2019moana
\refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U.,
Bernard, J., Jearls, C., Lux,  T. C. H., Hong, Y., Watson, L. T., and Butt,
A. R.}{2019}{MOANA: modeling and analyzing I/O variability in parallel
system experimental design}{IEEE Transactions on Parallel and Distributed
Systems}{30(8):1843--1856}

% fritsch1980monotone
\refj{Fritsch., F and Carlson., R.}{1980}{Monotone piecewise cubic
  interpolation}{SIAM Journal on Numerical Analysis}{17(2):238--246}

% fritsch1982LLNL
\reft{Fritsch, F. N.}{1982}{Piecewise cubic Hermite interpolation
  package, final specifications}{Computer Documentation UCID-30194,
  Lawrence Livermore National Laboratory}

% gregory1985shape
\reft{Gregory, J. A.}{1985}{Shape preserving spline
  interpolation}{19850020252, Brunel University}

% haralick1981facet
\refj{Haralick, R. M., and Watson, L. T.}{1981}{A facet model for
  image data.}{Computer Graphics and Image Processing}{15(2):113--129}

% he1998monotone
\refj{He, X. and Shi, P.}{1998}{Monotone B-spline smoothing}{Journal
  of the American Statistical Association}{93(442):643--650}

% herman2015techniques
\reft{Herman, D. L., and Oftedal, M. J.}{2015}{Techniques and
  workflows for computer graphics animation systems}{US Patent
  9,216,351}

% hess1994positive
\refj{He{\ss}, W. and Schmidt, J. W.}{1994}{Positive quartic, monotone
  quintic $\scriptstyle C^2$-spline interpolation in one and two
  dimensions}{Journal of Computational and Applied Mathematics}{55(1):51--67}

% knott2012interpolating
\reft{Knott, G. D.}{2012}{Interpolating cubic splines}{Springer Science
  \& Business Media, Volume 18}

% leitenstorfer2006generalized
\refj{Leitenstorfer, F. and Tutz, G.}{2006}{Generalized monotonic
  regression based on B-splines with an application to air pollution
  data}{Biostatistics}{8(3):654--673}

% lux2020thesis
\reft{Lux, T. C. H.}{2020}{Interpolants and error bounds for modeling
  and predicting variability in computer systems}{Thesis, Virginia
  Polytechnic Institute and State University, Blacksburg, Virginia}

% piah2011improved
\refj{Piah, A. R. M. and Unsworth, K.}{2011}{Improved sufficient
  conditions for monotonic piecewise rational quartic
  interpolation}{Sains Malaysiana}{40(10):1173--1178}

% quint2003scalable
\refj{Quint, A.}{2003}{Scalable vector graphics}{IEEE
  MultiMedia}{10(3):99--102}

% ramsay1988monotone
\refj{Ramsay, J. O.}{1988}{Monotone regression splines in
  action}{Statistical Science}{3(4):425--441}

% schmidt1988positivity
\refj{Schmidt, J. W. and He{\ss}, W.}{1988}{Positivity of cubic
  polynomials on intervals and positive spline interpolation}{BIT
  Numerical Mathematics}{28(2):340--352}

% ulrich1994positivity
\refj{Ulrich, G. and Watson, L. T.}{1994}{Positivity conditions for
  quartic polynomials}{SIAM Journal on Scientific
  Computing}{15(3):528--544}

% wang2004rational
\refj{Wang, Q. and Tan, J.}{2004}{Rational quartic spline involving
  shape parameters}{Journal of Information and Computational
  Science}{1(1):127--130}

% yao2018unconditionally
\refj{Yao, J. and Nelson, K. E.}{2018}{An unconditionally monotone
  $\scriptstyle C^2$ quartic spline method with nonoscillation
  derivatives}{Advances in Pure Mathematics}{8(LLNL-JRNL-742107)}

}
%% -------------------------------------------------------------------

\vfill\eject\end

### 2020-08-22 [2020-08]_ACMTOMS_(MQSI)/paper/3-spline-rep.tex
MQSI paper, making final changes.

\heading{3. SPLINE REPRESENTATION}
The monotone quintic spline interpolant $Q(x)$ is represented in terms of
a B-spline basis. The routine {\tt FIT\_SPLINE} in this package computes
the B-spline coefficients $\alpha_i$ of $Q(x)=\sum_{i=1}^{3n} \alpha_i
B_{i,6,t}(x)$ to match the piecewise quintic polynomial values and (first
two) derivatives at the breakpoints $X_i$, where the spline order is six
and the knot sequence $t$ has the breakpoint multiplicities (6, 3, $\ldots$,
3, 6).  The routine {\tt EVAL\_SPLINE} evaluates a spline represented in
terms of a B-spline basis. A Fortran 2003 implementation {\tt EVAL\_BSPLINE}
of the B-spline recurrence relation evaluation code by C. de Boor [1978]
for the value, derivatives, and integral of a B-spline is also provided.

### 2020-08-22 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
MQSI paper, making final changes.

\heading{4. COMPLEXITY AND SENSITIVITY}
Algorithms 1, 2, and 4 have ${\cal O}(1)$ runtime. Given a fixed
schedule for shrinking derivative values, Algorithm 3 has a ${\cal
  O}(n)$ runtime for $n$ data points. In execution, the majority of
the time, still ${\cal O}(n)$, is spent solving the banded linear
system of equations for the B-spline coefficients. Thus for $n$ data
points, the overall execution time is ${\cal O}(n)$. The quadratic
facet model produces a unique sensitivity to input perturbation, as
small changes in input may cause different quadratic facets to be
associated with a breakpoint, and thus different initial derivative
estimates can be produced. This phenomenon is depicted in Figure
1. Despite this sensitivity, the quadratic facet model is still
preferred because it generally provides aesthetically pleasing (low
{\it wiggle}) initial estimates of the first and second derivatives at
the breakpoints while perfectly capturing local low-order phenomena
(e.g., a run of points on a straight line) in the data. The binary
search for a point near the monotone boundary in $(\tau_1$, $\alpha$,
$\beta$, $\gamma)$ space is preferred because it results in monotone
quintic spline interpolants that are nearer to initial estimates than
a policy that strictly shrinks derivative values.

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
{\everymath={\scriptstyle}
\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
  facet model's sensitivity to small data perturbations. This example is
  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
  curvature. Given the exact five data points seen above, the quadratic
  facet model produces the slope seen in the solid blue line at $x = 5/2$.
  However, by subtracting the value of $f_3$ $= \epsilon(x-2)^2$ from
  points at $x = 3$, 4, where $\epsilon$ is the machine precision
  ($2^{-52}$ for an IEEE 64-bit real), the quadratic facet model produces
  the slope seen in the dashed red line at $x = 5/2$. This is the nature
  of a facet model and a side effect of associating data with local facets.
\par} \endinsert

### 2020-08-22 [2020-08]_ACMTOMS_(MQSI)/fortran/EVAL_BSPLINE.f90

@@ -1,0 +1,247 @@
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                        EVAL_BSPLINE.f90
+! 
+! DESCRIPTION:
+!   This file defines a subroutine EVAL_BSPLINE for computing the
+!   value, integral, or derivative(s) of a B-spline given its knot
+!   sequence.
+! 
+! CONTAINS:
+!   SUBROUTINE EVAL_BSPLINE(T, XY, D)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
+!     INTEGER, INTENT(IN), OPTIONAL :: D
+!   END SUBROUTINE EVAL_BSPLINE
+! 
+! EXTERNAL DEPENDENCIES:
+!   MODULE REAL_PRECISION
+!     INTEGER, PARAMETER :: R8
+!   END MODULE REAL_PRECISION
+! 
+! CONTRIBUTORS:
+!   Thomas C.H. Lux (tchlux@vt.edu)
+!   Layne T. Watson (ltwatson@computer.org)
+!   William I. Thacker (thackerw@winthrop.edu)
+! 
+! VERSION HISTORY:
+!   June 2020 -- (tchl) Created file, (ltw / wit) reviewed and revised.
+! 
+SUBROUTINE EVAL_BSPLINE(T, XY, D)
+! Subroutine for evaluating a B-spline with provided knot sequence, T.
+! W A R N I N G : This routine has NO ERROR CHECKING and assumes
+! informed usage for speed. It has undefined behavior for input that
+! doesn't match specifications.
+! 
+! INPUT:
+!   T(1:N) -- The nondecreasing sequence of N knots for the B-spline.
+! 
+! INPUT / OUTPUT:
+!   XY(1:M) -- On input, the locations at which the B-spline is evaluated;
+!     on output, holds the value of the Dth derivative of the B-spline
+!     with prescribed knots evaluated at the given locations, or the
+!     integral of the B-spline over [T(1), XY(.)] in XY(.).
+! 
+! OPTIONAL INPUT:
+!   D --  The order of the derivative to take of the B-spline. If omitted,
[…]

### 2020-08-22 [2018-10]_(metric_PCA)/paper.tex
Found issue in last table where accuracies were flipped, should have been one minus the values shown.

--- 2020-01-16 [2019-12-16]_HPC_(mpca)/paper.tex
@@
     \multirow{5}{*}{\textit{CIFAR10}}
-    & Delaunay & PCA & 31 & \underline{0.631} \\
-    & KNN (1) & MPCA & 31 & 0.681 \\
-    & KNN (10) & MPCA & 31 & 0.643 \\
-    & LSHEP & PCA & 31 & 0.650 \\
-    & MLPRegressor & raw & 3072 & 0.897 \\
+    & Delaunay & PCA & 31 & 0.369 \\
+    & KNN (1) & MPCA & 31 & 0.319 \\
+    & KNN (10) & MPCA & 31 & 0.357 \\
+    & LSHEP & PCA & 31 & 0.350 \\
+    & MLPRegressor & raw & 3072 & \underline{0.103} \\
     \hline

### 2020-08-22 [2019-11]_HPC_(quintic_spline)/paper/paper.tex
Deleted unnecessary files, renamed some things, cleaning up mostly.

@@
 Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by
-\begin{align*}
+\begin{equation}\begin{aligned}
 h_i &= k_{i+1} - k_{i}, \quad
@@
 p\big((x-k_i)/h_i\big).
-\end{align*}
+\end{aligned}\end{equation}
 Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$.
@@
   \centering
+  \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
+Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
+  \label{table:e3_results}
   \begin{tabular}{c|c c c|c c c|c c c}
@@
   \end{tabular}
-  \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
-Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
-  \label{table:e3_results}
 \end{table}
@@
 \textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has
-interests in numerical analysis, mathematical programming,
-bioinformatics, and data science.  He has been involved with the
-organization of HPCS since 2000.
+interests in numerical analysis, mathematical programming, and data
+science.  He has been involved with the organization of HPCS since
+2000. \email{ltw@vt.edu}
 
 \textbf{\uppercase{Tyler H. Chang}} is a Ph.D. candidate at VPI\&SU
-studying computer science under Dr. Layne Watson.
+studying under Dr. Layne Watson. \email{thchang@vt.edu}
 
 \textbf{\uppercase{Li Xu}} is a Ph.D. candidate at VPI\&SU studying
-statistics under Dr. Yili Hong.
+statistics under Dr. Yili Hong. \email{lix1992@vt.edu}
 
 \textbf{\uppercase{Yueyao Wang}} is a Ph.D. candidate at VPI\&SU
-studying statistics under Dr. Yili Hong.
+studying under Dr. Yili Hong. \email{yueyao94@vt.edu}
 
 \textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) is an
-associate professor of Statistics at VPI\&SU.
+associate professor of Statistics at VPI\&SU. \email{yilihong@vt.edu}
 

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
Initial response to TOMS editor.

@@
 (e.g., a run of points on a straight line) in the data. The binary
-search for a point near the monotone boundary in $(\tau_1$, $\alpha$,
+search for a point on the monotone boundary in $(\tau_1$, $\alpha$,
 $\beta$, $\gamma)$ space is preferred because it results in monotone

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/code/MQSI/EVAL_BSPLINE.f90

@@ -1,0 +1,1 @@
+/Users/thomaslux/Git/papers/[2020-08]_ACMTOMS_(MQSI)/code/EVAL_BSPLINE.f90

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/paper/response/response.tex
Initial response to TOMS editor.

%% General paragraph settings (no indent, big line skip).
\parindent=0pt
\parskip=12pt

%% Fonts (for title)
\font\titlebf=cmbx20
\font\titlett=cmtt20

%% Title
{\titlebf Response regarding {\titlett TOMS-2020-0061} \bigskip}

%% Body
Dr. Tim Hopkins,

After carefully reading and considering the comments from the reviewer, we would like to request additional reviews. Addressed individually below, we fear many of the comments made by the reviewer do not appreciably assess this work, nor its relevance to the community. Each numbered item and italicized text block below is a comment provided by the reviewer. Our responses immediately follow.

{\parindent=20pt \it

\item{(1)} A very similar algorithm for monotone quintic spline interpolation has already been published and detailed in pseudo-code:

\itemitem{} Thomas C.H. Lux, Layne T. Watson, et al: An Algorithm for Constructing Monotone Quintic Interpolating Splines. Proceedings of the 2020 Spring Simulation Conference, May 2020, Article No. 33, Pages 1-12.

\item{} The differences with respect to the current manuscript seem to be rather specific and minor, and at least I would have expected a discussion/comparison between the current approach and the one in the above published work.

}

That conference paper was an earlier version of this work and may appear similar at first glance, but there are many critical differences between the two (computations refactored for numerical stability, different method for estimating derivatives, using a binary search to find feasible derivative values for the entire spline rather than a single interval). Notably, there was no code published with that conference paper, which is one of the most important components of this submission.

{\parindent=20pt \it

\item{(2)} The overall procedure for computing a monotone spline interpolant is rather basic and quite standard. The only special ingredient is the use of a binary search in the last step, but the same idea has also been used in the paper mentioned in (1).

}

Whereas the overall procedure for computing a monotone spline is rather basic and standard for splines of lower order than quintic, there is no published algorithm with software (that we are aware of) that creates a piecewise quintic monotone spline interpolant using theoretically tight constraints. The submitted work resembles the methodology for constructing lower order spline interpolants because there are similar theoretical constructs. However, this work is distinctly different and more capable because the resulting monotone spline is quintic (twice continuously differentiable). A quick look at the theory involved in checking quintic monotonicity (cf. Ulrich and Watson) shows it is significantly more complex than the cubic theory, which also explains why no high quality quintic code has been produced before.

The binary search referred to in (1) is used to make a single interval of the interpolating spline monotone. That is distinctly different than the binary search in the present submission that searches for derivative values where {\it all} intervals are monotone. The only similarity between the papers in that regard is that they both utilize a binary search at some point.

{\parindent=20pt \it

\item{(3)} The literature study on monotone spline interpolation is incomplete and misses important work. As far as I am aware of, one of the most commonly used methods for shape-preserving interpolation is the quadratic spline method by Larry Schumaker (it does not only preserves monotonicity but also convexity!), at least in the spline approximation theory community:

\itemitem{} Larry L. Schumaker: On Shape Preserving Quadratic Spline Interpolation. SIAM Journal on Numerical Analysis, Vol. 20, No. 4 (1983), Pages 854-864.

}

The submission cites quadratic, cubic, and quartic previous works. The work by Larry Schumaker can readily be included in this list, and has been included in the attached update to the paper. It is our understanding that one of the most widely used shape-preserving spline algorithms is a cubic spline methodology proposed by Fritsch and Carlson (known as PCHIP), which is included in many mathematical software libraries and cited in the submission.

{\parindent=20pt \it

\item{(4)} The numerical section just shows few examples of data sets and the obtained spline results. The authors make some general statements about the algorithm like ``it generally provides aesthetically pleasing (low wiggle) initial estimates ... is preferred because it results in monotone quintic spline interpolants that are nearer to initial estimates'' but a deeper (more objective) discussion is lacking. Also a numerical comparison with other classical monotone spline methods is missing, at least the algorithm by Schumaker [1983] and also the one by Fritz/Carlson [1980]. This might help understand why the new method is of interest.

}

This is a reasonable criticism, but also easily addressed with minor revisions to the submission. While the reviewer may disagree with the language we used, there are objective reasons for the decisions made in this work in addition to the subjective comments made. Specifically, the quadratic facet model picks the minimum curvature local estimate for second derivatives which has the smallest magnitude second derivative, and the binary search identifies derivative values absolutely nearer to the initial estimates than the previous method of shrinking derivative values used in (1).

Comparisons with lower order methods can be included if greatly desired, noting that the comparisons will only be aesthetic because there is no published mathematical software that also constructs a monotone quintic spline interpolant. These comparisons were not originally included because subjective decisions must be made about what visuals to show and those decisions might draw criticisms that distract from the unique purpose and achievement of this work (producing $C^2$ monotonicity preserving approximations to data).

{\parindent=20pt \it

\item{(5)} The most time-consuming part of the algorithm is the conversion to the B-spline representation. This has been done by solving a global linear system. It would be more efficient when using a local quasi-interpolation spline scheme with projection property.

\item{} Moreover, spline evaluation by explicitly computing the linear combination of B-splines is not efficient; instead use de Boor-Cox algorithm.

}

The reviewer may have missed the reason for the global linear system, from the submission (section 5, paragraph 4), ``It would be faster to construct splines over intervals independently ... the single linear system is chosen here for the decreased redundancy''. To be clear, the de Boor-Cox algorithm {\it is} used to evaluate the underlying B-splines (section 5, paragraph 1). The linear combination in the submission is of the (quintic) B-splines that define the interpolant.

\bigskip

To the best of our knowledge, rejection based on a single review has not been a journal policy and we respectfully request additional reviews. We believe this algorithm and the accompanying code has great potential for TOMS, noting that the cubic work preceding this has remained state-of-the-art for forty years. Thank you for both your time and consideration.

Sincerely,

Thomas Lux, on behalf of all authors.

\bye

%% ----------------------------------------------------------------------
%%                        Useful TeX reference.
%% 
%% https://www.math.brown.edu/johsilve/ReferenceCards/TeXRefCard.v1.5.pdf
%% 
%% ----------------------------------------------------------------------

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
Initial response to TOMS editor.

@@
 The currently available software for monotone piecewise polynomial
-interpolation includes quadratic [He and Shi 1998], cubic [Fritsch and
-Carlson 1980], and (with limited application) quartic [Wang and Tan
-2004; Piah and Unsworth 2011; Yao and Nelson 2018] cases. In
-addition, a statistical method for bootstrapping the construction of
-an arbitrarily smooth monotone fit exists [Leitenstorfer and Tutz
-2006], but the method does not take advantage of known analytic
-properties of quintic polynomials. The code by Fritsch [1982] for
-$C^1$ cubic spline interpolation is the predominantly utilized code
-for constructing monotone interpolants at present. Theory has been
-provided for the quintic case [Ulrich and Watson 1994; He{\ss} and
-Schmidt 1994] and that theory was recently utilized in a proposed
-algorithm [Lux 2020] for monotone quintic spline construction, however
-no published mathematical software exists.
+interpolation includes quadratic [He and Shi 1998, McAllister and 
+Roulier 1981], cubic [Fritsch and Carlson 1980], and (with limited
+application) quartic [Wang and Tan 2004; Piah and Unsworth 2011; Yao
+and Nelson 2018] cases. In addition, a statistical method for
+bootstrapping the construction of an arbitrarily smooth monotone fit
+exists [Leitenstorfer and Tutz 2006], but the method does not take
+advantage of known analytic properties of quintic polynomials. 
+Conversely the work on quadratic shape preserving splines by Schumaker
+[1983] is widely utilized in theory for its conservation of both
+monotonicity and convexity, but a mathematical software is not readily
+available.
+
+The code by Fritsch [1982] for $C^1$ cubic spline interpolation is the
+predominantly utilized code for constructing monotone interpolants at
+present. Theory has been provided for the quintic case [Ulrich and
+Watson 1994; He{\ss} and Schmidt 1994] and that theory was recently
+utilized in a proposed algorithm [Lux 2020] for monotone quintic
+spline construction, however no published mathematical software
+exists. This work improves upon the algorithm presented by Lux et al.
+[2020] by refactoring computations for improved numerical stability,
+estimating minimum curvature derivatives at breakpoints with a
+quadratic facet model, and using a binary search to reduce the
+magnitude of the modifications made to initial derivative estimates
+when constructing a monotone spline interpolant.
 

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/code/MQSI/MQSI_c_wrapper.f90

@@ -1,0 +1,78 @@
+! This automatically generated Fortran wrapper file allows codes
+! written in Fortran to be called directly from C and translates all
+! C-style arguments into expected Fortran-style arguments (with
+! assumed size, local type declarations, etc.).
+
+
+SUBROUTINE C_MQSI(X_DIM_1, X, Y_DIM_1, Y, T_DIM_1, T, BCOEF_DIM_1, BCOEF, INFO, UV_PRESENT, UV_DIM_1, UV_DIM_2, UV) BIND(C)
+  USE REAL_PRECISION , ONLY : R8
+  IMPLICIT NONE
+  INTEGER, INTENT(IN) :: X_DIM_1
+  REAL(KIND=R8), INTENT(IN), DIMENSION(X_DIM_1) :: X
+  INTEGER, INTENT(IN) :: Y_DIM_1
+  REAL(KIND=R8), INTENT(INOUT), DIMENSION(Y_DIM_1) :: Y
+  INTEGER, INTENT(IN) :: T_DIM_1
+  REAL(KIND=R8), INTENT(OUT), DIMENSION(T_DIM_1) :: T
+  INTEGER, INTENT(IN) :: BCOEF_DIM_1
+  REAL(KIND=R8), INTENT(OUT), DIMENSION(BCOEF_DIM_1) :: BCOEF
+  INTEGER, INTENT(OUT) :: INFO
+  LOGICAL, INTENT(IN) :: UV_PRESENT
+  INTEGER, INTENT(IN) :: UV_DIM_1
+  INTEGER, INTENT(IN) :: UV_DIM_2
+  REAL(KIND=R8), INTENT(OUT), DIMENSION(UV_DIM_1,UV_DIM_2) :: UV
+
+  INTERFACE
+    SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
+      ! Computes a monotone quintic spline interpolant (MQSI), Q(X), to data
+      ! in terms of spline coefficients BCOEF for a B-spline basis defined by
+      ! knots T. Q(X) is theoretically guaranteed to be monotone increasing
+      ! (decreasing) over exactly the same intervals [X(I), X(J)] that the
+      ! data Y(.) is monotone increasing (decreasing).
+      !
+      ! INPUT:
+      !   X(1:ND) -- A real array of ND increasing values.
+      !   Y(1:ND) -- A real array of ND function values Y(I) = F(X(I)).
+      !
+      ! OUTPUT:
[…]

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/code/MQSI/__init__.py

@@ -1,0 +1,1 @@
+./MQSI_python_wrapper.py

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/code/MQSI/sample_main.f90

@@ -1,0 +1,1 @@
+/Users/thomaslux/Git/papers/[2020-08]_ACMTOMS_(MQSI)/code/sample_main.f90

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/code/SPLINE/MQSI.f90

@@ -1,0 +1,1 @@
+/Users/thomaslux/Git/papers/[2020-08]_ACMTOMS_(MQSI)/code/MQSI.f90

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/code/SPLINE/SPLINE_python_wrapper.py

@@ -1,0 +1,263 @@
+'''This Python code is an automatically generated wrapper
+for Fortran code made by 'fmodpy'. The original documentation
+for the Fortran source code follows.
+
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                           SPLINE.f90
+!
+! DESCRIPTION:
+!   This file defines the subroutines FIT_SPLINE for computing the
+!   coefficients of a B-spline basis necessary to reproduce given
+!   function and derivative values, and EVAL_SPLINE for evaluating the
+!   value, integral, and derivatives of a spline defined in terms of
+!   its B-spline basis.
+!
+! CONTAINS:
+!   SUBROUTINE FIT_SPLINE(XI, FX, T, BCOEF, INFO)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:)   :: XI
+!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:,:) :: FX
+!     REAL(KIND=R8), INTENT(OUT), DIMENSION(:)   :: T, BCOEF
+!     INTEGER, INTENT(OUT) :: INFO
+!   END SUBROUTINE FIT_SPLINE
+!
+!   SUBROUTINE EVAL_SPLINE(T, BCOEF, XY, INFO, D)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T, BCOEF
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
+!     INTEGER, INTENT(OUT) :: INFO
+!     INTEGER, INTENT(IN), OPTIONAL :: D
+!   END SUBROUTINE EVAL_SPLINE
+!
+! EXTERNAL DEPENDENCIES:
+!   MODULE REAL_PRECISION
+!     INTEGER, PARAMETER :: R8
+!   END MODULE REAL_PRECISION
+!
+!   SUBROUTINE EVAL_BSPLINE(T, XY, D)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
+!     INTEGER, INTENT(IN), OPTIONAL :: D
[…]

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/code/SPLINE/blas.f

@@ -1,0 +1,1 @@
+/Users/thomaslux/Git/papers/[2020-08]_ACMTOMS_(MQSI)/code/blas.f

### 2021-02-01 [2020-08]_ACMTOMS_(MQSI)/code/SPLINE/test_all.f90

@@ -1,0 +1,1 @@
+/Users/thomaslux/Git/papers/[2020-08]_ACMTOMS_(MQSI)/code/test_all.f90

### 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
Updated paper to responde to coauthor comments. Will use the updated diff with master to finish revising 'response.tex' in upcoming commits.

@@
 spline construction, however no published mathematical software
-exists. This work improves upon the algorithm presented by Lux et al.
-[2020] by refactoring computations for improved numerical stability,
-estimating minimum curvature derivatives at breakpoints with a
-quadratic facet model, and using a binary search to reduce the
-magnitude of the modifications made to initial derivative estimates
-when constructing a monotone spline interpolant.
+exists. The software presented here represents the first published
+software package for producing $C^2$ shape preserving splines. This
+work improves upon the algorithm presented by Lux et al. [2020] by
+refactoring computations for improved numerical stability, estimating
+minimum curvature derivatives at breakpoints with a quadratic facet
+model, and using a binary search to reduce the magnitude of the
+modifications made to initial derivative estimates when constructing a
+monotone spline interpolant.
 
-The importance of piecewise quintic interpolation over lower order
-approximations can be simply observed. In general, the order of a
-polynomial determines the number of function (and derivative) values
-it can interpolate, which in turn determines the growth rate of error
-away from interpolated values. $C^2$ quintic (order six) splines
-match the function value and two given derivatives at each breakpoint.
 This work provides a Fortran 2003 subroutine {\tt MQSI} based on the

### 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Updated paper to responde to coauthor comments. Will use the updated diff with master to finish revising 'response.tex' in upcoming commits.

@@
 
-Fritsch and Carlson [1980] originally proposed the use of central differences
-to estimate derivatives, however this often leads to extra and unnecessary
-{\it wiggles} in the spline when used to approximate second derivatives.
-In an attempt to capture the local shape of the data, this package uses a
-facet model from image processing [Haralick and Watson 1981] to estimate
-first and second derivatives at breakpoints. Rather than picking a local
-linear or quadratic fit with minimal residual, this work uses a quadratic
+Fritsch and Carlson [1980] originally proposed the use of central
+differences to estimate derivatives, however this often leads to extra
+and unnecessary {\it wiggles} in the spline when used to approximate
+second derivatives. Modern shape-preserving spline implementations use
+a weighted harmonic mean to estimate derivative values at breakpoints
+[Moler 2008], however this method also yields approximations with
+large {\it wiggle}, or whose second derivative functions have large
+$L2$ norm over the domain. In an attempt to capture the local shape of
+the data while minimizing wiggle, this package uses a facet model from
+image processing [Haralick and Watson 1981] to estimate first and
+second derivatives at breakpoints. Rather than picking a local linear
+or quadratic fit with minimal residual, this work uses a quadratic
 facet model that selects the local quadratic interpolant with minimum

### 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
Next round of response done, sending to coauthors.

@@
 system of equations for the B-spline coefficients. Thus for $n$ data
-points, the overall execution time is ${\cal O}(n)$. The quadratic
-facet model produces a unique sensitivity to input perturbation, as
-small changes in input may cause different quadratic facets to be
-associated with a breakpoint, and thus different initial derivative
-estimates can be produced. This phenomenon is depicted in Figure
-1. Despite this sensitivity, the quadratic facet model is still
-preferred because it generally provides aesthetically pleasing (low
-{\it wiggle}) initial estimates of the first and second derivatives at
-the breakpoints while perfectly capturing local low-order phenomena
-(e.g., a run of points on a straight line) in the data. The binary
-search for a point on the monotone boundary in $(\tau_1$, $\alpha$,
-$\beta$, $\gamma)$ space is preferred because it results in monotone
-quintic spline interpolants that are nearer to initial estimates than
-a policy that strictly shrinks derivative values.
+points, the overall execution time is ${\cal O}(n)$.
+
+The quadratic facet model produces a unique sensitivity to input
+perturbation, as small changes in input may cause different quadratic
+facets to be associated with a breakpoint, and thus different initial
+derivative estimates can be produced. This phenomenon is depicted in
+Figure 1. Despite this sensitivity, the quadratic facet model is still
+preferred because it exactly captures local linear and quadratic
+behavior while empirically producing final approximations with less
+wiggle ($L2$ norm of the second derivative) than other methods. A
+weighted harmonic mean estimate of first derivatives may be more
+accurate when the underlying function changes at a rate greater than a
+quadratic, but that method increases the sensitivity to small
+perturbations in data and empirically results in quintic splines with
+greater wiggle.
+
+The binary search for a point on the monotone boundary in $(\tau_1$,
+$\alpha$, $\beta$, $\gamma)$ space is performed because it results in
+monotone quintic spline interpolants with derivative values that are
+absolutely nearer to initial estimates than a search that strictly
+shrinks derivative values. Given the initial derivative estimates have
+desirable properties (capture low-order phenomena and are low wiggle),
+this search results in an approximation that is both monotone and has
+derivative values similar to the initial estimates.
 

### 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/response/response.tex
Next round of response done, sending to coauthors.

@@
 Notably, there was no code published with that conference paper, which is one of the most important components of this submission.
+To make these differences more clear, two sentences were added to the end of paragraph 4 in section 1.
 
@@
 The submission cites quadratic, cubic, and quartic previous works.
-The work by Larry Schumaker can readily be included in this list, and has been included in the attached update to the paper.
+The work by Larry Schumaker can readily be included in this list, and has been included in the attached update to the paper (section 1, paragraph 3).
 It is our understanding that one of the most widely used shape-preserving spline algorithms is a cubic spline methodology proposed by Fritsch and Carlson (known as PCHIP), which is included in many mathematical software libraries and cited in the submission.
@@
 Specifically, the quadratic facet model picks the minimum curvature local estimate for second derivatives which has the smallest magnitude second derivative, and the binary search identifies derivative values absolutely nearer to the initial estimates than the previous method of shrinking derivative values used in (1).
+These facts are now mentioned in the paper (section 4, paragraphs 2 and 3).
 
-Comparisons with lower order methods can be included if greatly desired, noting that the comparisons will only be aesthetic because there is no published mathematical software that also constructs a monotone quintic spline interpolant.
-These comparisons were not originally included because subjective decisions must be made about what visuals to show and those decisions might draw criticisms that distract from the unique purpose and achievement of this work (producing $C^2$ monotonicity preserving approximations to data).
+Comparisons with lower order methods are not appropriate for this work.
+There is no published mathematical software that also constructs a monotone quintic spline interpolant to data.
+If a user desires a monotonicity preserving $C^2$ approximation, this would be the only published software that achieves that goal.
+This paper is not intended to be a review and comparison of all available shape preserving spline packages.
+Including such comparisons would also require subjective decisions about what visuals to show, and those decisions might draw criticisms that distract from the unique purpose and achievement of this work (producing $C^2$ monotonicity preserving approximations to data).
+To emphasize that this work is not a replacement for previous works, but rather an additional method to consider, the first sentence of section 1, paragraph 5 that mentioned the importance of this work ``over'' previous methods was removed.
 
@@
 To the best of our knowledge, rejection based on a single review has not been a journal policy and we respectfully request additional reviews.
-We believe this algorithm and the accompanying code has great potential for TOMS, noting that the cubic work preceding this has remained state-of-the-art for forty years.
+We believe this algorithm and the accompanying code has great potential for TOMS, noting that the quadratic and cubic works preceding this have remained state-of-the-art for forty years.
 Thank you for both your time and consideration.

### 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Next round of response done, sending to coauthors.

@@
 a weighted harmonic mean to estimate derivative values at breakpoints
-[Moler 2008], however this method also yields approximations with
-large {\it wiggle}, or whose second derivative functions have large
-$L2$ norm over the domain. In an attempt to capture the local shape of
+[Moler 2008], however this method also yields approximations whose
+second derivative functions often have large $L2$ norm (approximations
+with large {\it wiggle}). In an attempt to capture the local shape of
 the data while minimizing wiggle, this package uses a facet model from

### 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/response/response.tex
Incorporating some most recent comments from Dr. Watson and Thacker

--- 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/response/response.tex
@@
 \parindent=0pt
-\parskip=12pt
+\parskip=9pt
 
@@
 The submission cites quadratic, cubic, and quartic previous works.
-The work by Larry Schumaker can readily be included in this list, and has been included in the attached update to the paper (section 1, paragraph 3).
+The work by Larry Schumaker is now referenced in the attached update to the paper (section 1, paragraph 3).
 It is our understanding that one of the most widely used shape-preserving spline algorithms is a cubic spline methodology proposed by Fritsch and Carlson (known as PCHIP), which is included in many mathematical software libraries and cited in the submission.
@@
 To emphasize that this work is not a replacement for previous works, but rather an additional method to consider, the first sentence of section 1, paragraph 5 that mentioned the importance of this work ``over'' previous methods was removed.
+The inherent difficulty of comparing this work with existing lower order methods is now mentioned in section 5, paragraph 5.
 
@@
 the single linear system is chosen here for the decreased redundancy''.
+Comments in the paper and the Fortran source code indicated an option for returning the derivatives at the data points, so one can use either the piecewise polynomial or the B-spline representations.
+Further, the time for the banded system solve is $O(n)$, reasonable for $n$ data points.
 To be clear, the de Boor-Cox algorithm {\it is} used to evaluate the underlying B-splines (section 5, paragraph 1).

### 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Incorporating some most recent comments from Dr. Watson and Thacker

--- 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
@@
 a weighted harmonic mean to estimate derivative values at breakpoints
-[Moler 2008], however this method also yields approximations with
-large {\it wiggle}, or whose second derivative functions have large
-$L2$ norm over the domain. In an attempt to capture the local shape of
+[Moler 2008], however this method also yields approximations whose
+second derivative functions often have large $L^2$ norm (approximations
+with large {\it wiggle}). In an attempt to capture the local shape of
 the data while minimizing wiggle, this package uses a facet model from
@@
 {\parindent 0mm
-{\bf Algorithm 1:} {
-\tt QUADRATIC\_FACET}$\bigl(X(1{:}n)$, $Y(1{:}n)$, $i \bigr)$
+{\bf Algorithm 1:}
+{\tt QUADRATIC\_FACET}$\bigl(X(1{:}n)$, $Y(1{:}n)$, $i \bigr)$
 
+\nobreak
 where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
@@
 {\bf Algorithm 2:}
-
 {\tt IS\_MONOTONE}$\bigl(x_0$, $x_1$, $f(x_0)$, $Df(x_0)$,
@@
 
+\nobreak
 where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
@@
 {\parindent 0mm
-{\bf Algorithm 3:} {\tt MQSI}$\bigl(X(1{:}n), Y(1{:}n) \bigr)$
+{\bf Algorithm 3:}
+{\tt MQSI}$\bigl(X(1{:}n), Y(1{:}n) \bigr)$
 
+\nobreak
 where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are data
@@
 %%              Algorithm 4: Global binary search
-\vskip 5mm {\parindent 0mm {\bf Algorithm 4:} {\tt
-    BINARY\_SEARCH}$\bigl(X(1{:}n), Y(1{:}n), u(1{:}n), v(1{:}n) \bigr)$
+\vskip 5mm {\parindent 0mm {\bf Algorithm 4:}
+{\tt BINARY\_SEARCH}$\bigl(X(1{:}n), Y(1{:}n), u(1{:}n), v(1{:}n) \bigr)$
 
+\nobreak
 where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are

### 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
Incorporating some most recent comments from Dr. Watson and Thacker

@@
 behavior while empirically producing final approximations with less
-wiggle ($L2$ norm of the second derivative) than other methods. A
+wiggle ($L^2$ norm of the second derivative) than other methods. A
 weighted harmonic mean estimate of first derivatives may be more
 accurate when the underlying function changes at a rate greater than a
-quadratic, but that method increases the sensitivity to small
-perturbations in data and empirically results in quintic splines with
-greater wiggle.
+quadratic, but that method increases the second derivative sensitivity
+to small perturbations in data and empirically results in quintic
+splines with greater wiggle.
 
@@
 absolutely nearer to initial estimates than a search that strictly
-shrinks derivative values. Given the initial derivative estimates have
-desirable properties (capture low-order phenomena and are low wiggle),
-this search results in an approximation that is both monotone and has
-derivative values similar to the initial estimates.
+shrinks derivative values. Given that the initial derivative estimates
+have desirable properties (capture low-order phenomena and are low
+wiggle), this search results in an approximation that is both monotone
+and has derivative values similar to the initial estimates.
 

### 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
Incorporating some most recent comments from Dr. Watson and Thacker

@@
 polynomial (piece of $Q(x)$) over each interval.
+
+Comparisons between this work and lower order spline algorithms are
+problematic, especially with sparse data. For exact data from smooth
+functions, comparing derivative values, higher order methods always win
+as the data density increases. This result is demonstrated in the work of
+Lux et al. [2020]. For sparse or synthetic data, what constitutes a
+better fit is either subjective (e.g., less wiggle) or dependent on the
+specific problem. If an application demands $C^2$ continuity and
+monotonicity, the target of {\tt MQSI}, then this software provides such
+solutions while no existing published spline algorithms suffice.

### 2021-03-19 [2020-08]_ACMTOMS_(MQSI)/code/SPLINE.f90
Fixed two numerical instabilities in the code related to large numbers of points. Updated based on minor feedback for response and paper.

@@ -166,8 +166,8 @@ END DO
 ! Assign the last knot to exist a small step outside the supported
 ! interval to ensure the B-spline basis functions are nonzero at the
 ! rightmost breakpoint.
-T(NK-DEGREE:NK) = MAX( XI(NB) + ABS(XI(NB))*SQRT(EPSILON(XI(NB))),  &
-                       XI(NB) + SQRT(EPSILON(XI(NB))) )
+T(NK-DEGREE:NK) = MAX( XI(NB) + ABS(XI(NB))*SQRT(SQRT(EPSILON(XI(NB)))),  &
+                       XI(NB) + SQRT(SQRT(EPSILON(XI(NB)))) )
 
 ! The next block of code evaluates each B-spline and it's derivatives
 ! at all breakpoints. The first and last elements of XI will be

### 2021-04-22 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex

@@
 [1983] is widely utilized in theory for its conservation of both
-monotonicity and convexity, but a mathematical software is not readily
-available.
+monotonicity and convexity, but no mathematical software is published.
 

### 2021-04-22 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex

--- 2020-08-22 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
@@
 polynomial (piece of $Q(x)$) over each interval.
+
+Comparisons between this work and lower order methods are problematic,
+especially with sparse data. For exact data from smooth functions,
+comparing derivative values, higher order methods always win (all
+other things being equal) as the data density increases. For sparse or
+synthetic data, what constitutes a better fit is either subjective or
+dependent on the specific problem. If an application demands $C^2$
+continuity and monotonicity, the target of {\tt MQSI}, then this
+package provides such interpolants while no existing published spline
+software suffices.

### 2021-04-22 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Final modifications to ACM TOMS paper before submission.
Initial response to TOMS editor.

@@
 % and William Thacker
-% ACM Trans. Math. Software, ?/?/2020.
+% ACM Trans. Math. Software, ?/?/202?.
 % 
@@
 
+%% 10002950.10003705
+%% Mathematics of computing~Mathematical software
+%% 
+%% 10002950.10003648.10003649.10003657.10003659
+%% Mathematics of computing~Spline models
+%% 
+%% 10002950.10003714.10003715.10003722
+%% Mathematics of computing~Interpolation
+%% 
+%% 10002950.10003714.10003715
+%% Mathematics of computing~Numerical analysis
+%% 
+%% 10002950.10003648.10003649.10003657
+%% Mathematics of computing~Nonparametric representations
+
 \medskip
@@
 
+% mcallister1981algorithm
+\refj{McAllister, D. F. and Roulier, J. A.}{1981}{An algorithm for 
+  computing a shape-preserving osculatory quadratic spline}{ACM 
+  Transactions on Mathematical Software (TOMS)}{7(3):331--347}
+
+% mcallister1981shape
+\refj{McAllister, D. F. and Roulier, J. A.}{1981}{Algorithm 574:
+  Shape-Preserving Osculatory Quadratic Splines [E1, E2]}{ACM
+  Transactions on Mathematical Software (TOMS)}{7(3):384--386}
+
+% moler2004numerical
+\refb{Moler, C. B.}{2008}{Numerical Computing with {\caps Matlab}}{
+  The MathWorks Inc., Natick, Massachusetts}
+
 % piah2011improved
@@
   Numerical Mathematics}{28(2):340--352}
+
+% schumaker1983shape
+\refj{Schumaker, Larry L.}{1983}{On shape preserving quadratic spline
+  interpolation}{SIAM Journal on Numerical Analysis}{20(4):854--864}
 

### 2022-03-07 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Working on revision. Added references to papers, making summaries. Initialized response document.

@@
 {\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
-monotone quintic spline interpolants to monotone data. Using sharp
+monotone quintic spline interpolants to univariate monotone data. Using sharp
 theoretical monotonicity constraints, first and second derivative estimates
 at data provided by a quadratic facet model are refined to produce a
-$\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
-details, complexity and sensitivity analyses, usage information, and a
-brief performance study are included.
+univariate $\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
+details, complexity and sensitivity analyses, usage information, a
+brief performance study, and comparisons with other spline approaches are included.
 
@@
 }
+
+\bigskip\hrule\bigskip
 
@@
 {\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\vskip 6pt minus 3pt}
+
+%% No extra knots, rational functions, sufficiency and necessity proven, works for unequally spaced data, no code associated with the paper, no algorithm provided for the initialization for free shape parameters.
+\refj{Abbas M., Majid, A. A. and Ali, J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for monotone data}{Applied Mathematics and Computation}{219(6):2885--2895}
+
+%% 
+\refj{Beatson, R. K.}{1982}{Monotone and convex approximation by splines: error estimates and a curve fitting algorithm}{SIAM Journal on Numerical Analysis}{19(6):1278--1285}
+
+%% Proves existence of splines that maintain monotonicity and convexity while having fixed continuity and order.
+\refj{Costantini, P.}{1986}{On monotone and convex spline interpolation}{Mathematics of Computation}{46(173):203--214}
+
+\refj{Cravero, I., C. Manni}{2003}{Shape-preserving interpolants with high smoothness}{Journal of computational and applied mathematics}{157, 383--405}
+
+\refj{Delbourgo, R., Gregory,  J. A.}{1983}{$C^2$ rational quadratic spline interpolation to monotonic data}{IMA J. Numer. Anal.}{3, 141--152}
+
+\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in tension}{SIAM J. Numer. Anal.}{30, 595--607}
+
+\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM J. Math. Anal.}{8, 891-905}
+
+\refj{DeVore, R. A.}{1977}{Monotone approximation by polynomials}{SIAM J. Math. Anal.}{8, 906-921}
+
+\refj{Dougherty, R. L., Edelman, A., Hyman, J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving cubic and quintic Hermite interpolation}{Math. Comp.}{52, 471--494}
+
+\refj{Gregory, J. A.,Delbourgo,  R.}{1982}{Piecewise rational quadratic interpolation to monotonic data}{IMA J. Numer. Anal.}{2, 123--130}
+
+\refj{Fiorot, J. C., Tabka, J.}{1991}{Shape-preserving $C^2$ cubic polynomial interpolating splines}{Math. Comp.}{57, 291--298}
+
+\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation with higher order continuity}{Applied Mathematics and Computation}{337, 1-13}
+
+\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM J. Numer. Anal.}{30, 57--100}
+
+\refj{Kvasov, B. I.}{2014}{Monotone and convex interpolation by weighted quadratic splines}{Adv. Comput. Math.}{40, 91-116}
+
+\refj{Manni, C., Sablonni\`ere, P.}{1997}{Monotone interpolation of order 3 by $C^2$ cubic splines}{IMA J. Numer.Anal.}{17, 305--320}
+
+\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite interpolation}{BIT Numerical Mathematics}{41, 127--148}
+
+\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline interpolation}{IMA J. Numer. Anal.}{13, 493--507}
+
+\refj{Zhu, Y.,Han, X.}{2015}{$C^2$ rational quartic interpolation spline with local shape preserving property}{Appl. Math. Lett.}{46, 57-63}
+
+\refj{Zhu, Y.,Han, X.}{2015}{Shape preserving $C^2$ rational quartic interpolation spline with two parameters}{International Journal of Computer Mathematics}{92, 2160-2177}
 

### 2022-03-07 [2020-08]_ACMTOMS_(MQSI)/revision/test.py

@@ -41,7 +41,7 @@ def toms_574_spline(x, y, pts=1000):
             py = eval_quadratic(px)
             f = fit_polynomial(px, py)
             return f.derivative()(z)
-    def deriv_2(z, s=.01):
+    def deriv_2(z, s=.001):
         try: return [deriv_2(v) for v in z]
         except:
             if (min(abs(z - x_min), abs(z - x_max)) <= s): return 0.0
@@ -75,8 +75,9 @@ def schumaker_spline(x, y):
     # Import the "schumaker" package from R.
     schumaker = importr('schumaker')
     # Calculate the response for the given x with R
+    edge_gradient = py2rpy(np.asarray([0.0, 0.0]))
     Spline, DerivativeSpline, SecondDerivativeSpline, IntervalTab = (
-        schumaker.Schumaker(py2rpy(x), py2rpy(y), Vectorised=True)
+        schumaker.Schumaker(py2rpy(x), py2rpy(y), edgeGradients=edge_gradient, Vectorised=True)
     )
     # Construct python functions for evaluating the fit spline.
     def second_derivative(x, ddf=SecondDerivativeSpline):
@@ -175,7 +176,8 @@ def bvspis_spline(x, y):
                                y2tab, erre, work, nwork)
         (x, y, np, n, k, xtab, ntab, sbopt, y0opt, y1opt, y2opt, 
          errc, d, d2, y0tab, y1tab, y2tab, erre, work, nwork) = output
-        assert (erre == 0), f"The DBVSSC routine reported a nonzero error {erre}. Check the documentation for more information about ERRE."
+        assert (erre == 0), f"The DBVSSC routine reported a nonzero error ERRE = {erre}. Check the documentation for more information about ERRE."
+        assert (errc == 0), f"The DBVSSC routine reported a nonzero error ERRC = {errc}. Check the documentation for more information about ERRC."
         return y0tab
     # First derivative evaluation.
     def df(xtab, x=x, y=y, np=np, n=n, k=k, sbopt=sbopt,
@@ -194,7 +196,8 @@ def bvspis_spline(x, y):
                                y2tab, erre, work, nwork)
         (x, y, np, n, k, xtab, ntab, sbopt, y0opt, y1opt, y2opt, 
          errc, d, d2, y0tab, y1tab, y2tab, erre, work, nwork) = output
[…]

### 2022-03-07 [2020-08]_ACMTOMS_(MQSI)/revision/response.tex
Working on revision. Added references to papers, making summaries. Initialized response document.

%% General paragraph settings (no indent, big line skip).
\parindent=0pt
\parskip=9pt

%% Fonts (for title)
\font\titlebf=cmbx20
\font\titlett=cmtt20

%% Title
{\titlebf Revision of {\titlett TOMS-2021-0028} \bigskip}

%% Body
Dear reviewers,

We are deeply grateful for the comments, questions, and references you have all provided. In accordance with your feedback we have made significant changes to the manuscript to improve clarity, expand the summary of related literature, and incorporate visual comparisons with existing spline software. We hope you find these changes have resulted in an overall better manuscript.

Each numbered item and italicized text block below is a reviewer comment (or summary of similar comments). Our responses immediately follow.

{\parindent=20pt \it

\item{(1)} The authors should include references to more related works and justify the differences between this work and the many prior works on $C^2$ interpolation. [given 23 references]

}

In order to properly satisfy this request, we've added a new subsection to the introduction that summarizes the individual tracks of research represented by the conglomerate of references provided. After reviewing all of those works we are still quite certain that this manuscript represents a novel and meaningful contribution to the literature, and have justified it's position amongst those works in subsection 1.1 of the revised manuscript.

{\parindent=20pt \it

\item{(2)} The authors should include references to and comparisons with the following software packages.

\itemitem{} Costantini, P. An algorithm for computing shape-preserving interpolating splines
of arbitrary degree, Journal of Computational and Applied Mathematics, 1988, 22(1), pp. 89-136.

\item{} While the concerns about subjectivity in comparisons are acknowledged, users will still want to know the context of this work and the trade-offs would they use this package instead of other spline softwares. For this reason a comparison with other methods including visuals should be provided.

}

We appreciate this perspective and have included such a comparison in the updated Section 5. The comparison shows methods from ...

{\parindent=20pt \it

\item{()} The monotonicity of Q(x) is not proved.

}

Per Section {\bf blank}, the theory proving the monotonicity of the resulting spline is quite dense and is formally covered by the cited references Ulrich and Watson, and {\bf blank} and {\bf blank}.

{\parindent=20pt \it

\item{()} The approximation order of the algorithm is not shown.

}

A new sentence has been added to Section {\bf blank} explaining that the approximation order of the method achieves second order convergence in almost all scenarios where the approximated function is twice continuously differentiable.

{\parindent=20pt \it

\item{()} Why use piecewise degree 5? In principle, piecewise degree 3 is enough for $C^2$. The additional degrees of freedom for making the interpolant monotone could be obtained by adding knots between the breakpoints. It would be great if you could motivate your choice in more detail.

}

In general it is not possible to form a $C^2$ approximation of arbitrary data without the incorporation of additional knots. The restriction of a fixed set of knots is somewhat arbitrary, but that is the application targeted by this line of research.

{\parindent=20pt \it

\item{()} Some statements depend on two variables being ``approximately equal'', but what does that mean? Equal up to some small threshold?

}

The definition of approximately equal for this work is when two IEEE 64-bit floating point numbers are equal up to 32 of 42 mantissa bits (precision bits, or significand in base two scientific notation).

{\parindent=20pt \it

\item{()} In Algorithm 1 (and also in the text), you use the term "curvature" to refer to the second derivative. I would avoid this, since "curvature" is a term reserved for curves, but you are dealing with functions. Similarly, the term "minimum curvature derivative" is inappropriate.

}

This has been rephrased to reduce confusion.

{\parindent=20pt \it

\item{()} In Algorithm 2, it might be better to say that f is a "degree 5" polynomial, instead of "order 6".

}

Noted and changed.

{\parindent=20pt \it

\item{()} Your binary search results in a particular set of derivatives, for which the local quintic piece is guaranteed to be monotone, but it is clearly not a unique choice. Alternatively, one could try to find, among all possible sets of derivatives and corresponding monotone quintic interpolants, the best, according to some criterion, e.g. least L2 norm. This would turn the problem into a constrained optimization problem, which is probably harder to solve, but it might, at the same time, give better results and overcome the limitation shown in Fig. 1. More motivating details would be appreciated.

}

This was a point of our own severe consideration before submitting the manuscript and these comments reflect our thoughts. In proper response to this we implemented naive $L2$ minimization algorithm that uses gradient descent over the spline representation to enforce local monotonicity constraints in a Lagrangian fashion (with an adaptive multiplier on monotonicity violations). The results we achieved were quite mixed and we've incorporated our commentary from this excursion into Section 4 of the manuscript. In truth, formally expanding on this trajectory would require an entirely new manuscript with deeper study of the problem, while it is not immediately obvious that the results would be {\it better} than the present approach. Included in this response (to reviewers only) are some of the preliminary visuals we produced in exploring this approach.

{\parindent=20pt \it

\item{()} It is quite probable that a user would want to convert the piecewise Hermite representation of the interpolant to quintic B-spline form. But the authors say in Sections 3 and 4 that they solve a banded linear system of length n (n the number of data points) to make this conversion. However, this is not necessary. At each data point, only three of the quintic B-splines are non-zero. Therefore, the three associated B-spline coefficients can be computed from the value, first, and second derivatives of the interpolant at that single point. Thus one can just solve a small 3*3 linear system to get one group of 3 B-spline coefficients. Well, it requires a bit of algebraic work to write out this 3*3 linear system, but it can be done.

}

As mentioned in the manuscript, the banded linear system representation is favored for two reasons in this package. The complexity order is unchanged by solving a single banded system, while the numerical stability is {\it theoretically} improved. In practice it is almost certainly slower than solving individual linear systems, but the increased stability and decreased redundancy in representation of the current approach is favored here. For users interested in constructing alternative representations, the optional argument {\it \bf UV} exists precisely for returning the computed first and second derivatives at knots.

{\parindent=20pt \it

\item{()} I'm confused by the O(1) and O(n) discussion in Section 4.
If there are n data points then surely the main algorithm (estimating first and second
derivatives at each data point) must require at least an O(n) algorithm, not O(1).

}

This was poorly worded on our part. The entire algorithm is O(n), however the binary search itself has a fixed number of loop iterations. The original intent of the statement was that the complexity order of the outer loop (the binary search) is constant, which ensures the overall complexity of O(n) is not changed. See the new phrasing in Section {\bf BLANK} paragraph {\bf BLANK}.

{\parindent=20pt \it

\item{()} The output are values -- why not the spline coefficients? If the output are values one would hope for a visualization package.

}

Response

{\parindent=20pt \it

\item{()} I was not familiar with the term/notion  "quadratic facet model". The term facet made me believe that the paper was on bivariate interpolation. Nothing in the abstract or title prevents this misconception.

}

\bigskip

To the best of our knowledge, rejection based on a single review has not been a journal policy and we respectfully request additional reviews. We believe this algorithm and the accompanying code has great potential for TOMS, noting that the quadratic and cubic works preceding this have remained state-of-the-art for forty years. Thank you for both your time and consideration.

Sincerely,

Thomas Lux, on behalf of all authors.

\bye

%% ----------------------------------------------------------------------
%%                        Useful TeX reference.
%% 
%% https://www.math.brown.edu/johsilve/ReferenceCards/TeXRefCard.v1.5.pdf
%% 
%% ----------------------------------------------------------------------

### 2022-03-07 [2020-08]_ACMTOMS_(MQSI)/visuals/cli.f90

@@ -1,0 +1,155 @@
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                             cli.f90
+! 
+! DESCRIPTION:
+!   This file defines a command line interface program to the MQSI
+!   package. The command line interface accepts data (to build a
+!   monotone quintic spline interpolant) and points (to approximate
+!   with MQSI) in files and outputs approximated values at points
+!   to a file. Compile with:
+! 
+!    $F03 $OPTS REAL_PRECISION.f90 EVAL_BSPLINE.f90 SPLINE.f90 MQSI.f90 \
+!       cli.f90 -o cli $LIB
+!
+!   where '$F03' is the name of the Fortran 2003 compiler, '$OPTS' are
+!   compiler options such as '-O3', and '$LIB' provides a flag to link
+!   BLAS and LAPACK. If the BLAS and LAPACK libraries are not
+!   available on your system, then replace $LIB with the filenames
+!   'blas.f lapack.f'; these files contain the routines from the BLAS
+!   and LAPACK libraries that are necessary for this package. Running
+!   the program with no arguments causes usage information to be
+!   printed to standard output.
+! 
+! CONTAINS:
+!   PROGRAM CLI
+! 
+! DEPENDENCIES:
+!   MODULE REAL_PRECISION
+!     INTEGER, PARAMETER :: R8
+!   END MODULE REAL_PRECISION
+! 
+!   SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: X
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: Y
+!     REAL(KIND=R8), INTENT(OUT),   DIMENSION(:) :: T, BCOEF
+!     INTEGER, INTENT(OUT) :: INFO
+!     REAL(KIND=R8), INTENT(OUT), DIMENSION(:,:), OPTIONAL :: UV
+!   END SUBROUTINE MQSI
+! 
+!   SUBROUTINE EVAL_SPLINE(T, BCOEF, XY, INFO, D)
+!     USE REAL_PRECISION, ONLY: R8
[…]

### 2022-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
Working on revision. Added references to papers, making summaries. Initialized response document.
Added summaries of all references. Still reworking the introduction section.

@@
-\bigskip\hrule\bigskip\medskip
 \heading{1. INTRODUCTION}
@@
 
-The currently available software for monotone piecewise polynomial
-interpolation includes quadratic [He and Shi 1998, McAllister and
-Roulier 1981], cubic [Fritsch and Carlson 1980], and (with limited
-application) quartic [Wang and Tan 2004; Piah and Unsworth 2011; Yao
-and Nelson 2018] cases. In addition, a statistical method for
-bootstrapping the construction of an arbitrarily smooth monotone fit
-exists [Leitenstorfer and Tutz 2006], but the method does not take
-advantage of known analytic properties of quintic polynomials.
-Conversely the work on quadratic shape preserving splines by Schumaker
-[1983] is widely utilized in theory for its conservation of both
-monotonicity and convexity, but no mathematical software is published.
+\heading{1.1 Related Works}
+
+[DeVore 1977] Proves approximation error bounds for monotone spline interpolation of data from monotone functions and accompanies the previous error bounds in the context of monotone polynomial interpolation of data from a monotone function.
+
+[Beatson 1982] develops a quadratic spline curve fitting algorithm that maintains local convexity while also proving some error bounds.
+This method avoids inserting additional knots, but does not interpolate the data.
+
+[Gregory and Delbourgo 1982] Proves necessary and sufficient conditions for a closed form solution to monotone C1 quadratic rational spline interpolation.
+Does not provide formal algorithm, nor code.
+Then again iterates towards a solution of nonlinear equations that ensure C2 continuity of quadratic rational splines.
+The solution they iterate towards exists and is unique.
+They provide proofs and methods for setting derivatives at the ends, no formal algorithms or code in [Delbourgo and Gregory 1983].
+
+[Costantini 1986] Proves existence of splines that maintain monotonicity and convexity while having fixed continuity and order.
+
+[Dougherty, Edelman, and Hyman 1989] Proposes an algorithm for constructing C2 Hermite monotone spline interpolants using proven sufficient, but not necessary conditions.
+As these bounds are not sharp, the space of allowable derivative and second derivative values at knots is unnecessarily shrunk.
+An algorithm is provided, but no code.
+Recommends against the direct optimization for smoothness parameters like L2, instead promoting problem specific definitions of 'geometric niceness'.
+
+Fiorot and Tabka 1991] Proves a simple method for determining the existence of a monotone C2 cubic interpolating spline, but functions of this kind do not always exist for arbitrary convex monotone data sets.
+
+[Delbourgo 1993] Continuous previous work by incorporating a cubic numerator, proves shape preservation when the correct value for a tension parameter is set but no method for setting that parameter.
+Accuracy depends on accurate derivative evaluations.
+No code nor formal algorithm is provided.
+
+[Huynh 1993] Presents several nonlinear boundary conditions based on the median function to establish a fourth order monotonicity preserving constraints relying on divided difference initial derivative estimates.
+Overall generates a monotone C1 cubic interpolant to data.
+
+[Pruess 1993] Proposes a cubic C2 shape preserving spline method with free parameters (mentions quintic splines are necessary for globally C2 monotone splines).
+Sufficient (but not necessary) conditions are provided for monotonicity.
+
+[Manni 1997] Constructs a C2 cubic spline with two additional knots per interval.
+Then [Cravero and Manni 2003] Constructs C3 interpolating splines by progressively increasing tension through Bezier control polygons until shape parameters are satisfied, no code is provided.
+
+[Abbas, Majid, and Ali 2012] No extra knots, rational functions, sufficiency and necessity proven, works for unequally spaced data, no code associated with the paper, no algorithm provided for the initialization for free shape parameters.
+
+[Wang and Tan 2004] Propose a sufficient quartic C2 method with two free parameters.
+
+[Piah and Unsworth 2011] improve upon the sufficient conditions of Wang and Tan for C2 quartic rational interpolation.
+
+[Kvasov 2014] use sufficient conditions for monotonicity and convexity to form a globally weighted combination of C1 quadratic splines that satisfy shape constraints.
+
+[Zhu and Han 2015] uses sufficient conditions over three free parameters to construct C2 rational quartic shape preserving splines.
+Finally [Han 2018] Proposes a method for making arbitrarily smooth shape preserving spline interpolants, but uses a centered quadratic to estimate derivatives and a sufficient set of bounding conditions to ensure shape preservation in practice.
+No necessary conditions are applied, no algorithm is provided, neither is any code provided.
+
+[Yao and Nelson 2018] Develop novel sufficient conditions for monotone C2 quartic spline interpolation and maintain shape with added control points.
+
+\heading{1.2 Available Software}
+
+The currently available peer reviewed and published software for
+monotone piecewise polynomial interpolation is severely lacking in
+comparison with the number of published methodologies mentioned
+above. This is indicative of how difficult it is to properly handle
+the numerical conditions that arise when constructing and evaluating
+precise shape preserving splines.
+
+The piecewise quadratic $C^1$ method of [McAllister and Roulier 1981]
+is available through TOMS, as well as the sufficient $C^2$ quintic
+method of BVSPIS. The sufficient $C^1$ cubic spline method of [Fritsch and Carlson 1980],
+is available through the SciPy Python package, and the $C^1$ rational
+quadratic method of Schumaker [1983] is available as an R package.
+
+cases. In addition, a statistical method for bootstrapping the
+construction of an arbitrarily smooth monotone fit exists
+[Leitenstorfer and Tutz 2006], but the method does not take advantage
+of known analytic properties of quintic polynomials.  Conversely the
+work on quadratic shape preserving splines by Schumaker [1983] is
+widely utilized in theory for its conservation of both monotonicity
+and convexity, but no mathematical software is published.
 
@@
 monotone spline interpolant.
+
+\heading{1.3 Overview}
 

### 2022-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Added summaries of all references. Still reworking the introduction section.

@@
 
-%% No extra knots, rational functions, sufficiency and necessity proven, works for unequally spaced data, no code associated with the paper, no algorithm provided for the initialization for free shape parameters.
-\refj{Abbas M., Majid, A. A. and Ali, J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for monotone data}{Applied Mathematics and Computation}{219(6):2885--2895}
+%% Proves approximation error bounds for monotone spline interpolation of data from monotone functions.
+\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM Journal on Mathematical Analysis}{8(5):891--905}
 
-%% 
+%% Accompanies the previous error bounds in the context of monotone polynomial interpolation of data from a monotone function.
+\refj{DeVore, R. A.}{1977}{Monotone approximation by polynomials}{SIAM Journal on Mathematical Analysis}{8(5):906--921}
+
+%% Develops a quadratic spline curve fitting algorithm that maintains local convexity while also proving some error bounds.
+This method avoids inserting additional knots, but does not interpolate the data.
 \refj{Beatson, R. K.}{1982}{Monotone and convex approximation by splines: error estimates and a curve fitting algorithm}{SIAM Journal on Numerical Analysis}{19(6):1278--1285}
+
+%% Proves necessary and sufficient conditions for a closed form solution to monotone C1 quadratic rational spline interpolation. Does not provide formal algorithm, nor code.
+\refj{Gregory , J. A.,Delbourgo,  R.}{1982}{Piecewise rational quadratic interpolation to monotonic data}{IMA Journal of Numerical Analysis}{2(2):123--130}
+
+%% Iterates towards a solution of nonlinear equations that ensure C2 continuity of quadratic rational splines.
+The solution they iterate towards exists and is unique.
+They provide proofs and methods for setting derivatives at the ends, no formal algorithms or code.
+\refj{Delbourgo, R., Gregory, J. A.}{1983}{$C^2$ rational quadratic spline interpolation to monotonic data}{IMA Journal of Numerical Analysis}{3(2):141--152}
 
@@
 
-\refj{Cravero, I., C. Manni}{2003}{Shape-preserving interpolants with high smoothness}{Journal of computational and applied mathematics}{157, 383--405}
+%% Proposes an algorithm for constructing C2 Hermite monotone spline interpolants using proven sufficient, but not necessary conditions.
+As these bounds are not sharp, the space of allowable derivative and second derivative values at knots is unnecessarily shrunk.
+An algorithm is provided, but no code.
+Recommends against the direct optimization for smoothness parameters like L2, instead promoting problem specific definitions of 'geometric niceness'.
+\refj{Dougherty, R. L., Edelman, A., Hyman, J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving cubic and quintic Hermite interpolation}{Mathematics of Computation}{52(186):471--494}
 
-\refj{Delbourgo, R., Gregory,  J. A.}{1983}{$C^2$ rational quadratic spline interpolation to monotonic data}{IMA J. Numer. Anal.}{3, 141--152}
+%% Proves a simple method for determining the existence of a monotone C2 cubic interpolating spline, but functions of this kind do not always exist for arbitrary convex monotone data sets.
+\refj{Fiorot, J. C., Tabka, J.}{1991}{Shape-preserving $C^2$ cubic polynomial interpolating splines}{Mathematics of Computation}{57(195):291--298}
 
-\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in tension}{SIAM J. Numer. Anal.}{30, 595--607}
+%% Continuous previous work by incorporating a cubic numerator, proves shape preservation when the correct value for a tension parameter is set but no method for setting that parameter.
+Accuracy depends on accurate derivative evaluations.
+No code nor formal algorithm is provided.
+\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}
 
-\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM J. Math. Anal.}{8, 891-905}
+%% Presents several nonlinear boundary conditions based on the median function to establish a fourth order monotonicity preserving constraints relying on divided difference initial derivative estimates.
+Overall generates a monotone C1 cubic interpolant to data.
+\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM Journal of Numerical Analysis}{30(1):57--100}
 
-\refj{DeVore, R. A.}{1977}{Monotone approximation by polynomials}{SIAM J. Math. Anal.}{8, 906-921}
+%% Proposes a cubic C2 shape preserving spline method with free parameters (mentions quintic splines are necessary for globally C2 monotone splines).
+Sufficient (but not necessary) conditions are provided for monotonicity.
+\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}
 
-\refj{Dougherty, R. L., Edelman, A., Hyman, J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving cubic and quintic Hermite interpolation}{Math. Comp.}{52, 471--494}
+%% Constructs a C2 cubic spline with two additional knots per interval. No code (paywall).
+\refj{Manni, C., Sablonni\`ere, P.}{1997}{Monotone interpolation of order 3 by $C^2$ cubic splines}{IMA Journal of Numerical Analysis}{17(2):305--320}
 
-\refj{Gregory, J. A.,Delbourgo,  R.}{1982}{Piecewise rational quadratic interpolation to monotonic data}{IMA J. Numer. Anal.}{2, 123--130}
+%% Uses sufficient conditions to generate C2 cubic spline. No code (paywall).
+\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite interpolation}{BIT Numerical Mathematics}{41(1):127--148}
 
-\refj{Fiorot, J. C., Tabka, J.}{1991}{Shape-preserving $C^2$ cubic polynomial interpolating splines}{Math. Comp.}{57, 291--298}
+%% Constructs C3 interpolating splines by progressively increasing tension through Bezier control polygons until shape parameters are satisfied, no code is provided.
+\refj{Cravero, I., C. Manni}{2003}{Shape-preserving interpolants with high smoothness}{Journal of computational and applied mathematics}{157(2):383--405}
 
-\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation with higher order continuity}{Applied Mathematics and Computation}{337, 1-13}
+%% No extra knots, rational functions, sufficiency and necessity proven, works for unequally spaced data, no code associated with the paper, no algorithm provided for the initialization for free shape parameters.
+\refj{Abbas M., Majid, A. A. and Ali, J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for monotone data}{Applied Mathematics and Computation}{219(6):2885--2895}
 
-\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM J. Numer. Anal.}{30, 57--100}
+%% Use sufficient conditions for monotonicity and convexity to form a globally weighted combination of C1 quadratic splines that satisfy shape constraints.
+\refj{Kvasov, B. I.}{2014}{Monotone and convex interpolation by weighted quadratic splines}{Advances in Computational Mathematics}{40(1):91--116}
 
-\refj{Kvasov, B. I.}{2014}{Monotone and convex interpolation by weighted quadratic splines}{Adv. Comput. Math.}{40, 91-116}
+%% Uses sufficient conditions over three free parameters to construct C2 rational quartic shape preserving splines.
+\refj{Zhu, Y.,Han, X.}{2015}{$C^2$ rational quartic interpolation spline with local shape preserving property}{Applied Mathematics Letters}{46:57--63}
 
-\refj{Manni, C., Sablonni\`ere, P.}{1997}{Monotone interpolation of order 3 by $C^2$ cubic splines}{IMA J. Numer.Anal.}{17, 305--320}
+%% Uses sufficient conditions over two free parameters to construct C2 rational quartic shape preserving splines.
+\refj{Zhu, Y.,Han, X.}{2015}{Shape preserving $C^2$ rational quartic interpolation spline with two parameters}{International Journal of Computer Mathematics}{92(10):2160--2177}
 
-\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite interpolation}{BIT Numerical Mathematics}{41, 127--148}
-
-\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline interpolation}{IMA J. Numer. Anal.}{13, 493--507}
-
-\refj{Zhu, Y.,Han, X.}{2015}{$C^2$ rational quartic interpolation spline with local shape preserving property}{Appl. Math. Lett.}{46, 57-63}
-
-\refj{Zhu, Y.,Han, X.}{2015}{Shape preserving $C^2$ rational quartic interpolation spline with two parameters}{International Journal of Computer Mathematics}{92, 2160-2177}
+%% Proposes a method for making arbitrarily smooth shape preserving spline interpolants, but uses a centered quadratic to estimate derivatives and a sufficient set of bounding conditions to ensure shape preservation in practice.
+No necessary conditions are applied, no algorithm is provided, neither is any code provided.
+\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation with higher order continuity}{Applied Mathematics and Computation}{337:1--13}
 
@@
   interpolation}{19850020252, Brunel University}
-
-% haralick1981facet
-\refj{Haralick, R. M., and Watson, L. T.}{1981}{A facet model for
-  image data.}{Computer Graphics and Image Processing}{15(2):113--129}
-
-% he1998monotone
-\refj{He, X. and Shi, P.}{1998}{Monotone B-spline smoothing}{Journal
-  of the American Statistical Association}{93(442):643--650}
 

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/response.tex
Finishing revision draft, few more red highlights to add.

@@
 \parskip=9pt
+\baselineskip=13pt plus 1pt minus 1pt \lineskiplimit=1pt
+
+%% Red coloring.
+\input colordvi
+\def\BEGINC{\textRed}
+\def\ENDC{\textBlack}
+\def\red#1{\BEGINC#1\ENDC}
+\let\beginred=\BEGINC
+\let\endred=\ENDC
 
@@
 \font\titlett=cmtt20
+\font\ttVIII=cmtt8
+
+%% EPS figures.
+\input epsf
 
@@
 
-We are deeply grateful for the comments, questions, and references you have all provided.
-In accordance with your feedback we have made significant changes to the manuscript to improve clarity, expand the summary of related literature, and incorporate visual comparisons with existing spline software.
-We hope you find these changes have resulted in an overall better manuscript.
+We are deeply grateful for the comments, questions, and references you
+have all provided. In accordance with your feedback we have made
+significant changes to the manuscript to improve clarity, expand the
+summary of related literature, and incorporate visual comparisons with
+existing spline software. We hope you find these changes have resulted
+in an overall better manuscript.
 
-Each numbered item and italicized text block below is a reviewer comment (or summary of similar comments). Our responses immediately follow.
+Each numbered item and italicized text block below is a reviewer
+comment (or summary of similar comments). Our responses immediately
+follow.
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{(1)} The authors should include references to more related works and justify the differences between this work and the many prior works on $C^2$ interpolation. [given 23 references]
+\item{(1)} The authors should include references to more related works
+  and justify the differences between this work and the many prior
+  works on $C^2$ interpolation. [given 23 references]
 
@@
 
-In order to properly satisfy this request, we've added a new subsection to the introduction that summarizes the individual tracks of research represented by the conglomerate of references provided.
-After reviewing all of those works we are still quite certain that this manuscript represents a novel and meaningful contribution to the literature, and have justified it's position amongst those works in subsection 1.1 of the revised manuscript.
+In order to properly satisfy this request, we have significantly
+expanded the review of relevant research in the introduction (Section
+1, paragraphs 3-6). This new content summarizes the individual tracks
+represented by the conglomerate of references provided. After
+reviewing all of those works we are still quite certain that this
+manuscript represents a novel and meaningful contribution to the
+literature, and have justified it's position among those works in the
+revised manuscript. The most outstanding contribution of this work
+remains the code, which applies sharp monotonicity constraints (unlike
+any other $C^2$ methods) to ensure a monotone quintic interpolant
+nearest to the initial nonmonotone (but otherwise {\it nice})
+approximation is achievable.
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{(2)} The authors should include references to and comparisons with the following software packages.
+\item{(2)} The authors should include references to and comparisons
+  with the following software package (which appears to be very
+  similar).
 
-\itemitem{} Costantini, P. An algorithm for computing shape-preserving interpolating splines
-of arbitrary degree, Journal of Computational and Applied Mathematics, 1988, 22(1), pp. 89-136.
+\itemitem{} Costantini, P. An algorithm for computing shape-preserving
+  interpolating splines of arbitrary degree, Journal of Computational
+  and Applied Mathematics, 1998, 22(1), pp. 89-136.
 
-\item{} While the concerns about subjectivity in comparisons are acknowledged, users will still want to know the context of this work and the trade-offs would they use this package instead of other spline softwares.
-For this reason a comparison with other methods including visuals should be provided.
+\item{} While the concerns about subjectivity in comparisons are
+  acknowledged, users will still want to know the context of this work
+  and the trade-offs would they use this package instead of other
+  spline algorithms. For this reason a comparison with other methods
+  including visuals should be provided.
 
@@
 
-We appreciate this perspective and have included such a comparison in the updated Section 5. The comparison shows methods from ...
+We appreciate this perspective and have included such comparisons in
+the updated Section 5. The comparison shows methods from four
+published spline software packages against MQSI for the same test
+problems as originally in the paper (and associated test code). An
+important point to note is that we were hesitant to include
+comparisons with the published R package for Schumaker spline
+interpolants, because (as is shown in the revised manuscript) the
+package often produces incorrectly nonmonotone approximations to
+data. In our opinion this undesirable outcome is a valid demonstration
+of the numerical difficulties that arise when writing codes for spline
+algorithms. They are not trivial to handle, and that underscores the
+importance of thoroughly and carefully peer reviewed implementations
+like what is provided by this work.
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} The monotonicity of Q(x) is not proved.
+\item{(3)} The monotonicity of Q(x) is not proved.
 
@@
 
-Per Section {\bf blank}, the theory proving the monotonicity of the resulting spline is quite dense and is formally covered by the cited references Ulrich and Watson, and {\bf blank} and {\bf blank}.
+Per Section 1 paragraph 5, Section 2 paragraphs 3, 4 and 6, and
+Algorithm 2 (all of the original submission) the theory proving the
+sharp necessary and sufficient conditions for monotonicity is provided
+by Ulrich and Watson [1994]. This code conforms precisely with the
+tight boundary conditions referenced therein. The theory is quite
+dense and it would seem inappropriate to reproduce all of that work in
+this manuscript as well. To further clarify that the conditions for
+monotonicity are proven, we have added an additional paragraph to
+Section 5 explaining this fact.
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} The approximation order of the algorithm is not shown.
+\item{(4)} The approximation order of the algorithm is not shown.
 
@@
 
-A new sentence has been added to Section {\bf blank} explaining that the approximation order of the method achieves second order convergence in almost all scenarios where the approximated function is twice continuously differentiable.
+In the same paragraph of Section 5 referenced above, we have explained
+that the approximation order of the method is that of any second order
+approximation (because a local quadratic interpolant is used to
+approximate initial derivative values), but also notes that the
+asymptotic approximation order is often of little importance in the
+context of (sparse) scattered data (the target application of most
+shape preserving techniques, especially those achieving higher levels
+of continuity).
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} Why use piecewise degree 5?
-In principle, piecewise degree 3 is enough for $C^2$.
-The additional degrees of freedom for making the interpolant monotone could be obtained by adding knots between the breakpoints.
-It would be great if you could motivate your choice in more detail.
+\item{(5)} Why use piecewise degree 5? In principle, piecewise degree 3
+  is enough for $C^2$. The additional degrees of freedom for making
+  the interpolant monotone could be obtained by adding knots between
+  the breakpoints. It would be great if you could motivate your choice
+  in more detail.
 
@@
 
-In general it is not possible to form a $C^2$ approximation of arbitrary data without the incorporation of additional knots.
-The restriction of a fixed set of knots is somewhat arbitrary, but that is the application targeted by this line of research.
+As stated in the comment, constructing polynomial pieces that
+interpolate arbitrary function values and two derivatives at the ends
+of an interval requires 6 degrees of freedom and hence an order 6
+polynomial (degree 5, quintic) without adding additional knots. This
+work builds on theory that does {\it not} incorporate new knots, and
+it is our opinion that it is generally favorable to not insert new
+knots. Without the insertion of additional knots, the ability to
+construct lower order monotone $C^2$ splines is restricted to specific
+arrangements of data. The restriction of having only a fixed set of
+knots is somewhat arbitrary, but that is the application targeted by
+this line of research.
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} Some statements depend on two variables being ``approximately equal'', but what does that mean? Equal up to some small threshold?
+\item{(6)} Some statements depend on two variables being
+  ``approximately equal'', but what does that mean? Equal up to some
+  small threshold?
 
@@
 
-The definition of approximately equal for this work is when two IEEE 64-bit floating point numbers are equal up to 32 of 42 mantissa bits (precision bits, or significand in base two scientific notation).
+The definition of approximately equal here is when two floating point
+numbers are within the machine epsilon (at their scale) of each other.
+This is a rather standard approach for ensuring safe separation of
+floating point numbers. We have added a sentence explaining this to
+the pretexts of the algorithms that reference this operation.
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} In Algorithm 1 (and also in the text), you use the term "curvature" to refer to the second derivative.
-I would avoid this, since "curvature" is a term reserved for curves, but you are dealing with functions.
-Similarly, the term "minimum curvature derivative" is inappropriate.
+\item{(7)} In Algorithm 1 (and also in the text), you use the term
+  ``curvature'' to refer to the second derivative. I would avoid this,
+  since ``curvature'' is a term reserved for curves, but you are dealing
+  with functions. Similarly, the term ``minimum curvature derivative''
+  is inappropriate.
 
@@
 
-This has been rephrased to reduce confusion.
+This has been rephrased throughout to reduce confusion.
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} In Algorithm 2, it might be better to say that f is a "degree 5" polynomial, instead of "order 6".
+\item{(8)} In Algorithm 2, it might be better to say that f is a
+  ``degree 5'' polynomial, instead of ``order 6''.
 
@@
 
-\item{()} Your binary search results in a particular set of derivatives, for which the local quintic piece is guaranteed to be monotone, but it is clearly not a unique choice.
-Alternatively, one could try to find, among all possible sets of derivatives and corresponding monotone quintic interpolants, the best, according to some criterion, e.g.
-least L2 norm.
-This would turn the problem into a constrained optimization problem, which is probably harder to solve, but it might, at the same time, give better results and overcome the limitation shown in Fig.
-1.
-More motivating details would be appreciated.
+\item{(9)} Your binary search results in a particular set of
+  derivatives, for which the local quintic piece is guaranteed to be
+  monotone, but it is clearly not a unique choice. Alternatively, one
+  could try to find, among all possible sets of derivatives and
+  corresponding monotone quintic interpolants, the best, according to
+  some criterion, e.g. least L2 norm. This would turn the problem into
+  a constrained optimization problem, which is probably harder to
+  solve, but it might, at the same time, give better results and
+  overcome the limitation shown in Fig. 1. More motivating details
+  would be appreciated.
 
@@
 
-This was a point of our own severe consideration before submitting the manuscript and these comments reflect our thoughts.
-In proper response to this we implemented naive $L2$ minimization algorithm that uses gradient descent over the spline representation to enforce local monotonicity constraints in a Lagrangian fashion (with an adaptive multiplier on monotonicity violations).
-The results we achieved were quite mixed and we've incorporated our commentary from this excursion into Section 4 of the manuscript.
-In truth, formally expanding on this trajectory would require an entirely new manuscript with deeper study of the problem, while it is not immediately obvious that the results would be {\it better} than the present approach.
-Included in this response (to reviewers only) are some of the preliminary visuals we produced in exploring this approach.
+This was a point of our own severe consideration before submitting the
+manuscript and this comment reflected our original thoughts. In due
+response to this we've implemented naive $L2$ minimization algorithm
+over the second derivative that uses gradient descent over the spline
+representation to enforce local monotonicity constraints in a
+Lagrangian fashion (with an adaptive multiplier on monotonicity
+violations). The results we achieved were quite mixed and we've
+incorporated our commentary from this excursion into Section 5 of the
+manuscript. In truth, formally expanding on this trajectory would
+require an entirely new manuscript with deeper study of the problem,
+while it is not immediately obvious that the results would be {\it
+  better} than the present approach. The global $L^2$ minimizer of the
+second derivative is simply the unique $C^2$ cubic spline interpolant,
+but that is known to produce large and unnecessary local variations in
+the approximation (along with not being shape preserving). Included in
+this response (to reviewers only) are some of the preliminary visuals
+we produced in exploring this approach. This is by no means final and
+conclusive, but the results were mixed enough for us to warrant not
+including them in the paper.
 
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/5-l2_minimizer.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/5-l2_minimizer.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/5-l2_minimizer.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/5-l2_minimizer.eps}}
+{\narrower\noindent {\bf $L^2$ Minimization Experiment }
+{\ttVIII MQSI} (light gray) compared with an experimental global $L^2$
+minimization approach (blue) on each of the test problems from the
+paper. It is {\it very} difficult to consistently construct fits that
+lack local oscillations as well as maintain monotonicity with this
+approach. The method relies on the B-spline basis as well, which
+affects the bias towards specific solutions when following the second
+derivative $L^2$ minimizing gradient.  This work would require a very
+significant line of research to properly resolve all these small
+issues in a theoretically sound way.
+\par}
+\vfill \eject
+
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} It is quite probable that a user would want to convert the piecewise Hermite representation of the interpolant to quintic B-spline form.
-But the authors say in Sections 3 and 4 that they solve a banded linear system of length n (n the number of data points) to make this conversion.
-However, this is not necessary.
-At each data point, only three of the quintic B-splines are non-zero.
-Therefore, the three associated B-spline coefficients can be computed from the value, first, and second derivatives of the interpolant at that single point.
-Thus one can just solve a small 3*3 linear system to get one group of 3 B-spline coefficients.
-Well, it requires a bit of algebraic work to write out this 3*3 linear system, but it can be done.
+\item{(10)} It is quite probable that a user would want to convert the
+  piecewise Hermite representation of the interpolant to quintic
+  B-spline form. But the authors say in Sections 3 and 4 that they
+  solve a banded linear system of length n (n the number of data
+  points) to make this conversion. However, this is not necessary. At
+  each data point, only three of the quintic B-splines are
+  non-zero. Therefore, the three associated B-spline coefficients can
+  be computed from the value, first, and second derivatives of the
+  interpolant at that single point. Thus one can just solve a small
+  3*3 linear system to get one group of 3 B-spline coefficients. Well,
+  it requires a bit of algebraic work to write out this 3*3 linear
+  system, but it can be done.
 
@@
 
-As mentioned in the manuscript, the banded linear system representation is favored for two reasons in this package.
-The complexity order is unchanged by solving a single banded system, while the numerical stability is {\it theoretically} improved.
-In practice it is almost certainly slower than solving individual linear systems, but the increased stability and decreased redundancy in representation of the current approach is favored here.
-For users interested in constructing alternative representations, the optional argument {\it \bf UV} exists precisely for returning the computed first and second derivatives at knots.
+As mentioned in the manuscript Section 5 paragraph 4, the banded
+linear system representation is favored for two reasons in this
+package. The computational complexity order is unchanged by solving a
+single banded system, while the global numerical stability is {\it
+  theoretically} improved. The system of equations and number of
+nonzero splines at each breakpoint is thoroughly documented in
+{\ttVIII SPLINE.f90}, lines 172--206. In practice this approach is
+almost certainly slower than solving individual linear systems, but
+the increased global stability and decreased redundancy in
+representation is favored here (since the banded linear system also
+does not increase computational complexity). For users interested in
+constructing alternative representations, the optional argument {\it
+  \bf UV} documented in the code exists precisely for returning the
+computed first and second derivatives at knots.
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} I'm confused by the O(1) and O(n) discussion in Section 4.
-If there are n data points then surely the main algorithm (estimating first and second
-derivatives at each data point) must require at least an O(n) algorithm, not O(1).
+\item{(11)} I'm confused by the O(1) and O(n) discussion in Section 4.
+  If there are n data points then surely the main algorithm
+  (estimating first and second derivatives at each data point) must
+  require at least an O(n) algorithm, not O(1).
 
@@
 
-This was poorly worded on our part.
-The entire algorithm is O(n), however the binary search itself has a fixed number of loop iterations.
-The original intent of the statement was that the complexity order of the outer loop (the binary search) is constant, which ensures the overall complexity of O(n) is not changed.
-See the new phrasing in Section {\bf BLANK} paragraph {\bf BLANK}.
+This was poorly worded on our part. The entire algorithm is O(n),
+however the binary search itself has a fixed number of loop
+iterations. The original intent of the statement was that the
+complexity order of the outer loop (the binary search) is constant,
+which ensures the overall complexity of O(n) is not changed. These
+sentences in Section 4 paragraph 1 have been rephrased.
 
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} The output are values -- why not the spline coefficients? If the output are values one would hope for a visualization package.
+\item{(12)} The output are values -- why not the spline coefficients? If
+  the output are values one would hope for a visualization package.
 
@@
 
-Response
+The code includes suitable documentation and examples to conform to
+any external visualization package desired by a user. Given the wide
+variety of mechanisms by which a user might want to visualize or use
+the results of this code, it is nearly impossible to provide a good
+generic solution to visualization.
 
+For the convenience of reviewers, we are providing the code with which
+we generated the visuals for this manuscript. Although we note it has
+very specific system requirements to operate effectively. Namely: (1)
+running a POSIX shell (on macOS, Ubuntu, $\ldots$); (2) having
+{\ttVIII python3} of at least version $3.7$ and the packages {\ttVIII
+  numpy}, {\ttVIII tlux}, and {\ttVIII rpy2}; (3) having installed
+{\ttVIII gfortran} (or otherwise reconfiguring the codes to use
+another Fortran compiler of choice); and (4) having installed {\ttVIII
+  R} (for Schumaker splines). The Fortran codes associated with this
+work as well as those from TOMS 574 (quadratic $C^1$) and BVSPIS from
+TOMS 770 are all wrapped to be callable directly from {\ttVIII
+  python3}, which will be the parent process for all calls. A simple
+test usage can be seen in the provided {\ttVIII python3} file {\ttVIII
+  test.py}. To be abundantly clear, this code is {\bf not} intended
+for review and publication, but rather convenient access to various
+codes for reviewers to use for visualizing and comparing results.
+
+\goodbreak
 {\parindent=20pt \it
 
-\item{()} I was not familiar with the term/notion  "quadratic facet model".
-The term facet made me believe that the paper was on bivariate interpolation.
-Nothing in the abstract or title prevents this misconception.
+\item{(13)} I was not familiar with the term/notion ``quadratic facet
+  model''. The term facet made me believe that the paper was on
+  bivariate interpolation. Nothing in the abstract or title prevents
+  this misconception.
 
@@
 
-\bigskip
+The term ``univariate'' has been added to the abstract for greater
+clarity on this point.
 
-To the best of our knowledge, rejection based on a single review has not been a journal policy and we respectfully request additional reviews.
-We believe this algorithm and the accompanying code has great potential for TOMS, noting that the quadratic and cubic works preceding this have remained state-of-the-art for forty years.
-Thank you for both your time and consideration.
+\vfill \eject
+
+Throughout the manuscript, all modified text has been marked in
+\red{red}. We hope that you all find these revisions to have improved
+the manuscript, and we eagerly await more feedback.
 

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Fixed two numerical instabilities in the code related to large numbers of points. Updated based on minor feedback for response and paper.
Finishing revision draft, few more red highlights to add.

@@
 [Moler 2008], however this method also yields approximations whose
-second derivative functions often have large $L^2$ norm (approximations
-with large {\it wiggle}). In an attempt to capture the local shape of
-the data while minimizing wiggle, this package uses a facet model from
-image processing [Haralick and Watson 1981] to estimate first and
-second derivatives at breakpoints. Rather than picking a local linear
-or quadratic fit with minimal residual, this work uses a quadratic
-facet model that selects the local quadratic interpolant with minimum
-magnitude curvature.
+second derivative functions often have large local $L^2$ norm
+(approximations with large {\it wiggle}). In an attempt to capture the
+local shape of the data while minimizing wiggle, this package uses a
+facet model from image processing [Haralick and Watson 1981] to
+estimate first and second derivatives at breakpoints. Rather than
+picking a local linear or quadratic fit with minimal residual, this
+work uses a quadratic facet model that selects the local quadratic
+interpolant with minimum magnitude second derivative.
 
@@
 where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
-\le n$, and $n \ge 3$. Returns the slope and curvature at $X_i$ of the
-local quadratic interpolant with minimum magnitude curvature.
+\le n$, and $n \ge 3$. Returns the first and second derivative at 
+$X_i$ of the local quadratic interpolant with minimum magnitude
+second derivative. Approximate equality is denoted with $\approx$ and
+states that two numbers are considered equal once they are within the
+machine precision $\epsilon$ of each other.
 
@@
 \itemitem{} The point $(X_i$, $Y_i)$ is an extreme point. The
-quadratic with minimum magnitude
-curvature that has slope zero at $X_i$ will be the facet chosen.
+quadratic with minimum magnitude second derivative that has slope
+zero at $X_i$ will be the facet chosen.
 \itemitem{} $f_1\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$, $(X_i,Y_i)$,
@@
 not exist because $i \in \{2, n-1\}$. In those cases, the minimum
-magnitude curvature among existing quadratics is chosen.}
+magnitude second derivative among existing quadratics is chosen.}
 \itemitem{} $f_1\:\hbox{interpolant to }(X_{i-2},Y_{i-2})$,
@@
 
-The estimated derivative values by the quadratic facet model are not
-guaranteed to produce monotone quintic polynomial segments. Ulrich and
-Watson [1994] established tight constraints on the monotonicity of a
-quintic polynomial piece, while deferring to He{\ss} and Schmidt
-[1994] for a relevant simplified case. The following algorithm
-implements a sharp check for monotonicity by considering the
-nondecreasing case. The nonincreasing case is handled similarly.
+For constructing a quadratic interpolant in $x$ over the interval
+$[L,R]$, the Chebyshev basis $1$, $z$, $2z^2-1$ is used, where
+$z={x-(L+R)/2 \over (R-L)/2}$. The estimated derivative values by the
+quadratic facet model are not guaranteed to produce monotone quintic
+polynomial segments. Ulrich and Watson [1994] established tight
+constraints on the monotonicity of a quintic polynomial piece, while
+deferring to He{\ss} and Schmidt [1994] for a relevant simplified
+case. The following algorithm implements a sharp check for
+monotonicity by considering the nondecreasing case. The nonincreasing
+case is handled similarly.
 
@@
 $Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
-increasing on $[x_0,x_1]$.
+increasing on $[x_0,x_1]$. Approximate equality is denoted with
+$\approx$ and states that two numbers are considered equal once they
+are within the machine precision $\epsilon$ of each other.
 
@@
   \sqrt{\alpha \delta}$. If none of the monotonicity conditions are
-  violated, then the order six piece is monotone and Step 12
+  violated, then the degree five piece is monotone and Step 12
   concludes.}
@@
 
-Given the minimum magnitude curvature nature of the initial derivative
-estimates, it is desirable to make the smallest necessary changes to
-the initial interpolating spline $Q$ while enforcing monotonicity. In
-practice a binary search for the boundary of monotonicity is used in
-place of solely shrinking $DQ$ and $D^2Q$ at breakpoints adjoining
-{\it active\/} intervals: intervals over which $Q$ is nonmonotone at
-least once during the search. The binary search considers a Boolean
-function $B_i(s)$, for $0 \le s \le 1$, that is true if the order six
-polynomial piece of $Q(x)$ on $[X_i, X_{i+1}]$ matching derivatives
+Given the minimum curvature (minimum magnitude of the second derivative)
+nature of the initial derivative estimates, it is desirable to make
+the smallest necessary changes to the initial interpolating spline $Q$
+while enforcing monotonicity. In practice a binary search for the
+boundary of monotonicity is used in place of solely shrinking $DQ$ and
+$D^2Q$ at breakpoints adjoining {\it active\/} intervals: intervals
+over which $Q$ is nonmonotone at least once during the search. The
+binary search considers a Boolean function $B_i(s)$, for $0 \le s \le
+1$, that is true if the order six polynomial piece of $Q(x)$ on
+$[X_i, X_{i+1}]$ matching derivatives
 $Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and 
 derivatives $Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, 
-$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false otherwise. 
-The binary search is only applied at those breakpoints adjoining intervals
-$[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and hence $B_i(1)$ is false.
-It is further assumed that there exists $0 \le s^* \le 1$ such that $B_i(s)$
-is true for $0 \le s \le s^*$ and false for some $1 > s > s^*$. Since the
-derivative conditions at interior breakpoints are shared by intervals left
-and right of the breakpoint, the binary search is performed at all
-breakpoints simultaneously.  Specifically, the monotonicity of $Q$ is
-checked on all active intervals in each step of the binary search to
-determine the next derivative modification at each breakpoint. The goal
-of this search is to converge on the boundary of the monotone region in
-the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
+$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false
+otherwise.  The binary search is only applied at those breakpoints
+adjoining intervals $[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and
+hence $B_i(1)$ is false.  It is further assumed that there exists
+$0 \le s^* \le 1$ such that $B_i(s)$ is true for $0 \le s \le s^*$ and
+false for some $1 > s > s^*$. Since the derivative conditions at
+interior breakpoints are shared by intervals left and right of the
+breakpoint, the binary search is performed at all breakpoints
+simultaneously.  Specifically, the monotonicity of $Q$ is checked on
+all active intervals in each step of the binary search to determine
+the next derivative modification at each breakpoint. The goal of this
+search is to converge on the boundary of the monotone region in the
+$(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
 Watson [1994]) for all intervals. This multiple-interval binary search
-allows the value zero to be obtained for all (first and second) derivative
-values in a fixed maximum number of computations, hence has no effect on
-computational complexity order. This binary search algorithm is outlined
-below.
+allows the value zero to be obtained for all (first and second)
+derivative values in a fixed maximum number of computations, hence has
+no effect on computational complexity order. This binary search
+algorithm is outlined below.
 

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
Finishing revision draft, few more red highlights to add.

@@
 \heading{4. COMPLEXITY AND SENSITIVITY}
-Algorithms 1, 2, and 4 have ${\cal O}(1)$ runtime. Given a fixed
+
+Algorithms 1 and 4 have ${\cal O}(n)$ runtime for $n$ data
+points. Algorithm 2 has a fixed cost ${\cal O}(1)$. Given a fixed
 schedule for shrinking derivative values, Algorithm 3 has a ${\cal
@@
 behavior while empirically producing final approximations with less
-wiggle ($L^2$ norm of the second derivative) than other methods. A
+wiggle (local $L^2$ norm of the second derivative) than other methods. A
 weighted harmonic mean estimate of first derivatives may be more
@@
 {\everymath={\scriptstyle}
+
 \narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
@@
   $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
-  curvature. Given the exact five data points seen above, the quadratic
-  facet model produces the slope seen in the solid blue line at $x = 5/2$.
-  However, by subtracting the value of $f_3$ $= \epsilon(x-2)^2$ from
-  points at $x = 3$, 4, where $\epsilon$ is the machine precision
-  ($2^{-52}$ for an IEEE 64-bit real), the quadratic facet model produces
-  the slope seen in the dashed red line at $x = 5/2$. This is the nature
-  of a facet model and a side effect of associating data with local facets.
+  curvature (equal second derivatives). Given the exact five data points
+  seen above, the quadratic facet model produces the slope seen in the
+  solid blue line at $x = 5/2$. However, by subtracting the value of
+  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
+  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
+  quadratic facet model produces the slope seen in the dashed red line
+  at $x = 5/2$. This is the nature of a facet model and a side effect
+  of associating data with local facets.
 \par} \endinsert

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
Finishing revision draft, few more red highlights to add.

@@
 \heading{5. PERFORMANCE AND APPLICATIONS}
-
-This section contains graphs of sample {\tt MQSI} results given various
-data configurations. Execution times for Algorithms 1 and 4 are given; the
-total execution time of {\tt MQSI} is utterly dominated by the time for a
-banded linear system solve computing the $3n$ $B$-spline coefficients,
-which is ${\cal O}(n)$.  The files {\tt sample\_main.f90} and {\tt
-sample\_main.dat} accompanying the subroutine {\tt MQSI} illustrate Fortran
-2003 subroutine usage with optional arguments and data points from a file.
-Compilation instructions and the full package contents are specified in
-the {\tt README} file.
 
@@
 
+This section contains graphs of sample {\tt MQSI} results given various
+data configurations. Execution times for Algorithms 1 and 4 are given; the
+total execution time of {\tt MQSI} is utterly dominated by the time for a
+banded linear system solve computing the $3n$ $B$-spline coefficients,
+which is ${\cal O}(n)$.  The files {\tt sample\_main.f90} and {\tt
+sample\_main.dat} accompanying the subroutine {\tt MQSI} illustrate Fortran
+2003 subroutine usage with optional arguments and data points from a file.
+Compilation instructions and the full package contents are specified in
+the {\tt README} file.
+
 Throughout, all visuals have points that are stylized by local
@@
 
+The binary search procedure provably converges on the boundary of the
+region of monotonicity precisely defined by Ulrich and Watson [1994]
+through the application of a boolean function that is guaranteed to be
+true at one end of an interval. Since a local quadratic interpolant is
+applied for initial function derivative estimates, the approximation
+order of the resulting fit is $O(h^3)$ (as for any second order
+approximation), but it should be noted that asymptotic approximation
+order is rarely of import when considering the scattered sparse data
+that $C^2$ approximations like MQSI provide.
+
 Comparisons between this work and lower order methods are problematic,
@@
 software suffices.
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/4-bvspis.eps}}
+{\narrower\noindent\rmVIII Fig.\ 5.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), PCHIP (third, green), and BVSPIS (fourth, purple)
+respectively on the {\itVIII piecewise polynomial} test function. MQSI
+is styled as a gray thin line in the background for comparison. The
+Schumaker code (second) fails to produce a monotone approximation for
+this problem while producing no errors or warnings. The primary
+comparative observation of MQSI otherwise is its exact reproduction of
+the linear segment between $0.3$ and $0.5$.
+\par}
+\endinsert
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/4-bvspis.eps}}
+{\narrower\noindent\rmVIII Fig.\ 5.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), PCHIP (third, green), and BVSPIS (fourth, purple)
+respectively on the {\itVIII large tangent} test function. MQSI is
+styled as a gray thin line in the background for comparison. This
+problems highlights the core weakness of the quadratic facet model
+approach to derivative estimation in MQSI, which consistently
+underestimates the actual curvature of this function. This weakness is
+accepted for its greater accuracy when estimating local lower order
+components of functions.
+\par}
+\endinsert
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/4-bvspis.eps}}
+{\narrower\noindent\rmVIII Fig.\ 6.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), PCHIP (third, green), and BVSPIS (fourth, purple)
+respectively on the {\itVIII signal decay} test function. MQSI is
+styled as a gray thin line in the background for comparison. The most
+notable differences between MQSI and other approaches can be observed
+near local extrema, where MQSI produces smaller magnitude second
+derivatives.
+\par}
+\endinsert
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/4-bvspis.eps}}
+{\narrower\noindent\rmVIII Fig.\ 7.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), PCHIP (third, green), and BVSPIS (fourth, purple)
+respectively on the {\itVIII random monotone} test function. MQSI is
+styled as a gray thin line in the background for comparison. Notice
+that the numerical conditions for this approximation problem are
+challenging enough that both the Schumaker and BVSPIS codes
+incorrectly produce nonmonotone segments (neither of which produced
+error codes or any form of indication that a failure
+occurred). Notably only MQSI and TOMS 574 satisfy tight monotonicity
+constraints while also correctly preserving monotonicity in their
+approximations. The main segment of divergence between the methods
+happens on the $(0.45, 0.5)$ interval, where MQSI favors maximizing
+the slope on the left side of the interval because the quadratic
+interpolant of the right sided points produces an (invalid) negative
+slope estimate on the left side of the interval.
+\par}
+\endinsert

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/TOMS_770/bvspis/__init__.py

@@ -1,1526 +1,0 @@
-'''This Python code is an automatically generated wrapper
-for Fortran code made by 'fmodpy'. The original documentation
-for the Fortran source code follows.
-
-
-'''
-
-import os
-import ctypes
-import numpy
-
-# --------------------------------------------------------------------
-#               CONFIGURATION
-# 
-_verbose = True
-_fort_compiler = "gfortran"
-_shared_object_name = "bvspis.so"
-_this_directory = os.path.dirname(os.path.abspath(__file__))
-_path_to_lib = os.path.join(_this_directory, _shared_object_name)
-_compile_options = ['-fPIC', '-shared', '-O3']
-_ordered_dependencies = ['bvspis.f', 'bvspis_c_wrapper.f90']
-# 
-# --------------------------------------------------------------------
-#               AUTO-COMPILING
-#
-# Try to import the existing object. If that fails, recompile and then try.
-try:
-    clib = ctypes.CDLL(_path_to_lib)
-except:
-    # Remove the shared object if it exists, because it is faulty.
-    if os.path.exists(_shared_object_name):
-        os.remove(_shared_object_name)
-    # Compile a new shared object.
-    _command = " ".join([_fort_compiler] + _compile_options + ["-o", _shared_object_name] + _ordered_dependencies)
-    if _verbose:
-        print("Running system command with arguments")
-        print("  ", _command)
-    # Run the compilation command.
-    import subprocess
-    subprocess.run(_command, shell=True, cwd=_this_directory)
-    # Import the shared object file as a C library with ctypes.
-    clib = ctypes.CDLL(_path_to_lib)
-# --------------------------------------------------------------------
-
-
-# ----------------------------------------------
-# Wrapper for the Fortran subroutine DBVSIS
-
-def dbvsis(x, y, np, n, k, opt, d0, dnp, d20, d2np, constr, eps, beta, betai, rho, rhoi, kmax, maxstp, xtab, ntab, sbopt, y0opt, y1opt, y2opt, errc, erre, d, d2, diagn, y0tab, y1tab, y2tab, work, nwork):
-    '''!  DBVSIS is merely a support routine which calls DBVSSC and DBVSSE
[…]

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/mqsi/MQSI_python_wrapper.py

@@ -1,0 +1,459 @@
+'''This Python code is an automatically generated wrapper
+for Fortran code made by 'fmodpy'. The original documentation
+for the Fortran source code is attached to each wrapped function.
+'''
+
+import os
+import ctypes
+import numpy
+
+# --------------------------------------------------------------------
+#               CONFIGURATION
+# 
+_verbose = True
+_link_lapack = True
+_fort_compiler = "gfortran"
+_shared_object_name = "MQSI.so"
+_this_directory = os.path.dirname(os.path.abspath(__file__))
+_path_to_lib = os.path.join(_this_directory, _shared_object_name)
+if _link_lapack:
+    _compile_options = ['-fPIC', '-shared', '-O3', '-lblas', '-llapack']
+    _ordered_dependencies = ['REAL_PRECISION.f90', 'EVAL_BSPLINE.f90', 'SPLINE.f90', 'MQSI.f90', 'MQSI_c_wrapper.f90']
+else:
+    _compile_options = ['-fPIC', '-shared', '-O3']
+    _ordered_dependencies = ['blas.f', 'lapack.f', 'REAL_PRECISION.f90', 'EVAL_BSPLINE.f90', 'SPLINE.f90', 'MQSI.f90', 'MQSI_c_wrapper.f90']
+# 
+# --------------------------------------------------------------------
+#               AUTO-COMPILING
+#
+# Try to import the existing object. If that fails, recompile and then try.
+try:
+    clib = ctypes.CDLL(_path_to_lib)
+except:
+    # Remove the shared object if it exists, because it is faulty.
+    if os.path.exists(_shared_object_name):
+        os.remove(_shared_object_name)
+    # Compile a new shared object.
+    _command = " ".join([_fort_compiler] + _compile_options + ["-o", _shared_object_name] + _ordered_dependencies)
+    if _verbose:
+        print("Running system command with arguments")
+        print("  ", _command)
+    # Run the compilation command.
+    import subprocess
+    subprocess.run(_command, shell=True, cwd=_this_directory)
+    # Import the shared object file as a C library with ctypes.
+    clib = ctypes.CDLL(_path_to_lib)
+# --------------------------------------------------------------------
+
+
+# ----------------------------------------------
+# Wrapper for the Fortran subroutine MQSI
+
+def mqsi(x, y, t, bcoef, uv=None):
+    '''! Computes a monotone quintic spline interpolant (MQSI), Q(X), to data
[…]

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/mqsi/sample_main.f90

@@ -1,0 +1,112 @@
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                          sample_main.f90
+! 
+! DESCRIPTION:
+!   This file (sample_main.f90) contains a sample main program that
+!   illustrates the usage of subroutines MQSI and EVAL_SPLINE to interpolate
+!   data read from a file and compute derivatives of the monotone quintic
+!   spline interpolant Q(x) at the data points (via the optional argument
+!   UV to MQSI) and at other points (via EVAL_SPLINE).  Compile with:
+! 
+!    $F03 $OPTS REAL_PRECISION.f90 EVAL_BSPLINE.f90 SPLINE.f90 MQSI.f90 \
+!       sample_main.f90 -o main $LIB
+!
+!   where '$F03' is the name of the Fortran 2003 compiler, '$OPTS' are
+!   compiler options such as '-O3', and '$LIB' provides a flag to link
+!   BLAS and LAPACK. If the BLAS and LAPACK libraries are not
+!   available on your system, then replace $LIB with the filenames
+!   'blas.f lapack.f'; these files contain the routines from the BLAS
+!   and LAPACK libraries that are necessary for this package.
+!
+! CONTAINS:
+!   PROGRAM SAMPLE_MAIN
+! 
+! DEPENDENCIES:
+!   MODULE REAL_PRECISION
+!     INTEGER, PARAMETER :: R8
+!   END MODULE REAL_PRECISION
+! 
+!   SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: X
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: Y
+!     REAL(KIND=R8), INTENT(OUT),   DIMENSION(:) :: T, BCOEF
+!     INTEGER, INTENT(OUT) :: INFO
[…]

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/mqsi/test_all.f90

@@ -1,0 +1,627 @@
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                           test_all.f90
+! 
+! DESCRIPTION:
+!   This file contains code for testing a local installation of the MQSI
+!   package. Here is an example command to compile and run all tests:
+! 
+!    $F03 $OPTS REAL_PRECISION.f90 EVAL_BSPLINE.f90 SPLINE.f90 MQSI.f90 \
+!      test_all.f90 -o test_all $LIB && ./test_all
+!
+!   where '$F03' is the name of the Fortran 2003 compiler, '$OPTS' are
+!   compiler options such as '-O3', and '$LIB' provides a flag to link
+!   BLAS and LAPACK. If the BLAS and LAPACK libraries are not
+!   available on your system, then replace $LIB with the filenames
+!   'blas.f lapack.f'; these files contain the routines from the BLAS
+!   and LAPACK libraries that are necessary for this package.
+! 
+!   All tests will be run and an error message will be printed for any
+!   failed test cases. Otherwise a message saying tests have passed
+!   will be directed to standard output. A small timing test will also
+!   be run to display the expected time required to fit and evaluate a
+!   monotone quintic spline interpolant to data of a given size.
+! 
+! CONTAINS:
+!   PROGRAM TEST_ALL
+! 
+! DEPENDENCIES:
+!   MODULE REAL_PRECISION
+!     INTEGER, PARAMETER :: R8
+!   END MODULE REAL_PRECISION
+! 
+!   SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: X
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: Y
+!     REAL(KIND=R8), INTENT(OUT),   DIMENSION(:) :: T, BCOEF
+!     INTEGER, INTENT(OUT) :: INFO
[…]

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/toms574/__init__.py

@@ -1,0 +1,780 @@
+'''This Python code is an automatically generated wrapper
+for Fortran code made by 'fmodpy'. The original documentation
+for the Fortran source code follows.
+
+!$$$C     PROGRAM MAINDR                                                    MAIN  10
+!$$$      REAL X(100),Y(100),XVAL(152),YVAL(152),M(100),XVAL1(200),         MAIN  20
+!$$$     *     DELTAH,EPS                                                   MAIN  30
+!$$$      INTEGER OUT,ERR                                                   MAIN  40
+!$$$C                                                                       MAIN  50
+!$$$C THIS PROGRAM IS A TEST DRIVER FOR THE SHAPE PRESERVING QUADRATIC      MAIN  60
+!$$$C SPLINES BY D.F.MCALLISTER AND J.A.ROULIER.                            MAIN  70
+!$$$C                                                                       MAIN  80
+!$$$C ON INPUT--                                                            MAIN  90
+!$$$C                                                                       MAIN 100
+!$$$C   X CONTAINS THE ABSCISSAS OF THE POINTS OF INTERPOLATION.            MAIN 110
+!$$$C                                                                       MAIN 120
+!$$$C   Y CONTAINS THE ORDINATES OF THE POINTS OF INTERPOLATION.            MAIN 130
+!$$$C                                                                       MAIN 140
+!$$$C   N IS THE NUMBER OF DATA POINTS.                                     MAIN 150
+!$$$C                                                                       MAIN 160
+!$$$C   K IS THE NUMBER OF POINTS AT WHICH THE SPLINE IS TO BE EVALUATED.   MAIN 170
+!$$$C                                                                       MAIN 180
+!$$$C                                                                       MAIN 190
+!$$$C UPON EXIT FROM SUBROUTINE 'SLOPES'--                                  MAIN 200
+!$$$C                                                                       MAIN 210
+!$$$C   M CONTAINS THE COMPUTED FIRST DERIVATIVES AT EACH DATA POINT.       MAIN 220
+!$$$C                                                                       MAIN 230
+!$$$C-----------------------------------------------------------------------MAIN 240
[…]

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
Finishing revision draft, few more red highlights to add.

--- 2021-04-22 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
@@
 
-The currently available software for monotone piecewise polynomial
-interpolation includes quadratic [He and Shi 1998, McAllister and
-Roulier 1981], cubic [Fritsch and Carlson 1980], and (with limited
-application) quartic [Wang and Tan 2004; Piah and Unsworth 2011; Yao
-and Nelson 2018] cases. In addition, a statistical method for
-bootstrapping the construction of an arbitrarily smooth monotone fit
-exists [Leitenstorfer and Tutz 2006], but the method does not take
-advantage of known analytic properties of quintic polynomials.
-Conversely the work on quadratic shape preserving splines by Schumaker
-[1983] is widely utilized in theory for its conservation of both
-monotonicity and convexity, but no mathematical software is published.
+The existing research in shape preserving interpolatory splines is
+rich and filled with many approaches for many different
+applications. In the context of this work, the unique quality of shape
+preserving spline algorithms can be observed through: (1) the type of
+spline (rational or Hermite), (2) the polynomial order and level of
+continuity, and (3) the sharpness, sufficiency, and necessity of the
+conditions used to establish shape preserving properties. While the
+choice of spline representation is less important, achieving higher
+levels of continuity is often more difficult than lower levels of
+continuity. And while all referenced algorithms establish sufficient
+conditions for shape preservation, it is less common and much more
+difficult to provide sharp {\it necessary} conditions for
+monotonicity. The following two paragraphs highlight a collection of
+research related to the construction of sufficient conditions for
+$C^1$ and $C^2$ shape preserving splines. From this foundation, the
+value and utility of the code presented here can be more readily
+appreciated.
 
-The code by Fritsch [1982] for $C^1$ cubic spline interpolation is the
-predominantly utilized code for constructing monotone interpolants at
-present. Theory has been provided for the quintic case [Ulrich and
-Watson 1994; He{\ss} and Schmidt 1994] and that theory was recently
+DeVore [1977] proves that monotone spline interpolants have improved
+accuracy over their nonmonotone counterparts when approximating
+monotone functions, while Constantini [1986] proves that monotone
+spline interpolants to data of fixed continuity and order exist in
+general. This groundwork establishes the potential for monotone spline
+interpolation. Schumaker [1983] and McAllister and Roulier [1981]
+(independently) establish necessary and sufficient conditions for
+monotone $C^1$ quadratic splines through the insertion of additional
+knots and Fritsch [1982] establishes simplified sufficient conditions
+for monotone $C^1$ cubic splines that require no knot insertion.
+Gregory and Delbourgo [1983] prove necessary and sufficient conditions
+for a closed form solution to monotone $C^1$ quadratic rational spline
+interpolation with nonlinear boundary equations and an iterative
+approach. Delbourgo and Gregory [1993] extend that same work to
+achieve sufficient conditions for $C^2$ continuity with a cubic
+rational spline given additional tension parameters defined by
+users. Huynh [1993] similarly arrives at monotone $C^1$ cubic spline
+interpolants and several necessary nonlinear boundary conditions on
+monotonicity.
+
+Continuing towards monotone $C^2$ spline interpolation, Fiorot and
+Tabka [1991] prove a simple method for determining the existence of a
+monotone $C^2$ cubic interpolating spline, but note that functions of
+this kind do not always exist for arbitrary convex monotone data sets.
+Pruess [1993] proposes sufficient conditions for a cubic $C^2$ shape
+preserving spline method while acknowledging that quintic splines are
+necessary for generally monotone $C^2$ spline interpolants that do not
+insert new knots. Similarly, Manni [1997] constructs sufficient
+conditions for a monotone $C^2$ cubic spline by adding two additional
+knots per interval, then Cravery and Manni [2003] extend that
+methodology to arrive at monotone $C^3$ interpolating splines by
+progressively increasing tension through Bezier control
+polygons. Constantini [1997] provides a method for monotone $C^2$
+quintic spline interpolation based on sufficient and necessary
+boundary value conditions, but the necessary conditions are not
+sharp. Dougherty, Edelman, and Hyman [1989] construct monotone $C^2$
+Hermite splines and in commentary recommend against the direct
+optimization for smoothness parameters like global $L^2$, rather
+promoting problem specific definitions of geometric niceness. Both
+Wang and Tan [2004] as well as Yao and Nelson [2018] propose
+sufficient conditions for $C^2$ quartic splines, while Piah and
+Unsworth [2011] improve upon those sufficient conditions for $C^2$
+quartic rational spline interpolation. Abbas, Majid, and Ali [2012]
+formulate a set of sufficient conditions for monotone $C^2$ cubic
+rational splines and similar work is extended to $C^2$ quartic
+rationals (and splines with arbitrary smoothness) by Zhu and Han
+[2015]. Ulrich and Watson [1994] arrived at sufficient and sharp
+necessary conditions for monotone $C^2$ quintic splines, as also did
+He{\ss} and Schmidt [1994] (independently). In addition, a statistical
+method for bootstrapping the construction of an arbitrarily smooth
+monotone fit exists [Leitenstorfer and Tutz 2006], but this is enabled
+by sufficient conditions only.
+
+The currently available peer reviewed and published software for
+monotone piecewise polynomial interpolation is severely reduced in
+comparison with the number of published approaches mentioned
+above. This is partially indicative of how difficult it is to properly
+handle the numerical conditions that arise when constructing and
+evaluating precise shape preserving splines, as well as how difficult
+it is to create code that correctly conforms with the theoretical
+expectations. The sufficient $C^1$ cubic spline method PCHIP of
+Fritsch and Carlson [1980] is available through the SciPy Python
+package ($>1$ billion downloads), and the $C^1$ quadratic method of
+Schumaker [1983] is available as an R package (300 thousand
+downloads). The piecewise quadratic $C^1$ method of McAllister and
+Roulier [1981] is available in FORTRAN (1000 downloads) and the
+sufficient $C^2$ quintic method of Costantini [1997] BVSPIS in FORTRAN
+as well (500 downloads). Based on publicly available download records,
+it is assumed that the code by Fritsch [1982] for monotone $C^1$ cubic
+spline interpolation is the predominant code for constructing monotone
+interpolants at present.
+
+The theory for sufficient and {\it sharp} necessary bounds for
+monotone quintic interpolation has been provided [Ulrich and Watson
+  1994; He{\ss} and Schmidt 1994] and that theory was recently
 utilized in a proposed algorithm [Lux 2020] for monotone quintic
-spline construction, however no published mathematical software
-exists. The software presented here represents the first published
-software package for producing $C^2$ shape preserving splines. This
-work improves upon the algorithm presented by Lux et al. [2020] by
-refactoring computations for improved numerical stability, estimating
-minimum curvature derivatives at breakpoints with a quadratic facet
-model, and using a binary search to reduce the magnitude of the
-modifications made to initial derivative estimates when constructing a
-monotone spline interpolant.
+spline construction, however no published mathematical software exists
+for the quintic case based on sharp monotonicity conditions. The
+software presented here represents the first published software
+package for producing $C^2$ shape preserving splines based on sharp
+monotonicity conditions. This work improves upon the algorithm
+presented by Lux et al. [2020] by refactoring computations for
+improved numerical stability, estimating minimum magnitude second
+derivatives at breakpoints with a quadratic facet model, and using a
+binary search to reduce the magnitude of the modifications made to
+initial derivative estimates when constructing a monotone spline
+interpolant.
+
+\heading{Overview}
 
 This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
-necessary and sufficient conditions in Ulrich and Watson [1994] for the
-construction of monotone quintic spline interpolants of monotone data.
-Precisely, the problem is, given a strictly increasing sequence $X_1<X_2<
-\cdots <X_n$ of breakpoints with corresponding monotone increasing
-function values $Y_1\le Y_2\le \cdots \le Y_n$, find a $C^2$ monotone
-increasing quintic spline $Q(x)$ with the same breakpoints satisfying
-$Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI} actually does something slightly
-more general, producing $Q(x)$ that is monotone increasing (decreasing)
-wherever the data is monotone increasing (decreasing).)
+sharp necessary and sufficient conditions in Ulrich and Watson [1994]
+for the construction of monotone quintic spline interpolants of
+monotone data. Precisely, the problem is, given a strictly increasing
+sequence $X_1<X_2< \cdots <X_n$ of breakpoints with corresponding
+monotone increasing function values $Y_1\le Y_2\le \cdots \le Y_n$,
+find a $C^2$ monotone increasing quintic spline $Q(x)$ with the same
+breakpoints satisfying $Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI}
+actually does something slightly more general, producing $Q(x)$ that
+is monotone increasing (decreasing) wherever the data is monotone
+increasing (decreasing).)
 

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Finishing revision draft, few more red highlights to add.

--- 2021-04-22 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
@@
 \magnification=1000 \hsize=30pc \vsize=47pc
-\parindent=10pt \parskip=0pt plus 3pt
-\baselineskip=13pt plus 2pt minus 1pt \lineskiplimit=2pt
+\parindent=10pt
+\parskip=0pt plus 3pt
+\baselineskip=13pt plus 2pt minus 1pt
+\lineskiplimit=2pt
 \lineskip=2pt plus 2pt %\raggedright
@@
 
+%% No extra knots, rational functions, sufficiency and necessity
+%% proven, works for unequally spaced data, no code associated with
+%% the paper, no algorithm provided for the initialization for free
+%% shape parameters.
+\refj{Abbas M., Majid, A. A. and Ali,
+  J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for
+  monotone data}{Applied Mathematics and
+  Computation}{219(6):2885--2895}
+
+%% Develops a quadratic spline curve fitting algorithm that maintains
+%% local convexity while also proving some error bounds. This method
+%% avoids inserting additional knots, but does not interpolate the
+%% data.
+%% \refj{Beatson, R. K.}{1982}{Monotone and convex approximation by
+%%   splines: error estimates and a curve fitting algorithm}{SIAM Journal
+%%   on Numerical Analysis}{19(6):1278--1285}
+
 % berglund2009planning
@@
 % deboor1978practical
-\refb{de Boor, C.}{1978}{A Practical Guide to Splines}{Springer, Verlag,
-New York}
+\refb{de Boor, C.}{1978}{A Practical Guide to Splines}{Springer,
+  Verlag, New York}
 
 % cameron2019moana
-\refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U.,
-Bernard, J., Jearls, C., Lux,  T. C. H., Hong, Y., Watson, L. T., and Butt,
-A. R.}{2019}{MOANA: modeling and analyzing I/O variability in parallel
-system experimental design}{IEEE Transactions on Parallel and Distributed
-Systems}{30(8):1843--1856}
+\refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth,
+  U., Bernard, J., Jearls, C., Lux, T. C. H., Hong, Y., Watson, L. T.,
+  and Butt, A. R.}{2019}{MOANA: modeling and analyzing I/O variability
+  in parallel system experimental design}{IEEE Transactions on
+  Parallel and Distributed Systems}{30(8):1843--1856}
+
+%% Proves existence of splines that maintain monotonicity and
+%% convexity while having fixed continuity and order.
+\refj{Costantini, P.}{1986}{On monotone and convex spline
+  interpolation}{Mathematics of Computation}{46(173):203--214}
+
+%% Source paper accompanying the BVSPIS publication.
+\refj{Costantini, P.}{1997}{Boundary-valued shape-preserving
+  interpolating splines}{ACM Transactions on Mathematical Software
+  (TOMS)}{23(2):229--251}
+
+%% BVSPIS Algorithm.
+\refj{Costantini, P.}{1997}{Algorithm 770: BVSPIS—a package for
+  computing boundary-valued shape-preserving interpolating
+  splines}{ACM Transactions on Mathematical Software (TOMS)}{23(2):252-–254}
+
+%% Constructs C3 interpolating splines by progressively increasing
+%% tension through Bezier control polygons until shape parameters are
+%% satisfied, no code is provided.
+\refj{Cravero, I., C. Manni}{2003}{Shape-preserving interpolants with
+  high smoothness}{Journal of computational and applied
+  mathematics}{157(2):383--405}
+
+%% Iterates towards a solution of nonlinear equations that ensure C2
+%% continuity of quadratic rational splines. The solution they iterate
+%% towards exists and is unique. They provide proofs and methods for
+%% setting derivatives at the ends, no formal algorithms or code.
+\refj{Delbourgo, R., Gregory, J. A.}{1983}{$C^2$ rational quadratic
+  spline interpolation to monotonic data}{IMA Journal of Numerical
+  Analysis}{3(2):141--152}
+
+%% Continuous previous work by incorporating a cubic numerator, proves
+%% shape preservation when the correct value for a tension parameter
+%% is set but no method for setting that parameter. Accuracy depends
+%% on accurate derivative evaluations. No code nor formal algorithm is
+%% provided.
+\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
+  tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}
+
+%% Proves approximation error bounds for monotone spline interpolation
+%% of data from monotone functions.
+\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
+  Journal on Mathematical Analysis}{8(5):891--905}
+
+%% Accompanies the previous error bounds in the context of monotone
+%% polynomial interpolation of data from a monotone function.
+\refj{DeVore, R. A.}{1977}{Monotone approximation by polynomials}{SIAM
+  Journal on Mathematical Analysis}{8(5):906--921}
+
+%% Proposes an algorithm for constructing C2 Hermite monotone spline
+%% interpolants using proven sufficient, but not necessary
+%% conditions. As these bounds are not sharp, the space of allowable
+%% derivative and second derivative values at knots is unnecessarily
+%% shrunk. An algorithm is provided, but no code. Recommends against
+%% the direct optimization for smoothness parameters like L2, instead
+%% promoting problem specific definitions of 'geometric niceness'.
+\refj{Dougherty, R. L., Edelman, A., Hyman,
+  J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving
+  cubic and quintic Hermite interpolation}{Mathematics of
+  Computation}{52(186):471--494}
+
+%% Proves a simple method for determining the existence of a monotone
+%% C2 cubic interpolating spline, but functions of this kind do not
+%% always exist for arbitrary convex monotone data sets.
+\refj{Fiorot, J. C., Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
+  polynomial interpolating splines}{Mathematics of
+  Computation}{57(195):291--298}
 
@@
 
+%% Proves necessary and sufficient conditions for a closed form
+%% solution to monotone C1 quadratic rational spline
+%% interpolation. Does not provide formal algorithm, nor code.
+\refj{Gregory , J. A.,Delbourgo, R.}{1982}{Piecewise rational
+  quadratic interpolation to monotonic data}{IMA Journal of Numerical
+  Analysis}{2(2):123--130}
+
 % gregory1985shape
@@
 
-% haralick1981facet
-\refj{Haralick, R. M., and Watson, L. T.}{1981}{A facet model for
-  image data.}{Computer Graphics and Image Processing}{15(2):113--129}
-
-% he1998monotone
-\refj{He, X. and Shi, P.}{1998}{Monotone B-spline smoothing}{Journal
-  of the American Statistical Association}{93(442):643--650}
+%% Proposes a method for making arbitrarily smooth shape preserving
+%% spline interpolants, but uses a centered quadratic to estimate
+%% derivatives and a sufficient set of bounding conditions to ensure
+%% shape preservation in practice. No necessary conditions are
+%% applied, no algorithm is provided, neither is any code provided.
+\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation
+  with higher order continuity}{Applied Mathematics and
+  Computation}{337:1--13}
 
@@
   quintic $\scriptstyle C^2$-spline interpolation in one and two
-  dimensions}{Journal of Computational and Applied Mathematics}{55(1):51--67}
+  dimensions}{Journal of Computational and Applied
+  Mathematics}{55(1):51--67}
+
+%% Presents several nonlinear boundary conditions based on the median
+%% function to establish a fourth order monotonicity preserving
+%% constraints relying on divided difference initial derivative
+%% estimates. Overall generates a monotone C1 cubic interpolant to
+%% data.
+\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
+  Journal of Numerical Analysis}{30(1):57--100}
 
@@
   \& Business Media, Volume 18}
+
+%% Use sufficient conditions for monotonicity and convexity to form a
+%% globally weighted combination of C1 quadratic splines that satisfy
+%% shape constraints.
+%% \refj{Kvasov, B. I.}{2014}{Monotone and convex interpolation by
+%%   weighted quadratic splines}{Advances in Computational
+%%   Mathematics}{40(1):91--116}
 
@@
   Polytechnic Institute and State University, Blacksburg, Virginia}
+
+%% Constructs a C2 cubic spline with two additional knots per
+%% interval. No code (paywall).
+\refj{Manni, C., Sablonni\`ere, P.}{1997}{Monotone interpolation of
+  order 3 by $C^2$ cubic splines}{IMA Journal of Numerical
+  Analysis}{17(2):305--320}
+
+%% Uses sufficient conditions to generate C2 cubic spline. No code
+%% (paywall).
+\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite
+  interpolation}{BIT Numerical Mathematics}{41(1):127--148}
 
@@
   interpolation}{Sains Malaysiana}{40(10):1173--1178}
+
+%% Proposes a cubic C2 shape preserving spline method with free
+%% parameters (mentions quintic splines are necessary for globally C2
+%% monotone splines). Sufficient (but not necessary) conditions are
+%% provided for monotonicity.
+\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
+  interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}
 
@@
 
+%% Uses sufficient conditions over three free parameters to construct
+%% C2 rational quartic shape preserving splines.
+\refj{Zhu, Y.,Han, X.}{2015}{$C^2$ rational quartic interpolation
+  spline with local shape preserving property}{Applied Mathematics
+  Letters}{46:57--63}
+
+%% Uses sufficient conditions over two free parameters to construct C2
+%% rational quartic shape preserving splines.
+\refj{Zhu, Y.,Han, X.}{2015}{Shape preserving $C^2$ rational quartic
+  interpolation spline with two parameters}{International Journal of
+  Computer Mathematics}{92(10):2160--2177}
+
 }

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/TOMS_574/new_formula.py

@@ -1,17 +1,0 @@
-# # Update derivative estimates by using the weighted harmonic mean.
-# DO I = 2, ND-1
-#    IP1 = I+1
-#    IM1 = I-1
-#    W1 = 2*X(IP1) - X(I) - X(IM1)
-#    W2 = -2*X(IM1) + X(I) + X(IP1)
-#    D1 = (Y(IP1) - Y(I)) / (X(IP1) - X(I))
-#    D2 = (Y(I) - Y(IM1)) / (X(I) - X(IM1))
-#    B = (W1 + W2) / (W1/D2 + W2/D1)
-#    A = B * ( 1.0_R8 / D1 - 1.0_R8 / D2 + &
-#         W1 / (Y(I) - Y(IM1)) - &
-#         W2 / (Y(IP1) - Y(I)) ) / (W1/D2 + W2/D1)
-#    IF (FX(I,2) .NE. 0.0_R8) &
-#         FX(I,2) = B
-#    IF (FX(I,3) .NE. 0.0_R8) &
-#         FX(I,3) = 0.0_R8
-# END DO

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/TOMS_574/toms574/__init__.py

@@ -1,724 +1,0 @@
-'''This Python code is an automatically generated wrapper
-for Fortran code made by 'fmodpy'. The original documentation
-for the Fortran source code follows.
-
-!$$$C     PROGRAM MAINDR                                                    MAIN  10
-!$$$      REAL X(100),Y(100),XVAL(152),YVAL(152),M(100),XVAL1(200),         MAIN  20
-!$$$     *     DELTAH,EPS                                                   MAIN  30
-!$$$      INTEGER OUT,ERR                                                   MAIN  40
-!$$$C                                                                       MAIN  50
-!$$$C THIS PROGRAM IS A TEST DRIVER FOR THE SHAPE PRESERVING QUADRATIC      MAIN  60
-!$$$C SPLINES BY D.F.MCALLISTER AND J.A.ROULIER.                            MAIN  70
-!$$$C                                                                       MAIN  80
-!$$$C ON INPUT--                                                            MAIN  90
-!$$$C                                                                       MAIN 100
-!$$$C   X CONTAINS THE ABSCISSAS OF THE POINTS OF INTERPOLATION.            MAIN 110
-!$$$C                                                                       MAIN 120
-!$$$C   Y CONTAINS THE ORDINATES OF THE POINTS OF INTERPOLATION.            MAIN 130
-!$$$C                                                                       MAIN 140
-!$$$C   N IS THE NUMBER OF DATA POINTS.                                     MAIN 150
-!$$$C                                                                       MAIN 160
-!$$$C   K IS THE NUMBER OF POINTS AT WHICH THE SPLINE IS TO BE EVALUATED.   MAIN 170
-!$$$C                                                                       MAIN 180
-!$$$C                                                                       MAIN 190
-!$$$C UPON EXIT FROM SUBROUTINE 'SLOPES'--                                  MAIN 200
-!$$$C                                                                       MAIN 210
-!$$$C   M CONTAINS THE COMPUTED FIRST DERIVATIVES AT EACH DATA POINT.       MAIN 220
-!$$$C                                                                       MAIN 230
-!$$$C-----------------------------------------------------------------------MAIN 240
[…]

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/TOMS_770/bvspis/bvspis.f

@@ -1,3183 +1,0 @@
-        SUBROUTINE DBVSIS (X,Y,NP,N,K,OPT,D0,DNP,D20,D2NP,CONSTR,EPS,
-     *                     BETA,BETAI,RHO,RHOI,KMAX,MAXSTP,XTAB,NTAB,
-     *                     SBOPT,Y0OPT,Y1OPT,Y2OPT,ERRC,ERRE,D,D2,
-     *                     DIAGN,Y0TAB,Y1TAB,Y2TAB,WORK,NWORK)
-
-C  DBVSIS is merely a support routine which calls DBVSSC and DBVSSE
-C  for the computation of the needed parameters and for the evaluation
-C  of a shape-preserving, C(k), k=1,2 , interpolating spline,
-C  optionally subject to boundary conditions.
-C  The use of DBVSIS is not recommended when more evaluations of the
-C  same spline are required; in this case it is better to separately
-C  call DBVSSC and then DBVSSE repeatedly.
-C  For an explanation of input and output parameters, the user is
-C  referred to the comments of DBVSSC and DBVSSE.
-
-
-        EXTERNAL BETA,BETAI,RHO,RHOI
-
-        INTEGER NP,N,K,OPT,CONSTR(0:NP-1),NTAB,KMAX,MAXSTP,SBOPT,
-     *          Y0OPT,Y1OPT,Y2OPT,DIAGN(0:NP-1),ERRC,ERRE,NWORK
-
-        DOUBLE PRECISION X(0:NP),Y(0:NP),D0,DNP,D20,D2NP,XTAB(0:NTAB),
-     *                   EPS,BETA,BETAI,RHO,RHOI,D(0:NP),D2(0:NP),
-     *                   Y0TAB(0:NTAB),Y1TAB(0:NTAB),Y2TAB(0:NTAB),
-     *                   WORK(1:NWORK)
-
-
-        CALL DBVSSC (X,Y,NP,N,K,OPT,D0,DNP,D20,D2NP,CONSTR,EPS,BETA,
-     *               BETAI,RHO,RHOI,KMAX,MAXSTP,ERRC,D,D2,DIAGN,
-     *               WORK,NWORK)
-
-        CALL DBVSSE (X,Y,NP,N,K,XTAB,NTAB,SBOPT,Y0OPT,Y1OPT,Y2OPT,
-     *               ERRC,D,D2,Y0TAB,Y1TAB,Y2TAB,ERRE,WORK,NWORK)
-
-        RETURN
-        END
-
[…]

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/l2_minimizer/monotone_fit_python_wrapper.py

@@ -1,0 +1,529 @@
+'''This Python code is an automatically generated wrapper
+for Fortran code made by 'fmodpy'. The original documentation
+for the Fortran source code follows.
+
+
+'''
+
+import os
+import ctypes
+import numpy
+
+# --------------------------------------------------------------------
+#               CONFIGURATION
+# 
+_verbose = True
+_fort_compiler = "gfortran"
+_shared_object_name = "monotone_fit.so"
+_this_directory = os.path.dirname(os.path.abspath(__file__))
+_path_to_lib = os.path.join(_this_directory, _shared_object_name)
+_compile_options = ['-fPIC', '-shared', '-O3', '-lblas', '-llapack']
+_ordered_dependencies = ['monotone_fit.f90', 'monotone_fit_c_wrapper.f90']
+# 
+# --------------------------------------------------------------------
+#               AUTO-COMPILING
+#
+# Try to import the existing object. If that fails, recompile and then try.
+try:
+    clib = ctypes.CDLL(_path_to_lib)
+except:
+    # Remove the shared object if it exists, because it is faulty.
+    if os.path.exists(_shared_object_name):
+        os.remove(_shared_object_name)
+    # Compile a new shared object.
+    _command = " ".join([_fort_compiler] + _compile_options + ["-o", _shared_object_name] + _ordered_dependencies)
+    if _verbose:
+        print("Running system command with arguments")
+        print("  ", _command)
+    # Run the compilation command.
+    import subprocess
+    subprocess.run(_command, shell=True, cwd=_this_directory)
+    # Import the shared object file as a C library with ctypes.
+    clib = ctypes.CDLL(_path_to_lib)
+# --------------------------------------------------------------------
+
+
+# ----------------------------------------------
+# Wrapper for the Fortran subroutine EVAL_BSPLINE
+
+def eval_bspline(t, xy, d=None):
+    '''! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                        EVAL_BSPLINE.f90
+!
+! DESCRIPTION:
+!   This file defines a subroutine EVAL_BSPLINE for computing the
[…]

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/schumaker/__init__.py

@@ -1,0 +1,34 @@
+
+# Construct and return a Schumaker spline fit to data.
+def spline_fit(x, y):
+    import numpy as np
+    # Update the "EDITOR" environment variable, otherwise rpy2 fails to
+    #  find standard libraries 'util' and 'stats'.
+    import os
+    os.environ.update({
+        'EDITOR': '/Applications/Emacs.app/Contents/MacOS/Emacs',
+    })
+    # Import rpy2 to call R packages from Python.
+    import rpy2
+    # For managing and executing R codes, as well as importing packages
+    from rpy2.robjects.packages import importr
+    # Import functions for converting to and from R vector formats.
+    from rpy2.robjects.numpy2ri import py2rpy, rpy2py
+    # Import the "schumaker" package from R.
+    schumaker = importr('schumaker')
+    # Calculate the response for the given x with R
+    edge_gradient = py2rpy(np.asarray([0.0, 0.0]))
+    Spline, DerivativeSpline, SecondDerivativeSpline, IntervalTab = (
+        schumaker.Schumaker(py2rpy(x), py2rpy(y), edgeGradients=edge_gradient, Vectorised=True)
+    )
+    # Construct python functions for evaluating the fit spline.
+    def second_derivative(x, ddf=SecondDerivativeSpline):
+        return rpy2py(ddf(py2rpy(x))).flatten()
+    def first_derivative(x, df=DerivativeSpline):
+        return rpy2py(df(py2rpy(x))).flatten()
+    def function(x, f=Spline):
+        return rpy2py(f(py2rpy(x))).flatten()
+    function.derivative = first_derivative
+    first_derivative.derivative = second_derivative
+    # Return the function (with derivatives).
+    return function

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/response.tex
Finished review draft, sending to coauthors.

@@
 Lagrangian fashion (with an adaptive multiplier on monotonicity
-violations). The results we achieved were quite mixed and we've
-incorporated our commentary from this excursion into Section 5 of the
-manuscript. In truth, formally expanding on this trajectory would
-require an entirely new manuscript with deeper study of the problem,
-while it is not immediately obvious that the results would be {\it
-  better} than the present approach. The global $L^2$ minimizer of the
-second derivative is simply the unique $C^2$ cubic spline interpolant,
-but that is known to produce large and unnecessary local variations in
-the approximation (along with not being shape preserving). Included in
-this response (to reviewers only) are some of the preliminary visuals
-we produced in exploring this approach. This is by no means final and
-conclusive, but the results were mixed enough for us to warrant not
-including them in the paper.
+violations). The results we achieved were quite mixed and we hestiate
+to incorporate commentary from this excursion into the manuscript. In
+truth, formally expanding on this trajectory would require an entirely
+new manuscript with deeper study of the problem, while it is not
+immediately obvious that the results would be {\it better} than the
+present approach. The global $L^2$ minimizer of the second derivative
+is simply the unique $C^2$ cubic spline interpolant, but that is known
+to produce large and unnecessary local variations in the approximation
+(along with not being shape preserving). Included in this response (to
+reviewers only) are some of the preliminary visuals we produced in
+exploring this approach. This is by no means final and conclusive, but
+the results were mixed enough for us to warrant not including them in
+the paper.
 
@@
 very specific system requirements to operate effectively. Namely: (1)
-running a POSIX shell (on macOS, Ubuntu, $\ldots$); (2) having
-{\ttVIII python3} of at least version $3.7$ and the packages {\ttVIII
-  numpy}, {\ttVIII tlux}, and {\ttVIII rpy2}; (3) having installed
-{\ttVIII gfortran} (or otherwise reconfiguring the codes to use
-another Fortran compiler of choice); and (4) having installed {\ttVIII
-  R} (for Schumaker splines). The Fortran codes associated with this
-work as well as those from TOMS 574 (quadratic $C^1$) and BVSPIS from
-TOMS 770 are all wrapped to be callable directly from {\ttVIII
-  python3}, which will be the parent process for all calls. A simple
-test usage can be seen in the provided {\ttVIII python3} file {\ttVIII
-  test.py}. To be abundantly clear, this code is {\bf not} intended
-for review and publication, but rather convenient access to various
-codes for reviewers to use for visualizing and comparing results.
+running a POSIX shell (on macOS, Ubuntu, $\ldots$); (2) having {\tt
+  python3} of at least version $3.7$ and the packages {\tt numpy},
+{\tt tlux}, and {\tt rpy2}; (3) having installed {\tt gfortran} (or
+otherwise reconfiguring the codes to use another Fortran compiler of
+choice); and (4) having installed {\tt R} (for Schumaker splines). The
+Fortran codes associated with this work as well as those from TOMS 574
+(quadratic $C^1$) and BVSPIS from TOMS 770 are all wrapped to be
+callable directly from {\tt python3}, which will be the parent process
+for all calls. A simple test usage can be seen in the provided {\tt
+  python3} file {\tt test.py}. To be abundantly clear, this code is
+{\bf not} intended for review and publication, but rather convenient
+access to various codes for reviewers to use for visualizing and
+comparing results.
 

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Finished review draft, sending to coauthors.

--- 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
@@
 [Moler 2008], however this method also yields approximations whose
-second derivative functions often have large $L^2$ norm (approximations
-with large {\it wiggle}). In an attempt to capture the local shape of
-the data while minimizing wiggle, this package uses a facet model from
-image processing [Haralick and Watson 1981] to estimate first and
-second derivatives at breakpoints. Rather than picking a local linear
-or quadratic fit with minimal residual, this work uses a quadratic
-facet model that selects the local quadratic interpolant with minimum
-magnitude curvature.
+second derivative functions often have large \red{local} $L^2$ norm
+(approximations with large {\it wiggle}). In an attempt to capture the
+local shape of the data while minimizing wiggle, this package uses a
+facet model from image processing [Haralick and Watson 1981] to
+estimate first and second derivatives at breakpoints. Rather than
+picking a local linear or quadratic fit with minimal residual, this
+work uses a quadratic facet model that selects the local quadratic
+interpolant with minimum magnitude \red{second derivative}.
 
@@
 where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
-\le n$, and $n \ge 3$. Returns the slope and curvature at $X_i$ of the
-local quadratic interpolant with minimum magnitude curvature.
+\le n$, and $n \ge 3$. Returns the \red{first and second derivative} at 
+$X_i$ of the local quadratic interpolant with minimum magnitude
+second derivative. \red{Approximate equality is denoted with
+$\approx$ and considers two numbers to be equal once they are within
+the machine precision $\epsilon$ of each other.}
 
@@
 \itemitem{} The point $(X_i$, $Y_i)$ is an extreme point. The
-quadratic with minimum magnitude
-curvature that has slope zero at $X_i$ will be the facet chosen.
+quadratic with minimum magnitude \red{second derivative} that has slope
+zero at $X_i$ will be the facet chosen.
 \itemitem{} $f_1\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$, $(X_i,Y_i)$,
@@
 not exist because $i \in \{2, n-1\}$. In those cases, the minimum
-magnitude curvature among existing quadratics is chosen.}
+magnitude \red{second derivative} among existing quadratics is chosen.}
 \itemitem{} $f_1\:\hbox{interpolant to }(X_{i-2},Y_{i-2})$,
@@
 \nobreak
+
 where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
@@
 $Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
-increasing on $[x_0,x_1]$.
+increasing on $[x_0,x_1]$. \red{Approximate equality is denoted with
+$\approx$ and considers two numbers to be equal once they are within
+the machine precision $\epsilon$ of each other.}
 
@@
   \sqrt{\alpha \delta}$. If none of the monotonicity conditions are
-  violated, then the order six piece is monotone and Step 12
+  violated, then the \red{degree five} piece is monotone and Step 12
   concludes.}
@@
 
-Given the minimum magnitude curvature nature of the initial derivative
-estimates, it is desirable to make the smallest necessary changes to
-the initial interpolating spline $Q$ while enforcing monotonicity. In
-practice a binary search for the boundary of monotonicity is used in
-place of solely shrinking $DQ$ and $D^2Q$ at breakpoints adjoining
-{\it active\/} intervals: intervals over which $Q$ is nonmonotone at
-least once during the search. The binary search considers a Boolean
-function $B_i(s)$, for $0 \le s \le 1$, that is true if the order six
-polynomial piece of $Q(x)$ on $[X_i, X_{i+1}]$ matching derivatives
+Given the minimum \red{curvature (minimum magnitude of the second derivative)}
+nature of the initial derivative estimates, it is desirable to make
+the smallest necessary changes to the initial interpolating spline $Q$
+while enforcing monotonicity. In practice a binary search for the
+boundary of monotonicity is used in place of solely shrinking $DQ$ and
+$D^2Q$ at breakpoints adjoining {\it active\/} intervals: intervals
+over which $Q$ is nonmonotone at least once during the search. The
+binary search considers a Boolean function $B_i(s)$, for $0 \le s \le
+1$, that is true if the order six polynomial piece of $Q(x)$ on
+$[X_i, X_{i+1}]$ matching derivatives
 $Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and 
 derivatives $Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, 
-$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false otherwise. 
-The binary search is only applied at those breakpoints adjoining intervals
-$[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and hence $B_i(1)$ is false.
-It is further assumed that there exists $0 \le s^* \le 1$ such that $B_i(s)$
-is true for $0 \le s \le s^*$ and false for some $1 > s > s^*$. Since the
-derivative conditions at interior breakpoints are shared by intervals left
-and right of the breakpoint, the binary search is performed at all
-breakpoints simultaneously.  Specifically, the monotonicity of $Q$ is
-checked on all active intervals in each step of the binary search to
-determine the next derivative modification at each breakpoint. The goal
-of this search is to converge on the boundary of the monotone region in
-the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
+$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false
+otherwise.  The binary search is only applied at those breakpoints
+adjoining intervals $[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and
+hence $B_i(1)$ is false.  It is further assumed that there exists
+$0 \le s^* \le 1$ such that $B_i(s)$ is true for $0 \le s \le s^*$ and
+false for some $1 > s > s^*$. Since the derivative conditions at
+interior breakpoints are shared by intervals left and right of the
+breakpoint, the binary search is performed at all breakpoints
+simultaneously.  Specifically, the monotonicity of $Q$ is checked on
+all active intervals in each step of the binary search to determine
+the next derivative modification at each breakpoint. The goal of this
+search is to converge on the boundary of the monotone region in the
+$(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
 Watson [1994]) for all intervals. This multiple-interval binary search
-allows the value zero to be obtained for all (first and second) derivative
-values in a fixed maximum number of computations, hence has no effect on
-computational complexity order. This binary search algorithm is outlined
-below.
+allows the value zero to be obtained for all (first and second)
+derivative values in a fixed maximum number of computations, hence has
+no effect on computational complexity order. This binary search
+algorithm is outlined below.
 

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Finished review draft, sending to coauthors.

@@
 \input colordvi
-\def\BEGINC{\textPeach} \def\ENDC{\textBlack}
+\def\BEGINC{\textRed}
+\def\ENDC{\textBlack}
 \def\red#1{\BEGINC#1\ENDC}
@@
 {\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
-monotone quintic spline interpolants to univariate monotone data. Using sharp
+monotone quintic spline interpolants to \red{univariate} monotone data. Using sharp
 theoretical monotonicity constraints, first and second derivative estimates
 at data provided by a quadratic facet model are refined to produce a
-univariate $\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
+\red{univariate} $\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
 details, complexity and sensitivity analyses, usage information, a

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
Finished review draft, sending to coauthors.

@@
 differentiable, which is desirable.
+
+\beginred
 
@@
 
-The theory for sufficient and {\it sharp} necessary bounds for
+\endred
+
+The theory for sufficient and \red{{\it sharp}} necessary bounds for
 monotone quintic interpolation has been provided [Ulrich and Watson
@@
 
+\beginred
+
 \heading{Overview}
+
+\endred
 
@@
 sensitivity of the algorithms in {\tt MQSI}, and Section 5 presents 
-timing data and some graphs of constructed interpolants.
+timing data and some graphs of constructed interpolants \red{as well
+  as a visual comparison with existing monotone spline packages}.

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
Finished review draft, sending to coauthors.

--- 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
@@
 \heading{4. COMPLEXITY AND SENSITIVITY}
-Algorithms 1, 2, and 4 have ${\cal O}(1)$ runtime. Given a fixed
+
+Algorithms \red{1 and 4 have ${\cal O}(n)$ runtime for $n$ data
+points. Algorithm 2 has a fixed cost ${\cal O}(1)$.} Given a fixed
 schedule for shrinking derivative values, Algorithm 3 has a ${\cal
@@
 behavior while empirically producing final approximations with less
-wiggle ($L^2$ norm of the second derivative) than other methods. A
+wiggle (\red{local} $L^2$ norm of the second derivative) than other methods. A
 weighted harmonic mean estimate of first derivatives may be more
@@
   $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
-  curvature. Given the exact five data points seen above, the quadratic
-  facet model produces the slope seen in the solid blue line at $x = 5/2$.
-  However, by subtracting the value of $f_3$ $= \epsilon(x-2)^2$ from
-  points at $x = 3$, 4, where $\epsilon$ is the machine precision
-  ($2^{-52}$ for an IEEE 64-bit real), the quadratic facet model produces
-  the slope seen in the dashed red line at $x = 5/2$. This is the nature
-  of a facet model and a side effect of associating data with local facets.
+  curvature \red{(equal second derivatives)}. Given the exact five data points
+  seen above, the quadratic facet model produces the slope seen in the
+  solid blue line at $x = 5/2$. However, by subtracting the value of
+  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
+  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
+  quadratic facet model produces the slope seen in the dashed red line
+  at $x = 5/2$. This is the nature of a facet model and a side effect
+  of associating data with local facets.
 \par} \endinsert

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
Finished review draft, sending to coauthors.

--- 2021-04-22 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
@@
 
-This section contains graphs of sample {\tt MQSI} results given various
-data configurations. Execution times for Algorithms 1 and 4 are given; the
-total execution time of {\tt MQSI} is utterly dominated by the time for a
-banded linear system solve computing the $3n$ $B$-spline coefficients,
-which is ${\cal O}(n)$.  The files {\tt sample\_main.f90} and {\tt
-sample\_main.dat} accompanying the subroutine {\tt MQSI} illustrate Fortran
-2003 subroutine usage with optional arguments and data points from a file.
-Compilation instructions and the full package contents are specified in
-the {\tt README} file.
-
 \topinsert
+\textBlack
 \centerline{\epsfxsize=4truein \epsffile{vis/2-piecewise-polynomial.eps}}
@@
 shown in the red dashed line, while the solid line with stylized
-markers denotes the approximation made with MQSI given equally spaced
-empirical distribution points from a sample of size 100.
+markers denotes the approximation made with {\tt MQSI} given
+equally spaced empirical distribution points from a sample of size 100.
 \par}
@@
 {\everymath={\scriptstyle} \narrower\noindent\rmVIII Fig.\ 5.
-Median total runtime of the routine {\ttVIII MQSI} for number of data points
+Median total runtime of the routine {\tt MQSI} for number of data points
 $n=10$
@@
 \endinsert
+
+This section contains graphs of sample {\tt MQSI} results given various
+data configurations. Execution times for Algorithms 1 and 4 are given; the
+total execution time of {\tt MQSI} is utterly dominated by the time for a
+banded linear system solve computing the $3n$ $B$-spline coefficients,
+which is ${\cal O}(n)$.  The files {\tt sample\_main.f90} and {\tt
+sample\_main.dat} accompanying the subroutine {\tt MQSI} illustrate Fortran
+2003 subroutine usage with optional arguments and data points from a file.
+Compilation instructions and the full package contents are specified in
+the {\tt README} file.
 
@@
 
-Comparisons between this work and lower order methods are problematic,
-especially with sparse data. For exact data from smooth functions,
-comparing derivative values, higher order methods always win (all
-other things being equal) as the data density increases. For sparse or
-synthetic data, what constitutes a better fit is either subjective or
-dependent on the specific problem. If an application demands $C^2$
-continuity and monotonicity, the target of {\tt MQSI}, then this
-package provides such interpolants while no existing published spline
-software suffices.
+\beginred
+
+The binary search procedure provably converges on the boundary of the
+region of monotonicity precisely defined by Ulrich and Watson [1994]
+through the application of a boolean function that is guaranteed to be
+true at one end of an interval. Since a local quadratic interpolant is
+applied for initial function derivative estimates, the approximation
+order of the resulting fit is $O(h^3)$ (as for any second order
+approximation), but it should be noted that asymptotic approximation
+order is rarely of import when considering the scattered sparse data
+that $C^2$ approximations like {\tt MQSI} provide. Were exact
+evaluations provided, higher order methods always win \textRed (all
+other things being equal) as the data density increases. For sparse
+data, what constitutes a better fit is either subjective or dependent
+on the problem.
+
+Lastly, comparisons between {\tt MQSI} and four published spline
+software packages are provided that allow subjective inspection on the
+same four test problems as presented above. In each of Figures 5
+through 8 the intricacies of the approximation created by {\tt MQSI}
+can be contrasted with two monotone $C^1$ quadratic splines (TOMS 574,
+Schumaker), the popular $C^1$ cubic method {\tt PCHIP}, and the $C^2$
+quintic method {\tt BVSPIS} of TOMS 770. Figures 5 and 8 demonstrate
+glaring numerical errors in the Schumaker and {\tt BVSPIS}
+implementations, and overall it can be observed that the minimal and
+tight conditions on monotonicity provided by TOMS 574 tend to be more
+visually appealing than the weaker sufficient conditions utilized by
+{\tt PCHIP}. In general the major determining factor for the quality
+of {\tt MQSI} is the local quadratic facet model. For applications
+where the underlying function is presumed to be a piecewise polynomial
+(of relatively low order), the choice of local quadratic interpolants
+is reasonable. However the consequence of this decision is that
+functions with superquadratic rates of change will tend to have
+consistently underestimated first and second derivatives. This
+limitation is accepted in favor of perfectly capturing lower order
+phenomenon. Were future research to apply alternative initializations
+for $C^2$ quintic splines, those could comfortably be made monotone by
+the procedures of {\tt MQSI} and most importantly by the application
+of these sharp monotonicity conditions.
+
+\endred
+
+If an application demands $C^2$ continuity and monotonicity, the
+target of {\tt MQSI}, then this package \red{uniquely} provides such
+interpolants \red{based on sharp monotonicity constraints.}
+
+\goodbreak
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/4-bvspis.eps}}
+{\narrower\noindent\rmVIII Fig.\ 5.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), {\tt PCHIP} (third, green), and {\tt BVSPIS} (fourth, purple)
+respectively on the {\itVIII piecewise polynomial} test function. {\tt MQSI}
+is styled as a gray thin line in the background for comparison. The
+Schumaker code (second) fails to produce a monotone approximation for
+this problem while producing no errors or warnings. The primary
+comparative observation of {\tt MQSI} otherwise is its exact reproduction of
+the linear segment between $0.3$ and $0.5$.
+\par}
+\endinsert
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/4-bvspis.eps}}
+{\narrower\noindent\rmVIII Fig.\ 6.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), {\tt PCHIP} (third, green), and {\tt BVSPIS} (fourth, purple)
+respectively on the {\itVIII large tangent} test function. {\tt MQSI} is
+styled as a gray thin line in the background for comparison. This
+problems highlights the core weakness of the quadratic facet model
+approach to derivative estimation in {\tt MQSI}, which consistently
+underestimates the actual curvature of this function. This weakness is
+accepted for its greater accuracy when estimating local lower order
+components of functions.
+\par}
+\endinsert
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/4-bvspis.eps}}
+{\narrower\noindent\rmVIII Fig.\ 7.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), {\tt PCHIP} (third, green), and {\tt BVSPIS} (fourth, purple)
+respectively on the {\itVIII signal decay} test function. {\tt MQSI} is
+styled as a gray thin line in the background for comparison. The most
+notable differences between {\tt MQSI} and other approaches can be observed
+near local extrema, where {\tt MQSI} produces smaller magnitude second
+derivatives.
+\par}
+\endinsert
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/4-bvspis.eps}}
+{\narrower\noindent\rmVIII Fig.\ 8.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), {\tt PCHIP} (third, green), and {\tt BVSPIS} (fourth, purple)
+respectively on the {\itVIII random monotone} test function. {\tt MQSI} is
+styled as a gray thin line in the background for comparison. Notice
+that the numerical conditions for this approximation problem are
+challenging enough that both the Schumaker and {\tt BVSPIS} codes
+incorrectly produce nonmonotone segments (neither of which produced
+error codes or any form of indication that a failure
+occurred). Notably only {\tt MQSI} and TOMS 574 satisfy tight monotonicity
+constraints while also correctly preserving monotonicity in their
+approximations. The main segment of divergence between the methods
+happens on the $(0.45, 0.5)$ interval, where {\tt MQSI} favors maximizing
+the slope on the left side of the interval because the quadratic
+interpolant of the right sided points produces an (invalid) negative
+slope estimate on the left side of the interval.
+\par}
+\endinsert

### 2022-03-12 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Added some missing red, moved location of visual files to be in the paper directory.

@@
 details, complexity and sensitivity analyses, usage information, a
-brief performance study, and comparisons with other spline approaches are included.
+brief performance study, \red{and comparisons with other spline approaches} are included.
 

### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/revisions/MQSI_testing/mqsi/__init__.py
All testing code and last comments implemented.

@@ -11,7 +11,7 @@ import numpy
 #               CONFIGURATION
 # 
 _verbose = True
-_link_lapack = True
+_link_lapack = False
 _fort_compiler = "gfortran"
 _shared_object_name = "MQSI.so"
 _this_directory = os.path.dirname(os.path.abspath(__file__))

### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/revisions/2022-03/response.tex
All testing code and last comments implemented.

--- 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/response.tex
@@
 \font\titlett=cmtt20
-\font\ttVIII=cmtt8
+%% Heading font.
+\font\headingbf=cmbx14
+\font\headingtt=cmtt14
+%% Caption font.
+\font\caprm=cmr8
+\font\captt=cmtt8
+\font\capit=cmmi8
+\font\capbf=cmbx8
+\font\capsy=cmsy8
+%% Caption text.
+\def\captext{
+  \narrower
+  \noindent
+  \caprm
+  \baselineskip=10pt plus 2pt minus 1pt
+  \lineskiplimit=2pt
+  \lineskip=2pt plus 1pt
+  \parskip=0pt
+  \textfont0=\caprm
+  \textfont1=\capit
+  \textfont2=\capsy
+}
 
@@
   and justify the differences between this work and the many prior
-  works on $C^2$ interpolation. [given 23 references]
+  works on $C^2$ interpolation. [given 22 references]
 
@@
 We appreciate this perspective and have included such comparisons in
-the updated Section 5. The comparison shows methods from four
-published spline software packages against MQSI for the same test
+the updated Section 5 (Figures 5-8). The comparison shows methods from
+four published spline software packages against MQSI for the same test
 problems as originally in the paper (and associated test code). An
@@
 monotonicity are proven, we have added an additional paragraph to
-Section 5 explaining this fact.
+Section 4 explaining this fact.
 
@@
 
-In the same paragraph of Section 5 referenced above, we have explained
+In the same paragraph of Section 4 referenced above, we have explained
 that the approximation order of the method is that of any second order
 approximation (because a local quadratic interpolant is used to
-approximate initial derivative values), but also notes that the
+approximate initial derivative values), but also note that the
 asymptotic approximation order is often of little importance in the
@@
 The definition of approximately equal here is when two floating point
-numbers are within the machine epsilon (at their scale) of each other.
+numbers are within the machine precision (at their scale) of each other.
 This is a rather standard approach for ensuring safe separation of
@@
 manuscript and this comment reflected our original thoughts. In due
-response to this we've implemented naive $L2$ minimization algorithm
+response to this we've implemented a naive $L^2$ minimization algorithm
 over the second derivative that uses gradient descent over the spline
@@
 Lagrangian fashion (with an adaptive multiplier on monotonicity
-violations). The results we achieved were quite mixed and we hestiate
+violations). The results we achieved were quite mixed and we hesitate
 to incorporate commentary from this excursion into the manuscript. In
@@
 
+\vfill
+
 \centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/5-l2_minimizer.eps}}
@@
 \centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/5-l2_minimizer.eps}}
-{\narrower\noindent {\bf $L^2$ Minimization Experiment }
-{\ttVIII MQSI} (light gray) compared with an experimental global $L^2$
-minimization approach (blue) on each of the test problems from the
-paper. It is {\it very} difficult to consistently construct fits that
-lack local oscillations as well as maintain monotonicity with this
-approach. The method relies on the B-spline basis as well, which
-affects the bias towards specific solutions when following the second
-derivative $L^2$ minimizing gradient.  This work would require a very
-significant line of research to properly resolve all these small
-issues in a theoretically sound way.
+{\captext {\capbf $L^2$ Minimization Experiment.} {\captt MQSI} (light
+  gray) compared with an experimental global $L^2$ minimization
+  approach (blue) on each of the test problems from the paper. It is
+  {\capit very} difficult to consistently construct fits that lack
+  local oscillations (i.e., minimize $L^2$) as well as maintain
+  monotonicity with this approach. The method relies on the B-spline
+  basis in this experiment, which affects the bias towards specific
+  solutions when following the second derivative $L^2$ minimizing
+  gradient. This work would warrant a significant line of research to
+  properly resolve these issues in a theoretically sound way.
 \par}
@@
 As mentioned in the manuscript Section 5 paragraph 4, the banded
-linear system representation is favored for two reasons in this
-package. The computational complexity order is unchanged by solving a
-single banded system, while the global numerical stability is {\it
+linear system representation is favored for it's decreased redundancy.
+The computational complexity order is unchanged by solving a single
+banded system, while the global numerical stability is {\it
   theoretically} improved. The system of equations and number of
 nonzero splines at each breakpoint is thoroughly documented in
-{\ttVIII SPLINE.f90}, lines 172--206. In practice this approach is
+{\tt SPLINE.f90}, lines 172--206. In practice this approach is
 almost certainly slower than solving individual linear systems, but
@@
 does not increase computational complexity). For users interested in
-constructing alternative representations, the optional argument {\it
-  \bf UV} documented in the code exists precisely for returning the
+constructing alternative representations, the optional argument {\tt
+  UV} documented in the code exists precisely for returning the
 computed first and second derivatives at knots.
@@
 access to various codes for reviewers to use for visualizing and
-comparing results.
+comparing results. Precise installation command line instructions are
+provided at the end of this document.
 
@@
 \red{red}. We hope that you all find these revisions to have improved
-the manuscript, and we eagerly await more feedback.
+the manuscript, and we look forward to your response.
 
@@
 Thomas Lux, on behalf of all authors.
+
+\vfill
+
+{\headingbf Installing dependencies and running tests in} {\hskip 2mm} {\headingtt MQSI\_testing.zip}
+
+\vskip 5mm
+
+{\tt Ubuntu 20+} command line instructions.
+
+{\parskip=2pt \parindent=20pt \tt
+\item{} \#() \string{ echo "\# \$@" \string}
+\item{} \textMaroon \# install gfortran \textBlack
+\item{} sudo apt install gfortran
+}
+
+\vskip 5mm
+
+{\tt macOS 12+} command line instructions (installs CommandLineTools and Homebrew first).
+
+{\parskip=2pt \parindent=20pt \tt
+\item{} \#() \string{ echo "\# \$@" \string}
+\item{} \textMaroon \# install CommandLineTools if you do not have it already \textBlack
+\item{} xcode-select -p || xcode-select --install
+\item{} \textMaroon \# install Homebrew for package management \textBlack
+\item{} brew --version || /bin/bash -c "\$(curl -fsSL $\backslash$
+\item{} \hskip 10mm https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
+\item{} \textMaroon \# install gfortran \textBlack
+\item{} brew install gfortran
+}
+
+\vskip 5mm
+
+Common instructions: Python package installation, unzipping archive, and running {\hskip 1mm} {\tt test.py}.
+
+{\parskip=2pt \parindent=20pt \tt
+\item{} \textMaroon \# install a visualization python package \textBlack
+\item{} python3 -m pip install --user tlux
+\item{}
+\item{} \textMaroon \# unzip archive, enter resulting directory, run tests \textBlack
+\item{} unzip MQSI\_testing.zip
+\item{} cd MQSI\_testing
+\item{} python3 test.py
+\item{}
+\item{} \textMaroon \# install R (if comparisons with Schumaker are desired) \textBlack
+\item{} \textMaroon \# install Schumaker package through R graphical user interface \textBlack
+\item{} python3 -m pip install --user rpy2
+\item{}
+\item{} \textMaroon \# make any desired modifications to test data at bottom of \textBlack test.py 
+}
 

### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
All testing code and last comments implemented.

--- 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
@@
 (splines) provide the smooth approximations for animation in graphics
-[Herman et al.\ 2015; Quint 2003], aesthetic structural support in
+[Herman and Oftedal\ 2015; Quint 2003], aesthetic structural support in
 architecture [Brennan 2020], efficient aerodynamic surfaces in
 automotive and aerospace engineering [Brennan 2020], prolonged
-effective operation of electric motors [Berglund et al.\ 2009], and
+effective operation of electric motors [Berglund et\ al.\ 2009], and
 accurate nonparametric approximations in statistics [Knott
@@
 Gregory 1985]. The specific properties the present algorithm will
-preserve in approximations are monotonicity and $C^2$ continuity. In
-addition to previously mentioned applications, these properties are
+preserve in approximations are monotonicity and $C^2$ continuity. 
+\beginred
+Notably this work does not consider convexity preservation, as Mulansky
+and Neamtu [1991] prove that there exists strictly convex data sets which
+do not allow convex interpolation with $C^2$ splines of fixed degree.
+\endred
+In addition to previously mentioned applications, \red{being $C^2$ and monotonicity preserving is}
 crucially important in statistics to the approximation of a cumulative
@@
 
+\beginred
+
 The existing research in shape preserving interpolatory splines is
@@
 levels of continuity is often more difficult than lower levels of
-continuity. And while all referenced algorithms establish sufficient
-conditions for shape preservation, it is less common and much more
-difficult to provide sharp {\it necessary} conditions for
-monotonicity. The following two paragraphs highlight a collection of
-research related to the construction of sufficient conditions for
-$C^1$ and $C^2$ shape preserving splines. From this foundation, the
-value and utility of the code presented here can be more readily
-appreciated.
+continuity. Furthermore, all referenced algorithms establish
+sufficient conditions for shape preservation, but it is less common
+and much more difficult to provide {\it sharp and necessary}
+conditions for monotonicity. The following two paragraphs highlight a
+collection of research related to the construction of $C^1$ and $C^2$
+shape preserving splines. From this foundation, the value and utility
+of the code presented here can be more readily appreciated.
 
@@
 accuracy over their nonmonotone counterparts when approximating
-monotone functions, while Constantini [1986] proves that monotone
+monotone functions, while Costantini [1986] proves that monotone
 spline interpolants to data of fixed continuity and order exist in
 general. This groundwork establishes the potential for monotone spline
-interpolation. Schumaker [1983] and McAllister and Roulier [1981]
+interpolation. Schumaker [1983] and McAllister and Roulier [1981a]
 (independently) establish necessary and sufficient conditions for
@@
 for monotone $C^1$ cubic splines that require no knot insertion.
-Gregory and Delbourgo [1983] prove necessary and sufficient conditions
+Gregory and Delbourgo [1982] prove necessary and sufficient conditions
 for a closed form solution to monotone $C^1$ quadratic rational spline
 interpolation with nonlinear boundary equations and an iterative
-approach. Delbourgo and Gregory [1993] extend that same work to
+approach. Delbourgo and Gregory [1983] extend that same work to
 achieve sufficient conditions for $C^2$ continuity with a cubic
@@
 preserving spline method while acknowledging that quintic splines are
-necessary for generally monotone $C^2$ spline interpolants that do not
-insert new knots. Similarly, Manni [1997] constructs sufficient
-conditions for a monotone $C^2$ cubic spline by adding two additional
-knots per interval, then Cravery and Manni [2003] extend that
-methodology to arrive at monotone $C^3$ interpolating splines by
+generally necessary for monotone $C^2$ spline interpolants that do not
+insert new knots. Similarly, Manni and Sablonni\`ere [1997] construct
+sufficient conditions for a monotone $C^2$ cubic spline by adding two
+additional knots per interval, then Cravero and Manni [2003] extend
+that methodology to arrive at monotone $C^3$ interpolating splines by
 progressively increasing tension through Bezier control
-polygons. Constantini [1997] provides a method for monotone $C^2$
+polygons. Costantini [1997] provides a method for monotone $C^2$
 quintic spline interpolation based on sufficient and necessary
 boundary value conditions, but the necessary conditions are not
-sharp. Dougherty, Edelman, and Hyman [1989] construct monotone $C^2$
-Hermite splines and in commentary recommend against the direct
-optimization for smoothness parameters like global $L^2$, rather
-promoting problem specific definitions of geometric niceness. Both
-Wang and Tan [2004] as well as Yao and Nelson [2018] propose
-sufficient conditions for $C^2$ quartic splines, while Piah and
-Unsworth [2011] improve upon those sufficient conditions for $C^2$
-quartic rational spline interpolation. Abbas, Majid, and Ali [2012]
-formulate a set of sufficient conditions for monotone $C^2$ cubic
-rational splines and similar work is extended to $C^2$ quartic
-rationals (and splines with arbitrary smoothness) by Zhu and Han
-[2015]. Ulrich and Watson [1994] arrived at sufficient and sharp
-necessary conditions for monotone $C^2$ quintic splines, as also did
-He{\ss} and Schmidt [1994] (independently). In addition, a statistical
-method for bootstrapping the construction of an arbitrarily smooth
-monotone fit exists [Leitenstorfer and Tutz 2006], but this is enabled
-by sufficient conditions only.
+sharp. Dougherty, et\ al.\ [1989] construct monotone $C^2$ Hermite
+splines and in commentary recommend against the direct optimization
+for smoothness parameters like global $L^2$, rather promoting problem
+specific definitions of geometric niceness. Both Wang and Tan [2004]
+as well as Yao and Nelson [2018] propose sufficient conditions for
+$C^2$ quartic splines, while Piah and Unsworth [2011] improve upon
+those sufficient conditions for $C^2$ quartic rational spline
+interpolation. Abbas, et\ al.\ [2012] formulate a set of sufficient
+conditions for monotone $C^2$ cubic rational splines and similar work
+is extended to $C^2$ quartic rationals (and splines with arbitrary
+smoothness) by Zhu and Han [2015a; 2015b]. Ulrich and Watson [1990]
+arrive at sharp necessary and sufficient conditions for monotone $C^2$
+quintic splines, while He{\ss} and Schmidt [1994] produce similar
+sufficient conditions for $C^2$ quintic splines. In addition, a method
+for constructing an arbitrarily smooth monotone fit exists
+[Leitenstorfer and Tutz 2006] as well as a method for arbitrarily
+smooth shape preservation [Han 2018], but these methods are enabled by
+sufficient conditions only.
 
 The currently available peer reviewed and published software for
-monotone piecewise polynomial interpolation is severely reduced in
+monotone piecewise polynomial interpolation is severely limited in
 comparison with the number of published approaches mentioned
@@
 Fritsch and Carlson [1980] is available through the SciPy Python
-package ($>1$ billion downloads), and the $C^1$ quadratic method of
-Schumaker [1983] is available as an R package (300 thousand
-downloads). The piecewise quadratic $C^1$ method of McAllister and
-Roulier [1981] is available in FORTRAN (1000 downloads) and the
-sufficient $C^2$ quintic method of Costantini [1997] BVSPIS in FORTRAN
-as well (500 downloads). Based on publicly available download records,
-it is assumed that the code by Fritsch [1982] for monotone $C^1$ cubic
-spline interpolation is the predominant code for constructing monotone
-interpolants at present.
+package ($>$1 billion downloads), and the $C^1$ piecewise quadratic
+method of Schumaker [1983] is available as an R package (300,000
+downloads). The $C^1$ piecewise quadratic method of McAllister and
+Roulier [1981a;\ 1981b] is available in FORTRAN (1000 downloads) and
+the sufficient $C^2$ piecewise quintic method of Costantini [1997a;
+  1997b] BVSPIS in FORTRAN as well (500 downloads). Based on publicly
+available download records, it is assumed that the code by Fritsch
+[1982] for monotone $C^1$ cubic spline interpolation is the
+predominant code for constructing monotone interpolants at present.
 
-The theory for sufficient and {\it sharp} necessary bounds for
-monotone quintic interpolation has been provided [Ulrich and Watson
-  1994; He{\ss} and Schmidt 1994] and that theory was recently
-utilized in a proposed algorithm [Lux 2020] for monotone quintic
-spline construction, however no published mathematical software exists
-for the quintic case based on sharp monotonicity conditions. The
-software presented here represents the first published software
-package for producing $C^2$ shape preserving splines based on sharp
-monotonicity conditions. This work improves upon the algorithm
-presented by Lux et al. [2020] by refactoring computations for
-improved numerical stability, estimating minimum magnitude second
-derivatives at breakpoints with a quadratic facet model, and using a
-binary search to reduce the magnitude of the modifications made to
-initial derivative estimates when constructing a monotone spline
-interpolant.
+\endred
+
+The theory for \red{{\it sharp}} necessary and sufficient bounds for
+monotone $C^2$ quintic interpolation has been provided [Ulrich and
+  Watson 1990; 1994] and that theory was recently utilized in a
+proposed algorithm [Lux et al. 2020; Lux 2020] for monotone $C^2$
+quintic spline construction, however no published mathematical
+software exists for the quintic case \red{based on sharp monotonicity
+conditions}. The software presented here represents the first published
+software package for producing $C^2$ shape preserving splines
+\red{based on sharp monotonicity conditions}. This work improves upon
+the algorithm presented by Lux et\ al.\ [2020] by refactoring
+computations for improved numerical stability, estimating minimum
+magnitude second derivatives at breakpoints with a quadratic facet
+model, and using a binary search to reduce the magnitude of the
+modifications made to initial derivative estimates when constructing a
+monotone spline interpolant.
+
+\beginred
 
@@
 
+\endred
+
 This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
 sharp necessary and sufficient conditions in Ulrich and Watson [1994]
-for the construction of monotone quintic spline interpolants of
+for the construction of monotone $C^2$ quintic spline interpolants of
 monotone data. Precisely, the problem is, given a strictly increasing
@@
 
-The remainder of this paper is structured as follows: Section 2 provides
-the algorithms for constructing a $C^2$ monotone quintic spline interpolant
-to monotone data, Section 3 outlines the method of spline representation
-($B$-spline basis) and evaluation, Section 4 analyzes the complexity and
-sensitivity of the algorithms in {\tt MQSI}, and Section 5 presents 
-timing data and some graphs of constructed interpolants.
+The remainder of this paper is structured as follows: Section 2
+provides the algorithms for constructing a monotone $C^2$ quintic
+spline interpolant to monotone data, Section 3 outlines the method of
+spline representation ($B$-spline basis) and evaluation, Section 4
+analyzes the complexity and sensitivity of the algorithms in {\tt
+  MQSI}, and Section 5 presents timing data and some graphs of
+constructed interpolants \red{as well as a visual comparison with
+  existing monotone spline packages}.

### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
All testing code and last comments implemented.

@@
 quadratic facet model are not guaranteed to produce monotone quintic
-polynomial segments. Ulrich and Watson [1994] established tight
-constraints on the monotonicity of a quintic polynomial piece, while
-deferring to He{\ss} and Schmidt [1994] for a relevant simplified
-case. The following algorithm implements a sharp check for
-monotonicity by considering the nondecreasing case. The nonincreasing
-case is handled similarly.
+polynomial segments. Ulrich and Watson [1990] established tight
+constraints on the monotonicity of a quintic polynomial piece and
+He{\ss} and Schmidt [1994] gave simpler conditions for a special case.
+The following algorithm implements a sharp check for monotonicity by
+considering the nondecreasing case. The nonincreasing case is handled
+similarly.
 

### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
Added some missing red, moved location of visual files to be in the paper directory.
All testing code and last comments implemented.

--- 2021-04-22 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
@@
 shown in the red dashed line, while the solid line with stylized
-markers denotes the approximation made with MQSI given equally spaced
-empirical distribution points from a sample of size 100.
+markers denotes the approximation made with {\ttVIII MQSI} given
+equally spaced empirical distribution points from a sample of size 100.
 \par}
@@
 \par}
-\iffalse
+\iffalse % ---- excluded timing visual (only appears in text) --------
 \vskip 15mm
@@
 \par}
-\fi
+\fi % ----------------------------------------------------------------
 \endinsert
+
+This section contains graphs of sample {\tt MQSI} results given various
+data configurations. Execution times for Algorithms 1 and 4 are given; the
+total execution time of {\tt MQSI} is utterly dominated by the time for a
+banded linear system solve computing the $3n$ $B$-spline coefficients,
+which is ${\cal O}(n)$.  The files {\tt sample\_main.f90} and {\tt
+sample\_main.dat} accompanying the subroutine {\tt MQSI} illustrate Fortran
+2003 subroutine usage with optional arguments and data points from a file.
+Compilation instructions and the full package contents are specified in
+the {\tt README} file.
 
@@
 
-Comparisons between this work and lower order methods are problematic,
-especially with sparse data. For exact data from smooth functions,
-comparing derivative values, higher order methods always win (all
-other things being equal) as the data density increases. For sparse or
-synthetic data, what constitutes a better fit is either subjective or
-dependent on the specific problem. If an application demands $C^2$
-continuity and monotonicity, the target of {\tt MQSI}, then this
-package provides such interpolants while no existing published spline
-software suffices.
+\beginred
+
+Lastly, comparisons between {\tt MQSI} and four published spline
+software packages are provided that allow subjective inspection on the
+same four test problems as presented above. In each of Figures 5
+through 8 the intricacies of the approximation created by {\tt MQSI}
+can be contrasted with two monotone $C^1$ quadratic splines (TOMS 574,
+and Schumaker), the popular $C^1$ piecewise cubic method {\tt PCHIP},
+and the $C^2$ piecewise quintic method {\tt BVSPIS} of TOMS
+770. Figures 5 and 8 demonstrate glaring numerical errors in the
+Schumaker and {\tt BVSPIS} implementations, and overall it can be
+observed that the minimal and tight conditions on monotonicity
+provided by TOMS 574 tend to be more visually appealing than the
+weaker sufficient conditions utilized by {\tt PCHIP}. In general the
+major determining factor for the quality of {\tt MQSI} is the local
+quadratic facet model. For applications where the underlying function
+is presumed to be a piecewise polynomial (of relatively low order),
+the choice of local quadratic interpolants is reasonable. However the
+consequence of this decision is that functions with superquadratic
+rates of change will tend to have consistently underestimated first
+and second derivatives. This limitation is accepted in favor of
+perfectly capturing lower order phenomena. Were future research to
+supply alternative initializations for $C^2$ quintic splines, those
+could comfortably be made monotone by the procedures of {\tt MQSI} and
+most importantly by the application of these sharp monotonicity
+conditions.
+
+\endred
+
+If an application demands $C^2$ continuity and monotonicity, the
+target of {\tt MQSI}, then this package \red{uniquely} provides such
+interpolants \red{based on sharp monotonicity constraints.}
+
+\goodbreak
+
+\beginred
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/4-bvspis.eps}}
+\beginred
+{\narrower\noindent\rmVIII Fig.\ 5.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), {\ttVIII PCHIP} (third, green), and {\ttVIII BVSPIS}
+(fourth, purple) respectively on the {\itVIII piecewise polynomial}
+test function. {\ttVIII MQSI} is styled as a gray thin line in the
+background for comparison. The Schumaker code (second) fails to
+produce a monotone approximation for this problem while producing no
+errors or warnings. The primary comparative observation of {\ttVIII
+  MQSI} otherwise is its exact reproduction of the linear segment
+between $0.3$ and $0.5$.
+\par}
+\endred
+\endinsert
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/4-bvspis.eps}}
+\beginred
+{\narrower\noindent\rmVIII Fig.\ 6.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), {\ttVIII PCHIP} (third, green), and {\ttVIII BVSPIS}
+(fourth, purple) respectively on the {\itVIII large tangent} test
+function. {\ttVIII MQSI} is styled as a gray thin line in the
+background for comparison. This problems highlights the core weakness
+of the quadratic facet model approach to derivative estimation in
+{\ttVIII MQSI}, which consistently underestimates the actual
+\red{second derivative} of this function. This weakness is accepted
+for its greater accuracy when estimating local lower order components
+of functions.
+\par}
+\endred
+\endinsert
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/4-bvspis.eps}}
+\beginred
+{\narrower\noindent\rmVIII Fig.\ 7.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), {\ttVIII PCHIP} (third, green), and {\ttVIII BVSPIS}
+(fourth, purple) respectively on the {\itVIII signal decay} test
+function. {\ttVIII MQSI} is styled as a gray thin line in the
+background for comparison. The most notable differences between
+{\ttVIII MQSI} and other approaches can be observed near local
+extrema, where {\ttVIII MQSI} produces continuous and smaller
+magnitude second derivatives.
+\par}
+\endred
+\endinsert
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/1-toms574.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/2-schumaker.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/3-pchip.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/4-bvspis.eps}}
+\beginred
+{\narrower\noindent\rmVIII Fig.\ 8.
+{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
+(second, red), {\ttVIII PCHIP} (third, green), and {\ttVIII BVSPIS}
+(fourth, purple) respectively on the {\itVIII random monotone} test
+function. {\ttVIII MQSI} is styled as a gray thin line in the
+background for comparison. Notice that the numerical conditions for
+this approximation problem are challenging enough that both the
+Schumaker and {\ttVIII BVSPIS} codes incorrectly produce nonmonotone
+segments (neither of which produced error codes or any indication that
+a failure occurred). Notably only {\ttVIII MQSI} and TOMS 574 satisfy
+tight monotonicity constraints while also correctly preserving
+monotonicity in their approximations, and only {\ttVIII MQSI} is
+$C^2$. The main segment of divergence between the methods happens on
+the $(0.45, 0.5)$ interval, where {\ttVIII MQSI} favors maximizing the
+slope on the left side of the interval because the quadratic
+interpolant of the right sided points produces a negative
+slope estimate (invalid) on the left side of the interval.
+\par}
+\endred
+\endinsert
+
+\endred

### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/revisions/MQSI_testing/test.py

@@ -1,0 +1,146 @@
+import numpy as np
+
+# Import all of the (python-wrapped) spline packages.
+from toms574 import spline_fit as toms_574_spline
+from schumaker import spline_fit as schumaker_spline
+from pchip import spline_fit as pchip_spline
+from bvspis import spline_fit as bvspis_spline
+from l2_minimizer import spline_fit as l2min_spline
+from mqsi import spline_fit as mqsi_spline
+
+
+# Given a spline fit function and some data, plot the fit over the
+#  interval covered by the data.
+def plot_spline_fit(x, y, spline_fit, name=None, p=None, color=None,
+                    dash=None, d1=False, d2=False, plot_points=5000):
+    spline = spline_fit(x, y)
+    # Evauate the spline.
+    zx = np.linspace(1,6, round(5 / 0.01))
+    zy = spline(zx)
+    dzy = spline.derivative(zx)
+    ddzy = spline.derivative.derivative(zx)
+    # Create a Plot object if one was not given.
+    if p is None:
+        p = Plot()
+    # Assign a color if none was given.
+    if color is None:
+        p.color_num += 1
+        color = p.color_num
+    # Assign a name if none was given.
+    if name is None:
+        name = str(spline_fit).split()[1].replace("_"," ").title()
+    # Get the bounds of the data to evaluate.
+    bounds = [x.min(), x.max()]
+    # Evaluate the fit and two derivatives.
+    p.add_func(f"{name}", spline, bounds, vectorized=True, color=color, group=color, dash=dash, plot_points=plot_points)
+    if d1:
+        p.add_func("First derivative", spline.derivative, bounds, vectorized=True,
+                   color=color, group=color, dash="dash", plot_points=plot_points)
+    if d2:
+        p.add_func("Second derivative", spline.derivative.derivative, bounds, vectorized=True,
+                   color=color, group=color, dash="dot", plot_points=plot_points)
+    # Return the plot.
+    return p
+
+
+# Show the various fits of the data.
+def show_fits(x, y, methods={}, d1=False, d2=False, **plot_kwargs):
+    # Ensure that the "x" and "y" are numpy arrays.
+    x = np.asarray(x, dtype=float).flatten()
+    y = np.asarray(y, dtype=float).flatten()
+    # Create a plot.
+    from tlux.plot import Plot
+    p = Plot()
[…]

### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/visuals/cli.f90

@@ -9,7 +9,7 @@
 !   to a file. Compile with:
 ! 
 !    $F03 $OPTS REAL_PRECISION.f90 EVAL_BSPLINE.f90 SPLINE.f90 MQSI.f90 \
-!       cli.f90 -o cli $LIB
+!       mqsi_cli.f90 -o mqsi $LIB
 !
 !   where '$F03' is the name of the Fortran 2003 compiler, '$OPTS' are
 !   compiler options such as '-O3', and '$LIB' provides a flag to link

### 2022-03-28 [2020-08]_ACMTOMS_(MQSI)/revisions/2022-03/response.tex
Updated response installation instructions, verified and tested on new computer.
Final for resubmission.

@@
 }
+%% Literal "underscore" character that can be copied.
+\font\underscorefont=ecrm1000
+\def\_{{\underscorefont\char"5F}}
 
@@
 
-\vskip 5mm
+\vskip 3mm
 
@@
 \item{} \textMaroon \# install Homebrew for package management \textBlack
-\item{} brew --version || /bin/bash -c "\$(curl -fsSL $\backslash$
+\item{} brew --version || /bin/bash -c "\$(curl -fsSL $ $
 \item{} \hskip 10mm https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
@@
 \item{}
-\item{} \textMaroon \# install R (if comparisons with Schumaker are desired) \textBlack
-\item{} \textMaroon \# install Schumaker package through R graphical user interface \textBlack
+\item{} \textMaroon \# download and install R (if comparisons with Schumaker are desired) \textBlack
+\item{} echo {\char13}Visit https://cran.r-project.org and follow installation instructions.{\char13}
+\item{} \textMaroon \# download install Schumaker package for R \textBlack
+\item{} curl --output schumaker.tar.gz $ $
+\item{} \hskip 10mm https://cran.r-project.org/src/contrib/schumaker\_1.2.1.tar.gz
+\item{} R CMD INSTALL schumaker.tar.gz
+\item{} \textMaroon \# install a Python package that wraps calls into R \textBlack
 \item{} python3 -m pip install --user rpy2

### 2022-03-30 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex

@@
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
+{\everymath={\scriptstyle}
+
+\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
+  facet model's sensitivity to small data perturbations. This example is
+  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
+  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
+  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
+  curvature \red{(equal second derivatives)}. Given the exact five data points
+  seen above, the quadratic facet model produces the slope seen in the
+  solid blue line at $x = 5/2$. However, by subtracting the value of
+  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
+  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
+  quadratic facet model produces the slope seen in the dashed red line
+  at $x = 5/2$. This is the nature of a facet model and a side effect
+  of associating data with local facets.
+\par} \endinsert
+
 \heading{4. COMPLEXITY AND SENSITIVITY}
@@
 
-\topinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
-{\everymath={\scriptstyle}
-
-\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
-  facet model's sensitivity to small data perturbations. This example is
-  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
-  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
-  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
-  curvature \red{(equal second derivatives)}. Given the exact five data points
-  seen above, the quadratic facet model produces the slope seen in the
-  solid blue line at $x = 5/2$. However, by subtracting the value of
-  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
-  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
-  quadratic facet model produces the slope seen in the dashed red line
-  at $x = 5/2$. This is the nature of a facet model and a side effect
-  of associating data with local facets.
-\par} \endinsert
+\beginred
+The binary search procedure provably converges on the boundary of the
+region of monotonicity precisely defined by Ulrich and Watson [1994]
+through the application of a Boolean function that is guaranteed to be
+true at one end of an interval. Since a local quadratic interpolant is
+applied for initial function derivative estimates, the approximation
+order of the resulting fit is $O(h^3)$ (as for any second order
+approximation), but it should be noted that asymptotic approximation
+order is rarely of importance when considering the scattered sparse
+data that $C^2$ approximations like {\tt MQSI} provide. Were exact
+evaluations provided, higher order methods always win \textRed (all
+other things being equal) as the data density increases. For sparse
+data, what constitutes a better fit is either subjective or dependent
+on the problem.
+\endred

### 2022-03-30 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex

@@
 %% shape parameters.
-\refj{Abbas M., Majid, A. A. and Ali,
+\refj{\beginred Abbas M., Majid, A. A. and Ali,
   J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for
@@
   Computation}{219(6):2885--2895}
+\endred
 
@@
 % berglund2009planning
-\refj{Berglund, T., Brodnik, A., Jonsson, H., Staffanson, M., and
+\refj{Berglund, T., Brodnik, A., Jonsson, H., Staffanson, M. and
   Soderkvist, I.}{2009}{Planning smooth and obstacle-avoiding B-spline
@@
 \refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth,
-  U., Bernard, J., Jearls, C., Lux, T. C. H., Hong, Y., Watson, L. T.,
+  U., Bernard, J., Jearls, C., Lux, T. C. H., Hong, Y., Watson, L. T.
   and Butt, A. R.}{2019}{MOANA: modeling and analyzing I/O variability
@@
 %% convexity while having fixed continuity and order.
-\refj{Costantini, P.}{1986}{On monotone and convex spline
+\refj{\beginred Costantini, P.}{1986}{On monotone and convex spline
   interpolation}{Mathematics of Computation}{46(173):203--214}
+\endred
 
 %% Source paper accompanying the BVSPIS publication.
-\refj{Costantini, P.}{1997}{Boundary-valued shape-preserving
+\refj{\beginred Costantini, P.}{1997a}{Boundary-valued shape-preserving
   interpolating splines}{ACM Transactions on Mathematical Software
   (TOMS)}{23(2):229--251}
+\endred
 
 %% BVSPIS Algorithm.
-\refj{Costantini, P.}{1997}{Algorithm 770: BVSPIS—a package for
+\refj{\beginred Costantini, P.}{1997b}{Algorithm 770: BVSPIS -- A package for
   computing boundary-valued shape-preserving interpolating
-  splines}{ACM Transactions on Mathematical Software (TOMS)}{23(2):252-–254}
+  splines}{ACM Transactions on Mathematical Software
+  (TOMS)}{23(2):252-–254}
+\endred
 
@@
 %% satisfied, no code is provided.
-\refj{Cravero, I., C. Manni}{2003}{Shape-preserving interpolants with
+\refj{\beginred Cravero, I. and C. Manni}{2003}{Shape-preserving interpolants with
   high smoothness}{Journal of computational and applied
   mathematics}{157(2):383--405}
+\endred
 
@@
 %% setting derivatives at the ends, no formal algorithms or code.
-\refj{Delbourgo, R., Gregory, J. A.}{1983}{$C^2$ rational quadratic
+\refj{\beginred Delbourgo, R. and Gregory, J. A.}{1983}{$C^2$ rational quadratic
   spline interpolation to monotonic data}{IMA Journal of Numerical
   Analysis}{3(2):141--152}
+\endred
 
@@
 %% provided.
-\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
+\refj{\beginred Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
   tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}
+\endred
 
@@
 %% of data from monotone functions.
-\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
+\refj{\beginred DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
   Journal on Mathematical Analysis}{8(5):891--905}
-
-%% Accompanies the previous error bounds in the context of monotone
-%% polynomial interpolation of data from a monotone function.
-\refj{DeVore, R. A.}{1977}{Monotone approximation by polynomials}{SIAM
-  Journal on Mathematical Analysis}{8(5):906--921}
+\endred
 
@@
 %% promoting problem specific definitions of 'geometric niceness'.
-\refj{Dougherty, R. L., Edelman, A., Hyman,
+\refj{\beginred Dougherty, R. L., Edelman, A. and Hyman,
   J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving
@@
   Computation}{52(186):471--494}
+\endred
 
@@
 %% always exist for arbitrary convex monotone data sets.
-\refj{Fiorot, J. C., Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
+\refj{\beginred Fiorot, J. C. and Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
   polynomial interpolating splines}{Mathematics of
-  Computation}{57(195):291--298}
+  Computation}{57(195):291--298} \endred \goodbreak
 
@@
 %% interpolation. Does not provide formal algorithm, nor code.
-\refj{Gregory , J. A.,Delbourgo, R.}{1982}{Piecewise rational
+\refj{\beginred Gregory, J. A. and Delbourgo, R.}{1982}{Piecewise rational
   quadratic interpolation to monotonic data}{IMA Journal of Numerical
   Analysis}{2(2):123--130}
+\endred
 
@@
 \reft{Gregory, J. A.}{1985}{Shape preserving spline
-  interpolation}{19850020252, Brunel University}
+  interpolation}{NASA, Langley Research Center Computational Geometry
+  and Computer-Aided Design}
+
+% The paper where a linear facet model is proposed.
+\refj{\beginred Haralick, R. M. and Watson, L. T.}{1981}{A facet model for image
+  data}{Computer Graphics Image Processing}{15:113--129}
+\endred
 
@@
 %% applied, no algorithm is provided, neither is any code provided.
-\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation
+\refj{\beginred Han, X.}{2018}{Shape-preserving piecewise rational interpolation
   with higher order continuity}{Applied Mathematics and
   Computation}{337:1--13}
+\endred
 
 % herman2015techniques
-\reft{Herman, D. L., and Oftedal, M. J.}{2015}{Techniques and
+\reft{Herman, D. L. and Oftedal, M. J.}{2015}{Techniques and
   workflows for computer graphics animation systems}{US Patent
@@
 %% data.
-\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
+\refj{\beginred Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
   Journal of Numerical Analysis}{30(1):57--100}
+\endred
 
@@
 \reft{Lux, T. C. H.}{2020}{Interpolants and error bounds for modeling
-  and predicting variability in computer systems}{Thesis, Virginia
-  Polytechnic Institute and State University, Blacksburg, Virginia}
+  and predicting variability in computer systems}{Ph.D. thesis,
+  Department of Computer Science, Virginia Polytechnic Institute and
+  State University, Blacksburg, Virginia}
+
+% Monotone quintic algorithm.
+\refj{\beginred Lux, T. C. H., Watson, L. T., Chang, T. H., Xu, L.,
+  Wang, Y. and Hong, Y.}{2020}{An algorithm for constructing monotone
+  quintic interpolating splines}{SCS {\sl Spring Simulation
+    Multiconference, High Performance Computing Symposium}}{Article
+  33, 1--12}
+\endred
 
@@
 %% interval. No code (paywall).
-\refj{Manni, C., Sablonni\`ere, P.}{1997}{Monotone interpolation of
+\refj{\beginred Manni, C. and Sablonni\`ere, P.}{1997}{Monotone interpolation of
   order 3 by $C^2$ cubic splines}{IMA Journal of Numerical
   Analysis}{17(2):305--320}
+\endred
 
@@
 %% (paywall).
-\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite
+\refj{\beginred Manni, C.}{2001}{On shape preserving $C^2$ Hermite
   interpolation}{BIT Numerical Mathematics}{41(1):127--148}
+\endred
 
 % mcallister1981algorithm
-\refj{McAllister, D. F. and Roulier, J. A.}{1981}{An algorithm for 
+\refj{McAllister, D. F. and Roulier, J. A.}{1981a}{An algorithm for 
   computing a shape-preserving osculatory quadratic spline}{ACM 
@@
 % mcallister1981shape
-\refj{McAllister, D. F. and Roulier, J. A.}{1981}{Algorithm 574:
+\refj{McAllister, D. F. and Roulier, J. A.}{1981b}{Algorithm 574:
   Shape-Preserving Osculatory Quadratic Splines [E1, E2]}{ACM
@@
   The MathWorks Inc., Natick, Massachusetts}
+
+% Proves that C2 convexity preserving splines do not always exist.
+\reft{\beginred Mulansky, B. and Neamtu, M.}{1991}{On the existence of shape
+  preserving interpolation operators}{Dissertation, University of
+  Twente, Vancouver}
+\endred
 
@@
 %% provided for monotonicity.
-\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
+\refj{\beginred Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
   interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}
+\endred
 
@@
 
+% ulrich1990positivity tech report
+\reft{\beginred Ulrich, G. and Watson, L. T.}{1990}{Positivity conditions for
+  quartic polynomials}{Technical Report 90-57, Department of Computer
+  Science, Virginia Polytechnic Institute and State University,
+  Blacksburg, Virginia}
+\endred
+
 % ulrich1994positivity
@@
   $\scriptstyle C^2$ quartic spline method with nonoscillation
-  derivatives}{Advances in Pure Mathematics}{8(LLNL-JRNL-742107)}
+  derivatives}{Advances in Pure Mathematics}{8(1):25--40}
 
@@
 %% C2 rational quartic shape preserving splines.
-\refj{Zhu, Y.,Han, X.}{2015}{$C^2$ rational quartic interpolation
+\refj{\beginred Zhu, Y. and Han, X.}{2015a}{$C^2$ rational quartic interpolation
   spline with local shape preserving property}{Applied Mathematics
   Letters}{46:57--63}
+\endred
 
@@
 %% rational quartic shape preserving splines.
-\refj{Zhu, Y.,Han, X.}{2015}{Shape preserving $C^2$ rational quartic
+\refj{\beginred Zhu, Y. and Han, X.}{2015b}{Shape preserving $C^2$ rational quartic
   interpolation spline with two parameters}{International Journal of
   Computer Mathematics}{92(10):2160--2177}
-
+\endred
 }

### 2022-08-23 [2020-08]_ACMTOMS_(MQSI)/paper/appendix.tex
Initial response to reviewers.

\vfil\break

\heading{APPENDIX}

The quiet failure of two published codes for shape
preserving spline interpolation is surprising, and accentuates how
difficult it is to correctly handle the numerical intricacies of
polynomial function evaluation. For the sake of reproducibility, coded
minimum working examples of the failure modes for the Schumaker spline
package in {\tt R} and the {\tt BVSPIP} package in Fortran are
provided.

\vskip 5mm

\noindent{\bf Schumaker error in {\tt R}}

\vskip 3mm

{\parskip=-1pt \parindent=10pt \ttVIII
\item{} \textMaroon \# Load the library and set up the problem data. \textBlack
\item{} library(schumaker)
\item{} x = c(0.0, 0.001, 0.02, 0.03)
\item{} y = c(0.0, 0.09, 0.1, 0.11)
\item{} \textMaroon \# Construct the spline fit, rename its evaluator as {\char13}f{\char13}. \textBlack
\item{} SchumakerSpline = schumaker::Schumaker(x, y)
\item{} f = SchumakerSpline\$Spline
\item{} \textMaroon \# Evaluate at 0.01 to get 0.28, greater than all {\char13}y{\char13}. \textBlack
\item{} f(0.01) > max(y)
\item{} \textMaroon \# Plot the results to see the failed approximation. \textBlack
\item{} plot(x, y, col=1, ylim=c(-0.1,0.35), main="Schumaker", ylab="y", xlab="x")
\item{} lines(seq(0,0.03,0.0001), f(seq(0,0.03,0.0001)), col=4)
}

\vskip 5mm

\noindent{\bf BVSPIS error in Fortran}

\vskip 3mm

{\parskip=0pt \parindent=10pt \ttVIII
\item{} \textMaroon ! Compile and execute the following program with: \textBlack
\item{} \textMaroon ! \$F03 bvspis.f \$FILENAME -o \$EXECNAME \textBlack
\item{} \textMaroon ! ./\$EXECNAME \textBlack
\item{} \textMaroon ! Where some placeholder values for the above shell variables might result in: \textBlack
\item{} \textMaroon ! gfortran bvspis.f test\_bvspis\_errors.f03 -o test\_bvspis\_errors \textBlack
\item{} \textMaroon ! ./test\_bvspis\_errors \textBlack
\item{} \textCyan PROGRAM \textBlue TEST\_BVSPIS \textBlack
\item{}   \textCyan IMPLICIT \textGreen NONE\textBlack
\item{}   \textCyan EXTERNAL \textBlack :: \textBlue DBVSSC\textBlack, \textBlue DBVSSE\textBlack
\item{}   \textGreen INTEGER \textBlack :: NP, N, K, R, Q, P, OPT, KMAX, MAXSTP, NWORK, SBOPT, NTAB, \textCyan\&\textBlack $ $
\item{}   \hskip 10mm Y0OPT, Y1OPT, Y2OPT, ERRC, ERRE
\item{}   \textGreen DOUBLE PRECISION \textBlack :: EPS, D0, DNP, D20, D2NP, BETA, BETAI, RHO, RHOI
\item{}   \textGreen INTEGER\textBlack, \textCyan ALLOCATABLE \textBlack :: CONSTR(:), DIAGN(:) 
\item{}   \textGreen DOUBLE PRECISION\textBlack, \textCyan ALLOCATABLE \textBlack :: X(:), Y(:), XTAB(:), D(:), D2(:), \textCyan\&\textBlack $ $
\item{}   \hskip 10mm WORK(:), Y0TAB(:), Y1TAB(:), Y2TAB(:)
\item{}   \textMaroon ! Set parameters. \textBlack
\item{}   EPS = \textCyan SQRT\textBlack(\textCyan EPSILON\textBlack(0.0D0)) \textMaroon ! Default precision. \textBlack
\item{}   NP = 3-1 \textMaroon ! Number of points minus one. \textBlack
\item{}   N = 6 \textMaroon ! Degree of the spline (minimum of 3*k). \textBlack
\item{}   K = 2 \textMaroon ! Class of continuity (C2). \textBlack
\item{}   R = 1 \textMaroon ! Enforce monotonicity constraints. \textBlack
\item{}   Q = 1 \textMaroon ! No boundary conditions are imposed. \textBlack
\item{}   P = 1 \textMaroon ! Derivative estimation by constraints only. \textBlack
\item{}   OPT = 100*P + 10*Q + R \textMaroon ! Contains PQR in decimal format. \textBlack
\item{}   NWORK = 5+(2+7)*NP+(N*(N+11))/2+9 \textMaroon ! Explained in DBVSSC comments. \textBlack
\item{}   SBOPT = 2 \textMaroon ! Do a binary search to find intervals for evaluation. \textBlack
\item{}   NTAB = 3-1 \textMaroon ! Number of evaluation points minus one. \textBlack
\item{}   Y0OPT = 1 \textMaroon ! Enable evaluation of the zeroth derivative. \textBlack
\item{}   Y1OPT = 0 \textMaroon ! Disable evaluation of the first derivative. \textBlack
\item{}   Y2OPT = 0 \textMaroon ! Disable evaluation of the second derivative. \textBlack
\item{}   ERRC = 0 \textMaroon ! Error code, initialize to 0, should be overwritten. \textBlack
\item{}   ERRE = 0 \textMaroon ! Error code, Initialize to 0, should be overwritten. \textBlack
\item{}   \textMaroon ! Allocate and initialize arrays. \textBlack
\item{}   \textCyan ALLOCATE\textBlack( \textCyan\&\textBlack $ $
\item{}   \hskip 10mm X(0:NP), \textCyan\& \textBlack \textMaroon ! X values that cause failure. \textBlack $ $
\item{}   \hskip 10mm Y(0:NP), \textCyan\& \textBlack \textMaroon ! Y values that cause failure. \textBlack $ $
\item{}   \hskip 10mm CONSTR(0:NP), \textCyan\& \textBlack \textMaroon ! Used for setting shape preserving constraints. \textBlack $ $
\item{}   \hskip 10mm XTAB(0:NTAB), \textCyan\& \textBlack \textMaroon ! Z evaluation point (known failure location). \textBlack $ $
\item{}   \hskip 10mm D(0:NP), \textCyan\& \textBlack \textMaroon ! Derivative values (on output). \textBlack $ $
\item{}   \hskip 10mm D2(0:NP), \textCyan\& \textBlack \textMaroon ! Referenced with K=2. \textBlack $ $
\item{}   \hskip 10mm DIAGN(0:NP-1), \textCyan\& \textBlack \textMaroon ! Diagnostic info. \textBlack $ $
\item{}   \hskip 10mm WORK(1:NWORK), \textCyan\& \textBlack \textMaroon ! Work space. \textBlack $ $
\item{}   \hskip 10mm Y0TAB(0:NTAB), \textCyan\& \textBlack \textMaroon ! Spline values. \textBlack $ $
\item{}   \hskip 10mm Y1TAB(0:NTAB), \textCyan\& \textBlack \textMaroon ! First derivative values. \textBlack $ $
\item{}   \hskip 10mm Y2TAB(0:NTAB) \textCyan\& \textBlack \textMaroon ! Second derivative values. \textBlack $ $
\item{}   )
\item{}   X(0:NP) = (/ 0.0D0, 0.09D0, 0.091D0 /)
\item{}   Y(0:NP) = (/ 0.0D0, 0.1D0, 0.2D0 /)
\item{}   XTAB(0:NTAB) = (/ 0.0D0, 0.06D0, 0.09D0 /)
\item{}   CONSTR(0:NP) = 1
\item{}   D(0:NP) = 0.0D0
\item{}   D2(0:NP) = 0.0D0
\item{}   DIAGN(0:NP-1) = 0.0D0
\item{}   WORK(1:NWORK) = 0.0D0
\item{}   Y0TAB(0:NTAB) = 0.0D0
\item{}   Y1TAB(0:NTAB) = 0.0D0
\item{}   Y2TAB(0:NTAB) = 0.0D0
\item{}   \textMaroon ! Unused parameters, ignored unless Q=3. \textBlack
\item{}   D0 = 0.0D0
\item{}   DNP = 0.0D0
\item{}   D20 = 0.0D0
\item{}   D2NP = 0.0D0
\item{}   \textMaroon ! Unused parameters, ignored unless Q=2. \textBlack
\item{}   BETA = 0.0D0
\item{}   BETAI = 0.0D0
\item{}   RHO = 0.0D0
\item{}   RHOI = 0.0D0
\item{}   KMAX = 0
\item{}   MAXSTP = 0
\item{}   \textMaroon ! Construct the fit. \textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Constructing fit.{\char13}\textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}X(:){\char13}\textBlack, X(:)
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Y(:){\char13}\textBlack, Y(:)
\item{}   \textCyan CALL \textBlue DBVSSC\textBlack( \textCyan\&\textBlack $ $
\item{}   \hskip 10mm X(:), Y(:), NP, N, K, OPT, D0, DNP, D20, D2NP, CONSTR(:), \textCyan\&\textBlack $ $
\item{}   \hskip 10mm EPS, BETA, BETAI, RHO, RHOI, KMAX, MAXSTP, ERRC, \textCyan\&\textBlack $ $
\item{}   \hskip 10mm D(:), D2(:), DIAGN(:), WORK(:), NWORK \textCyan\&\textBlack $ $
\item{}   )
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}ERRC {\char13}\textBlack, ERRC \textMaroon ! = 0 \textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}{\char13}\textBlack
\item{}   \textMaroon ! Evaluate the fit. \textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Evaluate the spline.{\char13}\textBlack
\item{}   \textCyan CALL \textBlue DBVSSE\textBlack( \textCyan\&\textBlack $ $
\item{}   \hskip 10mm X(:), Y(:), NP, N, K, XTAB(:), NTAB, SBOPT, Y0OPT, Y1OPT, \textCyan\&\textBlack $ $
\item{}   \hskip 10mm Y2OPT, ERRC, D(:), D2(:), Y0TAB(:), Y1TAB(:), Y2TAB(:), \textCyan\&\textBlack $ $
\item{}   \hskip 10mm ERRE, WORK(:), NWORK \textCyan\&\textBlack $ $
\item{}   )
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}XTAB(:) {\char13}\textBlack, XTAB(:) \textMaroon ! = 0.0, 0.06, 0.09 as expected.\textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Y0TAB(:){\char13}\textBlack, Y0TAB(:) \textMaroon ! = 0.0, 0.152, 0.1 not monotone, unexpected.\textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}ERRE {\char13}\textBlack, ERRE \textMaroon ! = 0 no error code, unexpected.\textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}ERRC {\char13}\textBlack, ERRC \textMaroon ! = 0 no error code, unexpected.\textBlack
\item{} \textCyan END PROGRAM \textBlue TEST\_BVSPIS \textBlack

}

### 2022-08-23 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Second round of internal revision.

--- 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
@@
 picking a local linear or quadratic fit with minimal residual, this
-work uses a quadratic facet model that selects the local quadratic
-interpolant with minimum magnitude second derivative.
+work \red{instead} % uses a quadratic facet model that
+selects the local quadratic
+interpolant with minimum magnitude second derivative \red{(hereafter
+referred to as a {\it quadratic facet model}). The term ``facet'' 
+refers to one piece of an image surface modelled by a two-dimensional
+piecewise polynomial.}.
 
@@
 $X_i$ of the local quadratic interpolant with minimum magnitude
-second derivative. Approximate equality is denoted with $\approx$ and
-states that two numbers are considered equal once they are within the
-machine precision $\epsilon$ of each other.
+second derivative. Approximate equality is denoted with
+$\approx$ and considers two numbers to be equal once they are within
+the machine precision $\epsilon$ of each other.
 
@@
 increasing on $[x_0,x_1]$. Approximate equality is denoted with
-$\approx$ and states that two numbers are considered equal once they
-are within the machine precision $\epsilon$ of each other.
+$\approx$ and considers two numbers to be equal once they are within
+the machine precision $\epsilon$ of each other.
 
@@
 
-Given the minimum curvature (minimum magnitude of the second derivative)
-nature of the initial derivative estimates, it is desirable to make
+Given \red{that} the \red{initial derivative estimates locally minimize 
+the magnitude of the second derivative}, it is desirable to make
 the smallest necessary changes to the initial interpolating spline $Q$

### 2022-08-23 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Second round of internal revision.

@@
 {\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
-monotone quintic spline interpolants to \red{univariate} monotone data. Using sharp
+monotone quintic spline interpolants to univariate monotone data. Using sharp
 theoretical monotonicity constraints, first and second derivative estimates
 at data provided by a quadratic facet model are refined to produce a
-\red{univariate} $\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
+univariate $\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
 details, complexity and sensitivity analyses, usage information, a
-brief performance study, \red{and comparisons with other spline approaches} are included.
+brief performance study, and comparisons with other spline approaches are included.
 
@@
 %% shape parameters.
-\refj{\beginred Abbas M., Majid, A. A. and Ali,
+\refj{Abbas M., Majid, A. A. and Ali,
   J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for
@@
   Computation}{219(6):2885--2895}
-\endred
 
@@
 %% convexity while having fixed continuity and order.
-\refj{\beginred Costantini, P.}{1986}{On monotone and convex spline
+\refj{Costantini, P.}{1986}{On monotone and convex spline
   interpolation}{Mathematics of Computation}{46(173):203--214}
-\endred
 
 %% Source paper accompanying the BVSPIS publication.
-\refj{\beginred Costantini, P.}{1997a}{Boundary-valued shape-preserving
+\refj{Costantini, P.}{1997a}{Boundary-valued shape-preserving
   interpolating splines}{ACM Transactions on Mathematical Software
   (TOMS)}{23(2):229--251}
-\endred
 
 %% BVSPIS Algorithm.
-\refj{\beginred Costantini, P.}{1997b}{Algorithm 770: BVSPIS -- A package for
+\refj{Costantini, P.}{1997b}{Algorithm 770: BVSPIS -- A package for
   computing boundary-valued shape-preserving interpolating
@@
   (TOMS)}{23(2):252-–254}
-\endred
 
@@
 %% satisfied, no code is provided.
-\refj{\beginred Cravero, I. and C. Manni}{2003}{Shape-preserving interpolants with
+\refj{Cravero, I. and C. Manni}{2003}{Shape-preserving interpolants with
   high smoothness}{Journal of computational and applied
   mathematics}{157(2):383--405}
-\endred
 
@@
 %% setting derivatives at the ends, no formal algorithms or code.
-\refj{\beginred Delbourgo, R. and Gregory, J. A.}{1983}{$C^2$ rational quadratic
+\refj{Delbourgo, R. and Gregory, J. A.}{1983}{$C^2$ rational quadratic
   spline interpolation to monotonic data}{IMA Journal of Numerical
   Analysis}{3(2):141--152}
-\endred
 
@@
 %% provided.
-\refj{\beginred Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
+\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
   tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}
-\endred
 
@@
 %% of data from monotone functions.
-\refj{\beginred DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
+\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
   Journal on Mathematical Analysis}{8(5):891--905}
-\endred
 
@@
 %% promoting problem specific definitions of 'geometric niceness'.
-\refj{\beginred Dougherty, R. L., Edelman, A. and Hyman,
+\refj{Dougherty, R. L., Edelman, A. and Hyman,
   J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving
@@
   Computation}{52(186):471--494}
-\endred
 
@@
 %% always exist for arbitrary convex monotone data sets.
-\refj{\beginred Fiorot, J. C. and Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
+\refj{Fiorot, J. C. and Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
   polynomial interpolating splines}{Mathematics of
-  Computation}{57(195):291--298} \endred \goodbreak
+  Computation}{57(195):291--298} \goodbreak
 
@@
 %% interpolation. Does not provide formal algorithm, nor code.
-\refj{\beginred Gregory, J. A. and Delbourgo, R.}{1982}{Piecewise rational
+\refj{Gregory, J. A. and Delbourgo, R.}{1982}{Piecewise rational
   quadratic interpolation to monotonic data}{IMA Journal of Numerical
   Analysis}{2(2):123--130}
-\endred
 
@@
 % The paper where a linear facet model is proposed.
-\refj{\beginred Haralick, R. M. and Watson, L. T.}{1981}{A facet model for image
+\refj{Haralick, R. M. and Watson, L. T.}{1981}{A facet model for image
   data}{Computer Graphics Image Processing}{15:113--129}
-\endred
 
@@
 %% applied, no algorithm is provided, neither is any code provided.
-\refj{\beginred Han, X.}{2018}{Shape-preserving piecewise rational interpolation
+\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation
   with higher order continuity}{Applied Mathematics and
   Computation}{337:1--13}
-\endred
 
@@
 %% data.
-\refj{\beginred Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
+\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
   Journal of Numerical Analysis}{30(1):57--100}
-\endred
 
@@
 % Monotone quintic algorithm.
-\refj{\beginred Lux, T. C. H., Watson, L. T., Chang, T. H., Xu, L.,
+\refj{Lux, T. C. H., Watson, L. T., Chang, T. H., Xu, L.,
   Wang, Y. and Hong, Y.}{2020}{An algorithm for constructing monotone
@@
   33, 1--12}
-\endred
 
@@
 %% interval. No code (paywall).
-\refj{\beginred Manni, C. and Sablonni\`ere, P.}{1997}{Monotone interpolation of
+\refj{Manni, C. and Sablonni\`ere, P.}{1997}{Monotone interpolation of
   order 3 by $C^2$ cubic splines}{IMA Journal of Numerical
   Analysis}{17(2):305--320}
-\endred
 
@@
 %% (paywall).
-\refj{\beginred Manni, C.}{2001}{On shape preserving $C^2$ Hermite
+\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite
   interpolation}{BIT Numerical Mathematics}{41(1):127--148}
-\endred
 
@@
 % Proves that C2 convexity preserving splines do not always exist.
-\reft{\beginred Mulansky, B. and Neamtu, M.}{1991}{On the existence of shape
+\reft{Mulansky, B. and Neamtu, M.}{1991}{On the existence of shape
   preserving interpolation operators}{Dissertation, University of
   Twente, Vancouver}
-\endred
 
@@
 %% provided for monotonicity.
-\refj{\beginred Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
+\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
   interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}
-\endred
 
@@
 % ulrich1990positivity tech report
-\reft{\beginred Ulrich, G. and Watson, L. T.}{1990}{Positivity conditions for
+\reft{Ulrich, G. and Watson, L. T.}{1990}{Positivity conditions for
   quartic polynomials}{Technical Report 90-57, Department of Computer
@@
   Blacksburg, Virginia}
-\endred
 
@@
 %% C2 rational quartic shape preserving splines.
-\refj{\beginred Zhu, Y. and Han, X.}{2015a}{$C^2$ rational quartic interpolation
+\refj{Zhu, Y. and Han, X.}{2015a}{$C^2$ rational quartic interpolation
   spline with local shape preserving property}{Applied Mathematics
   Letters}{46:57--63}
-\endred
 
@@
 %% rational quartic shape preserving splines.
-\refj{\beginred Zhu, Y. and Han, X.}{2015b}{Shape preserving $C^2$ rational quartic
+\refj{Zhu, Y. and Han, X.}{2015b}{Shape preserving $C^2$ rational quartic
   interpolation spline with two parameters}{International Journal of
   Computer Mathematics}{92(10):2160--2177}
-\endred
 }
@@
 
+\input appendix
+
 \vfill\eject\end

### 2022-08-25 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex

--- 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
@@
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
+{\everymath={\scriptstyle}
+
+\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
+  facet model's sensitivity to small data perturbations. This example is
+  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
+  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
+  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have
+  \red{equal second derivatives}. Given the exact five data points
+  seen above, the quadratic facet model produces the slope seen in the
+  solid blue line at $x = 5/2$. However, by subtracting the value of
+  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
+  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
+  quadratic facet model produces the slope seen in the dashed red line
+  at $x = 5/2$. This is the nature of a facet model and a side effect
+  of associating data with local facets.
+\par} \endinsert
+
 \heading{4. COMPLEXITY AND SENSITIVITY}
@@
 
-\topinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
-{\everymath={\scriptstyle}
-
-\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
-  facet model's sensitivity to small data perturbations. This example is
-  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
-  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
-  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
-  curvature (equal second derivatives). Given the exact five data points
-  seen above, the quadratic facet model produces the slope seen in the
-  solid blue line at $x = 5/2$. However, by subtracting the value of
-  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
-  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
-  quadratic facet model produces the slope seen in the dashed red line
-  at $x = 5/2$. This is the nature of a facet model and a side effect
-  of associating data with local facets.
-\par} \endinsert
+The binary search procedure provably converges on the boundary of the
+region of monotonicity precisely defined by Ulrich and Watson [1994]
+through the application of a Boolean function that is guaranteed to be
+true at one end of an interval. Since a local quadratic interpolant is
+applied for initial function derivative estimates, the approximation
+order of the resulting fit is $O(h^3)$ (as for any second order
+approximation), but it should be noted that asymptotic approximation
+order is rarely of importance when considering the scattered sparse
+data that $C^2$ approximations like {\tt MQSI} provide. Were exact
+evaluations provided, higher order methods always win (all
+other things being equal) as the data density increases. For sparse
+data, what constitutes a better fit is either subjective or dependent
+on the problem.

### 2022-08-25 [2020-08]_ACMTOMS_(MQSI)/paper/appendix.tex

@@
-\vfil\break
+\beginred
 
@@
 
-The quiet failure of two published codes for shape
-preserving spline interpolation is surprising, and accentuates how
-difficult it is to correctly handle the numerical intricacies of
-polynomial function evaluation. For the sake of reproducibility, coded
-minimum working examples of the failure modes for the Schumaker spline
-package in {\tt R} and the {\tt BVSPIP} package in Fortran are
-provided.
+The quiet failure of two published codes for shape preserving spline
+interpolation is surprising, and accentuates how difficult it is to
+correctly handle the numerical intricacies of polynomial function
+evaluation. For the sake of reproducibility, coded minimum working
+examples of the failure modes for the Schumaker spline package in {\tt
+  R} and the {\tt BVSPIP} package in Fortran are provided as
+supplemental files. The program {\tt schumaker\_test.r} demonstrates a
+sequence of four monotone increasing points that yields a nonmonotone
+approximation. The program {\tt bvspis\_test.f03} provides a sequence
+of three monotone increasing points for which {\tt BVSPIS} produces a
+nonmonotone approximation.
 
-\vskip 5mm
-
-\noindent{\bf Schumaker error in {\tt R}}
-
-\vskip 3mm
-
-{\parskip=-1pt \parindent=10pt \ttVIII
-\item{} \textMaroon \# Load the library and set up the problem data. \textBlack
-\item{} library(schumaker)
-\item{} x = c(0.0, 0.001, 0.02, 0.03)
-\item{} y = c(0.0, 0.09, 0.1, 0.11)
-\item{} \textMaroon \# Construct the spline fit, rename its evaluator as {\char13}f{\char13}. \textBlack
-\item{} SchumakerSpline = schumaker::Schumaker(x, y)
-\item{} f = SchumakerSpline\$Spline
-\item{} \textMaroon \# Evaluate at 0.01 to get 0.28, greater than all {\char13}y{\char13}. \textBlack
-\item{} f(0.01) > max(y)
-\item{} \textMaroon \# Plot the results to see the failed approximation. \textBlack
-\item{} plot(x, y, col=1, ylim=c(-0.1,0.35), main="Schumaker", ylab="y", xlab="x")
-\item{} lines(seq(0,0.03,0.0001), f(seq(0,0.03,0.0001)), col=4)
-}
-
-\vskip 5mm
-
-\noindent{\bf BVSPIS error in Fortran}
-
-\vskip 3mm
-
-{\parskip=0pt \parindent=10pt \ttVIII
-\item{} \textMaroon ! Compile and execute the following program with: \textBlack
-\item{} \textMaroon ! \$F03 bvspis.f \$FILENAME -o \$EXECNAME \textBlack
-\item{} \textMaroon ! ./\$EXECNAME \textBlack
-\item{} \textMaroon ! Where some placeholder values for the above shell variables might result in: \textBlack
-\item{} \textMaroon ! gfortran bvspis.f test\_bvspis\_errors.f03 -o test\_bvspis\_errors \textBlack
-\item{} \textMaroon ! ./test\_bvspis\_errors \textBlack
-\item{} \textCyan PROGRAM \textBlue TEST\_BVSPIS \textBlack
-\item{}   \textCyan IMPLICIT \textGreen NONE\textBlack
-\item{}   \textCyan EXTERNAL \textBlack :: \textBlue DBVSSC\textBlack, \textBlue DBVSSE\textBlack
-\item{}   \textGreen INTEGER \textBlack :: NP, N, K, R, Q, P, OPT, KMAX, MAXSTP, NWORK, SBOPT, NTAB, \textCyan\&\textBlack $ $
-\item{}   \hskip 10mm Y0OPT, Y1OPT, Y2OPT, ERRC, ERRE
-\item{}   \textGreen DOUBLE PRECISION \textBlack :: EPS, D0, DNP, D20, D2NP, BETA, BETAI, RHO, RHOI
-\item{}   \textGreen INTEGER\textBlack, \textCyan ALLOCATABLE \textBlack :: CONSTR(:), DIAGN(:) 
-\item{}   \textGreen DOUBLE PRECISION\textBlack, \textCyan ALLOCATABLE \textBlack :: X(:), Y(:), XTAB(:), D(:), D2(:), \textCyan\&\textBlack $ $
-\item{}   \hskip 10mm WORK(:), Y0TAB(:), Y1TAB(:), Y2TAB(:)
-\item{}   \textMaroon ! Set parameters. \textBlack
-\item{}   EPS = \textCyan SQRT\textBlack(\textCyan EPSILON\textBlack(0.0D0)) \textMaroon ! Default precision. \textBlack
-\item{}   NP = 3-1 \textMaroon ! Number of points minus one. \textBlack
-\item{}   N = 6 \textMaroon ! Degree of the spline (minimum of 3*k). \textBlack
-\item{}   K = 2 \textMaroon ! Class of continuity (C2). \textBlack
-\item{}   R = 1 \textMaroon ! Enforce monotonicity constraints. \textBlack
-\item{}   Q = 1 \textMaroon ! No boundary conditions are imposed. \textBlack
-\item{}   P = 1 \textMaroon ! Derivative estimation by constraints only. \textBlack
-\item{}   OPT = 100*P + 10*Q + R \textMaroon ! Contains PQR in decimal format. \textBlack
-\item{}   NWORK = 5+(2+7)*NP+(N*(N+11))/2+9 \textMaroon ! Explained in DBVSSC comments. \textBlack
-\item{}   SBOPT = 2 \textMaroon ! Do a binary search to find intervals for evaluation. \textBlack
-\item{}   NTAB = 3-1 \textMaroon ! Number of evaluation points minus one. \textBlack
-\item{}   Y0OPT = 1 \textMaroon ! Enable evaluation of the zeroth derivative. \textBlack
-\item{}   Y1OPT = 0 \textMaroon ! Disable evaluation of the first derivative. \textBlack
-\item{}   Y2OPT = 0 \textMaroon ! Disable evaluation of the second derivative. \textBlack
-\item{}   ERRC = 0 \textMaroon ! Error code, initialize to 0, should be overwritten. \textBlack
-\item{}   ERRE = 0 \textMaroon ! Error code, Initialize to 0, should be overwritten. \textBlack
-\item{}   \textMaroon ! Allocate and initialize arrays. \textBlack
-\item{}   \textCyan ALLOCATE\textBlack( \textCyan\&\textBlack $ $
-\item{}   \hskip 10mm X(0:NP), \textCyan\& \textBlack \textMaroon ! X values that cause failure. \textBlack $ $
-\item{}   \hskip 10mm Y(0:NP), \textCyan\& \textBlack \textMaroon ! Y values that cause failure. \textBlack $ $
-\item{}   \hskip 10mm CONSTR(0:NP), \textCyan\& \textBlack \textMaroon ! Used for setting shape preserving constraints. \textBlack $ $
-\item{}   \hskip 10mm XTAB(0:NTAB), \textCyan\& \textBlack \textMaroon ! Z evaluation point (known failure location). \textBlack $ $
-\item{}   \hskip 10mm D(0:NP), \textCyan\& \textBlack \textMaroon ! Derivative values (on output). \textBlack $ $
-\item{}   \hskip 10mm D2(0:NP), \textCyan\& \textBlack \textMaroon ! Referenced with K=2. \textBlack $ $
-\item{}   \hskip 10mm DIAGN(0:NP-1), \textCyan\& \textBlack \textMaroon ! Diagnostic info. \textBlack $ $
-\item{}   \hskip 10mm WORK(1:NWORK), \textCyan\& \textBlack \textMaroon ! Work space. \textBlack $ $
-\item{}   \hskip 10mm Y0TAB(0:NTAB), \textCyan\& \textBlack \textMaroon ! Spline values. \textBlack $ $
-\item{}   \hskip 10mm Y1TAB(0:NTAB), \textCyan\& \textBlack \textMaroon ! First derivative values. \textBlack $ $
-\item{}   \hskip 10mm Y2TAB(0:NTAB) \textCyan\& \textBlack \textMaroon ! Second derivative values. \textBlack $ $
-\item{}   )
-\item{}   X(0:NP) = (/ 0.0D0, 0.09D0, 0.091D0 /)
-\item{}   Y(0:NP) = (/ 0.0D0, 0.1D0, 0.2D0 /)
-\item{}   XTAB(0:NTAB) = (/ 0.0D0, 0.06D0, 0.09D0 /)
-\item{}   CONSTR(0:NP) = 1
-\item{}   D(0:NP) = 0.0D0
-\item{}   D2(0:NP) = 0.0D0
-\item{}   DIAGN(0:NP-1) = 0.0D0
-\item{}   WORK(1:NWORK) = 0.0D0
-\item{}   Y0TAB(0:NTAB) = 0.0D0
-\item{}   Y1TAB(0:NTAB) = 0.0D0
-\item{}   Y2TAB(0:NTAB) = 0.0D0
-\item{}   \textMaroon ! Unused parameters, ignored unless Q=3. \textBlack
-\item{}   D0 = 0.0D0
-\item{}   DNP = 0.0D0
-\item{}   D20 = 0.0D0
-\item{}   D2NP = 0.0D0
-\item{}   \textMaroon ! Unused parameters, ignored unless Q=2. \textBlack
-\item{}   BETA = 0.0D0
-\item{}   BETAI = 0.0D0
-\item{}   RHO = 0.0D0
-\item{}   RHOI = 0.0D0
-\item{}   KMAX = 0
-\item{}   MAXSTP = 0
-\item{}   \textMaroon ! Construct the fit. \textBlack
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Constructing fit.{\char13}\textBlack
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}X(:){\char13}\textBlack, X(:)
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Y(:){\char13}\textBlack, Y(:)
-\item{}   \textCyan CALL \textBlue DBVSSC\textBlack( \textCyan\&\textBlack $ $
-\item{}   \hskip 10mm X(:), Y(:), NP, N, K, OPT, D0, DNP, D20, D2NP, CONSTR(:), \textCyan\&\textBlack $ $
-\item{}   \hskip 10mm EPS, BETA, BETAI, RHO, RHOI, KMAX, MAXSTP, ERRC, \textCyan\&\textBlack $ $
-\item{}   \hskip 10mm D(:), D2(:), DIAGN(:), WORK(:), NWORK \textCyan\&\textBlack $ $
-\item{}   )
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}ERRC {\char13}\textBlack, ERRC \textMaroon ! = 0 \textBlack
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}{\char13}\textBlack
-\item{}   \textMaroon ! Evaluate the fit. \textBlack
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Evaluate the spline.{\char13}\textBlack
-\item{}   \textCyan CALL \textBlue DBVSSE\textBlack( \textCyan\&\textBlack $ $
-\item{}   \hskip 10mm X(:), Y(:), NP, N, K, XTAB(:), NTAB, SBOPT, Y0OPT, Y1OPT, \textCyan\&\textBlack $ $
-\item{}   \hskip 10mm Y2OPT, ERRC, D(:), D2(:), Y0TAB(:), Y1TAB(:), Y2TAB(:), \textCyan\&\textBlack $ $
-\item{}   \hskip 10mm ERRE, WORK(:), NWORK \textCyan\&\textBlack $ $
-\item{}   )
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}XTAB(:) {\char13}\textBlack, XTAB(:) \textMaroon ! = 0.0, 0.06, 0.09 as expected.\textBlack
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Y0TAB(:){\char13}\textBlack, Y0TAB(:) \textMaroon ! = 0.0, 0.152, 0.1 not monotone, unexpected.\textBlack
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}ERRE {\char13}\textBlack, ERRE \textMaroon ! = 0 no error code, unexpected.\textBlack
-\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}ERRC {\char13}\textBlack, ERRC \textMaroon ! = 0 no error code, unexpected.\textBlack
-\item{} \textCyan END PROGRAM \textBlue TEST\_BVSPIS \textBlack
-
-}
+\endred

### 2022-08-25 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex

@@
 
-\beginred
-
 Lastly, comparisons between {\tt MQSI} and four published spline
@@
 770. Figures 5 and 8 demonstrate glaring numerical errors in the
-Schumaker and {\tt BVSPIS} implementations, and overall it can be
-observed that the minimal and tight conditions on monotonicity
-provided by TOMS 574 tend to be more visually appealing than the
-weaker sufficient conditions utilized by {\tt PCHIP}. In general the
-major determining factor for the quality of {\tt MQSI} is the local
-quadratic facet model. For applications where the underlying function
-is presumed to be a piecewise polynomial (of relatively low order),
-the choice of local quadratic interpolants is reasonable. However the
-consequence of this decision is that functions with superquadratic
-rates of change will tend to have consistently underestimated first
-and second derivatives. This limitation is accepted in favor of
-perfectly capturing lower order phenomena. Were future research to
-supply alternative initializations for $C^2$ quintic splines, those
-could comfortably be made monotone by the procedures of {\tt MQSI} and
-most importantly by the application of these sharp monotonicity
-conditions.
-
-\endred
+Schumaker and {\tt BVSPIS} implementations, \textRed minimum working
+examples that reproduce similar errors are discussed in the Appendix.
+\textBlack Overall it can be observed that the minimal and tight
+conditions on monotonicity provided by TOMS 574 tend to be more
+visually appealing than the weaker sufficient conditions utilized by
+{\tt PCHIP}. In general the major determining factor for the quality
+of {\tt MQSI} is the local quadratic facet model. For applications
+where the underlying function is presumed to be a piecewise polynomial
+(of relatively low order), the choice of local quadratic interpolants
+is reasonable. However the consequence of this decision is that
+functions with superquadratic rates of change will tend to have
+consistently underestimated first and second derivatives. This
+limitation is accepted in favor of perfectly capturing lower order
+phenomena. Were future research to supply alternative initializations
+for $C^2$ quintic splines, those could comfortably be made monotone by
+the procedures of {\tt MQSI} and most importantly by the application
+of these sharp monotonicity conditions.
 
 If an application demands $C^2$ continuity and monotonicity, the
-target of {\tt MQSI}, then this package \red{uniquely} provides such
-interpolants \red{based on sharp monotonicity constraints.}
+target of {\tt MQSI}, then this package uniquely provides such
+interpolants based on sharp monotonicity constraints.
 
 \goodbreak
-
-\beginred
 
@@
 \centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/4-bvspis.eps}}
-\beginred
 {\narrower\noindent\rmVIII Fig.\ 5.
@@
 \par}
-\endred
 \endinsert
@@
 \centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/4-bvspis.eps}}
-\beginred
 {\narrower\noindent\rmVIII Fig.\ 6.
@@
 function. {\ttVIII MQSI} is styled as a gray thin line in the
-background for comparison. This problems highlights the core weakness
+background for comparison. This problem highlights the core weakness
 of the quadratic facet model approach to derivative estimation in
 {\ttVIII MQSI}, which consistently underestimates the actual
-\red{second derivative} of this function. This weakness is accepted
+second derivative of this function. This weakness is accepted
 for its greater accuracy when estimating local lower order components
@@
 \centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/4-bvspis.eps}}
-\beginred
 {\narrower\noindent\rmVIII Fig.\ 7.
@@
 \centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/4-bvspis.eps}}
-\beginred
 {\narrower\noindent\rmVIII Fig.\ 8.
@@
 \par}
-\endred
 \endinsert
-
-\endred

### 2022-09-27 [2020-08]_ACMTOMS_(MQSI)/code/test_all.f90
Responding to final requests made for changes to the content of the paper.

@@ -235,6 +235,11 @@ FUNCTION PASSES_TEST(F_NUM, N, EQUALLY_SPACED_X, TRIALS, EVAL_TIME, FIT_TIME)
   INTEGER :: I, INFO, J
   !  Storage for seeding the random number generator (for repeatability).
   INTEGER, DIMENSION(:), ALLOCATABLE :: SEED
+  ! Initialize random seed.
+  CALL RANDOM_SEED(SIZE=J)
+  ALLOCATE(SEED(J))
+  SEED(:) = 7919
+  CALL RANDOM_SEED(PUT=SEED)
   ! Initialize X values.
   IF (EQUALLY_SPACED_X) THEN
      DO I = 1, N
@@ -242,11 +247,6 @@ FUNCTION PASSES_TEST(F_NUM, N, EQUALLY_SPACED_X, TRIALS, EVAL_TIME, FIT_TIME)
      END DO
      X(:) = X(:) / REAL(N-1, KIND=R8)
   ELSE
-     ! Initialize random seed.
-     CALL RANDOM_SEED(SIZE=J)
-     ALLOCATE(SEED(J))
-     SEED(:) = 7919
-     CALL RANDOM_SEED(PUT=SEED)
      ! Generate random (increasing) data.
      CALL RANDOM_NUMBER(X)
      CALL SORT(X)
@@ -338,11 +338,11 @@ INTEGER, INTENT(IN) :: N_TEST
 ! Local variables.
 !  Maximum absolute observed error.
 REAL(KIND=R8) :: MAX_ERROR, SCALE
-!  Spline coefficients SC, spline knots SK, temporary
-!  input/output storage U, and test point storage Z.
-REAL(KIND=R8) :: SC(1:3*SIZE(X)), SK(1:3*SIZE(X)+6), U(1:SIZE(X)), Z(1:N_TEST)
+!  Random number R, spline coefficients SC, spline knots SK,
+!  temporary input/output storage U, and test point storage Z.
+REAL(KIND=R8) :: R, SC(1:3*SIZE(X)), SK(1:3*SIZE(X)+6), U(1:SIZE(X)), Z(1:N_TEST)
 !  Iteration variables I, J, and subroutine status integer INFO.
-INTEGER :: I, INFO, J
+INTEGER :: I, INFO, J, K, ORDER(1:SIZE(X))
[…]

### 2022-09-27 [2020-08]_ACMTOMS_(MQSI)/paper/appendix.tex
Responding to final requests made for changes to the content of the paper.

@@
 examples of the failure modes for the Schumaker spline package in {\tt
-  R} and the {\tt BVSPIP} package in Fortran are provided as
+  R} and the {\tt BVSPIS} package in Fortran are provided as
 supplemental files. The program {\tt schumaker\_test.r} demonstrates a

### 2022-09-27 [2020-08]_ACMTOMS_(MQSI)/revisions/MQSI_testing/mqsi/test_spline.py

@@ -1,0 +1,91 @@
+import fmodpy
+
+SPLINE = fmodpy.fimport(
+    "SPLINE.f90",
+    dependencies=["REAL_PRECISION.f90", "EVAL_BSPLINE.f90"],
+    blas=True,
+    lapack=True,
+    # verbose=True,
+    rebuild=True,
+)
+
+# Fit a spline.
+def fit(x, y):
+    import numpy as np
+    x = np.asarray(x, dtype="float64", order="F").flatten()
+    y = np.asarray(y, dtype="float64", order="F").reshape(x.size, -1)
+    assert x.shape[0] == y.shape[0], f"Number of points in x {x.shape} and y {y.shape} do not match."
+    nb = x.shape[0]
+    ncc = y.shape[1]
+    t = np.zeros(nb*ncc+2*ncc, dtype="float64", order="F")
+    bcoef = np.zeros(nb*ncc, dtype="float64", order="F")
+    t, bcoef, info = SPLINE.fit_spline(xi=x, fx=y, t=t, bcoef=bcoef)
+    assert (info == 0), f"FIT_SPLINE returned nonzero exit code {info}."
+    # Given a spline "f", evaluate it.
+    def evaluate(x, d=None, t=t, bcoef=bcoef):
+        import numpy as np
+        x = np.asarray(x, dtype="float64", order="F").flatten()
+        y = x.copy()
+        y, info = SPLINE.eval_spline(t=t, bcoef=bcoef, xy=y, d=d)
+        assert (info == 0), f"EVAL_SPLINE returned nonzero exit code {info}."
+        return y
+    # Return the evaluator.
+    return evaluate
+
+
+import numpy as np
+from tlux.plot import Plot, multiplot
+
+# Seed for repeatability.
+np.random.seed(0)
+
+# Set the number of points.
+# n = 4 
+n = 10
+
+# Set the number of continuity conditions.
+cc = 2
+
+# Set the number of evaluation points per interval.
+ne = 100
+randomized = True
+
+# Random X (sorted)
+x = np.random.random(size=(n,))
+x *= 10
+x.sort()
+
+# Random Y.
+y = np.random.random(size=(n,cc))*2-1
+
+# Fit the data.
+f = fit(x, y)
+
+# Evaluate the function at data points.
+fx = np.concatenate([np.linspace(x[i],x[i+1],ne) for i in range(n-1)])
+
+# Randomize the order of the points.
[…]

### 2022-10-26 [2020-08]_ACMTOMS_(MQSI)/paper/TeX/2-monotone-quintic.tex
Moved all paper files into a TeX directory. Creating LaTeX version next.

--- 2022-08-23 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
@@
 picking a local linear or quadratic fit with minimal residual, this
-work \red{instead} % uses a quadratic facet model that
-selects the local quadratic
-interpolant with minimum magnitude second derivative \red{(hereafter
-referred to as a {\it quadratic facet model}). The term ``facet'' 
-refers to one piece of an image surface modelled by a two-dimensional
-piecewise polynomial.}.
+work instead selects the local quadratic interpolant with minimum
+magnitude second derivative (hereafter referred to as a {\it quadratic
+  facet model}). The term ``facet'' refers to one piece of an image
+surface modelled by a two-dimensional piecewise polynomial.
 
@@
 
-Given \red{that} the \red{initial derivative estimates locally minimize 
-the magnitude of the second derivative}, it is desirable to make
+Given that the initial derivative estimates locally minimize 
+the magnitude of the second derivative, it is desirable to make
 the smallest necessary changes to the initial interpolating spline $Q$

### 2022-10-26 [2020-08]_ACMTOMS_(MQSI)/paper/LaTeX/3-spline-rep.tex
Adding LaTeX version of paper.

\section{SPLINE REPRESENTATION}
The monotone quintic spline interpolant $Q(x)$ is represented in terms
of a B-spline basis. The routine {\tt FIT\_SPLINE} in this package
computes the B-spline coefficients $\alpha_i$ of $Q(x)=\sum_{i=1}^{3n}
\alpha_i B_{i,6,t}(x)$ to match the piecewise quintic polynomial
values and (first two) derivatives at the breakpoints $X_i$, where the
spline order is six and the knot sequence $t$ has the breakpoint
multiplicities (6, 3, $\ldots$, 3, 6).  The routine {\tt EVAL\_SPLINE}
evaluates a spline represented in terms of a B-spline basis. A Fortran
2003 implementation {\tt EVAL\_BSPLINE} of the B-spline recurrence
relation evaluation code by C. de Boor [1978] for the value,
derivatives, and integral of a B-spline is also provided.

### 2022-10-26 [2020-08]_ACMTOMS_(MQSI)/paper/LaTeX/5-applications.tex
Adding LaTeX version of paper.

--- 2022-08-25 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
@@
-\heading{5. PERFORMANCE AND APPLICATIONS}
+\section{PERFORMANCE AND APPLICATIONS}
 
-\topinsert
-\textBlack
-\centerline{\epsfxsize=4truein \epsffile{vis/2-piecewise-polynomial.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/3-large-tangent.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/4-signal-decay.eps}}
-{\narrower\noindent\rmVIII Fig.\ 2.
+\begin{figure}[h]
+  \centering
+ \includegraphics[width=4in]{vis/2-piecewise-polynomial.eps}
+ \includegraphics[width=4in]{vis/3-large-tangent.eps}
+ \includegraphics[width=4in]{vis/4-signal-decay.eps}
+\caption{ % \narrower\noindent\rmVIII Fig.\ 2.
 {\ttVIII MQSI} results for three of the functions in the included test
@@
 The {\itVIII signal decay} (bottom) alternates between extreme values
-of steadily decreasing magnitude.
-\par}
-\vskip 6mm
-\centerline{\epsfxsize=4truein \epsffile{vis/5-real-data.eps}}
-{\narrower\noindent\rmVIII Fig.\ 3.
+of steadily decreasing magnitude.}
+\end{figure}
+  %% \vskip 6mm
+\begin{figure}
+  \centering
+  \includegraphics[width=4in]{vis/5-real-data.eps}
+\caption{ % \narrower\noindent\rmVIII Fig.\ 3.
 {\ttVIII MQSI} results when approximating the cumulative distribution
@@
 with a 3.2 GHz CPU performing file read operations from Cameron et al.
-[2019]. The empirical distribution of 30 thousand throughput values is
+\cite{cameron2019moana}. The empirical distribution of 30 thousand throughput values is
 shown in the red dashed line, while the solid line with stylized
 markers denotes the approximation made with {\ttVIII MQSI} given
-equally spaced empirical distribution points from a sample of size 100.
-\par}
-\endinsert
+equally spaced empirical distribution points from a sample of size 100.}
+\end{figure}
 
-\topinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/6-random-monotone.eps}}
-{\narrower\noindent\rmVIII  Fig.\ 4.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=4in]{vis/6-random-monotone.eps}
+\caption{ % \narrower\noindent\rmVIII  Fig.\ 4.
 The {\itVIII random monotone} test poses a particularly challenging
@@
 shifts in slope, the resulting monotone quintic spline interpolant
-provides smooth and reasonable estimates to function values between data.
-\par}
-\iffalse % ---- excluded timing visual (only appears in text) --------
-\vskip 15mm
-\centerline{\epsfxsize=4truein \epsffile{vis/7-runtimes-10.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/8-runtimes-100.eps}}
-{\everymath={\scriptstyle} \narrower\noindent\rmVIII Fig.\ 5.
-Median total runtime of the routine {\ttVIII MQSI} for number of data points
-$n=10$
-to 100 (top) and $n=100$ to 1000 (bottom), generated from 100
-repeated trials averaged over $14$ different testing functions. The
-timings are colored by major algorithmic component, but the vast
-majority of execution time is spent solving the banded linear system
-of equations in the {\ttVIII FIT\_SPLINE} routine. The runtimes for
-the quadratic facet (Algorithm 1) take roughly one microsecond
-($10^{-6}$ seconds) per breakpoint, while the binary search
-(Algorithm 4) takes roughly four microseconds per breakpoint.
-\par}
-\fi % ----------------------------------------------------------------
-\endinsert
+provides smooth and reasonable estimates to function values between data.}
+\end{figure}
 
@@
 banded linear system solve computing the $3n$ $B$-spline coefficients,
-which is ${\cal O}(n)$.  The files {\tt sample\_main.f90} and {\tt
+which is $\mathcal{O}(n)$.  The files {\tt sample\_main.f90} and {\tt
 sample\_main.dat} accompanying the subroutine {\tt MQSI} illustrate Fortran
@@
 cumulative distribution function made by {\tt MQSI} on a computer
-systems application by Cameron et al. [2019] that studies the
+systems application by Cameron et al. \cite{cameron2019moana} that studies the
 distribution of throughput (in bytes per second) when reading files
@@
 770. Figures 5 and 8 demonstrate glaring numerical errors in the
-Schumaker and {\tt BVSPIS} implementations, \textRed minimum working
-examples that reproduce similar errors are discussed in the Appendix.
-\textBlack Overall it can be observed that the minimal and tight
-conditions on monotonicity provided by TOMS 574 tend to be more
-visually appealing than the weaker sufficient conditions utilized by
-{\tt PCHIP}. In general the major determining factor for the quality
-of {\tt MQSI} is the local quadratic facet model. For applications
-where the underlying function is presumed to be a piecewise polynomial
-(of relatively low order), the choice of local quadratic interpolants
-is reasonable. However the consequence of this decision is that
-functions with superquadratic rates of change will tend to have
-consistently underestimated first and second derivatives. This
-limitation is accepted in favor of perfectly capturing lower order
-phenomena. Were future research to supply alternative initializations
-for $C^2$ quintic splines, those could comfortably be made monotone by
-the procedures of {\tt MQSI} and most importantly by the application
-of these sharp monotonicity conditions.
+Schumaker and {\tt BVSPIS} implementations. Minimum working examples
+that reproduce similar errors are provided in the associated software
+{\tt test\_schumaker.r} and {\tt test\_bvspis.f03}. The failure modes
+of these published codes are surprising, but also accentuate how
+difficult it is to correctly handle the numerical complexities of
+polynomial function evaluation.
+
+Overall it can be observed that the minimal and tight conditions on
+monotonicity provided by TOMS 574 tend to be more visually appealing
+than the weaker sufficient conditions utilized by {\tt PCHIP}. In
+general the major determining factor for the quality of {\tt MQSI} is
+the local quadratic facet model. For applications where the underlying
+function is presumed to be a piecewise polynomial (of relatively low
+order), the choice of local quadratic interpolants is
+reasonable. However the consequence of this decision is that functions
+with superquadratic rates of change will tend to have consistently
+underestimated first and second derivatives. This limitation is
+accepted in favor of perfectly capturing lower order phenomena. Were
+future research to supply alternative initializations for $C^2$
+quintic splines, those could comfortably be made monotone by the
+procedures of {\tt MQSI} and most importantly by the application of
+these sharp monotonicity conditions.
 
@@
 
-\topinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/1-toms574.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/2-schumaker.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/3-pchip.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/4-bvspis.eps}}
-{\narrower\noindent\rmVIII Fig.\ 5.
+\begin{figure}[h]
+  \centering
+\includegraphics[width=4in]{vis/comparisons/piecewise-polynomial/1-toms574.eps}
+\includegraphics[width=4in]{vis/comparisons/piecewise-polynomial/2-schumaker.eps}
+\includegraphics[width=4in]{vis/comparisons/piecewise-polynomial/3-pchip.eps}
+\includegraphics[width=4in]{vis/comparisons/piecewise-polynomial/4-bvspis.eps}
+\caption{ % \narrower\noindent\rmVIII Fig.\ 5.
 {\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
@@
   MQSI} otherwise is its exact reproduction of the linear segment
-between $0.3$ and $0.5$.
-\par}
-\endinsert
+between $0.3$ and $0.5$.}
+\end{figure}
 
-\topinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/1-toms574.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/2-schumaker.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/3-pchip.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/4-bvspis.eps}}
-{\narrower\noindent\rmVIII Fig.\ 6.
+\begin{figure}[h]
+  \centering
+\includegraphics[width=4in]{vis/comparisons/large-tangent/1-toms574.eps}
+\includegraphics[width=4in]{vis/comparisons/large-tangent/2-schumaker.eps}
+\includegraphics[width=4in]{vis/comparisons/large-tangent/3-pchip.eps}
+\includegraphics[width=4in]{vis/comparisons/large-tangent/4-bvspis.eps}
+\caption{ % \narrower\noindent\rmVIII Fig.\ 6.
 {\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
@@
 for its greater accuracy when estimating local lower order components
-of functions.
-\par}
-\endinsert
+of functions.}
+\end{figure}
 
-\topinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/1-toms574.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/2-schumaker.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/3-pchip.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/4-bvspis.eps}}
-{\narrower\noindent\rmVIII Fig.\ 7.
+\begin{figure}
+  \centering
+\includegraphics[width=4in]{vis/comparisons/signal-decay/1-toms574.eps}
+\includegraphics[width=4in]{vis/comparisons/signal-decay/2-schumaker.eps}
+\includegraphics[width=4in]{vis/comparisons/signal-decay/3-pchip.eps}
+\includegraphics[width=4in]{vis/comparisons/signal-decay/4-bvspis.eps}
+\caption{ % \narrower\noindent\rmVIII Fig.\ 7.
 {\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
@@
 extrema, where {\ttVIII MQSI} produces continuous and smaller
-magnitude second derivatives.
-\par}
-\endinsert
+magnitude second derivatives.}
+\end{figure}
 
-\topinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/1-toms574.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/2-schumaker.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/3-pchip.eps}}
-\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/4-bvspis.eps}}
-{\narrower\noindent\rmVIII Fig.\ 8.
+\begin{figure}
+  \centering
+\includegraphics[width=4in]{vis/comparisons/random-monotone/1-toms574.eps}
+\includegraphics[width=4in]{vis/comparisons/random-monotone/2-schumaker.eps}
+\includegraphics[width=4in]{vis/comparisons/random-monotone/3-pchip.eps}
+\includegraphics[width=4in]{vis/comparisons/random-monotone/4-bvspis.eps}
+\caption{ % \narrower\noindent\rmVIII Fig.\ 8.
 {\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
@@
 interpolant of the right sided points produces a negative
-slope estimate (invalid) on the left side of the interval.
-\par}
-\endinsert
+slope estimate (invalid) on the left side of the interval.}
+\end{figure}

### 2023-04-26 [2020-08]_ACMTOMS_(MQSI)/paper/LaTeX/1-introduction.tex

\section{INTRODUCTION}

Many domains of science rely on smooth approximations to real-valued
functions over a closed interval. Piecewise polynomial functions
(splines) provide the smooth approximations for animation in graphics
\cite{herman2015techniques,quint2003scalable}, aesthetic structural
support in architecture \cite{brennan2019measure}, efficient
aerodynamic surfaces in automotive and aerospace engineering
\cite{brennan2019measure}, prolonged effective operation of electric
motors \cite{berglund2009planning}, and accurate nonparametric
approximations in statistics \cite{knott2012interpolating}. While
polynomial interpolants and regressors apply broadly, splines are
often a good choice because they can approximate globally complex
functions while minimizing the local complexity of an approximation.

It is often the case that the true underlying function or phenomenon
being modeled has known properties like convexity, positivity, various
levels of continuity, or monotonicity. Given a reasonable amount of
data, it quickly becomes difficult to achieve desirable properties in
a single polynomial function. In general, the maintenance of function
properties through interpolation/regression is referred to as {\it
  shape preserving} \cite{fritsch1980monotone,gregory1985shape}. The
specific properties the present algorithm will preserve in
approximations are monotonicity and $C^2$ continuity.  Notably this
work does not consider convexity preservation, as Mulansky and Neamtu
\cite{mulansky1991existence} prove that there exists strictly convex
data sets which do not allow convex interpolation with $C^2$ splines
of fixed degree.  In addition to previously mentioned applications,
being $C^2$ and monotonicity preserving is crucially important in
statistics to the approximation of a cumulative distribution function
and subsequently the effective generation of random numbers from a
specified distribution \cite{ramsay1988monotone}.  A spline function
with these properties could approximate a cumulative distribution
function to a high level of accuracy with relatively few intervals. A
twice continuously differentiable approximation to a cumulative
distribution function (CDF) would produce a corresponding probability
density function (PDF) that is continuously differentiable, which is
desirable.

The existing research in shape preserving interpolatory splines is
rich and filled with many approaches for many different
applications. In the context of this work, the unique quality of shape
preserving spline algorithms can be observed through: (1) the type of
spline (rational or Hermite), (2) the polynomial order and level of
continuity, and (3) the sharpness, sufficiency, and necessity of the
conditions used to establish shape preserving properties. While the
choice of spline representation is less important, achieving higher
levels of continuity is often more difficult than lower levels of
continuity. Furthermore, all referenced algorithms establish
sufficient conditions for shape preservation, but it is less common
and much more difficult to provide {\it sharp and necessary}
conditions for monotonicity. The following two paragraphs highlight a
collection of research related to the construction of $C^1$ and $C^2$
shape preserving splines. From this foundation, the value and utility
of the code presented here can be more readily appreciated.

DeVore \cite{devore1977monotone} proves that monotone spline
interpolants have improved accuracy over their nonmonotone
counterparts when approximating monotone functions, while Costantini
\cite{constantini1986monotone} proves that monotone spline
interpolants to data of fixed continuity and order exist in
general. This groundwork establishes the potential for monotone spline
interpolation. Schumaker \cite{schumaker1983shape} and McAllister and
Roulier \cite{mcallister1981algorithm} (independently) establish
necessary and sufficient conditions for monotone $C^1$ quadratic
splines through the insertion of additional knots and Fritsch
\cite{fritsch1982LLNL} establishes simplified sufficient conditions
for monotone $C^1$ cubic splines that require no knot insertion.
Gregory and Delbourgo \cite{gregory1982piecewise} prove necessary and
sufficient conditions for a closed form solution to monotone $C^1$
quadratic rational spline interpolation with nonlinear boundary
equations and an iterative approach. Delbourgo and Gregory
\cite{delbourgo1983rational} extend that same work to achieve
sufficient conditions for $C^2$ continuity with a cubic rational
spline given additional tension parameters defined by users. Huynh
\cite{huynh1993accurate} similarly arrives at monotone $C^1$ cubic
spline interpolants and several necessary nonlinear boundary
conditions on monotonicity.

Continuing towards monotone $C^2$ spline interpolation, Fiorot and
Tabka \cite{fiorot1991shape} prove a simple method for determining the
existence of a monotone $C^2$ cubic interpolating spline, but note
that functions of this kind do not always exist for arbitrary convex
monotone data sets.  Pruess \cite{pruess1993shape} proposes sufficient
conditions for a cubic $C^2$ shape preserving spline method while
acknowledging that quintic splines are generally necessary for
monotone $C^2$ spline interpolants that do not insert new
knots. Similarly, Manni and Sablonni\`ere \cite{manni1997monotone}
construct sufficient conditions for a monotone $C^2$ cubic spline by
adding two additional knots per interval, then Cravero and Manni
\cite{cravero2003shape} extend that methodology to arrive at monotone
$C^3$ interpolating splines by progressively increasing tension
through Bezier control polygons. Costantini
\cite{constantini1997boundary} provides a method for monotone $C^2$
quintic spline interpolation based on sufficient and necessary
boundary value conditions, but the necessary conditions are not
sharp. Dougherty, et\ al.\ \cite{dougherty1989nonnegativity} construct
monotone $C^2$ Hermite splines and in commentary recommend against the
direct optimization for smoothness parameters like global $L^2$,
rather promoting problem specific definitions of geometric
niceness. Both Wang and Tan \cite{wang2004rational} as well as Yao and
Nelson \cite{yao2018unconditionally} propose sufficient conditions for
$C^2$ quartic splines, while Piah and Unsworth \cite{piah2011improved}
improve upon those sufficient conditions for $C^2$ quartic rational
spline interpolation. Abbas, et\ al.\ \cite{abbas2012monotonicity}
formulate a set of sufficient conditions for monotone $C^2$ cubic
rational splines and similar work is extended to $C^2$ quartic
rationals (and splines with arbitrary smoothness) by Zhu and Han
\cite{zhu2015rational,zhu2015shape}. Ulrich and Watson
\cite{ulrich1990positivity} arrive at sharp necessary and sufficient
conditions for monotone $C^2$ quintic splines, while He{\ss} and
Schmidt \cite{hess1994positive} produce similar sufficient conditions
for $C^2$ quintic splines. In addition, a method for constructing an
arbitrarily smooth monotone fit exists
\cite{leitenstorfer2006generalized} as well as a method for
arbitrarily smooth shape preservation \cite{han2018shape}, but these
methods are enabled by sufficient conditions only.

The currently available peer reviewed and published software for
monotone piecewise polynomial interpolation is severely limited in
comparison with the number of published approaches mentioned
above. This is partially indicative of how difficult it is to properly
handle the numerical conditions that arise when constructing and
evaluating precise shape preserving splines, as well as how difficult
it is to create code that correctly conforms with the theoretical
expectations. The sufficient $C^1$ cubic spline method PCHIP of
Fritsch and Carlson \cite{fritsch1980monotone} is available through
the SciPy Python package ($>$1 billion downloads), and the $C^1$
piecewise quadratic method of Schumaker \cite{schumaker1983shape} is
available as an R package (300,000 downloads). The $C^1$ piecewise
quadratic method of McAllister and Roulier
\cite{mcallister1981algorithm,mcallister1981shape} is available in
FORTRAN (1000 downloads) and the sufficient $C^2$ piecewise quintic
method of Costantini
\cite{constantini1997boundary,constantini1997bvspis} BVSPIS in FORTRAN
as well (500 downloads). Based on publicly available download records,
it is assumed that the code by Fritsch \cite{fritsch1980monotone} for
monotone $C^1$ cubic spline interpolation is the predominant code for
constructing monotone interpolants at present.

The theory for {\it sharp} necessary and sufficient bounds for
monotone $C^2$ quintic interpolation has been provided
\cite{ulrich1990positivity,ulrich1994positivity} and that theory was
recently utilized in a proposed algorithm
\cite{lux2020algorithm,lux2020thesis} for monotone $C^2$ quintic
spline construction, however no published mathematical software exists
for the quintic case based on sharp monotonicity conditions. The
software presented here represents the first published software
package for producing $C^2$ shape preserving splines based on sharp
monotonicity conditions. This work improves upon the algorithm
presented by Lux et\ al.\ \cite{lux2020algorithm} by refactoring
computations for improved numerical stability, estimating minimum
magnitude second derivatives at breakpoints with a quadratic facet
model, and using a binary search to reduce the magnitude of the
modifications made to initial derivative estimates when constructing a
monotone spline interpolant.

\subsection*{Overview}
This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
sharp necessary and sufficient conditions in Ulrich and Watson
\cite{ulrich1994positivity} for the construction of monotone $C^2$
quintic spline interpolants of monotone data. Precisely, the problem
is, given a strictly increasing sequence $X_1<X_2< \cdots <X_n$ of
breakpoints with corresponding monotone increasing function values
$Y_1\le Y_2\le \cdots \le Y_n$, find a $C^2$ monotone increasing
quintic spline $Q(x)$ with the same breakpoints satisfying
$Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI} actually does something
slightly more general, producing $Q(x)$ that is monotone increasing
(decreasing) wherever the data is monotone increasing (decreasing).)

The remainder of this paper is structured as follows: Section 2
provides the algorithms for constructing a monotone $C^2$ quintic
spline interpolant to monotone data, Section 3 outlines the method of
spline representation ($B$-spline basis) and evaluation, Section 4
analyzes the complexity and sensitivity of the algorithms in {\tt
  MQSI}, and Section 5 presents timing data and some graphs of
constructed interpolants as well as a visual comparison with existing
monotone spline packages.

### 2023-04-26 [2020-08]_ACMTOMS_(MQSI)/paper/LaTeX/2-monotone-quintic.tex

\section{MONOTONE QUINTIC INTERPOLATION}

In order to construct a monotone quintic interpolating spline, two
primary problems must be solved. First, reasonable derivative values
at data points need to be estimated. Second, the estimated derivative
values need to be modified to enforce monotonicity on all polynomial
pieces.

Fritsch and Carlson \cite{fritsch1980monotone} originally proposed the
use of central differences to estimate derivatives, however this often
leads to extra and unnecessary {\it wiggles} in the spline when used
to approximate second derivatives. Modern shape-preserving spline
implementations use a weighted harmonic mean to estimate derivative
values at breakpoints \cite{moler2004numerical}, however this method
also yields approximations whose second derivative functions often
have large local $L^2$ norm (approximations with large {\it
  wiggle}). In an attempt to capture the local shape of the data while
minimizing wiggle, this package uses a facet model from image
processing \cite{haralick1981facet} to estimate first and second
derivatives at breakpoints. Rather than picking a local linear or
quadratic fit with minimal residual, this work instead selects the
local quadratic interpolant with minimum magnitude second derivative
(hereafter referred to as a {\it quadratic facet model}). The term
``facet'' refers to one piece of an image surface modelled by a
two-dimensional piecewise polynomial.

%% ===================================================================
%%                   Algorithm 1: QUADRATIC_FACET
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 1:}
{\tt QUADRATIC\_FACET}$\bigl(X(1{:}n)$, $Y(1{:}n)$, $i \bigr)$

\nobreak
where $X_j$, $Y_j$ $\in \mathbf{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
\le n$, and $n \ge 3$. Returns the first and second derivative at 
$X_i$ of the local quadratic interpolant with minimum magnitude
second derivative. Approximate equality is denoted with
$\approx$ and considers two numbers to be equal once they are within
the machine precision $\epsilon$ of each other.

}
{\parindent=3mm
\begin{itemize}
%% -------------------------------------------------------------------
\item[]{ {\tt if} $\bigl((i\ne1\wedge Y_i \approx Y_{i-1})$ {\tt or}
$(i\ne n\wedge Y_i \approx Y_{i+1})\bigr)$ {\tt then return $(0,0)$}}

\item[]{ {\tt else if} $i=1$ {\tt then}}
\begin{itemize}
\item[]{ $f_1\:\hbox{interpolant to }(X_1,Y_1)$,}
  $(X_2,Y_2)$, and $(X_3,Y_3)$.
\item[]{ {\tt if} $\bigl(Df_1(X_1)(Y_2-Y_1)<0\bigr)$ {\tt then return}
  $\bigl(0,0\bigr)$}
\item[]{ {\tt else return} $\bigl(Df_1(X_1),D^2f_1\bigr)$}
\item[]{ {\tt endif}
}
\end{itemize}
\item[]{ {\tt else if} $i=n$ {\tt then}}
\begin{itemize}
\item[]{ $f_1\:\hbox{interpolant to }(X_{n-2},Y_{n-2})$,
  $(X_{n-1},Y_{n-1})$, and $(X_n,Y_n)$.}
\item[]{ {\tt if} $\bigl(Df_1(X_n)(Y_n-Y_{n-1})<0\bigr)$ {\tt then return}
  $\bigl(0,0\bigr)$}
\item[]{ {\tt else return} $\bigl(Df_1(X_n),D^2f_1\bigr)$}
\item[]{ {\tt endif}}
\end{itemize}
\item[]{ {\tt else if} $\bigl(1<i<n\wedge (Y_{i+1} - Y_i)(Y_i - Y_{i-1}) < 0
\bigr)$ {\tt then}}
\begin{itemize}
\item[]{ The point $(X_i$, $Y_i)$ is an extreme point. The
quadratic with minimum magnitude second derivative that has slope
zero at $X_i$ will be the facet chosen.}
\item[]{ $f_1\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$, $(X_i,Y_i)$,
  and $Df_1(X_i) = 0$.}
\item[]{ $f_2\:\hbox{interpolant to }(X_i,Y_i)$, $(X_{i+1},Y_{i+1})$,
  and $Df_2(X_i) = 0$.}
\item[]{ {\tt if} $\bigl(|D^2f_1| \leq |D^2f_2|\bigr)$ {\tt then
  return} $\bigl(0$, $D^2f_1\bigr)$}
\item[]{ {\tt else return} $\bigl(0$, $D^2f_2\bigr)$}
\item[]{ {\tt endif}}
\end{itemize}
\item[]{ {\tt else}}
\begin{itemize}
\item[]{ {The point $(X_i$, $Y_i)$ is in a monotone segment of
data. In the following, it is possible that $f_1$ or $f_3$ do
not exist because $i \in \{2, n-1\}$. In those cases, the minimum
magnitude second derivative among existing quadratics is chosen.}}
\item[]{ $f_1\:\hbox{interpolant to }(X_{i-2},Y_{i-2})$,
  $(X_{i-1},Y_{i-1})$, and $(X_i,Y_i)$.}
\item[]{ $f_2\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$,
  $(X_i,Y_i)$, and $(X_{i+1},Y_{i+1})$.}
\item[]{ $f_3\:\hbox{interpolant to }(X_i,Y_i)$,
  $(X_{i+1},Y_{i+1})$, and $(X_{i+2},Y_{i+2})$.}
\item[]{ {\tt if} $\bigl(Df_1(X_i)(Y_i-Y_{i-1}) \ge 0 \wedge
  |D^2f_1| = \min\bigl\{ |D^2f_1|,
  |D^2f_2|, |D^2f_3|\bigr\} \bigr)$ {\tt then}}
\item[]{ \codent {\tt return} $\bigl(Df_1(X_i),D^2f_1\bigr)$}
\item[]{ {\tt else if} $\bigl(Df_2(X_i)(Y_i-Y_{i-1}) \ge 0 \wedge
  |D^2f_2| = \min\bigl\{ |D^2f_1|,
  |D^2f_2|, |D^2f_3|\bigr\} \bigr)$}
\item[]{ \codent {\tt then return} $\bigl(Df_2(X_i),D^2f_2\bigr)$}
\item[]{ {\tt else if} $\bigl(Df_3(X_i)(Y_{i+1}-Y_i) \ge 0\bigr)$ {\tt then}}
\item[]{ \codent {\tt return} $\bigl(Df_3(X_i),D^2f_3\bigr)$}
\item[]{ {\tt else return} $(0, 0)$}
\item[]{ {\tt endif}}
\end{itemize}
\item[]{ {\tt endif}}
\end{itemize}
}
\vskip 5mm
%% ----------------------------------------------------------------------

For constructing a quadratic interpolant in $x$ over the interval
$[L,R]$, the Chebyshev basis $1$, $z$, $2z^2-1$ is used, where
$z={x-(L+R)/2 \over (R-L)/2}$. The estimated derivative values by the
quadratic facet model are not guaranteed to produce monotone quintic
polynomial segments. Ulrich and Watson \cite{ulrich1990positivity}
established tight constraints on the monotonicity of a quintic
polynomial piece and He{\ss} and Schmidt \cite{hess1994positive} gave
simpler conditions for a special case.  The following algorithm
implements a sharp check for monotonicity by considering the
nondecreasing case. The nonincreasing case is handled similarly.

%% ===================================================================
%%                    Algorithm 2: IS_MONOTONE
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 2:}
{\tt IS\_MONOTONE}$\bigl(x_0$, $x_1$, $f(x_0)$, $Df(x_0)$,
$D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1) \bigr)$

\nobreak

where $x_0$, $x_1 \in {\mathbf R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$. Approximate equality is denoted with
$\approx$ and considers two numbers to be equal once they are within
the machine precision $\epsilon$ of each other.

}
{\parindent=6mm
\begin{itemize}
%% -------------------------------------------------------------------
\item[1.]{ {\tt if} $\bigl(f(x_0) \approx f(x_1)\bigr)$ {\tt then}}
\item[2.]{ \codent {\tt return} $\bigl( 0 = Df(x_0) = Df(x_1)
  = D^2f(x_0) = D^2f(x_1) \bigr)$}
\item[3.]{ {\tt endif}}
\item[4.]{ {\tt if} $\bigl(Df(x_0) < 0$ {\tt or} $Df(x_1) < 0\bigr)$ {\tt
  then return FALSE endif}}
\item[5.]{ $w \: x_1 - x_0$}
\item[6.]{ $z \: f(x_1) - f(x_0)$
}
\item[]{ {The necessity of Steps 1--4 follows directly from the
  fact that $f$ is $C^2$. The following Steps 7--13 coincide with a
  simplified condition for quintic monotonicity that reduces to one of
  cubic positivity studied by Schmidt and He{\ss} [1988]. Given
  $\alpha$, $\beta$, $\gamma$, and $\delta$ as defined by Schmidt and
  He{\ss}, monotonicity results when $\alpha \geq 0$, $\delta \geq 0$,
  $\beta \geq \alpha - 2 \sqrt{\alpha \delta}$, and $\gamma \geq
  \delta - 2 \sqrt{\alpha \delta}$.  Step 4 checked for $\delta < 0$,
  Step 8 checks $\alpha < 0$, Step 10 checks $\beta < \alpha - 2
  \sqrt{\alpha \delta}$, and Step 11 checks $\gamma < \delta - 2
  \sqrt{\alpha \delta}$. If none of the monotonicity conditions are
  violated, then the degree five piece is monotone and Step 12
  concludes.}
}
\item[7.]{ {\tt if} $\bigl(Df(x_0) \approx 0$ {\tt or} $Df(x_1) \approx
0\bigr)$ {\tt then}}
\item[8.]{ \codent {\tt if} $\bigl(D^2f(x_1)w > 4Df(x_1)$ {\tt then
return FALSE endif}}
\item[9.]{ \codent $t \: 2 \sqrt{Df(x_0) (4Df(x_1) - D^2f(x_1) w) }$}
\item[10.]{ \codent {\tt if} $\bigl(t + 3Df(x_0) + D^2f(x_0)w < 0 \bigr)$
  {\tt then return FALSE endif}}
\item[11.]{ \codent {\tt if} $\bigl(60z - w\bigl(24Df(x_0) + 32Df(x_1) - 2t
  + w(3D^2f(x_0) - 5D^2f(x_1))\bigr) < 0\bigr)$}
\item[]{   \codent \codent {\tt then return FALSE endif}}
\item[12.]{ \codent {\tt return TRUE}}
\item[13.]{ {\tt endif}
}
\item[]{ {The following code considers the full quintic monotonicity
case studied by Ulrich and Watson \cite{ulrich1994positivity}. Given $\tau_1$, $\alpha$,
$\beta$, and $\gamma$ as defined by Ulrich and Watson, a quintic
piece is proven to be monotone if and only if
$\tau_1 > 0$, and $\alpha, \gamma > -(\beta+2)/2$ when $\beta \leq 6$,
and $\alpha, \gamma > -2 \sqrt{\beta-2}$ when $\beta > 6$.
Step 14 checks $\tau_1 \le 0$, Steps 19 and 20 determine monotonicity based
on $\alpha$, $\beta$, and $\gamma$.}
}
\item[14.]{ {\tt if} $\bigl( w\bigl(2\sqrt{Df(x_0)\,Df(x_1)} - 3(Df(x_0) +
  Df(x_1))\bigr) - 24z \leq 0 \bigr)$}
\item[]{ \codent {\tt then return FALSE endif}}
\item[15.]{ $t \: \bigl(Df(x_0)\, Df(x_1)\bigr)^{3/4}$}
\item[16.]{ $\alpha \: (4 Df(x_1) - D^2f(x_1)w) \sqrt{Df(x_0)} / t$}
\item[17.]{ $\gamma \: (4 Df(x_0) - D^2f(x_0)w) \sqrt{Df(x_1)} / t$}
\item[18.]{ $\displaystyle \beta \: {60z/w + 3\bigl(w(D^2f(x_1) -
  D^2f(x_0)) - 8(Df(x_0) + Df(x_1))\bigr) \over 2 \sqrt{Df(x_0)\,Df(x_1)}}$}
\item[19.]{ {\tt if} $(\beta \leq 6)$ {\tt then return}
$\bigl( \min\{\alpha,\gamma\} > - (\beta + 2) / 2 \bigr)$}
\item[20.]{ {\tt else} {\tt return}
$\bigl( \min\{\alpha,\gamma\} > -2 \sqrt{\beta - 2}\,\bigr)$}
\item[21.]{ {\tt endif}}
\end{itemize}
}
\vskip 5mm
%% ----------------------------------------------------------------------

It is shown by Ulrich and Watson \cite{ulrich1994positivity} that when
$0 = DQ(X_i) = DQ(X_{i+1}) = D^2Q(X_i) = D^2Q(X_{i+1})$, the quintic
polynomial over $[X_i$, $X_{i+1}]$ is guaranteed to be monotone. Using
this fact, the following algorithm shrinks (in magnitude) initial
derivative estimates until a monotone spline is achieved and outlines
the core routine in the accompanying package.

%% ===================================================================
%%       Algorithm 3: MQSI - Monotone quintic spline interpolant
\vskip 3mm
{\parindent 0mm
{\bf Algorithm 3:}
{\tt MQSI}$\bigl(X(1{:}n), Y(1{:}n) \bigr)$

\nobreak
where $(X_i,Y_i) \in \mathbf{R}\times\mathbf{R}$, $i = 1$, $\ldots$, $n$ are data
points. Returns monotone quintic spline interpolant $Q(x)$ such that
$Q(X_i) = Y_i$ and is monotone increasing (decreasing) on all
intervals that $Y_i$ is monotone increasing (decreasing).

}
{\parindent=6mm
\begin{itemize}
\item[]{ Approximate first and second derivatives at $X_i$ with {\tt
  QUADRATIC\_FACET}.}
\item[]{ {\tt for} $i\:1$ {\tt step} 1 {\tt until} $n$ {\tt do}}
\item[]{ \codent $(u_i$, $v_i)\:$ {\tt QUADRATIC\_FACET}$(X$, $Y$, $i)$}
\item[]{ {\tt enddo}}
\item[]{ Identify and store all intervals where $Q$ is nonmonotone in a
queue $\mathcal{Q}$.}
\item[]{ {\tt for} $i\:1$ {\tt step} 1 {\tt until} $n-1$ {\tt do}}
\item[]{ \codent {\tt if not IS\_MONOTONE}$(X_i$, $X_{i+1}$, $Y_i$, $u_i$,
$v_i$, $Y_{i+1}$, $u_{i+1}$, $v_{i+1})$ {\tt then}}
\item[]{ \codent\codent Add interval $\bigl(X_i$, $X_{i+1}\bigr)$ to
queue $\mathcal{Q}$.}
\item[]{ \codent {\tt endif}}
\item[]{ {\tt enddo}}
\item[]{ {\tt do while} $\bigl($ queue $\mathcal{Q}$ of intervals is nonempty $\bigr)$}
\begin{itemize}
\item[]{ Shrink (in magnitude) $DQ$ (in $u$) and $D^2Q$ (in $v$)
  that border intervals where $Q$ is nonmonotone.  }
\item[]{ Identify and store remaining intervals where $Q$ is
  nonmonotone in queue $\mathcal{Q}$.}
\end{itemize}
\item[]{ {\tt enddo}}
\item[]{ Construct and return a $B$-spline representation of $Q(x)$.}
\end{itemize}
}
\vskip 3mm
%% ----------------------------------------------------------------------

Since {\tt IS\_MONOTONE} can handle both nondecreasing and
nonincreasing simultaneously by taking into account the sign of $z$,
Algorithm 3 produces $Q(x)$ that is monotone increasing (decreasing)
over exactly the same intervals that the data $(X_i$, $Y_i)$ is
monotone increasing (decreasing).

Given that the initial derivative estimates locally minimize the
magnitude of the second derivative, it is desirable to make the
smallest necessary changes to the initial interpolating spline $Q$
while enforcing monotonicity. In practice a binary search for the
boundary of monotonicity is used in place of solely shrinking $DQ$ and
$D^2Q$ at breakpoints adjoining {\it active\/} intervals: intervals
over which $Q$ is nonmonotone at least once during the search. The
binary search considers a Boolean function $B_i(s)$, for $0 \le s \le
1$, that is true if the order six polynomial piece of $Q(x)$ on $[X_i,
  X_{i+1}]$ matching derivatives $Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$,
$D^2Q(X_i)=s \,v_i$ at $X_i$, and derivatives $Q(X_{i+1})=Y_{i+1}$,
$DQ(X_{i+1})=s \,u_{i+1}$, $D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is
monotone, and false otherwise.  The binary search is only applied at
those breakpoints adjoining intervals $[X_i, X_{i+1}]$ over which $Q$
is nonmonotone and hence $B_i(1)$ is false.  It is further assumed
that there exists $0 \le s^* \le 1$ such that $B_i(s)$ is true for $0
\le s \le s^*$ and false for some $1 > s > s^*$. Since the derivative
conditions at interior breakpoints are shared by intervals left and
right of the breakpoint, the binary search is performed at all
breakpoints simultaneously.  Specifically, the monotonicity of $Q$ is
checked on all active intervals in each step of the binary search to
determine the next derivative modification at each breakpoint. The
goal of this search is to converge on the boundary of the monotone
region in the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described
in Ulrich and Watson \cite{ulrich1994positivity}) for all
intervals. This multiple-interval binary search allows the value zero
to be obtained for all (first and second) derivative values in a fixed
maximum number of computations, hence has no effect on computational
complexity order. This binary search algorithm is outlined below.

%% ===================================================================
%%              Algorithm 4: Global binary search
\vskip 5mm {\parindent 0mm {\bf Algorithm 4:}
{\tt BINARY\_SEARCH}$\bigl(X(1{:}n), Y(1{:}n), u(1{:}n), v(1{:}n) \bigr)$

\nobreak
where $(X_i,Y_i) \in \mathbf{R}\times\mathbf{R}$, $i = 1$, $\ldots$, $n$ are
data points, and $Q(x)$ is a quintic spline interpolant such that
$Q(X_i) = Y_i$, $DQ(X_i) = u_i$, $D^2Q(X_i) = v_i$. Modifies
derivative values ($u$ and $v$) of $Q$ at data points to ensure
{\tt IS\_MONOTONE} is true for all intervals defined by adjacent data
points, given a desired precision $\mu \in {\mathbf R}$. proj$\bigl(w$,
int$(a,b)\bigr)$ denotes the projection of $w$ onto the closed
interval with endpoints $a$ and $b$.

}
{\parindent=6mm
\begin{itemize}
\item[]{ Initialize the step size $s$, make a copy of data defining $Q$,
and construct three queues necessary for the multiple-interval binary search.}
\item[]{ $s \: 1$}
\item[]{ $(\hat u,\hat v) \: (u,v)$}
\item[]{ {\tt searching} $\:$ {\tt TRUE}}
\item[]{ {\tt checking} $\:$ empty queue for holding left indices of {\it intervals}}
\item[]{ {\tt growing} $\:$ empty queue for holding indices of {\it data points}}
\item[]{ {\tt shrinking} $\:$ empty queue for holding indices of {\it data points}}
\item[]{ {\tt for} $i\:1$ {\tt step} 1 {\tt until} $n-1$ {\tt do}}
\begin{itemize}
\item[]{ {\tt if not IS\_MONOTONE}$\bigl( X_i$, $X_{i+1}$, $Y_i$, $u_i$,
$v_i$, $Y_{i+1}$, $u_{i+1}$, $v_{i+1} \bigr)$ {\tt then}}
\item[]{ \codent Add data indices $i$ and $i+1$ to queue {\tt shrinking}.}
\item[]{ {\tt endif}}
\end{itemize}
\item[]{ {\tt enddo}}
\item[]{ {\tt do while} $\bigl(${\tt searching or} ({\tt shrinking} is
nonempty)$\bigr)$}
\begin{itemize}
\item[]{ Compute the {\it step size} $s$ for this iteration of the search.}
\item[]{ {\tt if  searching  then} $s \: \max\{\mu, s/2\}$
{\tt else} $s \: 3s/2$ {\tt endif}}
\item[]{ {\tt if} $(s=\mu)$ {\tt then searching} $\:$ {\tt FALSE};
  clear queue {\tt growing endif}}
\item[]{ Increase in magnitude $u_i$ and $v_i$ for all data indices
  $i$ in {\tt growing} such that the points $X_i$ are strictly adjoining
  intervals over which $Q$ is monotone.}
\item[]{ {\tt for} ($i \in$ {\tt growing}) {\tt and} ($i \not\in$
{\tt shrinking}) {\tt do}}
\item[]{ \codent $u_i \: \hbox{proj}\bigl(u_i + s\,\hat u_i, \hbox{int}
  (0,\hat u_i)\bigr)$}
\item[]{ \codent $v_i \: \hbox{proj}\bigl(v_i + s\,\hat v_i, \hbox{int}
  (0,\hat v_i)\bigr)$}
\item[]{ \codent Add data indices $i-1$ (if not 0) and $i$ (if not $n$) to queue
  {\tt checking}.}
\item[]{ {\tt enddo}}
\item[]{ Decrease in magnitude $u_i$ and $v_i$ for all data indices
  $i$ in {\tt shrinking} and ensure those data point indices are placed
  into {\tt growing} when {\tt searching}.}
\item[]{ {\tt for} $i \in$ {\tt shrinking do}}
\item[]{ \codent If {\tt searching}, then add index $i$ to queue
  {\tt growing} if not already present.}
\item[]{ \codent $u_i \: \hbox{proj}\bigl(u_i - s\,\hat u_i, \hbox{int}
  (0,\hat u_i)\bigr)$}
\item[]{ \codent $v_i \: \hbox{proj}\bigl(v_i - s\,\hat v_i, \hbox{int}
  (0,\hat v_i)\bigr)$}
\item[]{ \codent Add data indices $i-1$ (if not 0) and $i$ (if not $n$) to queue
  {\tt checking}.}
\item[]{ {\tt enddo}}
\item[]{ Empty queue {\tt shrinking}, then check all intervals left-indexed
  in queue {\tt checking} for monotonicity with {\tt IS\_MONOTONE},
  placing data endpoint indices of intervals over which $Q$ is nonmonotone
  into queue {\tt shrinking}.}
\item[]{ Clear queue {\tt shrinking}.}
\item[]{ {\tt for} $i \in$ {\tt checking do}}
\item[]{ \codent
  {\tt if not IS\_MONOTONE}$\bigl( X_i$, $X_{i+1}$, $Y_i$, $u_i$,
  $v_i$, $Y_{i+1}$, $u_{i+1}$, $v_{i+1} \bigr)$ {\tt then}}
\item[]{ \codent \codent Add data indices $i$ and $i+1$ to {\tt shrinking}.}
\item[]{ \codent {\tt endif}}
\item[]{ {\tt enddo}}
\item[]{ Clear queue {\tt checking}.}
\end{itemize}
\item[]{ {\tt enddo }}
\end{itemize}
}
\vskip 5mm
%% ----------------------------------------------------------------------

%% \itemitem{} Take search steps either modifying derivative values
%%   towards monotonicity (zero) or towards the original values over all
%%   intervals that are either nonmonotone or previously modified.

In the subroutine {\tt MQSI}, $\mu = 2^{-26}$, which results in 26
guaranteed search steps for all intervals that are initially
nonmonotone. An additional 43 steps could be required to reduce a
derivative magnitude to zero with step size growth rate of $3/2$. This
can only happen when $Q$ becomes nonmonotone on an interval for the
first time while the step size equals $\mu$, but for which the only
viable solution is a derivative value of zero. The maximum number of
steps is due to the fact that $\sum_{i=0}^{42} \mu (3/2)^i > 1$. In
total {\tt BINARY\_SEARCH} search could require 69 steps.

### 2023-04-26 [2020-08]_ACMTOMS_(MQSI)/paper/LaTeX/4-analysis.tex

--- 2022-08-25 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
@@
-\topinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
-{\everymath={\scriptstyle}
+\begin{figure}
+\includegraphics[width=4in]{vis/1-sensitivity.eps}
+\caption{ % \everymath={\scriptstyle}  % \narrower\noindent\rmVIII Fig.\ 1.
+  A demonstration of the quadratic facet model's sensitivity to small
+  data perturbations. This example is composed of two quadratic
+  functions $f_1(x) = x^2$ over points $\{1$, $2$, $5/2\}$, and
+  $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$, $4\}$. Notably,
+  $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have equal second
+  derivatives. Given the exact five data points seen above, the
+  quadratic facet model produces the slope seen in the solid blue line
+  at $x = 5/2$. However, by subtracting the value of $f_3$ $=
+  \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$ is the
+  machine precision ($2^{-52}$ for an IEEE 64-bit real), the quadratic
+  facet model produces the slope seen in the dashed red line at $x =
+  5/2$. This is the nature of a facet model and a side effect of
+  associating data with local facets.}
+\end{figure}
 
-\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
-  facet model's sensitivity to small data perturbations. This example is
-  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
-  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
-  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have
-  \red{equal second derivatives}. Given the exact five data points
-  seen above, the quadratic facet model produces the slope seen in the
-  solid blue line at $x = 5/2$. However, by subtracting the value of
-  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
-  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
-  quadratic facet model produces the slope seen in the dashed red line
-  at $x = 5/2$. This is the nature of a facet model and a side effect
-  of associating data with local facets.
-\par} \endinsert
+\section{COMPLEXITY AND SENSITIVITY}
 
-\heading{4. COMPLEXITY AND SENSITIVITY}
-
-Algorithms 1 and 4 have ${\cal O}(n)$ runtime for $n$ data
-points. Algorithm 2 has a fixed cost ${\cal O}(1)$. Given a fixed
-schedule for shrinking derivative values, Algorithm 3 has a ${\cal
-  O}(n)$ runtime for $n$ data points. In execution, the majority of
-the time, still ${\cal O}(n)$, is spent solving the banded linear
-system of equations for the B-spline coefficients. Thus for $n$ data
-points, the overall execution time is ${\cal O}(n)$.
+Algorithms 1 and 4 have $\mathcal{O}(n)$ runtime for $n$ data
+points. Algorithm 2 has a fixed cost $\mathcal{O}(1)$. Given a fixed
+schedule for shrinking derivative values, Algorithm 3 has a
+$\mathcal{O}(n)$ runtime for $n$ data points. In execution, the
+majority of the time, still $\mathcal{O}(n)$, is spent solving the banded
+linear system of equations for the B-spline coefficients. Thus for $n$
+data points, the overall execution time is $\mathcal{O}(n)$.
 
@@
 The binary search procedure provably converges on the boundary of the
-region of monotonicity precisely defined by Ulrich and Watson [1994]
-through the application of a Boolean function that is guaranteed to be
-true at one end of an interval. Since a local quadratic interpolant is
-applied for initial function derivative estimates, the approximation
-order of the resulting fit is $O(h^3)$ (as for any second order
-approximation), but it should be noted that asymptotic approximation
-order is rarely of importance when considering the scattered sparse
-data that $C^2$ approximations like {\tt MQSI} provide. Were exact
-evaluations provided, higher order methods always win (all
-other things being equal) as the data density increases. For sparse
-data, what constitutes a better fit is either subjective or dependent
-on the problem.
+region of monotonicity precisely defined by Ulrich and Watson
+\cite{ulrich1994positivity} through the application of a Boolean
+function that is guaranteed to be true at one end of an
+interval. Since a local quadratic interpolant is applied for initial
+function derivative estimates, the approximation order of the
+resulting fit is $O(h^3)$ (as for any second order approximation), but
+it should be noted that asymptotic approximation order is rarely of
+importance when considering the scattered sparse data that $C^2$
+approximations like {\tt MQSI} provide. Were exact evaluations
+provided, higher order methods always win (all other things being
+equal) as the data density increases. For sparse data, what
+constitutes a better fit is either subjective or dependent on the
+problem.

### 2023-04-26 [2020-08]_ACMTOMS_(MQSI)/paper/TeX/1-introduction.tex

--- 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
@@
 preserve in approximations are monotonicity and $C^2$ continuity. 
-\beginred
 Notably this work does not consider convexity preservation, as Mulansky
@@
 do not allow convex interpolation with $C^2$ splines of fixed degree.
-\endred
-In addition to previously mentioned applications, \red{being $C^2$ and monotonicity preserving is}
+In addition to previously mentioned applications, being $C^2$ and monotonicity preserving is
 crucially important in statistics to the approximation of a cumulative
@@
 differentiable, which is desirable.
-
-\beginred
 
@@
 
-\endred
-
-The theory for \red{{\it sharp}} necessary and sufficient bounds for
+The theory for {\it sharp} necessary and sufficient bounds for
 monotone $C^2$ quintic interpolation has been provided [Ulrich and
@@
 quintic spline construction, however no published mathematical
-software exists for the quintic case \red{based on sharp monotonicity
-conditions}. The software presented here represents the first published
+software exists for the quintic case based on sharp monotonicity
+conditions. The software presented here represents the first published
 software package for producing $C^2$ shape preserving splines
-\red{based on sharp monotonicity conditions}. This work improves upon
+based on sharp monotonicity conditions. This work improves upon
 the algorithm presented by Lux et\ al.\ [2020] by refactoring
@@
 
-\beginred
-
 \heading{Overview}
-
-\endred
-
 This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
@@
   MQSI}, and Section 5 presents timing data and some graphs of
-constructed interpolants \red{as well as a visual comparison with
-  existing monotone spline packages}.
+constructed interpolants as well as a visual comparison with
+  existing monotone spline packages.
